---
title: "00_validation_prediction"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: /

# Validation

Data analysis scripts for data analysis in the validation cohorts. 

# Support Script: Machine learning framework

Here, we will train a classifier to predict tumor metacluster from Fischer et al. (Zenodo) to assign metacluster in Fischer et al. (processed with discovery to avoid batch effects).

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SpatialExperiment)
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scran)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(caret)
library(caretEnsemble)
library(doParallel)
library(tidyverse)
library(ggridges)
library(ggsci)
library(ggVennDiagram)
library(circlize)
library(imcdatasets)
library(survival)
library(survminer)
library(edgeR)
library(wesanderson)
library(gridExtra)
library(dendextend)
library(Rphenoannoy)
```

## 1. Read in data (train)

```{r}
sce_ref <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
```

```{r}
sce_T <- readRDS("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_Jana/zenodo/analysis/SCEs/SCE_T.rds")
sce_T

# Add cell_id
sce_T$cell_id <- colnames(sce_T)
```

```{r}
sce_fischer_zenodo <- readRDS("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_Jana/zenodo/analysis/SCEs/SCE.rds")
sce_fischer_zenodo

# Add cell_id
sce_fischer_zenodo$cell_id <- colnames(sce_fischer_zenodo)
```

```{r}
sce_fischer_zenodo$tumor_cluster_100 <- sce_T$tumor_cluster_100[match(sce_fischer_zenodo$cell_id, sce_T$cell_id)]
unique(sce_fischer_zenodo$tumor_cluster_100)
```

```{r}
# Give metacluster names
sce_fischer_zenodo$tumor_metacluster_fischer <- unfactor(sce_fischer_zenodo$tumor_cluster_100)
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(16, 12, 17)] <- "HER2hi_ARpos_CKpos_ERneg_PRneg"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(20, 3, 49, 8, 13)] <- "HRpos_CKhi"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(14, 24, 28, 21)] <- "ERhi_HRpos_CKpos"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(53, 9, 35, 47)] <- "GATA3hi_HRpos_CKpos"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(10, 43, 6, 42, 33)] <- "Basal_CKs"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(18, 31, 50, 36)] <- "CD15pos"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(4, 2, 1)] <- "Ecadpos_CKpos_ARpos_pmTORpos"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(27, 46, 32)] <- "CK7pos_CKneg"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(30, 51, 26)] <- "HER2pos"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(23, 34)] <- "ER_PRlow"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(11, 22, 38, 7)] <- "Luminal_low"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(29, 19)] <- "TNBC_p53hi"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(5, 55)] <- "TNBC_EGFRhi"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(41, 37)] <- "TNBC_Ki67hi_Vimentinpos"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(44, 45)] <- "Hypoxic"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(15)] <- "TNBC_cMYChi"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(48, 40, 56)] <- "Apoptotic"
sce_fischer_zenodo$tumor_metacluster_fischer[sce_fischer_zenodo$tumor_metacluster_fischer %in% c(25, 58, 39, 52, 54, 57, 59)] <- "Epithelial_low"

length(unique(sce_fischer_zenodo$tumor_metacluster_fischer)) #18 Metacluster
```

```{r}
sce_fischer_zenodo$PID <- unfactor(sce_fischer_zenodo$PID)

sce_fischer_zenodo$grade <- str_extract(sce_fischer_zenodo$Grade, "[:digit:]")

sce_fischer_zenodo$pT_simple <- str_extract(sce_fischer_zenodo$pT_4gr, "[:digit:]")
sce_fischer_zenodo$pN_simple <- str_extract(sce_fischer_zenodo$pN_4gr, "[:digit:]")
sce_fischer_zenodo$pN_simple[sce_fischer_zenodo$pN_simple == "7"] <- NA
sce_fischer_zenodo$pM_simple <- sce_fischer_zenodo$M1_BRASI_new

sce_fischer_zenodo$status_OS <- sce_fischer_zenodo$Livestatus1dead
sce_fischer_zenodo$OS_months <- sce_fischer_zenodo$OS

sce_fischer_zenodo$status_OS <- ifelse(!is.na(sce_fischer_zenodo$OS_months) & is.na(sce_fischer_zenodo$status_OS), 0, sce_fischer_zenodo$status_OS)
```

```{r}
# Perform arcsinh-transformation 
assay(sce_fischer_zenodo, "exprs") <- asinh(assay(sce_fischer_zenodo, "raw")/1)

# Update rowData
rowData(sce_fischer_zenodo)$full_target <- rownames(sce_fischer_zenodo)
rowData(sce_fischer_zenodo)$clean_target <- c("AR", "CD45", "GATA3", 
"CD20", "CAIX", "Ecad", 
"Ki67", "S6", "HER2", "pH3", 
"Bcl2", "ER", "PR", 
"CD44", "CD8a", "H3", "DNA1", 
"DNA2", "H3K27me3", "panCK", 
"Fibronectin", "HLADR", "CK8_18", 
"CD15", "CD68", "SMA", "cMYC", "CK5", 
"KRT14", "Vimentin", "CD3", "CD11c", 
"p53", "EGFR", "CD4", "CD13_vWF", "p_mTOR", 
"CK7", "c_Cas3_PARP")
rownames(sce_fischer_zenodo) <- rowData(sce_fischer_zenodo)$clean_target

rowData(sce_fischer_zenodo)$tumor_cluster_fischer <- rowData(sce_fischer_zenodo)$clean_target %in% c("GATA3", "ER", "c_Cas3_PARP", "PR", "panCK", "CK8_18", "p_mTOR", "CK7", "p53", "cMYC", "CAIX", "Ecad", "AR", "HER2", "EGFR", "Ki67", "CK5", "KRT14", "CD15")
```

```{r}
# Filter for primary tumor samples
sce_fischer_zenodo <- sce_fischer_zenodo[,sce_fischer_zenodo$ArrayNr != "ZTMA27" & sce_fischer_zenodo$TissueType %in% c("primary tumor","tumor recurrence") & (!is.na(sce_fischer_zenodo$tumor_metacluster_fischer))]

# Option: Filter for primary tumor samples
#sce_fischer_zenodo <- sce_fischer_zenodo[,sce_fischer_zenodo$ArrayNr != "ZTMA27" & sce_fischer_zenodo$TissueType %in% c("primary tumor") & (!is.na(sce_fischer_zenodo$tumor_metacluster_fischer))]
```

```{r}
## Correct double-assigned MolSig patients == Undefined
check <- colData(sce_fischer_zenodo) %>% as.data.frame %>% select(PID, Mol_Signature) %>% unique()
PID2 <- check %>% count(PID) %>% filter(n != 1) %>% pull(PID)
check_PID2 <- check %>% filter(PID %in% PID2)
check_PID2$Mol_Signature_cor <- "Undefined"
sce_fischer_zenodo$Mol_Signature <- ifelse(!is.na(check_PID2$Mol_Signature_cor[match(sce_fischer_zenodo$PID, check_PID2$PID)]), check_PID2$Mol_Signature_cor[match(sce_fischer_zenodo$PID, check_PID2$PID)], sce_fischer_zenodo$Mol_Signature)

colData(sce_fischer_zenodo) %>% as.data.frame %>% select(PID, Mol_Signature) %>% unique() %>% group_by(Mol_Signature) %>% count() %>% ungroup() %>% mutate(sum = sum(n))
```

```{r}
saveRDS(sce_fischer_zenodo,"/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/ML_sce_fischer_zenodo_prediction.rds")
#sce_fischer_zenodo <- readRDS("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/ML_sce_fischer_zenodo_prediction.rds")
```

### Filter for tumor cells

```{r}
sce <- sce_fischer_zenodo
sce$metacluster_combined <- factor(sce$tumor_metacluster_fischer)
sce
```

## 2. Train the model

### Random forest 

#### All markers 

```{r}
set.seed(2201)

# Register parallel backend 
cl <- makePSOCKcluster(5, setup_strategy = "sequential")
registerDoParallel(cl)

#Set seeds for each iteration for parallel processing
num_folds <- 5
num_combinations <- 10  # Default is 10 for ranger
seeds <- vector(mode = "list", length = num_folds + 1)
for (i in 1:num_folds) {
    seeds[[i]] <- sample.int(5000, num_combinations)
}
seeds[[num_folds + 1]] <- sample.int(5000, 1)  # Final model seed


# Randomly split into train and test data
trainIndex <- createDataPartition(sce$metacluster_combined, p = 0.75)
train_sce <- sce[,trainIndex$Resample1]
test_sce <- sce[,-trainIndex$Resample1]

colData(train_sce) %>% as.data.frame %>% group_by(metacluster_combined) %>% count() %>% ungroup %>% mutate(fra = n/sum(n))
colData(test_sce) %>% as.data.frame %>% group_by(metacluster_combined) %>% count() %>% ungroup %>% mutate(fra = n/sum(n))

# Train data
train_data <- t(assay(train_sce, "exprs")[rowData(train_sce)$tumor_cluster_fischer,])

# Set folds for CV steps based on PID
folds <- groupKFold(train_sce$PID, k=5) 

# Specify train parameters for 5-fold cross validation
fitControl <- trainControl(method = "cv",
                           number = 5,
                           seeds = seeds, 
                           index = folds)

# Calculate class weights
class_freq <- table(train_sce$metacluster_combined)
class_weights <- 1 / class_freq
class_weights <- class_weights / sum(class_weights) 
sample_weights <- class_weights[as.character(train_sce$metacluster_combined)]

# Train a random forest model for predicting cell labels
# This call also performs parameter tuning
rffit <- train(x = train_data, 
               method = "ranger", 
               importance = "impurity",
               y = train_sce$metacluster_combined,
               metric = 'Accuracy',
               tuneLength = 5,
               trControl = fitControl,
               weights = sample_weights)

rffit
stopCluster(cl)

saveRDS(rffit, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_fischerzenondo_metacluster_pred_folds_allmarkers_withweights.rds")
```

```{r}
rffit <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_fischerzenondo_metacluster_pred_folds_allmarkers_withweights.rds")
```

### Assess feature importance

```{r}
# Classification accuracy during parameter tuning
ggplot(rffit) + 
  geom_errorbar(data = rffit$results,
                aes(ymin = Accuracy - AccuracySD,
                    ymax = Accuracy + AccuracySD),
                width = 0.4) +
    theme_classic(base_size = 15)

# Scaled feature importance (Decrease in Gini index)
plot(varImp(rffit))
```


```{r}
out_var <- varImp(rffit, scale = TRUE)

out_comb <- rbind(
  out_var$importance %>% rownames_to_column("marker") %>% mutate(category = "all_markers")
)


ggplot(out_comb, aes(x = reorder(marker, desc(Overall)), y = Overall, fill = category))+
  facet_wrap(~category, scales = "free_x")+
  geom_col()+
  theme_classic()+
  scale_fill_brewer(palette = "Paired")+
  coord_flip()+
  ylab("Scaled feature importance")+
  xlab("Markers")
```


## 3. Predict on test data

```{r}
# Select test data
mat_test <- t(assay(test_sce, "exprs")[rowData(test_sce)$tumor_cluster_fischer,])

# Predict cell phenotypes in test data
cur_pred <- predict(rffit, 
                    newdata = mat_test)
```

### Assess classifier performance

```{r}
cm <- confusionMatrix(data = cur_pred, 
                      reference = test_sce$metacluster_combined, 
                      mode = "everything")

cm

cm_stats <- as.data.frame(cm$byClass)
mean(cm_stats$`Balanced Accuracy`) # OR report this
mean(cm_stats$F1) # Report this
```


```{r}
df_out <- 
  rbind(
    data.frame(cm$byClass) %>%
      mutate(class = sub("Class: ", "", rownames(cm$byClass))) %>% mutate(category = "all_markers")
    )

ggplot(df_out) + 
  facet_wrap(~category) +
  geom_point(aes(1-Specificity, Sensitivity, 
                 size = Detection.Rate,
                 fill = class),
             shape = 21) + 
  #scale_fill_manual(values = metadata(sce_ref)$colors$metacluster_combined, limits = force) +
  theme_classic(base_size = 15) + 
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")
```

```{r}
data.frame(cm$byClass) %>%
  mutate(class = sub("Class: ", "", rownames(cm$byClass))) %>%
  ggplot() + 
  geom_point(aes(1-Specificity, Sensitivity, 
                 size = Detection.Rate,
                 fill = class),
             shape = 21) + 
  #scale_fill_manual(values = metadata(sce)$colors$metacluster_combined, limits = force) +
  theme_classic(base_size = 15) + 
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")+
  ggtitle(paste0("Accuracy: ",round(cm$overall[1],3)," [",round(cm$overall[3],3)," - ",round(cm$overall[4],3),"]"))
```


## 4. Read and predict on independent dataset 

### Fischer et al. - Processed with discovery

First, we will read in the `SingleCellExperiment` object containing the cleaned, spillover-corrected single-cell data. 
**Note:** We pre-processed the main cohort together with the validation cohort (including segmentation) to avoid strong batch effects. 

```{r read-data-batch-correction, message=FALSE}
sce_fischer <- readRDS("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/04_sce_TNBC_Clus_com.rds")
```

Filter sce for validation cohort and tumor cells

```{r filter sce, message=FALSE}
sce_fischer <- sce_fischer[,sce_fischer$analysis_cat == "tumor" & sce_fischer$TMANr %in% c("TMA21", "TMA26")]
sce_fischer

# Add rowData 
rowData(sce_fischer)$tumor_cluster_fischer <- 
  rownames(sce_fischer) %in% rownames(sce_fischer_zenodo)[rowData(sce_fischer_zenodo)$tumor_cluster_fischer]
```

Correct survival information 

```{r}
sce_fischer$status_OS <- ifelse(is.na(sce_fischer$status_OS) & !is.na(sce_fischer$OS_months), 0,sce_fischer$status_OS)
```

#### Full marker set

```{r}
set.seed(22)

# Select validation expression matrix
mat_valid <- t(assay(sce_fischer, "exprs")[rowData(sce_fischer)$tumor_cluster_fischer,])

# Predict metacluster
cell_class <- as.character(predict.train(rffit, 
                                         newdata = mat_valid, 
                                         type = "raw"))

names(cell_class) <- rownames(mat_valid)

table(cell_class)
```

Add predicted metacluster to sce and save 

```{r}
# Store labels in SCE object
sce_fischer$tumor_metacluster_fischer <- factor(cell_class[match(sce_fischer$cell_id, names(cell_class))])

cur_table <- table(sce_fischer$tumor_metacluster_fischer, sce_fischer$PID)
cur_table
```

# Save objects 

```{r}
# Save sce objects after prediction
saveRDS(sce_fischer,"/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/ML_sce_fischer_prediction.rds")
```

# Session Info 

```{r}
sessionInfo()
```
