---
title: "combined_discovery_fischer"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: sce_fischer_ALL.rds, sce_ALL.rds

# Validation

Data analysis scripts for data analysis in the validation cohorts. 

# Script: Combined analysis of Discovery and Fischer cohorts

Here, we pre-process the discovery cohort together with the Fischer et al (including improved segmentation) to avoid strong batch effects and allow better comparison. 

# Software requirements

```{r}
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scran)
library(bluster)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(VennDiagram)
library(easyalluvial)
library(scuttle)
library(mclust)
library(Rphenoannoy)
library(UpSetR)
library(cowplot)
```

# 1. Images / Masks

## Read in images/masks

```{r read-images}
#Load images from TMA21/26
images_jana <- loadImages("/mnt/tnbc_volume/TNBC_Jana/img/",on_disk = TRUE,h5FilesPath = "/mnt/tnbc_volume/TNBC_Jana/h5_img/",BPPARAM = MulticoreParam(workers = 20,progressbar = TRUE))
#set channel-names for multi-channel images
#double-check that the channel order is correct
panel <- read_csv("/mnt/tnbc_volume/TNBC_all/panel.csv")
cytomapper::channelNames(images_jana)<- panel$clean_target

images_jana

# Load remaining images 
images <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/images_segtest.rds") #Load images from previous analysis
mcols(images) <- NULL

## Combine images 
images_com <- append(images, images_jana)

# Load masks
masks <- loadImages("/mnt/tnbc_volume/TNBC_all/masks/", as.is = TRUE)
```

Add additional metadata to the `elementMetadata` slot of the `CytoImageList` (accessible via mcols)

```{r add-image-metadata}
masks <- masks[match(names(images_com), names(masks))]
all.equal(names(images_com), names(masks))

mcols(images_com) <- mcols(masks) <- DataFrame(sample_id = names(masks),
                                               PID = sce$PID[match(names(masks), sce$sample_id)])
```

## Save image and mask objects for downstream analysis

```{r save-objects-read-data}
saveRDS(images_com, "/mnt/tnbc_volume/TNBC_all/data_analysis/images.rds")
saveRDS(masks, "/mnt/tnbc_volume/TNBC_all/data_analysis/masks.rds")
```


# 2. SCE

### steinbock generated data

```{r read and inspect steinbock}
sce <- read_steinbock("/mnt/tnbc_volume/TNBC_all/",intensities_folder = "intensities",regionprops_folder = "regionprops", graphs_folder = "neighbors",return_as = "sce")
sce
```

## Single-cell processing

```{r set-colnames}
colnames(sce) <- sce$cell_id <- paste0(sce$sample_id, "_", sce$ObjectNumber)

rownames(sce) <- rowData(sce)$clean_target
```

```{r add-meta}
meta <- read.csv("/mnt/tnbc_volume/TNBC_all/data_analysis/03_data/Metadata_Combined_TNBC_LM.csv",row.names = 1)

meta_249 <- read.csv("/mnt/rcc_volume/TNBC/data_analysis/03_data/ZTMA249_TNBC_LM_processed_bysample.csv",row.names = 1) %>% select(ID_Number,PID)

unique(sce$sample_id)

## Add PID (different way of sample annotation for 174, 249 and 21/26)
PID_df <- data.frame(sample_id = unique(sce$sample_id),
           ZTMA174_ID = str_split(unique(sce$sample_id),"_",simplify = TRUE)[,3],
           ZTMA249_ID = str_split(unique(sce$sample_id),"_",simplify = TRUE)[,4], 
           ZTMA2126_ID = str_replace(paste0(str_split(unique(sce$sample_id),"_",simplify = TRUE)[,1], "_",
                                            str_split(unique(sce$sample_id),"_",simplify = TRUE)[,3]), "Z",""))

meta$TMA_ID <- paste0(meta$TMANr, "_",meta$ID_Number)
  
PID_df$ZTMA249_PID <- meta_249$PID[match(PID_df$ZTMA249_ID,meta_249$ID_Number)]
PID_df <- PID_df %>% mutate(PID = ifelse(ZTMA174_ID == "ID",ZTMA249_PID,ZTMA174_ID)) 

PID_df$PID <- ifelse(str_detect(PID_df$PID,"B"),PID_df$PID, meta$PID[match(PID_df$ZTMA2126_ID, meta$TMA_ID)])

sce$PID <- PID_df$PID[match(sce$sample_id,PID_df$sample_id)]

# Remove sample with no available PID (ZTMA249.1_ZTMA249_ID_247_CX7Y1_99)
sce <- sce[,!is.na(sce$PID)]

## Add other relevant metadata based on PID
for (entry in colnames(meta)[!colnames(meta) %in% c("PID","ID_Number","OS_data","DFS_data","TMA_ID")]) {
  colData(sce)[,entry] <- meta[,entry][match(colData(sce)[,"PID"],meta[,"PID"])]
}
```

```{r select-features}
rowData(sce)$use_channel <- !grepl("DNA|Histone", rowData(sce)$clean_target)
```

```{r color-scheme}
colors <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/color_meta.rds")
colors

metadata(sce)$colors <- colors
```

```{r transform-counts, message=FALSE, fig.width=7,fig.height=10}
set.seed(22)
cur_id <- sample(sce$sample_id,5)

#Raw counts
multi_dittoPlot(sce[,sce$sample_id %in% cur_id], vars = rownames(sce)[rowData(sce)$clean_target %in% c("panCK","Vimentin","CD45","CD3","CD68")],
               group.by = "PID", plots = c("ridgeplot"), 
               assay = "counts", 
               color.panel = metadata(sce)$colors$PID)

#Transformed counts
assay(sce, "exprs") <- asinh(counts(sce)/1)

multi_dittoPlot(sce[,sce$sample_id %in% cur_id], vars = rownames(sce)[rowData(sce)$clean_target %in% c("panCK","Vimentin","CD45","CD3","CD68")],
               group.by = "PID", plots = c("ridgeplot"), 
               assay = "exprs", 
               color.panel = metadata(sce)$colors$PID)

```

## Save single cell object for downstream analysis

```{r save-objects-read-data}
saveRDS(sce, "/mnt/tnbc_volume/TNBC_all/data_analysis/01_sce_TNBC_com.rds")
```


# 3. Spillover correction 

## Generate the spillover matrix

### Read in the data

```{r read-txts-transform, message=FALSE}
sce_spill <- readSCEfromTXT("/mnt/rcc_volume/TNBC/spillover/spillover_renamed/")

#Transformation
assay(sce_spill, "exprs") <- asinh(counts(sce_spill)/5)
```

### Quality control

```{r QC-heatmap, message = FALSE, fig.width=7, fig.height=7}
# Log10 median pixel counts per spot and channel
plotSpotHeatmap(sce_spill)

# Thresholded on 200 pixel counts
plotSpotHeatmap(sce_spill, log = FALSE, threshold = 200)
```


### Optional pixel binning {#pixel_binning}

```{r binning, message=FALSE, fig.width=7, fig.height=7}
# Define grouping
bin_size = 5

sce_spill_2 <- binAcrossPixels(sce_spill, bin_size = bin_size)

assay(sce_spill_2, "exprs") <- asinh(counts(sce_spill_2)/5)

# Log10 median pixel counts per spot and channel
plotSpotHeatmap(sce_spill_2)

# Thresholded on 200 pixel counts
plotSpotHeatmap(sce_spill_2, log = FALSE, threshold = 200)
```

### Filtering incorrectly assigned pixels

```{r debarcoding, message=FALSE}
bc_key <- as.numeric(unique(sce_spill$sample_mass))
bc_key <- bc_key[order(bc_key)]

#Preliminary metal mass for pixel
sce_spill_2 <- assignPrelim(sce_spill_2, bc_key = bc_key)

#estimate and apply cut-offs
sce_spill_2 <- estCutoffs(sce_spill_2)
sce_spill_2 <- applyCutoffs(sce_spill_2)
```

```{r assignment-heatmap, fig.width=7, fig.height=7}
cur_table <- table(sce_spill_2$bc_id, sce_spill_2$sample_mass)

pheatmap::pheatmap(log10(cur_table + 1),
         cluster_rows = FALSE,
         cluster_cols = FALSE)
```

```{r pixel-filtering}
sce_spill_2 <- filterPixels(sce_spill_2, minevents = 40, correct_pixels = TRUE)
```

### Compute spillover matrix

```{r compute-spillover, fig.width=7, fig.height=7, warning=FALSE}
sce_spill_2 <- computeSpillmat(sce_spill_2)
sm <- as.data.frame(metadata(sce_spill_2)$spillover_matrix)

isotope_list <- CATALYST::isotope_list
isotope_list$Ar <- 80

#Load Janas spillover matrix for the same panel (aquired at similiar time-points and the same panel)
sm_jana <- read.csv("/mnt/rcc_volume/TNBC/data_analysis/03_data/spillover_matrix_Jana.csv",row.names = 1)
metadata(sce_spill_2)$sm_jana <- sm_jana

#Plot both spillover matrices
plotSpillmat(sce_spill_2,sm_jana, isotope_list = isotope_list) #Janas
plotSpillmat(sce_spill_2, isotope_list = isotope_list) #My - with pixel binning

#Check that spillover matrices contain are metals that are included in the study
panel <- read.csv("/mnt/rcc_volume/TNBC/panel.csv")
channels <- paste0(panel$channel[panel$keep == 1], "Di")

channels[which(!channels %in% rownames(sm))]
channels[which(!channels %in% rownames(sm_jana))]

#Add missing Ho165 to Dy164/Er166 spillover estimates from my experiment to Janas spillover matrix (used for downstream analysis)
sm_jana <- rbind(sm_jana,sm[rownames(sm) == "Ho165Di",])
```

### Save spillover matrix as csv

```{r write_sm_csv}
#sm
#write.csv(sm, "/mnt/rcc_volume/TNBC/data_analysis/03_data/spillover_matrix_LM_mod.csv")

sm_jana
write.csv(sm, "/mnt/tnbc_volume/TNBC_all/data_analysis/03_data/spillover_matrix_Jana_mod.csv")
```

## Single-cell data compensation

```{r single-cell-compensation}
sce <- readRDS("/mnt/tnbc_volume/TNBC_all/data_analysis/01_sce_TNBC_com.rds")

rowData(sce)$channel_name <- paste0(rowData(sce)$channel, "Di")

# #sce <- compCytof(sce, sm, 
#                  transform = TRUE, cofactor = 1,
#                  isotope_list = isotope_list, 
#                  overwrite = FALSE)
sce_jana <- compCytof(sce, sm_jana, 
                 transform = TRUE, cofactor = 1,
                 isotope_list = isotope_list, 
                 overwrite = FALSE)
```

```{r visualize-single-cell-spillover, message=FALSE}
#Simple scatterplot
before <- dittoScatterPlot(sce, x.var = "H3K27me3", y.var = "Bcl2",
                           assay.x = "exprs", assay.y = "exprs") +
    ggtitle("Before compensation")

# after <- dittoScatterPlot(sce, x.var = "H3K27me3", y.var = "Bcl2",
#                           assay.x = "compexprs", assay.y = "compexprs") +
#     ggtitle("After compensation")

after_1 <- dittoScatterPlot(sce_jana, x.var = "H3K27me3", y.var = "Bcl2",
                          assay.x = "compexprs", assay.y = "compexprs") +
    ggtitle("After compensation - Jana")

before+after_1

#Smoothed density scatterplot
buylrd = c("#313695", "#4575B4", "#74ADD1", "#ABD9E9", "#E0F3F8", "#FFFFBF",
           "#FEE090", "#FDAE61", "#F46D43", "#D73027", "#A50026") 
myColRamp = colorRampPalette(c(buylrd))


smoothScatter(x=assay(sce,"exprs")[rownames(sce) == "H3K27me3"],
              y=assay(sce,"exprs")[rownames(sce) == "Bcl2"],
              colramp=myColRamp,
              main="Before compensation",
              xlab="H3K27me3 (La139)",
              ylab="Bcl2 (Gd155)")
  
smoothScatter(x=assay(sce_jana,"compexprs")[rownames(sce_jana) == "H3K27me3"],
              y=assay(sce_jana,"compexprs")[rownames(sce_jana) == "Bcl2"],
              colramp=myColRamp,
              main="After compensation - Jana",
              xlab="H3K27me3 (La139)",
              ylab="Bcl2 (Gd155)")
```

```{r overwrite-assays}
assay(sce, "counts") <- assay(sce_jana, "compcounts") 
assay(sce, "exprs") <- assay(sce_jana, "compexprs") 
assay(sce, "compcounts") <- assay(sce, "compexprs") <- NULL
```

## Save objects

```{r save, message = F}
saveRDS(sce, "/mnt/tnbc_volume/TNBC_all/data_analysis/02_sce_TNBC_spillcor_com.rds")
#saveRDS(images_comp, "/mnt/rcc_volume/TNBC/data_analysis/images_TNBC_spillcor.rds")
```

# 4. Quality Control 

## Read in the data

```{r read-data, message=FALSE}
images <- readRDS("/mnt/tnbc_volume/TNBC_all/data_analysis/images.rds")
masks <- readRDS("/mnt/tnbc_volume/TNBC_all/data_analysis/masks.rds")
sce <- readRDS("/mnt/tnbc_volume/TNBC_all/data_analysis/02_sce_TNBC_spillcor_com.rds")
```

## Image-level quality control

**HER2/ER cases exclusion** 

```{r HER2/ER-image-mean}
cur_sce <- sce

# GMM for panCK (tumor cells)
GMM_CK <- Mclust(assay(cur_sce, "exprs")["panCK",], G = 2)

cur_sce$GMM_CK <- GMM_CK$classification

cur_sce <- cur_sce[,cur_sce$GMM_CK == "2"]

image_mean <- aggregateAcrossCells(cur_sce, 
                                   ids = cur_sce$PID, 
                                   statistics="mean",
                                   use.assay.type = "counts")

## 1. HER2 
dat <- dittoBoxPlot(image_mean, var = "HER2", group.by = "HER2_case", assay = "counts", data.out = TRUE)
dat_df <- dat$data

p2 <- ggplot(dat_df, aes(x = grouping, y = var.data))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = grouping), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_brewer(palette = "Set1")+
  xlab("TMA")+
  ylab("HER2 exprs (panCK+ cells)")+
  theme_classic()

p2 

#pdf("/mnt/tnbc_volume/TNBC_all/data_analysis/01_figures/HER2exprs_TMAs.pdf",width = 8,height = 6)
#p2
#dev.off()

## remove HER2 positive cases
## kick out 4 cases with suspiciously high HER2 expression

sce <- sce[,sce$HER2_case == FALSE]
sce <- sce[,!sce$PID %in% c("B08.32458","45","167","962")]

## 2. ER
dat <- dittoBoxPlot(image_mean, var = "ER", group.by = "HER2_case", assay = "counts", data.out = TRUE)
dat_df <- dat$data

ggplot(dat_df, aes(x = grouping, y = var.data))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = grouping), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_brewer(palette = "Set1")+
  xlab("TMA")+
  ylab("ER exprs (panCK+ cells)")+
  theme_classic()

## kick out 8 cases with suspiciously high ER expression
sce <- sce[,!sce$PID %in% c("B14.59347","B07.07453","B13.30305","B08.41783","784","1501","1520","1362")]
sce
```

**Kick out incorrect images** 

```{r}
# images with aquisition between cores
sce <- sce[,!sce$sample_id == "ZTMA21_93_48_AX8Y6"]
```

**Area QC**

```{r cell-density}
#Load TMA tissue masks
tissue_area <- read_csv("/mnt/tnbc_volume/TNBC_all/tissue_mask/tissue_area_TissueRescaled.csv") 
sce$tissue_area <- as.numeric(tissue_area$AreaShape_Area[match(colData(sce)$sample_id,str_split(tissue_area$FileName_Tissue_Mask, "_Pro",simplify = TRUE)[,1])])

cell_density <- colData(sce) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    summarize(cell_area = sum(area),
              tissue_area = mean(tissue_area)) %>%
  mutate(covered_area = (cell_area / tissue_area)*100) %>%
  mutate(n_cells = as.numeric(table(colData(sce)$sample_id))) %>%
  arrange(covered_area)

ggplot(cell_density) +
  geom_point(aes(reorder(sample_id,covered_area), covered_area, color = covered_area))+ 
  ylim(c(0, 100))+
  geom_hline(yintercept = median(cell_density$covered_area), linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$covered_area)[2],linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$covered_area)[4],linetype="dashed", color = "grey")+
  theme_classic(base_size = 15)+
  theme(axis.text.x = element_blank())+
  ylab("% covered area")+ 
  xlab("")+
  scale_color_viridis()

ggplot(cell_density) +
  geom_point(aes(reorder(sample_id,n_cells), n_cells, color = n_cells))+
  geom_hline(yintercept = median(cell_density$n_cells), linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$n_cells)[2],linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$n_cells)[4],linetype="dashed", color = "grey")+
  theme_classic(base_size = 15)+
  theme(axis.text.x = element_blank())+
  ylab("Number of cells per ROI")+ 
  xlab("")+
  scale_color_viridis()

## Visualize critical images
#Kick-out criteria:
#n_cells < 300, covered_area < 8, tissue_area < 150000?
#split acquisitions should also be excluded #str_detect(cell_density$sample_id, "split")
image_id_cri <- cell_density %>% filter(n_cells < 300 | covered_area < 10 | tissue_area < 150000) %>% pull(sample_id)

# Normalize and clip images
cur_images <- images[names(images) %in% image_id_cri]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

plotPixels(cur_images,
           mask = masks[names(masks) %in% image_id_cri],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "CD3","CD68","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "CD3" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))

#keep one image
image_id_cri <- image_id_cri[image_id_cri != "ZTMA249.2_ZTMA249_ID_34_AX2Y5_188"]

#Remove outlying images and cells (n=3043 cells)
summary(colData(sce)$sample_id %in% image_id_cri)
sce <- sce[,!colData(sce)$sample_id %in% image_id_cri]
sce
```

## Cell-level quality control

**Cell size QC**
Next, we observe the distributions of cell size across the individual images. Differences in cell size distributions can indicate segmentation biases due to differences in cell density or can indicate biological differences due to cell type compositions (tumor cells tend to be larger than immune cells).
We also detect very small cells in the dataset and will remove them. The chosen threshold is arbitrary and needs to be adjusted per dataset.

```{r cellsize_dis, message=FALSE,warning=FALSE}
colData(sce) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    ggplot() +
        geom_boxplot(aes(reorder(sample_id,area), area)) +
        theme_classic(base_size = 15) + 
        theme(axis.text.x = element_blank()) +
        ylab(~ "Cell area " (mu * m^2)) + xlab("")+
  geom_hline(yintercept = median(colData(sce)$area), linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(colData(sce)$area)[2],linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(colData(sce)$area)[4],linetype="dashed", color = "grey")

#Cell area range
summary(sce$area)

#Only keep cells with an area above 5 (n=3040 cells)
sum(sce$area <= 5)
sce <- sce[,sce$area > 5]
sce
```

## Save objects

The modified `SingleCellExperiment` object is saved for further downstream analysis.

```{r save-objects-quality-control}
saveRDS(sce, "/mnt/tnbc_volume/TNBC_all/data_analysis/03_sce_TNBC_QC_com.rds")
```



# 5. Tumor / Non-Tumor

## Read in data 

```{r read-data-batch-correction, message=FALSE}
sce <- readRDS("/mnt/tnbc_volume/TNBC_all/data_analysis/03_sce_TNBC_QC_com.rds")
sce
```

## Perform "overclustering"

**NOTE**: Please be aware that some analyse including unsupervised clustering involves random processes and may not be full reproducible across systems. We did, however, set seeds for all such analysis. To ensure full reproducibility, please use the information stored in the SCE objects available on Zenodo upon publication.  

```{r}
set.seed(22)

mat <- t(assay(sce, "exprs")[rowData(sce)$use_channel==TRUE,])

#k=30
out <- Rphenoannoy(mat, k = 30)

clusters <- factor(membership(out[[2]]))

sce$cluster <- clusters

unique(sce$cluster)

#Save SCEs after clustering
saveRDS(sce, "/mnt/tnbc_volume/TNBC_all/data_analysis/04_sce_TNBC_Clus_com.rds")
```

## Use GMM for panCK

We will use a two component gaussian mixture model to distinguish between tumor (panCK+) and stromal (panCK-) cells. 

```{r GMM_CK, message=FALSE,warning=FALSE}
## Gaussian mixture model to distinguish tumor/stroma based on panCK expression
GMM_CK <- Mclust(assay(sce, "exprs")["panCK",], G = 2)

GMM_df <- data.frame(class_CK = ifelse(GMM_CK$classification == 2 & GMM_CK$uncertainty == 0, "panCK+", "panCK-"), 
                     panCK_exprs = assay(sce, "exprs")["panCK",],
                     uncertainty = GMM_CK$uncertainty)

#Visualize GMM
GMM_plot_exp <- ggplot(GMM_df, aes(x = panCK_exprs, fill = class_CK))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")
  
GMM_plot_unc <- ggplot(GMM_df, aes(x = panCK_exprs, y = uncertainty, color = class_CK))+
  geom_line()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_color_brewer(palette = "Set1")

GMM_plot_exp/GMM_plot_unc

#Add GMM CK class to SCE
sce$GMM_CK <- GMM_df$class_CK

## Gaussian mixture model to distinguish tumor/stroma based on Ecad expression
GMM_Ecad <- Mclust(assay(sce, "exprs")["Ecad",], G = 2)

GMM_df_1 <- data.frame(class_Ecad = ifelse(GMM_Ecad$classification == 2, 
                                           "Ecad+", "Ecad-"), 
                     Ecad_exprs = assay(sce, "exprs")["Ecad",],
                     uncertainty = GMM_Ecad$uncertainty)

GMM_plot_unc <- ggplot(GMM_df_1, aes(x = Ecad_exprs, y = uncertainty, color = class_Ecad))+
  geom_line()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_color_brewer(palette = "Set1")

GMM_plot_unc

#Add GMM CK class to SCE
sce$GMM_Ecad <- GMM_df_1$class_Ecad

#Save sce after clustering and GMM
saveRDS(sce, "/mnt/tnbc_volume/TNBC_all/data_analysis/04_sce_TNBC_Clus_com.rds")
```

## Check correlation of panCK and Ecad as epithelial cell markers

```{r marker correlations, message = FALSE}
#sce <- readRDS("/mnt/tnbc_volume/TNBC_all/data_analysis/04_sce_TNBC_Clus_com.rds")

GOI <- c("panCK","Ecad")
cor_sce <- as.data.frame(t(assay(sce,"exprs")[rowData(sce)$clean_target %in% GOI,]))

# For specific markers of interest (also shows the value distribution which is recommended)
pairs.panels(cor_sce,
             method = "spearman",
             hist.col = "#A6CEE3",
             digits = 2)
```

## Astir for automated cell type assignment 

see [here](https://www.sciencedirect.com/science/article/pii/S2405471221003355?via%3Dihub) for full publication (Geuenich et al.,2021, Cell Systems).

```{r astir-expo}
#Export arsinh normalized expression matrix as csv
#Add sampleID as covariate in design matrix
expr_astir <- assay(sce,"exprs") %>% t() %>% as.data.frame() %>% mutate(sample = sce$sample_id)

write.csv(expr_astir, file = "/mnt/tnbc_volume/TNBC_all/data_analysis/04_astir/expr_TNBC_astir.csv",row.names = TRUE)
```

`Next: Run Astir in python - See 00_astir_TNBC.ipynb`

```{r astir-ana, message = FALSE,warning=FALSE}
#Import Astir cell type assignments
astir_type <- read_csv("/mnt/tnbc_volume/TNBC_all/data_analysis/04_astir/cell_types_TNBC.csv",col_names = c("ID","cell_type"),skip = 1)

astir_sum <- astir_type %>% as.data.frame %>% group_by(cell_type) %>% dplyr::count()

sce$astir_type <- astir_type$cell_type

#Mark channels used for cell type assignment 
astir_channel = c("panCK","Ecad","CD3","CD8a","CD4","CD20","CD68","SMA","CD11c")

rowData(sce)$use_astir <- rowData(sce) %>% as.data.frame() %>% mutate(use_astir = ifelse(rownames(sce) %in% astir_channel,TRUE,FALSE)) %>% pull(use_astir)
```

## Visualization 

### Rich Heatmap 

Heatmap of clusters with GMM and ASTIR information 

```{r heatmap, message=FALSE, warning=FALSE}
#Read sce after clustering and GMM
#sce <- readRDS("/mnt/tnbc_volume/TNBC_all/data_analysis/04_sce_TNBC_Clus_com.rds")

### 1. Heatmap body ###

# Min-Max [0,1] normalization of arcsinh-transformed counts
assay(sce, "min_max") <- assay(sce, "exprs") - rowMins(assay(sce, "exprs"))
assay(sce, "min_max") <- assay(sce, "min_max") / (rowMaxs(assay(sce, "exprs")) - rowMins(assay(sce, "exprs")))

# Aggregrate across cells with the mean
mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce)$use_channel)

# Heatmap body color 
col_main = viridis(100)

### 2. Heatmap annotation ###

### 2.1 Spatial features

#Add number of neighbors to sce object (saved in ColPair)
n_neighbors <- colPair(sce) %>% as.data.frame() %>% group_by(from) %>% dplyr::count() %>% arrange(desc(n))
sce$n_neighbors <- n_neighbors$n[match(seq_along(colnames(sce)),n_neighbors$from)]
sce$n_neighbors <- sce$n_neighbors %>% replace_na(0)

#Select spatial features and average over clusters 
spatial <- colData(sce) %>% as.data.frame %>% select(area,eccentricity,cluster,n_neighbors)
spatial <- spatial %>% select(-cluster) %>% aggregate(by = list(cluster = spatial$cluster), FUN = mean) %>% column_to_rownames("cluster")

#Define color schemes for spatial features and add to metadata
area_col <- colorRamp2(c(min(spatial$area),median(spatial$area),max(spatial$area)), c(brewer.pal(3,"Blues")[1],brewer.pal(3,"Blues")[2],brewer.pal(3,"Blues")[3]))
metadata(sce)$colors$area <- area_col

eccentricity_col <- colorRamp2(c(min(spatial$eccentricity),median(spatial$eccentricity),max(spatial$eccentricity)), c(brewer.pal(3,"Purples")[1],brewer.pal(3,"Purples")[2],brewer.pal(3,"Purples")[3]))
metadata(sce)$colors$eccentricity <- eccentricity_col

n_neighbors_col <- colorRamp2(c(min(spatial$n_neighbors),median(spatial$n_neighbors),max(spatial$n_neighbors)), c(brewer.pal(3,"RdPu")[1],brewer.pal(3,"RdPu")[2],brewer.pal(3,"RdPu")[3]))
metadata(sce)$colors$n_neighbors <- n_neighbors_col

#Create Spatial Heatmap Annotation
ha_spatial <- HeatmapAnnotation(
    area = spatial$area,
    eccentricity = spatial$eccentricity,
    n_neighbors = spatial$n_neighbors, 
    col = list(area = metadata(sce)$colors$area,
               eccentricity = metadata(sce)$colors$eccentricity,
               n_neighbors = metadata(sce)$colors$n_neighbors),
    border = TRUE,
    which = "row")

### 2.2 Metadata features

#Number of cells per cluster 
n_cells <- colData(mean_sce) %>% as.data.frame %>% select(cluster, ncells)

#GMM - panCK
nr_tum <- colData(sce) %>% as.data.frame %>% select(cluster, GMM_CK) %>% table()
fr_tum <- as.data.frame(nr_tum) %>% pivot_wider(names_from = GMM_CK, values_from = Freq) %>% column_to_rownames("cluster")
fr_tum <- fr_tum %>% transmute(fr_tum = `panCK+`/rowSums(fr_tum)) %>% as.data.frame()

# #Astir
nr_tum_2 <- colData(sce) %>% as.data.frame %>% select(cluster, astir_type) %>% table()
fr_tum_2 <- as.data.frame(nr_tum_2) %>% pivot_wider(names_from = astir_type, values_from = Freq) %>% column_to_rownames("cluster")
fr_tum_2 <- fr_tum_2 %>% transmute(fr_tum = `Epithelial`/rowSums(fr_tum_2)) %>% as.data.frame()

#GMM - Ecad
nr_tum_3 <- colData(sce) %>% as.data.frame %>% select(cluster, GMM_Ecad) %>% table()
fr_tum_3 <- as.data.frame(nr_tum_3) %>% pivot_wider(names_from = GMM_Ecad, values_from = Freq) %>% column_to_rownames("cluster")
fr_tum_3 <- fr_tum_3 %>% transmute(fr_tum_3 = `Ecad+`/rowSums(fr_tum_3)) %>% as.data.frame()

ha_meta <- HeatmapAnnotation(GMM_CK = anno_barplot(fr_tum, width = unit(10, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            GMM_Ecad = anno_barplot(fr_tum_3, width = unit(10, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            Astir_Epi = anno_barplot(fr_tum_2, width = unit(10, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            n_cells = anno_barplot(n_cells$ncells, width = unit(10, "mm"),gp = gpar(fill = "violet", col = 1)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(1,"mm"),
                            which = "row")

### 3. Define tumor, mixed and stroma cluster (based on GMM_CK)

anno_cluscat <- fr_tum %>% 
  rownames_to_column(var = "cluster") %>% 
  mutate(cluster_category = ifelse(fr_tum >= 0.65,"Epithelial",ifelse(fr_tum < 0.65 & fr_tum > 0.2,"Mixed","Non_Epithelial")))

#Set colors for cluster_category and add to metdata
cluster_category <- setNames(c("#1F78B4", "#A6CEE3", "#B2DF8A"),sort(unique(anno_cluscat$cluster_category)))                                                                  
metadata(sce)$colors$cluster_category <- cluster_category               

#Visualize GMM_panCK density plots
ggplot(anno_cluscat, aes(fr_tum))+
  geom_density()+
  geom_rug(aes(color = cluster_category))+
  geom_vline(xintercept = 0.65, color = "red3")+
  geom_vline(xintercept = 0.2, color = "red3")+
  theme_classic()+
  theme(axis.title.x = element_blank())+
  scale_color_manual(values = cluster_category)

ha_cluscat <- HeatmapAnnotation(cluster_category = anno_cluscat$cluster_category,
                                col = list(cluster_category = metadata(sce)$colors$cluster_category),
                                border = TRUE,
                                which = "row")



### 4. Plot combined Heatmap
h <- Heatmap(t(assay(mean_sce, "min_max")),
        col = col_main,
        name= "min-max",
        km = 1,
        show_row_names = T, 
        show_column_names =  T, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")+
  ha_spatial+
  ha_meta+
  ha_cluscat

draw(h)

### Add cluster_categories to sce
sce$cluster_category <- anno_cluscat$cluster_category[match(sce$cluster,anno_cluscat $cluster)]

#Save sce
saveRDS(sce, "/mnt/tnbc_volume/TNBC_all/data_analysis/04_sce_TNBC_Clus_com.rds")
```

## Re-analyse mixed clusters based on GMM

```{r clus_mixed, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
sce_mixed <- sce[,colData(sce)$cluster_category == "Mixed"] #n=60973

#Select cells that are either panCK or Ecad positive based on gaussian mixture model
sce_mixed$cluster_category_2 <- ifelse(sce_mixed$GMM_CK == "panCK+" | sce_mixed$GMM_Ecad == "Ecad+" ,"Epithelial","Non_Epithelial")

sce_mixed$cluster_category_2 %>% table()

spatial_mixed <- colData(sce_mixed) %>% as.data.frame %>% select(area,eccentricity,cluster_category_2,n_neighbors)
spatial_mixed <- spatial_mixed %>% select(-cluster_category_2) %>% aggregate(by = list(cluster_category_2 = spatial_mixed$cluster_category_2), FUN = mean) %>% column_to_rownames("cluster_category_2")

sce_mixed$cluster_category_2_mean_area <- as.character(round(spatial_mixed$area[match(sce_mixed$cluster_category_2, rownames(spatial_mixed))],2))

#Single-cell Heatmap 
dittoHeatmap(sce_mixed, genes = rownames(sce_mixed)[rowData(sce_mixed)$use_channel],
             assay = "exprs", cluster_cols = FALSE, order.by = c("cluster_category_2","GMM_CK","GMM_Ecad","cluster"), scale = "none",
             heatmap.colors = viridis(100), annot.by = c("cluster","GMM_CK","GMM_Ecad","cluster_category_2","cluster_category_2_mean_area"))


# Add cluster_categories_2 to main sce
sce$cluster_category_2 <- sce_mixed$cluster_category_2[match(colnames(sce),colnames(sce_mixed))]
```

## Define categories for downstream analysis
Use all epithelial clusters and mixed clusters with panCK+/Ecad+ GMM for downstream tumor analysis.

```{r ana_cat, message = FALSE, warning=FALSE}
sce$analysis_cat <- colData(sce) %>% as.data.frame %>% 
  select(cluster_category, cluster_category_2) %>% 
  mutate(analysis_cat = ifelse(sce$cluster_category == "Epithelial" |(sce$cluster_category == "Mixed" & sce$cluster_category_2 == "Epithelial"),"tumor","non_tumor")) %>% 
  pull(analysis_cat)

#color scheme for analysis_cat
analysis_cat <- setNames(c("#E41A1C", "#377EB8"),sort(unique(sce$analysis_cat)))                                                                  
metadata(sce)$colors$analysis_cat <- analysis_cat   

colData(sce) %>% as.data.frame %>% select(analysis_cat) %>% group_by(analysis_cat) %>% dplyr::count()
#627677 tumor cells
#525537 non-tumor cells

#Visualize in alluvial plot 
colData(sce) %>% as.data.frame %>% select(cluster_category, analysis_cat)

alluvial_wide(data = colData(sce) %>% as.data.frame %>% select(cluster_category, analysis_cat), 
              max_variables = 4,
              fill_by = 'all_flows',
              order_levels = c("Non_Epithelial","Epithelial","Mixed","tumor","non_tumor"))+
  theme_classic(base_size = 12, base_line_size = 0.5)


#Intersection plot
## Upset plot
#1. tumor   
tumor = list(
analysis_cat_tumor = colData(sce) %>% as.data.frame %>% filter(analysis_cat == "tumor") %>% rownames_to_column() %>% pull("rowname"),
panCK_positive_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_CK == "panCK+") %>% rownames_to_column() %>% pull("rowname"),
Ecad_positive_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_Ecad == "Ecad+") %>% rownames_to_column() %>% pull("rowname")
)

upset <- upset(fromList(tumor), order.by = "freq")
upset

non_tumor = list(
analysis_cat_nontumor = colData(sce) %>% as.data.frame %>% filter(analysis_cat == "non_tumor") %>% rownames_to_column() %>% pull("rowname"),
panCK_negative_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_CK == "panCK-") %>% rownames_to_column() %>% pull("rowname"),
Ecad_negative_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_Ecad == "Ecad-") %>% rownames_to_column() %>% pull("rowname")
)

upset_1 <- upset(fromList(non_tumor), order.by = "freq")
upset_1
```

## Visualization of results on UMAP/TSNE

```{r viz_sub, message = FALSE, warning=FALSE, fig.width=20, fig.height=20}
## Subsample 10% of cells from each core 
#to save time and not get super crowded dimensionality reduction plots
set.seed(22)
sub <- colData(sce) %>% as.data.frame() %>% group_by("sample_id") %>% slice_sample(prop = 0.1) %>% arrange("sample_id") %>% mutate(sub_id = paste(sample_id,ObjectNumber,sep="_")) %>% pull(sub_id)
sce_sub = sce[,sub]
sce_sub$cluster = as.factor(sce_sub$cluster)  

#Run UMAP/TSNE
sce_sub = runUMAP(sce_sub[rowData(sce_sub)$use_channel,], n_neighbors = 30, pca = 50,exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam((detectCores()-2)))
sce_sub = runTSNE(sce_sub[rowData(sce_sub)$use_channel,], exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam((detectCores()-2)))

## 1. Visualize UMAP
#panCK and GMM_CK
p1 <- plotReducedDim(sce_sub, dimred = "UMAP", by_exprs_values = "exprs",colour_by = "panCK")

p2 <- plotReducedDim(sce_sub, dimred = "UMAP", by_exprs_values = "exprs",colour_by = "Ecad")

p3 <- plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "cluster_category")+
  scale_color_manual(values = metadata(sce)$colors$cluster_category)

p4 <- plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$analysis_cat)

(p1+p2)/(p3+p4)

#Visualize all marker expression
all_plots_UMAP <- lapply(c(rownames(sce_sub)[rowData(sce_sub)$use_channel]),
                    function(x){
                      p <- plotReducedDim(sce_sub, dimred = "UMAP", 
                                          colour_by = x, 
                                          by_exprs_values = "exprs",
                                          point_size = 0.5)
                      return(p)                    
                    })

plots_UMAP_exp <- plot_grid(plotlist = all_plots_UMAP)
plots_UMAP_exp

## 2. Visualize TSNE
#panCK and GMM_CK
p5 <- plotReducedDim(sce_sub, dimred = "TSNE", by_exprs_values = "exprs",colour_by = "panCK")

p6 <- plotReducedDim(sce_sub, dimred = "TSNE", by_exprs_values = "exprs",colour_by = "Ecad")

p7 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "cluster_category")+
  scale_color_manual(values = metadata(sce)$colors$cluster_category)

p8 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$analysis_cat)

(p5+p6)/(p7+p8)

#Visualize all marker expression
all_plots_TSNE <- lapply(c(rownames(sce_sub)[rowData(sce_sub)$use_channel]),
                    function(x){
                      p <- plotReducedDim(sce_sub, dimred = "TSNE", 
                                          colour_by = x, 
                                          by_exprs_values = "exprs",
                                          point_size = 0.5)
                      return(p)                    
                    })

plots_TSNE_exp <- plot_grid(plotlist = all_plots_TSNE)
plots_TSNE_exp

#Visualize TMANr and PID 
p9 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "TMANr")+
  scale_color_manual(values = metadata(sce)$colors$TMANr)

p10 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "PID")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  guides(color = "none")
p9+p10
```

## Save objects

```{r save_objects, message = FALSE, warning=FALSE}
#Save sce
saveRDS(sce, "/mnt/tnbc_volume/TNBC_all/data_analysis/04_sce_TNBC_Clus_com.rds")
#Save sce_sub
saveRDS(sce_sub, "/mnt/tnbc_volume/TNBC_all/data_analysis/04_sce_SUB_TNBC_Clus_com.rds")
```

# Session Info 

```{r}
sessionInfo()
```
