---
title: null
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

# Figure 6 and Figure S9/10

Data analysis scripts for re-creating Figure 6 and Figure S9/10. 

# Software requirements

```{r, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(tidyverse)
library(ComplexHeatmap)
library(scuttle)
library(RColorBrewer)
library(circlize)
library(conflicted)
library(scater)
library(survival)
library(survminer)
library(viridis)
library(foreach)
library(bluster)
library(patchwork)
library(ggpubr)
library(glmnet)
library(doMC)
library(imcRtools)
library(edgeR)
library(ggnewscale)
library(epitools)
library(spicyR)
```

```{r, message=FALSE,warning=FALSE}
conflicts_prefer(dplyr::desc)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::count)
conflicts_prefer(dplyr::select)
conflicts_prefer(stats::dist)
conflicts_prefer(stats::cor)
```

# Read data

```{r}
# Available from Zenodo - DOI: 10.5281/zenodo.15396067
# File paths have to be adjusted by the user

## SCE objects
sce_TME <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_TME.rds")
sce_ALL <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_ALL.rds")
sce_ALL_sub <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_ALL_sub.rds")
sce_wang <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_wang.rds")

## Other
PDL1_IHC <- read_csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/PDL1_IHC.csv")
spicyR_spatialgroup <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/spicyR_spatialgroup.rds")
```

# Figure 6

## 6A

```{r, fig.width=12, fig.height=7}
### 1. Heatmap body ###
cluster_mean_sce <- aggregateAcrossCells(sce_TME, ids = sce_TME$cluster_130,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_TME)$cluster_channel)

# Heatmap body colors 
col_exprs <- colorRamp2(c(0,1,2,3), c("#440154FF", "#31688EFF", "#35B779FF", "#FDE725FF"))

### 2. Heatmap annotation ###

### 2.1 Spatial features

#Select spatial features and average over clusters 
spatial <- colData(sce_TME) %>% as.data.frame %>% select(cluster_130, area, eccentricity, n_neighbors)
spatial <- spatial %>% select(-cluster_130) %>% aggregate(by = list(cluster_130 = spatial$cluster_130), FUN = mean) %>% column_to_rownames("cluster_130")

#Create Spatial Heatmap Annotation
ha_spatial <- HeatmapAnnotation(
    area = spatial$area,
    eccentricity = spatial$eccentricity,
    n_neighbors = spatial$n_neighbors, 
    col = list(area = metadata(sce_TME)$colors$area,
               eccentricity = metadata(sce_TME)$colors$eccentricity,
               n_neighbors = metadata(sce_TME)$colors$n_neighbors),
    border = TRUE,
    which = "row")


### 2.2 Immune Metacluster 
anno <- colData(cluster_mean_sce) %>% as.data.frame %>% select(cluster_130, immune_metacluster, ncells)

### 2.3 Astir immune cell types 

## Create ASTIR annotation
# Astir cell types per cluster
astir_anno <- colData(sce_TME) %>% as.data.frame() %>% select(cluster_130,astir_TME) %>% group_by(cluster_130) %>% table() %>% as.data.frame
astir_anno <- astir_anno %>% group_by(cluster_130) %>% mutate(fra = Freq/sum(Freq)) 
astir_anno <- astir_anno %>% select(-Freq) %>% pivot_wider(id_cols = cluster_130,names_from = astir_TME,values_from = fra) %>% column_to_rownames("cluster_130")

ha_astir <- HeatmapAnnotation(astir_TME = anno_barplot(astir_anno, width = unit(20, "mm"), gp = gpar(fill = metadata(sce_TME)$colors$astir_col)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(3,"mm"),
                            which = "row")

###2.3 Metadata features

#Number of cells per cluster 
n_cells <- colData(cluster_mean_sce) %>% as.data.frame %>% select(cluster_130, ncells)

# Proportion of grade per cluster
grade <- colData(sce_TME) %>% as.data.frame() %>% select(cluster_130,grade) %>% group_by(cluster_130) %>% table() %>% as.data.frame
grade <- grade %>% group_by(cluster_130) %>% mutate(fra = Freq/sum(Freq)) 
grade <- grade %>% select(-Freq) %>% pivot_wider(id_cols = cluster_130,names_from = grade,values_from = fra) %>% column_to_rownames("cluster_130")

# Proportion of pT stage per cluster
pT <- colData(sce_TME) %>% as.data.frame() %>% select(cluster_130,pT_simple) %>% group_by(cluster_130) %>% table() %>% as.data.frame
pT <- pT %>% group_by(cluster_130) %>% mutate(fra = Freq/sum(Freq)) 
pT <- pT %>% select(-Freq) %>% pivot_wider(id_cols = cluster_130,names_from = pT_simple,values_from = fra) %>% column_to_rownames("cluster_130")

# Proportion of pN stage per cluster
pN <- colData(sce_TME) %>% as.data.frame() %>% select(cluster_130,pN_simple) %>% group_by(cluster_130) %>% table() %>% as.data.frame
pN <- pN %>% group_by(cluster_130) %>% mutate(fra = Freq/sum(Freq)) 
pN <- pN %>% select(-Freq) %>% pivot_wider(id_cols = cluster_130,names_from = pN_simple,values_from = fra) %>% column_to_rownames("cluster_130")

# Number of contributing patients
cluster_PID <- colData(sce_TME) %>% as.data.frame() %>% select(cluster_130,PID) %>% group_by(cluster_130) %>% table() %>% as.data.frame
n_PID <- cluster_PID %>% filter(Freq>0) %>% group_by(cluster_130) %>% count(name = "n_PID") %>% column_to_rownames("cluster_130")

# Metadata annotation
ha_meta <- HeatmapAnnotation(cluster_130 = anno_simple(unfactor(anno$cluster_130), width = unit(8, "mm"), pch = unfactor(anno$cluster_130),border=TRUE,gp = gpar(fill = "white")),
                            metacluster = anno_simple(unfactor(anno$immune_metacluster),border=TRUE,col=metadata(sce_TME)$colors$immune_metacluster),
                            n_cells = anno_barplot(n_cells$ncells, width = unit(10, "mm"),gp = gpar(fill = "#440154FF")),
                            n_PID = anno_barplot(n_PID, width = unit(10, "mm"),gp = gpar(fill = "#440154FF")),
                            grade = anno_barplot(grade, width = unit(12, "mm"),gp = gpar(fill = metadata(sce_TME)$colors$grade)),
                            pT = anno_barplot(pT, width = unit(12, "mm"),gp = gpar(fill = metadata(sce_TME)$colors$pT_simple)),
                            pN = anno_barplot(pN, width = unit(12, "mm"),gp = gpar(fill = metadata(sce_TME)$colors$pN_simple)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(1,"mm"),
                            which = "row")

### 3. Plot heatmap ###
h <- Heatmap(t(assay(cluster_mean_sce, "exprs")),
        column_title = "cluster_channels",
        col = col_exprs,
        name = "mean exprs",
        row_title = NULL, 
        show_column_names =  TRUE, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2",
        row_split = anno$immune_metacluster,
        cluster_row_slices = FALSE)+
  ha_astir+
  ha_meta+
  ha_spatial

lgd_astir <- Legend(title = "Astir_TME", at = names(metadata(sce_TME)$colors$astir_col), 
              legend_gp = gpar(fill = metadata(sce_TME)$colors$astir_col))

lgd_meta <- Legend(title = "Immune_Metacluster", at = names(metadata(sce_TME)$colors$immune_metacluster), 
              legend_gp = gpar(fill = metadata(sce_TME)$colors$immune_metacluster))          

draw(h,annotation_legend_list = list(lgd_astir, lgd_meta))
```

## 6B

```{r}
tumor_color <- c("grey")
names(tumor_color) <- "tumor"
TME_color <- c("grey")
names(TME_color) <- "TME"

color_TME <- c(metadata(sce_ALL_sub)$colors$metacluster_combined, tumor_color, TME_color)

plotReducedDim(sce_ALL_sub, dimred = "TSNE", colour_by = "metacluster_tumor_combined", point_size = 1)+
  scale_color_manual(values = color_TME, limits = force)
```

## 6D

```{r}
# Kaplan-Meier
## Spatial groups
cur_surv <- colData(sce_ALL) %>% as.data.frame() %>% select(PID, spatial_CD8_pheno, DFS_months, status_DFS) %>% unique()

fit <- survfit(Surv(DFS_months, status_DFS) ~ spatial_CD8_pheno, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c("steelblue1","navy","firebrick"))
```

## 6E

```{r}
sce_ALL_tumor <- sce_ALL[,sce_ALL$analysis_cat == "tumor"]

# Aggregate expression per PID 
spatialpheno_mean_PID <- aggregateAcrossCells(sce_ALL_tumor, ids = sce_ALL_tumor$PID,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_ALL_tumor)$tumor_cluster_channel)

# Test for differential expression per category 
test <- t(assay(spatialpheno_mean_PID, "exprs")[rowData(spatialpheno_mean_PID)$tumor_cluster_channel,]) %>% as.data.frame %>% mutate(spatial_CD8_pheno = spatialpheno_mean_PID$spatial_CD8_pheno) 
test_result <- compare_means(c(H3K27me3, CK5, HLADR, CK8_18, CD15, KRT14, Vimentin, 
cMYC, HER2, pH3, ER, PR, p53, AR, GATA3, CAIX, 
Ki67, EGFR, S6, p_mTOR, CK7, c_Cas3_PARP)~spatial_CD8_pheno, method = "kruskal.test",data = test)
test_result$p.adj <- p.adjust(test_result$p, method = "BH") 
test_result$psignif.adj <- as.matrix(symnum(test_result$p.adj,
       cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), 
       symbols = c("****", "***", "**", "*", "ns")))[,1]
rownames(test_result) <- test_result$.y.
sig.poteins <- unfactor(test_result$.y.[test_result$p.adj < 0.05])


# Aggregate expression per spatial phenotype
spatialpheno_mean_sce_ALL <- aggregateAcrossCells(sce_ALL_tumor, ids = sce_ALL_tumor$spatial_CD8_pheno,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_ALL_tumor)$tumor_cluster_channel)

mat_scaled <- t(scale(t(assay(spatialpheno_mean_sce_ALL, "exprs"))))
highest_cat <- mat_scaled %>% as.data.frame() %>% rownames_to_column("gene") %>% pivot_longer(cols = 2:4, names_to = "spatial_CD8", values_to = "exprs") %>% group_by(gene) %>% mutate(max = max(exprs)) %>% filter(exprs == max) %>% select(gene, spatial_CD8) %>% as.data.frame() %>% column_to_rownames("gene")

## Visualize 
h <- Heatmap(assay(spatialpheno_mean_sce_ALL, "exprs"), col = viridis(100), name = "exprs", cluster_columns = FALSE, row_names_side = "left", show_row_dend = FALSE)+
  rowAnnotation(cat = anno_simple(highest_cat, col = metadata(sce_ALL)$colors$spatial_CD8_pheno, width = unit(10, "mm")))+
  rowAnnotation(test = anno_text(test_result$psignif.adj))

lgd_cat <- Legend(title = "Highest_exprs", at = names(metadata(sce_ALL)$colors$spatial_CD8_pheno), 
              legend_gp = gpar(fill = metadata(sce_ALL)$colors$spatial_CD8_pheno))

draw(h,annotation_legend_list = list(lgd_cat))
```

## 6F

```{r}
## Number of B patches per PID 
n_Bpatch_PID <- colData(sce_ALL) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  unique() %>% 
  group_by(spatial_CD8_pheno_merg, PID) %>% 
  count()
plot_df <- colData(sce_ALL) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg) %>% 
  unique() 
rownames(plot_df) <- NULL
plot_df$n_Bpatch <- n_Bpatch_PID$n[match(plot_df$PID, n_Bpatch_PID$PID)]
plot_df$n_Bpatch[is.na(plot_df$n_Bpatch)] <- 0
plot_df$spatial_CD8_pheno <- sce_ALL$spatial_CD8_pheno[match(plot_df$PID, sce_ALL$PID)]

## Size of B cell patches 
size_Bpatch_PID <- colData(sce_ALL) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  group_by(patch_B_cell, PID) %>% 
  count()
size_Bpatch_PID$spatial_CD8_pheno_merg <- sce_ALL$spatial_CD8_pheno_merg[match(size_Bpatch_PID$PID, sce_ALL$PID)]
size_Bpatch_PID$spatial_CD8_pheno <- sce_ALL$spatial_CD8_pheno[match(size_Bpatch_PID$PID, sce_ALL$PID)]

# Define cut-off for small and large B cell follicles - use 30 B cells per patch
size_Bpatch_PID <- size_Bpatch_PID %>% mutate(patch_score = ifelse(n > 30, "large", "small"))
PID_large_patches <- size_Bpatch_PID %>% filter(patch_score == "large") %>% pull(PID) %>% unique()
size_Bpatch_PID <- size_Bpatch_PID %>% mutate(large_Bpatch = ifelse(PID %in% PID_large_patches, TRUE, FALSE))

## Create B cell score per PID 
B_cells_PID <- colData(sce_ALL) %>% as.data.frame %>% select(PID, metacluster_combined) %>% filter(metacluster_combined == "B_cell") %>% group_by(PID, metacluster_combined) %>% count()
plot_df$n_Bcells <- B_cells_PID$n[match(plot_df$PID, B_cells_PID$PID)]
plot_df$n_Bcells[is.na(plot_df$n_Bcells)] <- 0
plot_df$large_Bpatch <- size_Bpatch_PID$large_Bpatch[match(plot_df$PID, size_Bpatch_PID$PID)]

plot_df <- plot_df %>% mutate(B_cell_score = ifelse(n_Bcells == 0, "No_B_cell", 
                                         ifelse(n_Bpatch == 0, "No_B_patch", 
                                                ifelse(large_Bpatch == FALSE, "Small_B_patch", "Large_B_patch"))))

plot_df$B_cell_score <- factor(plot_df$B_cell_score, levels = c("No_B_cell","No_B_patch", "Small_B_patch", "Large_B_patch"))
plot <- plot_df %>%
  group_by(B_cell_score, spatial_CD8_pheno) %>%
  summarise(n=n()) %>%
  group_by(spatial_CD8_pheno) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup()

plot_test <- plot %>% pivot_wider(names_from = spatial_CD8_pheno, id_cols = B_cell_score, values_from = n) %>% column_to_rownames("B_cell_score")
plot_test_out <- fisher.test(plot_test, simulate.p.value = TRUE)

ggplot(plot) + 
  geom_col(aes(y=spatial_CD8_pheno, x=fraction, fill=B_cell_score), color = "black") +
  theme_classic() + 
  xlab("Fraction of PID") +
  ylab("")+
  scale_fill_brewer(palette = "Paired")+
  coord_flip()+
  ggtitle(paste0("p (Fisher) = ", plot_test_out$p.value))
```

## 6G

Pixel and single-cell level IMC images were generated using [cytoviewer](https://github.com/BodenmillerGroup/cytoviewer).

## 6H

```{r}
out_PDL1 <- PDL1_IHC %>% group_by(CPS_results, spatial_CD8_pheno) %>% dplyr::count() %>% group_by(spatial_CD8_pheno) %>% mutate(fra = n/sum(n))

fisher_out <- fisher.test(xtabs(n ~ spatial_CD8_pheno + CPS_results, data = out_PDL1))
fisher_out

ggplot(out_PDL1, aes(x = spatial_CD8_pheno, y = fra, fill = CPS_results))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  ggtitle(paste0("p (Fisher) = ", round(fisher_out$p.value, 4)))
```

## 6I

```{r}
## Immunotherapy response comparison
out_check <- colData(sce_wang) %>% as.data.frame %>% dplyr::select(PID_phase, BiopsyPhase, spatial_CD8_pheno_merg, spatial_CD8_pheno, Arm, pCR) %>% unique() 

out_C <- out_check %>% filter(BiopsyPhase == "Baseline") %>% group_by(spatial_CD8_pheno, Arm, BiopsyPhase, pCR) %>% dplyr::count() %>% group_by(spatial_CD8_pheno, BiopsyPhase, Arm) %>% mutate(fra = n/sum(n))

out_C$pCR <- factor(out_C$pCR, levels = c("RD","pCR"))

ggplot(out_C, aes(x = spatial_CD8_pheno, y = fra, fill = pCR))+
  facet_wrap(~Arm + BiopsyPhase)+
  geom_bar(stat = "identity", color = "black")+
  geom_hline(yintercept = 0.5, linetype = "dashed")+
  theme_classic()+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  coord_flip()

fisher_out <- chisq.test(xtabs(n ~ spatial_CD8_pheno + pCR, data = out_C %>% filter(Arm == "C&I")))
fisher_out

fisher_out_2 <- chisq.test(xtabs(n ~ spatial_CD8_pheno + pCR, data = out_C %>% filter(Arm == "C")))
fisher_out_2
```

## 6J

```{r}
## Analyse tumor compartment expression 
sce_tumor <- sce_wang[,sce_wang$isEpithelial == "TRUE" & sce_wang$BiopsyPhase == "Baseline"]

# Aggregate expression per PID 
spatialpheno_mean_PID <- aggregateAcrossCells(sce_tumor, ids = sce_tumor$PID,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_tumor)$tumor_markers_1)

# Test for differential expression per category 
test <- t(assay(spatialpheno_mean_PID, "exprs")[rowData(spatialpheno_mean_PID)$tumor_markers_1,]) %>% as.data.frame %>% mutate(spatial_CD8_pheno = spatialpheno_mean_PID$spatial_CD8_pheno) 
test_result <- compare_means(c(PDL1_SP, CD56, Helios, GATA3, IDO, AR, PDL1_73, Ki67, CK5_14, HLA_ABC, cPARP, CK8_18, CD15, HLA_DR, pH2AX, CA9)~spatial_CD8_pheno, method = "kruskal.test",data = test)
test_result$p.adj <- p.adjust(test_result$p, method = "BH") 
test_result$psignif.adj <- as.matrix(symnum(test_result$p.adj,
       cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), 
       symbols = c("****", "***", "**", "*", "ns")))[,1]

rownames(test_result) <- test_result$.y.
sig.proteins <- unfactor(test_result$.y.[test_result$p.adj < 0.05])

# Aggregate expression per spatial phenotype
spatialpheno_mean_sce <- aggregateAcrossCells(sce_tumor, ids = sce_tumor$spatial_CD8_pheno,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_tumor)$tumor_markers_1)

mat_scaled<- t(scale(t(assay(spatialpheno_mean_sce, "exprs"))))
highest_cat <- mat_scaled %>% as.data.frame() %>% rownames_to_column("gene") %>% pivot_longer(cols = 2:4, names_to = "spatial_CD8", values_to = "exprs") %>% group_by(gene) %>% mutate(max = max(exprs)) %>% filter(exprs == max) %>% select(gene, spatial_CD8) %>% as.data.frame() %>% column_to_rownames("gene")

## Visualize 
h <- Heatmap(assay(spatialpheno_mean_sce, "exprs"), col = viridis(100), name = "exprs", cluster_columns = FALSE, row_names_side = "left", show_row_dend = FALSE)+
  rowAnnotation(cat = anno_simple(highest_cat, col = metadata(sce_wang)$colors$spatial_CD8_pheno, width = unit(10, "mm")))+
  rowAnnotation(test = anno_text(test_result$psignif.adj))

lgd_cat <- Legend(title = "Highest_exprs", at = names(metadata(sce_wang)$colors$spatial_CD8_pheno), 
              legend_gp = gpar(fill = metadata(sce_wang)$colors$spatial_CD8_pheno))

draw(h,annotation_legend_list = list(lgd_cat))
```

## 6K

```{r}
sce_wang_sub <- sce_wang[,sce_wang$BiopsyPhase == "Baseline"]

## 1. Number of patches per PID 
n_Bpatch_PID <- colData(sce_wang_sub) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  unique() %>% 
  group_by(spatial_CD8_pheno_merg, PID) %>% 
  dplyr::count()

plot_df <- colData(sce_wang_sub) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg) %>% 
  unique() 
rownames(plot_df) <- NULL
plot_df$n_Bpatch <- n_Bpatch_PID$n[match(plot_df$PID, n_Bpatch_PID$PID)]
plot_df$n_Bpatch[is.na(plot_df$n_Bpatch)] <- 0
plot_df$spatial_CD8_pheno <- sce_wang_sub$spatial_CD8_pheno[match(plot_df$PID, sce_wang_sub$PID)]
  
## 2. Size of B cell patches 
size_Bpatch_PID <- colData(sce_wang_sub) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  group_by(patch_B_cell, PID) %>% 
  dplyr::count()

size_Bpatch_PID$spatial_CD8_pheno_merg <- plot_df$spatial_CD8_pheno_merg[match(size_Bpatch_PID$PID, plot_df$PID)]
size_Bpatch_PID$spatial_CD8_pheno <- sce_wang_sub$spatial_CD8_pheno[match(size_Bpatch_PID$PID, sce_wang_sub$PID)]

# Define cut-off for small and large B cell follicles - use 30 B cells per patch (same cutoff as discovery)
size_Bpatch_PID <- size_Bpatch_PID %>% mutate(patch_score = ifelse(n > 30, "large", "small"))
PID_large_patches <- size_Bpatch_PID %>% filter(patch_score == "large") %>% pull(PID) %>% unique()
size_Bpatch_PID <- size_Bpatch_PID %>% mutate(large_Bpatch = ifelse(PID %in% PID_large_patches, TRUE, FALSE))

## 3. Create B cell score per PID 
B_cells_PID <- colData(sce_wang_sub) %>% as.data.frame %>% select(PID, celltype_broad) %>% filter(celltype_broad == "B_cell") %>% group_by(PID, celltype_broad) %>% dplyr::count()

plot_df$n_Bcells <- B_cells_PID$n[match(plot_df$PID, B_cells_PID$PID)]
plot_df$n_Bcells[is.na(plot_df$n_Bcells)] <- 0
plot_df$large_Bpatch <- size_Bpatch_PID$large_Bpatch[match(plot_df$PID, size_Bpatch_PID$PID)]
plot_df <- plot_df %>% mutate(B_cell_score = ifelse(n_Bcells == 0, "No_B_cell", 
                                         ifelse(n_Bpatch == 0, "No_B_patch", 
                                                ifelse(large_Bpatch == FALSE, "Small_B_patch", "Large_B_patch"))))
plot_df$B_cell_score <- factor(plot_df$B_cell_score, levels = c("No_B_cell","No_B_patch", "Small_B_patch", "Large_B_patch"))
plot <- plot_df %>%
  group_by(B_cell_score, spatial_CD8_pheno) %>%
  dplyr::summarise(n=n()) %>%
  group_by(spatial_CD8_pheno) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup()
plot_test <- plot %>% pivot_wider(names_from = spatial_CD8_pheno, id_cols = B_cell_score, values_from = n) %>% column_to_rownames("B_cell_score")
plot_test[is.na(plot_test)] <- 0
plot_test_out <- fisher.test(plot_test, simulate.p.value = TRUE)

ggplot(plot) + 
  geom_col(aes(y=spatial_CD8_pheno, x=fraction, fill=B_cell_score), color = "black") +
  theme_classic() + 
  xlab("Fraction of PID") +
  ylab("")+
  scale_fill_brewer(palette = "Paired")+
  coord_flip()+
  ggtitle(paste0("p (Fisher) = ", plot_test_out$p.value))
```

# Figure S9

## S9A

```{r}
clusters <- colnames(colData(sce_TME))[str_detect(colnames(colData(sce_TME)),"cluster_[:digit:]")]
df <- foreach(i = seq_along(clusters),.combine = rbind) %do% {
num_clus <- length(unique(colData(sce_TME)[,clusters[i]]))
sil <- approxSilhouette(t(assay(sce_TME, "exprs")), clusters=colData(sce_TME)[,clusters[i]])
rmsd <- clusterRMSD(t(assay(sce_TME, "exprs")), clusters=colData(sce_TME)[,clusters[i]])
data.frame(num_clus = num_clus,
           k = as.numeric(str_split(clusters[i],"_",simplify = TRUE)[,2]), 
           sil = mean(sil$width), 
           rmsd = sum(rmsd))
}


p1 <- ggplot(df, aes(x=k, y=num_clus))+
  geom_point(aes(k, num_clus, col = num_clus))+
  geom_line(aes(k, num_clus,col=num_clus))+
  theme_classic()+
  scale_color_continuous(type = "viridis")

p2 <- ggplot(df, aes(x=k, y=rmsd))+
  geom_point(aes(k, rmsd, col = rmsd))+
  geom_line(aes(k, rmsd,col=rmsd))+
  theme_classic()+
  scale_color_continuous(type = "viridis")

p3 <- ggplot(df, aes(x=k, y=sil))+
  geom_point(aes(k, sil, col = sil))+
  geom_line(aes(k, sil,col=sil))+
  theme_classic()+
  scale_color_continuous(type = "viridis")

p1+p2+p3
```

## S9B

```{r, fig.height=15}
## PID immune_metacluster fraction dataframe
PID_immune_metacluster <- colData(sce_TME) %>% as.data.frame() %>% select(immune_metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_immune_metacluster <- PID_immune_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
cur_df <- PID_immune_metacluster %>% pivot_wider(id_cols = "PID", names_from = "immune_metacluster",values_from = "fra") %>% column_to_rownames("PID")

# Perform hierarchical clustering on fractions
dd <- dist(cur_df, method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
row_sorted <- hc$order

# Define annotation dataframe
cur_df <- cur_df[row_sorted,]
anno <- colData(sce_TME) %>% as.data.frame %>% select(PID, age_bin, grade, pT_simple, pN_simple, pM_simple, pretreatment, shannon_div) %>% unique()
anno <- anno[match(rownames(cur_df), anno$PID),]

# Plot annotation dataframe with ComplexHeatmap
fraction_anno <- HeatmapAnnotation(MC_fractions = anno_barplot(cur_df, gp=gpar(fill=metadata(sce_TME)$colors$immune_metacluster[colnames(cur_df)]),
                                   bar_width = 1,
                                   height = unit(25,"cm"),
                                   width = unit(11,"cm"),
                                   show_row_names = FALSE), 
                          age = anno_simple(anno$age_bin, col = metadata(sce_TME)$colors$age_bin, border = TRUE),
                          grade = anno_simple(anno$grade, col = metadata(sce_TME)$colors$grade, border = TRUE),
                          pT = anno_simple(anno$pT_simple, col=metadata(sce_TME)$colors$pT_simple, border=TRUE),
                          pN = anno_simple(anno$pN_simple, col=metadata(sce_TME)$colors$pN_simple, border=TRUE),
                          pM = anno_simple(anno$pM_simple, col=metadata(sce_TME)$colors$pM_simple, border = TRUE),
                          pretreatment = anno_simple(as.character(anno$pretreatment), col=metadata(sce_TME)$colors$pretreatment, border = TRUE),
                         diversity = anno_simple(anno$shannon_div, col = metadata(sce_TME)$colors$diversity_col, border = TRUE),
                          border = TRUE, 
                          gap = unit(1,"mm"),
                          which = "row")

draw(fraction_anno + NULL)
```

## S9C

```{r}
SID_immune_metacluster <- colData(sce_TME) %>% as.data.frame() %>% select(immune_metacluster,sample_id) %>% group_by(sample_id) %>% table() %>% as.data.frame
SID_immune_metacluster_cor <- SID_immune_metacluster %>% group_by(sample_id) %>% mutate(fra = Freq/sum(Freq))
SID_immune_metacluster_cor$PID <- sce_TME$PID[match(SID_immune_metacluster$sample_id,sce_TME$sample_id)]

SID_immune_metacluster_cor <- foreach(i = unique(colData(sce_TME)$PID),.combine = "rbind")%do%{
SID <- SID_immune_metacluster_cor %>% filter(PID %in% i) %>% select(-Freq,-PID) %>% pivot_wider(id_cols = immune_metacluster,names_from = sample_id,values_from = fra) %>% column_to_rownames("immune_metacluster")
cor <- cor(SID,method = "spearman")
n <- length(colnames(SID))
if(n > 1){
mean(cor[2:n,1])
}else{NA}
}

SID_immune_metacluster_cor <- as.data.frame(SID_immune_metacluster_cor) 
SID_immune_metacluster_cor$PID <- unique(colData(sce_TME)$PID) 
SID_immune_metacluster_cor$category <- "PID"

SID_immune_metacluster_cor_plot <- ggplot(SID_immune_metacluster_cor, aes(x = category, y = V1))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce_TME)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  ylab("Correlation of immune_metacluster fractions per PID")+
  ggtitle(paste("Spearman rho = ",round(median(SID_immune_metacluster_cor$V1, na.rm = TRUE), digits = 2)))

SID_immune_metacluster_cor_plot
```

## S9D

```{r}
# Lasso-regularized Cox proportional hazards model
## MULTIVARIATE

PID_metacluster_combined <- colData(sce_ALL) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster_combined <- PID_metacluster_combined %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metacluster_combined %>% pivot_wider(id_cols = "PID",names_from = "metacluster_combined",values_from = "fra") %>% column_to_rownames("PID")

### Add clinical features
Cox_df <- PID_metagroups
Cox_df$status_DFS <- sce_ALL$status_DFS[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$DFS_months <- sce_ALL$DFS_months[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$age <- sce_ALL$age[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$grade <- factor(sce_ALL$grade[match(rownames(Cox_df),sce_ALL$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce_ALL$pT_simple[match(rownames(Cox_df),sce_ALL$PID)])
Cox_df$pN_simple <- as.factor(sce_ALL$pN_simple[match(rownames(Cox_df),sce_ALL$PID)])
Cox_df$pM_simple <- as.factor(sce_ALL$pM_simple[match(rownames(Cox_df),sce_ALL$PID)]) 
Cox_df <- Cox_df[complete.cases(Cox_df),]
one_hot <- as.data.frame(model.matrix(~ . - 1, data = Cox_df %>% select(grade,pT_simple, pM_simple, pN_simple)))
Cox_df <- cbind(Cox_df %>% select(-grade,-pT_simple, -pM_simple, -pN_simple),one_hot)
Cox_df <- Cox_df %>% filter(DFS_months != 0)
x <-  data.matrix(Cox_df %>% select(-status_DFS,-DFS_months)) #predictor variables
y <- Surv(time = Cox_df$DFS_months, event = Cox_df$status_DFS) #response variables

### Run lasso
registerDoMC(cores = 20)
set.seed(220195)
cv.fit <- cv.glmnet(x, y, family="cox", type.measure = "C", nfolds = 50, parallel = TRUE)
fit <- glmnet(x, y, family="cox")

### Extract output
lasso_df <- as.data.frame(as.matrix(coef(fit, s = cv.fit$lambda.min))) %>% rownames_to_column("predictor")
colnames(lasso_df) <- c("predictor","coef")

### Visualize hazard ratios
ggplot(lasso_df, aes(x = reorder(predictor,-exp(coef)), y = exp(coef))) +
  geom_point(size = 4, aes(color = exp(coef)>1)) +
  geom_hline(yintercept = 1,color = "black")+
  ylab("Lasso HR")+
  xlab("Predictors")+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90))+
  scale_y_continuous(trans="log10")+
  ggtitle("DFS")+
  scale_color_brewer(palette = "Set1")
```

## S9E

Pixel and single-cell level IMC images were generated using [cytoviewer](https://github.com/BodenmillerGroup/cytoviewer).

## S9F

```{r}
cur_sce <- sce_ALL[, sce_ALL$metacluster_tumor_combined %in% c("CD8_T_cell")]
cur_dat <- t(assay(cur_sce, "exprs")) %>% as.data.frame
cur_dat$CD8Tcell_type <- sce_ALL$CD8Tcell_type[match(rownames(cur_dat),colnames(sce_ALL))]
cur_dat$CD8Tcell_type <- factor(cur_dat$CD8Tcell_type, levels = c("ignored","bystanding","infiltrating"))
cur_dat$Ki67_GMM <- sce_ALL$GMM_Ki67[match(rownames(cur_dat),colnames(sce_ALL))]

cur_out <- cur_dat %>% select(CD8Tcell_type, Ki67_GMM) %>% group_by(CD8Tcell_type, Ki67_GMM) %>% count()
cur_plot <- cur_out %>% group_by(CD8Tcell_type) %>% mutate(fra = n/sum(n))
cur_test <- cur_plot %>% pivot_wider(names_from = CD8Tcell_type, id_cols = Ki67_GMM, values_from = n) %>% column_to_rownames("Ki67_GMM")
cur_test_out <- chisq.test(cur_test)

ggplot(cur_plot) +
  geom_tile(color = "black",aes(CD8Tcell_type, fra, fill = Ki67_GMM), position = "stack")+
  scale_fill_brewer(palette = "Paired")+
  theme_classic(base_size = 15)+
  ylab("Density")+
  theme(axis.title.x = element_blank())+
  ggtitle(paste("p (Chi-Square) = ",cur_test_out$p.value))
```

## S9G

```{r}
# Kaplan-Meier
## Spatial groups
cur_surv <- colData(sce_ALL) %>% as.data.frame() %>% select(PID, spatial_CD8_pheno_merg, DFS_months, status_DFS) %>% unique()

fit <- survfit(Surv(DFS_months, status_DFS) ~ spatial_CD8_pheno_merg, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c("steelblue1","firebrick"))
```

## S9H

```{r}
# Cox proportional hazards model
## MULTIVARIATE 

PID_CD8 <- colData(sce_ALL) %>% as.data.frame() %>% select(spatial_CD8_pheno,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_CD8 <- PID_CD8 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_CD8 %>% pivot_wider(id_cols = "PID",names_from = "spatial_CD8_pheno",values_from = "fra") %>% column_to_rownames("PID")

Cox_df <- PID_metagroups

### Add clinical covariates
Cox_df$status_DFS <- sce_ALL$status_DFS[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$DFS_months <- sce_ALL$DFS_months[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$age <- sce_ALL$age[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$grade <- factor(sce_ALL$grade[match(rownames(Cox_df),sce_ALL$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce_ALL$pT_simple[match(rownames(Cox_df),sce_ALL$PID)])
Cox_df$pN_simple <- as.factor(sce_ALL$pN_simple[match(rownames(Cox_df),sce_ALL$PID)])
Cox_df$pM_simple <- as.factor(sce_ALL$pM_simple[match(rownames(Cox_df),sce_ALL$PID)])

covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

### Plot hazard ratios
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0.1,63))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))
```

## S9I

```{r}
# Kaplan-Meier
cur_surv <- colData(sce_ALL) %>% as.data.frame() %>% select(PID, CD8_number_category, DFS_months, status_DFS) %>% unique()

fit <- survfit(Surv(DFS_months, status_DFS) ~ CD8_number_category, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = "Set1")
```

## S9J

```{r}
# Cox proportional hazards model
## MULTIVARIATE 

PID_CD8 <- colData(sce_ALL) %>% as.data.frame() %>% select(CD8_number_category,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_CD8 <- PID_CD8 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_CD8 %>% pivot_wider(id_cols = "PID",names_from = "CD8_number_category",values_from = "fra") %>% column_to_rownames("PID")

Cox_df <- PID_metagroups

### Add clinical covariates
Cox_df$status_DFS <- sce_ALL$status_DFS[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$DFS_months <- sce_ALL$DFS_months[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$age <- sce_ALL$age[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$grade <- factor(sce_ALL$grade[match(rownames(Cox_df),sce_ALL$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce_ALL$pT_simple[match(rownames(Cox_df),sce_ALL$PID)])
Cox_df$pN_simple <- as.factor(sce_ALL$pN_simple[match(rownames(Cox_df),sce_ALL$PID)])
Cox_df$pM_simple <- as.factor(sce_ALL$pM_simple[match(rownames(Cox_df),sce_ALL$PID)])

covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -high) %>% colnames
multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+high"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

### Plot hazard ratios
ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0.1,100))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))
```

## S9K

```{r}
out <- colData(sce_ALL) %>% as.data.frame %>% select(PID, sample_id, spatial_CD8_pheno, metacluster_tumor_combined) %>% filter(metacluster_tumor_combined == "CD8_T_cell") %>% group_by(sample_id, spatial_CD8_pheno, PID) %>% dplyr::count() %>% group_by(PID, spatial_CD8_pheno) %>% mutate(n = mean(n)) %>% select(PID, spatial_CD8_pheno, n) %>% unique()

out$CD8_category <- sce_ALL$CD8_number_category[match(out$PID, sce_ALL$PID)]
out_viz <- out %>% group_by(CD8_category, spatial_CD8_pheno) %>% dplyr::count() %>% ungroup() %>% group_by(CD8_category) %>% mutate(fra = n/sum(n))

out_viz$CD8_category <- factor(out_viz$CD8_category, levels = c("high","medium","low"))

ggplot(out_viz, aes(x = CD8_category, y = fra, fill = spatial_CD8_pheno))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = metadata(sce_ALL)$colors$spatial_CD8_pheno)
```

## S9L

```{r}
### Differential abundance analysis based on edgeR
abundances <- table(sce_ALL$metacluster_combined, sce_ALL$PID)
abundances <- unclass(abundances) 
meta <- colData(sce_ALL)[match(colnames(abundances), sce_ALL$PID),]
meta$spatial_CD8_pheno_merg <- relevel(x=factor(meta$spatial_CD8_pheno_merg), ref = "cold/excluded")

#filter
meta <- meta[!is.na(meta$spatial_CD8_pheno_merg),]
abundances <- abundances[,colnames(abundances) %in% meta$PID]

#create DGEList - based on edgeR package
y.ab <- DGEList(abundances, samples=meta, group = meta$spatial_CD8_pheno_merg)

#Design matrix
design <- model.matrix(~spatial_CD8_pheno_merg, data = y.ab$samples)

# Estimate NB and QL dispersion 
y.ab <- estimateDisp(y.ab, design, trend="none")
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)

# Test for differential abundance o
res <- glmQLFTest(fit.ab, coef=ncol(design))
DA <- res$table
DA <- topTags(res, n = nrow(DA), adjust.method = "BH")$table
DA <- DA %>% mutate(cluster = rownames(DA), sign = DA$FDR<0.1, result = ifelse(DA$logFC > 0 & DA$FDR < 0.1, "UP",ifelse(DA$logFC < 0 & DA$FDR < 0.1, "DOWN", "NOTSIG"))) %>% arrange(desc(logFC)) %>% mutate(category = ifelse(str_detect(cluster,"tumor"),"tumor","non_tumor"))
DA$cluster <- factor(DA$cluster, levels = DA$cluster)
sig <- DA %>% filter(FDR<0.1) %>% pull(cluster) %>% unfactor

## Visualize
ggplot(DA, aes(x = cluster, y = category, col = logFC, size = -log10(FDR))) +
  geom_point() +
  theme_classic() +
  scale_color_gradient2(high = "#E41A1C", mid = "white", low = "#377EB8") +
  scale_size(range = c(2, 10)) +
  new_scale_colour() +
  geom_point(data = DA, shape = 1, aes(size = -log10(FDR), col = result), stroke = 1.5) +
  scale_color_manual(values = c("UP" = "#E41A1C", "NOTSIG" = "snow2", "DOWN" = "#377EB8"), name = "Direction") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## S9M

Pixel and single-cell level IMC images were generated using [cytoviewer](https://github.com/BodenmillerGroup/cytoviewer).

# Figure S10

## S10A

```{r}
results <- spicyR_spatialgroup
marksToPlot <- unfactor(unique(sce_ALL$metacluster_combined))
breaks <- NULL
comparisonGroup <- NULL
colours <- c("#4575B4", "white", "#D73027")
cutoff <- 0.05
fdr <- TRUE

if (is.null(comparisonGroup)) {
    coef <- 2
  } else {
    coef <- which(levels(results$condition) == comparisonGroup)
  }
  marks <- unique(results$comparisons$from)
  
  if (is.null(breaks)) breaks <- c(-3, 3, 0.5)
  breaks <- seq(from = breaks[1], to = breaks[2], by = breaks[3])
  pal <- grDevices::colorRampPalette(colours)(length(breaks))
  col <- colorRamp2(breaks, pal)

  pVal <- results$p.value[, coef]

  if (min(pVal, na.rm = TRUE) == 0) {
    pVal[pVal == 0] <-
      pVal[pVal == 0] + 10^floor(log10(min(pVal[pVal > 0], na.rm = TRUE)))
  }

  if (fdr) {
    pVal <- stats::p.adjust(pVal, method = "fdr")
  }

  isGreater <- results$coefficient[, coef] > 0

  pVal <- log10(pVal)

  pVal[isGreater] <- abs(pVal[isGreater])

  pVal <- matrix(pVal, nrow = length(marks))
  colnames(pVal) <- marks
  rownames(pVal) <- marks
  
  pVal[abs(pVal) < -log10(0.05)] <- 0
  
  anno <- data.frame(celltypes = marksToPlot)
  
  anno$type <- ifelse(str_detect(anno$celltypes, "tumor"),"tumor","TME")
  
Heatmap(pVal,
          col = col,
          name = "log10(p)",
          row_split = anno$type,
          column_split = anno$type,
          cluster_columns = FALSE, 
          cluster_rows = FALSE, 
          border = TRUE, 
          row_title = "From", 
          column_title = "To")
```

## S10B

```{r}
## Number of B patches per PID 
n_Bpatch_PID <- colData(sce_ALL) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  unique() %>% 
  group_by(spatial_CD8_pheno_merg, PID) %>% 
  count()
plot_df <- colData(sce_ALL) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg) %>% 
  unique() 
rownames(plot_df) <- NULL
plot_df$n_Bpatch <- n_Bpatch_PID$n[match(plot_df$PID, n_Bpatch_PID$PID)]
plot_df$n_Bpatch[is.na(plot_df$n_Bpatch)] <- 0
plot_df$spatial_CD8_pheno <- sce_ALL$spatial_CD8_pheno[match(plot_df$PID, sce_ALL$PID)]

ggplot(plot_df, aes(x = spatial_CD8_pheno, y = n_Bpatch))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = spatial_CD8_pheno), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce_ALL)$colors$spatial_CD8_pheno)+
  theme_classic(base_size = 12)+
  coord_cartesian(ylim = c(0,15))+
  theme(axis.title.x = element_blank())+
  ylab("Number of B cell patches per PID")+
  stat_compare_means(method = "kruskal.test")
```

## S10C

```{r}
PDL1_IHC$PID <- factor(PDL1_IHC$PID, levels = PDL1_IHC %>% arrange(desc(CPS_score),desc(spatial_CD8_pheno)) %>% pull(PID))

ggplot(PDL1_IHC, aes(x = PID, y = CPS_score, fill = CPS_score_fine))+
  geom_bar(stat = "identity", color = "black")+
  geom_rug(aes(color = spatial_CD8_pheno), sides = "b", linewidth = 3)+
  geom_hline(yintercept = 10, linetype = "dashed")+
  theme_classic()+
  scale_color_manual(values = metadata(sce_ALL)$colors$spatial_CD8_pheno)+
  scale_fill_manual(values = c("#ECD2FC","#8019C0", "#BF8CDF"))+
  theme(axis.text.x = element_text(angle = 90))
```

## S10E

```{r}
out_check <- colData(sce_wang) %>% as.data.frame %>% dplyr::select(PID_phase, BiopsyPhase, spatial_CD8_pheno_merg, spatial_CD8_pheno, Arm, pCR) %>% unique() 
out_C <- out_check %>% filter(BiopsyPhase == "Baseline") %>% group_by(spatial_CD8_pheno, Arm, BiopsyPhase, pCR) %>% dplyr::count() %>% group_by(spatial_CD8_pheno, BiopsyPhase, Arm) %>% mutate(fra = n/sum(n))
out_C$pCR <- factor(out_C$pCR, levels = c("RD","pCR"))

res_CI <- as.data.frame(epitab(xtabs(n ~ spatial_CD8_pheno + pCR, data = out_C %>% filter(Arm == "C&I")), method = "oddsratio", pvalue = "chi2")$tab)
res_CI <- res_CI %>% rownames_to_column("spatial_CD8_pheno")
res_CI$category <- "CI"

res_C <- as.data.frame(epitab(xtabs(n ~ spatial_CD8_pheno + pCR, data = out_C %>% filter(Arm == "C")), method = "oddsratio", pvalue = "chi2")$tab)
res_C <- res_C %>% rownames_to_column("spatial_CD8_pheno")
res_C$category <- "C"

res_comb <- rbind(res_CI, res_C)
res_comb$sig <- ifelse(res_comb$p.value < 0.05, TRUE, FALSE)

ggplot(res_comb, aes(x = spatial_CD8_pheno, y = oddsratio, color = sig))+
facet_wrap(~category)+
geom_errorbar(aes(ymax = upper, ymin = lower))+
  geom_point(size = 4)+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Spatial groups")+
ylab("Odds ratio (pCR/RD)")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))+
coord_flip()
```

## S10F

```{r}
out_check <- colData(sce_wang) %>% as.data.frame %>% dplyr::select(PID_phase, BiopsyPhase, CD8_number_category_merg, CD8_number_category, Arm, pCR) %>% unique() 

out_C <- out_check %>% filter(BiopsyPhase == "Baseline") %>% group_by(CD8_number_category, Arm, BiopsyPhase, pCR) %>% dplyr::count() %>% group_by(CD8_number_category, BiopsyPhase, Arm) %>% mutate(fra = n/sum(n))

out_C$CD8_number_category <- factor(out_C$CD8_number_category, levels = c("low","medium","high"))

out_C$pCR <- factor(out_C$pCR, levels = c("RD","pCR"))

ggplot(out_C, aes(x = CD8_number_category, y = fra, fill = pCR))+
  facet_wrap(~Arm + BiopsyPhase)+
  geom_bar(stat = "identity", color = "black")+
  geom_hline(yintercept = 0.5, linetype = "dashed")+
  theme_classic()+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  coord_flip()

fisher_out <- chisq.test(xtabs(n ~ CD8_number_category + pCR, data = out_C %>% filter(Arm == "C&I")))
fisher_out

fisher_out_2 <- chisq.test(xtabs(n ~ CD8_number_category + pCR, data = out_C %>% filter(Arm == "C")))
fisher_out_2
```

## S10G

```{r}
out_check <- colData(sce_wang) %>% as.data.frame %>% dplyr::select(PID_phase, BiopsyPhase, CD8_number_category_merg, CD8_number_category, Arm, pCR) %>% unique() 
out_C <- out_check %>% filter(BiopsyPhase == "Baseline") %>% group_by(CD8_number_category, Arm, BiopsyPhase, pCR) %>% dplyr::count() %>% group_by(CD8_number_category, BiopsyPhase, Arm) %>% mutate(fra = n/sum(n))
out_C$pCR <- factor(out_C$pCR, levels = c("RD","pCR"))
out_C$CD8_number_category <- factor(out_C$CD8_number_category, levels = c("low","medium","high"))

res_CI <- as.data.frame(epitab(xtabs(n ~ CD8_number_category + pCR, data = out_C %>% filter(Arm == "C&I")), method = "oddsratio", pvalue = "chi2")$tab)
res_CI <- res_CI %>% rownames_to_column("CD8_number_category")
res_CI$category <- "CI"

res_C <- as.data.frame(epitab(xtabs(n ~ CD8_number_category + pCR, data = out_C %>% filter(Arm == "C")), method = "oddsratio", pvalue = "chi2")$tab)
res_C <- res_C %>% rownames_to_column("CD8_number_category")
res_C$category <- "C"

res_comb <- rbind(res_CI, res_C)
res_comb$sig <- ifelse(res_comb$p.value < 0.05, TRUE, FALSE)
res_comb$CD8_number_category <- factor(res_comb$CD8_number_category, levels = c("low","medium","high"))

ggplot(res_comb, aes(x = CD8_number_category, y = oddsratio, color = sig))+
facet_wrap(~category)+
geom_errorbar(aes(ymax = upper, ymin = lower))+
  geom_point(size = 4)+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Spatial groups")+
ylab("Odds ratio (pCR/RD)")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))+
coord_flip()
```

## S10H

```{r}
colors <- colData(sce_wang) %>% as.data.frame %>% select(celltype_fine, Colour) %>% unique()
colors_celltype_fine <- setNames(colors$Colour, colors$celltype_fine)

out_T_cell_type <- colData(sce_wang) %>% as.data.frame %>% filter(BiopsyPhase == "Baseline" & !is.na(CD8Tcell_type)) %>% as.data.frame %>% select(CD8Tcell_type, celltype_fine) %>% group_by(CD8Tcell_type, celltype_fine) %>% dplyr::count() %>% group_by(CD8Tcell_type) %>% mutate(fra = n/sum(n))

out_T_cell_type$CD8Tcell_type <- factor(out_T_cell_type$CD8Tcell_type, levels = c("ignored", "bystanding", "infiltrating"))

ggplot(out_T_cell_type, aes(x = CD8Tcell_type, y = fra, fill = celltype_fine))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = colors_celltype_fine)+
  theme_classic()
```

## S10I

```{r}
sce_wang_CD8 <- sce_wang[,sce_wang$celltype_broad == "CD8_T" & sce_wang$BiopsyPhase == "Baseline"]

# Aggregate expression per PID 
spatialpheno_mean_PID <- aggregateAcrossCells(sce_wang_CD8, ids = sce_wang_CD8$PID,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_wang_CD8)$CD8_T_cell_markers)

# Test for differential expression per category 
test <- t(assay(spatialpheno_mean_PID, "exprs")[rowData(spatialpheno_mean_PID)$CD8_T_cell_markers,]) %>% as.data.frame %>% mutate(spatial_CD8_pheno = spatialpheno_mean_PID$spatial_CD8_pheno) 
test_result <- compare_means(c(TOX, TCF1, PD1, ICOS, GZMB)~spatial_CD8_pheno, method = "kruskal.test",data = test)
test_result$p.adj <- p.adjust(test_result$p, method = "BH") 
test_result$psignif.adj <- as.matrix(symnum(test_result$p.adj,
       cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), 
       symbols = c("****", "***", "**", "*", "ns")))[,1]
rownames(test_result) <- test_result$.y.
sig.proteins <- unfactor(test_result$.y.[test_result$p.adj < 0.05])

# Aggregate expression per spatial phenotype
spatialpheno_mean_sce_wang <- aggregateAcrossCells(sce_wang_CD8, ids = sce_wang_CD8$spatial_CD8_pheno,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_wang_CD8)$CD8_T_cell_markers)
mat_scaled<- t(scale(t(assay(spatialpheno_mean_sce_wang, "exprs"))))
highest_cat <- mat_scaled %>% as.data.frame() %>% rownames_to_column("gene") %>% pivot_longer(cols = 2:4, names_to = "spatial_CD8", values_to = "exprs") %>% group_by(gene) %>% mutate(max = max(exprs)) %>% filter(exprs == max) %>% select(gene, spatial_CD8) %>% as.data.frame() %>% column_to_rownames("gene")

## Visualize 
h <- Heatmap(assay(spatialpheno_mean_sce_wang, "exprs"), col = viridis(100), name = "exprs", cluster_columns = FALSE, row_names_side = "left", show_row_dend = FALSE)+
  rowAnnotation(cat = anno_simple(highest_cat, col = metadata(sce_wang)$colors$spatial_CD8_pheno, width = unit(10, "mm")))+
  rowAnnotation(test = anno_text(test_result$psignif.adj))

lgd_cat <- Legend(title = "Highest_exprs", at = names(metadata(sce_wang)$colors$spatial_CD8_pheno), 
              legend_gp = gpar(fill = metadata(sce_wang)$colors$spatial_CD8_pheno))

draw(h,annotation_legend_list = list(lgd_cat))
```

## S10J

```{r}
### Differential abundance analysis based on edgeR
sce_wang_sub <- sce_wang[,sce_wang$BiopsyPhase == "Baseline"]

abundances <- table(sce_wang_sub$celltype_fine, sce_wang_sub$PID)
abundances <- unclass(abundances) 
meta <- colData(sce_wang_sub)[match(colnames(abundances), sce_wang_sub$PID),]
meta$spatial_CD8_pheno_merg <- relevel(x=factor(meta$spatial_CD8_pheno_merg), ref = "cold_excluded")

#filter
meta <- meta[!is.na(meta$spatial_CD8_pheno_merg),]
abundances <- abundances[,colnames(abundances) %in% meta$PID]

#create DGEList - based on edgeR package
y.ab <- DGEList(abundances, samples=meta, group = meta$spatial_CD8_pheno_merg)

#Design matrix
design <- model.matrix(~spatial_CD8_pheno_merg, data = y.ab$samples)

# Estimate NB and QL dispersion 
y.ab <- estimateDisp(y.ab, design, trend="none")
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)

# Test for differential abundance
res <- glmQLFTest(fit.ab, coef=ncol(design))
DA <- res$table
DA <- topTags(res, n = nrow(DA), adjust.method = "BH")$table
DA <- DA %>% mutate(cluster = rownames(DA), sign = DA$FDR<0.05, result = ifelse(DA$logFC > 0 & DA$FDR < 0.05, "UP",ifelse(DA$logFC < 0 & DA$FDR < 0.05, "DOWN", "NOTSIG"))) %>% arrange(desc(logFC)) 
DA$category <- sce_wang$isEpithelial[match(DA$cluster, sce_wang$celltype_fine)]
DA$cluster <- factor(DA$cluster, levels = DA$cluster)
sig <- DA %>% filter(FDR<0.05) %>% pull(cluster) %>% unfactor

## Visualize
ggplot(DA, aes(x = cluster, y = category, col = logFC, size = -log10(FDR))) +
  geom_point() +
  theme_classic() +
  scale_color_gradient2(high = "#E41A1C", mid = "white", low = "#377EB8") +
  scale_size(range = c(2, 10)) +
  new_scale_colour() +
  geom_point(data = DA, shape = 1, aes(size = -log10(FDR), col = result), stroke = 1.5) +
  scale_color_manual(values = c("UP" = "#E41A1C", "NOTSIG" = "snow2", "DOWN" = "#377EB8"), name = "Direction") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Session Info 

```{r}
sessionInfo()
```