---
title: null
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

# Figure 7 and Figure S11-S15

Data analysis scripts for re-creating Figure 7 and Figure S11-15. 

# Software requirements

```{r, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(tidyverse)
library(ComplexHeatmap)
library(survival)
library(survminer)
library(conflicted)
library(easyalluvial)
library(wesanderson)
library(scuttle)
library(RColorBrewer)
library(CATALYST)
library(dittoSeq)
library(gridExtra)
library(patchwork)
library(ggnewscale)
```

```{r, message=FALSE,warning=FALSE}
conflicts_prefer(dplyr::desc)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::count)
conflicts_prefer(dplyr::select)
```

# Read data

```{r}
# Available from Zenodo - DOI: 10.5281/zenodo.15396067
# File paths have to be adjusted by the user

## SCE objects
sce_ALL <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_ALL.rds")
sce_fischer_ALL <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_fischer_ALL.rds")

## Other
TCGA_DET <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/TCGA_DET.rds")
TCGA_DS <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/TCGA_DS.rds")
TCGA_QS <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/TCGA_QS.rds")
MDACC_ES <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/MDACC_ES.rds")
MDACC_EXP <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/MDACC_EXP.rds")
ISPY2_ES <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/ISPY2_ES.rds")
ISPY2_BM <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/ISPY2_BM.rds")
GeneSignatures <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/GeneSignatures.rds") 
```

# Figure 7

## 7A

```{r, fig.width=14, fig.height=5}
### 1. Heatmap body ###

# Tumor metacluster
PID_spatial_tumor <- colData(sce_ALL) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_spatial_tumor <- PID_spatial_tumor %>% filter(str_detect(metacluster_combined, "tumor")) %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_spatial_tumor$BL_CD8_group_merg <- sce_ALL$BL_CD8_group_merg[match(PID_spatial_tumor$PID, sce_ALL$PID)]
PID_spatial_tumor <- PID_spatial_tumor[!is.na(PID_spatial_tumor$BL_CD8_group_merg),]

hb_tumor <- PID_spatial_tumor %>% ungroup %>% select(metacluster_combined,BL_CD8_group_merg,fra) %>% group_by(BL_CD8_group_merg, metacluster_combined) %>% summarize(mean = mean(fra)) %>% pivot_wider(names_from = metacluster_combined, values_from = mean) %>% column_to_rownames("BL_CD8_group_merg")
hb_tumor_scaled <- scale(as.matrix(hb_tumor))

# TME metacluster
PID_spatial_TME <- colData(sce_ALL) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_spatial_TME <- PID_spatial_TME %>% filter(!str_detect(metacluster_combined, "tumor")) %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_spatial_TME$BL_CD8_group_merg <- sce_ALL$BL_CD8_group_merg[match(PID_spatial_TME$PID, sce_ALL$PID)]
PID_spatial_TME <- PID_spatial_TME[!is.na(PID_spatial_TME$BL_CD8_group_merg),]

hb_TME <- PID_spatial_TME %>% ungroup %>% select(metacluster_combined,BL_CD8_group_merg,fra) %>% group_by(BL_CD8_group_merg, metacluster_combined) %>% summarize(mean = mean(fra)) %>% pivot_wider(names_from = metacluster_combined, values_from = mean) %>% column_to_rownames("BL_CD8_group_merg")
hb_TME_scaled <- scale(as.matrix(hb_TME))

### 2. Heatmap annotation ###

###2.1 Row - Metadata features

# Annotation dataframe 
anno <- colData(sce_ALL) %>% as.data.frame %>% select(BL_CD8_group_merg) %>% unique() %>% filter(!is.na(BL_CD8_group_merg))
rownames(anno) <- anno$BL_CD8_group_merg
  
# Proportion of grade per cluster
grade <- colData(sce_ALL) %>% as.data.frame() %>% select(BL_CD8_group_merg,grade) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
grade <- grade %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq)) 
grade <- grade %>% select(-Freq) %>% pivot_wider(id_cols = BL_CD8_group_merg,names_from = grade,values_from = fra) %>% column_to_rownames("BL_CD8_group_merg")

# Proportion of pT stage per cluster
pT <- colData(sce_ALL) %>% as.data.frame() %>% select(BL_CD8_group_merg,pT_simple) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
pT <- pT %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq)) 
pT <- pT %>% select(-Freq) %>% pivot_wider(id_cols = BL_CD8_group_merg,names_from = pT_simple,values_from = fra) %>% column_to_rownames("BL_CD8_group_merg")

# Proportion of pN stage per cluster
pN <- colData(sce_ALL) %>% as.data.frame() %>% select(BL_CD8_group_merg,pN_simple) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
pN <- pN %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq)) 
pN <- pN %>% select(-Freq) %>% pivot_wider(id_cols = BL_CD8_group_merg,names_from = pN_simple,values_from = fra) %>% column_to_rownames("BL_CD8_group_merg")

# Number of contributing patients
cluster_PID <- colData(sce_ALL) %>% as.data.frame() %>% select(BL_CD8_group_merg,PID) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
n_PID <- cluster_PID %>% filter(Freq>0) %>% group_by(BL_CD8_group_merg) %>% count(name = "n_PID") %>% column_to_rownames("BL_CD8_group_merg")

# Risk group 
anno$BL_CD8_group_merg <- anno$BL_CD8_group_merg[match(rownames(n_PID), rownames(anno))]
rownames(anno) <- anno$BL_CD8_group_merg

anno$PID_group_risk <- sce_ALL$PID_group_risk[match(anno$BL_CD8_group_merg, sce_ALL$BL_CD8_group_merg)]


# Metadata annotation
ha_group <- HeatmapAnnotation(BL_CD8_group_merg = anno_simple(unfactor(anno$BL_CD8_group_merg),border=TRUE,col=metadata(sce_ALL)$colors$group_color),
                              PID_group_risk = anno_simple(unfactor(anno$PID_group_risk),border=TRUE,col=metadata(sce_ALL)$colors$risk_color),
                             border = TRUE, 
                             annotation_name_rot = 90,
                             gap = unit(2,"mm"),
                             which = "row")

ha_meta <- HeatmapAnnotation(n_PID = anno_barplot(n_PID, width = unit(10, "mm"),gp = gpar(fill = "#440154FF")),
                             grade = anno_barplot(grade, width = unit(10, "mm"),gp = gpar(fill = metadata(sce_ALL)$colors$grade)),
                             pT = anno_barplot(pT, width = unit(10, "mm"),gp = gpar(fill = metadata(sce_ALL)$colors$pT_simple)),
                             pN = anno_barplot(pN, width = unit(10, "mm"),gp = gpar(fill = metadata(sce_ALL)$colors$pN_simple)),
                             border = TRUE, 
                             annotation_name_rot = 90,
                             gap = unit(2,"mm"),
                             which = "row")


hb_tumor_scaled <- hb_tumor_scaled[match(rownames(anno),rownames(hb_tumor_scaled)),]
hb_TME_scaled <- hb_TME_scaled[match(rownames(anno),rownames(hb_TME_scaled)),]

###2.2 Row - Score features

# Expression of CK5,CK7, CD8a 
cluster_mean_sce_ALL <- aggregateAcrossCells(sce_ALL, ids = sce_ALL$BL_CD8_group_merg,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_ALL)$clean_target %in% c("CK5","CK7","HLADR"))

exprs_out <- scale(t(assay(cluster_mean_sce_ALL, "exprs")))

# Number of B cell patches per PID 
n_Bpatch_PID <- colData(sce_ALL) %>% as.data.frame %>% select(PID, BL_CD8_group_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  unique() %>% 
  group_by(BL_CD8_group_merg, PID) %>% 
  count()

plot_df <- colData(sce_ALL) %>% as.data.frame %>% select(PID, BL_CD8_group_merg) %>% 
  unique() 
rownames(plot_df) <- NULL
plot_df$n_Bpatch <- n_Bpatch_PID$n[match(plot_df$PID, n_Bpatch_PID$PID)]
plot_df$n_Bpatch[is.na(plot_df$n_Bpatch)] <- 0
plot_df$BL_CD8_group_merg <- sce_ALL$BL_CD8_group_merg[match(plot_df$PID, sce_ALL$PID)]

# Proportion of B cell score per group
Bscore <- colData(sce_ALL) %>% as.data.frame() %>% select(BL_CD8_group_merg,B_cell_score) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
Bscore <- Bscore %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq)) 
Bscore <- Bscore %>% select(-Freq) %>% pivot_wider(id_cols = BL_CD8_group_merg,names_from = B_cell_score,values_from = fra) %>% column_to_rownames("BL_CD8_group_merg") 

B_out <- plot_df %>% pivot_wider(names_from = PID, values_from = n_Bpatch) %>% column_to_rownames("BL_CD8_group_merg")
B_out <- B_out[match(rownames(Bscore),rownames(B_out)),]

# Proportion of proliferation score per group
prolif_PID <- colData(sce_ALL) %>% as.data.frame() %>% select(PID,GMM_Ki67) %>% group_by(PID) %>% table() %>% as.data.frame
prolif_PID <- prolif_PID %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
prolif_PID$BL_CD8_group_merg <- sce_ALL$BL_CD8_group_merg[match(prolif_PID$PID, sce_ALL$PID)]
prolif_PID <- prolif_PID %>% filter(GMM_Ki67 == "Ki67+")

prolif_out <- prolif_PID %>% select(PID, fra, BL_CD8_group_merg) %>% pivot_wider(names_from = PID, values_from = fra) %>% column_to_rownames("BL_CD8_group_merg")
prolif_out <- prolif_out[match(rownames(Bscore),rownames(prolif_out)),]

col_patch <- setNames(brewer.pal(4,"Paired"),colnames(Bscore))

ha_score <- HeatmapAnnotation(
  prolif_fraction = anno_boxplot(as.matrix(prolif_out), gp = gpar(fill = "pink"), which = "row",outline = FALSE,width = unit(15, "mm")),
  B_cell_score = anno_barplot(Bscore, gp = gpar(fill = col_patch),width = unit(10, "mm")),
  B_cell_patch = anno_boxplot(as.matrix(B_out), gp = gpar(fill = "#B2DF8A"), which = "row",outline = FALSE,width = unit(15, "mm")),
  border = TRUE, 
  annotation_name_rot = 90,
  gap = unit(2,"mm"),
  which = "row")

### 3. Plot heatmap ###
order <- c("BL","L","B","N","inflamed")

h <- ha_group+
  Heatmap(hb_tumor_scaled,
        name = "Mean tumor MC fraction",
        show_row_dend = FALSE,
        clustering_method_rows = "ward.D2", 
        clustering_method_columns = "ward.D2",
        row_order = order,
        row_title = NULL)+
  Heatmap(hb_TME_scaled,
          name = "Mean TME MC fraction")+
  ha_meta+
  Heatmap(exprs_out, 
          name = "Mean expression")+
  ha_score

#Create legends for annotation 
lgd <- Legend(title = "Patient groups", at = names(metadata(sce_ALL)$colors$group_color), 
              legend_gp = gpar(fill = metadata(sce_ALL)$colors$group_color))

lgd_1 <- Legend(title = "Risk groups", at = names(metadata(sce_ALL)$colors$risk_color), 
              legend_gp = gpar(fill = metadata(sce_ALL)$colors$risk_color))

# Draw heatmap 
draw(h,annotation_legend_list = list(lgd, lgd_1))
```

## 7B

```{r}
df_alluvial <- colData(sce_ALL) %>% as.data.frame %>% dplyr::select(PID, CK5_7_group_name, spatial_CD8_pheno_merg,BL_CD8_group_merg, PID_group_risk) %>% unique %>% dplyr::select(CK5_7_group_name, spatial_CD8_pheno_merg,BL_CD8_group_merg, PID_group_risk)

alluvial_wide(df_alluvial, 
              colorful_fill_variable_stratum = FALSE,
              col_vector_flow = c("#85D4E3","grey", "#9C964A", "#F4B5BD"), 
              order_levels = c("CK5/7+","CK5/7-","CK7+","CK5+","cold/excluded","BL","N","L","B","high","medium","low"))+
  theme_classic()
```

## 7C/D

```{r}
# Kaplan-Meier
cur_surv <- colData(sce_ALL) %>% as.data.frame() %>% 
  dplyr::select(PID, BL_CD8_group_merg, PID_group_risk, DFS_months, status_DFS) %>% 
  unique()

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ BL_CD8_group_merg, data = cur_surv)
group_color <- metadata(sce_ALL)$colors$group_color

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = unname(group_color))

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ PID_group_risk, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c(wes_palette("Zissou1")[1] ,"gold", wes_palette("Zissou1")[5]))
```

## 7E

```{r}
df_alluvial <- colData(TCGA_DET) %>% as.data.frame %>% dplyr::select(KRT5_7_score, HLADR_group,PID_group_test,PID_group_risk)
alluvial_wide(df_alluvial, 
              colorful_fill_variable_stratum = FALSE,
              col_vector_flow = c("#85D4E3","grey", "#9C964A","#F4B5BD"),
              order_levels = c("KRT5/7+","KRT5/7-","KRT7+","KRT5+","high","medium","low"))+
  theme_classic()
```

## 7F

```{r}
GOI <- c("KRT5","KRT7","HLA-DRA")

assayNames(TCGA_DET) <- "vst"

GOI_mean_sce <- aggregateAcrossCells(TCGA_DET, ids = TCGA_DET$PID_group_test,
                                 statistics = "mean",
                                 use.assay.type = "vst",
                                 subset.row = rowData(TCGA_DET)$gene_name %in% GOI)

Heatmap(scale(t(assay(GOI_mean_sce, "vst"))))
```
## 7G

```{r}
df_alluvial <- pData(MDACC_ES) %>% as.data.frame %>% dplyr::select(KRT5_7_score,GMM_HLADRA,PID_groups, PID_group_risk)

alluvial_wide(df_alluvial, 
              colorful_fill_variable_stratum = FALSE,
              col_vector_flow = c("#85D4E3","grey", "#9C964A", "#F4B5BD"), 
              order_levels = c("KRT5/7+","KRT5/7-","KRT7+","KRT5+", "high", "medium", "low"))+
  theme_classic()
```

## 7H

```{r}
key_marker <- c("KRT5","KRT7","HLA-DRA")

## Create heatmap body 
df <- t(MDACC_EXP[key_marker,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")
df$PID_groups <- MDACC_ES$PID_groups[match(df$PID, colnames(MDACC_ES))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

## Create row annotation 
anno <- df %>% select(PID_groups) %>% group_by(PID_groups) %>% dplyr::count()
anno <- anno[match(anno$PID_groups, rownames(ht)),]

## Print heatmap 
Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(anno = anno_text(anno$PID_groups))
```

## 7I/J

```{r}
# Kaplan-Meier
cur_surv <- colData(TCGA_DET) %>% as.data.frame() %>%
  dplyr::select(PID, PID_group_test, PID_group_risk, DFI, DFI.time) 

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ PID_group_test, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE,
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"),
          log.rank.weights = "n",
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ PID_group_risk, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(),
          palette = c("#3B9AB2", "gold","#F21A00"),
          log.rank.weights = "n",
          xlim = c(0, 4000))
```

## 7K/L 

```{r}
#Assess survival impact 
cur_surv <- pData(MDACC_ES) %>% as.data.frame() %>% dplyr::select(PID, PID_groups, PID_group_risk, DRFS, DRFS_time) 

#Plot survival curves
fit <- survfit(Surv(DRFS_time, DRFS) ~ PID_groups, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"))

#Plot survival curves
fit <- survfit(Surv(DRFS_time, DRFS) ~ PID_group_risk, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE,
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c(wes_palette("Zissou1")[5],"gold",wes_palette("Zissou1")[1]), 
          conf.int = FALSE)
```

# Figure S11

## S11A 

```{r}
sce_ALL_cat <- sce_ALL 
sce_ALL_cat$sample_id <- factor(sce_ALL$PID)
sce_ALL_cat$cluster_id <- factor(sce_ALL$BL_CD8_group_merg)
metadata(sce_ALL_cat)$cluster_codes <- data.frame(PID_SC_groups = factor(sce_ALL_cat$BL_CD8_group_merg))

# MDS pseudobulk by metacluster
pbMDS(sce_ALL_cat, by = "cluster_id", 
      features = rownames(sce_ALL_cat)[rowData(sce_ALL_cat)$use_channel], 
      label_by = "cluster_id", k = "PID_SC_groups") +
  scale_color_manual(values = metadata(sce_ALL_cat)$colors$group_color)+
  theme_classic()
```

## S11B

```{r}
# Cox proportional hazards model
## UNIVARIATE 
BL_CD8_group <- colData(sce_ALL) %>% as.data.frame() %>% dplyr::select(BL_CD8_group_merg,PID) %>% group_by(PID) %>% table() %>% as.data.frame
BL_CD8_group <- BL_CD8_group %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
BL_CD8_group <- BL_CD8_group %>% pivot_wider(id_cols = "PID",names_from = "BL_CD8_group_merg",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- BL_CD8_group

# Add Clinical Covariates
Cox_df$status_DFS <- sce_ALL$status_DFS[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$DFS_months <- sce_ALL$DFS_months[match(rownames(Cox_df),sce_ALL$PID)]

covariates <- Cox_df %>% dplyr::select(-status_DFS,-DFS_months, -inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~", paste(covariates, collapse = " + "), "+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)
plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

# Plot hazard ratios
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))
```
```{r}
# Cox proportional hazards model
## MULTIVARIATE 
BL_CD8_group <- colData(sce_ALL) %>% as.data.frame() %>% dplyr::select(BL_CD8_group_merg,PID) %>% group_by(PID) %>% table() %>% as.data.frame
BL_CD8_group <- BL_CD8_group %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
BL_CD8_group <- BL_CD8_group %>% pivot_wider(id_cols = "PID",names_from = "BL_CD8_group_merg",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- BL_CD8_group

# Add clinical covariates
Cox_df$status_DFS <- sce_ALL$status_DFS[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$DFS_months <- sce_ALL$DFS_months[match(rownames(Cox_df),sce_ALL$PID)]

Cox_df$age <- sce_ALL$age[match(rownames(Cox_df),sce_ALL$PID)]
Cox_df$grade <- factor(sce_ALL$grade[match(rownames(Cox_df),sce_ALL$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce_ALL$pT_simple[match(rownames(Cox_df),sce_ALL$PID)])
Cox_df$pN_simple <- as.factor(sce_ALL$pN_simple[match(rownames(Cox_df),sce_ALL$PID)])
Cox_df$pM_simple <- as.factor(sce_ALL$pM_simple[match(rownames(Cox_df),sce_ALL$PID)])

covariates <- Cox_df %>% dplyr::select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "), "+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)
plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

# Plot hazard ratios
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
coord_cartesian(ylim=c(0.1,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))
```

## S11C

```{r}
# Kaplan-Meier
cur_surv <- colData(sce_ALL) %>% as.data.frame() %>% 
  dplyr::select(PID, BL_CD8_group_merg, PID_group_risk, DFS_months, status_DFS, pretreatment) %>% filter(pretreatment == FALSE) %>%
  unique()

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ BL_CD8_group_merg, data = cur_surv)
group_color <- metadata(sce_ALL)$colors$group_color

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = unname(group_color))
```

## S11D

```{r}
df_alluvial <- colData(sce_fischer_ALL) %>% as.data.frame %>% dplyr::select(PID, CK5_7_group_name, spatial_CD8_pheno_merg,BL_CD8_group_merg, PID_group_risk) %>% unique %>% dplyr::select(CK5_7_group_name, spatial_CD8_pheno_merg,BL_CD8_group_merg, PID_group_risk)

alluvial_wide(df_alluvial, 
              colorful_fill_variable_stratum = FALSE,
              col_vector_flow = c("#85D4E3","grey","#9C964A", "#F4B5BD"), 
              order_levels = c("CK5/7+","CK5/7-","CK7+","CK5+","cold/excluded","BL","N","L","B","high","medium","low"))+
  theme_classic()
```

## S11E

```{r}
fischer_over <- colData(sce_fischer_ALL) %>% as.data.frame %>% select(PID, PAM50_clean, TNBC, BL_CD8_group_merg) %>% unique() %>% remove_rownames()

fischer_PAM50_group <- fischer_over %>% group_by(BL_CD8_group_merg, PAM50_clean) %>% count() %>% ungroup() %>% group_by(PAM50_clean) %>% mutate(fra = n/sum(n))

ggplot(fischer_PAM50_group, aes(x = PAM50_clean, y = fra, fill = BL_CD8_group_merg))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = group_color)+
  theme(axis.text.x = element_text(angle = 90))

fischer_TNBC_group <- fischer_over %>% group_by(BL_CD8_group_merg, TNBC) %>% count() %>% ungroup() %>% group_by(TNBC) %>% mutate(fra = n/sum(n)) 

ggplot(fischer_TNBC_group, aes(x = TNBC, y = fra, fill = BL_CD8_group_merg))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = group_color)+
  theme(axis.text.x = element_text(angle = 90))
```

## S11F

```{r}
# Kaplan-Meier
cur_surv <- colData(sce_fischer_ALL) %>% as.data.frame() %>%
  dplyr::select(PID, BL_CD8_group_merg, OS_months, status_OS) %>% 
  unique()

group_color <- setNames(c("#3B9AB2", "gold", "pink", "orange","#F21A00"), c("inflamed","N","B","L","BL"))

#Plot survival curves
fit <- survfit(Surv(OS_months, status_OS) ~ BL_CD8_group_merg, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = unname(group_color))
```

## S11G

```{r}
# Cox proportional hazards model
## UNIVARIATE 

PID_CK57 <- colData(sce_fischer_ALL) %>% as.data.frame() %>%
  select(BL_CD8_group_merg,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_CK57 <- PID_CK57 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_CK57 %>% pivot_wider(id_cols = "PID",names_from = "BL_CD8_group_merg",values_from = "fra") %>% column_to_rownames("PID")

# Add clinical covariates
Cox_df <- PID_metagroups
Cox_df$status_OS <- sce_fischer_ALL$status_OS[match(rownames(Cox_df),sce_fischer_ALL$PID)]
Cox_df$OS_months <- sce_fischer_ALL$OS_months[match(rownames(Cox_df),sce_fischer_ALL$PID)]

covariates <- Cox_df %>% select(-status_OS,-OS_months, -inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~", paste(covariates, collapse = " + "),"+inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## MULTIVARIATE 

PID_CK57 <- colData(sce_fischer_ALL) %>% as.data.frame() %>%
  select(BL_CD8_group_merg,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_CK57 <- PID_CK57 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_CK57 %>% pivot_wider(id_cols = "PID",names_from = "BL_CD8_group_merg",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

# Add clinical covariates
Cox_df$status_OS <- sce_fischer_ALL$status_OS[match(rownames(Cox_df),sce_fischer_ALL$PID)]
Cox_df$OS_months <- sce_fischer_ALL$OS_months[match(rownames(Cox_df),sce_fischer_ALL$PID)]

Cox_df$age <- sce_fischer_ALL$age[match(rownames(Cox_df),sce_fischer_ALL$PID)]
Cox_df$grade <- factor(sce_fischer_ALL$grade[match(rownames(Cox_df),sce_fischer_ALL$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce_fischer_ALL$pT_simple[match(rownames(Cox_df),sce_fischer_ALL$PID)])
Cox_df$pN_simple <- as.factor(sce_fischer_ALL$pN_simple[match(rownames(Cox_df),sce_fischer_ALL$PID)])
Cox_df$pM_simple <- as.factor(sce_fischer_ALL$pM_simple[match(rownames(Cox_df),sce_fischer_ALL$PID)])

covariates <- Cox_df %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))


#Plot hazard ratios
hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0.1,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

## S11H

```{r}
# Kaplan-Meier
cur_surv <- colData(sce_fischer_ALL) %>% as.data.frame() %>%
  dplyr::select(PID, PID_group_risk, OS_months, status_OS) %>% 
  unique()

#Plot survival curves
fit <- survfit(Surv(OS_months, status_OS) ~ PID_group_risk, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c(wes_palette("Zissou1")[1] ,"gold", wes_palette("Zissou1")[5]))
```

## S11I

```{r}
# Cox proportional hazards model
## UNIVARIATE 

PID_CK57 <- colData(sce_fischer_ALL) %>% as.data.frame() %>%
  select(PID_group_risk,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_CK57 <- PID_CK57 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_CK57 %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")

# Add clinical covariates
Cox_df <- PID_metagroups
Cox_df$status_OS <- sce_fischer_ALL$status_OS[match(rownames(Cox_df),sce_fischer_ALL$PID)]
Cox_df$OS_months <- sce_fischer_ALL$OS_months[match(rownames(Cox_df),sce_fischer_ALL$PID)]

covariates <- Cox_df %>% select(-status_OS,-OS_months, -low) %>% colnames
multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~", paste(covariates, collapse = " + "),"+low"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

# Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## MULTIVARIATE 

PID_CK57 <- colData(sce_fischer_ALL) %>% as.data.frame() %>% 
  select(PID_group_risk,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_CK57 <- PID_CK57 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_CK57 %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")

# Add clinical covariates
Cox_df <- PID_metagroups
Cox_df$status_OS <- sce_fischer_ALL$status_OS[match(rownames(Cox_df),sce_fischer_ALL$PID)]
Cox_df$OS_months <- sce_fischer_ALL$OS_months[match(rownames(Cox_df),sce_fischer_ALL$PID)]

Cox_df$age <- sce_fischer_ALL$age[match(rownames(Cox_df),sce_fischer_ALL$PID)]
Cox_df$grade <- factor(sce_fischer_ALL$grade[match(rownames(Cox_df),sce_fischer_ALL$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce_fischer_ALL$pT_simple[match(rownames(Cox_df),sce_fischer_ALL$PID)])
Cox_df$pN_simple <- as.factor(sce_fischer_ALL$pN_simple[match(rownames(Cox_df),sce_fischer_ALL$PID)])
Cox_df$pM_simple <- as.factor(sce_fischer_ALL$pM_simple[match(rownames(Cox_df),sce_fischer_ALL$PID)])

covariates <- Cox_df %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -low) %>% colnames
multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+low"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))


# Plot hazard ratios
hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0.1,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

# Figure S12

## S12A

```{r}
GMM_df <- data.frame(class_epi = TCGA_DET$HLADR_group, 
                     HLADR_exprs = counts(TCGA_DS, normalized = TRUE)["HLA-DRA",])

ggplot(GMM_df, aes(x = HLADR_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")
```

## S12B

```{r}
GMM_df <- data.frame(class_epi = ifelse(MDACC_EXP["HLA-DRA",] > 13, "HLA-DRA+", "HLA-DRA-"),
                     HLADRA_exprs = MDACC_EXP["HLA-DRA",])

ggplot(GMM_df, aes(x = HLADRA_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")
```

## S12C

```{r}
GOI_plus <- c("CD8A","CD274","PDCD1","GZMB","PDCD1LG2","IDO1", "ICOS", "TOX")

GOI_mean_sce_2 <- aggregateAcrossCells(TCGA_DET, ids = TCGA_DET$PID_group_test,
                                 statistics = "mean",
                                 use.assay.type = "vst",
                                 subset.row = rowData(TCGA_DET)$gene_name %in% GOI_plus)

Heatmap(scale(t(assay(GOI_mean_sce_2, "vst"))))
```

## S12D

```{r}
IFNG_sig <- GeneSignatures$INFG_Ayers 

## IFN signature per PID group 
IFNG_sig_rel <- IFNG_sig[IFNG_sig %in% rownames(assay(TCGA_DET))]
df <- t(assay(TCGA_DET)[IFNG_sig_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_group_test <- TCGA_DET$PID_group_test[match(rownames(df), colnames(TCGA_DET))]

ggplot(df, aes(x = PID_group_test, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = PID_group_test), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(TCGA_DET)$colors$PID_group_test)+
  theme_classic()+
  ylab("IFN-gamma signature score")+
  stat_compare_means()
```

## S12E

```{r}
# Z-scoring of QS results by cell type
TCGA_QS_1 <- TCGA_QS %>% filter(cell_type != "uncharacterized cell") %>% column_to_rownames("cell_type")
TCGA_QS_z <- as.data.frame(t(scale(t(TCGA_QS_1), center=TRUE, scale=TRUE)))
TCGA_QS_z <- TCGA_QS_z %>% rownames_to_column("cell_type")
TCGA_QS_plot <- TCGA_QS_z %>% pivot_longer(2:ncol(TCGA_QS_z), names_to = "PID", values_to = "score")

TCGA_QS_plot$PID_group_test <- TCGA_DET$PID_group_test[match(TCGA_QS_plot$PID, colnames(TCGA_DET))]
  
# Heatmap visualization  
ht_QS <- TCGA_QS_plot %>% group_by(PID_group_test, cell_type) %>% dplyr::summarize(mean = mean(score)) %>% pivot_wider(names_from = PID_group_test, values_from = mean) %>% column_to_rownames("cell_type")

Heatmap(ht_QS, name = "mean QS-Z-score")
```

## S12F

```{r}
dittoBarPlot(TCGA_DET, var = "PID_group_test", group.by = "PAM50Call_RNAseq",retain.factor.levels = TRUE)+
  scale_fill_manual(values = metadata(TCGA_DET)$colors$PID_group_test)+
  ggtitle("")

dittoBarPlot(TCGA_DET, var = "PID_group_test", group.by = "TNBC",retain.factor.levels = TRUE)+
  scale_fill_manual(values = metadata(TCGA_DET)$colors$PID_group_test)+
  ggtitle("")
```

## S12G

```{r}
plot <- as.data.frame(t(MDACC_EXP[rownames(MDACC_EXP) %in% c("KRT5", "KRT7", "KRT18","KRT14","HLA-DRA"),]))
plot$KRT5_7_score <- MDACC_ES$KRT5_7_score[match(rownames(plot),colnames(MDACC_ES))]
plot$PAM50 <- MDACC_ES$`pam50_class:ch1`[match(rownames(plot),colnames(MDACC_ES))]
plot$TNBC <- MDACC_ES$TNBC_assign_lehmann[match(rownames(plot),colnames(MDACC_ES))]
plot$PID_groups <- MDACC_ES$PID_groups[match(rownames(plot),colnames(MDACC_ES))]
group_color <- setNames(c("#3B9AB2", "gold", "pink", "orange","#F21A00"), 
                        c("inflamed","KRT5/7-","KRT5+","KRT7+","KRT5/7+"))

# PID groups 

## PAM50 subtype 
bar_out <- plot %>% as.data.frame() %>% group_by(PID_groups, PAM50) %>% dplyr::count() %>% filter(PAM50 != "NA") %>% ungroup %>% group_by(PAM50) %>% mutate(fra = n/sum(n)) %>% arrange(PAM50)

ggplot(bar_out, aes(x = PAM50, y = fra, fill = PID_groups))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = group_color)+
  theme_classic()

## TNBC 
bar_out <- plot %>% as.data.frame() %>% group_by(PID_groups, TNBC) %>% dplyr::count() %>% ungroup %>% group_by(TNBC) %>% mutate(fra = n/sum(n)) %>% arrange(TNBC)

ggplot(bar_out, aes(x = TNBC, y = fra, fill = PID_groups))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = group_color)+
  theme_classic()
```

## S12H

```{r}
# Kaplan-Meier
cur_surv <- colData(TCGA_DET) %>% as.data.frame() %>%
  dplyr::select(PID, PID_group_test, PID_group_risk, PFI, PFI.time) 

#Plot survival curves
fit <- survfit(Surv(PFI.time, PFI) ~ PID_group_test, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE,
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"),
          log.rank.weights = "n",
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(PFI.time, PFI) ~ PID_group_risk, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(),
          palette = c("#3B9AB2", "gold","#F21A00"),
          log.rank.weights = "n",
          xlim = c(0, 4000))
```

## S12I

```{r}
# Cox proportional hazards model
## UNIVARIATE 

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% 
  dplyr::select(PID_group_test,PID) %>% 
group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_test",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")
Cox_df <- PID_metagroups

# Add clinical covariates
Cox_df$DFI <- TCGA_DET$DFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$DFI.time <- TCGA_DET$DFI.time[match(rownames(Cox_df),TCGA_DET$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

# Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## MULTIVARIATE 

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% 
  dplyr::select(PID_group_test, PID) %>% 
  group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_test",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")
Cox_df <- PID_metagroups

# Add clinical covariates
Cox_df$DFI <- TCGA_DET$DFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$DFI.time <- TCGA_DET$DFI.time[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$age <- as.numeric(TCGA_DET$age[match(rownames(Cox_df),TCGA_DET$PID)])
Cox_df$pT_simple <- as.factor(TCGA_DET$Tumor_nature2012[match(rownames(Cox_df),TCGA_DET$PID)])
Cox_df$pN_simple <- as.factor(TCGA_DET$Node_nature2012[match(rownames(Cox_df),TCGA_DET$PID)])
Cox_df$pM_simple <- as.factor(TCGA_DET$Metastasis_nature2012[match(rownames(Cox_df),TCGA_DET$PID)])

covariates <- Cox_df %>% dplyr::select(-DFI,-DFI.time, -age, -pT_simple, -pN_simple, -pM_simple, - inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~ age + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "), "+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

# Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_log10()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

## S12J

```{r}
# Cox proportional hazards model
## UNIVARIATE 

PID_metagroups <- pData(MDACC_ES) %>% as.data.frame() %>%
  dplyr::select(PID_groups,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_groups",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")
Cox_df <- PID_metagroups

# Add clinical covariates
Cox_df$DFI.time <- as.numeric(MDACC_ES$`drfs_even_time_years:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]
Cox_df$DFI <- as.numeric(MDACC_ES$`drfs_1_event_0_censored:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```
```{r}
# Cox proportional hazards model
## MULTIVARIATE 

PID_metagroups <- pData(MDACC_ES) %>% 
  as.data.frame() %>% 
  dplyr::select(PID_groups,PID) %>% 
  group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_groups",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")
Cox_df <- PID_metagroups

# Add clinical covariates
Cox_df$DFI.time <- as.numeric(MDACC_ES$`drfs_even_time_years:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]
Cox_df$DFI <- as.numeric(MDACC_ES$`drfs_1_event_0_censored:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]
Cox_df$age <- as.numeric(MDACC_ES$`age_years:ch1`[match(rownames(Cox_df),MDACC_ES$PID)])
Cox_df$grade <- as.factor(MDACC_ES$`grade:ch1`[match(rownames(Cox_df),MDACC_ES$PID)])
Cox_df$grade[Cox_df$grade == "NA"] <- NA 
Cox_df$grade <- factor(Cox_df$grade)
Cox_df$pT_simple <- as.factor(MDACC_ES$`clinical_t_stage:ch1`[match(rownames(Cox_df),MDACC_ES$PID)])
Cox_df$pN_simple <- as.factor(MDACC_ES$`clinical_nodal_status:ch1`[match(rownames(Cox_df),MDACC_ES$PID)])

covariates <- Cox_df %>% select(-DFI,-DFI.time, -age, -pT_simple, -pN_simple, -grade, - inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~ age + pT_simple + pN_simple + grade +", paste(covariates, collapse = " + "), "+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

# Figure S13

## S13A

```{r}
# Kaplan-Meier
## Disease free survival 
cur_surv <- colData(TCGA_DET) %>% as.data.frame() %>% filter(TNBC == "TNBC") %>% 
  dplyr::select(PID, PID_group_test, PID_group_risk, TNBCtype, TNBCtype_4, DFI, DFI.time) 

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ PID_group_test, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          log.rank.weights = "n",
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"),
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ PID_group_risk, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          log.rank.weights = "n",
          palette = c("#3B9AB2", "gold","#F21A00"),
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ TNBCtype_4, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          log.rank.weights = "n",
          palette = "Set1",
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ TNBCtype, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(),
          log.rank.weights = "n",
          palette = "Set1",
          xlim = c(0, 4000))
```

## S13B

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## PID group

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>% 
  dplyr::select(PID_group_test,PID) %>% 
group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_test",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")
Cox_df <- PID_metagroups

#Add clinical covariates
Cox_df$DFI <- TCGA_DET$DFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$DFI.time <- TCGA_DET$DFI.time[match(rownames(Cox_df),TCGA_DET$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## PID risk

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>% 
  dplyr::select(PID_group_risk,PID) %>% 
group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

#Add clinical covariates
Cox_df$DFI <- TCGA_DET$DFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$DFI.time <- TCGA_DET$DFI.time[match(rownames(Cox_df),TCGA_DET$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - low) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ low"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## TNBCtype_4 

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% dplyr::select(TNBCtype_4,PID) %>% filter(!is.na(TNBCtype_4)) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype_4",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

# Add clincical covariates
Cox_df$DFI <- TCGA_DET$DFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$DFI.time <- TCGA_DET$DFI.time[match(rownames(Cox_df),TCGA_DET$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - BL1) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ BL1"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

# Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## TNBCtype 

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% dplyr::select(TNBCtype,PID) %>% filter(!is.na(TNBCtype)) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

# Add clincical covariates
Cox_df$DFI <- TCGA_DET$DFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$DFI.time <- TCGA_DET$DFI.time[match(rownames(Cox_df),TCGA_DET$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - BL1) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ BL1"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

# Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```


## S13C

```{r}
# Kaplan-Meier
## Progression free survival 
cur_surv <- colData(TCGA_DET) %>% as.data.frame() %>% filter(TNBC == "TNBC") %>% 
  dplyr::select(PID, PID_group_test, PID_group_risk, TNBCtype, TNBCtype_4, PFI, PFI.time) 

#Plot survival curves
fit <- survfit(Surv(PFI.time, PFI) ~ PID_group_test, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          log.rank.weights = "n",
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"),
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(PFI.time, PFI) ~ PID_group_risk, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          log.rank.weights = "n",
          palette = c("#3B9AB2", "gold","#F21A00"),
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(PFI.time, PFI) ~ TNBCtype_4, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          log.rank.weights = "n",
          palette = "Set1",
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(PFI.time, PFI) ~ TNBCtype, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(),
          log.rank.weights = "n",
          palette = "Set1",
          xlim = c(0, 4000))
```

## S13D

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## PID group

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>% 
  dplyr::select(PID_group_test,PID) %>% 
group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_test",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")
Cox_df <- PID_metagroups

#Add clinical covariates
Cox_df$PFI <- TCGA_DET$PFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$PFI.time <- TCGA_DET$PFI.time[match(rownames(Cox_df),TCGA_DET$PID)]

covariates <- Cox_df %>% select(-PFI,-PFI.time, - inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## PID risk

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>% 
  dplyr::select(PID_group_risk,PID) %>% 
group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

#Add clinical covariates
Cox_df$PFI <- TCGA_DET$PFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$PFI.time <- TCGA_DET$PFI.time[match(rownames(Cox_df),TCGA_DET$PID)]

covariates <- Cox_df %>% select(-PFI,-PFI.time, - low) %>% colnames
multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ low"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## TNBCtype_4 

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% dplyr::select(TNBCtype_4,PID) %>% filter(!is.na(TNBCtype_4)) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype_4",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

# Add clincical covariates
Cox_df$PFI <- TCGA_DET$PFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$PFI.time <- TCGA_DET$PFI.time[match(rownames(Cox_df),TCGA_DET$PID)]

covariates <- Cox_df %>% select(-PFI,-PFI.time, - BL1) %>% colnames
multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ BL1"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

# Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## TNBCtype 

PID_metagroups <- colData(TCGA_DET) %>% as.data.frame() %>% dplyr::select(TNBCtype,PID) %>% filter(!is.na(TNBCtype)) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

# Add clincical covariates
Cox_df$PFI <- TCGA_DET$PFI[match(rownames(Cox_df),TCGA_DET$PID)]
Cox_df$PFI.time <- TCGA_DET$PFI.time[match(rownames(Cox_df),TCGA_DET$PID)]

covariates <- Cox_df %>% select(-PFI,-PFI.time, - BL1) %>% colnames
multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ BL1"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

# Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

## S13E

```{r}
# Kaplan-Meier
## Distant relapse free survival 
cur_surv <- pData(MDACC_ES) %>% as.data.frame() %>% dplyr::select(PID, KRT5_7_score, PID_groups, PID_group_risk, TNBC, `pam50_class:ch1`,TNBC_assign_lehmann, TNBCtype,TNBCtype_4, DRFS_time, DRFS) %>% filter(TNBC_assign_lehmann == "TNBC") 

#Plot survival curves
fit <- survfit(Surv(DRFS_time, DRFS) ~ PID_groups, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"))

#Plot survival curves
fit <- survfit(Surv(DRFS_time, DRFS) ~ PID_group_risk, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c(wes_palette("Zissou1")[5],"gold",wes_palette("Zissou1")[1]), 
          conf.int = FALSE)

#Plot survival curves
fit <- survfit(Surv(DRFS_time, DRFS) ~ TNBCtype_4, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = "Set1", 
          conf.int = FALSE)

#Plot survival curves
fit <- survfit(Surv(DRFS_time, DRFS) ~ TNBCtype, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE,
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(),
          palette = "Set1", 
          conf.int = FALSE)
```

## S13F

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## PID group

PID_metagroups <- pData(MDACC_ES) %>% as.data.frame() %>%
  filter(TNBC_assign_lehmann == "TNBC") %>% 
  dplyr::select(PID_groups,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_groups",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")
Cox_df <- PID_metagroups

#Add clinical covariates 
Cox_df$DFI.time <- as.numeric(MDACC_ES$`drfs_even_time_years:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]
Cox_df$DFI <- as.numeric(MDACC_ES$`drfs_1_event_0_censored:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - inflamed) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ inflamed"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```
```{r}
# Cox proportional hazards model
## UNIVARIATE 
## PID risk

PID_metagroups <- pData(MDACC_ES) %>% 
  as.data.frame() %>% 
  filter(TNBC_assign_lehmann == "TNBC") %>% 
  dplyr::select(PID_group_risk,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

#add clinical covariates
Cox_df$DFI.time <- as.numeric(MDACC_ES$`drfs_even_time_years:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]
Cox_df$DFI <- as.numeric(MDACC_ES$`drfs_1_event_0_censored:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - low) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ low"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)
plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_manual(values = "#377EB8")+
ylab("HR for PID groups")+
xlab("Covariates")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## TNBCtype_4

PID_metagroups <- pData(MDACC_ES) %>% 
  as.data.frame() %>% 
  filter(TNBC_assign_lehmann == "TNBC") %>% 
  dplyr::select(TNBCtype_4,PID) %>% filter(!is.na(TNBCtype_4)) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype_4",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

#Add clinical covariates
Cox_df$DFI.time <- as.numeric(MDACC_ES$`drfs_even_time_years:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]
Cox_df$DFI <- as.numeric(MDACC_ES$`drfs_1_event_0_censored:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - BL1) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ BL1"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# Cox proportional hazards model
## UNIVARIATE 
## TNBCtype

PID_metagroups <- pData(MDACC_ES) %>% 
  as.data.frame() %>% 
  filter(TNBC_assign_lehmann == "TNBC") %>% 
  dplyr::select(TNBCtype,PID) %>% filter(!is.na(TNBCtype)) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype",values_from = "fra") %>% column_to_rownames("PID")
Cox_df <- PID_metagroups

#Add clinical covariates
Cox_df$DFI.time <- as.numeric(MDACC_ES$`drfs_even_time_years:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]
Cox_df$DFI <- as.numeric(MDACC_ES$`drfs_1_event_0_censored:ch1`)[match(rownames(Cox_df),MDACC_ES$PID)]

covariates <- Cox_df %>% select(-DFI,-DFI.time, - BL1) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ BL1"))
multi_model <- coxph(multi_formula, data = Cox_df)
multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

# Figure S14

## S14A

```{r}
### 1. Normalized counts
plot <- as.data.frame(t(exprs(ISPY2_ES)[rownames(exprs(ISPY2_ES)) %in% c("KRT5", "KRT7", "KRT18","KRT14","HLA-DRA"),]))
plot$KRT5_7_score <- ISPY2_ES$KRT5_7_score[match(rownames(plot),colnames(ISPY2_ES))]
plot$GMM_KRT5 <- ISPY2_ES$GMM_KRT5[match(rownames(plot),colnames(ISPY2_ES))]
plot$GMM_KRT7 <- ISPY2_ES$GMM_KRT7[match(rownames(plot),colnames(ISPY2_ES))]
col_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("KRT5/7-", "KRT5/7+","KRT5+", "KRT7+"))

main <- ggplot(plot, aes(x=KRT7, y=KRT5, color = KRT5_7_score)) +
  geom_point(size = 2, alpha = 0.5)+
  theme_classic()+
  xlab("CK7 log counts")+
  ylab("CK5 log counts")+
  scale_color_manual(values = col_score)+
  guides(color = "none")

y_side <- ggplot(plot, aes(x = KRT5, fill=GMM_KRT5)) +
  geom_density(alpha=0.8) +
  scale_fill_manual(values = c('#999999','#F4B5BD')) +
  theme_classic()+
  theme(legend.position = "none", axis.title.y = element_blank())+
  coord_flip()

x_side <- ggplot(plot, aes(x = KRT7, fill=GMM_KRT7)) +
  geom_density(alpha=0.8) +
  scale_fill_manual(values = c('#999999','#9C964A')) +
  theme_classic()+
  theme(legend.position = "none", axis.title.x = element_blank())

blankPlot <- ggplot()+geom_blank(aes(1,1)) +
  cowplot::theme_nothing()

grid.arrange(x_side, blankPlot, main, y_side,
        ncol=2, nrow=2, widths=c(4, 1.4), heights=c(1.4, 4))
```

## S14B

```{r}
# Define keratins of interest
keratins <- c("KRT5","KRT7","KRT15","KRT8","KRT18","KRT19", "KRT14","KRT17","ITGA6","EPCAM")

keratins_rel <- keratins[keratins %in% rownames(exprs(ISPY2_ES))]

## Create heatmap body 
df <- t(exprs(ISPY2_ES)[keratins_rel,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")
df$KRT5_7_score <- ISPY2_ES$KRT5_7_score[match(df$PID, colnames(ISPY2_ES))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% dplyr::count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

```{r}
# Define stem markers of interest
stem <- c("SOX9","KIT","PROM1","HDAC2","NOTCH1","CD44","ALDH1A3","BRCA1","MUC1","EZH2", "YAP1","ANXA8", "TGFB2", "CD24", "TNFAIP3", "ERBB3", "TNFRSF11A")
stem_rel <- stem[stem %in% rownames(exprs(ISPY2_ES))]

## Create heatmap body 
df <- t(exprs(ISPY2_ES)[stem_rel,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")
df$KRT5_7_score <- ISPY2_ES$KRT5_7_score[match(df$PID, colnames(ISPY2_ES))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% dplyr::summarize(mean = mean(exprs, na.rm = TRUE))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% dplyr::count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

## S14C

```{r}
BL_sig_up <- GeneSignatures$BL_Gray
BL_sig_up_rel <- BL_sig_up[BL_sig_up %in% rownames(exprs(ISPY2_ES))]

df <- t(exprs(ISPY2_ES)[BL_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- ISPY2_ES$KRT5_7_score[match(rownames(df), colnames(ISPY2_ES))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = col_score)+
  theme_classic()+
  ylab("Basoluminal signature")+
  stat_compare_means()
```
```{r}
BL_sig_up <- GeneSignatures$BL_Gray
BL_sig_up_rel <- BL_sig_up[BL_sig_up %in% rownames(exprs(ISPY2_ES))]

df <- t(exprs(ISPY2_ES[,ISPY2_ES$TNBC == "TNBC"])[BL_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- ISPY2_ES$KRT5_7_score[match(rownames(df), colnames(ISPY2_ES))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = col_score)+
  theme_classic()+
  ylab("Basoluminal signature")+
  stat_compare_means()
```

## S14D

```{r}
df_alluvial <- pData(ISPY2_ES) %>% as.data.frame %>% dplyr::select(KRT5_7_score,GMM_HLADRA,PID_groups, PID_group_risk)
alluvial_wide(df_alluvial, 
              colorful_fill_variable_stratum = FALSE,
              col_vector_flow = c("#85D4E3","grey", "#9C964A", "#F4B5BD"), 
              order_levels = c("KRT5/7+","KRT5/7-","KRT7+","KRT5+", "high", "medium", "low"))+
  theme_classic()
```

## S14E

```{r}
# Define key markers of interest
key_marker <- c("KRT5","KRT7","HLA-DRA")

## Create heatmap body 
df <- t(exprs(ISPY2_ES)[key_marker,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")
df$PID_groups <- ISPY2_ES$PID_groups[match(df$PID, colnames(ISPY2_ES))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

## Create row annotation 
anno <- df %>% select(PID_groups) %>% group_by(PID_groups) %>% dplyr::count()
anno <- anno[match(anno$PID_groups, rownames(ht)),]

## Print heatmap 
Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(anno = anno_text(anno$PID_groups))
```

## S14F 

```{r}
plot <- as.data.frame(t(exprs(ISPY2_ES)[rownames(exprs(ISPY2_ES)) %in% c("KRT5", "KRT7", "KRT18","KRT14","HLA-DRA"),]))
plot$PAM50 <- ISPY2_ES$PAM50.Subtype[match(rownames(plot),colnames(ISPY2_ES))]
plot$TNBC <- ISPY2_ES$TNBC[match(rownames(plot),colnames(ISPY2_ES))]
plot$PID_groups <- ISPY2_ES$PID_groups[match(rownames(plot),colnames(ISPY2_ES))]
group_color <- setNames(c("#3B9AB2", "gold", "pink", "orange","#F21A00"), 
                        c("inflamed","KRT5/7-","KRT5+","KRT7+","KRT5/7+"))

# PID groups 
## PAM50 subtype 
bar_out <- plot %>% as.data.frame() %>% group_by(PID_groups, PAM50) %>% dplyr::count() %>% filter(PAM50 != "NA") %>% ungroup %>% group_by(PAM50) %>% mutate(fra = n/sum(n)) %>% arrange(PAM50)

ggplot(bar_out, aes(x = PAM50, y = fra, fill = PID_groups))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = group_color)+
  theme_classic()

## TNBC 
bar_out <- plot %>% as.data.frame() %>% group_by(PID_groups, TNBC) %>% dplyr::count() %>% ungroup %>% group_by(TNBC) %>% mutate(fra = n/sum(n)) %>% arrange(TNBC)

ggplot(bar_out, aes(x = TNBC, y = fra, fill = PID_groups))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = group_color)+
  theme_classic()
```

## S14G

```{r, fig.height = 10}
# Biomarker association

# Heatmap body 
mat <- t(pData(ISPY2_ES) %>% select(25:50, MP_index) %>% mutate_all(as.numeric))

test <- as.data.frame(t(mat))

anno <- pData(ISPY2_ES) 
test$PID_groups <- anno$PID_groups[match(rownames(test), rownames(anno))]

test_result <- as.data.frame(compare_means(c(ICS5_score, Chemokine12_score, Module5_TcellBcell_score, 
STAT1_sig, Module3_IFN_score, Dendritic_cells, B_cells, 
Mast_cells, Module11_Prolif_score, MP_index, Basal_Index, 
PARPi7_score, PARPi7_plus_MP2, VCpred_TN, STMN1_dat, 
HER2_Index, Mod7_ERBB2, ERBB2.Y1248, EGFR.Y1173, mTOR.S2448, 
IGF1R_dat, TIE2.Y992, Mod10_ECM, RPL24_dat, LYMPHS_PCA_16704732, 
Luminal_Index, ER_PGR_avg)~PID_groups, method = "kruskal.test",data = test))

test_result$p.adj <- p.adjust(test_result$p, method = "BH") 

test_result$psignif.adj <- as.matrix(symnum(test_result$p.adj,
       cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), 
       symbols = c("****", "***", "**", "*", "ns")))[,1]

rownames(test_result) <- test_result$.y.

mat_mean <- test %>% pivot_longer(cols = 1:27, names_to = "cat", values_to = "score") %>% group_by(PID_groups, cat) %>% summarize(mean_score = mean(score, na.rm = TRUE)) %>% pivot_wider(names_from = "cat", values_from = "mean_score") %>% column_to_rownames("PID_groups")

# Row annotation 
highest_cat <- test %>% pivot_longer(cols = 1:27, names_to = "cat", values_to = "score") %>% group_by(PID_groups, cat) %>% summarize(mean_score = mean(score, na.rm = TRUE)) %>% ungroup() %>% group_by(cat) %>% mutate(max = max(mean_score)) %>% filter(mean_score == max) %>% select(cat, PID_groups) %>% as.data.frame %>% column_to_rownames("cat")

ISPY2_BM$PID_groups <- highest_cat$PID_groups[match(rownames(ISPY2_BM), rownames(highest_cat))]

# Column annotation 
col_anno <- anno %>% group_by(PID_groups) %>% count() %>% as.data.frame()
rownames(col_anno) <- col_anno$PID_groups 

# Align order
ISPY2_BM <- ISPY2_BM[match(colnames(mat_mean), rownames(ISPY2_BM)),]
test_result <- test_result[match(colnames(mat_mean), rownames(test_result)),]
col_anno <- col_anno[match(rownames(mat_mean), rownames(col_anno)),]

ca <- columnAnnotation(n_PID = anno_barplot(col_anno$n, gp = gpar(fill = "#440154FF")))

## Visualize 
# Heatmap 
h <- 
  rowAnnotation(bio = anno_text(ISPY2_BM$biomarker))+
  rowAnnotation(path = anno_simple(ISPY2_BM$pathway, col = setNames(brewer.pal(8, "Paired"), unique(ISPY2_BM$pathway)), width = unit(5, "mm")))+
  rowAnnotation(type = anno_simple(ISPY2_BM$type, col = setNames(brewer.pal(3, "Set1")[1:2], unique(ISPY2_BM$type)), width = unit(5, "mm")))+
  Heatmap(t(mat_mean), name = "mean_score", cluster_columns = TRUE, row_names_side = "left", show_row_dend = FALSE, top_annotation = ca)+
  rowAnnotation(cat = anno_simple(unfactor(ISPY2_BM$PID_groups), col = group_color, width = unit(5, "mm")))+
  rowAnnotation(test = anno_text(test_result$psignif.adj))

# Legends
lgd_cat <- Legend(title = "Highest_exprs", at = names(group_color), 
                  legend_gp = gpar(fill = group_color))
lgd_path <- Legend(title = "Pathway", at = names(setNames(brewer.pal(8, "Paired"), unique(ISPY2_BM$pathway))), 
                  legend_gp = gpar(fill = setNames(brewer.pal(8, "Paired"), unique(ISPY2_BM$pathway))))
lgd_type <- Legend(title = "Highest_exprs", at = names(setNames(brewer.pal(3, "Set1")[1:2], unique(ISPY2_BM$type))), 
                  legend_gp = gpar(fill = setNames(brewer.pal(3, "Set1")[1:2], unique(ISPY2_BM$type))))

# FINAL
draw(h,annotation_legend_list = list(lgd_cat, lgd_path, lgd_type))
```

## S14H

```{r}
plot <- pData(ISPY2_ES) %>% as.data.frame %>% select(PID_groups, response)

## General chemotherapy response 
bar_out <- plot %>% as.data.frame() %>% group_by(PID_groups, response) %>% dplyr::count() %>% filter(response != "NA") %>% ungroup %>% group_by(PID_groups) %>% mutate(fra = n/sum(n))
test <- chisq.test(xtabs(n ~ response + PID_groups, data = bar_out))
bar_out$PID_groups <- factor(bar_out$PID_groups, levels = bar_out %>% filter(response == "1") %>% arrange(fra) %>% pull(PID_groups) %>% unfactor())

ggplot(bar_out, aes(x = PID_groups, y = fra, fill = response))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  theme_classic()+
  ggtitle(paste0("p = ", test$p.value))
```

## S14I

```{r}
# Full cohort
ISPY2_ES$arm_short <- ISPY2_ES$`Arm (short name)`
treatment <- pData(ISPY2_ES) %>% select(PID, arm_short, response, PID_groups, TNBC)

out_check <- treatment
out_C <- out_check %>% group_by(PID_groups, arm_short, response) %>% dplyr::count() %>% group_by(PID_groups, arm_short) %>% mutate(fra = n/sum(n))
out_C$response <- factor(out_C$response, levels = c("1", "0"))
out_C$PID_groups <- factor(out_C$PID_groups, levels = c("inflamed", "KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

res_final <- list()
for(i in unique(out_C$PID_groups)){
  data <- out_C %>% mutate(PID_binary = ifelse(PID_groups == i, i, "Other"))
  for(y in unique(out_C$arm_short)){
    data_test <- data %>% filter(arm_short == y)
    if(length(unique(data_test$PID_binary)) > 1){
      res <- fisher.test(xtabs(n ~ PID_binary + response, data = data_test))
      out <- data.frame(odds_ratio = res$estimate, p = res$p.value, arm_short = y, group = i)
      res_final[[paste0(y, i)]] <- out
    }else{
      res_final[[paste0(y, i)]] <- data.frame(odds_ratio = NA, p = NA, arm_short = y, group = i)
    }
  }
}
res_final <- do.call("rbind", res_final)
isGreater <- res_final$odds_ratio > 1 & !is.na(res_final$odds_ratio)
pVal <- log10(res_final$p)
pVal[isGreater] <- abs(pVal[isGreater])
res_final$p_out <- pVal
colnames(res_final) <- c("odds_ratio", "p_value", "arm","category", "p_out")
res_final$result <- ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.05, "UP",
                         ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.1, "UP_trend",
                                ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.05, "DOWN",
                                       ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.1, "DOWN_trend", "NOTSIG"))))
res_final$odds_ratio[res_final$odds_ratio == Inf] <- 1

cur_df <- res_final %>% select(arm, category, odds_ratio) %>% pivot_wider(values_from = odds_ratio, names_from = arm)
res_final$category <- factor(res_final$category, levels = c("inflamed", "KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))
dd <- dist(t(cur_df[2:ncol(cur_df)]), method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
column_sorted <- hc$order
res_final$arm <- factor(res_final$arm, levels = colnames(cur_df[2:ncol(cur_df)][,column_sorted]))

res_final_full <- res_final
res_final_full$subset <- "Full"
```

```{r}
# TNBC subset
ISPY2_ES$arm_short <- ISPY2_ES$`Arm (short name)`
treatment <- pData(ISPY2_ES) %>% select(PID, arm_short, response, PID_groups, TNBC)
treatment <- treatment %>% filter(TNBC == "TNBC")

out_check <- treatment
out_C <- out_check %>% group_by(PID_groups, arm_short, response) %>% dplyr::count() %>% group_by(PID_groups, arm_short) %>% mutate(fra = n/sum(n))
out_C$response <- factor(out_C$response, levels = c("1", "0"))
out_C$PID_groups <- factor(out_C$PID_groups, levels = c("inflamed", "KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))
out_C_TNBC<- out_C

res_final <- list()
for(i in unique(out_C$PID_groups)){
  data <- out_C %>% mutate(PID_binary = ifelse(PID_groups == i, i, "Other"))
  for(y in unique(out_C$arm_short)){
    data_test <- data %>% filter(arm_short == y)
    if(length(unique(data_test$PID_binary)) > 1){
      res <- fisher.test(xtabs(n ~ PID_binary + response, data = data_test))
      out <- data.frame(odds_ratio = res$estimate, p = res$p.value, arm_short = y, group = i)
      res_final[[paste0(y, i)]] <- out
    }else{
      res_final[[paste0(y, i)]] <- data.frame(odds_ratio = NA, p = NA, arm_short = y, group = i)
    }
  }
}
res_final <- do.call("rbind", res_final)
isGreater <- res_final$odds_ratio > 1 & !is.na(res_final$odds_ratio)
pVal <- log10(res_final$p)
pVal[isGreater] <- abs(pVal[isGreater])
res_final$p_out <- pVal
colnames(res_final) <- c("odds_ratio", "p_value", "arm","category", "p_out")
res_final$result <- ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.05, "UP",
                         ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.1, "UP_trend",
                                ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.05, "DOWN",
                                       ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.1, "DOWN_trend", "NOTSIG"))))

res_final$odds_ratio[res_final$odds_ratio == Inf] <- 1

cur_df <- res_final %>% select(arm, category, odds_ratio) %>% pivot_wider(values_from = odds_ratio, names_from = arm)
res_final$category <- factor(res_final$category, levels = c("inflamed", "KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))
dd <- dist(t(cur_df[2:ncol(cur_df)]), method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
column_sorted <- hc$order
res_final$arm <- factor(res_final$arm, levels = colnames(cur_df[2:ncol(cur_df)][,column_sorted]))

res_final_TNBC <- res_final 
res_final_TNBC$subset <- "TNBC"
```

```{r}
# Visualization
res_final_full <- rbind(res_final_TNBC, res_final_full)

ggplot(res_final_full, aes(y=category, x = arm,col = p_out, size = abs(p_out)))+
  facet_wrap(~subset)+
  geom_point()+
  theme_classic()+
  scale_color_gradient2(high = "#E41A1C", mid = "white", low = "#377EB8", midpoint = 0)+
  scale_size(range = c(2,10))+
  new_scale_colour() +
  geom_point(data = res_final_full,shape = 1,aes(size = abs(p_out), col = result), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214", "UP_trend" = "#E41A1C", "NOTSIG" = "snow2","DOWN_trend" = "#377EB8", "DOWN" = "#25567d"), name = "Direction")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Figure S15

## S15A

Pixel and single-cell level IMC images were generated using [cytoviewer](https://github.com/BodenmillerGroup/cytoviewer).

## S15B

```{r}
IMSUP_up <- c("PVR", "CD276", "TGFB2", "PROM1", "TGFB1")
IMSUP_up_rel <- IMSUP_up[IMSUP_up %in% rownames(assay(TCGA_DET))]

df <- t(assay(TCGA_DET)[IMSUP_up_rel,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_group_test <- TCGA_DET$PID_group_test[match(rownames(df), colnames(TCGA_DET))]

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_group_test, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_group_test")
Heatmap(ht, name = "Mean signature Z-scores")
```

## S15C

```{r}
# IFNg
IFNG_sig <- GeneSignatures$INFG_Ayers
IFNG_sig_rel <- IFNG_sig[IFNG_sig %in% rownames(assay(TCGA_DET))]
df <- t(assay(TCGA_DET)[IFNG_sig_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_group_test <- TCGA_DET$PID_group_test[match(rownames(df), colnames(TCGA_DET))]
mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_group_test, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
INFg <- mat %>% group_by(PID_group_test) %>% summarize(INFg = mean(mean)) %>% column_to_rownames("PID_group_test")

# IFNab
INFab_sig_up <- GeneSignatures$INFAB_Reactome
INFab_sig_rel <- INFab_sig_up[INFab_sig_up %in% rownames(assay(TCGA_DET))]
df <- t(assay(TCGA_DET)[INFab_sig_rel  ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_group_test <- TCGA_DET$PID_group_test[match(rownames(df), colnames(TCGA_DET))]
mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_group_test, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
INFab <- mat %>% group_by(PID_group_test) %>% summarize(INFab = mean(mean)) %>% column_to_rownames("PID_group_test")

# CTantigen
CTA_sig_up <- GeneSignatures$CTA_PNAS
CTA_sig_rel <- CTA_sig_up[CTA_sig_up %in% rownames(assay(TCGA_DET))]
df <- t(assay(TCGA_DET)[CTA_sig_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_group_test <- TCGA_DET$PID_group_test[match(rownames(df), colnames(TCGA_DET))]
mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_group_test, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
CTA <- mat %>% group_by(PID_group_test) %>% summarize(CTA = mean(mean)) %>% column_to_rownames("PID_group_test")

## Combined visualization
Heatmap(CTA, name = "CTA", cluster_rows = TRUE)+
  Heatmap(INFab, name = "INFab", cluster_rows = FALSE)+
  Heatmap(INFg, name = "INFg", cluster_rows = FALSE)
```

## S15D

```{r}
key_marker <- c("TGFB2", "TGFB1", "PROM1", "PVRL1")
key_marker <- key_marker[key_marker %in% rownames(MDACC_EXP)]

## Create heatmap body 
df <- t(MDACC_EXP[key_marker,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")
df$PID_groups <- MDACC_ES$PID_groups[match(df$PID, colnames(MDACC_ES))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

## Create row annotation 
anno <- df %>% select(PID_groups) %>% group_by(PID_groups) %>% dplyr::count()
anno <- anno[match(anno$PID_groups, rownames(ht)),]

## Print heatmap 
Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(anno = anno_text(anno$PID_groups))
```

## S15E

```{r}
# IFNg
INFg_sig_up <- GeneSignatures$INFG_Ayers
INFg_sig_up_rel <- INFg_sig_up[INFg_sig_up %in% rownames(MDACC_EXP)]
df <- t(MDACC_EXP[INFg_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_groups <- MDACC_ES$PID_groups[match(rownames(df), colnames(MDACC_ES))]
mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
INFg <- mat %>% group_by(PID_groups) %>% summarize(INFg = mean(mean)) %>% column_to_rownames("PID_groups")

#IFNab
INFab_sig_up <- GeneSignatures$INFAB_Reactome
INFab_sig_up_rel <- INFab_sig_up[INFab_sig_up %in% rownames(MDACC_EXP)]
df <- t(MDACC_EXP[INFab_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_groups <- MDACC_ES$PID_groups[match(rownames(df), colnames(MDACC_ES))]
mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
INFab <- mat %>% group_by(PID_groups) %>% summarize(INFab = mean(mean)) %>% column_to_rownames("PID_groups")

#CTA
CTA_sig_up <- GeneSignatures$CTA_PNAS
CTA_sig_up_rel <- CTA_sig_up[CTA_sig_up %in% rownames(MDACC_EXP)]
df <- t(MDACC_EXP[CTA_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_groups <- MDACC_ES$PID_groups[match(rownames(df), colnames(MDACC_ES))]
mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
CTA <- mat %>% group_by(PID_groups) %>% summarize(CTA = mean(mean)) %>% column_to_rownames("PID_groups")

##Combined visualization
Heatmap(CTA, name = "CTA", cluster_rows = TRUE)+
  Heatmap(INFab, name = "INFab", cluster_rows = FALSE)+
  Heatmap(INFg, name = "INFg", cluster_rows = FALSE)
```

## S15F

```{r}
# Define key immune markers of interest
key_marker <- c("TGFB2", "TGFB1", "PROM1", "PVR", "PVRL1","CD276")
key_marker <- key_marker[key_marker %in% rownames(exprs(ISPY2_ES))]

## Create heatmap body 
df <- t(exprs(ISPY2_ES)[key_marker,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")
df$PID_groups <- ISPY2_ES$PID_groups[match(df$PID, colnames(ISPY2_ES))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

## Create row annotation 
anno <- df %>% select(PID_groups) %>% group_by(PID_groups) %>% dplyr::count()
anno <- anno[match(anno$PID_groups, rownames(ht)),]

## Print heatmap 
Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(anno = anno_text(anno$PID_groups))
```

## S15G

```{r}
#IFNg
INFg_sig_up <- GeneSignatures$INFG_Ayers
INFg_sig_up_rel <- INFg_sig_up[INFg_sig_up %in% rownames(exprs(ISPY2_ES))]
df <- t(exprs(ISPY2_ES)[INFg_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_groups <- ISPY2_ES$PID_groups[match(rownames(df), colnames(ISPY2_ES))]
mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
INFg <- mat %>% group_by(PID_groups) %>% summarize(INFg = mean(mean)) %>% column_to_rownames("PID_groups")

#IFNab
INFab_sig_up <- GeneSignatures$INFAB_Reactome
INFab_sig_up_rel <- INFab_sig_up[INFab_sig_up %in% rownames(exprs(ISPY2_ES))]
df <- t(exprs(ISPY2_ES)[INFab_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_groups <- ISPY2_ES$PID_groups[match(rownames(df), colnames(ISPY2_ES))]
mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")
INFab <- mat %>% group_by(PID_groups) %>% summarize(INFab = mean(mean)) %>% column_to_rownames("PID_groups")

#CTA
## 2. BL signature per PID group 
CTA_sig_up <- GeneSignatures$CTA_PNAS
CTA_sig_up_rel <- CTA_sig_up[CTA_sig_up %in% rownames(exprs(ISPY2_ES))]
df <- t(exprs(ISPY2_ES)[CTA_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))
df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_groups <- ISPY2_ES$PID_groups[match(rownames(df), colnames(ISPY2_ES))]
mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
CTA <- mat %>% group_by(PID_groups) %>% summarize(CTA = mean(mean)) %>% column_to_rownames("PID_groups")

##Combined visualization
Heatmap(CTA, name = "CTA", cluster_rows = TRUE)+
  Heatmap(INFab, name = "INFab", cluster_rows = FALSE)+
  Heatmap(INFg, name = "INFg", cluster_rows = FALSE)
```

# Session Info 

```{r}
sessionInfo()
```