---
title: null
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

# Figure 1 and Figure S1

Data analysis scripts for re-creating Figure 1 and Figure S1. 

# Software requirements

```{r, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(tidyverse)
library(ComplexHeatmap)
library(RColorBrewer)
library(scater)
library(viridis)
```

# Read data

```{r}
# Available from Zenodo - DOI: 10.5281/zenodo.15396067
# File paths have to be adjusted by the user

## SCE objects
sce_ALL_sub <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_ALL_sub.rds")
sce_ALL <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_ALL.rds")

## Other
Meta_viz <- read.csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/TNBC_discovery_metadata.csv")
```

# Figure 1

## 1B

```{r}
# Arrange
Meta_viz[is.na(Meta_viz)] <- "NA"
Meta_viz <- Meta_viz %>% arrange(desc(pT_simple),desc(pN_simple), desc(grade))

# Colors
color_vectors <- list(
TMANr = setNames(colorRampPalette(brewer.pal(3, "Dark2"))(length(unique(Meta_viz$TMANr))), 
                sort(unique(Meta_viz$TMANr))),
PID = setNames(colorRampPalette(brewer.pal(12, "Paired"))(length(unique(Meta_viz$PID))), 
                sort(unique(Meta_viz$PID))),
n_samples = setNames(colorRampPalette(brewer.pal(3, "Greens"))(length(unique(Meta_viz$n_samples))), 
                sort(unique(Meta_viz$n_samples))),
grade = setNames(c(colorRampPalette(brewer.pal(3, "Oranges"))(length(unique(Meta_viz$grade))-1),"black"), 
                sort(unique(Meta_viz$grade))),
pT_simple = setNames(c(colorRampPalette(brewer.pal(9, "Blues"))(length(unique(Meta_viz$pT_simple))-1),"black"), 
                sort(unique(Meta_viz$pT_simple), na.last = TRUE)),
pN_simple = setNames(c(colorRampPalette(brewer.pal(9, "Purples"))(length(unique(Meta_viz$pN_simple))-1),"black"), 
                sort(unique(Meta_viz$pN_simple))),
pM_simple = setNames(c(colorRampPalette(brewer.pal(3, "PuRd"))(length(unique(Meta_viz$pM_simple))-1),"black"), 
                sort(unique(Meta_viz$pM_simple))),
pretreatment = setNames(c(colorRampPalette(brewer.pal(4, "Set3")[3:4])(length(unique(Meta_viz$pretreatment))-1),"black"), 
                c("FALSE","TRUE","NA")),
OS_data = setNames(c("black","grey"),sort(unique(Meta_viz$OS_data))),
DFS_data = setNames(c("black","grey"),sort(unique(Meta_viz$DFS_data))),
age_bin = setNames(c(colorRampPalette(brewer.pal(11, "BrBG"))(length(unique(Meta_viz$age_bin))-1),"black"), 
                sort(unique(Meta_viz$age_bin)))
)

# Heatmap Annotation
sample_anno <- HeatmapAnnotation(TMANr = Meta_viz$TMANr,
                                 PID = Meta_viz$PID,
                                 age_bin = Meta_viz$age_bin,
                                 n_samples = Meta_viz$n_samples,
                                 grade = Meta_viz$grade,
                                 pT_simple = Meta_viz$pT_simple,
                                pN_simple = Meta_viz$pN_simple,
                                pM_simple = Meta_viz$pM_simple,
                                pretreatment = Meta_viz$pretreatment,
                                OS_data = Meta_viz$OS_data,
                                DFS_data = Meta_viz$DFS_data,
                                col = color_vectors,
                                border = T,
                                na_col = "black",
                                gap = unit(5, "points"),
                                show_annotation_name = T,
                                show_legend = c(PID = FALSE)
)%v%NULL

# Display plot
draw(sample_anno)
```

## 1D

Pixel and single-cell level IMC images were generated using [cytoviewer](https://github.com/BodenmillerGroup/cytoviewer).

# Figure S1

## S1A

```{r}
# Plot TSNE
plotReducedDim(sce_ALL_sub, dimred = "TSNE", colour_by = "TMANr", point_size = 1)+
  scale_color_manual(values = metadata(sce_ALL_sub)$colors$TMANr)

plotReducedDim(sce_ALL_sub, dimred = "TSNE", colour_by = "PID", point_size = 1)+
  scale_color_manual(values = metadata(sce_ALL_sub)$colors$PID)+
  guides(color = "none")

plotReducedDim(sce_ALL_sub, dimred = "TSNE", colour_by = "pretreatment", point_size = 1)+
  scale_color_manual(values = metadata(sce_ALL_sub)$colors$pretreatment)
```

## S1B

```{r}
plotReducedDim(sce_ALL_sub, dimred = "TSNE", colour_by = "cluster_category", point_size = 1)+
  scale_color_manual(values = metadata(sce_ALL_sub)$colors$cluster_category)
```

## S1C/D

```{r, fig.width = 9}
### 1. Heatmap body ###

# Aggregate across cells
mean_sce <- aggregateAcrossCells(sce_ALL, ids = sce_ALL$cluster_broad,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce_ALL)$use_channel)

# Heatmap body color 
col_main <- viridis(100)

### 2. Heatmap annotation ###

### 2.1 Spatial features

#Select spatial features and average over clusters 
spatial <- colData(sce_ALL) %>% as.data.frame %>% select(area,eccentricity,cluster_broad,n_neighbors)
spatial <- spatial %>% select(-cluster_broad) %>% aggregate(by = list(cluster_broad = spatial$cluster_broad), FUN = mean) %>% column_to_rownames("cluster_broad")

#Create Spatial Heatmap Annotation
ha_spatial <- HeatmapAnnotation(
    area = spatial$area,
    eccentricity = spatial$eccentricity,
    n_neighbors = spatial$n_neighbors, 
    col = list(area = metadata(sce_ALL)$colors$area,
               eccentricity = metadata(sce_ALL)$colors$eccentricity,
               n_neighbors = metadata(sce_ALL)$colors$n_neighbors),
    border = TRUE,
    which = "row")

### 2.2 Metadata features

#Number of cells per cluster_broad 
n_cells <- colData(mean_sce) %>% as.data.frame %>% select(cluster_broad, ncells)

# GMM - panCK
nr_tum <- colData(sce_ALL) %>% as.data.frame %>% select(cluster_broad, GMM_CK) %>% table()
fr_tum <- as.data.frame(nr_tum) %>% pivot_wider(names_from = GMM_CK, values_from = Freq) %>% column_to_rownames("cluster_broad")
fr_tum <- fr_tum %>% transmute(fr_tum = `panCK+`/rowSums(fr_tum)) %>% as.data.frame()

# Astir
nr_tum_2 <- colData(sce_ALL) %>% as.data.frame %>% select(cluster_broad, astir_type) %>% table()
fr_tum_2 <- as.data.frame(nr_tum_2) %>% pivot_wider(names_from = astir_type, values_from = Freq) %>% column_to_rownames("cluster_broad")
fr_tum_2 <- fr_tum_2 %>% transmute(fr_tum = `Epithelial`/rowSums(fr_tum_2)) %>% as.data.frame()

# GMM - Ecad
nr_tum_3 <- colData(sce_ALL) %>% as.data.frame %>% select(cluster_broad, GMM_Ecad) %>% table()
fr_tum_3 <- as.data.frame(nr_tum_3) %>% pivot_wider(names_from = GMM_Ecad, values_from = Freq) %>% column_to_rownames("cluster_broad")
fr_tum_3 <- fr_tum_3 %>% transmute(fr_tum_3 = `Ecad+`/rowSums(fr_tum_3)) %>% as.data.frame()

ha_meta <- HeatmapAnnotation(GMM_CK = anno_barplot(fr_tum, width = unit(15, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            GMM_Ecad = anno_barplot(fr_tum_3, width = unit(15, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            Astir_Epi = anno_barplot(fr_tum_2, width = unit(15, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            n_cells = anno_barplot(n_cells$ncells, width = unit(15, "mm"),gp = gpar(fill = "violet", col = 1)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(1,"mm"),
                            which = "row")

### 3. Define cluster_broad category
anno_cluscat <- fr_tum %>% 
  rownames_to_column(var = "cluster_broad") %>% 
  mutate(cluster_category = ifelse(fr_tum >= 0.7,"Epithelial",ifelse(fr_tum < 0.7 & fr_tum > 0.2,"Mixed","Non_Epithelial")))


ha_cluscat <- HeatmapAnnotation(cluster_category = anno_cluscat$cluster_category,
                                col = list(cluster_category = metadata(sce_ALL)$colors$cluster_category),
                                border = TRUE,
                                which = "row")



### 4. Plot combined Heatmap - Figure S1C
h <- Heatmap(t(assay(mean_sce, "min_max")),
        col = col_main,
        name= "min-max",
        km = 1,
        show_row_names = T, 
        show_column_names =  T, row_split = anno_cluscat$cluster_category,
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")+
  ha_spatial+
  ha_meta+
  ha_cluscat

draw(h)

# Visualize GMM_panCK density plots - Figure S1D
ggplot(anno_cluscat, aes(fr_tum))+
  geom_density()+
  geom_rug(aes(color = cluster_category))+
  geom_vline(xintercept = 0.7, color = "red3")+
  geom_vline(xintercept = 0.2, color = "red3")+
  theme_classic()+
  theme(axis.title.x = element_blank())+
  scale_color_manual(values = metadata(sce_ALL)$colors$cluster_category)
```

## S1E

```{r}
plotReducedDim(sce_ALL_sub, dimred = "TSNE", colour_by = "analysis_cat", point_size = 1)+
  scale_color_manual(values = metadata(sce_ALL_sub)$colors$analysis_cat)

plotReducedDim(sce_ALL_sub, dimred = "TSNE", colour_by = "panCK", by_exprs_values = "exprs", point_size = 1)
plotReducedDim(sce_ALL_sub, dimred = "TSNE", colour_by = "Ecad", by_exprs_values = "exprs", point_size = 1)
```
## S1F 

Pixel and single-cell level IMC images were generated using [cytoviewer](https://github.com/BodenmillerGroup/cytoviewer).

# Session Info 

```{r}
sessionInfo()
```