---
title: null
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

# Figure 4 and Figure S5/6

Data analysis scripts for re-creating Figure 4 and Figure S5/6. 

# Software requirements

```{r, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(tidyverse)
library(ComplexHeatmap)
library(RColorBrewer)
library(conflicted)
library(survival)
library(survminer)
library(imcRtools)
library(BiocParallel)
library(wesanderson)
library(dittoSeq)
library(scater)
library(patchwork)
```

```{r, message=FALSE,warning=FALSE}
conflicts_prefer(dplyr::desc)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::count)
conflicts_prefer(base::intersect)
conflicts_prefer(dplyr::select)
```

# Read data

```{r}
# Available from Zenodo - DOI: 10.5281/zenodo.15396067
# File paths have to be adjusted by the user

## SCE objects
sce_ALL <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_ALL.rds")
sce_fischer_TNBC <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_fischer_TNBC.rds")
sce_fischer_ALL <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_fischer_ALL.rds")
sce_PalWu <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_PalWu.rds")
sce_PalWu_ALL <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_PalWu_ALL.rds")
sce_organoid <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/SCE/sce_organoid.rds")

## Other
INT_MC8 <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/INT_MC8.rds")
INT_fischer_MC8 <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/INT_fischer_MC8.rds")
IF <- read_csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/IF.csv")
TCGA_DET <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/TCGA_DET.rds")
MDACC_ES <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/MDACC_ES.rds")
MDACC_EXP <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/MDACC_EXP.rds")
```

# Figure 4

## 4A

Pixel and single-cell level IMC images were generated using [cytoviewer](https://github.com/BodenmillerGroup/cytoviewer).

## 4B

```{r}
patch_MC8_PID <- colData(sce_ALL) %>% as.data.frame() %>% filter(!is.na(patch_tumor_8)) %>% pull(PID) %>% unique()
cells_MC8_PID <- colData(sce_ALL) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_patch <- colData(sce_ALL) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_patch <- PID_patch %>% group_by(PID) %>% filter(PID %in% patch_MC8_PID) %>% filter(str_detect(metacluster_combined,"tumor")) %>% mutate(fra = Freq/sum(Freq)) 

PID_dom_MC8 <- PID_patch %>% mutate(max = max(fra)) %>% filter(fra == max) %>% ungroup() %>% filter(metacluster_combined == "tumor_8") %>% pull(PID)

PID_overall <- colData(sce_ALL) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_MC8_overall <- PID_overall %>% group_by(PID) %>% filter(metacluster_combined == "tumor_8" & Freq > 5) %>% mutate(patch = ifelse(PID %in% patch_MC8_PID, TRUE, FALSE)) %>% mutate(dominated = ifelse(PID %in% PID_dom_MC8, TRUE, FALSE)) %>% mutate(PID_category = ifelse(patch && dominated, "dominated", ifelse(patch, "patch", "cells")))
PID_MC8_overall_plot <- PID_MC8_overall %>% group_by(PID_category) %>% count() %>% ungroup() %>% mutate(fra = n/sum(n)) %>% mutate(category = "MC8")
PID_MC8_overall_plot$PID_category <- factor(PID_MC8_overall_plot$PID_category, levels = c("cells","patch","dominated"))

ggplot(PID_MC8_overall_plot) +
  geom_tile(color = "black",aes(category, n, fill = PID_category), position = "stack")+
  scale_fill_brewer(palette = "Blues")+
  theme_classic(base_size = 15)+
  ylab("Density")+
  coord_flip()
```

## 4C

```{r}
PID_patch_abundant <- PID_patch %>%
  mutate(max = max(fra)) %>% filter(fra == max) %>% ungroup() %>% count(metacluster_combined) %>% mutate(fra = n/sum(n), category = "cat") %>% arrange(fra)

PID_patch_abundant$metacluster_combined <- factor(PID_patch_abundant$metacluster_combined, levels = unfactor(PID_patch_abundant$metacluster_combined))

ggplot(PID_patch_abundant) +
  geom_tile(color = "black",aes(category, n, fill = metacluster_combined), position = "stack")+
  scale_fill_manual(values=metadata(sce_ALL)$colors$metacluster_combined) + 
  theme_classic(base_size = 15)+
  ylab("Density")+
  xlab("PID")+
  coord_flip()
```

## 4D

```{r}
# Interaction testing 
## Visualization
interaction_tumor_plot <- INT_MC8 %>% as_tibble() %>% filter(from_label %in% c("tumor_8")) %>% filter(to_label != "tumor_8") %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

ggplot(interaction_tumor_plot) +
        geom_tile(aes(reorder(to_label,sum_sigval), from_label, fill = sum_sigval)) +
        scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_classic()
```

## 4F

```{r}
cur_IF <- IF %>% select("CK7","CK7:CK19","CK19","CK5","CK5:CK7","CK5:CK7:CK19","CK5:CK19", "PID") %>% pivot_longer(1:7,values_to = "n", names_to = "CK_group")

# Focus on CK5/7 
cur_in <- cur_IF %>% ungroup() %>% mutate(CK5_7_group = ifelse(CK_group %in% c("CK5","CK5:CK19"),"CK5+", 
                                         ifelse(CK_group %in% c("CK7","CK7:CK19"),"CK7+", 
                                                ifelse(CK_group %in% c("CK5:CK7","CK5:CK7:CK19"),"CK5/7+", NA)))) %>% filter(!is.na(CK5_7_group))

cur_in <- cur_in %>% group_by(PID, CK_group) %>% mutate(mean_n = mean(n)) %>% 
  group_by(PID, CK5_7_group) %>% mutate(total_n = sum(mean_n)) 

cur_in <- cur_in %>% select(PID, CK5_7_group, total_n) %>% unique() %>% group_by(PID) %>% mutate(fra = total_n/sum(total_n))

cur_CK <- cur_in %>% pivot_wider(id_cols = "PID", names_from = "CK5_7_group",values_from = "fra") %>% column_to_rownames("PID")

# Define annotation
row_sorted_fra <- cur_in %>% filter(CK5_7_group == "CK5/7+") %>% arrange(desc(fra)) %>% pull(PID)
cur_CK <- cur_CK[row_sorted_fra,]
cur_CK <- cur_CK[,c("CK5/7+", "CK5+", "CK7+")]

anno <- cur_in %>% pivot_wider(id_cols = "PID", names_from = "CK5_7_group", values_from = "total_n") %>% column_to_rownames("PID")
anno <- anno[match(rownames(cur_CK), rownames(anno)),]
anno <- anno[,match(colnames(cur_CK), colnames(anno))]

cur_color_CK <- setNames(wes_palette("Moonrise3")[1:3], c("CK5/7+", "CK5+", "CK7+"))

# Plot
fraction_anno <- HeatmapAnnotation(PID = anno_text(rownames(anno)), 
                                   CK_fractions = anno_barplot(cur_CK, gp=gpar(fill=cur_color_CK),
                                                               bar_width = 1, height = unit(25,"cm"), width = unit(11,"cm"),
                                                               show_row_names = TRUE, row_names_side = "left"),
                                   total_n = anno_barplot(rowSums(anno), border = TRUE, gp = gpar(fill = "#440154FF"), width = unit(3,"cm")),
                                   gap = unit(1,"mm"),
                                   which = "row")
h_1 <- fraction_anno + NULL

lgd_1 <- Legend(title = "CK5_7_group", at = colnames(cur_CK), 
              legend_gp = gpar(fill = cur_color_CK))
             
draw(h_1,annotation_legend_list = list(lgd_1))
```

## 4G

```{r}
# IMC fraction
PID_CK57 <- colData(sce_ALL) %>% as.data.frame() %>% select(CK5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_CK57 <- PID_CK57 %>% group_by(PID) %>% filter(CK5_7_score != "CK5/7-") %>% mutate(fra = Freq/sum(Freq))
PID_CK57_df <- PID_CK57 %>% pivot_wider(id_cols = "PID",names_from = "CK5_7_score",values_from = "fra") %>% column_to_rownames("PID")

# IF fraction 
cur_CK_out <- cur_CK
rownames(cur_CK_out) <- str_replace(rownames(cur_CK_out),"_",".")

# Combined
cur_cor <- PID_CK57_df[rownames(PID_CK57_df) %in% rownames(cur_CK_out),]
cur_cor <- cur_cor %>% select("CK5/7+")
cur_cor$`CK5/7+_IF` <- cur_CK_out$`CK5/7+`[match(rownames(cur_cor), rownames(cur_CK_out))]
cur_cor$PID <- rownames(cur_cor)

spearman_test <- cor.test(cur_cor[,1],cur_cor[,2], method = "spearman")

ggplot(cur_cor, aes(x = `CK5/7+`, y = `CK5/7+_IF`))+
  geom_smooth(method='lm', formula= y~x, se = TRUE, color = "darkgrey")+
  geom_point(aes(color = PID), size = 3)+
  ylab("CK5/7 Fraction IF")+
  xlab("CK5/7 Fraction IMC")+
  theme_classic()+
  scale_color_manual(values=metadata(sce_ALL)$colors$PID)+
  ggtitle(paste("Spearman rho = ",round(spearman_test$estimate, digits = 2),"/ p =",round(spearman_test$p.value, digits = 4)))
```

## 4H

```{r}
cur_BL <- cur_IF %>% filter(CK_group %in% c("CK5:CK7","CK5:CK7:CK19", "CK5:CK19")) %>% group_by(PID, CK_group) %>%
  mutate(mean_n = mean(n)) %>% group_by(PID, CK_group) %>% mutate(total_n = sum(mean_n)) %>% select(PID, CK_group, total_n) %>% unique() %>% group_by(PID) %>% mutate(fra = total_n/sum(total_n))
  
ggplot(cur_BL, aes(x=CK_group, y=fra))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce_ALL)$colors$PID)+
  theme_classic()+
  ylab("Fraction of basoluminal cells per PID")+
  stat_compare_means(method = "kruskal.test", paired = TRUE)
```

# Figure S5

## S5A/B

```{r}
for(x in c("patch_tumor_3", "patch_tumor_8")){
patch_PID <- unique(sce_ALL$PID[!is.na(colData(sce_ALL)[,x])])
colData(sce_ALL)[paste0(x, "_binary")] <- ifelse(sce_ALL$PID %in% patch_PID,1,0)
}
patch_combined <- colData(sce_ALL) %>% as.data.frame() %>% select(patch_tumor_3_binary, patch_tumor_8_binary, PID) %>% unique() %>% as.data.frame %>% remove_rownames() %>% column_to_rownames("PID") %>% mutate(sum = rowSums(.)) %>% mutate(patch_tumor_3or8_binary = ifelse(sum > 0, 1, 0)) 
sce_ALL$patch_tumor_3or8_binary <- patch_combined$patch_tumor_3or8_binary[match(sce_ALL$PID, rownames(patch_combined))]

# Kaplan-Meier
cur_surv <- colData(sce_ALL) %>% as.data.frame() %>% select(PID, patch_tumor_8_binary, patch_tumor_3_binary, patch_tumor_3or8_binary, OS_months, DFS_months, status_DFS, status_OS) %>% unique() 

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ patch_tumor_8_binary, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c("#377EB8", "#E41A1C")
)

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ patch_tumor_3_binary, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c("#377EB8", "#E41A1C")
          )

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ patch_tumor_3or8_binary, data = cur_surv)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c("#377EB8", "#E41A1C")
)
```

## S5C

```{r}
patch_MC8_PID <- colData(sce_fischer_TNBC) %>% as.data.frame() %>% filter(!is.na(patch_MC8)) %>% pull(PID) %>% unique()
cells_MC8_PID <- colData(sce_fischer_TNBC) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(PID) %>% table() %>% as.data.frame

sce_fischer_TNBC$patch_MC8_classification <- ifelse(sce_fischer_TNBC$PID %in% patch_MC8_PID,"patch_MC8","no_patch_MC8")

PID_patch <- colData(sce_fischer_TNBC) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_patch <- PID_patch %>% group_by(PID) %>% filter(PID %in% patch_MC8_PID) %>% filter(str_detect(metacluster_predict,"tumor")) %>% mutate(fra = Freq/sum(Freq)) 
```

```{r}
PID_dom_MC8 <- PID_patch %>% mutate(max = max(fra)) %>% filter(fra == max) %>% ungroup() %>% filter(metacluster_predict == "tumor_8") %>% pull(PID)
PID_overall <- colData(sce_fischer_TNBC) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_MC8_overall <- PID_overall %>% group_by(PID) %>% filter(metacluster_predict == "tumor_8" & Freq > 5) %>% mutate(patch = PID %in% patch_MC8_PID, dominated = PID %in% PID_dom_MC8, PID_category = ifelse(patch & dominated, "dominated",ifelse(patch, "patch", "cells")))               
PID_MC8_overall_plot <- PID_MC8_overall %>% group_by(PID_category) %>% count() %>% ungroup() %>% mutate(fra = n/sum(n)) %>% mutate(category = "MC8")
PID_MC8_overall_plot$PID_category <- factor(PID_MC8_overall_plot$PID_category, levels = c("cells","patch","dominated"))

ggplot(PID_MC8_overall_plot) +
  geom_tile(color = "black",aes(category, n, fill = PID_category), position = "stack")+
  scale_fill_brewer(palette = "Blues")+
  theme_classic(base_size = 15)+
  ylab("Number of patients")+
  theme(axis.title.x = element_blank())
```

## S5D

```{r}
PID_patch_abundant <- PID_patch %>%
  mutate(max = max(fra)) %>% filter(fra == max) %>% ungroup() %>% count(metacluster_predict) %>% mutate(fra = n/sum(n), category = "cat") %>% arrange(fra)

PID_patch_abundant$metacluster_predict <- factor(PID_patch_abundant$metacluster_predict, levels = unfactor(PID_patch_abundant$metacluster_predict))

ggplot(PID_patch_abundant) +
  geom_tile(color = "black",aes(category, n, fill = metacluster_predict), position = "stack")+
  scale_fill_manual(values=metadata(sce_fischer_TNBC)$colors$metacluster_combined) + 
  theme_classic(base_size = 15)+
  ylab("Density")+
  xlab("PID")+
  theme(axis.text.x = element_blank())
```

## S5E

```{r}
# Interaction testing
## Visualization

interaction_tumor_plot <- INT_fischer_MC8 %>% as_tibble() %>% filter(from_label %in% c("tumor_8")) %>% filter(to_label != "tumor_8") %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

ggplot(interaction_tumor_plot) +
        geom_tile(aes(reorder(to_label,sum_sigval), from_label, fill = sum_sigval)) +
        scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_classic()
```

## S5F

Pixel and single-cell level IMC images were generated using [cytoviewer](https://github.com/BodenmillerGroup/cytoviewer).

## S5G

```{r}
# Define keratins of interest
keratins <- c("KRT5","KRT7","KRT15","KRT8","KRT18","KRT19", "KRT14","KRT17","ITGA6","EPCAM")

## Create heatmap body 
df <- t(assay(TCGA_DET)[keratins,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")
df$KRT5_7_score <- TCGA_DET$KRT5_7_score[match(df$PID, colnames(TCGA_DET))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

## S5H

```{r}
# Define keratins of interest
keratins <- c("KRT5","KRT7","KRT15","KRT8","KRT18","KRT19", "KRT14","KRT17","ITGA6","EPCAM")
keratins_rel <- keratins[keratins %in% rownames(MDACC_EXP)]

## Create heatmap body 
df <- t(MDACC_EXP[keratins_rel,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")
df$KRT5_7_score <- MDACC_ES$KRT5_7_score[match(df$PID, colnames(MDACC_ES))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% dplyr::count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

## S5I

```{r}
dittoBarPlot(TCGA_DET, var = "KRT5_7_score", group.by = "PAM50Call_RNAseq")+
    scale_fill_manual(values = metadata(TCGA_DET)$colors$KRT57_score)

dittoBarPlot(TCGA_DET, var = "KRT5_7_score", group.by = "TNBC")+
    scale_fill_manual(values = metadata(TCGA_DET)$colors$KRT57_score)
```

## S5J

```{r}
MDACC_ES$TNBC <- MDACC_ES$TNBC_assign_lehmann
MDACC_ES$PAM50 <- MDACC_ES$`pam50_class:ch1`
col_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("KRT5/7-", "KRT5/7+","KRT5+", "KRT7+"))

bar_out <- pData(MDACC_ES) %>% as.data.frame() %>% group_by(KRT5_7_score, PAM50) %>% dplyr::count() %>% ungroup %>% group_by(PAM50) %>% mutate(fra = n/sum(n)) %>% arrange(PAM50)

ggplot(bar_out, aes(x = PAM50, y = fra, fill = KRT5_7_score))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = col_score)+
  theme_classic()

bar_out <- pData(MDACC_ES) %>% as.data.frame() %>% group_by(KRT5_7_score, TNBC) %>% dplyr::count() %>% ungroup %>% group_by(TNBC) %>% mutate(fra = n/sum(n)) %>% arrange(TNBC)

ggplot(bar_out, aes(x = TNBC, y = fra, fill = KRT5_7_score))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = col_score)+
  theme_classic()
```

## S5K

```{r}
fischer_over <- colData(sce_fischer_ALL) %>% as.data.frame %>% select(PID, PAM50_clean, TNBC, CK5_7_group_name) %>% unique() %>% remove_rownames()
fischer_PAM50_group <- fischer_over %>% group_by(CK5_7_group_name, PAM50_clean) %>% count() %>% ungroup() %>% group_by(PAM50_clean) %>% mutate(fra = n/sum(n)) %>% mutate(dataset = "fischer")

ggplot(fischer_PAM50_group, aes(x = PAM50_clean, y = fra, fill = CK5_7_group_name))+
  facet_wrap(~dataset)+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = c("grey","#85D4E3","#F4B5BD","#9C964A"))+
  theme(axis.text.x = element_text(angle = 90))

fischer_TNBC_group <- fischer_over %>% group_by(CK5_7_group_name, TNBC) %>% count() %>% ungroup() %>% group_by(TNBC) %>% mutate(fra = n/sum(n)) %>% mutate(dataset = "fischer")

ggplot(fischer_TNBC_group, aes(x = TNBC, y = fra, fill = CK5_7_group_name))+
  facet_wrap(~dataset)+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = c("grey","#85D4E3","#F4B5BD","#9C964A"))+
  theme(axis.text.x = element_text(angle = 90))
```

# Figure S6

## S6A/B

```{r, fig.width=10}
plotReducedDim(sce_PalWu, dimred="TSNE", colour_by="dataset", point_size = 1)+
  scale_color_manual(values = c("#1F78B4", "#B2DF8A"))+

plotReducedDim(sce_PalWu, dimred="TSNE", colour_by="subtype", point_size = 1)+
  scale_color_manual(values = c("#4DAF4A", "#377EB8", "#E41A1C"))


plotReducedDim(sce_PalWu, dimred="TSNE", colour_by="KRT5_7_score", point_size = 1)+
  scale_color_manual(values = metadata(sce_PalWu)$colors$KRT57_score)+
plotReducedDim(sce_PalWu, dimred = "TSNE", 
               colour_by = "KRT5", 
               by_exprs_values = "logcounts", point_size = 1)+
plotReducedDim(sce_PalWu, dimred = "TSNE", 
               colour_by = "KRT7", 
               by_exprs_values = "logcounts", point_size = 1)
```

## S6C

```{r}
PID_KRT5_7_score <- colData(sce_PalWu) %>% as.data.frame() %>% dplyr::select(KRT5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_KRT5_7_score <- PID_KRT5_7_score %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
cur_df <- PID_KRT5_7_score %>% pivot_wider(id_cols = "PID", names_from = "KRT5_7_score",values_from = "fra") %>% column_to_rownames("PID")

# Perform hierarchical clustering on fractions
dd <- dist(cur_df, method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
row_sorted <- hc$order

# Define annotation dataframe
cur_df <- cur_df[row_sorted,]
anno <- colData(sce_PalWu) %>% as.data.frame %>% dplyr::select(PID, TNBC, subtype, dataset) %>% unique()
anno <- anno[match(rownames(cur_df), anno$PID),]
n_cells_PID <- PID_KRT5_7_score %>% group_by(PID) %>% summarize(n_cells = sum(Freq))  
anno$ncells <- n_cells_PID$n_cells[match(anno$PID, n_cells_PID$PID)]

# Define colors 
sub_col <- setNames(rev(brewer.pal(3, "Set1")), unique(sce_PalWu$subtype))
dataset_col <- setNames(brewer.pal(3, "Paired")[2:3], unique(sce_PalWu$dataset))

# Plot annotation dataframe
fraction_anno <- HeatmapAnnotation(
  n_cells = anno_barplot(anno$ncells, gp=gpar(fill="#440154FF"),
                                   bar_width = 1,
                                   height = unit(3,"cm"), axis_param = list(direction = "reverse")),
  score_fractions = anno_barplot(cur_df, gp=gpar(fill=metadata(sce_PalWu)$colors$KRT57_score[colnames(cur_df)]),
                                   bar_width = 1,
                                   height = unit(25,"cm"),
                                   width = unit(11,"cm"),
                                   show_row_names = FALSE), 
                          subtype = anno_simple(anno$subtype, col = sub_col, border = TRUE),
                          dataset = anno_simple(anno$dataset, col = dataset_col, border = TRUE),
                          border = TRUE, 
                          gap = unit(1,"mm"),
                          which = "row")

lgd <- Legend(title = "subtype", at = names(sub_col), 
               legend_gp = gpar(fill = sub_col))               
lgd_1 <- Legend(title = "dataset", at = names(dataset_col), 
               legend_gp = gpar(fill = dataset_col))  

draw(fraction_anno + NULL, annotation_legend_list = list(lgd, lgd_1))
```

## S6D

```{r}
dittoBarPlot(sce_PalWu, var = "KRT5_7_score", group.by = "TNBC")+
  scale_fill_manual(values = metadata(sce_PalWu)$colors$KRT57_score)+
  guides(fill = "none")
```

## S6E

```{r}
keratins <- c("KRT5","KRT7","KRT15","KRT8","KRT18","KRT19", "KRT14","KRT17","ITGA6","EPCAM")

cluster_mean_sce_all <- aggregateAcrossCells(sce_PalWu_ALL, ids = sce_PalWu_ALL$KRT5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "logcounts",
                                 subset.row = rownames(sce_PalWu_ALL) %in% keratins)

scaled_mat <- t(scale(t(assay(cluster_mean_sce_all, "logcounts"))))

anno <- colData(sce_PalWu_ALL) %>% as.data.frame() %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% dplyr::count()
anno <- anno[match(anno$KRT5_7_score, rownames(t(scaled_mat))),]

## Print heatmap 
Heatmap(t(scaled_mat), name = "Scaled mean expression", clustering_method_columns = "ward.D2", clustering_method_rows = "ward.D2")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

## S6G

```{r}
plotReducedDim(sce_organoid, dimred = "TSNE", 
               colour_by = "KRT5", 
               by_exprs_values = "logcounts")+
plotReducedDim(sce_organoid, dimred = "TSNE", 
               colour_by = "KRT7", 
               by_exprs_values = "logcounts")

plotTSNE(sce_organoid, colour_by="KRT5_7_score")+
  scale_color_manual(values = metadata(sce_organoid)$colors$KRT57_score)
```

## S6H

```{r}
dittoBarPlot(sce_organoid, var = "KRT5_7_score", group.by = "Patient_ID",scale="percent")+
    scale_fill_manual(values = metadata(sce_organoid)$colors$KRT57_score)
```

## S6I

```{r}
keratins <- c("KRT5","KRT7","KRT15","KRT8","KRT18","KRT19", "KRT14","KRT17","ITGA6","EPCAM")

cluster_mean_sce_organoid <- aggregateAcrossCells(sce_organoid, ids = sce_organoid$KRT5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "logcounts",
                                 subset.row = rowData(sce_organoid)$Symbol %in% keratins)

scaled_mat <- t(scale(t(assay(cluster_mean_sce_organoid, "logcounts"))))

anno <- colData(sce_organoid) %>% as.data.frame() %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% dplyr::count()
anno <- anno[match(anno$KRT5_7_score, rownames(t(scaled_mat))),]

## Print heatmap 
Heatmap(t(scaled_mat), name = "Scaled mean expression", clustering_method_columns = "ward.D2", clustering_method_rows = "ward.D2")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```


# Session Info 

```{r}
sessionInfo()
```