---
title: "01_lib_read_data"
author: "Lasse Meyer"
date: "1/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Software requirements

```{r lib, message=FALSE,warning=FALSE}
#Data containers
library(SingleCellExperiment)

#Data processing
library(CATALYST)
library(scuttle)
library(scater)
library(imcRtools)

#Data visualization
library(cytomapper)
library(dittoSeq)
library(viridis)

#Utilities
library(tidyverse)
library(openxlsx)
library(stringr)
library(BiocParallel)
```

## Read in single-cell information from TMA-based TNBC study
based on [SingleCellExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html)
class (or [SpatialExperiment]())

### steinbock generated data
The `read_steinbock` function provided by `imcRtools` can now be used to read in
`steinbock` generated data. 

```{r read and inspect steinbock}
sce <- read_steinbock("/mnt/rcc_volume/TNBC/",intensities_folder = "intensities",regionprops_folder = "regionprops", graphs_folder = "neighbors",return_as = "sce")
sce
```

## Single-cell processing

We can set the `colnames` of the object to generate unique identifiers per cell:

```{r set-colnames}
colnames(sce) <- sce$cell_id <- paste0(sce$sample_id, "_", sce$ObjectNumber)

rownames(sce) <- rowData(sce)$clean_target
```

It is also often the case that sample-specific `metadata` are available externally. This metadata is available as an external csv file:

```{r add-meta}
meta <- read.csv("/mnt/rcc_volume/TNBC/data_analysis/03_data/Metadata_Combined_TNBC_LM.csv",row.names = 1)
meta_249 <- read.csv("/mnt/rcc_volume/TNBC/data_analysis/03_data/ZTMA249_TNBC_LM_processed_bysample.csv",row.names = 1) %>% select(ID_Number,PID)

## Add PID (different way of sample annotation for 174 and 249)
PID_df <- data.frame(sample_id = unique(sce$sample_id),
           ZTMA174_ID = str_split(unique(sce$sample_id),"_",simplify = TRUE)[,3],
           ZTMA249_ID = str_split(unique(sce$sample_id),"_",simplify = TRUE)[,4])

PID_df$ZTMA249_PID <- meta_249$PID[match(PID_df$ZTMA249_ID,meta_249$ID_Number)]

PID_df <- PID_df %>% mutate(PID = ifelse(ZTMA174_ID == "ID",ZTMA249_PID,ZTMA174_ID))

sce$PID <- PID_df$PID[match(sce$sample_id,PID_df$sample_id)]

# Remove sample with no available PID (ZTMA249.1_ZTMA249_ID_247_CX7Y1_99)
sce <- sce[,!is.na(sce$PID)]

## Add other relevant metadata based on PID
for (entry in colnames(meta)[!colnames(meta) %in% c("PID","ID_Number","OS_data","DFS_data")]) {
  colData(sce)[,entry] <- meta[,entry][match(colData(sce)[,"PID"],meta[,"PID"])]
}

#Mark cases that are HER2+ (no metadata available)
sce$HER2_case <- ifelse(!is.na(sce$TMANr),FALSE,TRUE)
```

**Define interesting channels**

Subset markers of interest for downstream analysis.

```{r select-features}
rowData(sce)$use_channel <- !grepl("DNA|Histone", rowData(sce)$clean_target)
```

**Define color schemes**

We will already at this point define color schemes for different metadata
entries which will be helpful for downstream data visualizations. 

```{r color-scheme}
colors <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/color_meta.rds")
colors

metadata(sce)$colors <- colors
```

**Transform counts**

Problem: Right-skewed distribution (bias from highly expressing cells)
Solution: Count transformation using an inverse hyperbolic sine function (arcsinh; common in flow cytometry) = leads to linear-like scale near zero and log-like scale beyond a threshold (`cofactor`)
We apply a `cofactor` of `1`. 

```{r transform-counts, message=FALSE, fig.width=7,fig.height=10}
set.seed(22)
cur_id <- sample(sce$sample_id,5)

#Raw counts
multi_dittoPlot(sce[,sce$sample_id %in% cur_id], vars = rownames(sce)[rowData(sce)$clean_target %in% c("panCK","Vimentin","CD45","CD3","CD68")],
               group.by = "PID", plots = c("ridgeplot"), 
               assay = "counts", 
               color.panel = metadata(sce)$colors$PID)

#Transformed counts
assay(sce, "exprs") <- asinh(counts(sce)/1)

multi_dittoPlot(sce[,sce$sample_id %in% cur_id], vars = rownames(sce)[rowData(sce)$clean_target %in% c("panCK","Vimentin","CD45","CD3","CD68")],
               group.by = "PID", plots = c("ridgeplot"), 
               assay = "exprs", 
               color.panel = metadata(sce)$colors$PID)

```

## Save single cell object for downstream analysis

Note that masks and images were loaded, modified and saved in the [00_seg_mesmer_QC.Rmd]() script. 

```{r save-objects-read-data}
saveRDS(sce, "/mnt/rcc_volume/TNBC/data_analysis/01_sce_TNBC.rds")
```


