---
title: "03_QC"
author: "Lasse Meyer"
date: "1/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Software requirements

```{r lib, message=FALSE,warning=FALSE}
#Data containers
library(SpatialExperiment)
library(SingleCellExperiment)

#Data processing
library(scuttle)
library(scater)
library(imcRtools)

#Data visualization
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)

#Utilities
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(BiocParallel)
library(BiocNeighbors)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
```

# Image and cell-level quality control

The following section discusses possible quality indicators for data obtained
by IMC and other highly multiplexed imaging technologies.
Due to the complexity of the data, the quality metric range across the image
and single-cell levels and the chapter is sectioned as such.

## Read in the data

We will first read in the data processed in previous sections:

```{r read-data, message=FALSE}
images <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/images_TNBC_spillcor.rds")
masks <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/masks_segtest_4.rds")
sce <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/sce_TNBC_spillcor.rds")
```

## Segmentation quality control

Was performed before. See [00_seg_mesmer_QC]().

Here, we again pre-select 5 images of interest (e.g. high immune infiltration, TLS, high epithelial cell number). We then perform image- and channel-wise
normalization (channels are first min-max normalized and scaled to a range of
0-1 before clipping the maximum intensity to 0.2).

```{r overlay-masks, message=FALSE}
#10 preselected images
img_files <- list.files("/mnt/rcc_volume/TNBC/img/")
img_ids<- c(which(str_detect(img_files ,"ZTMA249_ID_271_CX7Y4")),which(str_detect(img_files ,"ZTMA249_ID_82_AX2Y11")), which(str_detect(img_files ,"ZTMA249_ID_179_BX3Y8")), which(str_detect(img_files ,"ZTMA249_ID_183_BX7Y8")), which(str_detect(img_files ,"ZTMA174_B08.26868_79_AX7Y10")))

# Normalize and clip image
cur_images <- images[img_ids]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#by epithelial channels
plotPixels(cur_images,
           mask = masks[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "Vimentin", "SMA","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "Vimentin" = c("black", "cyan"),
                         "SMA" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = "/mnt/rcc_volume/TNBC/data_analysis/spillcor_selectedimages_epi.png", scale = 2))

#by immune channels
plotPixels(cur_images,
           mask = masks[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD3", "CD8a", "CD68", "CD20", "CD44"),
           colour = list("CD3" = c("black", "yellow"),
                         "CD20" = c("black", "magenta"),
                         "CD8a" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = "/mnt/rcc_volume/TNBC/data_analysis/spillcor_selectedimages_immune.png", scale = 2))
```

## Image-level quality control

Besides GUI for image-level quality control (e.g. HistoCAT), we will here specifically focus on quantitative metrics to assess image
quality.

It is often of interest to calculate the signal-to-noise ratio (SNR) for
individual channels and markers. Here, we define the SNR as:

$$SNR = I_s/I_n$$

where $I_s$ is the intensity of the signal (mean intensity of pixels with true
signal) and $I_n$ is the intensity of the noise (mean intensity of pixels
containing noise). Finding a threshold that separates pixels containing signal
and pixels containing noise is not trivial and different approaches can be
chosen. Here, we use the `otsu` thresholding approach to find pixels of the
"foreground" (i.e., signal) and "background" (i.e., noise). The SNR is then
defined as the mean intensity of foreground pixels divided by the mean intensity
of background pixels. We then plot the SNR versus mean signal intensity.

**We will perform Image-QC on images before spillover correction**

```{r image-snr, message=FALSE, warning=FALSE}
#Load uncompensated images
images <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/images_segtest.rds")

#Randomly select 100 images
set.seed(22011995)
img_random <- sample(seq_len(length(images)), 100)

cur_snr <- lapply(images[img_random], function(img){
    mat <- apply(img, 3, function(ch){
        # Otsu threshold
        thres <- otsu(ch, range = c(min(ch), max(ch)))
        # Signal-to-noise ratio
        snr <- mean(ch[ch > thres]) / mean(ch[ch <= thres])
        # Signal intensity
        ps <- mean(ch[ch > thres])
        
        return(c(snr = snr, ps = ps))
    })
    t(mat) %>% as.data.frame() %>% 
        mutate(marker = colnames(mat)) %>% 
        pivot_longer(cols = c(snr, ps))
})

cur_snr <- do.call(rbind, cur_snr)

cur_snr <- cur_snr %>% 
    group_by(marker, name) %>%
    summarize(mean = mean(value),
              ci = qnorm(0.975)*sd(value)/sqrt(n())) %>%
    pivot_wider(names_from = name, values_from = c(mean, ci))

#NA Values for EGFR/DNA_1/DNA_2?

    ggplot(cur_snr) +
#    geom_errorbar(aes(y = log2(mean_snr), xmin = log2(mean_ps - ci_ps), 
#                      xmax = log2(mean_ps + ci_ps))) +
#    geom_errorbar(aes(x = log2(mean_ps), ymin = log2(mean_snr - ci_snr), 
#                      ymax = log2(mean_snr + ci_snr))) +
    geom_point(aes(log2(mean_ps), log2(mean_snr), color = log2(mean_snr))) +
    geom_label_repel(aes(log2(mean_ps), log2(mean_snr), label = marker)) +
    theme_classic(base_size = 15) + ylab("Signal-to-noise ratio") +
    geom_vline(xintercept = 0,linetype="dashed", color = "grey")+
    #coord_cartesian(ylim = c(0,100))+
    xlab("Signal intensity")+
  ggtitle("Image-level QC")

#Inf for p53, panCK, ki67
```

Another quality indicator is the image area covered by cells (or biological tissue).
This metric identifies ROIs where little cells are present, possibly hinting
at incorrect selection of the ROI. We can compute the percentage of image 
are covered by cells using the TMA masks measured with ilastik and cellprofiler. 

We can then identify ROIs that have unusually low cell coverage or number of cells. These images can now be visualized using `cytomapper`.

```{r cell-density}
#Load TMA tissue masks
tissue_area <- read_csv("/mnt/rcc_volume/TNBC/tissue_mask/tissue_area_TissueRescaled.csv") 
sce$tissue_area <- as.numeric(tissue_area$AreaShape_Area[match(colData(sce)$sample_id,str_split(tissue_area$FileName_Tissue_Mask, "_Pro",simplify = TRUE)[,1])])

cell_density <- colData(sce) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    summarize(cell_area = sum(area),
              tissue_area = mean(tissue_area)) %>%
  mutate(covered_area = (cell_area / tissue_area)*100) %>%
  mutate(n_cells = as.numeric(table(colData(sce)$sample_id))) %>%
  arrange(covered_area)

median(cell_density$covered_area)
quantile(cell_density$covered_area)[2]

ggplot(cell_density) +
  geom_point(aes(reorder(sample_id,covered_area), covered_area, color = covered_area))+ 
  ylim(c(0, 100))+
  geom_hline(yintercept = median(cell_density$covered_area), linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$covered_area)[2],linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$covered_area)[4],linetype="dashed", color = "grey")+
  theme_classic(base_size = 15)+
  theme(axis.text.x = element_blank())+
  ylab("% covered area")+ 
  xlab("")

ggplot(cell_density) +
  geom_point(aes(reorder(sample_id,n_cells), n_cells, color = n_cells))+
  geom_hline(yintercept = median(cell_density$n_cells), linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$n_cells)[2],linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$n_cells)[4],linetype="dashed", color = "grey")+
  theme_classic(base_size = 15)+
  theme(axis.text.x = element_blank())+
  ylab("Number of cells per ROI")+ 
  xlab("")

## Visualize critical images
#Kick-out criteria:
#n_cells < 300, covered_area < 8, tissue_area < 150000?
#split acquisitions should also be excluded
image_id_cri <- cell_density %>% filter(n_cells < 300 | covered_area < 8 | tissue_area < 150000 | str_detect(cell_density$sample_id, "split")) %>% pull(sample_id)

# Normalize and clip images
cur_images <- images[names(images) %in% image_id_cri]
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, inputRange = c(0, 0.2))

plotPixels(cur_images,
           mask = masks[names(masks) %in% image_id_cri],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "CD3","CD68","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "CD3" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))

#ZTMA249.3_ZTMA249_ID_16_AX8Y2 - split_364 looks fine - keep

#Remove outlying images and cells (n=1843 cells)
image_id_cri <- image_id_cri[!image_id_cri %in% "ZTMA249.3_ZTMA249_ID_16_AX8Y2 - split_364"]

summary(colData(sce)$sample_id %in% image_id_cri)
sce <- sce[,!colData(sce)$sample_id %in% image_id_cri]
```

## Cell-level quality control

Related to the signal-to-noise ratio (SNR) calculated above on the pixel-level, a similar measure can be derived on the single-cell level. Here, we will use a two component Gaussian mixture model for each marker to find cells with positive and negative expression. The SNR is defined as:
$$SNR=Is/In$$ 

where **Is** is the intensity of the signal (mean intensity of cells with positive signal) and **In** is the intensity of the noise (mean intensity of cells lacking expression). We calculate the SNR and signal intensity by fitting the mixture model across the transformed counts of all cells contained in the sce.

```{r GMM cell level, message = FALSE, warning=FALSE}
assays(sce)

set.seed(22)
mat <- apply(assay(sce, "exprs"), 1, function(x){
    cur_model <- Mclust(x, G = 2)
    mean1 <- mean(x[cur_model$classification == 1])
    mean2 <- mean(x[cur_model$classification == 2])
    
    signal <- ifelse(mean1 > mean2, mean1, mean2)
    noise <- ifelse(mean1 > mean2, mean2, mean1)
    
    return(c(snr = signal/noise, ps = signal))
})
    
cell_snr <- t(mat) %>% as.data.frame() %>% 
        mutate(marker = colnames(mat))

cell_snr %>% ggplot() +
    geom_point(aes(log2(ps), log2(snr),color = log2(snr))) +
    geom_label_repel(aes(log2(ps), log2(snr), label = marker)) +
    theme_classic(base_size = 15) + ylab("Signal-to-noise ratio") +
      geom_vline(xintercept = 0,linetype="dashed", color = "grey")+
    xlab("Signal intensity")+
  ggtitle("Cell-level QC")
```

Next, we observe the distributions of cell size across the individual images. Differences in cell size distributions can indicate segmentation biases due to differences in cell density or can indicate biological differences due to cell type compositions (tumor cells tend to be larger than immune cells).
We also detect very small cells in the dataset and will remove them. The chosen threshold is arbitrary and needs to be adjusted per dataset.

```{r cellsize_dis, message=FALSE,warning=FALSE}
colData(sce) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    ggplot() +
        geom_boxplot(aes(reorder(sample_id,area), area)) +
        theme_classic(base_size = 15) + 
        theme(axis.text.x = element_blank()) +
        ylab(~ "Cell area " (mu * m^2)) + xlab("")+
  geom_hline(yintercept = median(colData(sce)$area), linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(colData(sce)$area)[2],linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(colData(sce)$area)[4],linetype="dashed", color = "grey")

#Cell area range
summary(sce$area)

#Only keep cells with an area above 5 (n=3223 cells)
sum(sce$area <= 5)
sce <- sce[,sce$area > 5]
```

**Staining artifacts**
Observing potential staining differences (e.g. between samples/batches) can be crucial to assess data quality. We will use ridgeline visualizations to check differences in staining patterns:

```{r stainingarti, message=FALSE, warning=FALSE, fig.width=20, fig.height=20}
#Randomly select 10 sample Ids for visualization
set.seed(22)
sampleid_random <- sample(unique(sce$sample_id), 10)

dittoRidgePlot(sce[,sce$sample_id %in% sampleid_random], var = "panCK", group.by = "sample_id", assay = "exprs")
dittoRidgePlot(sce[,1:10000], var = "CD44", group.by = "sample_id", assay = "exprs")
dittoRidgePlot(sce[,1:10000], var = "CD45", group.by = "sample_id", assay = "exprs")

ridge_plots <- lapply(c(rownames(sce)[rowData(sce)$use_channel]),
                    function(x){
                      p <- dittoRidgePlot(sce[,sce$sample_id %in% sampleid_random], 
                                          var = x, 
                                          group.by = "sample_id", 
                                          assay = "exprs")
                      return(p)                    
                    })

plot_grid(plotlist = ridge_plots)
```

**Dimensionality reduction**
Finally, we will use non-linear dimensionality reduction methods to project cells from a high-dimensional (40) down to a low-dimensional (2) space. For this the scater package provides the runUMAP (and runTSNE) function. To ensure reproducibility, we will set a seed. For computing time, we will subsample to 10% of the cells.

**Goal:** Visualization of the low-dimensional embedding facilitates assessment of potential "batch effects". Should not be present in present cohort.

```{r dim_red, fig.width=15,fig.height=15}
## Subsample 10% of cells from each core 
#to save time and not get super crowded dimensionality reduction plots
set.seed(22)
sub <- colData(sce) %>% as.data.frame() %>% group_by("sample_id") %>% slice_sample(prop = 0.1) %>% arrange("sample_id") %>% mutate(sub_id = paste(sample_id,ObjectNumber,sep="_")) %>% pull(sub_id)
sce_sub = sce[,sub]

#Run umap and tsne and save output directly into subsampled SCE
sce_sub = runUMAP(sce_sub, exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam(workers=14))
sce_sub = runTSNE(sce_sub, exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam(workers=14))

#Visualize
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "sample_id")+
  guides(color="none")
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "sample_id")+
  guides(color="none")

#Visualize all marker expression
all_plots <- lapply(c(rownames(sce_sub)[rowData(sce_sub)$use_channel]),
                    function(x){
                      p <- plotReducedDim(sce_sub, dimred = "TSNE", 
                                          colour_by = x, 
                                          by_exprs_values = "exprs",
                                          point_size = 0.5)
                      return(p)                    
                    })

library(cowplot)
plot_grid(plotlist = all_plots)

# ## Also possible to use all cells, but more time intensive
# #set.seed(22)
# #sce <- runUMAP(sce, subset_row = rowData(sce)$use_channel, exprs_values = "exprs", BPPARAM = MulticoreParam(workers = 14)) 
# #
# #Visualize sample IDs
# #plotReducedDim(sce, dimred = "UMAP", colour_by = "sample_id")+
# #  guides(color="none")
# 
# #Visualize marker of interest
# dittoDimPlot(sce, var = "panCK", reduction.use = "UMAP", 
#                    assay = "exprs", size = 0.2) +
#     scale_color_viridis(name = "panCK") +
#     ggtitle("panCK expression on UMAP")
# 
# 
# #Visualize all marker expression
# all_plots <- lapply(c(rownames(sce)[rowData(sce)$use_channel]),
#                     function(x){
#                       p <- plotReducedDim(sce, dimred = "UMAP", 
#                                           colour_by = x, 
#                                           by_exprs_values = "exprs",
#                                           point_size = 0.5)
#                       return(p)                    
#                     })
# 
# library(cowplot)
# plot_grid(plotlist = all_plots)
```

## Save objects

The modified `SingleCellExperiment` object is saved for further downstream analysis.

```{r save-objects-quality-control}
saveRDS(sce, "/mnt/rcc_volume/TNBC/data_analysis/sce_TNBC_QC.rds")
saveRDS(sce_sub, "/mnt/rcc_volume/TNBC/data_analysis/sce_TNBC_QC_sub.rds")
```










