---
title: "04_SingleCell_InitialClustering"
output: html_document
date: '2022-03-30'
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Software requirements

```{r lib, message=FALSE,warning=FALSE}
#Data containers
library(SingleCellExperiment)

#Data processing
library(scuttle)
library(scater)
library(scran)
library(bluster)
library(imcRtools)

#Data visualization
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)

#Utilities
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(VennDiagram)
library(easyalluvial)

#Batch correction
library(batchelor)
```

# Single-cell analysis - Pipeline 1 - Distinguish tumor and immune cells

Here, we will perform the first single cell analysis steps, including distinguishing tumor and stromal cells and dimensionality reduction.

## Read in data 
First, we will read in the `SingleCellExperiment` object containing the cleaned, spillover- and quality-controlled single-cell data. 

```{r read-data-batch-correction, message=FALSE}
sce <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/03_sce_TNBC_QC.rds")
sce
```

## Perform "overclustering"

Next, we will run a granular Rphenograph (Jaccard-based weights, Louvain clustering as community detection algorithm) clustering. 

```{r overcluster, message=FALSE}
set.seed(22)

clust <- clusterCells(sce[rowData(sce)$use_channel==TRUE,],
                        assay.type = "exprs",
                        full = TRUE,
                        BLUSPARAM = NNGraphParam(k=30,
                                                 cluster.fun ="louvain",
                                                 type="jaccard",
                                                 BPPARAM= MulticoreParam(workers = 20),
                                                 BNPARAM = AnnoyParam()))

sce$cluster <- clust$clusters

#Save sce after clustering 
saveRDS(sce, "/mnt/rcc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
```


```{r}
library(Rphenoannoy)

set.seed(22)

mat <- t(assay(sce, "exprs")[rowData(sce)$use_channel==TRUE,])

#k=30
out <- Rphenoannoy(mat, k = 30)

clusters <- factor(membership(out[[2]]))

sce$cluster <- clusters

unique(sce$cluster)

#Save SCEs after clustering
saveRDS(sce, "/mnt/rcc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
```

## Use GMM for panCK

We will use a two component gaussian mixture model to distinguish between tumor (panCK+) and stromal (panCK-) cells. 

```{r GMM_CK, message=FALSE,warning=FALSE}
## Gaussian mixture model to distinguish tumor/stroma based on panCK expression
GMM_CK <- Mclust(assay(sce, "exprs")["panCK",], G = 2)

GMM_df <- data.frame(class_CK = ifelse(GMM_CK$classification == 2 & GMM_CK$uncertainty == 0, "panCK+", "panCK-"), 
                     panCK_exprs = assay(sce, "exprs")["panCK",],
                     uncertainty = GMM_CK$uncertainty)

#Visualize GMM
GMM_plot_exp <- ggplot(GMM_df, aes(x = panCK_exprs, fill = class_CK))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")
  
GMM_plot_unc <- ggplot(GMM_df, aes(x = panCK_exprs, y = uncertainty, color = class_CK))+
  geom_line()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_color_brewer(palette = "Set1")

GMM_plot_exp/GMM_plot_unc

#Add GMM CK class to SCE
#sce$GMM_CK <- GMM_df$class_CK

## Gaussian mixture model to distinguish tumor/stroma based on Ecad expression
GMM_Ecad <- Mclust(assay(sce, "exprs")["Ecad",], G = 2)

GMM_df_1 <- data.frame(class_Ecad = ifelse(GMM_Ecad$classification == 2, 
                                           "Ecad+", "Ecad-"), 
                     Ecad_exprs = assay(sce, "exprs")["Ecad",],
                     uncertainty = GMM_Ecad$uncertainty)

GMM_plot_exp_1 <- ggplot(GMM_df_1, aes(x = Ecad_exprs, fill = class_Ecad))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")

GMM_plot_unc_1 <- ggplot(GMM_df_1, aes(x = Ecad_exprs, y = uncertainty, color = class_Ecad))+
  geom_line()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_color_brewer(palette = "Set1")

GMM_plot_exp_1/GMM_plot_unc_1

#Add GMM CK class to SCE
#sce$GMM_Ecad <- GMM_df_1$class_Ecad

#Save sce after clustering and GMM
#saveRDS(sce, "/mnt/rcc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
```

## Check correlation of panCK and Ecad as epithelial cell markers

```{r marker correlations, message = FALSE}
sce <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")

GOI <- c("panCK","Ecad")
cor_sce <- as.data.frame(t(assay(sce,"exprs")[rowData(sce)$clean_target %in% GOI,]))
library(psych)
# For specific markers of interest (also shows the value distribution which is recommended)
pairs.panels(cor_sce,
             method = "spearman",
             hist.col = "#A6CEE3",
             digits = 2)
```

## Astir for automated cell type assignment 

## 2. ASTIR model

see [here](https://www.sciencedirect.com/science/article/pii/S2405471221003355?via%3Dihub) for full publication (Geuenich et al.,2021, Cell Systems).

```{r astir-expo}
#Export arsinh normalized expression matrix as csv
#Add sampleID as covariate in design matrix
expr_astir <- assay(sce,"exprs") %>% t() %>% as.data.frame() %>% mutate(sample = sce$sample_id)

write.csv(expr_astir, file = "/mnt/rcc_volume/TNBC/data_analysis/04_astir/expr_TNBC_astir.csv",row.names = TRUE)
```

`Next: Run Astir in conda-environment.`

```{r astir-ana, message = FALSE,warning=FALSE}
#Import Astir cell type assignments
astir_type <- read_csv("/mnt/rcc_volume/TNBC/data_analysis/04_astir/cell_types_TNBC.csv",col_names = c("ID","cell_type"),skip = 1)

astir_sum <- astir_type %>% as.data.frame %>% group_by(cell_type) %>% dplyr::dplyr::count()

sce$astir_type <- astir_type$cell_type

#Mark channels used for cell type assignment 
astir_channel = c("panCK","Ecad","CD3","CD8a","CD4","CD20","CD68","SMA","CD11c")

rowData(sce)$use_astir <- rowData(sce) %>% as.data.frame() %>% mutate(use_astir = ifelse(rownames(sce) %in% astir_channel,TRUE,FALSE)) %>% pull(use_astir)
```


## Heatmap 

Heatmap of clusters with GMM and ASTIR information 

```{r heatmap, message=FALSE, warning=FALSE}
#Read sce after clustering and GMM
#sce <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")

### 1. Heatmap body ###

# Min-Max [0,1] normalization of arcsinh-transformed counts
assay(sce, "min_max") <- assay(sce, "exprs") - rowMins(assay(sce, "exprs"))
assay(sce, "min_max") <- assay(sce, "min_max") / (rowMaxs(assay(sce, "exprs")) - rowMins(assay(sce, "exprs")))

# Aggregrate across cells with the mean
mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce)$use_channel)

# Heatmap body color 
col_main = viridis(100)

### 2. Heatmap annotation ###

### 2.1 Spatial features

#Add number of neighbors to sce object (saved in ColPair)
n_neighbors <- colPair(sce) %>% as.data.frame() %>% group_by(from) %>% dplyr::count() %>% arrange(desc(n))
sce$n_neighbors <- n_neighbors$n[match(seq_along(colnames(sce)),n_neighbors$from)]
sce$n_neighbors <- sce$n_neighbors %>% replace_na(0)

#Select spatial features and average over clusters 
spatial <- colData(sce) %>% as.data.frame %>% select(area,eccentricity,cluster,n_neighbors)
spatial <- spatial %>% select(-cluster) %>% aggregate(by = list(cluster = spatial$cluster), FUN = mean) %>% column_to_rownames("cluster")

#Define color schemes for spatial features and add to metadata
area_col <- colorRamp2(c(min(spatial$area),median(spatial$area),max(spatial$area)), c(brewer.pal(3,"Blues")[1],brewer.pal(3,"Blues")[2],brewer.pal(3,"Blues")[3]))
metadata(sce)$colors$area <- area_col

eccentricity_col <- colorRamp2(c(min(spatial$eccentricity),median(spatial$eccentricity),max(spatial$eccentricity)), c(brewer.pal(3,"Purples")[1],brewer.pal(3,"Purples")[2],brewer.pal(3,"Purples")[3]))
metadata(sce)$colors$eccentricity <- eccentricity_col

n_neighbors_col <- colorRamp2(c(min(spatial$n_neighbors),median(spatial$n_neighbors),max(spatial$n_neighbors)), c(brewer.pal(3,"RdPu")[1],brewer.pal(3,"RdPu")[2],brewer.pal(3,"RdPu")[3]))
metadata(sce)$colors$n_neighbors <- n_neighbors_col

#Create Spatial Heatmap Annotation
ha_spatial <- HeatmapAnnotation(
    area = spatial$area,
    eccentricity = spatial$eccentricity,
    n_neighbors = spatial$n_neighbors, 
    col = list(area = metadata(sce)$colors$area,
               eccentricity = metadata(sce)$colors$eccentricity,
               n_neighbors = metadata(sce)$colors$n_neighbors),
    border = TRUE,
    which = "row")

### 2.2 Metadata features

#Number of cells per cluster 
n_cells <- colData(mean_sce) %>% as.data.frame %>% select(cluster, ncells)

#GMM - panCK
nr_tum <- colData(sce) %>% as.data.frame %>% select(cluster, GMM_CK) %>% table()
fr_tum <- as.data.frame(nr_tum) %>% pivot_wider(names_from = GMM_CK, values_from = Freq) %>% column_to_rownames("cluster")
fr_tum <- fr_tum %>% transmute(fr_tum = `panCK+`/rowSums(fr_tum)) %>% as.data.frame()

# #Astir
nr_tum_2 <- colData(sce) %>% as.data.frame %>% select(cluster, astir_type) %>% table()
fr_tum_2 <- as.data.frame(nr_tum_2) %>% pivot_wider(names_from = astir_type, values_from = Freq) %>% column_to_rownames("cluster")
fr_tum_2 <- fr_tum_2 %>% transmute(fr_tum = `Epithelial`/rowSums(fr_tum_2)) %>% as.data.frame()

#GMM - Ecad
nr_tum_3 <- colData(sce) %>% as.data.frame %>% select(cluster, GMM_Ecad) %>% table()
fr_tum_3 <- as.data.frame(nr_tum_3) %>% pivot_wider(names_from = GMM_Ecad, values_from = Freq) %>% column_to_rownames("cluster")
fr_tum_3 <- fr_tum_3 %>% transmute(fr_tum_3 = `Ecad+`/rowSums(fr_tum_3)) %>% as.data.frame()

ha_meta <- HeatmapAnnotation(GMM_CK = anno_barplot(fr_tum, width = unit(10, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            GMM_Ecad = anno_barplot(fr_tum_3, width = unit(10, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            Astir_Epi = anno_barplot(fr_tum_2, width = unit(10, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            n_cells = anno_barplot(n_cells$ncells, width = unit(10, "mm"),gp = gpar(fill = "violet", col = 1)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(1,"mm"),
                            which = "row")

### 3. Define tumor, mixed and stroma cluster (based on GMM_CK)

anno_cluscat <- fr_tum %>% 
  rownames_to_column(var = "cluster") %>% 
  mutate(cluster_category = ifelse(fr_tum >= 0.7,"Epithelial",ifelse(fr_tum < 0.7 & fr_tum > 0.2,"Mixed","Non_Epithelial")))

#Set colors for cluster_category and add to metdata
cluster_category <- setNames(c("#1F78B4", "#A6CEE3", "#B2DF8A"),sort(unique(anno_cluscat$cluster_category)))                                                                  
metadata(sce)$colors$cluster_category <- cluster_category               

#Visualize GMM_panCK density plots
ggplot(anno_cluscat, aes(fr_tum))+
  geom_density()+
  geom_rug(aes(color = cluster_category))+
  geom_vline(xintercept = 0.7, color = "red3")+
  geom_vline(xintercept = 0.2, color = "red3")+
  theme_classic()+
  theme(axis.title.x = element_blank())+
  scale_color_manual(values = cluster_category)

ha_cluscat <- HeatmapAnnotation(cluster_category = anno_cluscat$cluster_category,
                                col = list(cluster_category = metadata(sce)$colors$cluster_category),
                                border = TRUE,
                                which = "row")



### 4. Plot combined Heatmap
h <- Heatmap(t(assay(mean_sce, "min_max")),
        col = col_main,
        name= "min-max",
        km = 1,
        show_row_names = T, 
        show_column_names =  T, row_split = anno_cluscat$cluster_category,
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")+
  ha_spatial+
  ha_meta+
  ha_cluscat

draw(h)

### Add cluster_categories to sce
#sce$cluster_category <- anno_cluscat$cluster_category[match(sce$cluster,anno_cluscat $cluster)]

#Save sce
#saveRDS(sce, "/mnt/rcc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
```

## Venn diagrams / Pie plots for categories

```{r prop_tumor, message=FALSE, warning=FALSE}
## Pie plots 
#panCK+ per cluster category
prop_tumor <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(GMM_CK == "panCK+") %>% group_by(cluster_category) %>% dplyr::count()
prop_tumor <- prop_tumor %>% mutate(fr_tumor_clus = n/sum(prop_tumor$n))

pie(prop_tumor$fr_tumor_clus, labels = paste0(prop_tumor$cluster_category," - ",round(prop_tumor$fr_tumor_clus,3),"%"),col=metadata(sce)$colors$cluster_category, main = "panCK+")

#panCK- per cluster category
prop_tumor_1 <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(GMM_CK == "panCK-") %>% group_by(cluster_category) %>% dplyr::count()
prop_tumor_1 <- prop_tumor_1 %>% mutate(fr_tumor_clus = n/sum(prop_tumor_1$n))
pie(prop_tumor_1$fr_tumor_clus, labels = paste0(prop_tumor_1$cluster_category," - ",round(prop_tumor_1$fr_tumor_clus,3),"%"),col=metadata(sce)$colors$cluster_category, main = "panCK-")

#Tumor per GMM_CK
prop_tumor_2 <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(cluster_category == "Epithelial") %>% group_by(GMM_CK) %>% dplyr::count()
prop_tumor_2 <- prop_tumor_2 %>% mutate(fr_tumor_clus = n/sum(prop_tumor_2$n))
pie(prop_tumor_2$fr_tumor_clus, labels = paste0(prop_tumor_2$GMM_CK," - ",round(prop_tumor_2$fr_tumor_clus,3),"%"), col =c("#E41A1C","#377EB8"), main = "Epithelial")

prop_tumor_3 <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(cluster_category == "Mixed") %>% group_by(GMM_CK) %>% dplyr::count()
prop_tumor_3 <- prop_tumor_3 %>% mutate(fr_tumor_clus = n/sum(prop_tumor_3$n))
pie(prop_tumor_3$fr_tumor_clus, labels = paste0(prop_tumor_3$GMM_CK," - ",round(prop_tumor_3$fr_tumor_clus,3),"%"), col =c("#E41A1C","#377EB8"), main = "Mixed")

prop_tumor_4 <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(cluster_category == "Non_Epithelial") %>% group_by(GMM_CK) %>% dplyr::count()
prop_tumor_4 <- prop_tumor_4 %>% mutate(fr_tumor_clus = n/sum(prop_tumor_4$n))
pie(prop_tumor_4$fr_tumor_clus, labels = paste0(prop_tumor_4$GMM_CK," - ",round(prop_tumor_4$fr_tumor_clus,3),"%"), col =c("#E41A1C","#377EB8"), main = "Non_Epithelial")
```

## Re-analyse mixed clusters based on GMM

```{r clus_mixed, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
sce_mixed <- sce[,colData(sce)$cluster_category == "Mixed"] #n=60973

#Select cells that are either panCK or Ecad positive based on gaussian mixture model
sce_mixed$cluster_category_2 <- ifelse(sce_mixed$GMM_CK == "panCK+" | sce_mixed$GMM_Ecad == "Ecad+" ,"Epithelial","Non_Epithelial")

sce_mixed$cluster_category_2 %>% table()

spatial_mixed <- colData(sce_mixed) %>% as.data.frame %>% select(area,eccentricity,cluster_category_2,n_neighbors)
spatial_mixed <- spatial_mixed %>% select(-cluster_category_2) %>% aggregate(by = list(cluster_category_2 = spatial_mixed$cluster_category_2), FUN = mean) %>% column_to_rownames("cluster_category_2")

sce_mixed$cluster_category_2_mean_area <- as.character(round(spatial_mixed$area[match(sce_mixed$cluster_category_2, rownames(spatial_mixed))],2))

#Single-cell Heatmap 
dittoHeatmap(sce_mixed, genes = rownames(sce_mixed)[rowData(sce_mixed)$use_channel],
             assay = "exprs", cluster_cols = FALSE, order.by = c("cluster_category_2","GMM_CK","GMM_Ecad","cluster"), scale = "none",
             heatmap.colors = viridis(100), annot.by = c("cluster","GMM_CK","GMM_Ecad","cluster_category_2","cluster_category_2_mean_area"))


# Add cluster_categories_2 to main sce
sce$cluster_category_2 <- sce_mixed$cluster_category_2[match(colnames(sce),colnames(sce_mixed))]
```

## Visualize mixed clusters on images 

```{r viz critical clus, message=FALSE, warning=FALSE}
#Load images and masks
images <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/images_segtest.rds")
masks <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/masks_segtest_4.rds")

##Visualize mixed clusters on images - select 10 images for each category with highest number of cells
mixed_epi_samples <- colData(sce_mixed) %>% as.data.frame %>% group_by(cluster_category_2) %>% filter(cluster_category_2 == "Epithelial") %>% dplyr::count(sample_id) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(sample_id)

mixed_non_epi_samples <- colData(sce_mixed) %>% as.data.frame %>% group_by(cluster_category_2) %>% filter(cluster_category_2 == "Non_Epithelial") %>% dplyr::count(sample_id) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(sample_id)

mixed_samples_viz <- c(mixed_epi_samples,mixed_non_epi_samples)

mixed_samples_viz <- mixed_samples_viz %>% unique() #total of 20 unique images

#Prepare images
cur_images <- images[mixed_samples_viz]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

plotPixels(cur_images,
           mask = masks[mixed_samples_viz],
           object = sce_mixed,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "Vimentin", "SMA"),
           outline_by = "cluster_category_2",
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "Vimentin" = c("black", "cyan"),
                         "SMA" = c("black", "green"),
                         "cluster_category_2" = c("Epithelial" = "#377EB8", "Non_Epithelial" = "#E41A1C")
                         ),
           thick = TRUE,
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = paste0("/mnt/rcc_volume/TNBC/data_analysis/01_figures/mixed_clusters_viz/cluster_mixed_epi_non_epi_bycluscat.png"), scale = 2)
           )
```

## Define categories for downstream analysis
Use all epithelial clusters and mixed clusters with panCK+/Ecad+ GMM for downstream tumor analysis.

```{r ana_cat, message = FALSE, warning=FALSE}
library(easyalluvial)

sce$analysis_cat <- colData(sce) %>% as.data.frame %>% 
  select(cluster_category, cluster_category_2) %>% 
  mutate(analysis_cat = ifelse(sce$cluster_category == "Epithelial" |(sce$cluster_category == "Mixed" & sce$cluster_category_2 == "Epithelial"),"tumor","non_tumor")) %>% 
  pull(analysis_cat)

#color scheme for analysis_cat
analysis_cat <- setNames(c("#E41A1C", "#377EB8"),sort(unique(sce$analysis_cat)))                                                                  
metadata(sce)$colors$analysis_cat <- analysis_cat   

colData(sce) %>% as.data.frame %>% select(analysis_cat) %>% group_by(analysis_cat) %>% dplyr::count()
#492712 tumor cells
#436281 non-tumor cells

#Visualize in alluvial plot 
colData(sce) %>% as.data.frame %>% select(cluster_category, analysis_cat)

alluvial_wide(data = colData(sce) %>% as.data.frame %>% select(cluster_category, analysis_cat), 
              max_variables = 4,
              fill_by = 'all_flows',
              order_levels = c("Non_Epithelial","Epithelial","Mixed","tumor","non_tumor"))+
  theme_classic(base_size = 12, base_line_size = 0.5)


#Intersection plot
## Upset plot
#1. tumor   
library(UpSetR)
tumor = list(
analysis_cat_tumor = colData(sce) %>% as.data.frame %>% filter(analysis_cat == "tumor") %>% rownames_to_column() %>% pull("rowname"),
panCK_positive_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_CK == "panCK+") %>% rownames_to_column() %>% pull("rowname"),
Ecad_positive_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_Ecad == "Ecad+") %>% rownames_to_column() %>% pull("rowname")
)

upset <- upset(fromList(tumor), order.by = "freq")
upset

non_tumor = list(
analysis_cat_nontumor = colData(sce) %>% as.data.frame %>% filter(analysis_cat == "non_tumor") %>% rownames_to_column() %>% pull("rowname"),
panCK_negative_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_CK == "panCK-") %>% rownames_to_column() %>% pull("rowname"),
Ecad_negative_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_Ecad == "Ecad-") %>% rownames_to_column() %>% pull("rowname")
)

upset_1 <- upset(fromList(non_tumor), order.by = "freq")
upset_1
```

## Visualization of results on UMAP/TSNE

```{r viz_sub, message = FALSE, warning=FALSE, fig.width=20, fig.height=20}
# ## Subsample 10% of cells from each core 
# #to save time and not get super crowded dimensionality reduction plots
# set.seed(22)
# sub <- colData(sce) %>% as.data.frame() %>% group_by("sample_id") %>% slice_sample(prop = 0.1) %>% arrange("sample_id") %>% mutate(sub_id = paste(sample_id,ObjectNumber,sep="_")) %>% pull(sub_id)
# sce_sub = sce[,sub]
# sce_sub$cluster = as.factor(sce_sub$cluster)  
# 
# #Run UMAP/TSNE
# sce_sub = runUMAP(sce_sub[rowData(sce_sub)$use_channel,], n_neighbors = 30, pca = 50,exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam((detectCores()-2)))
# sce_sub = runTSNE(sce_sub[rowData(sce_sub)$use_channel,], exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam((detectCores()-2)))

sce_sub <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/04_sce_SUB_TNBC_Clus.rds")


## 1. Visualize TSNE
#panCK and GMM_CK
p1 <- plotReducedDim(sce_sub, dimred = "TSNE", by_exprs_values = "exprs",colour_by = "panCK")

p2 <- plotReducedDim(sce_sub, dimred = "TSNE", by_exprs_values = "exprs",colour_by = "Ecad")

p3 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "cluster_category")+
  scale_color_manual(values = metadata(sce)$colors$cluster_category)

p4 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$analysis_cat)

p5 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "PID")+#shape_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  guides(color = "none")

p6 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "TMANr")+#shape_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$TMANr)


p1+p2+p3+p4+p5+p6


#Visualize all marker expression
all_plots_TSNE <- lapply(c(rownames(sce_sub)[rowData(sce_sub)$use_channel]),
                    function(x){
                      p <- plotReducedDim(sce_sub, dimred = "TSNE", 
                                          colour_by = x, 
                                          by_exprs_values = "exprs",
                                          point_size = 0.5)
                      return(p)                    
                    })

library(cowplot)
plots_TSNE_exp <- plot_grid(plotlist = all_plots_TSNE)
plots_TSNE_exp
```

## Estimate fraction of tumor cells for each sample/image

```{r tumor cell fraction}
#Per sampleID
tumor_cell_SID <- colData(sce) %>% as.data.frame() %>% select(sample_id,analysis_cat) %>% group_by(sample_id) %>% table() %>% as.data.frame
tumor_cell_SID <- tumor_cell_SID %>% group_by(sample_id) %>% mutate(fra = ((Freq/sum(Freq))*100)) %>% filter(analysis_cat == "tumor")

ggplot(tumor_cell_SID, aes(x = analysis_cat,y = fra))+
  geom_boxplot()+
  geom_jitter(aes(color = sample_id), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color=FALSE)

tumor_cell_SID_cut_off <- foreach(i = 0:100, .combine = rbind) %do%{
data.frame(cut_off = i, n_PID = tumor_cell_SID %>% filter(fra > i) %>% ungroup %>% dplyr::count())
}

ggplot(tumor_cell_SID_cut_off, aes(x=cut_off,y = n))+
  geom_bar(stat = "identity",color = "black")+
  theme_classic()


#By patient
tumor_cell_PID <- colData(sce) %>% as.data.frame() %>% select(PID,analysis_cat) %>% group_by(PID) %>% table() %>% as.data.frame
tumor_cell_PID <- tumor_cell_PID %>% group_by(PID) %>% mutate(fra = ((Freq/sum(Freq))*100)) %>% filter(analysis_cat == "tumor")

ggplot(tumor_cell_PID, aes(x=analysis_cat,y = fra))+
  geom_boxplot()+
  geom_jitter(aes(color = PID, size = Freq), position=position_jitter(w=0.1,h=0.1))+
  guides(color=FALSE)+
  theme_classic()

ggplot(tumor_cell_PID, aes(x=analysis_cat,y = Freq))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = PID, size = fra), position=position_jitter(w=0.1,h=0.1))+
  guides(color=FALSE)+
  theme_classic()

tumor_cell_PID_cut_off <- foreach(i = 0:100, .combine = rbind) %do%{
data.frame(cut_off = i, n_PID = tumor_cell_PID %>% filter(fra > i) %>% ungroup %>% dplyr::count())
}

ggplot(tumor_cell_PID_cut_off, aes(x=cut_off,y = n))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()


#Choose filter criteria
#here total frequency of tumor cells per patient > 30 (100?)
PID_select_tumor_content <- tumor_cell_PID %>% filter(Freq < 100) %>% pull(PID) %>% unfactor 

summary(tumor_cell_PID$Freq)
sce$PID_low_tumor <- sce$PID %in% PID_select_tumor_content
```

## Extract representative images

```{r}
images <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/images_TNBC_spillcor.rds")
masks <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/masks_segtest_4.rds")
sce <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/07_sce_combined.rds")
```

```{r}
library(cytomapper)
#Randomly select 20 images
set.seed(22)
img_random <- sample(seq_len(length(images)), 10)

img_files <- list.files("/mnt/rcc_volume/TNBC/img/")
img_selected <- c(which(str_detect(img_files ,"ZTMA249_ID_320_CX8Y10")),which(str_detect(img_files ,"ZTMA249_ID_289_CX1Y7")),which(str_detect(img_files ,"ZTMA249_ID_36_AX4Y5")),which(str_detect(img_files ,"ZTMA249_ID_7_AX7Y1")),which(str_detect(img_files ,"ZTMA249_ID_220_BX4Y13")),which(str_detect(img_files ,"ZTMA249_ID_271_CX7Y4")),which(str_detect(img_files ,"ZTMA249_ID_82_AX2Y11")), which(str_detect(img_files ,"ZTMA249_ID_179_BX3Y8")), which(str_detect(img_files ,"ZTMA249_ID_183_BX7Y8")), which(str_detect(img_files ,"ZTMA174_B08.26868_79_AX7Y10")))

img_ids <- c(img_selected,img_random)

# Normalize and clip image
cur_images <- images[img_ids]
#cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
#cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#Subset mask and object
cur_mask <- masks[img_ids]
cur_sce <- sce[,sce$sample_id %in% names(cur_images)]
```

We will use cytoviewer for image exploratiion and download

```{r}
library(cytoviewer)
app <- cytoviewer(image = cur_images, mask = cur_mask, object = cur_sce, img_id = "sample_id", cell_id = "ObjectNumber")

shiny::runApp(app, launch.browser = TRUE)
```


## Save sce objects

```{r save_objects, message = FALSE, warning=FALSE}
#Save sce
saveRDS(sce, "/mnt/rcc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
#Save sce_sub
saveRDS(sce_sub, "/mnt/rcc_volume/TNBC/data_analysis/04_sce_SUB_TNBC_Clus.rds")
```

## Save plots as pdf

```{r save_plots}
##IMPORTANT: change to chunk output in console for this last part 

##save plots as pdf (vector graphics)
pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_EcadpanCKCor.pdf")
pairs.panels(cor_sce,
             method = "spearman",
             hist.col = "#A6CEE3",
             digits = 2)
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_GMMpanCK.pdf", width = 5)
GMM_plot_exp/GMM_plot_unc
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_GMMEcad.pdf",width = 5)
GMM_plot_exp_1/GMM_plot_unc_1
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_EpiNonEpi_HeatmapGMMCC.pdf",width = 12,height = 7)
draw(h)
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_EpiNonEpi_DensityCC.pdf", width = 5, height = 5)
ggplot(anno_cluscat, aes(fr_tum))+
  geom_density(size = 1)+
  geom_rug(aes(color = cluster_category))+
  geom_vline(xintercept = 0.7, color = "red3", size = 1)+
  geom_vline(xintercept = 0.2, color = "red3", size = 1)+
  theme_classic()+
  theme(axis.title.x = element_blank())+
  scale_color_manual(values = cluster_category)
dev.off()


pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_AlluvialGMMCC.pdf",width = 8,height = 5)
alluvial_wide(data = colData(sce) %>% as.data.frame %>% select(cluster_category, analysis_cat), 
              max_variables = 4,
              fill_by = 'all_flows',
              order_levels = c("Non_Epithelial","Epithelial","Mixed","tumor","non_tumor"))+
  theme_classic(base_size = 12, base_line_size = 0.5)
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_UpsetCC_TUMOR.pdf")
upset
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_UpsetCC_TME.pdf")
upset_1
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_TSNE_collection.pdf", width = 12)
p1+p2+p3+p4+p5+p6
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/04_initial_clus/TNBC_TSNE_exprsmarkers.pdf", height = 20, width = 12)
plots_TSNE_exp
dev.off()

##IMPORTANT: change back to chunk output inline
```
