---
title: "06_singlecell_tumor"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Software requirements

```{r lib, message=FALSE,warning=FALSE}
#Data containers
library(SpatialExperiment)
library(SingleCellExperiment)

#Data processing
library(scuttle)
library(scater)
library(scran)
library(imcRtools)

#Data visualization
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)

#Utilities
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(psych)

#Batch correction
library(batchelor)
```

# Single-cell analysis - Pipeline 2 - TNBC tumor phenotypes

Here, we will perform the second single cell analysis steps, including phenotyping tumor cells.

## Read in data
First, we will read in the `SingleCellExperiment` object containing the cleaned, spillover- and (batch-corrected) single-cell data. 

```{r read-data-batch-correction, message=FALSE}
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
```

## Filter sce for tumor cells
Here, we will filter for tumor cells based on the cluster category and GMM approach.

```{r filter sce, message=FALSE}
#based on cluster categories (n=492712)
sce <- sce[,colData(sce)$analysis_cat == "tumor"]
sce
```

# 1. Clustering

## Define markers for clustering

```{r tumor_markers, message = FALSE, warning = FALSE}
cluster_channel = c("CK5","CK8_18","CD15","KRT14","cMYC","HER2","ER","PR","p53","AR","GATA3","EGFR","p_mTOR","CK7","Vimentin","Ki67","HLADR")

#cell_state_channel = c()

other_immune_channel = c("Bcl2","CD44","panCK","Ecad","SMA","Fibronectin","CD68","CD3","CD20","CD11c","CD45","CD20","CD8a","CD4","CD31_vWF", "CAIX", "S6","c_Cas3_PARP", "pH3", "H3K27me3")

#Add to sce
rowData(sce)$cluster_channel <- (rowData(sce)$clean_target %in% cluster_channel)
#rowData(sce)$cell_state_channel <- (rowData(sce)$clean_target %in% cell_state_channel)
rowData(sce)$other_immune_channel <- (rowData(sce)$clean_target %in% other_immune_channel)
```

## Run clustering
Next, we will run Rphenoannoy (Jaccard-based weights, Louvain clustering as community detection algorithm) clustering with different resolution parameters. 

```{r overcluster, message=FALSE}
library(Rphenoannoy)
set.seed(22)

mat <- t(assay(sce, "exprs")[rowData(sce)$cluster_channel,])

#k=30
out <- Rphenoannoy(mat, k = 30)

clusters <- factor(membership(out[[2]]))

sce$cluster_30 <- clusters

unique(sce$cluster_30)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor.rds")

gc()

#k=50
out <- Rphenoannoy(mat, k = 50)

clusters <- factor(membership(out[[2]]))

sce$cluster_50 <- clusters

unique(sce$cluster_50)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor.rds")

gc()

#k=70
out <- Rphenoannoy(mat, k = 70)

clusters <- factor(membership(out[[2]]))

sce$cluster_70 <- clusters

unique(sce$cluster_70)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor.rds")

gc()

#k=90
out <- Rphenoannoy(mat, k = 90)

clusters <- factor(membership(out[[2]]))

sce$cluster_90 <- clusters

unique(sce$cluster_90)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor.rds")

gc()

#k=110
out <- Rphenoannoy(mat, k = 110)

clusters <- factor(membership(out[[2]]))

sce$cluster_110 <- clusters

unique(sce$cluster_110)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor.rds")

gc()

#k=130
out <- Rphenoannoy(mat, k = 130)

clusters <- factor(membership(out[[2]]))

sce$cluster_130 <- clusters

unique(sce$cluster_130)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor.rds")

gc()
```

## Cluster stability testing and comparison

```{r clustree, message = FALSE, warning=FALSE}
#Read sce after clustering and GMM
#sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor_new.rds")

#Pairwise rand index 
#Agreement between two clustering approaches - ARI > 0.5 = "good" similarity
pairwiseRand(sce$cluster_30,sce$cluster_50, mode = "index")
pairwiseRand(sce$cluster_50,sce$cluster_70, mode = "index")
pairwiseRand(sce$cluster_70,sce$cluster_90, mode = "index")
pairwiseRand(sce$cluster_90,sce$cluster_110, mode = "index")
pairwiseRand(sce$cluster_110,sce$cluster_130, mode = "index")

#Over different resolution parameters - Silhouette width [-1,1,max], Purity[0,1,max], Root-mean-squared difference [min]
clusters <- colnames(colData(sce))[str_detect(colnames(colData(sce)),"cluster_[:digit:]")]
df <- foreach(i = seq_along(clusters),.combine = rbind) %do% {
num_clus <- length(unique(colData(sce)[,clusters[i]]))
sil <- approxSilhouette(t(assay(sce, "min_max")), clusters=colData(sce)[,clusters[i]])
#purity <- neighborPurity(t(assay(sce, "exprs")), clusters=colData(sce)[,clusters[i]]) #takes very long to compute - do once and then save figure
rmsd <- clusterRMSD(t(assay(sce, "min_max")), clusters=colData(sce)[,clusters[i]])
data.frame(num_clus = num_clus,
           k = as.numeric(str_split(clusters[i],"_",simplify = TRUE)[,2]), 
           sil = mean(sil$width), 
#           purity = mean(purity$purity),
           rmsd = sum(rmsd))
}

p1 <- ggplot(df, aes(x=k, y=num_clus))+
  geom_point(aes(k, num_clus, col = num_clus))+
  geom_line(aes(k, num_clus,col=num_clus))+
  theme_classic()+
  scale_color_continuous(type = "viridis")

p2 <- ggplot(df, aes(x=k, y=rmsd))+
  geom_point(aes(k, rmsd, col = rmsd))+
  geom_line(aes(k, rmsd,col=rmsd))+
  theme_classic()+
  scale_color_continuous(type = "viridis")

p3 <- ggplot(df, aes(x=k, y=sil))+
  geom_point(aes(k, sil, col = sil))+
  geom_line(aes(k, sil,col=sil))+
  theme_classic()+
  scale_color_continuous(type = "viridis")
  
p1+p2+p3 

#Chosen method => k=110

# #Cluster colors 
# col_clus <- colorRampPalette(brewer.pal(9,"Paired"))(length(unique(sce$cluster_110)))
# 
# #By clusters for one method (here k=110)
# #1. Silhouette width [-1,1]
# #Clusters with large positive silhouette widths are well-separated from other clusters
# sil.approx <- approxSilhouette(t(assay(sce, "exprs")), clusters=sce$cluster_110)
# sil.data <- as.data.frame(sil.approx)
# sil.data$closest <- factor(ifelse(sil.data$width > 0, sce$cluster_110, sil.data$other))
# sil.data$cluster <- sce$cluster_110
# 
# p4 <- ggplot(sil.data, aes(x=cluster, y=width, colour=closest)) +
#     ggbeeswarm::geom_quasirandom(method="smiley")+
#   theme_classic()+
#   scale_color_manual(values = col_clus)
# p4

```
## Add missing metadata 

```{r}
# Missing metadata for 10 ZTMA174 patients
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor.rds")

meta <- read.csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Metadata_Combined_TNBC_LM.csv",row.names = 1)

colData <- colData(sce) %>% as.data.frame() %>% select("PID","age","grade","pT", "pT_simple", "pN", "pN_simple", "status_OS","status_DFS", "OS_months","DFS_months") %>% distinct()

## Add relevant metadata based on PID
for (entry in colnames(meta)[colnames(meta) %in% c("age","grade","pT", "pT_simple", "pN", "pN_simple", "status_OS","status_DFS", "OS_months","DFS_months")]) {
  colData(sce)[,entry] <- meta[,entry][match(colData(sce)[,"PID"],meta[,"PID"])]
}
```

# 2. Heatmap visualization

## Heatmap with different scales 

```{r heatmap scales, message=FALSE,warning=FALSE}
#Read sce after clustering
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor_new.rds")

#Heatmap body colors
col_exprs <- colorRamp2(c(0,1,2,3), c("#440154FF", "#31688EFF", "#35B779FF", "#FDE725FF"))
col_min_max <- colorRamp2(c(0,0.2,0.4,0.6,0.8), c("#440154FF", "#33628DFF", "#28AE80FF", "#DBE318FF", "#FDE725FF"))

### 1. Mean exprs counts
#Aggregrate across cells with the mean

cluster_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_110,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$cluster_channel)

cellstate_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_110,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$cell_state_channel)

other_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_110,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$other_immune_channel)

h_exprs <- Heatmap(t(assay(cluster_mean_sce, "exprs")),
        column_title = "1. mean exprs",
        col = col_exprs,
        name= "mean exprs",
        show_row_names = T, 
        show_column_names =  T, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")+
Heatmap(t(assay(cellstate_mean_sce, "exprs")),
        col = col_exprs,
        name= "mean exprs",
        show_row_names = T,
        show_column_names =  T,
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")+
  Heatmap(t(assay(other_mean_sce, "exprs")),
        col = col_exprs,
        name= "mean exprs",
        show_row_names = T, 
        show_column_names =  T, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")


h_exprs 

### 2. Mean min_max counts
#Aggregrate across cells with the mean
cluster_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_110,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce)$cluster_channel)

cellstate_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_110,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce)$cell_state_channel)

other_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_110,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce)$other_immune_channel)

h_min_max <- Heatmap(t(assay(cluster_mean_sce, "min_max")),
        column_title = "2. mean min_max",
        col = col_min_max,
        name= "min_max",
        show_row_names = T, 
        show_column_names =  T, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")+
Heatmap(t(assay(cellstate_mean_sce, "min_max")),
        col = col_min_max,
        name= "min_max",
        show_row_names = T,
        show_column_names =  T,
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")+
Heatmap(t(assay(other_mean_sce, "min_max")),
        col = col_min_max,
        name= "min_max",
        show_row_names = T, 
        show_column_names =  T, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")

h_exprs+
h_min_max
```

### Optional: Cluster merging of similar clusters - NOT PERFORMED HERE

```{r cluster merg}
sce$cluster_110_merg <- sce$cluster_110
sce$cluster_110_merg <- ifelse(sce$cluster_110_merg %in% c("10","11", "24"), "10_11_24", sce$cluster_110_merg)
sce$cluster_110_merg <- as.factor(sce$cluster_110_merg)

#define new and sorted cluster names
cluster_110_merg_df <- data.frame(cluster_110_merg = unique(sce$cluster_110_merg), cluster_110_merg_names = seq_along(unique(sce$cluster_110_merg))) 
sce$cluster_110_merg <- cluster_110_merg_df$cluster_110_merg_names[match(sce$cluster_110_merg,cluster_110_merg_df$cluster_110_merg)]
sce$cluster_110_merg <- as.factor(sce$cluster_110_merg)
```

# 3. Metaclustering and rich heatmap 

## Rich heatmap of clusters with additional information (other channels, spatial, grade, subtype)

```{r heatmap sce_1, message=FALSE, warning=FALSE}
### 1. Heatmap body ###

#Aggregrate across cells with the mean (median is also possible)
cluster_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_110,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$cluster_channel)

other_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_110,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$other_immune_channel)

# Heatmap body colors 
col_exprs <- colorRamp2(c(0,1,2,3), c("#440154FF", "#31688EFF", "#35B779FF", "#FDE725FF"))


### 2. Heatmap annotation ###
###2.1 Metacluster 

library(clValid)

#Internal measures
set.seed(22)
testCL<- clValid(t(assay(cluster_mean_sce, "exprs")),nClust = 4:20,clMethods = c("hierarchical","agnes","kmeans","pam"),validation="internal",metric="euclidean", method = "ward")
summary(testCL)
plot(testCL, legend = FALSE)

#Stability measures 
testCL_1<- clValid(t(assay(cluster_mean_sce, "exprs")),nClust = 4:20,clMethods = c("hierarchical","agnes","kmeans","pam"),validation="stability",metric="euclidean", method = "ward")
summary(testCL_1)
plot(testCL_1, legend = FALSE)

#chosen method: hierarchical, k=8

library(dendextend)
avg_dend_obj_MCP <- as.dendrogram(agnes(t(assay(cluster_mean_sce, "exprs")),metric = "euclidean",method = "ward"))
avg_col_dend_MCP <- color_branches(avg_dend_obj_MCP, k = 8)
plot(avg_col_dend_MCP)

#retrieve row_order
set.seed(22)
ht_kmeans <- Heatmap(t(assay(cluster_mean_sce, "exprs")),
        column_title = "cluster_channels",
        col = col_exprs,
        name= "mean exprs",
        show_row_names = T, 
        show_column_names =  T, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2",
        #cluster_rows = avg_col_dend_MCP, 
        row_split = 9,
        #row_km = 8, row_km_repeats = 10000,
        cluster_row_slices = FALSE)
  
ht_kmeans <- draw(ht_kmeans)
row_order_ht <- row_order(ht_kmeans)

anno <- colData(cluster_mean_sce) %>% as.data.frame %>% select(cluster_110,ncells)
anno$pre_metacluster <- NA
foreach(i = seq_along(row_order_ht))%do%{
anno$pre_metacluster <- as.factor((ifelse(anno$cluster_110 %in% row_order_ht[[i]], i,anno$pre_metacluster)))
}

## pre_metacluster_1 - splitting 
library(dendextend)
dend_1 <- row_dend(ht_kmeans)[[1]]
dend_1 <- color_branches(dend_1, k=3)
plot(dend_1)

anno <- anno %>% mutate(metacluster = ifelse(pre_metacluster == "1" & cluster_110 %in% c("39","43","40","42"), "10", 
                                       ifelse(pre_metacluster == "1" & cluster_110 %in% c("8","28","15","11","5","38"), "11", 
                                              ifelse(pre_metacluster == "1" & cluster_110 %in% c("13","22","24","12","53","54"), "1", pre_metacluster))))
anno$metacluster <- factor(anno$metacluster)

#add (pre) metacluster to sce
sce$pre_metacluster <- anno$pre_metacluster[match(sce$cluster_110,anno$cluster_110)]
sce$metacluster <- anno$metacluster[match(sce$cluster_110,anno$cluster_110)]

### 2.2 Spatial features

#Add number of neighbors to sce object (saved in ColPair)
n_neighbors <- colPair(sce) %>% as.data.frame %>% group_by(from) %>% dplyr::count() %>% arrange(desc(n))

sce$n_neighbors <- n_neighbors$n[match(seq_along(colnames(sce)),n_neighbors$from)]
sce$n_neighbors <- sce$n_neighbors %>% replace_na(0)

#Select spatial features and average over clusters 
spatial <- colData(sce) %>% as.data.frame %>% select(cluster_110, area, eccentricity, n_neighbors)
spatial <- spatial %>% select(-cluster_110) %>% aggregate(by = list(cluster_110 = spatial$cluster_110), FUN = mean) %>% column_to_rownames("cluster_110")

#Create Spatial Heatmap Annotation
ha_spatial <- HeatmapAnnotation(
    area = spatial$area,
    eccentricity = spatial$eccentricity,
    n_neighbors = spatial$n_neighbors, 
    col = list(area = metadata(sce)$colors$area,
               eccentricity = metadata(sce)$colors$eccentricity,
               n_neighbors = metadata(sce)$colors$n_neighbors),
    border = TRUE,
    which = "row")


###2.3 Metadata features

#Number of cells per cluster 
n_cells <- colData(cluster_mean_sce) %>% as.data.frame %>% select(cluster_110, ncells)

# Proportion of grade per cluster
grade <- colData(sce) %>% as.data.frame() %>% select(cluster_110,grade) %>% group_by(cluster_110) %>% table() %>% as.data.frame
grade <- grade %>% group_by(cluster_110) %>% mutate(fra = Freq/sum(Freq)) 
grade <- grade %>% select(-Freq) %>% pivot_wider(id_cols = cluster_110,names_from = grade,values_from = fra) %>% column_to_rownames("cluster_110")

# Proportion of pT stage per cluster
pT <- colData(sce) %>% as.data.frame() %>% select(cluster_110,pT_simple) %>% group_by(cluster_110) %>% table() %>% as.data.frame
pT <- pT %>% group_by(cluster_110) %>% mutate(fra = Freq/sum(Freq)) 
pT <- pT %>% select(-Freq) %>% pivot_wider(id_cols = cluster_110,names_from = pT_simple,values_from = fra) %>% column_to_rownames("cluster_110")

# Proportion of pN stage per cluster
pN <- colData(sce) %>% as.data.frame() %>% select(cluster_110,pN_simple) %>% group_by(cluster_110) %>% table() %>% as.data.frame
pN <- pN %>% group_by(cluster_110) %>% mutate(fra = Freq/sum(Freq)) 
pN <- pN %>% select(-Freq) %>% pivot_wider(id_cols = cluster_110,names_from = pN_simple,values_from = fra) %>% column_to_rownames("cluster_110")

# Number of contributing patients
cluster_PID <- colData(sce) %>% as.data.frame() %>% select(cluster_110,PID) %>% group_by(cluster_110) %>% table() %>% as.data.frame
n_PID <- cluster_PID %>% filter(Freq>0) %>% group_by(cluster_110) %>% count(name = "n_PID") %>% column_to_rownames("cluster_110")

# Cluster categories contributing to clusters
cluster_CC <- colData(sce) %>% as.data.frame() %>% select(cluster_110,cluster_category) %>% group_by(cluster_110) %>% table() %>% as.data.frame
cluster_CC <- cluster_CC %>% group_by(cluster_110) %>% mutate(fra = Freq/sum(Freq)) 
cluster_CC <- cluster_CC %>% select(-Freq) %>% pivot_wider(id_cols = cluster_110,names_from = cluster_category,values_from = fra) %>% column_to_rownames("cluster_110")

# Define color scheme for clusters and metaclusters
row_order_clusters <- row_order(Heatmap(t(assay(cluster_mean_sce, "exprs")),
        column_title = "cluster_channels",
        col = col_exprs,
        name= "mean exprs",
        km = 1,
        show_row_names = T, 
        show_column_names =  T, 
        cluster_row_slices = FALSE,
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2"))

cluster_col <- setNames(colorRampPalette(brewer.pal(9, "Paired"))(length(unique(anno$cluster_110))),row_order_clusters)
metacluster_col <- setNames(colorRampPalette(brewer.pal(9, "Paired"))(length(unique(anno$metacluster))),order(unique(anno$metacluster)))                                             
metadata(sce)$colors$cluster_110 <- cluster_col
metadata(sce)$colors$metacluster <- metacluster_col

# Metadata annotation
ha_meta <- HeatmapAnnotation(cluster_110 = anno_simple(unfactor(anno$cluster_110), pch = unfactor(anno$cluster_110),width = unit(8, "mm"), border=TRUE,col=metadata(sce)$colors$cluster_110),
                            metacluster = anno_simple(unfactor(anno$metacluster), pch = unfactor(anno$metacluster),width = unit(8, "mm"), border=TRUE,col=metadata(sce)$colors$metacluster),
                            n_cells = anno_barplot(n_cells$ncells, width = unit(12, "mm"),gp = gpar(fill = "#440154FF")),
                            n_PID = anno_barplot(n_PID, width = unit(12, "mm"),gp = gpar(fill = "#440154FF")),
                            grade = anno_barplot(grade, width = unit(15, "mm"),gp = gpar(fill = metadata(sce)$colors$grade)),
                            pT = anno_barplot(pT, width = unit(15, "mm"),gp = gpar(fill = metadata(sce)$colors$pT_simple)),
                            pN = anno_barplot(pN, width = unit(15, "mm"),gp = gpar(fill = metadata(sce)$colors$pN_simple)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(2,"mm"),
                            which = "row")
                            
### 3. Plot heatmap ###

h_meta <- Heatmap(t(assay(cluster_mean_sce, "exprs")),
        column_title = "cluster_channels",
        col = col_exprs,
        name = "mean exprs",
        row_title = NULL, 
        show_column_names =  TRUE, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2",
        row_split = anno$metacluster,
        cluster_row_slices = FALSE)+
  ha_meta+
  ha_spatial

draw(h_meta)

#6.3 Cluster annotation heatmap
# h_cluster_anno <- HeatmapAnnotation(cluster_110 = anno_simple(row_order_clusters, pch = as.character(row_order_clusters),border=TRUE,col=metadata(sce)$colors$cluster_110),
#                             metacluster = anno_simple(unfactor(anno$metacluster[row_order_clusters]), 
#                                                       pch= unfactor(anno$metacluster[row_order_clusters]),border=TRUE,col=metadata(sce)$colors$metacluster),
#                             border = TRUE, 
#                             annotation_name_rot = 90,
#                             gap = unit(1,"mm"),
#                             which = "row")+NULL
# 
# draw(h_cluster_anno)
```

```{r save object}
#Save sce after metacluster assignment
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor_new.rds")

sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor_new.rds")
```

# 4. Single-cell visualization

## Dimensionality reduction 

```{r dimred, message=FALSE,warning=FALSE}
## Subsample 10% of cells from each core 
#to save time and not get super crowded dimensionality reduction plots
set.seed(22)
sub <- colData(sce) %>% as.data.frame() %>% group_by("sample_id") %>% slice_sample(prop = 0.1) %>% arrange("sample_id") %>% mutate(sub_id = paste(sample_id,ObjectNumber,sep="_")) %>% pull(sub_id)
sce_sub <- sce[,sub]

#Run UMAP/TSNE on cluster channels
sce_sub <- runUMAP(sce_sub[rowData(sce_sub)$cluster_channel,], n_neighbors = 100, pca = 50, exprs_values = "exprs",external_neighbors=TRUE, BPPARAM = MulticoreParam())

sce_sub <- runTSNE(sce_sub[rowData(sce_sub)$cluster_channel,], exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam())


## 1. Visualize UMAP and TSNE
UMAP_clus <- plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "cluster_110")+
  scale_color_manual(values = metadata(sce)$colors$cluster_10)
UMAP_metaclus <- plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "metacluster")+
  scale_color_manual(values = metadata(sce)$colors$metacluster)

UMAP_clus+UMAP_metaclus
  
TSNE_clus <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "cluster_110")+
  scale_color_manual(values = metadata(sce)$colors$cluster_110)
TSNE_metaclus <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "metacluster")+
  scale_color_manual(values = metadata(sce)$colors$metacluster)

TSNE_clus+TSNE_metaclus

#Visualize all marker expression
all_plots_UMAP <- lapply(c(rownames(sce_sub)[rowData(sce_sub)$cluster_channel]),
                    function(x){
                      p <- plotReducedDim(sce_sub, dimred = "UMAP", 
                                          colour_by = x, 
                                          by_exprs_values = "exprs",
                                          point_size = 0.5)
                      return(p)                    
                    })

library(cowplot)
exp_plots_UMAP <- plot_grid(plotlist = all_plots_UMAP)

all_plots_TSNE <- lapply(c(rownames(sce_sub)[rowData(sce_sub)$cluster_channel]),
                    function(x){
                      p <- plotReducedDim(sce_sub, dimred = "TSNE", 
                                          colour_by = x, 
                                          by_exprs_values = "exprs",
                                          point_size = 0.5)
                      return(p)                    
                    })

library(cowplot)
exp_plots_TSNE <- plot_grid(plotlist = all_plots_TSNE)
```

## Heatmap for subset of cells

```{r heatmap sub}
#Heatmap visualization
dittoHeatmap(sce_sub, 
             genes = rownames(sce_sub)[rowData(sce_sub)$cluster_channel],
             assay = "exprs", 
             scale = "none", 
             order.by = c("metacluster", "cluster_110"), 
             heatmap.colors = viridis(100), 
             annot.by = c("metacluster", "cluster_110"),
             annotation_colors = list(metacluster = metadata(sce_sub)$colors$metacluster, 
                                      cluster_110 = metadata(sce_sub)$colors$cluster_110)
)
             
```

## Violin plot visualization

```{r celltype violin, message=FALSE, fig.width=6, fig.height=10}
#Violin Plot - plotExpression
plotExpression(sce_sub, 
               features = rownames(sce_sub)[rowData(sce_sub)$cluster_channel],
               x = "metacluster", exprs_values = "exprs", 
               colour_by = "metacluster") +
    theme(axis.text.x =  element_text(angle = 90))+
    scale_color_manual(values = metadata(sce_sub)$colors$metacluster)
```

## Barplot visualization 

### Fractions - Visualization and correlation of cluster fractions

```{r cluster-pro, message = FALSE, warning = FALSE}
## 1. Fraction of metacluster/cluster per sample/PID
d_SID <- dittoBarPlot(sce, var = "metacluster", group.by = "sample_id")+
  scale_fill_manual(values = metadata(sce)$colors$metacluster)+
  theme(axis.text.x = element_blank())

d_PID <- dittoBarPlot(sce, var = "metacluster", group.by = "PID")+
  scale_fill_manual(values = metadata(sce)$colors$metacluster)+
  theme(axis.text.x = element_blank())

dittoBarPlot(sce, var = "cluster_110", group.by = "sample_id")+
  scale_fill_manual(values = metadata(sce)$colors$cluster_110)+
  theme(axis.text.x = element_blank())

dittoBarPlot(sce, var = "cluster_110", group.by = "PID")+
  scale_fill_manual(values = metadata(sce)$colors$cluster_110)+
  theme(axis.text.x = element_blank())

## 2. Fraction of sample/PID per metacluster/cluster
dittoBarPlot(sce, var = "PID", group.by = "metacluster")+
  scale_fill_manual(values = metadata(sce)$colors$PID)+
  guides(fill = "none")

dittoBarPlot(sce, var = "PID", group.by = "cluster_110")+
  scale_fill_manual(values = metadata(sce)$colors$PID)+
  guides(fill = "none")

dittoBarPlot(sce, var = "sample_id", group.by = "metacluster")+
  guides(fill = "none")

dittoBarPlot(sce, var = "sample_id", group.by = "cluster_110")+
  guides(fill = "none")


## 3. Correlation of metacluster/cluster fractions for each PID (2-4 samples per PID)
SID_cluster <- colData(sce) %>% as.data.frame() %>% select(cluster_110,sample_id) %>% group_by(sample_id) %>% table() %>% as.data.frame
SID_cluster <- SID_cluster %>% group_by(sample_id) %>% mutate(fra = Freq/sum(Freq))

SID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,sample_id) %>% group_by(sample_id) %>% table() %>% as.data.frame
SID_metacluster <- SID_metacluster %>% group_by(sample_id) %>% mutate(fra = Freq/sum(Freq))

# Cluster fraction correlation for each PID (2 samples per PID)
## by cluster
SID_cluster_cor <- SID_cluster
SID_cluster_cor$PID <- sce$PID[match(SID_cluster$sample_id,sce$sample_id)]

SID_cluster_cor <- foreach(i = unique(colData(sce)$PID),.combine = "rbind")%do%{
SID <- SID_cluster_cor %>% filter(PID %in% i) %>% select(-Freq,-PID) %>% pivot_wider(id_cols = cluster_110,names_from = sample_id,values_from = fra) %>% column_to_rownames("cluster_110")
cor <- cor(SID,method = "spearman")
cor[2]
}

SID_cluster_cor <- as.data.frame(SID_cluster_cor) 
SID_cluster_cor$PID <- unique(colData(sce)$PID) 
SID_cluster_cor$category <- "PID"

SID_cluster_cor_plot <- ggplot(SID_cluster_cor, aes(x = category, y = V1))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  ylab("Correlation of cluster fractions per PID (n=2-4)")+
  ggtitle(paste("Spearman rho = ",round(median(SID_cluster_cor$V1, na.rm = TRUE), digits = 2)))

SID_cluster_cor_plot

## by metacluster
SID_metacluster_cor <- SID_metacluster
SID_metacluster_cor$PID <- sce$PID[match(SID_metacluster$sample_id,sce$sample_id)]

SID_metacluster_cor <- foreach(i = unique(colData(sce)$PID),.combine = "rbind")%do%{
SID <- SID_metacluster_cor %>% filter(PID %in% i) %>% select(-Freq,-PID) %>% pivot_wider(id_cols = metacluster,names_from = sample_id,values_from = fra) %>% column_to_rownames("metacluster")
cor <- cor(SID,method = "spearman")
cor[2]
}

SID_metacluster_cor <- as.data.frame(SID_metacluster_cor) 
SID_metacluster_cor$PID <- unique(colData(sce)$PID) 
SID_metacluster_cor$category <- "PID"

SID_metacluster_cor_plot <- ggplot(SID_metacluster_cor, aes(x = category, y = V1))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  ylab("Correlation of metacluster fractions per PID (n=2-4)")+
  ggtitle(paste("Spearman rho = ",round(median(SID_metacluster_cor$V1, na.rm = TRUE), digits = 2)))

SID_metacluster_cor_plot

d_PID+SID_metacluster_cor_plot

```
### Complex Heatmap based fraction stacked barplot

```{r}
## PID metacluster fraction dataframe
PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster <- PID_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
cur_df <- PID_metacluster %>% pivot_wider(id_cols = "PID", names_from = "metacluster",values_from = "fra") %>% column_to_rownames("PID")

# Perform hierarchical clustering on fractions
dd <- dist(cur_df, method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
plot(hc)
row_sorted <- hc$order

# Define annotation dataframe
cur_df <- cur_df[row_sorted,]

anno <- colData(sce) %>% as.data.frame %>% select(PID, age_bin, grade, pT_simple, pN_simple, pM_simple) %>% unique()
anno <- anno[match(rownames(cur_df), anno$PID),]

# Plot annotation dataframe with ComplexHeatmap
fraction_anno <- HeatmapAnnotation(MC_fractions = anno_barplot(cur_df, gp=gpar(fill=metadata(sce)$colors$metacluster[colnames(cur_df)]),
                                   bar_width = 1,
                                   height = unit(25,"cm"),
                                   width = unit(11,"cm"),
                                   show_row_names = FALSE), 
                          PID = anno_simple(anno$PID, col=metadata(sce)$colors$PID, border=TRUE),
                          age = anno_simple(anno$age_bin, col = metadata(sce)$colors$age_bin, border = TRUE),
                          grade = anno_simple(anno$grade, col = metadata(sce)$colors$grade, border = TRUE),
                          pT = anno_simple(anno$pT_simple, col=metadata(sce)$colors$pT_simple, border=TRUE),
                          pN = anno_simple(anno$pN_simple, col=metadata(sce)$colors$pN_simple, border=TRUE),
                          pM = anno_simple(anno$pM_simple, col=metadata(sce)$colors$pM_simple, border = TRUE),
                          border = TRUE, 
                          #annotation_name_rot = 90,
                          gap = unit(1,"mm"),
                          which = "row")

fraction_anno + NULL
```



### Densities - Visualization and correlation of cluster densities

```{r density cor}
# 1. By metaclusters

## Density barplot

### For each sample
SID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,sample_id) %>% group_by(sample_id) %>% table() %>% as.data.frame

SID_metacluster$tissue_area <- sce$tissue_area[match(SID_metacluster$sample_id,sce$sample_id)]

SID_metacluster <- SID_metacluster %>% mutate(density = Freq/(tissue_area/1000000))

ggplot(SID_metacluster) +
  geom_tile(color = "black",aes(sample_id, density, fill = metacluster), position = "stack")+
  scale_fill_manual(values=metadata(sce)$colors$metacluster) + 
  theme_classic(base_size = 15)+
  ylab("Density")+
  xlab("PID")+
  theme(axis.text.x = element_blank())

### For each PID
PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metacluster$tissue_area <- sce$tissue_area[match(PID_metacluster$PID,sce$PID)]

PID_metacluster <- PID_metacluster %>% mutate(density = Freq/(tissue_area/1000000))

d_dens_PID <- ggplot(PID_metacluster) +
  geom_tile(color = "black",aes(PID, density, fill = metacluster), position = "stack")+
  scale_fill_manual(values=metadata(sce)$colors$metacluster) + 
  theme_classic(base_size = 15)+
  ylab("Density")+
  xlab("PID")+
  theme(axis.text.x = element_blank())


## Correlation of metacluster/cluster densities for each PID (2-4 samples per PID) - SAME AS FOR FRACTION CORRELATION?

### By metacluster density
SID_metacluster_cor <- SID_metacluster

SID_metacluster_cor$PID <- sce$PID[match(SID_metacluster_cor$sample_id,sce$sample_id)]

SID_metacluster_cor <- foreach(i = unique(colData(sce)$PID),.combine = "rbind")%do%{
SID <- SID_metacluster_cor %>% filter(PID %in% i) %>% select(-Freq,-PID) %>% pivot_wider(id_cols = metacluster,names_from = sample_id,values_from = density) %>% column_to_rownames("metacluster")
cor <- cor(SID,method = "pearson")
cor[2]
}

SID_metacluster_cor <- as.data.frame(SID_metacluster_cor) 
SID_metacluster_cor$PID <- unique(colData(sce)$PID) 
SID_metacluster_cor$category <- "PID"

SID_metacluster_dens_cor_plot <- ggplot(SID_metacluster_cor, aes(x = category, y = V1))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  ylab("Correlation of metacluster densities per PID (n=2-4)")+
  ggtitle(paste("Spearman rho = ",round(median(SID_metacluster_cor$V1, na.rm = TRUE), digits = 2)))

SID_metacluster_dens_cor_plot


### By cluster density
SID_immune_cluster <- colData(sce) %>% as.data.frame() %>% select(cluster_130,sample_id) %>% group_by(sample_id) %>% table() %>% as.data.frame

SID_immune_cluster$tissue_area <- sce$tissue_area[match(SID_immune_cluster$sample_id,sce$sample_id)]

SID_immune_cluster <- SID_immune_cluster %>% mutate(density = Freq/(tissue_area/1000000))

SID_cluster_dens_cor <- SID_immune_cluster

SID_cluster_dens_cor$PID <- sce$PID[match(SID_cluster_dens_cor$sample_id,sce$sample_id)]

SID_cluster_dens_cor <- foreach(i = unique(colData(sce)$PID),.combine = "rbind")%do%{
SID <- SID_cluster_dens_cor %>% filter(PID %in% i) %>% select(-Freq,-PID,-tissue_area) %>% pivot_wider(id_cols = cluster_130,names_from = sample_id,values_from = density) %>% column_to_rownames("cluster_130")
cor <- cor(SID,method = "spearman")
cor[2]
}

SID_cluster_dens_cor <- as.data.frame(SID_cluster_dens_cor) 
SID_cluster_dens_cor$PID <- unique(colData(sce)$PID) 
SID_cluster_dens_cor$category <- "PID"

SID_cluster_dens_cor_plot <- ggplot(SID_cluster_dens_cor, aes(x = category, y = V1))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  ylab("Correlation of cluster densities per PID (n=2-4)")+
  ggtitle(paste("Spearman rho = ",round(median(SID_cluster_dens_cor$V1, na.rm = TRUE), digits = 2)))

SID_cluster_dens_cor_plot

(d_PID+SID_metacluster_cor_plot)/
(d_dens_PID+SID_metacluster_dens_cor_plot)

```
## Scatter-plot visualization 

```{r}
# Focus on basoluminal cells
cur_dat <- t(assay(sce[,sce$metacluster == "8"],"exprs")) %>% as.data.frame %>% select("CK5","CK7","CK8_18","CD15")

scatter_CK5_7 <- ggplot(cur_dat, aes(x=CK7, y=CK5)) +
  geom_point(alpha = 0.5)+
  stat_density_2d(aes(fill = ..level..), geom="polygon", contour_var = "count", alpha = 0.5)+
  theme_classic()

scatter_CK5_8_18 <- ggplot(cur_dat, aes(x=CK8_18, y=CK5)) +
  geom_point(alpha = 0.5)+
  stat_density_2d(aes(fill = ..level..), geom="polygon", contour_var = "count", alpha = 0.5)+
  theme_classic()

scatter_CK5_7 + scatter_CK5_8_18
```

## CATALYST-based visualization

We will first convert the sce into a CATALYST-compatible format.

```{r CATALYST}
library(CATALYST)
# save spe in CATALYST-compatible object with renamed colData entries and 
# new metadata information
sce_cat <- sce 
sce_cat$sample_id <- factor(sce$PID)
sce_cat$cluster_id <- factor(sce$metacluster)

#add celltype information to metadata
metadata(sce_cat)$cluster_codes <- data.frame(metacluster = factor(sce_cat$metacluster))
```

```{r CATALYST pbmds and clrDR, message=FALSE}
# MDS pseudobulk by metacluster
pbMDS(sce_cat, by = "cluster_id", 
      features = rownames(sce_cat)[rowData(sce_cat)$cluster_channel], 
      label_by = "cluster_id", k = "metacluster") +
  scale_color_manual(values = metadata(sce_cat)$colors$metacluster)+
  theme_classic()

# MDS pseudobulk by metacluster
pbMDS(sce_cat, by = "both", 
      features = rownames(sce_cat)[rowData(sce_cat)$cluster_channel], 
      k = "metacluster",
      size_by = TRUE) +
  scale_color_manual(values = metadata(sce_cat)$colors$metacluster)+
  theme_classic()

# CLR of patient proportions per metacluster
clrDR(sce_cat, dr = "PCA", 
      by = "cluster_id", k = "metacluster", 
      label_by = "cluster_id", arrow_col = "PID", 
      point_pal = metadata(sce_cat)$colors$metacluster, 
      arrows = FALSE) +
  scale_color_manual(values = metadata(sce_cat)$colors$PID)+
  guides(color = "none")+
  theme_classic()

# CLR of metacluster proportions per patient
clrDR(sce_cat, dr = "PCA", 
      by = "sample_id", point_col = "sample_id",
      k = "metacluster", point_pal = metadata(sce_cat)$colors$PID)+
  scale_color_manual(values = metadata(sce_cat)$colors$metacluster)+
  guides(fill = "none")+
  theme_classic()
```

## Check correlation of markers

```{r marker correlations}
# Cluster channel marker correlation (all markers - does not show distribution which is usually recommended)
cor_sce <- as.data.frame(t(assay(sce,"exprs")[rowData(sce)$cluster_channel,]))

corrplot(cor(cor_sce,method = "spearman"), method="color", col=brewer.pal(n=8, name="RdBu"),  
         type="upper", order="hclust",hclust.method = "ward.D2", 
         addCoef.col = "black", addCoefasPercent = F,# Add coefficient of correlation
         tl.col="black", tl.srt=40, #Text label color and rotation
         # Combine with significance
         #p.mat = p.mat, sig.level = 0.01, insig = "blank", 
         # hide correlation coefficient on the principal diagonal
         diag=F
)

# For specific markers of interest (also shows the value distribution which is recommended)
#pairs.panels(cor_sce[,c("CK5","KRT14")],
#             method = "spearman",
#             hist.col = "royalblue1",
#             digits = 2)
```

## 5. Tumor heterogeneity analysis 

### Heatmap for cell count per cluster/metacluster

```{r PID/cluster cells viz}
# 1. Cluster
library(tidyverse)
for_plot <- colData(sce) %>% as_tibble() %>%
    group_by(PID, cluster_110) %>%
    summarize(count = n()) %>%
    filter(count >= 5) %>%
    mutate(log_count = log(count)) %>%
    pivot_wider(id_cols = PID, names_from = cluster_110, 
                values_from = log_count, values_fill = 0) %>%
    ungroup() %>%
    select(-PID)

pheatmap(for_plot, #color = colorRampPalette(c("dark blue", "white", "dark red"))(100), 
         scale = "none", show_rownames = TRUE)

# 2. Metacluster
library(tidyverse)
for_plot_MC <- colData(sce) %>% as_tibble() %>%
    group_by(PID, metacluster) %>%
    summarize(count = n()) %>%
      filter(count >= 5) %>%
    mutate(log_count = log(count)) %>%
    pivot_wider(id_cols = PID, names_from = metacluster, 
                values_from = log_count, values_fill = 0) %>%
    ungroup() %>%
    select(-PID)

#Heatmap(for_plot_MC, 
#        name = "log(cell_count)",
#        col = viridis(100)
#        )
```

### Heterogeneity scores

```{r}
## 1. For Metaclusters
PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metacluster <- PID_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Number of metaclusters per PID 
PID_metacluster_p <- PID_metacluster %>% filter(Freq > 5) %>% group_by(PID) %>% count(name = "n_MC") %>% mutate(category = "PID")

n_MC_PID <- ggplot(PID_metacluster_p, aes(x = category, y = n_MC))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  coord_cartesian(ylim = c(0,11))+
  theme(axis.title.x = element_blank())+
  ylab("Number of tumor metaclusters per PID")

n_MC_PID

### Calculate shannon entropy per PID
library(vegan)

shannon_df <- PID_metacluster %>% pivot_wider(id_cols = "PID",names_from = "metacluster",values_from = "Freq") %>% column_to_rownames("PID")

shannon_plot <- data.frame(shannon_div = diversity(shannon_df),
                           number = 11,
                           shannon_equi = (diversity(shannon_df)/log(11)),
                           PID = rownames(shannon_df), 
                           category = "PID")

shannon_plot <- shannon_plot %>% arrange(desc(shannon_equi)) %>% mutate(heterogeneity = ntile(shannon_equi,3)) %>% mutate(heterogeneity_score = ifelse(heterogeneity == 3, "High",                                                                                                                                        ifelse(heterogeneity == 2, "Medium", "Low")))

shannon_MC_PID <- ggplot(shannon_plot, aes(x = category, y = shannon_equi))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  coord_cartesian(ylim = c(0,1))+
  ylab("Shannon Equitability Index of tumor metaclusters per PID")

n_MC_PID/shannon_MC_PID
```


```{r}
## 2. For clusters
PID_cluster_110 <- colData(sce) %>% as.data.frame() %>% select(cluster_110,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_cluster_110 <- PID_cluster_110 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Number of clusters per PID 
PID_cluster_110_p <- PID_cluster_110 %>% filter(Freq > 5) %>% group_by(PID) %>% count(name = "n_MC") %>% mutate(category = "PID")

n_clus_PID <- ggplot(PID_cluster_110_p, aes(x = category, y = n_MC))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  coord_cartesian(ylim = c(0,60))+
  theme(axis.title.x = element_blank())+
  ylab("Number of tumor clusters per PID")

n_clus_PID

### Calculate shannon entropy per PID
library(vegan)

shannon_df <- PID_cluster_110 %>% pivot_wider(id_cols = "PID",names_from = "cluster_110",values_from = "Freq") %>% column_to_rownames("PID")

shannon_plot <- data.frame(shannon_div = diversity(shannon_df),
                           number = 11,
                           shannon_equi = (diversity(shannon_df)/log(11)),
                           PID = rownames(shannon_df), 
                           category = "PID")

shannon_plot <- shannon_plot %>% arrange(desc(shannon_equi)) %>% mutate(heterogeneity = ntile(shannon_equi,3)) %>% mutate(heterogeneity_score = ifelse(heterogeneity == 3, "High",                                                                                                                                        ifelse(heterogeneity == 2, "Medium", "Low")))

shannon_clus_PID <- ggplot(shannon_plot, aes(x = category, y = shannon_equi))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  coord_cartesian(ylim = c(0,1))+
  ylab("Shannon Equitability Index of tumor clusters per PID")

n_clus_PID/shannon_clus_PID
```


### Euclidian distance analysis 

```{r}
## Euclidian distance analysis

### 1. NOT clr transformed metacluster fractions

SID_metacluster_cor <- SID_metacluster
SID_metacluster_cor$PID <- sce$PID[match(SID_metacluster$sample_id,sce$sample_id)]

## Euclidian metacluster distance between metacluster fractions of the same PID 
PID_image_1 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 1) %>% pull(PID)
PID_image_2 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 2) %>% pull(PID)
PID_image_4 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 4) %>% pull(PID)

out <- sapply(unique(colData(sce)$PID), function(x){
  
  SID <- SID_metacluster_cor %>% filter(PID %in% x) %>% select(-Freq,-PID) %>% pivot_wider(id_cols = metacluster,names_from = sample_id,values_from = fra) %>%
    column_to_rownames("metacluster")
  
  if(x %in% PID_image_2){
    eu <- dist(t(SID))
    }else if(x %in% PID_image_4){
      eu <- dist(t(SID))
      eu <- mean(eu)
      }
  
  return(eu)
})

out_vec <- unlist(out, use.names = FALSE)
out_image <- data_frame(dist = out_vec, category = "Image")

## Mean euclidian distance between PID and all other PIDs
test <- PID_metacluster %>% pivot_wider(id_cols = "PID",names_from = "metacluster",values_from = "fra") %>% column_to_rownames("PID")
test_1 <- as.data.frame(as.matrix(dist(test)))
test_1[test_1 == 0] <- NA 

out_PID <- rowMeans(test_1, na.rm = TRUE)
out_PID <- data.frame(dist = out_PID, category = "PID")

## Plot 
out_plot <- rbind(out_image, out_PID)

ggplot(out_plot, aes(x = dist))+
  geom_density(aes(fill = category), alpha = 0.5)+
  geom_rug(aes(color = category))+
  theme_classic()+
  ggtitle("Mean euclidian distance of metacluster fractions", 
          subtitle = paste0("p = ",compare_means(dist ~ category, out_plot)$p.format, " (Wilcoxon)"))+
  scale_fill_brewer(palette = "Paired")+
  scale_color_brewer(palette = "Paired")

### 2. clr-transformed metacluster fractions

SID_metacluster_cor <- SID_metacluster
SID_metacluster_cor$PID <- sce$PID[match(SID_metacluster$sample_id,sce$sample_id)]

## Euclidian metacluster distance between images of the same PID 

PID_image_1 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 1) %>% pull(PID)
PID_image_2 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 2) %>% pull(PID)
PID_image_4 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 4) %>% pull(PID)

out <- sapply(unique(colData(sce)$PID), function(x){
  
  SID <- SID_metacluster_cor %>% filter(PID %in% x) %>% select(-Freq,-PID) %>% pivot_wider(id_cols = metacluster,names_from = sample_id,values_from = fra) %>%
    column_to_rownames("metacluster")
  SID <- clr(t(SID))
  
  if(x %in% PID_image_2){
    eu <- dist(SID)
    }else if(x %in% PID_image_4){
      eu <- dist(SID)
      eu <- mean(eu)
      }
  
  return(eu)
})

out_vec <- unlist(out, use.names = FALSE)
out_image <- data_frame(dist = out_vec, category = "Image", PID = names(out_vec))

## Mean euclidian distance between PID and all other PIDs
test <- PID_metacluster %>% pivot_wider(id_cols = "PID",names_from = "metacluster",values_from = "fra") %>% column_to_rownames("PID")
test <- clr(test)
test_1 <- as.data.frame(as.matrix(dist(test)))
test_1[test_1 == 0] <- NA 

out_PID <- rowMeans(test_1, na.rm = TRUE)
out_PID <- data.frame(dist = out_PID, category = "PID", PID = names(out_PID))


## Plot 
out_plot_1 <- rbind(out_image, out_PID)

ggplot(out_plot_1, aes(x = dist))+
  geom_density(aes(fill = category), color = "black",alpha = 0.5)+
  #geom_rug(aes(color = PID))+
  geom_rug(aes(color = category), alpha = 0.5)+
  theme_classic()+
  ggtitle("Mean pairwise euclidian distance of metacluster fractions", 
          subtitle = paste0("p = ",compare_means(dist ~ category, out_plot,paired = TRUE)$p.format, " (Wilcoxon)"))+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")+
  guides(color = "none")+
  coord_cartesian(clip = "off")
```

### Dominant phenotype analysis 

```{r}
SID_metacluster_dom <- SID_metacluster
SID_metacluster_dom$PID <- sce$PID[match(SID_metacluster$sample_id,sce$sample_id)]

## Dominant metacluster analysis 
PID_image_1 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 1) %>% pull(PID)
PID_image_2 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 2) %>% pull(PID)
PID_image_4 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 4) %>% pull(PID)


out_test <- foreach(i = unique(sce$sample_id),.combine = "rbind")%do%{
out <- SID_metacluster_dom %>% filter(sample_id %in% i) %>% select(-Freq) %>% arrange(desc(fra)) %>% slice(1)
}

out_two <- out_test %>% arrange(PID) %>% filter(PID %in% PID_image_2) %>% group_by(PID) %>% distinct(metacluster) %>% count(PID) %>% mutate(dominant = ifelse(n == "1", TRUE, FALSE)) %>% ungroup() 

out_four <- out_test %>% arrange(PID) %>% filter(PID %in% PID_image_4) %>% group_by(PID) %>% distinct(metacluster) %>% count(PID) %>% mutate(dominant = ifelse(n %in% c("1","2"), TRUE, FALSE)) %>% ungroup()

out_all <- rbind(out_two, out_four)

out_bar <- out_all %>% count(dominant) %>% mutate(fra = n/sum(n), category = "PID")

ggplot(out_bar, aes(x = category, y = n, fill = dominant))+
  geom_tile(position = "stack",color = "black")+#+alpha = 0.5)+
  theme_classic()+
  ylab("Number of PID")+
  scale_fill_brewer(palette = "Set1")


PID_true <- out_all %>% filter(dominant == TRUE) %>% pull(PID)

test <- out_test %>% arrange(PID) %>% filter(PID %in% PID_true) %>% group_by(PID) %>% mutate(mean_fra = mean(fra), dif_fra = max(fra)-min(fra))

test_plot <- test %>% select(PID, mean_fra, dif_fra)


ggplot(test_plot, aes(mean_fra, dif_fra, color = PID))+
  geom_point(size = 3)+
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  guides(color = "none")+
  theme_classic()
  
```

```{r}
SID_cluster_dom <- SID_cluster
SID_cluster_dom$PID <- sce$PID[match(SID_cluster$sample_id,sce$sample_id)]


## Dominant cluster analysis 
PID_image_1 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 1) %>% pull(PID)
PID_image_2 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 2) %>% pull(PID)
PID_image_4 <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id) %>% unique() %>% group_by(PID) %>% count() %>% filter(n == 4) %>% pull(PID)


out_test <- foreach(i = unique(sce$sample_id),.combine = "rbind")%do%{
out <- SID_cluster_dom %>% filter(sample_id %in% i) %>% select(-Freq) %>% arrange(desc(fra)) %>% slice(1)
}

out_two <- out_test %>% arrange(PID) %>% filter(PID %in% PID_image_2) %>% group_by(PID) %>% distinct(cluster_110) %>% count(PID) %>% mutate(dominant = ifelse(n == "1", TRUE, FALSE)) %>% ungroup() 

out_four <- out_test %>% arrange(PID) %>% filter(PID %in% PID_image_4) %>% group_by(PID) %>% distinct(cluster_110) %>% count(PID) %>% mutate(dominant = ifelse(n %in% c("1","2"), TRUE, FALSE)) %>% ungroup()

out_all <- rbind(out_two, out_four)

out_bar <- out_all %>% count(dominant) %>% mutate(fra = n/sum(n), category = "PID")

ggplot(out_bar, aes(x = category, y = n, fill = dominant))+
  geom_tile(position = "stack", color = "black",alpha = 0.5)+
  theme_classic()+
  ylab("Number of PID")+
  scale_fill_brewer(palette = "Set1")


PID_true <- out_all %>% filter(dominant == TRUE) %>% pull(PID)

test <- out_test %>% arrange(PID) %>% filter(PID %in% PID_true) %>% group_by(PID) %>% mutate(mean_fra = mean(fra), dif_fra = max(fra)-min(fra))

test_plot <- test %>% select(PID, mean_fra, dif_fra)


ggplot(test_plot, aes(mean_fra, dif_fra, color = PID))+
  geom_point(size = 3)+
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  guides(color = "none")+
  theme_classic()
  
```


### Age and fraction correlation - MC8 / CK5/7 scores

```{r}
library(ggpmisc)
# Metacluster 8 
PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster <- PID_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_MC8 <- PID_metacluster %>% filter(metacluster == "8", fra > 0.05, Freq > 10) %>% arrange(desc(fra))

PID_MC8$age <- sce$age[match(PID_MC8$PID, sce$PID)]

ggplot(PID_MC8, aes(x = fra, y=age))+
  geom_point()+
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth", 
              color = "firebrick")+
  stat_poly_eq()+
  stat_fit_glance(method = 'lm',
                       method.args = list(formula = y ~ x),
                       geom = 'text',
                       aes(label = paste("P-value = ", signif(..p.value.., digits = 4), sep = "")))+
  theme_classic()
```

```{r}
library(ggpmisc)
# CK5/7+
PID_CK57 <- colData(sce) %>% as.data.frame() %>% select(CK5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_CK57 <- PID_CK57 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_CK57_pos <- PID_CK57 %>% filter(CK5_7_score == "CK5/7+", fra > 0.05, Freq > 10) %>% arrange(desc(fra))

PID_CK57_pos$age <- sce$age[match(PID_CK57_pos$PID, sce$PID)]

ggplot(PID_CK57_pos, aes(x = fra, y=age))+
  geom_point()+
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth", 
              color = "firebrick")+
  stat_poly_eq()+
  stat_fit_glance(method = 'lm',
                       method.args = list(formula = y ~ x),
                       geom = 'text',
                       aes(label = paste("P-value = ", signif(..p.value.., digits = 4), sep = "")))+
  theme_classic()
```

## 6. CK5/7 Signature 

### Define signature / PID groups 

```{r}
## 5. CK5-7 signature
# 1. CK5
GMM_CK5 <- Mclust(assay(sce, "exprs")["CK5",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK5$classification == 2 & GMM_CK5$uncertainty == 0, 
                                        "CK5+",
                                        "CK5-"), 
                     KRT5_exprs = assay(sce, "exprs")["CK5",],
                     uncertainty = GMM_CK5$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = KRT5_exprs, fill = class_epi))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")
  
#Add GMM epi class to SCE
sce$GMM_tumor_CK5 <- GMM_df$class_epi

# 2. CK7
GMM_CK7 <- Mclust(assay(sce, "exprs")["CK7",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK7$classification == 2 & GMM_CK7$uncertainty == 0, 
                                        "CK7+",
                                        "CK7-"), 
                     KRT7_exprs = assay(sce, "exprs")["CK7",],
                     uncertainty = GMM_CK7$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = KRT7_exprs, fill = class_epi))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")
  
#Add GMM epi class to SCE
sce$GMM_tumor_CK7 <- GMM_df$class_epi

## Combine CK5/7 GMM scores 
sce$CK5_7_score <- ifelse(sce$GMM_tumor_CK5 == "CK5+" & sce$GMM_tumor_CK7 == "CK7+", "CK5/7+", 
       ifelse(sce$GMM_tumor_CK5 == "CK5+" & sce$GMM_tumor_CK7 == "CK7-", "CK5+",
              ifelse(sce$GMM_tumor_CK5 == "CK5-" & sce$GMM_tumor_CK7 == "CK7+", "CK7+","CK5/7-")))

## Visualize 
cur_dat <- t(assay(sce,"exprs")) %>% as.matrix %>% as.data.frame %>% dplyr::select("CK5","CK7")
cur_dat$CK5_7_score <- sce$CK5_7_score[match(rownames(cur_dat), sce$cell_id)]
cur_dat$GMM_tumor_CK5 <- sce$GMM_tumor_CK5[match(rownames(cur_dat), sce$cell_id)]
cur_dat$GMM_tumor_CK7 <- sce$GMM_tumor_CK7[match(rownames(cur_dat), sce$cell_id)]


library(wesanderson)

metadata(sce)$colors$CK5_7_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("CK5/7-", "CK5/7+", "CK5+", "CK7+"))

main <- ggplot(cur_dat, aes(x=CK7, y=CK5, color = CK5_7_score)) +
  geom_point(alpha = 0.5)+
  theme_classic()+
  xlab("CK7 exprs")+
  ylab("CK5 exprs")+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$CK5_7_score)

y_side <- ggplot(cur_dat, aes(x = CK5, fill=GMM_tumor_CK5)) + 
  geom_density(alpha=0.8) + 
  scale_fill_manual(values = c('#999999','#F4B5BD')) + 
  theme_classic()+
  theme(legend.position = "none", axis.title.x = element_blank())+
  coord_flip()

x_side <- ggplot(cur_dat, aes(x = CK7, fill=GMM_tumor_CK7)) + 
  geom_density(alpha=0.8) + 
  scale_fill_manual(values = c('#999999','#9C964A')) + 
  theme_classic()+
  theme(legend.position = "none", axis.title.x = element_blank())

blankPlot <- ggplot()+geom_blank(aes(1,1)) + 
  cowplot::theme_nothing()

library(gridExtra)
scatter_final <- grid.arrange(x_side, blankPlot, main, y_side, 
        ncol=2, nrow=2, widths=c(4, 1.4), heights=c(1.4, 4))

plot(scatter_final)
```


```{r}
## Define patient groups 
plot_score_PID <- colData(sce) %>% as.data.frame %>% group_by(PID,CK5_7_score) %>% count() %>% group_by(PID) %>% mutate(fra = n/sum(n))

ggplot(plot_score_PID, aes(x=PID, y=fra, fill = CK5_7_score))+
  geom_tile(position = "fill")

plot_heat_PID <- plot_score_PID %>% pivot_wider(id_cols = PID, values_from = fra, names_from = CK5_7_score) %>% column_to_rownames("PID")
plot_heat_PID[is.na(plot_heat_PID)] <- 0

testCL<- clValid(plot_heat_PID,nClust = 2:20,clMethods = c("hierarchical","agnes"),validation="internal",metric="euclidean", method = "ward")
plot(testCL, legend = FALSE)

dev.off()

h <- Heatmap(plot_heat_PID,
             name = "Proportion",
             col = viridis(100), 
             row_split=4,
             show_row_names = FALSE)

h <- draw(h)

row_order_h <- row_order(h)

anno <- colData(sce) %>% as.data.frame %>% dplyr::select(PID) %>% unique()
anno$CK5_7_group <- NA

for(i in seq_along(row_order_h)){
anno$CK5_7_group <- as.factor(ifelse(anno$PID %in% rownames(plot_heat_PID)[row_order_h[[i]]], i, anno$CK5_7_group))
}

sce$CK5_7_group <- anno$CK5_7_group[match(sce$PID, anno$PID)]
```

```{r}
## Dominant CK5/7 score per CK5/7 group 
# Dominant == highest fraction 
score_PID <- colData(sce) %>% as.data.frame %>% group_by(PID,CK5_7_score) %>% count() %>% group_by(PID) %>% mutate(fra = n/sum(n))
score_PID <- score_PID %>% group_by(PID) %>% mutate(dom = max(fra)) %>% filter(dom == fra) 

# Add to sce 
sce$dom_CK57_score <- score_PID$CK5_7_score[match(sce$PID, score_PID$PID)]

# Subset sce
cell_id_out <- colData(sce) %>% as.data.frame %>% select(PID,dom_CK57_score, CK5_7_group) %>% unique() %>% rownames_to_column("cell_id") %>% pull("cell_id")
sce_PID <- sce[,sce$cell_id %in% cell_id_out]

# Plot 
dittoBarPlot(sce_PID, var = "dom_CK57_score", group.by = "CK5_7_group")+
  scale_fill_manual(values = metadata(sce)$colors$CK5_7_score)+
  ylab("Percent of patients")+
  ggtitle(NULL)
```


```{r}
## Plot fraction of each CK5/7 score per CK5/7 group 
plot_score_PID$CK5_7_group <- as.character(sce$CK5_7_group[match(plot_score_PID$PID,sce$PID)])

set.seed(10)
ggplot(plot_score_PID, aes(x = CK5_7_group, y = fra))+
  facet_wrap(~CK5_7_score)+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = CK5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce)$colors$CK5_7_score)+
  theme_classic()

set.seed(10)
ggplot(plot_score_PID, aes(x = CK5_7_score, y = fra))+
  facet_wrap(~CK5_7_group)+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = CK5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce)$colors$CK5_7_score)+
  stat_compare_means()+
  theme_classic()
```


### Kaplan Meier

```{r}
#Assess survival impact 
library(survival)
library(survminer)

cur_surv <- colData(sce) %>% as.data.frame() %>% select(PID, CK5_7_group, OS_months, DFS_months, status_DFS, status_OS) %>% unique() #%>% filter(CK5_7_group %in% c("1","4"))

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ CK5_7_group, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          #palette = "Set1",
          #log.rank.weights = "n"
)
```

### Cox for CK5/7 score

```{r}
library(survival)
library(survminer)

PID_CK57 <- colData(sce) %>% as.data.frame() %>% select(CK5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_CK57 <- PID_CK57 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_CK57$CK5_7_score <- str_replace(PID_CK57$CK5_7_score,"\\+","positive")
PID_CK57$CK5_7_score <- str_replace(PID_CK57$CK5_7_score,"\\/","_")
PID_CK57$CK5_7_score <- str_replace(PID_CK57$CK5_7_score,"\\-","negative")

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_CK57 %>% pivot_wider(id_cols = "PID",names_from = "CK5_7_score",values_from = "fra") %>% column_to_rownames("PID")
#colnames(PID_metagroups) <- paste0("CK57_",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -CK5_7negative) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+CK5_7negative"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratiDFS
#hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0,185))+
#scale_y_continuous(trans = "log10")
theme(axis.text.x = element_text(angle = 90))

#hr_prop
```

### Cox for CK5/7 groups

Multivariate 

```{r}
library(survival)
library(survminer)

PID_CK57 <- colData(sce) %>% as.data.frame() %>% select(CK5_7_group,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_CK57 <- PID_CK57 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_CK57 %>% pivot_wider(id_cols = "PID",names_from = "CK5_7_group",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- paste0("CK57_",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -CK57_4) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+CK57_4"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratiDFS
#hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0,40))+
#scale_y_continuous(trans = "log10")
theme(axis.text.x = element_text(angle = 90))

#hr_prop
```

Univariate 

```{r}
library(survival)
library(survminer)

PID_CK57 <- colData(sce) %>% as.data.frame() %>% select(CK5_7_group,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_CK57 <- PID_CK57 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_CK57 %>% pivot_wider(id_cols = "PID",names_from = "CK5_7_group",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- paste0("CK57_",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -CK57_4) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~", paste(covariates, collapse = " + "),"+CK57_4"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratiDFS
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0,10))+
#scale_y_continuous(trans = "log10")
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

# 7. Survival analysis 

## Cox-proportonal hazards model for metaclusters

### For OS on metaclusters

```{r cox}
library(survival)
library(survminer)

PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster <- PID_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_metacluster %>% pivot_wider(id_cols = "PID",names_from = "metacluster",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- paste0("MC",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_OS <- sce$status_OS[match(rownames(Cox_df),sce$PID)]
Cox_df$OS_months <- sce$OS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -MC1) %>% colnames

multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+ MC1"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop

## 2. Based on categories (mean division of cluster proportions)
Cox_df_mean <- foreach(i = Cox_df %>% select(-status_OS,-OS_months, -status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames,.combine = "cbind")%do%{
Cox_df %>% select(i) %>% mutate(mean = ifelse(Cox_df[,i] > mean(Cox_df[,i]),"B","A")) %>% pull(mean)
}

colnames(Cox_df_mean) <- Cox_df %>% select(-status_OS,-OS_months, -status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames
rownames(Cox_df_mean) <- Cox_df %>% select(-status_OS,-OS_months, -status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% rownames
Cox_df_mean <- as.data.frame(Cox_df_mean)

identical(rownames(Cox_df), rownames(Cox_df_mean))

Cox_df_mean <- cbind(Cox_df_mean, Cox_df %>% select(status_OS,OS_months, status_OS, OS_months, age, grade, pT_simple, pN_simple, pM_simple))

#Based on proportions
covariates <- Cox_df_mean %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_mean)

summary(multi_model)

multi_res <- summary(multi_model)

plot_df_mean <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_mean <- ggplot(plot_df_mean, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for mean")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_mean

## 3. Based on categories (GMM division of cluster proportions)
Cox_df_GMM <- foreach(i = Cox_df %>% select(-status_OS,-OS_months, -status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames,.combine = "cbind")%do%{
GMM_Cox <- Mclust(Cox_df[,i], G = 2)
as.character(GMM_Cox$classification)
}

colnames(Cox_df_GMM) <- Cox_df %>% select(-status_OS,-OS_months, -status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames
rownames(Cox_df_GMM) <- Cox_df %>% select(-status_OS,-OS_months, -status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% rownames
Cox_df_GMM <- as.data.frame(Cox_df_GMM,stringsAsFactors=TRUE)

identical(rownames(Cox_df), rownames(Cox_df_GMM))

Cox_df_GMM <- cbind(Cox_df_GMM, Cox_df %>% select(status_OS,OS_months, status_OS, OS_months, age, grade, pT_simple, pN_simple, pM_simple))

#Based on proportions
covariates <- Cox_df_GMM %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_GMM)

summary(multi_model)

summary(Cox_df_GMM)

multi_res <- summary(multi_model)

plot_df_GMM <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_GMM <- ggplot(plot_df_GMM, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for GMM")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop/
hr_mean/
hr_GMM
```


```{r OS survival_curves}
#Plot survival curves
fit <- survfit(Surv(OS_months, status_OS) ~ MC6, data = Cox_df_mean)
print(fit)

p1 <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = "Set1",
          log.rank.weights = "n")+
  ggtitle("Mean proportion")

fit <- survfit(Surv(OS_months, status_OS) ~ MC2, data = Cox_df_mean)
print(fit)

p2 <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = "Set1",
          log.rank.weights = "n")+
  ggtitle("Mean proportion")

arrange_ggsurvplots(list(p1,p2))
```

### For DFS on metaclusters

Multivariate 

```{r cox dfs}
library(survival)
library(survminer)

PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster <- PID_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_metacluster %>% pivot_wider(id_cols = "PID",names_from = "metacluster",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- paste0("MC",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -MC1) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+ MC1"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratiDFS
#hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0,185))+
#scale_y_continuous(trans = "log10")
theme(axis.text.x = element_text(angle = 90))

hr_prop

## 2. Based on categories (mean division of cluster proportions)

Cox_df_mean <- foreach(i = Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames,.combine = "cbind")%do%{
Cox_df %>% select(i) %>% mutate(mean = ifelse(Cox_df[,i] > mean(Cox_df[,i]),"B","A")) %>% pull(mean)
}

colnames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames
rownames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% rownames
Cox_df_mean <- as.data.frame(Cox_df_mean)

identical(rownames(Cox_df), rownames(Cox_df_mean))

Cox_df_mean <- cbind(Cox_df_mean, Cox_df %>% select(status_DFS,DFS_months, status_DFS, DFS_months, age, grade, pT_simple, pN_simple, pM_simple))

#Based on proportions
covariates <- Cox_df_mean %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_mean)

summary(multi_model)

multi_res <- summary(multi_model)

plot_df_mean <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_mean <- ggplot(plot_df_mean, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for mean")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_mean

## 3. Based on categories (GMM division of cluster proportions)
Cox_df_GMM <- foreach(i = Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames,.combine = "cbind")%do%{
GMM_Cox <- Mclust(Cox_df[,i], G = 2)
as.character(GMM_Cox$classification)
}

colnames(Cox_df_GMM) <- Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames
rownames(Cox_df_GMM) <- Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% rownames
Cox_df_GMM <- as.data.frame(Cox_df_GMM,stringsAsFactors=TRUE)

identical(rownames(Cox_df), rownames(Cox_df_GMM))

Cox_df_GMM <- cbind(Cox_df_GMM, Cox_df %>% select(status_DFS,DFS_months, status_DFS, DFS_months, age, grade, pT_simple, pN_simple, pM_simple))

#Based on proportions
covariates <- Cox_df_GMM %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_GMM)

summary(multi_model)

summary(Cox_df_GMM)

multi_res <- summary(multi_model)

plot_df_GMM <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratio
hr_GMM <- ggplot(plot_df_GMM, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for GMM")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_mean

hr_prop/
hr_mean/
hr_GMM
```

Univariate 

```{r cox dfs}
library(survival)
library(survminer)

PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster <- PID_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_metacluster %>% pivot_wider(id_cols = "PID",names_from = "metacluster",values_from = "fra") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- paste0("MC",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -MC1) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ ", paste(covariates, collapse = " + "),"+ MC1"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratiDFS
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0,50))+
#scale_y_continuous(trans = "log10")
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r OS survival_curves}
#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ MC8, data = Cox_df_mean)
print(fit)

p1 <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = "Set1")
p1
 
# Find best categorical cut-off 
res.cut <- surv_cutpoint(Cox_df, time = "DFS_months", event = "status_DFS",
                         variables = c("MC8"))
summary(res.cut)
plot(res.cut, "MC8", palette = "npg")

res.cat <- surv_categorize(res.cut)
fit <- survfit(Surv(DFS_months, status_DFS) ~MC8, data = res.cat)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = "Set1") 
  
  
  
  
  
  
  
fit <- survfit(Surv(DFS_months, status_DFS) ~ MC9, data = Cox_df_mean)
print(fit)

p2 <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = "Set1",
          log.rank.weights = "S2")+
  ggtitle("GMM proportion")

arrange_ggsurvplots(list(p1,p2))
p1
```

### For OS on metacluster density 

```{r OS metacluster density}
library(survival)
library(survminer)

PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_area <- colData(sce) %>% as.data.frame() %>% select(tissue_area,PID) %>% unique() %>% group_by(PID) %>% mutate(mean_tissue_area = mean(tissue_area))

PID_metacluster$mean_tissue_area <- PID_area$mean_tissue_area[match(PID_metacluster$PID, PID_area$PID)]
  
PID_metacluster <- PID_metacluster %>% group_by(PID) %>% mutate(density = Freq/(mean_tissue_area/1000000))  

### Multivariate cox proportional hazard model on multiple covariates (here density of phenotypes)
PID_metagroups <- PID_metacluster %>% pivot_wider(id_cols = "PID",names_from = "metacluster",values_from = "density") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- paste0("MC",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_OS <- sce$status_OS[match(rownames(Cox_df),sce$PID)]
Cox_df$OS_months <- sce$OS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#mean division 
Cox_df_mean <- foreach(i = Cox_df %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames,.combine = "cbind")%do%{
Cox_df %>% select(i) %>% mutate(mean = ifelse(Cox_df[,i] > mean(Cox_df[,i]),"B","A")) %>% pull(mean)
}

colnames(Cox_df_mean) <- Cox_df %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames
rownames(Cox_df_mean) <- Cox_df %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% rownames
Cox_df_mean <- as.data.frame(Cox_df_mean)

identical(rownames(Cox_df), rownames(Cox_df_mean))

Cox_df_mean <- cbind(Cox_df_mean, Cox_df %>% select(status_OS,OS_months, age, grade, pT_simple, pN_simple, pM_simple))

#run the model
covariates <- Cox_df_mean %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_mean)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

multi_res$coefficients
#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for mean_density")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

### For DFS on metacluster density 

```{r DFS metacluster density}
library(survival)
library(survminer)

PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_area <- colData(sce) %>% as.data.frame() %>% select(tissue_area,PID) %>% unique() %>% group_by(PID) %>% mutate(mean_tissue_area = mean(tissue_area))

PID_metacluster$mean_tissue_area <- PID_area$mean_tissue_area[match(PID_metacluster$PID, PID_area$PID)]
  
PID_metacluster <- PID_metacluster %>% group_by(PID) %>% mutate(density = Freq/(mean_tissue_area/1000000))  

### Multivariate cox proportional hazard model on multiple covariates (here density of phenotypes)
PID_metagroups <- PID_metacluster %>% pivot_wider(id_cols = "PID",names_from = "metacluster",values_from = "density") %>% column_to_rownames("PID")
colnames(PID_metagroups) <- paste0("MC",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#mean division 
Cox_df_mean <- foreach(i = Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames,.combine = "cbind")%do%{
Cox_df %>% select(i) %>% mutate(mean = ifelse(Cox_df[,i] > mean(Cox_df[,i]),"B","A")) %>% pull(mean)
}

colnames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames
rownames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% rownames
Cox_df_mean <- as.data.frame(Cox_df_mean)

identical(rownames(Cox_df), rownames(Cox_df_mean))

Cox_df_mean <- cbind(Cox_df_mean, Cox_df %>% select(status_DFS,DFS_months, age, grade, pT_simple, pN_simple, pM_simple))

#run the model
covariates <- Cox_df_mean %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_mean)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

multi_res$coefficients

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Immune metaclusters")+
ylab("HR for mean_density")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

### Plot proportion density curves for all metaclusters

Add lines for mean and GMM division.

```{r}
#mean division
mean_metacluster <- PID_metacluster %>% group_by(metacluster) %>% transmute(metacluster,mean = mean(fra)) %>% unique()

#GMM division
GMM_div <- Cox_df_GMM %>% select(colnames(Cox_df_GMM)[str_detect(colnames(Cox_df_GMM),"MC")]) %>% rownames_to_column("PID") %>% pivot_longer(cols = 2:9,names_to = "metacluster", values_to = "GMM")
PID_metacluster$GMM <- GMM_div$GMM
GMM_1_max_metacluster <- PID_metacluster %>% filter(GMM == "1") %>% group_by(metacluster) %>% summarise(max = max(fra))

# Plot proportion density curves for all metaclusters
ggplot(PID_metacluster, aes(fra))+
  facet_wrap(~metacluster)+
  geom_density(aes(fill = metacluster))+
  geom_rug()+
  coord_cartesian(ylim = c(0,10))+
  scale_fill_manual(values = metadata(sce)$colors$metacluster)+
  theme_classic()+
  geom_vline(data = mean_metacluster, aes(xintercept = mean),color = "#E41A1C", linetype = "dashed") #red
  #geom_vline(data = GMM_1_max_metacluster, aes(xintercept = max),color = "#377EB8", linetype = "dashed") #blue
```

## Lasso-regularized Cox-proportonal hazards model for clusters

### OS

```{r lasso OS clusters}
library(glmnet)
library(broom) 

PID_cluster <- colData(sce) %>% as.data.frame() %>% select(cluster_110,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_cluster <- PID_cluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_groups <- PID_cluster %>% pivot_wider(id_cols = "PID",names_from = "cluster_110",values_from = "fra") %>% column_to_rownames("PID")

### Lasso regularization
### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)

## 1. Based on continuous cluster proportions
Cox_df <- PID_groups

#add covariates for correction
Cox_df$status_OS <- sce$status_OS[match(rownames(Cox_df),sce$PID)]
Cox_df$OS_months <- sce$OS_months[match(rownames(Cox_df),sce$PID)]

#Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
#Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
#Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
#Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
#Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)]) #remove from model otherwise coefficients are shrinked to 0 for all others? 

## Run lasso regularized coxph model
Cox_df <- Cox_df[complete.cases(Cox_df),] #only have complete rows as input

x <- data.matrix(Cox_df %>% select(-status_OS,-OS_months)) #predictor variables

y <- Surv(time = Cox_df$OS_months, event = Cox_df$status_OS) #response variables

## Run lasso 
set.seed(22)
cv.fit <- cv.glmnet(x, y, family="cox", type.measure = "C")
plot(cv.fit)

fit <- glmnet(x, y, family="cox")

coef(fit, s = cv.fit$lambda.min)

## Run lasso 100 times and average error curves to be more robust
MSEs <- NULL
for (i in 1:10){
                 cv <- cv.glmnet(x, y, family="cox", type.measure = "C")
                 MSEs <- cbind(MSEs, cv$cvm)
             }

rownames(MSEs) <- cv$lambda
lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

fit <- glmnet(x, y, family="cox")
coef(fit, s = lambda.min)

#Extract output
lasso_df <- tidy(coef(fit, s = cv.fit$lambda.min))
colnames(lasso_df) <- c("predictor","column","coef")

#Visualize hazard ratios
ggplot(lasso_df, aes(x = predictor, y = exp(coef))) +
  geom_point(size = 4, aes(color = predictor)) +
  geom_hline(yintercept = 1,color = "black")+
  scale_y_continuous(trans="log")+
  ylab("Lasso HR")+
  theme_classic()+
  ggtitle("OS")+
  scale_color_manual(values = metadata(sce)$colors$cluster_110, limits = force, na.value = "black")

```

### DFS

```{r lasso DFS clusters}
library(glmnet)
library(broom) 

PID_cluster <- colData(sce) %>% as.data.frame() %>% select(cluster_110,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_cluster <- PID_cluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_groups <- PID_cluster %>% pivot_wider(id_cols = "PID",names_from = "cluster_110",values_from = "fra") %>% column_to_rownames("PID")

### Lasso regularization
### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)

## 1. Based on continuous cluster proportions
Cox_df <- PID_groups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)]) #remove from model otherwise coefficients are shrinked to 0 for all others? 

## Run lasso regularized coxph model
Cox_df <- Cox_df[complete.cases(Cox_df),] #only have complete rows as input

Cox_df <- Cox_df %>% filter(DFS_months != 0)

x <-  data.matrix(Cox_df %>% select(-status_DFS,-DFS_months)) #predictor variables

y <- Surv(time = Cox_df$DFS_months, event = Cox_df$status_DFS) #response variables

## Run lasso 
require(doMC)
registerDoMC(cores = 20)
set.seed(22)
cv.fit <- cv.glmnet(x, y, family="cox", type.measure = "C", nfolds = 50, parallel = TRUE)
plot(cv.fit)

fit <- glmnet(x, y, family="cox")

coef(fit, s = cv.fit$lambda.min)

# ## Run lasso 10 times (with 10 fold validation each run) and average error curves to be more robust
# MSEs <- NULL
# for (i in 1:10){
#                   cv <- cv.glmnet(x, y, family="cox", type.measure = "C")
#                   MSEs <- cbind(MSEs, cv$cvm)
#                   }
#  
# rownames(MSEs) <- cv$lambda
# lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))
#  
# fit <- glmnet(x, y, family="cox")
# coef(fit, s = lambda.min)

#Extract output
lasso_df <- tidy(coef(fit, s = cv.fit$lambda.min))
 colnames(lasso_df) <- c("predictor","column","coef")

#Visualize hazard ratios
ggplot(lasso_df, aes(x = reorder(predictor, coef), y = exp(coef))) +
  geom_point(size = 4, aes(color = predictor)) +
  geom_hline(yintercept = 1,color = "black")+
  scale_y_continuous(trans="log")+
  ylab("Lasso HR")+
  ggtitle("DFS")+
  theme_classic()+
  scale_color_manual(values = metadata(sce)$colors$cluster_110, limits = force, na.value = "black")

lasso_plot <- ggplot(lasso_df, aes(x = reorder(predictor,-exp(coef)), y = exp(coef))) +
  geom_point(size = 4, aes(color = exp(coef)>1)) +
  geom_hline(yintercept = 1,color = "black")+
  scale_y_continuous(trans="log")+
  ylab("Lasso HR")+
  xlab("Predictors")+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90))+
  ggtitle("DFS")+
  scale_color_brewer(palette = "Set1")
  scale_color_manual(values = metadata(sce)$colors$cluster_110, limits = force, na.value = "black")
lasso_plot
```

# 8. Differential abundance testing 

## pN 

```{r}
#Assess survival impact 
library(patchwork)
library(survival)
library(survminer)


#also possible to use spatial_CD8_pheno_merg (cold/excluded vs inflamed)
cur_surv <- colData(sce) %>% as.data.frame() %>% 
  select(PID, pN_merg,OS_months, DFS_months, status_DFS, status_OS) %>% unique()

#Plot survival curves
fit <- survfit(Surv(OS_months, status_OS) ~ pN_merg, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          #palette = c("steelblue1","navy","firebrick"),
          log.rank.weights = "n")

fit <- survfit(Surv(DFS_months, status_DFS) ~ pN_merg, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("grey","purple"),
          log.rank.weights = "n")
```


### For metaclusters

```{r}
### Differential abundance analysis based on edgeR
library(edgeR)

sce$pN_merg <- colData(sce) %>% 
  as.data.frame %>% 
  select(pN_simple) %>% 
  mutate(pN_merg = ifelse(pN_simple %in% c("1","2","3"),"1/2/3", pN_simple)) %>% pull(pN_merg)

sce <- sce
abundances <- table(unfactor(sce$metacluster), sce$PID)
abundances <- unclass(abundances) 
head(abundances)

#relevant patient metadata (Relevel so that pN 1/2 is baseline)
meta <- colData(sce)[match(colnames(abundances), sce$PID),]
meta$pN_merg <- relevel(x=factor(meta$pN_merg), ref = "0")

#filter
meta <- meta[!is.na(meta$pN_merg),]
abundances <- abundances[,colnames(abundances) %in% meta$PID]

#create DGEList - based on edgeR package
y.ab <- DGEList(abundances, samples=meta, group = meta$pN_merg)

#Design matrix
design <- model.matrix(~pN_merg, data = y.ab$samples)

#Visualize
# logCPM of abundances per cluster
cur_mat <- cpm(abundances,prior.count = 1,log = TRUE)

Heatmap(cur_mat, 
        col = viridis(100),
        column_split = unfactor(meta$pN_merg))%v%
  columnAnnotation(pN_merg = anno_simple(unfactor(meta$pN_merg)))


# Estimate NB and QL dispersion 
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
plotBCV(y.ab, cex=1)

fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)
plotQLDisp(fit.ab, cex=1)

# Test for differential abundance of clusters between tumor stages
res <- glmQLFTest(fit.ab, coef=ncol(design))

res$coefficients
summary(decideTests(res))

DA <- res$table

DA <- topTags(res, n = nrow(DA), adjust.method = "BH")$table

DA <- DA %>% mutate(cluster = rownames(DA), sign = DA$FDR<0.1, result = ifelse(DA$logFC > 0 & DA$FDR < 0.1, "UP",ifelse(DA$logFC < 0 & DA$FDR < 0.1, "DOWN", "NOTSIG"))) %>% 
  arrange(desc(logFC))

DA$cluster <- factor(DA$cluster, levels = DA$cluster)

sig <- DA %>% filter(FDR<0.1) %>% pull(cluster) %>% unfactor

## Visualize
#Barplot
ggplot(DA, aes(cluster, y = logFC))+
  geom_col(aes(fill=result), col = "black")+
  scale_fill_manual(values = c("#377EB8","lightgrey","#E41A1C"))+
  ylab("logFC")+
  xlab("Cluster")+
  labs(fill = "FDR<0.1")+
  ggtitle("pN_stage (0 vs 1/2/3)")+
  theme_classic()

#Volcano plot  
ggplot(DA, aes(x = logFC, y = -log10(FDR)))+
  geom_point(aes(col=cluster, size = -log10(FDR)))+
    geom_label_repel(data = DA %>% filter(FDR<0.1), aes(label = cluster))+
  scale_color_manual(values = metadata(sce)$colors$metacluster)+
  xlab("logFC (pN: 0 vs 1/2/3)")+
  ylab("-log10(FDR)")+
  theme_classic()+
  geom_hline(yintercept = -log10(0.1), linetype = "dashed")+
  geom_vline(xintercept = 0, linetype = "dashed")+
  labs(color = "cluster")

#Abundance vs log fold change (best?)
ggplot(DA, aes(x = logFC, y = logCPM))+
  geom_point(aes(col= cluster, 
                 size = -log10(FDR)))+
  geom_label_repel(data = DA %>% filter(FDR<0.1), aes(label = cluster))+
  scale_color_manual(values = metadata(sce)$colors$metacluster)+
  xlab("logFC (pN: 0 vs 1/2/3)")+
  ylab("Abundance (logCPM)")+
  theme_classic()+
  geom_vline(xintercept = 0, linetype = "dashed")+
  labs(color = "cluster")
```

### For clusters

```{r}
### Differential abundance analysis based on edgeR
library(edgeR)

sce$pN_merg <- colData(sce) %>% 
  as.data.frame %>% 
  select(pN_simple) %>% 
  mutate(pN_merg = ifelse(pN_simple %in% c("1","2","3"),"1/2/3", pN_simple)) %>% pull(pN_merg)

sce <- sce
abundances <- table(unfactor(sce$cluster_110), sce$PID)
abundances <- unclass(abundances) 
head(abundances)

#relevant patient metadata (Relevel so that pN 1/2 is baseline)
meta <- colData(sce)[match(colnames(abundances), sce$PID),]
meta$pN_merg <- relevel(x=factor(meta$pN_merg), ref = "0")

#filter
meta <- meta[!is.na(meta$pN_merg),]
abundances <- abundances[,colnames(abundances) %in% meta$PID]

#create DGEList - based on edgeR package
y.ab <- DGEList(abundances, samples=meta, group = meta$pN_merg)

#Design matrix
design <- model.matrix(~pN_merg, data = y.ab$samples)

#Visualize
# logCPM of abundances per cluster
cur_mat <- cpm(abundances,prior.count = 1,log = TRUE)

Heatmap(cur_mat, 
        col = viridis(100),
        column_split = unfactor(meta$pN_merg))%v%
  columnAnnotation(pN_merg = anno_simple(unfactor(meta$pN_merg)))


# Estimate NB and QL dispersion 
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
plotBCV(y.ab, cex=1)

fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)
plotQLDisp(fit.ab, cex=1)

# Test for differential abundance of clusters between tumor stages
res <- glmQLFTest(fit.ab, coef=ncol(design))

res$coefficients
summary(decideTests(res))

DA <- res$table

DA <- topTags(res, n = nrow(DA), adjust.method = "BH")$table

DA <- DA %>% mutate(cluster = rownames(DA), sign = DA$FDR<0.05, result = ifelse(DA$logFC > 0 & DA$FDR < 0.05, "UP",ifelse(DA$logFC < 0 & DA$FDR < 0.05, "DOWN", "NOTSIG"))) %>% 
  arrange(desc(logFC))

DA$cluster <- factor(DA$cluster, levels = DA$cluster)

sig <- DA %>% filter(FDR<0.05) %>% pull(cluster) %>% unfactor

## Visualize
#Barplot
ggplot(DA, aes(cluster, y = logFC))+
  geom_col(aes(fill=result), col = "black")+
  scale_fill_manual(values = c("#377EB8","lightgrey","#E41A1C"))+
  ylab("logFC")+
  xlab("Cluster")+
  labs(fill = "FDR<0.1")+
  ggtitle("pN_stage (0 vs 1/2/3)")+
  theme_classic()

#Volcano plot  
ggplot(DA, aes(x = logFC, y = -log10(FDR)))+
  geom_point(aes(col=cluster, size = -log10(FDR)))+
  geom_label_repel(data = DA %>% filter(FDR<0.05), aes(label = cluster))+
  scale_color_manual(values = metadata(sce)$colors$cluster_110)+
  xlab("logFC (pN: 0 vs 1/2/3)")+
  ylab("-log10(FDR)")+
  theme_classic()+
  geom_hline(yintercept = -log10(0.05), linetype = "dashed")+
  geom_vline(xintercept = 0, linetype = "dashed")+
  labs(color = "cluster")

#Abundance vs log fold change (best?)
ggplot(DA, aes(x = logFC, y = logCPM))+
  geom_point(aes(col= cluster, 
                 size = -log10(FDR)))+
  geom_label_repel(data = DA %>% filter(FDR<0.05), aes(label = cluster))+
  scale_color_manual(values = metadata(sce)$colors$cluster_110)+
  xlab("logFC (pN: 0 vs 1/2/3)")+
  ylab("Abundance (logCPM)")+
  theme_classic()+
  geom_vline(xintercept = 0, linetype = "dashed")+
  labs(color = "cluster")
```

# 9. Image visualization

```{r load images, message = FALSE}
#Load images and masks
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_segtest.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")
```

## Visualize metaclusters on images 

```{r viz clus, message=FALSE, warning=FALSE}
clus_select <- unique(sce$metacluster)

library(doParallel)
registerDoParallel(cores = 7)

foreach(i = clus_select)%dopar%{

##Visualize clusters on images - select 10 images for each category with highest number of cells
cur_sample <- colData(sce) %>% as.data.frame %>% group_by(metacluster) %>% filter(metacluster == i) %>% count(sample_id) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(sample_id)

#Prepare images
cur_images <- images[cur_sample]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#Subset sce
cur_sce <- sce[,sce$metacluster == i]

#Plot 
plotPixels(cur_images,
           mask = masks[cur_sample],
           object = cur_sce,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           missing_colour = "white",
           outline_by = "metacluster",
           colour_by = c("CK5","Vimentin","CK7","CD15","EGFR","KRT14"),
           thick = FALSE,
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = paste0("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/Tumor_clusters_viz/Tumor_viz_metacluster_",i,".png"), scale = 2)
           )
 
  }

```


## Visualize metaclusters on images

```{r viz metaclus, message=FALSE, warning=FALSE}

# 1. Part 1 of immune_metacluster viz 
clus_select <- anno %>% filter(!metacluster %in% c("Endothelial","Stromal","Neutrophil")) %>% pull(metacluster) %>% unfactor()

for (i in clus_select) {
  
##Visualize clusters on images - select 10 images for each category with highest number of cells
cur_sample <- colData(sce) %>% as.data.frame %>% group_by(immune_metacluster) %>% filter(immune_metacluster == i) %>% count(sample_id) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(sample_id)

#Prepare images
cur_images <- images[cur_sample]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#Subset sce
cur_sce <- sce[,sce$immune_metacluster == i]

#Plot 
plotPixels(cur_images,
           mask = masks[cur_sample],
           object = cur_sce,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           missing_colour = "white",
           outline_by = "cluster_130",
           colour_by = c("CD4", "HLADR", "CD68","CD8a", "CD20","CD3"),
           #colour = list("SMA" = c("black", "green"),
           #              "CD3" = c("black", "yellow"),
           #              "CD31_vWF" = c("black", "magenta"),
           #              "CD20" = c("black", "cyan"),
           #              "CD68" = c("black", "red"),
           #              "cluster_130" = metadata(test_sce)$colors$cluster_130
           #              ),
           thick = FALSE,
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = paste0("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/TME_clusters_viz/TME_viz_immune_metacluster_",i,".png"), scale = 2)
           )
 
  }


# 2. Part 2 of immune_metacluster viz 

clus_select <- anno %>% filter(metacluster %in% c("Endothelial","Stromal","Neutrophil")) %>% pull(metacluster) %>% unfactor()

for (i in clus_select) {
  
##Visualize clusters on images - select 10 images for each category with highest number of cells
cur_sample <- colData(sce) %>% as.data.frame %>% group_by(immune_metacluster) %>% filter(immune_metacluster == i) %>% count(sample_id) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(sample_id)

#Prepare images
cur_images <- images[cur_sample]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#Subset sce
cur_sce <- sce[,sce$immune_metacluster == i]

#Plot 
plotPixels(cur_images,
           mask = masks[cur_sample],
           object = cur_sce,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           missing_colour = "white",
           outline_by = "cluster_130",
           colour_by = c("CD31_vWF", "SMA", "CD68","CD15", "CD20","CD3"),
           #colour = list("SMA" = c("black", "green"),
           #              "CD3" = c("black", "yellow"),
           #              "CD31_vWF" = c("black", "magenta"),
           #              "CD20" = c("black", "cyan"),
           #              "CD68" = c("black", "red"),
           #              "cluster_130" = metadata(test_sce)$colors$cluster_130
           #              ),
           thick = FALSE,
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = paste0("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/TME_clusters_viz/TME_viz_immune_metacluster_",i,".png"), scale = 2)
           )
 
  }
```


## Save sce objects

```{r save_objects, message = FALSE, warning=FALSE}
#Save sce
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor_new.rds")
```

## Save plots as pdf

```{r save_plots, include=FALSE}
##IMPORTANT: change to chunk output in console for this last part - DO NOT RUN LIKE THIS DUE TO DUPLICATED VARIABLE NAMES

##save plots as pdf (vector graphics)

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Heatmap_ClusterTesting.pdf",height = 4, width = 9)
p1+p2+p3 
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Heatmap_Cluster_Metacluster.pdf",width = 13,height = 13)
draw(h_meta)
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Heatmap_Cluster_Metacluster_QC.pdf",width = 14,height = 9)
draw(h_meta)
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Metacluster_Fra_Cor.pdf", height = 8)
SID_metacluster_cor_plot
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_CLRDR_Metacluster.pdf")
clrDR(sce_cat, dr = "PCA", 
      by = "cluster_id", k = "metacluster", 
      label_by = "cluster_id", arrow_col = "PID", 
      point_pal = metadata(sce_cat)$colors$metacluster, 
      arrows = FALSE) +
  scale_color_manual(values = metadata(sce_cat)$colors$PID)+
  guides(color = "none")+
  theme_classic()
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_FractionHeatmap.pdf", width = 10, height = 16, onefile = FALSE)
fraction_anno + NULL
dev.off()


pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_EucDistanceCLR_Fra.pdf", height = 6,width = 9)
ggplot(out_plot_1, aes(x = dist))+
  geom_density(aes(fill = category), color = "black",alpha = 0.5)+
  #geom_rug(aes(color = PID))+
  geom_rug(aes(color = category), alpha = 0.5)+
  theme_classic()+
  ggtitle("Mean pairwise euclidian distance of metacluster fractions", 
          subtitle = paste0("p = ",compare_means(dist ~ category, out_plot,paired = TRUE)$p.format, " (Wilcoxon)"))+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")+
  guides(color = "none")+
  coord_cartesian(clip = "off")
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_DominantBar.pdf", width = 5)
ggplot(out_bar, aes(x = category, y = n, fill = dominant))+
  geom_tile(position = "stack", alpha = 0.5, color = "black", size = 0.7)+
  theme_classic()+
  ylab("Number of PID")+
  scale_fill_brewer(palette = "Set1")
dev.off()


pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Dominant_Scatter.pdf", height = 6, width = 9)
ggplot(test_plot, aes(mean_fra, dif_fra, color = PID))+
  geom_point(size = 3)+
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  guides(color = "none")+
  theme_classic()
dev.off()


pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Cox_HR_DFS.pdf", height = 4, width = 8)
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0,185))+
#scale_y_continuous(trans = "log10")
theme(axis.text.x = element_text(angle = 90))
dev.off()


pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Kaplan_DFS_bestcutoff.pdf", width = 6, height = 6, onefile = FALSE)
ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = "Set1") 
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_DiffAbun_Node.pdf", width = 5, height = 4, onefile = FALSE)
ggplot(DA, aes(x = logFC, y = logCPM))+
  geom_point(aes(col= cluster, 
                 size = -log10(FDR)))+
  geom_label_repel(data = DA %>% filter(FDR<0.1), aes(label = cluster))+
  scale_color_manual(values = metadata(sce)$colors$metacluster)+
  xlab("logFC (pN: 0 vs 1/2/3)")+
  ylab("Abundance (logCPM)")+
  theme_classic()+
  geom_vline(xintercept = 0, linetype = "dashed")+
  labs(color = "cluster")
dev.off()


pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_CK57score_Heatmap.pdf", width = 11, onefile = FALSE)
h
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_CK57score_Cox.pdf",width = 8,height = 4)
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0,40))+
#scale_y_continuous(trans = "log10")
theme(axis.text.x = element_text(angle = 90))
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Cluster_Fra_Cor.pdf", height = 8)
SID_cluster_cor_plot
dev.off()


pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_CLRDR_PID.pdf")
clrDR(sce_cat, dr = "PCA", 
      by = "sample_id", point_col = "sample_id",
      k = "metacluster", point_pal = metadata(sce_cat)$colors$PID)+
  scale_color_manual(values = metadata(sce_cat)$colors$metacluster)+
  guides(fill = "none")+
  theme_classic()
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_n_metacluster_PID.pdf", height = 8)
n_MC_PID
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_n_cluster_PID.pdf", height = 8)
n_clus_PID
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_DominantBar_Clus.pdf", width = 5)
ggplot(out_bar, aes(x = category, y = n, fill = dominant))+
  geom_tile(position = "stack", color = "black",alpha = 0.5)+
  theme_classic()+
  ylab("Number of PID")+
  scale_fill_brewer(palette = "Set1")
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Basoluminal_Scatter.pdf", height = 4,width = 8)
scatter_CK5_7 + scatter_CK5_8_18
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Cox_HR_OS_Meta.pdf", height = 4, width = 8)
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0,60))+
#scale_y_continuous(trans = "log10")
theme(axis.text.x = element_text(angle = 90))
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Cox_HR_DFS_Meta_Uni.pdf", height = 4, width = 6)
hr_prop
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Cox_HR_DFS_CK57PID_Uni.pdf", height = 4, width = 4)
hr_prop
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Cox_HR_DFS_Clus_Lasso.pdf", height = 5, width = 8)
lasso_plot
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_KaplanMeier_pNmerg.pdf", width = 6, height = 6, onefile = FALSE)
ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("grey","purple"),
          log.rank.weights = "n")
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_DiffAbun_Node_Clus.pdf", width = 6, height = 5, onefile = FALSE)
ggplot(DA, aes(x = logFC, y = logCPM))+
  geom_point(aes(col= cluster, 
                 size = -log10(FDR)))+
  geom_label_repel(data = DA %>% filter(FDR<0.05), aes(label = cluster))+
  scale_color_manual(values = metadata(sce)$colors$cluster_110)+
  xlab("logFC (pN: 0 vs 1/2/3)")+
  ylab("Abundance (logCPM)")+
  theme_classic()+
  geom_vline(xintercept = 0, linetype = "dashed")+
  labs(color = "cluster")
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_CK57score_scater.pdf", width = 5, height = 5)
ggplot(cur_dat, aes(x=CK7, y=CK5, color = CK5_7_score)) +
  geom_point(alpha = 0.5)+
  theme_classic()+
  xlab("CK7 exprs")+
  ylab("CK5 exprs")+
  scale_color_manual(values = metadata(sce)$colors$CK5_7_score)
dev.off()


pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Age_MC8Abundance_scater.pdf", width = 5, height = 5)
ggplot(PID_MC8, aes(x = fra, y=age))+
  geom_point()+
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth", 
              color = "firebrick")+
  stat_poly_eq()+
  stat_fit_glance(method = 'lm',
                       method.args = list(formula = y ~ x),
                       geom = 'text',
                       aes(label = paste("P-value = ", signif(..p.value.., digits = 4), sep = "")))+
  theme_classic()
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_Age_CK57pos_scater.pdf", width = 5, height = 5)
ggplot(PID_CK57_pos, aes(x = fra, y=age))+
  geom_point()+
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth", 
              color = "firebrick")+
  stat_poly_eq()+
  stat_fit_glance(method = 'lm',
                       method.args = list(formula = y ~ x),
                       geom = 'text',
                       aes(label = paste("P-value = ", signif(..p.value.., digits = 4), sep = "")))+
  theme_classic()
dev.off()


pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_CK57score_scater_density.pdf", width = 6, height = 5)
plot(scatter_final)
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/05_tumor/TNBC_Tumor_CK57group_domScore_bar.pdf", width = 5, height = 5)
dittoBarPlot(sce_PID, var = "dom_CK57_score", group.by = "CK5_7_group")+
  scale_fill_manual(values = metadata(sce)$colors$CK5_7_score)+
  ylab("Percent of patients")+
  ggtitle(NULL)
dev.off()




##IMPORTANT: change back to chunk output inline
```


