---
title: "09_combined_spatial"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here, we will attend to perform analysis akin to what has been proposed as **Tissue schematics (TS)** by Bhate et al., Cell Systems, 2022.

### 1. Cellular neighborhoods 

First, we will find cellular neighborhoods based on all cell type clusters (metaclusters) - "Tumor ecosystem communities"

To this end, we will define cellular neighborhoods by clustering cells based on cell type composition of their k-nearest neighbors. (k=20, euclidian distance between X/Y)


```{r neighborhood-1, message=FALSE, warning=FALSE}
library(imcRtools)
library(bluster)

##Load sce and colors
sce <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/sce_TNBC_249_08.rds")
col_clusters <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/color_phenocombined.rds")
sce

# Build 10-nearest neighbor graph (max distance between neighbors 30 um)
sce <- buildSpatialGraph(sce, img_id = "sample_id", type = "knn", k = 10, 
                             coords = c("Pos_X",  "Pos_Y"), max_dist = 30,
                         BPPARAM = MulticoreParam(progressbar = TRUE))

colPairNames(sce) #now includes knn_interaction_graph

# Aggregate across neighbors
sce <- aggregateNeighbors(sce, colPairName = "knn_interaction_graph", 
                              aggregate_by = "metadata", count_by = "metacluster_TME")

head(sce$aggregatedNeighbors)

# Perform cluster sweep for kmeans (silhouette width, wcss)
set.seed(22)
clust_list <- lapply(2:20, function(k){
  cur_km <- kmeans(sce$aggregatedNeighbors, centers = k)
  return(list(wss = sum(cur_km$withinss),
           clusters = cur_km$cluster))
})

sil <- vapply(lapply(clust_list, `[[`, 2), 
              function(x) mean(approxSilhouette(sce$aggregatedNeighbors, x)$width), 
              0)

ggplot(data.frame(k = 2:20, sil = sil))+
  geom_point(aes(k, sil, col = sil))+
  geom_line(aes(k, sil,col=sil))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  xlab("k")+
  ylab("Mean silhouette width")+
  scale_color_continuous(type = "viridis")

ggplot(data.frame(k = 2:20,wss = unlist(lapply(clust_list, `[[`, 1))))+
  geom_point(aes(k, wss, col = wss))+
  geom_line(aes(k, wss, col = wss))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("k")+
  ylab("Total WCSS")+
  scale_color_continuous(type = "viridis")

# Select 11 clusters
cur_clusters <- kmeans(sce$aggregatedNeighbors, centers = 11, nstart = 100)$cluster

sce$cellular_neighborhood <- as.factor(cur_clusters)

##Save sce 
#saveRDS(sce, "/mnt/rcc_volume/TNBC/data_analysis/sce_TNBC_249_08.rds")
```

We can now visualize a few example images and highlight the type of cells within each neighborhoood.

```{r neighborhood-2, fig.width=15, fig.height=12}
set.seed(22)
cur_sam <- sample(unique(sce$sample_id), 6)

plotSpatial(sce[,sce$sample_id %in% cur_sam], 
            img_id = "sample_id",
            coords = c("Pos_X", "Pos_Y"),
            node_color_by = "cellular_neighborhood",
            draw_edges = TRUE,
            colPairName = "knn_interaction_graph", 
            nodes_first = FALSE,
            directed = FALSE)+
  scale_color_brewer(palette = "Set3")+
  theme_classic()

colPairNames(sce)

plotSpatial(sce[,sce$sample_id %in% cur_sam], 
            img_id = "sample_id",
            coords = c("Pos_X", "Pos_Y"),
            node_color_by = "cluster_combined",
            node_shape_by = "cluster_origin",
            draw_edges = TRUE,
            directed = FALSE,
            colPairName = "knn_interaction_graph", 
            nodes_first = FALSE) +
  scale_color_manual(values = col_clusters$cluster_combined)+
  theme_classic()
```

```{r}
# Visualize cell type proportions of cellular neighborhoods
cn_prop <- table(cellular_neighborhood = sce$cellular_neighborhood,metacluster_TME = sce$metacluster_TME) %>% as.data.frame %>% group_by(cellular_neighborhood) %>% mutate(fra = Freq/sum(Freq))

## 1. Heatmap body
cn_prop_wide <- cn_prop %>% pivot_wider(id_cols = cellular_neighborhood, names_from = metacluster_TME, values_from = fra) %>% column_to_rownames(var = "cellular_neighborhood")

#Option: min_max normalization per neighborhood
cn_prop_wide <- as.matrix(cn_prop_wide)
cn_prop_wide_min_max <- (cn_prop_wide - rowMins(cn_prop_wide))/(rowMaxs(cn_prop_wide)-rowMins(cn_prop_wide))

#Heatmap body color
col_main = viridis(100)

## 2. Row annotation - number of cells/PID
cluster_PID <- colData(sce) %>% as.data.frame() %>% select(cellular_neighborhood,PID) %>% group_by(cellular_neighborhood) %>% table() %>% as.data.frame
n_PID <- cluster_PID %>% filter(Freq>0) %>% group_by(cellular_neighborhood) %>% count(name = "n_PID") %>% column_to_rownames("cellular_neighborhood")
ncells <- colData(sce) %>% as.data.frame() %>% select(cellular_neighborhood) %>% group_by(cellular_neighborhood) %>% count() 

row_anno <- cbind(n_PID,ncells)

#check and correct order
row_anno <- row_anno[match(rownames(cn_prop_wide),row_anno$cellular_neighborhood),]

## 3. Column annotation 
col_anno <- colData(sce) %>% as.data.frame() %>% select(metacluster_TME,metacluster) %>% unique()
col_anno$metacluster_TME <- as.character(col_anno$metacluster_TME)
col_anno$metacluster <- as.character(col_anno$metacluster)

#check and correct order
col_anno <- col_anno[match(colnames(cn_prop_wide),col_anno$metacluster_TME),]


col_ha <- HeatmapAnnotation(metacluster_TME = col_anno$metacluster_TME,
                            metacluster = col_anno$metacluster,
                  #col = col_clusters, 
                  border = TRUE, 
                  na_col = "black",
                  annotation_name_side = "left")

## 4. Plot heatmap
h <- Heatmap(cn_prop_wide_min_max,
             name = "min_max cluster proportion",
        col = col_main, 
        bottom_annotation = col_ha)+
  rowAnnotation(neighborhood = anno_simple(unfactor(row_anno$cellular_neighborhood),pch = unfactor(row_anno$cellular_neighborhood), border = TRUE))+
  rowAnnotation(n_PID = anno_barplot(n_PID, width = unit(10, "mm"),gp = gpar(fill = "#440154FF"), col = 1),annotation_name_rot = 90)+
  rowAnnotation(n_cells = anno_barplot(ncells$n, width = unit(10, "mm"),gp = gpar(fill = "#440154FF"), col = 1),annotation_name_rot = 90)

draw(h)
```


### 2. Spatial contexts

First, we will find spatial contexts based on cellular neighborhoods - "Microenvironments of CN"

To this end, we will define spatial contexts for each cell by the frequency of cellular neighborhood assigments in the cell neighborhood (k=50). We will then use a minimal combination of CN assignments for each window (>90% of cells are assigned to a given CN combination). 


```{r neighborhood-1, message=FALSE, warning=FALSE}
library(imcRtools)
library(bluster)
library(doMC)

#Subset sce object to 1 image for trial code
sce <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/sce_TNBC_249_08.rds")
cur_sam = c("ZTMA249.2_ZTMA249_ID_317_CX5Y10_159")
sce <- sce[,sce$sample_id %in% cur_sam]
sce

## 1. Build 50-nearest neighbor graph (max dist?)
sce <- buildSpatialGraph(sce, img_id = "sample_id", type = "knn", k = 50, 
                             coords = c("Pos_X",  "Pos_Y"), 
                         name = "knn_spatialcontext_graph",
                         BPPARAM = MulticoreParam(progressbar = TRUE))

colPairNames(sce) #now includes knn_spatialcontext_graph

# Aggregate across neighbors
sce <- aggregateNeighbors(sce, colPairName = "knn_spatialcontext_graph", 
                              aggregate_by = "metadata", count_by = "cellular_neighborhood")

## 2. Find spatial contexts using minimal combination approach
# Prepare dataframe from aggregatedNeighbors
sce_sub <- sce$aggregatedNeighbors
sce_sub_t <- sce_sub %>% as.data.frame %>% t() %>% as.data.frame
rownames(sce_sub_t) <- colnames(sce_sub) #rows: cellular neighborhoods, columns: single cells, data = fraction of cellular neighborhoods in cell windows (knn graph)

# Define combination approach threshold
threshold = 0.9

registerDoMC(cores=detectCores()-2) #perform analysis on multiple cores with %dopar% from foreach

SC <- foreach(i = colnames(sce_sub_t), .combine = "rbind")%dopar%{ 
#Create cumulative sum
test <- sce_sub_t %>% select(1) %>% arrange
test <- test %>% arrange(desc(test[,1])) %>% cumsum

#Filter rows that reach cumulative sum threshold (user-defined) - then sort numbers
test <- test %>% slice(1:(length(which(test <=threshold))+1)) %>% rownames() %>% as.numeric() %>% sort()

#Create spatial community identifier
paste(test, collapse = "_")
}

SC <- as.data.frame(SC)

#add to sce 
sce$spatial_context <- SC$V1 

#TO DO:
#find filter criteria for abundant spatial contexts (TODO)
#also filter TMA cores that have too few cells for spatial context assignment
#max_dist for spatial graph? 

##Save sce 
#saveRDS(sce, "/mnt/rcc_volume/TNBC/data_analysis/sce_TNBC_249_08.rds")
```


# Visualize spatial contexts on example images 

```{r neighborhood-2}
# set.seed(22)
# cur_sam <- sample(unique(sce$sample_id), 5)

#by cell types (clusters)
p1 <- plotSpatial(sce, 
            img_id = "sample_id",
            coords = c("Pos_X", "Pos_Y"),
            node_color_by = "cluster_combined",
            draw_edges = FALSE,
            colPairName = "knn_interaction_graph", 
            nodes_first = FALSE,
            node_size_fix = 2,
            node_shape_by = "analysis_cat",
            directed = FALSE)+
  theme_classic()

#by cellular neighborhoods
p2 <- plotSpatial(sce, 
            img_id = "sample_id",
            coords = c("Pos_X", "Pos_Y"),
            node_color_by = "cellular_neighborhood",
            draw_edges = FALSE,
            colPairName = "knn_interaction_graph", 
            nodes_first = FALSE,
            node_size_fix = 2,
            node_shape_by = "analysis_cat",
            directed = FALSE)+
  theme_classic()

#by spatial contexts
p3 <- plotSpatial(sce, 
            img_id = "sample_id",
            coords = c("Pos_X", "Pos_Y"),
            node_color_by = "spatial_context",
            draw_edges = FALSE,
            colPairName = "knn_spatialcontext_graph", 
            nodes_first = FALSE,
            node_size_fix = 2,
            node_shape_by = "analysis_cat",
            directed = FALSE)+
  theme_classic()

p1/p2/p3
```

```{r viz spatial contexts}
## 1. Total number of cells in spatial context
SC_count <- colData(sce) %>% as.data.frame %>% select(spatial_context) %>% group_by(spatial_context) %>% count() %>% arrange(desc(n))

## 2. Fraction of sample_id per spatial
SC_SID <- colData(sce) %>% as.data.frame() %>% select(spatial_context,sample_id) %>% group_by(spatial_context) %>% table() %>% as.data.frame
SC_SID <- SC_SID %>% group_by(spatial_context) %>% mutate(fra = Freq/sum(Freq))

ggplot(SC_SID) +
  geom_tile(color = "black",aes(spatial_context, fra, fill = sample_id), position = "stack")+
  theme_classic(base_size = 15)+
  ylab("SID fraction")+
  theme(axis.title.x = element_blank())+
  guides(fill = "none")

## 3. Fraction of spatial context per PID
SID_SC <- colData(sce) %>% as.data.frame() %>% select(spatial_context,sample_id) %>% group_by(sample_id) %>% table() %>% as.data.frame
SID_SC <- SID_SC %>% group_by(sample_id) %>% mutate(fra = Freq/sum(Freq))

ggplot(SID_SC) +
  geom_tile(color = "black",aes(sample_id, fra, fill = spatial_context), position = "stack")+
  theme_classic(base_size = 15)+
  ylab("SC Fraction")+
  xlab("Sample_ID")+
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())


#TO DO: Find filter criteria for abundant spatial contexts - Make workflow more efficient/faster (talk to Nils)
```

```{r cn comb map}
#Plot CN combination map from each sample_id 

## 1. Create edgelist
list <- str_split(SC_SID$spatial_context, "_")
list_length <- sapply(list, length)

registerDoMC(cores=detectCores()-2) #perform analysis on multiple cores with %dopar% from foreach

edges <- foreach(i = seq_along(list), .combine = "rbind")%dopar%{

list_options <- list[length(list[[i]])+1 == list_length]

if(length(list_options) != 0){
  
list_select <- list_options[sapply(list_options, function(x){length(intersect(x,list[[i]])) == length(list[[i]])})]

if(length(list_select) != 0){ 
data.frame("from" = sapply(list[i], paste, collapse = "_"),
                     "to" = sapply(list_select, paste, collapse = "_")
)
} else {
    NULL
  } 
} else {
  NULL
  }
}

#add layer information 
SC_SID$layer <- as.character(list_length)

## 2. Create graph from edgelist using igraph and ggnetwork
g1 <- graph_from_data_frame(edges, directed = FALSE, vertices = SC_SID)

#specify vertical layout using sugiyama
LO <- layout.sugiyama(g1, V(g1)$layer)

#convert to ggnetwork object
n <- ggnetwork(g1, layout = LO$layout)

#plot
p4 <- ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "black")+
  geom_nodes(aes(color = name,size = Freq))+
  geom_nodelabel_repel(aes(color = name, label = name))+
  theme_blank()+
  ggtitle("")+
  guides(color = FALSE)

p4

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/09_spatial/spatialcontext_CNMap_example.pdf",width = 16,height = 14)
(p1+p2)/(p3+p4)
dev.off()

```
```{r test interaction function}





```










