---
title: "IF_Basoluminal_Quantification"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Software requirements

```{r lib, message=FALSE,warning=FALSE}
#Data containers
library(SpatialExperiment)
library(SingleCellExperiment)

#Data processing
library(scuttle)
library(scater)
library(scran)
library(imcRtools)

#Data visualization
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)

#Utilities
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(psych)

#Batch correction
library(batchelor)
```

```{r}
sce <- readRDS("/mnt/rcc_volume/TNBC/data_analysis/05_sce_TNBC_tumor_new.rds")
```

# 1. IF Basoluminal - Quantification 

Data generated with QuPath. 
Steps included tumor region designation, single-cell segmentation (nuclei detection + expansion), thresholding on markers (CK5/7/19)

```{r}
# read in data 
dat <- read_csv("/mnt/rcc_volume/TNBC/data_analysis/03_data/QuPath_Basoluminal_Quantification_Annotation.csv")

colnames(dat) <- str_replace(colnames(dat),"Num ","")
colnames(dat) <- str_replace(colnames(dat),"Cy3","CK7")
colnames(dat) <- str_replace(colnames(dat),"Cy5","CK19")
colnames(dat) <- str_replace(colnames(dat),"Cy7","CK5")
colnames(dat) <- str_replace(colnames(dat)," ","")
colnames(dat) <- str_replace(colnames(dat)," ","")

dat$PID <- str_split(dat$Image,".czi", simplify = TRUE)[,1]


# Optional: Kick out samples for which tumor selection was difficult in QuPath / low basal/luminal cells present

dat <- dat %>% filter(!PID %in% c("B10_23314","B07_41959","B10_59762"))

```


```{r}
cur_dat <- dat %>% select("CK7","CK7:CK19","CK19","CK5","CK5:CK7","CK5:CK7:CK19","CK5:CK19", "PID") %>% pivot_longer(1:7,values_to = "n", names_to = "CK_group")
```

```{r}
# CK group fraction dataframe
cur_in <- cur_dat %>% as.data.frame %>% group_by(PID, CK_group) %>% summarise(mean_n = mean(n)) %>% group_by(PID) %>% mutate(total_n = sum(mean_n))

cur_df <- cur_in %>% group_by(PID) %>% mutate(fra = mean_n/sum(mean_n))
cur_df <- cur_df %>% pivot_wider(id_cols = "PID", names_from = "CK_group",values_from = "fra") %>% column_to_rownames("PID")

# Option: Perform hierarchical clustering on fractions
dd <- dist(cur_df, method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
plot(hc)
row_sorted <- hc$order

# Option: order by triple-positive cell fraction
row_sorted_fra <- cur_in %>% group_by(PID) %>% mutate(fra = mean_n/sum(mean_n)) %>% filter(CK_group == "CK5:CK7:CK19") %>% arrange(desc(fra)) %>% pull(PID)

# Define annotation dataframe
cur_df <- cur_df[row_sorted_fra,]
cur_df <- cur_df[,c("CK5:CK7:CK19", "CK5:CK7","CK7:CK19","CK19","CK7","CK5","CK5:CK19")]

anno <- cur_in
anno <- anno[match(rownames(cur_df), anno$PID),]

# Color
cur_color <- setNames(brewer.pal(7, "Paired"), colnames(cur_df))

# Plot annotation dataframe with ComplexHeatmap
fraction_anno <- HeatmapAnnotation(PID = anno_text(anno$PID), 
                                   CK_fractions = anno_barplot(cur_df, gp=gpar(fill=cur_color),
                                                               bar_width = 1, height = unit(25,"cm"), width = unit(11,"cm"),
                                                               show_row_names = TRUE, row_names_side = "left"),
                                   total_n = anno_barplot(anno$total_n, border = TRUE),
                                   gap = unit(1,"mm"),
                                   which = "row")
h <- fraction_anno + NULL

lgd <- Legend(title = "CK_group", at = colnames(cur_df), 
              legend_gp = gpar(fill = cur_color))
             
# Plot
draw(h,annotation_legend_list = list(lgd))
```
```{r}
# Focus on CK5/7
cur_dat <- cur_dat %>% ungroup() %>% mutate(CK5_7_group = ifelse(CK_group %in% c("CK5","CK5:CK19"),"CK5+", 
                                         ifelse(CK_group %in% c("CK7","CK7:CK19"),"CK7+", 
                                                ifelse(CK_group %in% c("CK5:CK7","CK5:CK7:CK19"),"CK5/7+", NA)))) %>% filter(!is.na(CK5_7_group))

cur_dat <- cur_dat %>% as.data.frame %>% group_by(PID, CK5_7_group) %>% summarise(mean_n = mean(n)) %>% group_by(PID) %>% mutate(total_n = sum(mean_n))                                     

cur_CK <- cur_dat %>% group_by(PID) %>% mutate(fra = mean_n/sum(mean_n))

cur_CK <- cur_CK %>% pivot_wider(id_cols = "PID", names_from = "CK5_7_group",values_from = "fra") %>% column_to_rownames("PID")

# Option: order by triple-positive cell fraction
row_sorted_fra <- cur_dat %>% group_by(PID) %>% mutate(fra = mean_n/sum(mean_n)) %>% filter(CK5_7_group == "CK5/7+") %>% arrange(desc(fra)) %>% pull(PID)

# Define annotation dataframe
cur_CK <- cur_CK[row_sorted_fra,]
cur_CK <- cur_CK[,c("CK5/7+", "CK5+", "CK7+")]

anno <- cur_dat 
anno <- anno[match(rownames(cur_CK), anno$PID),]

# Color
cur_color_CK <- setNames(wes_palette("Moonrise3")[1:3], c("CK5/7+", "CK5+", "CK7+"))

# Plot annotation dataframe with ComplexHeatmap
fraction_anno <- HeatmapAnnotation(PID = anno_text(anno$PID), 
                                   CK_fractions = anno_barplot(cur_CK, gp=gpar(fill=cur_color_CK),
                                                               bar_width = 1, height = unit(25,"cm"), width = unit(11,"cm"),
                                                               show_row_names = TRUE, row_names_side = "left"),
                                   total_n = anno_barplot(anno$total_n, border = TRUE),
                                   gap = unit(1,"mm"),
                                   which = "row")
h_1 <- fraction_anno + NULL

lgd_1 <- Legend(title = "CK_group", at = colnames(cur_CK), 
              legend_gp = gpar(fill = cur_color_CK))
             
# Plot
draw(h_1,annotation_legend_list = list(lgd_1))
```
Combined 

```{r}
fraction_combined <- HeatmapAnnotation(PID = anno_text(anno$PID), 
                                       CK57_fractions = anno_barplot(cur_CK, gp=gpar(fill=cur_color_CK),
                                                               bar_width = 1, height = unit(25,"cm"), width = unit(11,"cm"),
                                                               show_row_names = TRUE, row_names_side = "left"),
                                       CK_fractions = anno_barplot(cur_df, gp=gpar(fill=cur_color),
                                                               bar_width = 1, height = unit(25,"cm"), width = unit(11,"cm"),
                                                               show_row_names = TRUE, row_names_side = "left"),
                                       total_n = anno_barplot(anno$total_n, border = TRUE),
                                       gap = unit(2,"mm"),
                                       which = "row")

h_comb <- fraction_combined + NULL
draw(h_comb,annotation_legend_list = list(lgd,lgd_1))
```

Correlation between IMC measurements 

```{r}
PID_CK57 <- colData(sce) %>% as.data.frame() %>% select(CK5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_CK57 <- PID_CK57 %>% group_by(PID) %>% filter(CK5_7_score != "CK5/7-") %>% mutate(fra = Freq/sum(Freq))

PID_CK57_df <- PID_CK57 %>% pivot_wider(id_cols = "PID",names_from = "CK5_7_score",values_from = "fra") %>% column_to_rownames("PID")


cur_CK_out <- cur_CK
rownames(cur_CK_out) <- str_replace(rownames(cur_CK_out),"_",".")

rownames(cur_CK_out)[rownames(cur_CK_out) == "B13.6577"] <- "B13.06577"
rownames(cur_CK_out)[rownames(cur_CK_out) == "B07.8412"] <- "B07.08412"

cur_cor <- PID_CK57_df[rownames(PID_CK57_df) %in% rownames(cur_CK_out),]

cur_cor <- cur_cor %>% select("CK5/7+")

cur_cor$`CK5/7+_IF` <- cur_CK_out$`CK5/7+`[match(rownames(cur_cor), rownames(cur_CK_out))]
cur_cor$PID <- rownames(cur_cor)

spearman_test <- cor.test(cur_cor[,1],cur_cor[,2], method = "pearson")

ggplot(cur_cor, aes(x = `CK5/7+`, y = `CK5/7+_IF`))+
  geom_smooth(method='lm', formula= y~x, se = TRUE, color = "darkgrey")+
  geom_point(aes(color = PID), size = 3)+
  ylab("CK5/7 Fraction IF")+
  xlab("CK5/7 Fraction IMC")+
  theme_classic()+
  scale_color_manual(values=metadata(sce)$colors$PID)+
  ggtitle(paste("Pearson r = ",round(spearman_test$estimate, digits = 2),"/ p =",round(spearman_test$p.value, digits = 4)))
```

Check overlap of CK19 and CK7 positive cells

```{r}
cur_719 <- cur_in %>% filter(CK_group != "CK5") %>% mutate(CK7_19_group = ifelse(CK_group %in% c("CK5:CK19","CK19"),"CK19+", 
                                                ifelse(CK_group %in% c("CK5:CK7","CK7"),"CK7+", "CK7_19+"))) 

cur_719 <- cur_719 %>% as.data.frame %>% group_by(PID, CK7_19_group) %>% summarise(sum_n = sum(mean_n))                               

cur_719 <- cur_719 %>% group_by(PID) %>% mutate(fra = sum_n/sum(sum_n))

cur_719$PID <- str_replace(cur_719$PID, "_",".")
cur_719$PID[cur_719$PID == "B13.6577"] <- "B13.06577"
cur_719$PID[cur_719$PID == "B07.8412"] <- "B07.08412"

ggplot(cur_719, aes(x=CK7_19_group, y=fra))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic()
```

# 2. IF BL Stem - Quantification 

Data generated with QuPath. 
Steps included tumor region designation, single-cell segmentation (nuclei detection + expansion), thresholding on markers (CK5/19/SOX9)

```{r}
# read in data 
#dat <- read_csv("/mnt/rcc_volume/TNBC/data_analysis/03_data/TNBC_IF_BasolumStem_Anno.csv")
#dat <- read_csv("/mnt/rcc_volume/TNBC/data_analysis/03_data/IF_TNBC_BasoLum_Stem_2.csv")

colnames(dat) <- str_replace(colnames(dat),"Num ","")
colnames(dat) <- str_replace(colnames(dat),"Cy3","CK5")
colnames(dat) <- str_replace(colnames(dat),"Cy5","CK19")
colnames(dat) <- str_replace(colnames(dat),"Cy7","SOX9")
colnames(dat) <- str_replace(colnames(dat)," ","")
colnames(dat) <- str_replace(colnames(dat)," ","")

dat$PID <- str_split(dat$Image,".czi", simplify = TRUE)[,1]

dat$PID

# Optional: Kick out samples for which tumor selection was difficult in QuPath / low basal/luminal cells present
dat <- dat %>% filter(!PID %in% c("B10_23314","B07_1959","B10_59762")) %>% filter(!is.na(PID))
```


```{r}
cur_dat <- dat %>% select(c("CK5", "CK5:CK19", "CK5:CK19:SOX9", 
"CK5:SOX9", "CK19", "CK19:SOX9", "SOX9", "PID")) %>% pivot_longer(1:7,values_to = "n", names_to = "CK_group")
```

```{r}
#Focus: Stem-like basoluminal cells
cur_out <- cur_dat %>% filter(CK_group %in% c("CK5:CK19:SOX9", "CK5:CK19")) %>% group_by(PID) %>% mutate(sum = sum(n)) %>% mutate(fra = n/sum) %>% group_by(PID, CK_group) %>% mutate(fra = sum(n)/sum) %>% filter(CK_group == "CK5:CK19:SOX9") %>% select(PID, fra) %>% unique()

cur_out$PID <- str_replace(cur_out$PID, "_", ".")

# Fraction of stem-like vs. not stem-like
ggplot(cur_out, aes(x=CK_group, y=fra))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  coord_cartesian(ylim = c(0,1))+
  theme_classic()
```

```{r}
#General: Stem-like cells of basal, luminal and basoluminal 
cur_out <- cur_dat %>% 
  mutate(CK_subgroup = ifelse(CK_group %in% c("CK5:CK19:SOX9", "CK5:CK19"), "BL", 
                                                   ifelse(CK_group %in% c("CK5:SOX9", "CK5"), "B",
                                                          ifelse(CK_group %in% c("CK19:SOX9", "CK19"), "L",NA)))) %>% 
  filter(!is.na(CK_subgroup)) %>% 
  group_by(PID, CK_subgroup) %>% 
  mutate(sum = sum(n)) %>% 
  group_by(PID, CK_subgroup, CK_group) %>% 
  mutate(fra = sum(n)/sum) %>% 
  select(PID, CK_group, CK_subgroup, sum,fra) %>% filter(str_detect(CK_group, "SOX9")) %>% unique()

cur_out$PID <- str_replace(cur_out$PID, "_", ".")
cur_out$CK_subgroup <- factor(cur_out$CK_subgroup, levels = c("BL","L","B"))

# Fraction of stem-like vs. not stem-like
ggplot(cur_out, aes(x=CK_subgroup, y=fra))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  coord_cartesian(ylim = c(0,1))+
  theme_classic()+
  stat_compare_means(method = "kruskal.test") # group differences

compare_means(fra ~ CK_subgroup, cur_out, method = "wilcox.test") # between group differences

# Number of cells per PID analyzed 
ggplot(cur_out, aes(x=CK_subgroup, y=sum))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  #coord_cartesian(ylim = c(0,1))+
  theme_classic()

```


```{r}
# Heatmap for basal-luminal overview
cur_in <- cur_dat %>% 
  mutate(CK_subgroup = ifelse(CK_group %in% c("CK5:CK19:SOX9", "CK5:CK19"), "BL", 
                                                   ifelse(CK_group %in% c("CK5:SOX9", "CK5"), "B",
                                                          ifelse(CK_group %in% c("CK19:SOX9", "CK19"), "L",NA)))) %>% 
  filter(!is.na(CK_subgroup)) %>% 
  group_by(PID, CK_subgroup) %>% 
  mutate(sum_sub = sum(n)) %>% 
  group_by(PID) %>% 
  mutate(fra = sum_sub/sum(n)) %>% 
  select(PID, CK_subgroup ,fra) %>% 
  unique()

cur_df <- cur_in %>% pivot_wider(id_cols = "PID", names_from = "CK_subgroup",values_from = "fra") %>% column_to_rownames("PID")

# Option: Perform hierarchical clustering on fractions
dd <- dist(cur_df, method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
plot(hc)
row_sorted <- hc$order

# Option: order by triple-positive cell fraction
row_sorted_fra <- cur_in %>% filter(CK_subgroup == "BL") %>% arrange(desc(fra)) %>% pull(PID)

# Define annotation dataframe
cur_df <- cur_df[row_sorted_fra,]
cur_df <- cur_df[,c("BL","L","B")]

anno <- cur_in
anno <- anno[match(rownames(cur_df), anno$PID),]

# Color
cur_color <- setNames(brewer.pal(7, "Paired"), colnames(cur_df))

# Plot annotation dataframe with ComplexHeatmap
fraction_anno <- HeatmapAnnotation(PID = anno_text(anno$PID), 
                                   CK_fractions = anno_barplot(cur_df, gp=gpar(fill=cur_color),
                                                               bar_width = 1, height = unit(25,"cm"), width = unit(11,"cm"),
                                                               show_row_names = TRUE, row_names_side = "left"),
                                   #total_n = anno_barplot(anno$total_n, border = TRUE),
                                   gap = unit(1,"mm"),
                                   which = "row")
h <- fraction_anno + NULL

lgd <- Legend(title = "CK_group", at = colnames(cur_df), 
              legend_gp = gpar(fill = cur_color))
             
# Plot
draw(h,annotation_legend_list = list(lgd))
```

## Save plots

```{r}
## Save plots as pdf

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/IF_Basoluminal/TNBC_QuPath_Basoluminal_Heatmap_combined.pdf",width = 12, height = 4, onefile = FALSE)
draw(h_comb,annotation_legend_list = list(lgd,lgd_1))
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/IF_Basoluminal/TNBC_QuPath_Basoluminal_IF_IMC_cor.pdf", height = 5, width = 6)
ggplot(cur_cor, aes(x = `CK5/7+`, y = `CK5/7+_IF`))+
  geom_smooth(method='lm', formula= y~x, se = TRUE, color = "darkgrey")+
  geom_point(aes(color = PID), size = 3)+
  ylab("CK5/7 Fraction IF")+
  xlab("CK5/7 Fraction IMC")+
  theme_classic()+
  scale_color_manual(values=metadata(sce)$colors$PID)+
  ggtitle(paste("Pearson r = ",round(spearman_test$estimate, digits = 2),"/ p =",round(spearman_test$p.value, digits = 4)))
dev.off()

pdf("/mnt/rcc_volume/TNBC/data_analysis/01_figures/IF_Basoluminal/TNBC_QuPath_Basoluminal_CK7_19.pdf", height = 5, width = 6)
ggplot(cur_719, aes(x=CK7_19_group, y=fra))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic()
dev.off()
```

