---
title: "01_lib_read_data"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: / 

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Script 1: Read in data

## Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(CATALYST)
library(scuttle)
library(scater)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(viridis)
library(tidyverse)
library(openxlsx)
library(stringr)
library(BiocParallel)
```

## Read in single-cell information

based on [SingleCellExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) and IMC data preprocessed with the [steinbock](https://bodenmillergroup.github.io/steinbock/latest/) framework. 

```{r read and inspect steinbock}
sce <- read_steinbock("/mnt/central_nas/tnbc_volume/TNBC/",intensities_folder = "intensities",regionprops_folder = "regionprops", graphs_folder = "neighbors",return_as = "sce")
sce
```

## SCE processing

```{r set-colnames}
colnames(sce) <- sce$cell_id <- paste0(sce$sample_id, "_", sce$ObjectNumber)

rownames(sce) <- rowData(sce)$clean_target
```

Read in metadata

```{r add-meta}
meta <- read.csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Metadata_Combined_TNBC_LM.csv",row.names = 1)
meta_249 <- read.csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/ZTMA249_TNBC_LM_processed_bysample.csv",row.names = 1) %>% select(ID_Number,PID)

## Add PID
PID_df <- data.frame(sample_id = unique(sce$sample_id),
           ZTMA174_ID = str_split(unique(sce$sample_id),"_",simplify = TRUE)[,3],
           ZTMA249_ID = str_split(unique(sce$sample_id),"_",simplify = TRUE)[,4])

PID_df$ZTMA249_PID <- meta_249$PID[match(PID_df$ZTMA249_ID,meta_249$ID_Number)]

PID_df <- PID_df %>% mutate(PID = ifelse(ZTMA174_ID == "ID",ZTMA249_PID,ZTMA174_ID))

sce$PID <- PID_df$PID[match(sce$sample_id,PID_df$sample_id)]

# Remove sample with no available PID (ZTMA249.1_ZTMA249_ID_247_CX7Y1_99)
sce <- sce[,!is.na(sce$PID)]

## Add other relevant metadata based on PID
for (entry in colnames(meta)[!colnames(meta) %in% c("PID","ID_Number","OS_data","DFS_data")]) {
  colData(sce)[,entry] <- meta[,entry][match(colData(sce)[,"PID"],meta[,"PID"])]
}
```

Subset markers of interest for downstream analysis.

```{r select-features}
rowData(sce)$use_channel <- !grepl("DNA|Histone", rowData(sce)$clean_target)
```

Define color schemes

```{r color-scheme}
colors <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/color_meta.rds")
colors

metadata(sce)$colors <- colors
```

### Count transformation

```{r transform-counts, message=FALSE, fig.width=7,fig.height=10}
set.seed(22)
cur_id <- sample(sce$sample_id,5)

#Raw counts
multi_dittoPlot(sce[,sce$sample_id %in% cur_id], vars = rownames(sce)[rowData(sce)$clean_target %in% c("panCK","Vimentin","CD45","CD3","CD68")],
               group.by = "PID", plots = c("ridgeplot"), 
               assay = "counts", 
               color.panel = metadata(sce)$colors$PID)

#Transform counts with acsinh and a cofactor of 1
assay(sce, "exprs") <- asinh(counts(sce)/1)

multi_dittoPlot(sce[,sce$sample_id %in% cur_id], vars = rownames(sce)[rowData(sce)$clean_target %in% c("panCK","Vimentin","CD45","CD3","CD68")],
               group.by = "PID", plots = c("ridgeplot"), 
               assay = "exprs", 
               color.panel = metadata(sce)$colors$PID)

```

# Save objects

Note that masks and images were loaded, modified and saved in the [00_seg_mesmer_QC.Rmd]() script. 

```{r save-objects-read-data}
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_sce_TNBC.rds")
```

# Session Info 

```{r}
sessionInfo()
```

