---
title: "02_spillover_matrix"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: / 

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Script 2: Spillover correction

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(CATALYST)
library(scuttle)
library(scater)
library(imcRtools)
library(dittoSeq)
library(tidyverse)
library(openxlsx)
library(stringr)
library(BiocParallel)
library(graphics)
```

## Single-cell data compensation

We use functionality of the `CATALYST` R/Bioconductor package for spillover estimation and correction.

**NOTE**: We used the spillover matrix from Fischer et al. for downstream analysis (DOI: 10.5281/zenodo.7494509, see Methods). 

```{r}
sm_jana <- read.csv(sm, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/spillover_matrix_Jana_mod.csv")
```

```{r single-cell-compensation}
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_sce_TNBC.rds")

rowData(sce)$channel_name <- paste0(rowData(sce)$channel, "Di")

sce_jana <- compCytof(sce, sm_jana, 
                 transform = TRUE, cofactor = 1,
                 isotope_list = isotope_list, 
                 overwrite = FALSE)
```

```{r visualize-single-cell-spillover, message=FALSE}
#Simple scatterplot
before <- dittoScatterPlot(sce, x.var = "H3K27me3", y.var = "Bcl2",
                           assay.x = "exprs", assay.y = "exprs") +
    ggtitle("Before compensation")

after_1 <- dittoScatterPlot(sce_jana, x.var = "H3K27me3", y.var = "Bcl2",
                          assay.x = "compexprs", assay.y = "compexprs") +
    ggtitle("After compensation - Jana")

before+after+after_1

#Smoothed density scatterplot
buylrd = c("#313695", "#4575B4", "#74ADD1", "#ABD9E9", "#E0F3F8", "#FFFFBF",
           "#FEE090", "#FDAE61", "#F46D43", "#D73027", "#A50026") 
myColRamp = colorRampPalette(c(buylrd))


smoothScatter(x=assay(sce,"exprs")[rownames(sce) == "H3K27me3"],
              y=assay(sce,"exprs")[rownames(sce) == "Bcl2"],
              colramp=myColRamp,
              main="Before compensation",
              xlab="H3K27me3 (La139)",
              ylab="Bcl2 (Gd155)")
  
smoothScatter(x=assay(sce_jana,"compexprs")[rownames(sce_jana) == "H3K27me3"],
              y=assay(sce_jana,"compexprs")[rownames(sce_jana) == "Bcl2"],
              colramp=myColRamp,
              main="After compensation - Jana",
              xlab="H3K27me3 (La139)",
              ylab="Bcl2 (Gd155)")
```

```{r overwrite-assays}
assay(sce, "counts") <- assay(sce_jana, "compcounts") 
assay(sce, "exprs") <- assay(sce_jana, "compexprs") 
assay(sce, "compcounts") <- assay(sce, "compexprs") <- NULL
```

## Save objects

```{r save, message = F}
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/02_sce_TNBC_spillcor.rds")
```

# Session Info 

```{r}
sessionInfo()
```