---
title: "03_QC"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: / 

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Script 3: Quality control

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(BiocParallel)
library(BiocNeighbors)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(scuttle)
library(mclust)
```

## Read in the data

```{r read-data, message=FALSE}
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_TNBC_new.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/02_sce_TNBC_spillcor.rds")
```

## Segmentation quality control

Was performed before - See `00_seg_QC.Rmd`.

```{r overlay-masks, message=FALSE}
#10 preselected images
img_files <- list.files("/mnt/central_nas/tnbc_volume/TNBC/img/")
img_ids<- c(which(str_detect(img_files ,"ZTMA249_ID_271_CX7Y4")),which(str_detect(img_files ,"ZTMA249_ID_82_AX2Y11")), which(str_detect(img_files ,"ZTMA249_ID_179_BX3Y8")), which(str_detect(img_files ,"ZTMA249_ID_183_BX7Y8")), which(str_detect(img_files ,"ZTMA174_B08.26868_79_AX7Y10")))

# Normalize and clip image
cur_images <- images[img_ids]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#by epithelial channels
plotPixels(cur_images,
           mask = masks[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "Vimentin", "SMA","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "Vimentin" = c("black", "cyan"),
                         "SMA" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/spillcor_selectedimages_epi.png", scale = 2))

#by immune channels
plotPixels(cur_images,
           mask = masks[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD3", "CD8a", "CD68", "CD20", "CD44"),
           colour = list("CD3" = c("black", "yellow"),
                         "CD20" = c("black", "magenta"),
                         "CD8a" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/spillcor_selectedimages_immune.png", scale = 2))
```

## Image-level QC

### HER2 case exclusion 

We will exclude patients from ZTMA174 that were later identified as HER2-positive cases or have outlying HER2/ER expression. 

```{r HER2/ER-image-mean}
cur_sce <- sce

# GMM for panCK (tumor cells)
GMM_CK <- Mclust(assay(cur_sce, "exprs")["panCK",], G = 2)

cur_sce$GMM_CK <- GMM_CK$classification

cur_sce <- cur_sce[,cur_sce$GMM_CK == "2"]

image_mean <- aggregateAcrossCells(cur_sce, 
                                   ids = cur_sce$PID, 
                                   statistics="mean",
                                   use.assay.type = "counts")

## 1. HER2 
dat <- dittoBoxPlot(image_mean, var = "HER2", group.by = "HER2_case", assay = "counts", data.out = TRUE)
dat_df <- dat$data

p2 <- ggplot(dat_df, aes(x = grouping, y = var.data))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = grouping), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_brewer(palette = "Set1")+
  xlab("TMA")+
  ylab("HER2 exprs (panCK+ cells)")+
  theme_classic()

p2 

## remove HER2 positive cases
## kick out 1 case with very high HER2 expression
sce <- sce[,sce$HER2_case == FALSE]
sce <- sce[,!sce$PID %in% c("B08.32458")]

## 2. ER
dat <- dittoBoxPlot(image_mean, var = "ER", group.by = "HER2_case", assay = "counts", data.out = TRUE)
dat_df <- dat$data

ggplot(dat_df, aes(x = grouping, y = var.data))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = grouping), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_brewer(palette = "Set1")+
  xlab("TMA")+
  ylab("ER exprs (panCK+ cells)")+
  theme_classic()

## kick out 4 cases with suspiciously high ER expression
sce <- sce[,!sce$PID %in% c("B14.59347","B07.07453","B13.30305","B08.41783")]
sce
```

### Tissue area and number of cells

We will identify images that have unusually low cell coverage, small tissue areas or little number of cells.

```{r cell-density}
#Load TMA tissue masks
tissue_area <- read_csv("/mnt/central_nas/tnbc_volume/TNBC/tissue_mask/tissue_area_TissueRescaled.csv") 
sce$tissue_area <- as.numeric(tissue_area$AreaShape_Area[match(colData(sce)$sample_id,str_split(tissue_area$FileName_Tissue_Mask, "_Pro",simplify = TRUE)[,1])])

cell_density <- colData(sce) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    summarize(cell_area = sum(area),
              tissue_area = mean(tissue_area)) %>%
  mutate(covered_area = (cell_area / tissue_area)*100) %>%
  mutate(n_cells = as.numeric(table(colData(sce)$sample_id))) %>%
  arrange(covered_area)

ggplot(cell_density) +
  geom_point(aes(reorder(sample_id,covered_area), covered_area, color = covered_area))+ 
  ylim(c(0, 100))+
  geom_hline(yintercept = median(cell_density$covered_area), linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$covered_area)[2],linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$covered_area)[4],linetype="dashed", color = "grey")+
  theme_classic(base_size = 15)+
  theme(axis.text.x = element_blank())+
  ylab("% covered area")+ 
  xlab("")+
  scale_color_viridis()

ggplot(cell_density) +
  geom_point(aes(reorder(sample_id,n_cells), n_cells, color = n_cells))+
  geom_hline(yintercept = median(cell_density$n_cells), linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$n_cells)[2],linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(cell_density$n_cells)[4],linetype="dashed", color = "grey")+
  theme_classic(base_size = 15)+
  theme(axis.text.x = element_blank())+
  ylab("Number of cells per ROI")+ 
  xlab("")+
  scale_color_viridis()

## Visualize critical images
# Criteria: n_cells < 300, covered_area < 8, tissue_area < 150000
# We will also exclude split acquisitions via #str_detect(cell_density$sample_id, "split")
image_id_cri <- cell_density %>% filter(n_cells < 300 | covered_area < 8 | tissue_area < 150000) %>% pull(sample_id)

# Normalize and clip images
cur_images <- images[names(images) %in% image_id_cri]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

plotPixels(cur_images,
           mask = masks[names(masks) %in% image_id_cri],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "CD3","CD68","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "CD3" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))

# Retain one more image that looks fine 
image_id_cri <- image_id_cri[!image_id_cri == c("ZTMA249.3_ZTMA249_ID_323_CX3Y11_328")]

#Remove outlying images and cells (n=1728 cells)
summary(colData(sce)$sample_id %in% image_id_cri)
sce <- sce[,!colData(sce)$sample_id %in% image_id_cri]
sce
```

## Cell-level QC

### Cell size QC

We will remove very small cells from downstream analysis.

```{r cellsize_dis, message=FALSE,warning=FALSE}
colData(sce) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    ggplot() +
        geom_boxplot(aes(reorder(sample_id,area), area)) +
        theme_classic(base_size = 15) + 
        theme(axis.text.x = element_blank()) +
        ylab(~ "Cell area " (mu * m^2)) + xlab("")+
  geom_hline(yintercept = median(colData(sce)$area), linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(colData(sce)$area)[2],linetype="dashed", color = "grey")+
  geom_hline(yintercept = quantile(colData(sce)$area)[4],linetype="dashed", color = "grey")

#Cell area range
summary(sce$area)

#Only keep cells with an area above 5 (n=2909 cells)
sum(sce$area <= 5)
sce <- sce[,sce$area > 5]
sce
```

## Save objects

The modified `SingleCellExperiment` object is saved for further downstream analysis.

```{r save-objects-quality-control}
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_sce_TNBC_QC.rds")
```

# Session Info 

```{r}
sessionInfo()
```