---
title: "04_SingleCell_InitialClustering"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: sce_ALL.rds, sce_ALL_sub.rds 

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Script 4: Distinguish Tumor/Non-Tumor cells

Here, we will distinguish between epithelial (tumor) and non-epithelial (non-tumor) cells.

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scran)
library(bluster)
library(imcRtools)
library(cytomapper)
library(cytoviewer)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(VennDiagram)
library(easyalluvial)
library(psych)
library(Rphenoannoy)
library(easyalluvial)
library(UpSetR)
library(cowplot)
```

## Read in data 

```{r read-data-batch-correction, message=FALSE}
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_sce_TNBC_QC.rds")
sce
```

## Perform "overclustering"

Next, we will run a granular Rphenograph (Jaccard-based weights, Louvain clustering as community detection algorithm) clustering. 

**NOTE**: Please be aware that some analyse including unsupervised clustering involves random processes and may not be full reproducible across systems. We did, however, set seeds for all such analysis. To ensure full reproducibility, please use the information stored in the SCE objects available on Zenodo and the figure generation scripts. 

```{r}
set.seed(22)

mat <- t(assay(sce, "exprs")[rowData(sce)$use_channel==TRUE,])

#k=30
out <- Rphenoannoy(mat, k = 30)

clusters <- factor(membership(out[[2]]))

sce$cluster <- clusters

unique(sce$cluster)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
```

## Use GMM for panCK

We will use a two component gaussian mixture model to distinguish between panCK+ and panCK- cells. 

```{r GMM_CK, message=FALSE,warning=FALSE}
## Gaussian mixture model to distinguish tumor/stroma based on panCK expression
GMM_CK <- Mclust(assay(sce, "exprs")["panCK",], G = 2)

GMM_df <- data.frame(class_CK = ifelse(GMM_CK$classification == 2 & GMM_CK$uncertainty == 0, "panCK+", "panCK-"), 
                     panCK_exprs = assay(sce, "exprs")["panCK",],
                     uncertainty = GMM_CK$uncertainty)

#Visualize GMM
GMM_plot_exp <- ggplot(GMM_df, aes(x = panCK_exprs, fill = class_CK))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")
  
GMM_plot_unc <- ggplot(GMM_df, aes(x = panCK_exprs, y = uncertainty, color = class_CK))+
  geom_line()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_color_brewer(palette = "Set1")

GMM_plot_exp/GMM_plot_unc

#Add GMM CK class to SCE
sce$GMM_CK <- GMM_df$class_CK

## Gaussian mixture model to distinguish tumor/stroma based on Ecad expression
GMM_Ecad <- Mclust(assay(sce, "exprs")["Ecad",], G = 2)

GMM_df_1 <- data.frame(class_Ecad = ifelse(GMM_Ecad$classification == 2, 
                                           "Ecad+", "Ecad-"), 
                     Ecad_exprs = assay(sce, "exprs")["Ecad",],
                     uncertainty = GMM_Ecad$uncertainty)

GMM_plot_exp_1 <- ggplot(GMM_df_1, aes(x = Ecad_exprs, fill = class_Ecad))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")

GMM_plot_unc_1 <- ggplot(GMM_df_1, aes(x = Ecad_exprs, y = uncertainty, color = class_Ecad))+
  geom_line()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_color_brewer(palette = "Set1")

GMM_plot_exp_1/GMM_plot_unc_1

#Add GMM CK class to SCE
sce$GMM_Ecad <- GMM_df_1$class_Ecad

#Save sce after clustering and GMM
#saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
```

### Check correlation of panCK and Ecad as epithelial cell markers

```{r marker correlations, message = FALSE}
GOI <- c("panCK","Ecad")
cor_sce <- as.data.frame(t(assay(sce,"exprs")[rowData(sce)$clean_target %in% GOI,]))
pairs.panels(cor_sce,
             method = "spearman",
             hist.col = "#A6CEE3",
             digits = 2)
```

## Astir for automated cell type assignment 

```{r astir-expo}
#Export arsinh normalized expression matrix as csv
#Add sampleID as covariate in design matrix
expr_astir <- assay(sce,"exprs") %>% t() %>% as.data.frame() %>% mutate(sample = sce$sample_id)

write.csv(expr_astir, file = "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_astir/expr_TNBC_astir.csv",row.names = TRUE)
```

`Next: Run Astir in python - See 00_astir_TNBC.ipynb`

```{r astir-ana, message = FALSE,warning=FALSE}
#Import Astir cell type assignments
astir_type <- read_csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_astir/cell_types_TNBC.csv",col_names = c("ID","cell_type"),skip = 1)

astir_sum <- astir_type %>% as.data.frame %>% group_by(cell_type) %>% dplyr::dplyr::count()

sce$astir_type <- astir_type$cell_type

#Mark channels used for cell type assignment 
astir_channel = c("panCK","Ecad","CD3","CD8a","CD4","CD20","CD68","SMA","CD11c")

rowData(sce)$use_astir <- rowData(sce) %>% as.data.frame() %>% mutate(use_astir = ifelse(rownames(sce) %in% astir_channel,TRUE,FALSE)) %>% pull(use_astir)
```

## Visualization 

### Rich Heatmap

Heatmap of clusters with GMM and ASTIR information 

```{r heatmap, message=FALSE, warning=FALSE}
#Read sce after clustering and GMM
#sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")

### 1. Heatmap body ###

# Min-Max [0,1] normalization of arcsinh-transformed counts
assay(sce, "min_max") <- assay(sce, "exprs") - rowMins(assay(sce, "exprs"))
assay(sce, "min_max") <- assay(sce, "min_max") / (rowMaxs(assay(sce, "exprs")) - rowMins(assay(sce, "exprs")))

# Aggregrate across cells with the mean
mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce)$use_channel)

# Heatmap body color 
col_main = viridis(100)

### 2. Heatmap annotation ###

### 2.1 Spatial features

#Add number of neighbors to sce object (saved in ColPair)
n_neighbors <- colPair(sce) %>% as.data.frame() %>% group_by(from) %>% dplyr::count() %>% arrange(desc(n))
sce$n_neighbors <- n_neighbors$n[match(seq_along(colnames(sce)),n_neighbors$from)]
sce$n_neighbors <- sce$n_neighbors %>% replace_na(0)

#Select spatial features and average over clusters 
spatial <- colData(sce) %>% as.data.frame %>% select(area,eccentricity,cluster,n_neighbors)
spatial <- spatial %>% select(-cluster) %>% aggregate(by = list(cluster = spatial$cluster), FUN = mean) %>% column_to_rownames("cluster")

#Define color schemes for spatial features and add to metadata
area_col <- colorRamp2(c(min(spatial$area),median(spatial$area),max(spatial$area)), c(brewer.pal(3,"Blues")[1],brewer.pal(3,"Blues")[2],brewer.pal(3,"Blues")[3]))
metadata(sce)$colors$area <- area_col

eccentricity_col <- colorRamp2(c(min(spatial$eccentricity),median(spatial$eccentricity),max(spatial$eccentricity)), c(brewer.pal(3,"Purples")[1],brewer.pal(3,"Purples")[2],brewer.pal(3,"Purples")[3]))
metadata(sce)$colors$eccentricity <- eccentricity_col

n_neighbors_col <- colorRamp2(c(min(spatial$n_neighbors),median(spatial$n_neighbors),max(spatial$n_neighbors)), c(brewer.pal(3,"RdPu")[1],brewer.pal(3,"RdPu")[2],brewer.pal(3,"RdPu")[3]))
metadata(sce)$colors$n_neighbors <- n_neighbors_col

#Create Spatial Heatmap Annotation
ha_spatial <- HeatmapAnnotation(
    area = spatial$area,
    eccentricity = spatial$eccentricity,
    n_neighbors = spatial$n_neighbors, 
    col = list(area = metadata(sce)$colors$area,
               eccentricity = metadata(sce)$colors$eccentricity,
               n_neighbors = metadata(sce)$colors$n_neighbors),
    border = TRUE,
    which = "row")

### 2.2 Metadata features

#Number of cells per cluster 
n_cells <- colData(mean_sce) %>% as.data.frame %>% select(cluster, ncells)

#GMM - panCK
nr_tum <- colData(sce) %>% as.data.frame %>% select(cluster, GMM_CK) %>% table()
fr_tum <- as.data.frame(nr_tum) %>% pivot_wider(names_from = GMM_CK, values_from = Freq) %>% column_to_rownames("cluster")
fr_tum <- fr_tum %>% transmute(fr_tum = `panCK+`/rowSums(fr_tum)) %>% as.data.frame()

# #Astir
nr_tum_2 <- colData(sce) %>% as.data.frame %>% select(cluster, astir_type) %>% table()
fr_tum_2 <- as.data.frame(nr_tum_2) %>% pivot_wider(names_from = astir_type, values_from = Freq) %>% column_to_rownames("cluster")
fr_tum_2 <- fr_tum_2 %>% transmute(fr_tum = `Epithelial`/rowSums(fr_tum_2)) %>% as.data.frame()

#GMM - Ecad
nr_tum_3 <- colData(sce) %>% as.data.frame %>% select(cluster, GMM_Ecad) %>% table()
fr_tum_3 <- as.data.frame(nr_tum_3) %>% pivot_wider(names_from = GMM_Ecad, values_from = Freq) %>% column_to_rownames("cluster")
fr_tum_3 <- fr_tum_3 %>% transmute(fr_tum_3 = `Ecad+`/rowSums(fr_tum_3)) %>% as.data.frame()

ha_meta <- HeatmapAnnotation(GMM_CK = anno_barplot(fr_tum, width = unit(15, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            GMM_Ecad = anno_barplot(fr_tum_3, width = unit(15, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            Astir_Epi = anno_barplot(fr_tum_2, width = unit(15, "mm"),gp = gpar(fill = "#377EB8", col = 1)),
                            n_cells = anno_barplot(n_cells$ncells, width = unit(15, "mm"),gp = gpar(fill = "violet", col = 1)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(1,"mm"),
                            which = "row")

### 3. Define tumor, mixed and stroma cluster (based on GMM_CK)

anno_cluscat <- fr_tum %>% 
  rownames_to_column(var = "cluster") %>% 
  mutate(cluster_category = ifelse(fr_tum >= 0.7,"Epithelial",ifelse(fr_tum < 0.7 & fr_tum > 0.2,"Mixed","Non_Epithelial")))

#Set colors for cluster_category and add to metdata
cluster_category <- setNames(c("#1F78B4", "#A6CEE3", "#B2DF8A"),sort(unique(anno_cluscat$cluster_category)))                                                                  
metadata(sce)$colors$cluster_category <- cluster_category               

#Visualize GMM_panCK density plots
ggplot(anno_cluscat, aes(fr_tum))+
  geom_density()+
  geom_rug(aes(color = cluster_category))+
  geom_vline(xintercept = 0.7, color = "red3")+
  geom_vline(xintercept = 0.2, color = "red3")+
  theme_classic()+
  theme(axis.title.x = element_blank())+
  scale_color_manual(values = cluster_category)

ha_cluscat <- HeatmapAnnotation(cluster_category = anno_cluscat$cluster_category,
                                col = list(cluster_category = metadata(sce)$colors$cluster_category),
                                border = TRUE,
                                which = "row")



### 4. Plot combined Heatmap
h <- Heatmap(t(assay(mean_sce, "min_max")),
        col = col_main,
        name= "min-max",
        km = 1,
        show_row_names = T, 
        show_column_names =  T, row_split = anno_cluscat$cluster_category,
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")+
  ha_spatial+
  ha_meta+
  ha_cluscat

draw(h)

### Add cluster_categories to sce
#sce$cluster_category <- anno_cluscat$cluster_category[match(sce$cluster,anno_cluscat $cluster)]

#Save sce
#saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
```

### Venn diagrams / Pie plots for categories

```{r prop_tumor, message=FALSE, warning=FALSE}
## Pie plots 
#panCK+ per cluster category
prop_tumor <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(GMM_CK == "panCK+") %>% group_by(cluster_category) %>% dplyr::count()
prop_tumor <- prop_tumor %>% mutate(fr_tumor_clus = n/sum(prop_tumor$n))

pie(prop_tumor$fr_tumor_clus, labels = paste0(prop_tumor$cluster_category," - ",round(prop_tumor$fr_tumor_clus,3),"%"),col=metadata(sce)$colors$cluster_category, main = "panCK+")

#panCK- per cluster category
prop_tumor_1 <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(GMM_CK == "panCK-") %>% group_by(cluster_category) %>% dplyr::count()
prop_tumor_1 <- prop_tumor_1 %>% mutate(fr_tumor_clus = n/sum(prop_tumor_1$n))
pie(prop_tumor_1$fr_tumor_clus, labels = paste0(prop_tumor_1$cluster_category," - ",round(prop_tumor_1$fr_tumor_clus,3),"%"),col=metadata(sce)$colors$cluster_category, main = "panCK-")

#Tumor per GMM_CK
prop_tumor_2 <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(cluster_category == "Epithelial") %>% group_by(GMM_CK) %>% dplyr::count()
prop_tumor_2 <- prop_tumor_2 %>% mutate(fr_tumor_clus = n/sum(prop_tumor_2$n))
pie(prop_tumor_2$fr_tumor_clus, labels = paste0(prop_tumor_2$GMM_CK," - ",round(prop_tumor_2$fr_tumor_clus,3),"%"), col =c("#E41A1C","#377EB8"), main = "Epithelial")

prop_tumor_3 <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(cluster_category == "Mixed") %>% group_by(GMM_CK) %>% dplyr::count()
prop_tumor_3 <- prop_tumor_3 %>% mutate(fr_tumor_clus = n/sum(prop_tumor_3$n))
pie(prop_tumor_3$fr_tumor_clus, labels = paste0(prop_tumor_3$GMM_CK," - ",round(prop_tumor_3$fr_tumor_clus,3),"%"), col =c("#E41A1C","#377EB8"), main = "Mixed")

prop_tumor_4 <- colData(sce) %>% as.data.frame() %>%
  select(GMM_CK,cluster_category) %>% filter(cluster_category == "Non_Epithelial") %>% group_by(GMM_CK) %>% dplyr::count()
prop_tumor_4 <- prop_tumor_4 %>% mutate(fr_tumor_clus = n/sum(prop_tumor_4$n))
pie(prop_tumor_4$fr_tumor_clus, labels = paste0(prop_tumor_4$GMM_CK," - ",round(prop_tumor_4$fr_tumor_clus,3),"%"), col =c("#E41A1C","#377EB8"), main = "Non_Epithelial")
```


## Re-analyse Mixed Clusters

```{r clus_mixed, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
sce_mixed <- sce[,colData(sce)$cluster_category == "Mixed"] #n=60973

#Select cells that are either panCK or Ecad positive based on gaussian mixture model
sce_mixed$cluster_category_2 <- ifelse(sce_mixed$GMM_CK == "panCK+" | sce_mixed$GMM_Ecad == "Ecad+" ,"Epithelial","Non_Epithelial")

sce_mixed$cluster_category_2 %>% table()

spatial_mixed <- colData(sce_mixed) %>% as.data.frame %>% select(area,eccentricity,cluster_category_2,n_neighbors)
spatial_mixed <- spatial_mixed %>% select(-cluster_category_2) %>% aggregate(by = list(cluster_category_2 = spatial_mixed$cluster_category_2), FUN = mean) %>% column_to_rownames("cluster_category_2")

sce_mixed$cluster_category_2_mean_area <- as.character(round(spatial_mixed$area[match(sce_mixed$cluster_category_2, rownames(spatial_mixed))],2))

#Single-cell Heatmap 
dittoHeatmap(sce_mixed, genes = rownames(sce_mixed)[rowData(sce_mixed)$use_channel],
             assay = "exprs", cluster_cols = FALSE, order.by = c("cluster_category_2","GMM_CK","GMM_Ecad","cluster"), scale = "none",
             heatmap.colors = viridis(100), annot.by = c("cluster","GMM_CK","GMM_Ecad","cluster_category_2","cluster_category_2_mean_area"))


# Add cluster_categories_2 to main sce
sce$cluster_category_2 <- sce_mixed$cluster_category_2[match(colnames(sce),colnames(sce_mixed))]
```

### Visualize mixed clusters on images

```{r viz critical clus, message=FALSE, warning=FALSE}
#Load images and masks
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_segtest.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")

##Visualize mixed clusters on images - select 10 images for each category with highest number of cells
mixed_epi_samples <- colData(sce_mixed) %>% as.data.frame %>% group_by(cluster_category_2) %>% filter(cluster_category_2 == "Epithelial") %>% dplyr::count(sample_id) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(sample_id)

mixed_non_epi_samples <- colData(sce_mixed) %>% as.data.frame %>% group_by(cluster_category_2) %>% filter(cluster_category_2 == "Non_Epithelial") %>% dplyr::count(sample_id) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(sample_id)

mixed_samples_viz <- c(mixed_epi_samples,mixed_non_epi_samples)

mixed_samples_viz <- mixed_samples_viz %>% unique() #total of 20 unique images

#Prepare images
cur_images <- images[mixed_samples_viz]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

plotPixels(cur_images,
           mask = masks[mixed_samples_viz],
           object = sce_mixed,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "Vimentin", "SMA"),
           outline_by = "cluster_category_2",
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "Vimentin" = c("black", "cyan"),
                         "SMA" = c("black", "green"),
                         "cluster_category_2" = c("Epithelial" = "#377EB8", "Non_Epithelial" = "#E41A1C")
                         ),
           thick = TRUE,
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = paste0("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/mixed_clusters_viz/cluster_mixed_epi_non_epi_bycluscat.png"), scale = 2)
           )
```

## Define categories for downstream analysis

Use all epithelial clusters and mixed clusters with panCK+/Ecad+ GMM for downstream tumor analysis.

```{r ana_cat, message = FALSE, warning=FALSE}
sce$analysis_cat <- colData(sce) %>% as.data.frame %>% 
  select(cluster_category, cluster_category_2) %>% 
  mutate(analysis_cat = ifelse(sce$cluster_category == "Epithelial" |(sce$cluster_category == "Mixed" & sce$cluster_category_2 == "Epithelial"),"tumor","non_tumor")) %>% 
  pull(analysis_cat)

#color scheme for analysis_cat
analysis_cat <- setNames(c("#E41A1C", "#377EB8"),sort(unique(sce$analysis_cat)))                                                                  
metadata(sce)$colors$analysis_cat <- analysis_cat   

colData(sce) %>% as.data.frame %>% select(analysis_cat) %>% group_by(analysis_cat) %>% dplyr::count()

#Visualize in alluvial plot 
colData(sce) %>% as.data.frame %>% select(cluster_category, analysis_cat)

alluvial_wide(data = colData(sce) %>% as.data.frame %>% select(cluster_category, analysis_cat), 
              max_variables = 4,
              fill_by = 'all_flows',
              order_levels = c("Non_Epithelial","Epithelial","Mixed","tumor","non_tumor"))+
  theme_classic(base_size = 12, base_line_size = 0.5)


#Intersection plot
## Upset plot
#1. tumor   
tumor = list(
analysis_cat_tumor = colData(sce) %>% as.data.frame %>% filter(analysis_cat == "tumor") %>% rownames_to_column() %>% pull("rowname"),
panCK_positive_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_CK == "panCK+") %>% rownames_to_column() %>% pull("rowname"),
Ecad_positive_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_Ecad == "Ecad+") %>% rownames_to_column() %>% pull("rowname")
)

upset <- upset(fromList(tumor), order.by = "freq")
upset

non_tumor = list(
analysis_cat_nontumor = colData(sce) %>% as.data.frame %>% filter(analysis_cat == "non_tumor") %>% rownames_to_column() %>% pull("rowname"),
panCK_negative_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_CK == "panCK-") %>% rownames_to_column() %>% pull("rowname"),
Ecad_negative_GMM = colData(sce) %>% as.data.frame %>% filter(GMM_Ecad == "Ecad-") %>% rownames_to_column() %>% pull("rowname")
)

upset_1 <- upset(fromList(non_tumor), order.by = "freq")
upset_1
```

### Dimensionality reduction

```{r viz_sub, message = FALSE, warning=FALSE, fig.width=20, fig.height=20}
# ## Subsample 10% of cells from each core 
# #to save time and not get super crowded dimensionality reduction plots
# set.seed(22)
# sub <- colData(sce) %>% as.data.frame() %>% group_by("sample_id") %>% slice_sample(prop = 0.1) %>% arrange("sample_id") %>% mutate(sub_id = paste(sample_id,ObjectNumber,sep="_")) %>% pull(sub_id)
# sce_sub = sce[,sub]
# sce_sub$cluster = as.factor(sce_sub$cluster)  
# 
# #Run UMAP/TSNE
# sce_sub = runUMAP(sce_sub[rowData(sce_sub)$use_channel,], n_neighbors = 30, pca = 50,exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam((detectCores()-2)))
# sce_sub = runTSNE(sce_sub[rowData(sce_sub)$use_channel,], exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam((detectCores()-2)))

sce_sub <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_SUB_TNBC_Clus.rds")

## 1. Visualize TSNE
#panCK and GMM_CK
p1 <- plotReducedDim(sce_sub, dimred = "TSNE", by_exprs_values = "exprs",colour_by = "panCK")

p2 <- plotReducedDim(sce_sub, dimred = "TSNE", by_exprs_values = "exprs",colour_by = "Ecad")

p3 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "cluster_category")+
  scale_color_manual(values = metadata(sce)$colors$cluster_category)

p4 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$analysis_cat)

p5 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "PID")+#shape_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  guides(color = "none")

p6 <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "TMANr")+#shape_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$TMANr)


p1+p2+p3+p4+p5+p6


#Visualize all marker expression
all_plots_TSNE <- lapply(c(rownames(sce_sub)[rowData(sce_sub)$use_channel]),
                    function(x){
                      p <- plotReducedDim(sce_sub, dimred = "TSNE", 
                                          colour_by = x, 
                                          by_exprs_values = "exprs",
                                          point_size = 0.5)
                      return(p)                    
                    })

plots_TSNE_exp <- plot_grid(plotlist = all_plots_TSNE)
plots_TSNE_exp
```

```{r}
# Mark by pre-treatment information
sce_sub$pretreatment <- str_detect(sce_sub$pT, "y")
plotReducedDim(sce_sub, dimred = "TSNE",colour_by = "pretreatment")
```

### Extract representative images

```{r}
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_TNBC_spillcor.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")
#sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
```

```{r}
#Randomly select 20 images
set.seed(22)
img_random <- sample(seq_len(length(images)), 10)

img_files <- list.files("/mnt/central_nas/tnbc_volume/TNBC/img/")
img_selected <- c(which(str_detect(img_files ,"ZTMA249_ID_320_CX8Y10")),which(str_detect(img_files ,"ZTMA249_ID_289_CX1Y7")),which(str_detect(img_files ,"ZTMA249_ID_36_AX4Y5")),which(str_detect(img_files ,"ZTMA249_ID_7_AX7Y1")),which(str_detect(img_files ,"ZTMA249_ID_220_BX4Y13")),which(str_detect(img_files ,"ZTMA249_ID_271_CX7Y4")),which(str_detect(img_files ,"ZTMA249_ID_82_AX2Y11")), which(str_detect(img_files ,"ZTMA249_ID_179_BX3Y8")), which(str_detect(img_files ,"ZTMA249_ID_183_BX7Y8")), which(str_detect(img_files ,"ZTMA174_B08.26868_79_AX7Y10")))

img_ids <- c(img_selected,img_random)

# Normalize and clip image
cur_images <- images[img_ids]
#cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
#cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#Subset mask and object
cur_mask <- masks[img_ids]
cur_sce <- sce[,sce$sample_id %in% names(cur_images)]
```

We will use cytoviewer for image exploration and download

```{r}
app <- cytoviewer(image = cur_images, mask = cur_mask, object = cur_sce, img_id = "sample_id", cell_id = "ObjectNumber")

shiny::runApp(app, launch.browser = TRUE)
```


## Save objects

```{r save_objects, message = FALSE, warning=FALSE}
#Save sce
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
#Save sce_sub
saveRDS(sce_sub, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_SUB_TNBC_Clus.rds")
```

# Session Info 

```{r}
sessionInfo()
```