---
title: "06_singlecell_TME"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: sce_TME.rds 

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Script 6: TME analysis

Here, we will perform comprehensive single cell immune and stromal cell analysis.

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scran)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(wesanderson)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(Rphenoannoy)
library(vegan)
library(ggpubr)
library(survival)
library(survminer)
library(cowplot)
library(CATALYST)
library(corrplot)
```


## Read in data

```{r read-data-batch-correction, message=FALSE}
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_TNBC_Clus_com.rds")
```

## Filter sce for tumor cells

```{r filter sce, message=FALSE}
#based on cluster categories (n=525537)
sce <- sce[,colData(sce)$analysis_cat == "non_tumor"]
sce
```

# 1A. Clustering 

**NOTE**: Please be aware that some analyse including unsupervised clustering involves random processes and may not be full reproducible across systems. We did, however, set seeds for all such analysis. To ensure full reproducibility, please use the information stored in the SCE objects available on Zenodo and the figure generation scripts. 

## Define markers for clustering

```{r tumor_markers, message = FALSE, warning = FALSE}
cluster_channel = c("SMA","CD68","CD3","CD20","CD11c","CD20","CD8a","CD4","CD31_vWF","CD15", "HLADR")

other_epi_channel = c("pH3","H3K27me3","Bcl2", "c_Cas3_PARP","S6", "CD44","CAIX","Ki67", "Vimentin","Fibronectin","panCK","Ecad","CK5","CK8_18","KRT14","cMYC","HER2","ER","PR","p53","AR","EGFR","p_mTOR","CK7","GATA3","CD45")

#Add to sce
rowData(sce)$cluster_channel <- (rowData(sce)$clean_target %in% cluster_channel)
rowData(sce)$other_epi_channel <- (rowData(sce)$clean_target %in% other_epi_channel)
```

## Run unsupervised clustering

Next, we will run Rphenoannoy (Jaccard-based weights, Louvain clustering as community detection algorithm) clustering with different resolution parameters. 

```{r overcluster, message=FALSE}
set.seed(22)

mat <- t(assay(sce, "exprs")[rowData(sce)$cluster_channel,])

#k=30
out <- Rphenoannoy(mat, k = 30)

clusters <- factor(membership(out[[2]]))

sce$cluster_30 <- clusters

unique(sce$cluster_30)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME_com.rds")

gc()

#k=50
out <- Rphenoannoy(mat, k = 50)

clusters <- factor(membership(out[[2]]))

sce$cluster_50 <- clusters

unique(sce$cluster_50)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME_com.rds")

gc()

#k=70
out <- Rphenoannoy(mat, k = 70)

clusters <- factor(membership(out[[2]]))

sce$cluster_70 <- clusters

unique(sce$cluster_70)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME_com.rds")

gc()

#k=90
out <- Rphenoannoy(mat, k = 90)

clusters <- factor(membership(out[[2]]))

sce$cluster_90 <- clusters

unique(sce$cluster_90)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME_com.rds")

gc()

#k=110
out <- Rphenoannoy(mat, k = 110)

clusters <- factor(membership(out[[2]]))

sce$cluster_110 <- clusters

unique(sce$cluster_110)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME_com.rds")

gc()

#k=130
out <- Rphenoannoy(mat, k = 130)

clusters <- factor(membership(out[[2]]))

sce$cluster_130 <- clusters

unique(sce$cluster_130)

#Save SCEs after clustering
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME_com.rds")

gc()
```

## Cluster stability testing and comparison

```{r clustree, message = FALSE, warning=FALSE}
#Read sce after clustering and GMM
#sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME.rds")
#sce

#Pairwise rand index 
#Agreement between two clustering approaches - ARI > 0.5 = "good" similarity
pairwiseRand(sce$cluster_30,sce$cluster_50, mode = "index")
pairwiseRand(sce$cluster_50,sce$cluster_70, mode = "index")
pairwiseRand(sce$cluster_70,sce$cluster_90, mode = "index")
pairwiseRand(sce$cluster_90,sce$cluster_110, mode = "index")
pairwiseRand(sce$cluster_110,sce$cluster_130, mode = "index")

#Over different resolution parameters - Silhouette width [-1,1,max], Root-mean-squared difference [min]
clusters <- colnames(colData(sce))[str_detect(colnames(colData(sce)),"cluster_[:digit:]")]
df <- foreach(i = seq_along(clusters),.combine = rbind) %do% {
num_clus <- length(unique(colData(sce)[,clusters[i]]))
sil <- approxSilhouette(t(assay(sce, "exprs")), clusters=colData(sce)[,clusters[i]])
rmsd <- clusterRMSD(t(assay(sce, "exprs")), clusters=colData(sce)[,clusters[i]])
data.frame(num_clus = num_clus,
           k = as.numeric(str_split(clusters[i],"_",simplify = TRUE)[,2]), 
           sil = mean(sil$width), 
           rmsd = sum(rmsd))
}


p1 <- ggplot(df, aes(x=k, y=num_clus))+
  geom_point(aes(k, num_clus, col = num_clus))+
  geom_line(aes(k, num_clus,col=num_clus))+
  theme_classic()+
  scale_color_continuous(type = "viridis")

p2 <- ggplot(df, aes(x=k, y=rmsd))+
  geom_point(aes(k, rmsd, col = rmsd))+
  geom_line(aes(k, rmsd,col=rmsd))+
  theme_classic()+
  scale_color_continuous(type = "viridis")

p3 <- ggplot(df, aes(x=k, y=sil))+
  geom_point(aes(k, sil, col = sil))+
  geom_line(aes(k, sil,col=sil))+
  theme_classic()+
  scale_color_continuous(type = "viridis")
  
p1+p2+p3 

#Chosen method => k=130
```


# 1B. Astir model

```{r astir-expo}
#Export arsinh normalized expression matrix as csv
#Add sampleID as covariate in design matrix
expr_astir <- assay(sce,"exprs") %>% t() %>% as.data.frame() %>% mutate(sample = sce$sample_id)

#write.csv(expr_astir, file = "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_astir/expr_TNBC_TME_astir.csv",row.names = TRUE)
```

`Next: Run Astir in python.`

```{r astir-ana, message = FALSE,warning=FALSE}
# Import Astir results 
astir_type <- read_csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_astir/cell_types_TNBC_TME.csv",col_names = c("ID","cell_type"),skip = 1)

astir_sum <- astir_type %>% group_by(cell_type) %>% count()

sce$astir_TME <- astir_type$cell_type
```


# 2. Heatmap visualization

## Rich heatmap

```{r heatmap sce_1, message=FALSE,warning=FALSE}
### 1. Heatmap body ###

#Aggregrate across cells with the mean (median is also possible)
cluster_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_130,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$cluster_channel)

other_mean_sce <- aggregateAcrossCells(sce, ids = sce$cluster_130,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$other_epi_channel)

# Heatmap body colors 
col_exprs <- colorRamp2(c(0,1,2,3), c("#440154FF", "#31688EFF", "#35B779FF", "#FDE725FF"))


### 2. Heatmap annotation ###

### 2.1 Spatial features

#Add number of neighbors to sce object (saved in ColPair)
n_neighbors <- colPair(sce) %>% as.data.frame %>% group_by(from) %>% dplyr::count() %>% arrange(desc(n))

sce$n_neighbors <- n_neighbors$n[match(seq_along(colnames(sce)),n_neighbors$from)]
sce$n_neighbors <- sce$n_neighbors %>% replace_na(0)

#Select spatial features and average over clusters 
spatial <- colData(sce) %>% as.data.frame %>% select(cluster_130, area, eccentricity, n_neighbors)
spatial <- spatial %>% select(-cluster_130) %>% aggregate(by = list(cluster_130 = spatial$cluster_130), FUN = mean) %>% column_to_rownames("cluster_130")

#Create Spatial Heatmap Annotation
ha_spatial <- HeatmapAnnotation(
    area = spatial$area,
    eccentricity = spatial$eccentricity,
    n_neighbors = spatial$n_neighbors, 
    col = list(area = metadata(sce)$colors$area,
               eccentricity = metadata(sce)$colors$eccentricity,
               n_neighbors = metadata(sce)$colors$n_neighbors),
    border = TRUE,
    which = "row")


### 2.2 Immune Metacluster 

anno <- colData(cluster_mean_sce) %>% as.data.frame %>% select(cluster_130, ncells)

anno <- anno %>% mutate(metacluster = ifelse(cluster_130 %in% c("4","10"),"CD8_T_cell",
                                     ifelse(cluster_130 %in% c("14","17","9"),"CD4_T_cell",
                                            #ifelse(cluster_130 %in% c("21"),"BnT_cell",
                                                   ifelse(cluster_130 %in% c("20","21"),"B_cell",
                                                          ifelse(cluster_130 %in% c("16","2","12"),"Macrophage",
                                                                 ifelse(cluster_130 %in% c("1"),"HLA_DR",
                                                                        ifelse(cluster_130 %in% c("19","18"),"Neutrophil",
                                                                               ifelse(cluster_130 %in% c("5","7","3"),"Endothelial",
                                                                                      ifelse(cluster_130 %in% c("13","8","6","11"),"Stromal",
                                                                                             ifelse(cluster_130 %in% c("15"),"Immune_low",NA)
                                                                                             ))))))))) %>% mutate(metacluster = as.factor(metacluster))

sce$immune_metacluster <- anno$metacluster[match(sce$cluster_130, anno$cluster_130)]


### 2.3 Astir immune cell types 

## Create ASTIR annotation
# Astir cell types per cluster
astir_anno <- colData(sce) %>% as.data.frame() %>% select(cluster_130,astir_TME) %>% group_by(cluster_130) %>% table() %>% as.data.frame
astir_anno <- astir_anno %>% group_by(cluster_130) %>% mutate(fra = Freq/sum(Freq)) 
astir_anno <- astir_anno %>% select(-Freq) %>% pivot_wider(id_cols = cluster_130,names_from = astir_TME,values_from = fra) %>% column_to_rownames("cluster_130")

#Create colors for astir cell types 
astir_col <- setNames(c("#7FC97F","#E1B8A8","#FDD58C","#E8EE9B","#DA2950","#666666","white"),sort(unique(sce$astir_TME)))

metadata(sce)$colors$astir_col <- astir_col

ha_astir <- HeatmapAnnotation(astir_TME = anno_barplot(astir_anno, width = unit(20, "mm"), gp = gpar(fill = metadata(sce)$colors$astir_col)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(3,"mm"),
                            which = "row")

###2.3 Metadata features

#Number of cells per cluster 
n_cells <- colData(cluster_mean_sce) %>% as.data.frame %>% select(cluster_130, ncells)

# Proportion of grade per cluster
grade <- colData(sce) %>% as.data.frame() %>% select(cluster_130,grade) %>% group_by(cluster_130) %>% table() %>% as.data.frame
grade <- grade %>% group_by(cluster_130) %>% mutate(fra = Freq/sum(Freq)) 
grade <- grade %>% select(-Freq) %>% pivot_wider(id_cols = cluster_130,names_from = grade,values_from = fra) %>% column_to_rownames("cluster_130")

# Proportion of pT stage per cluster
pT <- colData(sce) %>% as.data.frame() %>% select(cluster_130,pT_simple) %>% group_by(cluster_130) %>% table() %>% as.data.frame
pT <- pT %>% group_by(cluster_130) %>% mutate(fra = Freq/sum(Freq)) 
pT <- pT %>% select(-Freq) %>% pivot_wider(id_cols = cluster_130,names_from = pT_simple,values_from = fra) %>% column_to_rownames("cluster_130")

# Proportion of pN stage per cluster
pN <- colData(sce) %>% as.data.frame() %>% select(cluster_130,pN_simple) %>% group_by(cluster_130) %>% table() %>% as.data.frame
pN <- pN %>% group_by(cluster_130) %>% mutate(fra = Freq/sum(Freq)) 
pN <- pN %>% select(-Freq) %>% pivot_wider(id_cols = cluster_130,names_from = pN_simple,values_from = fra) %>% column_to_rownames("cluster_130")

# Number of contributing patients
cluster_PID <- colData(sce) %>% as.data.frame() %>% select(cluster_130,PID) %>% group_by(cluster_130) %>% table() %>% as.data.frame
n_PID <- cluster_PID %>% filter(Freq>0) %>% group_by(cluster_130) %>% count(name = "n_PID") %>% column_to_rownames("cluster_130")


# Define color scheme for immune clusters/metaclusters
row_order_clusters <- row_order(Heatmap(t(assay(cluster_mean_sce, "exprs")),
        column_title = "cluster_channels",
        cluster_row_slices = TRUE,
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2"))


cluster_col <- setNames(colorRampPalette(brewer.pal(9, "Accent"))(length(unique(anno$cluster_130))), row_order_clusters)                                        
metacluster_col <- setNames(colorRampPalette(brewer.pal(9, "Accent"))(length(unique(anno$metacluster))), sort(unique(anno$metacluster)))                                          
metadata(sce)$colors$cluster_130 <- cluster_col
metadata(sce)$colors$immune_metacluster <- metacluster_col

# Metadata annotation
ha_meta <- HeatmapAnnotation(cluster_130 = anno_simple(unfactor(anno$cluster_130), width = unit(8, "mm"), pch = unfactor(anno$cluster_130),border=TRUE,gp = gpar(fill = "white")),
                            metacluster = anno_simple(unfactor(anno$metacluster),border=TRUE,col=metadata(sce)$colors$immune_metacluster),
                            n_cells = anno_barplot(n_cells$ncells, width = unit(10, "mm"),gp = gpar(fill = "#440154FF")),
                            n_PID = anno_barplot(n_PID, width = unit(10, "mm"),gp = gpar(fill = "#440154FF")),
                            grade = anno_barplot(grade, width = unit(12, "mm"),gp = gpar(fill = metadata(sce)$colors$grade)),
                            pT = anno_barplot(pT, width = unit(12, "mm"),gp = gpar(fill = metadata(sce)$colors$pT_simple)),
                            pN = anno_barplot(pN, width = unit(12, "mm"),gp = gpar(fill = metadata(sce)$colors$pN_simple)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(1,"mm"),
                            which = "row")

### 3. Plot heatmap ###

h <- Heatmap(t(assay(cluster_mean_sce, "exprs")),
        column_title = "cluster_channels",
        col = col_exprs,
        name = "mean exprs",
        row_title = NULL, 
        show_column_names =  TRUE, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2",
        row_split = anno$metacluster,
        cluster_row_slices = FALSE)+
  # Heatmap(t(assay(other_mean_sce, "exprs")),
  #       column_title = "other_channels",
  #       col = col_exprs,
  #       name= "mean exprs",
  #       show_row_names = FALSE, 
  #       show_column_names =  TRUE, 
  #       clustering_method_rows = "ward.D2",
  #       clustering_method_columns = "ward.D2")+
  ha_astir+
  ha_meta+
  ha_spatial

# Add customized legend for anno_barplot()
lgd_astir <- Legend(title = "Astir_TME", at = names(metadata(sce)$colors$astir_col), 
              legend_gp = gpar(fill = metadata(sce)$colors$astir_col))

lgd_meta <- Legend(title = "Immune_Metacluster", at = names(metadata(sce)$colors$immune_metacluster), 
              legend_gp = gpar(fill = metadata(sce)$colors$immune_metacluster))          
# Plot
draw(h,annotation_legend_list = list(lgd_astir, lgd_meta))
```

```{r save_objects, message = FALSE, warning=FALSE}
#Save sce
#saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME_com.rds")
```


# 3. TME heterogeneity analysis 

## Shannon diversity index

```{r}

PID_immune_metacluster <- colData(sce) %>% as.data.frame() %>% select(immune_metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_immune_metacluster <- PID_immune_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

shannon_df <- PID_immune_metacluster %>% pivot_wider(id_cols = "PID",names_from = "immune_metacluster",values_from = "Freq") %>% column_to_rownames("PID")

diversity(shannon_df)

## 1.For fraction-based
#Calculate shannon-diversity/equitability for each patient separately

shannon_plot <- data.frame(shannon_div = diversity(shannon_df),
                           shannon_equi = (diversity(shannon_df)/log(length(unique(sce$immune_metacluster)))),
                           PID = rownames(shannon_df))

# Split diversity into high/low
shannon_plot$shannon_div_mean <- ifelse(shannon_plot$shannon_div >= mean(shannon_plot$shannon_div), "high", "low")

sce$shannon_div <- shannon_plot$shannon_div[match(sce$PID, shannon_plot$PID)]
sce$shannon_div_mean <- shannon_plot$shannon_div_mean[match(sce$PID, shannon_plot$PID)]
sce$shannon_equi <- shannon_plot$shannon_equi[match(sce$PID, shannon_plot$PID)]

# Create col for shannon_div
div_col <- colorRamp2(c(min(shannon_plot$shannon_div),median(shannon_plot$shannon_div),max(shannon_plot$shannon_div)), c(brewer.pal(3,"Greens")[1],brewer.pal(3,"Greens")[2],brewer.pal(3,"Greens")[3]))
metadata(sce)$colors$diversity_col <- div_col
```


```{r}
#Assess survival impact 
cur_surv <- colData(sce) %>% as.data.frame() %>% 
  dplyr::select(PID, shannon_div_mean, OS_months, DFS_months, status_DFS, status_OS, pretreatment) %>% 
  #filter(pretreatment == "FALSE") %>%
  unique()

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ shannon_div_mean, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = "Set1")
```

```{r}
div_pre <- colData(sce) %>% as.data.frame() %>% 
  dplyr::select(PID, shannon_div, pretreatment) %>% 
  unique() 

ggplot(div_pre, aes(x = pretreatment, y = shannon_div))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = pretreatment), position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$pretreatment)+
  theme_classic(base_size = 12)+
  stat_compare_means()
```

# 4. Single-cell visualization

## Dimensionality reduction 

```{r dimred, message=FALSE,warning=FALSE}
## Subsample 10% of cells from each core 
#to save time and not get super crowded dimensionality reduction plots
set.seed(22)
sub <- colData(sce) %>% as.data.frame() %>% group_by("sample_id") %>% slice_sample(prop = 0.1) %>% arrange("sample_id") %>% mutate(sub_id = paste(sample_id,ObjectNumber,sep="_")) %>% pull(sub_id)
sce_sub = sce[,sub]

#Run UMAP/TSNE on cluster channels
sce_sub = runUMAP(sce_sub[rowData(sce_sub)$cluster_channel,], n_neighbors = 50, pca = 50, exprs_values = "exprs",external_neighbors=TRUE, BPPARAM = MulticoreParam((detectCores()-2)))
sce_sub = runTSNE(sce_sub[rowData(sce_sub)$cluster_channel,], exprs_values = "exprs", external_neighbors=TRUE, BPPARAM = MulticoreParam((detectCores()-2)))


## 1. Visualize UMAP and TSNE
#panCK and GMM_CK
UMAP_clus <- plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "cluster_50_renamed")+
  scale_color_manual(values = col_clusters$cluster_50_renamed)

TSNE_clus <- plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "cluster_50_renamed")+
  scale_color_manual(values = col_clusters$cluster_50_renamed)

(UMAP_clus+TSNE_clus)

#Visualize all marker expression
all_plots_UMAP <- lapply(c(rownames(sce_sub)[rowData(sce_sub)$cluster_channel]),
                    function(x){
                      p <- plotReducedDim(sce_sub, dimred = "UMAP", 
                                          colour_by = x, 
                                          by_exprs_values = "exprs",
                                          point_size = 0.5)
                      return(p)                    
                    })


exp_plots_UMAP <- plot_grid(plotlist = all_plots_UMAP)

all_plots_TSNE <- lapply(c(rownames(sce_sub)[rowData(sce_sub)$cluster_channel]),
                    function(x){
                      p <- plotReducedDim(sce_sub, dimred = "TSNE", 
                                          colour_by = x, 
                                          by_exprs_values = "exprs",
                                          point_size = 0.5)
                      return(p)                    
                    })

exp_plots_TSNE <- plot_grid(plotlist = all_plots_TSNE)

#Save subsampled sce
#saveRDS(sce_sub,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/sce_sub_TNBC_249_TME_CC_ClusterChannelClus_k.rds")
```


## Barplot visualization 

### Fractions - Visualization and correlation of cluster fractions

```{r cluster-pro, message = FALSE, warning = FALSE}
## 1. Fraction of immune_immune_metacluster/cluster per sample/PID
dittoBarPlot(sce, var = "immune_metacluster", group.by = "sample_id")+
  scale_fill_manual(values = metadata(sce)$colors$immune_metacluster)+
  theme(axis.text.x = element_blank())

d_PID <- dittoBarPlot(sce, var = "immune_metacluster", group.by = "PID")+
  scale_fill_manual(values = metadata(sce)$colors$immune_metacluster)+
  theme(axis.text.x = element_blank())

dittoBarPlot(sce, var = "cluster_130", group.by = "sample_id")+
  scale_fill_manual(values = metadata(sce)$colors$cluster_130)+
  theme(axis.text.x = element_blank())

dittoBarPlot(sce, var = "cluster_130", group.by = "PID")+
  scale_fill_manual(values = metadata(sce)$colors$cluster_130)+
  theme(axis.text.x = element_blank())

## 2. Fraction of sample/PID per immune_metacluster/cluster
dittoBarPlot(sce, var = "PID", group.by = "immune_metacluster")+
  scale_fill_manual(values = metadata(sce)$colors$PID)+
  guides(fill = "none")

dittoBarPlot(sce, var = "PID", group.by = "cluster_130")+
  scale_fill_manual(values = metadata(sce)$colors$PID)+
  guides(fill = "none")

dittoBarPlot(sce, var = "sample_id", group.by = "immune_metacluster")+
  guides(fill = "none")

dittoBarPlot(sce, var = "sample_id", group.by = "cluster_130")+
  guides(fill = "none")


## 3. Correlation of immune_metacluster/cluster fractions for each PID (2-4 samples per PID)
SID_cluster <- colData(sce) %>% as.data.frame() %>% select(cluster_130,sample_id) %>% group_by(sample_id) %>% table() %>% as.data.frame
SID_cluster <- SID_cluster %>% group_by(sample_id) %>% mutate(fra = Freq/sum(Freq))

SID_immune_metacluster <- colData(sce) %>% as.data.frame() %>% select(immune_metacluster,sample_id) %>% group_by(sample_id) %>% table() %>% as.data.frame
SID_immune_metacluster <- SID_immune_metacluster %>% group_by(sample_id) %>% mutate(fra = Freq/sum(Freq))

# Cluster fraction correlation for each PID (2 samples per PID)
## by cluster
SID_cluster_cor <- SID_cluster
SID_cluster_cor$PID <- sce$PID[match(SID_cluster$sample_id,sce$sample_id)]

SID_cluster_cor <- foreach(i = unique(colData(sce)$PID),.combine = "rbind")%do%{
SID <- SID_cluster_cor %>% filter(PID %in% i) %>% select(-Freq,-PID) %>% pivot_wider(id_cols = cluster_130,names_from = sample_id,values_from = fra) %>% column_to_rownames("cluster_130")
cor <- cor(SID,method = "pearson")
cor[2]
}

SID_cluster_cor <- as.data.frame(SID_cluster_cor) 
SID_cluster_cor$PID <- unique(colData(sce)$PID) 
SID_cluster_cor$category <- "PID"

SID_cluster_cor_plot <- ggplot(SID_cluster_cor, aes(x = category, y = V1))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  ylab("Correlation of cluster fractions per PID (n=2-4)")+
  ggtitle(paste("Spearman rho = ",round(median(SID_cluster_cor$V1, na.rm = TRUE), digits = 3)))

SID_cluster_cor_plot

## by immune_metacluster
SID_immune_metacluster_cor <- SID_immune_metacluster
SID_immune_metacluster_cor$PID <- sce$PID[match(SID_immune_metacluster$sample_id,sce$sample_id)]

SID_immune_metacluster_cor <- foreach(i = unique(colData(sce)$PID),.combine = "rbind")%do%{
SID <- SID_immune_metacluster_cor %>% filter(PID %in% i) %>% select(-Freq,-PID) %>% pivot_wider(id_cols = immune_metacluster,names_from = sample_id,values_from = fra) %>% column_to_rownames("immune_metacluster")
cor <- cor(SID,method = "spearman")
cor[2]
}

SID_immune_metacluster_cor <- as.data.frame(SID_immune_metacluster_cor) 
SID_immune_metacluster_cor$PID <- unique(colData(sce)$PID) 
SID_immune_metacluster_cor$category <- "PID"

SID_immune_metacluster_cor_plot <- ggplot(SID_immune_metacluster_cor, aes(x = category, y = V1))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  ylab("Correlation of immune_metacluster fractions per PID (n=2-4)")+
  ggtitle(paste("Spearman rho = ",round(median(SID_immune_metacluster_cor$V1, na.rm = TRUE), digits = 2)))


SID_immune_metacluster_cor_plot

d_PID + SID_immune_metacluster_cor_plot
```


## Complex Heatmap based fraction stacked barplot

```{r}
## PID immune_metacluster fraction dataframe
PID_immune_metacluster <- colData(sce) %>% as.data.frame() %>% select(immune_metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_immune_metacluster <- PID_immune_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
cur_df <- PID_immune_metacluster %>% pivot_wider(id_cols = "PID", names_from = "immune_metacluster",values_from = "fra") %>% column_to_rownames("PID")

# Perform hierarchical clustering on fractions
dd <- dist(cur_df, method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
plot(hc)
row_sorted <- hc$order

# Define annotation dataframe
cur_df <- cur_df[row_sorted,]

anno <- colData(sce) %>% as.data.frame %>% select(PID, age_bin, grade, pT_simple, pN_simple, pM_simple, pretreatment, shannon_div) %>% unique()
anno <- anno[match(rownames(cur_df), anno$PID),]

# Plot annotation dataframe with ComplexHeatmap
fraction_anno <- HeatmapAnnotation(MC_fractions = anno_barplot(cur_df, gp=gpar(fill=metadata(sce)$colors$immune_metacluster[colnames(cur_df)]),
                                   bar_width = 1,
                                   height = unit(25,"cm"),
                                   width = unit(11,"cm"),
                                   show_row_names = FALSE), 
                         # PID = anno_simple(anno$PID, col=metadata(sce)$colors$PID, border=TRUE),
                          age = anno_simple(anno$age_bin, col = metadata(sce)$colors$age_bin, border = TRUE),
                          grade = anno_simple(anno$grade, col = metadata(sce)$colors$grade, border = TRUE),
                          pT = anno_simple(anno$pT_simple, col=metadata(sce)$colors$pT_simple, border=TRUE),
                          pN = anno_simple(anno$pN_simple, col=metadata(sce)$colors$pN_simple, border=TRUE),
                          pM = anno_simple(anno$pM_simple, col=metadata(sce)$colors$pM_simple, border = TRUE),
                          pretreatment = anno_simple(as.character(anno$pretreatment), col=metadata(sce)$colors$pretreatment, border = TRUE),
                         diversity = anno_simple(anno$shannon_div, col = metadata(sce)$colors$diversity_col, border = TRUE),
                          border = TRUE, 
                          #annotation_name_rot = 90,
                          gap = unit(1,"mm"),
                          which = "row")

lgd <- Legend(col_fun = metadata(sce)$colors$diversity_col, title = "Diversity")


draw(fraction_anno + NULL, annotation_legend_list = list(lgd))
```

## CATALYST-based visualization

```{r CATALYST}
# save spe in CATALYST-compatible object with renamed colData entries and 
# new metadata information
sce_cat <- sce 
sce_cat$sample_id <- factor(sce$PID)
sce_cat$cluster_id <- factor(sce$immune_metacluster)

#add celltype information to metadata
metadata(sce_cat)$cluster_codes <- data.frame(immune_metacluster = factor(sce_cat$immune_metacluster))
```

```{r CATALYST pbmds and clrDR, message=FALSE}
# MDS pseudobulk by metacluster
pbMDS(sce_cat, by = "cluster_id", 
      features = rownames(sce_cat)[rowData(sce_cat)$cluster_channel], 
      label_by = "cluster_id", k = "immune_metacluster") +
  scale_color_manual(values = metadata(sce_cat)$colors$immune_metacluster)+
  theme_classic()

# MDS pseudobulk by metacluster
pbMDS(sce_cat, by = "both", 
      features = rownames(sce_cat)[rowData(sce_cat)$cluster_channel], 
      k = "immune_metacluster",
      size_by = TRUE) +
  scale_color_manual(values = metadata(sce_cat)$colors$immune_metacluster)+
  theme_classic()

# CLR on metacluster proportions across samples
clrDR(sce_cat, dr = "PCA", 
      by = "cluster_id", k = "immune_metacluster", 
      label_by = "cluster_id", arrow_col = "PID", 
      point_pal = metadata(sce_cat)$colors$immune_metacluster, 
      arrows = FALSE) +
  scale_color_manual(values = metadata(sce_cat)$colors$PID)+
  guides(color = "none")+
  theme_classic()

# CLR on sample proportions across metaclusters
clrDR(sce_cat, dr = "PCA", 
      by = "sample_id", point_col = "sample_id",
      k = "immune_metacluster", point_pal = metadata(sce_cat)$colors$PID)+
  scale_color_manual(values = metadata(sce_cat)$colors$immune_metacluster)+
  guides(fill = "none")+
  theme_classic()
```


# 5. Image visualization

```{r load images, message = FALSE}
#Load images and masks
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_segtest.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")
```

## Visualize metaclusters on images

```{r viz metaclus, message=FALSE, warning=FALSE}

# 1. Part 1 of immune_metacluster viz 
clus_select <- anno %>% filter(!metacluster %in% c("Endothelial","Stromal","Neutrophil")) %>% pull(metacluster) %>% unfactor()

for (i in clus_select) {
  
##Visualize clusters on images - select 10 images for each category with highest number of cells
cur_sample <- colData(sce) %>% as.data.frame %>% group_by(immune_metacluster) %>% filter(immune_metacluster == i) %>% count(sample_id) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(sample_id)

#Prepare images
cur_images <- images[cur_sample]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#Subset sce
cur_sce <- sce[,sce$immune_metacluster == i]

#Plot 
plotPixels(cur_images,
           mask = masks[cur_sample],
           object = cur_sce,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           missing_colour = "white",
           outline_by = "cluster_130",
           colour_by = c("CD4", "HLADR", "CD68","CD8a", "CD20","CD3"),
           #colour = list("SMA" = c("black", "green"),
           #              "CD3" = c("black", "yellow"),
           #              "CD31_vWF" = c("black", "magenta"),
           #              "CD20" = c("black", "cyan"),
           #              "CD68" = c("black", "red"),
           #              "cluster_130" = metadata(test_sce)$colors$cluster_130
           #              ),
           thick = FALSE,
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = paste0("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/TME_clusters_viz/TME_viz_immune_metacluster_",i,".png"), scale = 2)
           )
 
  }


# 2. Part 2 of immune_metacluster viz 

clus_select <- anno %>% filter(metacluster %in% c("Endothelial","Stromal","Neutrophil")) %>% pull(metacluster) %>% unfactor()

for (i in clus_select) {
  
##Visualize clusters on images - select 10 images for each category with highest number of cells
cur_sample <- colData(sce) %>% as.data.frame %>% group_by(immune_metacluster) %>% filter(immune_metacluster == i) %>% count(sample_id) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(sample_id)

#Prepare images
cur_images <- images[cur_sample]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#Subset sce
cur_sce <- sce[,sce$immune_metacluster == i]

#Plot 
plotPixels(cur_images,
           mask = masks[cur_sample],
           object = cur_sce,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           missing_colour = "white",
           outline_by = "cluster_130",
           colour_by = c("CD31_vWF", "SMA", "CD68","CD15", "CD20","CD3"),
           #colour = list("SMA" = c("black", "green"),
           #              "CD3" = c("black", "yellow"),
           #              "CD31_vWF" = c("black", "magenta"),
           #              "CD20" = c("black", "cyan"),
           #              "CD68" = c("black", "red"),
           #              "cluster_130" = metadata(test_sce)$colors$cluster_130
           #              ),
           thick = FALSE,
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = paste0("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/TME_clusters_viz/TME_viz_immune_metacluster_",i,".png"), scale = 2)
           )
  }
```

# 6. Survival analysis

## Cox-proportonal hazards model for metaclusters

### For DFS on metaclusters

Multivariate 

```{r cox dfs}

PID_metacluster <- colData(sce) %>% as.data.frame() %>% select(immune_metacluster,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster <- PID_metacluster %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_metacluster %>% pivot_wider(id_cols = "PID",names_from = "immune_metacluster",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -Immune_low) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+ Immune_low"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

write.csv(plot_df_prop, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/supp_tables/Cox_Multi_TMEMC.csv")


#Plot hazard ratiDFS
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0.001,100))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Immune metaclusters")+
ylab("HR for con_prop")+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop

## 2. Based on categories (mean division of cluster proportions)

Cox_df_mean <- foreach(i = Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames,.combine = "cbind")%do%{
Cox_df %>% select(i) %>% mutate(mean = ifelse(Cox_df[,i] > mean(Cox_df[,i]),"B","A")) %>% pull(mean)
}

colnames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames
rownames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% rownames
Cox_df_mean <- as.data.frame(Cox_df_mean)

identical(rownames(Cox_df), rownames(Cox_df_mean))

Cox_df_mean <- cbind(Cox_df_mean, Cox_df %>% select(status_DFS,DFS_months, status_DFS, DFS_months, age, grade, pT_simple, pN_simple, pM_simple))

#Based on proportions
covariates <- Cox_df_mean %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_mean)

summary(multi_model)

multi_res <- summary(multi_model)

plot_df_mean <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

  
#Plot hazard ratios
hr_mean <- ggplot(plot_df_mean, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Immune metaclusters")+
ylab("HR for mean")+
coord_cartesian(ylim=c(0.1,200))+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_mean

## 3. Based on categories (GMM division of cluster proportions)
Cox_df_GMM <- foreach(i = Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames,.combine = "cbind")%do%{
GMM_Cox <- Mclust(Cox_df[,i], G = 2)
as.character(GMM_Cox$classification)
}

colnames(Cox_df_GMM) <- Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames
rownames(Cox_df_GMM) <- Cox_df %>% select(-status_DFS,-DFS_months, -status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% rownames
Cox_df_GMM <- as.data.frame(Cox_df_GMM,stringsAsFactors=TRUE)

identical(rownames(Cox_df), rownames(Cox_df_GMM))

Cox_df_GMM <- cbind(Cox_df_GMM, Cox_df %>% select(status_DFS,DFS_months, status_DFS, DFS_months, age, grade, pT_simple, pN_simple, pM_simple))

#Based on proportions
covariates <- Cox_df_GMM %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_GMM)

summary(multi_model)

summary(Cox_df_GMM)

multi_res <- summary(multi_model)

plot_df_GMM <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratiDFS
hr_GMM <- ggplot(plot_df_GMM, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Immune metaclusters")+
ylab("HR for GMM")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_GMM

hr_prop/
hr_mean/
hr_GMM
```

## Save objects

```{r save_objects, message = FALSE, warning=FALSE}
#Save sce
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME_new.rds")
```

# Session Info 

```{r}
sessionInfo()
```
