---
title: "07_combined_tumor_TME"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: sce_ALL.rds, sce_ALL_sub.rds, spicyR_spatialgroup.rds, INT_MC8.rds, INT_fischer_MC8.rds 

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Script 7: Combined Tumor and TME analysis

Here, we will perform combined analysis of tumor and TME phenotypes including spatial analysis. 

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scran)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(glmnet)
library(broom) 
library(lisaClust)
library(spicyR)
library(survival)
library(survminer)
library(ggnewscale)
library(scales)
library(ggpubr)
library(wesanderson)
library(cytoviewer)
library(edgeR)
```

# 1. Data integration

## Load data

```{r load data, message = FALSE, warning = FALSE}
#tumor sce
sce_tumor <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor_new.rds") 
sce_tumor

#TME sce
sce_TME <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/06_sce_TNBC_TME_new.rds") 
sce_TME

#sce with all cells
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_TNBC_Clus.rds")
sce
```

## Combine tumor and TME cluster/metacluster/color information

```{r combine tumor TME, message=FALSE,warning=FALSE}
## 1. Cluster 
tumor_cluster <- colData(sce_tumor) %>% as.data.frame %>% mutate(cluster_combined = paste0("tumor_",sce_tumor$cluster_110)) %>% select(cell_id,cluster_combined)

TME_cluster <- colData(sce_TME) %>% as.data.frame %>% mutate(cluster_combined = paste0("TME_",sce_TME$cluster_130)) %>% select(cell_id,cluster_combined)

tumor_TME_cluster <- rbind(tumor_cluster, TME_cluster)

### Add to sce
sce$cluster_combined <- tumor_TME_cluster$cluster_combined[match(sce$cell_id, tumor_TME_cluster$cell_id)]
sce$cluster_combined <- factor(sce$cluster_combined)


## 2. Metacluster 
tumor_metacluster <- colData(sce_tumor) %>% as.data.frame %>% mutate(metacluster_combined = paste0("tumor_",sce_tumor$metacluster)) %>% select(cell_id,metacluster_combined)

TME_metacluster <- colData(sce_TME) %>% as.data.frame %>% mutate(metacluster_combined = sce_TME$immune_metacluster) %>% select(cell_id,metacluster_combined)

tumor_TME_metacluster <- rbind(tumor_metacluster, TME_metacluster)

### Add to sce
sce$metacluster_combined <- tumor_TME_metacluster$metacluster_combined[match(sce$cell_id, tumor_TME_metacluster$cell_id)]
sce$metacluster_combined <- factor(sce$metacluster_combined)


## 3. Color mapping
names(metadata(sce_tumor)$colors$cluster_110) <- paste0("tumor_", names(metadata(sce_tumor)$colors$cluster_110))
names(metadata(sce_TME)$colors$cluster_130) <- paste0("TME_", names(metadata(sce_TME)$colors$cluster_130))

names(metadata(sce_tumor)$colors$metacluster) <- paste0("tumor_", names(metadata(sce_tumor)$colors$metacluster))

#Add to sce 
metadata(sce)$colors$cluster_combined <- c(metadata(sce_tumor)$colors$cluster_110, metadata(sce_TME)$colors$cluster_130)
metadata(sce)$colors$metacluster_combined <- c(metadata(sce_tumor)$colors$metacluster, metadata(sce_TME)$colors$immune_metacluster)


## 4. Cluster channels 
rowData(sce)$TME_cluster_channel <- rowData(sce_TME)$cluster_channel
rowData(sce)$tumor_cluster_channel <- rowData(sce_tumor)$cluster_channel
```

```{r save_objects, message = FALSE, warning=FALSE}
#Save sce
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
```

## Dimensionality reduction (TSNE)

```{r combined tSNE}
#Read subsampled sce from pipeline 1
#sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
sce_sub <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/04_sce_SUB_TNBC_Clus.rds")

#Add cluster_combined info
sce_sub$cluster_combined <- sce$cluster_combined[match(sce_sub$cell_id, sce$cell_id)]
sce_sub$metacluster_combined <- sce$metacluster_combined[match(sce_sub$cell_id, sce$cell_id)]
sce_sub$pretreatment <- sce$pretreatment[match(sce_sub$cell_id, sce$cell_id)]

#sub cluster categories
sce_sub$metacluster_tumor_combined <- sce$metacluster_tumor_combined[match(sce_sub$cell_id, sce$cell_id)]
sce_sub$metacluster_TME_combined <- unfactor(sce_sub$metacluster_combined)
sce_sub$metacluster_TME_combined[sce_sub$analysis_cat == "non_tumor"] <- "TME"

tumor_color <- c("grey")
names(tumor_color) <- "tumor"
TME_color <- c("grey")
names(TME_color) <- "TME"

color_TME <- c(metadata(sce)$colors$metacluster_combined, tumor_color, TME_color)
color_TME

#Plot TSNE
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "PID", point_size = 1)+#shape_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$PID)+
  guides(color = "none")

plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "TMANr", point_size = 1)+#shape_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$TMANr)

plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "pretreatment", point_size = 1)+#shape_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$pretreatment)

plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "analysis_cat", point_size = 1)+#shape_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$analysis_cat)

plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "metacluster_tumor_combined", point_size = 1)+#shape_by = "analysis_cat")+
  scale_color_manual(values = color_TME, limits = force)

plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "metacluster_TME_combined", point_size = 1)+#shape_by = "analysis_cat")+
  scale_color_manual(values = color_TME, limits = force)
  
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "metacluster_TME_combined", point_size = 1)+#shape_by = "analysis_cat")+
  scale_color_manual(values = color_TME, limits = force)

plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "panCK", by_exprs_values = "exprs", point_size = 1)
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "Ecad", by_exprs_values = "exprs", point_size = 1)

```

# 2. Combined analysis

## Combined survival analysis 

### Lasso Cox model for all phenotypes 

Multivariate 

```{r}
PID_metacluster_combined <- colData(sce) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster_combined <- PID_metacluster_combined %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_metacluster_combined %>% pivot_wider(id_cols = "PID",names_from = "metacluster_combined",values_from = "fra") %>% column_to_rownames("PID")


### Lasso regularization
### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)]) 

## Run lasso regularized coxph model
Cox_df <- Cox_df[complete.cases(Cox_df),] #only have complete rows as input

# One-hot encoding of clinical variables
one_hot <- as.data.frame(model.matrix(~ . - 1, data = Cox_df %>% select(grade,pT_simple, pM_simple, pN_simple)))
Cox_df <- cbind(Cox_df %>% select(-grade,-pT_simple, -pM_simple, -pN_simple),one_hot)

Cox_df <- Cox_df %>% filter(DFS_months != 0)

x <-  data.matrix(Cox_df %>% select(-status_DFS,-DFS_months)) #predictor variables

y <- Surv(time = Cox_df$DFS_months, event = Cox_df$status_DFS) #response variables

## Run lasso (This performs 50-fold cross-validation for the lambda value)
require(doMC)
registerDoMC(cores = 20)
set.seed(220195)
cv.fit <- cv.glmnet(x, y, family="cox", type.measure = "C", nfolds = 50, parallel = TRUE)
plot(cv.fit)

fit <- glmnet(x, y, family="cox")

coef(fit, s = cv.fit$lambda.min)

# Extract output
lasso_df <- as.data.frame(as.matrix(coef(fit, s = cv.fit$lambda.min))) %>% rownames_to_column("predictor")
colnames(lasso_df) <- c("predictor","coef")

 #Visualize hazard ratios
ggplot(lasso_df, aes(x = reorder(predictor, coef), y = exp(coef))) +
  geom_point(size = 4, aes(color = predictor)) +
  geom_hline(yintercept = 1,color = "black")+
  scale_y_continuous(trans="log")+
  ylab("Lasso HR")+
  ggtitle("DFS")+
  theme_classic()+
  scale_color_manual(values = metadata(sce)$colors$metacluster_combined)

lasso_plot <- ggplot(lasso_df, aes(x = reorder(predictor,-exp(coef)), y = exp(coef))) +
  geom_point(size = 4, aes(color = exp(coef)>1)) +
  geom_hline(yintercept = 1,color = "black")+
  ylab("Lasso HR")+
  xlab("Predictors")+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90))+
  scale_y_continuous(trans="log10")+
  ggtitle("DFS")+
  scale_color_brewer(palette = "Set1")
lasso_plot
```

# 3. Spatial analysis

```{r load combined sce}
##Load sce
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
```

# 3.1 Spatial Graph 

```{r}
# Build spatial knn graph
sce <- buildSpatialGraph(sce, img_id = "sample_id", type = "knn", k = 20, 
                         coords = c("Pos_X",  "Pos_Y"),
                         #max_dist = 30,
                         BPPARAM = MulticoreParam())

colPairNames(sce) #now includes knn_interaction_graph
```

# 3.2 Patch detection

## All phenotypes 

```{r patch-milieu}
# Patch detection for all phenotypes

for(i in unfactor(unique(sce$metacluster_combined))){ 
  sce <- patchDetection(sce,
               patch_cells = sce$metacluster_combined == i,
               colPairName = "knn_interaction_graph",
               min_patch_size = 10,
               name = paste0("patch_",i),
               expand_by = 0, 
               img_id = "sample_id",
               BPPARAM = MulticoreParam(workers = 20))
}
```

### MC8 patches

```{r}
## Patch detection
sce <- patchDetection(sce,
               patch_cells = sce$metacluster_combined == "tumor_8",
               colPairName = "knn_interaction_graph",
               min_patch_size = 10,
               name = "patch_MC8",
               expand_by = 0, 
               img_id = "sample_id", 
               BPPARAM = MulticoreParam(workers = 10))
gc()
```

# 3.3 MC8 / CK5/7+ focus

## Metacluster 8 

### Heatmap for PID with patches 

```{r}
patch_MC8_PID <- colData(sce) %>% as.data.frame() %>% filter(!is.na(patch_tumor_8)) %>% pull(PID) %>% unique()
cells_MC8_PID <- colData(sce) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame

sce$patch_MC8_classification <- ifelse(sce$PID %in% patch_MC8_PID,"patch_MC8","no_patch_MC8")

PID_patch <- colData(sce) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_patch <- PID_patch %>% group_by(PID) %>% filter(PID %in% patch_MC8_PID) %>% filter(str_detect(metacluster_combined,"tumor")) %>% mutate(fra = Freq/sum(Freq)) 

cur_df <- PID_patch %>% pivot_wider(id_cols = "PID", names_from = "metacluster_combined",values_from = "fra") %>% column_to_rownames("PID")

# Perform hierarchical clustering on fractions
dd <- dist(cur_df, method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
plot(hc)
row_sorted <- hc$order

cur_df <- cur_df[row_sorted,]

# Define annotation dataframe
anno <- colData(sce) %>% as.data.frame %>% select(PID, age_bin, grade, pT_simple, pN_simple, pM_simple) %>% unique()
anno <- anno[match(rownames(cur_df), anno$PID),]


# Plot annotation dataframe with ComplexHeatmap
fraction_anno <- HeatmapAnnotation(MC_fractions = anno_barplot(cur_df, gp=gpar(fill=metadata(sce)$colors$metacluster[colnames(cur_df)]),
                                   bar_width = 1,
                                   height = unit(25,"cm"),
                                   width = unit(11,"cm"),
                                   show_row_names = FALSE), 
                          PID = anno_simple(anno$PID, col=metadata(sce)$colors$PID, border=TRUE),
                          age = anno_simple(anno$age_bin, col = metadata(sce)$colors$age_bin, border = TRUE),
                          grade = anno_simple(anno$grade, col = metadata(sce)$colors$grade, border = TRUE),
                          pT = anno_simple(anno$pT_simple, col=metadata(sce)$colors$pT_simple, border=TRUE),
                          pN = anno_simple(anno$pN_simple, col=metadata(sce)$colors$pN_simple, border=TRUE),
                          pM = anno_simple(anno$pM_simple, col=metadata(sce)$colors$pM_simple, border = TRUE),
                          border = TRUE, 
                          #annotation_name_rot = 90,
                          gap = unit(1,"mm"),
                          which = "row")

fraction_anno + NULL
```

### Most abundant phenotype 

```{r}
# most abundant phenotype in patients with MC8 patches 
PID_patch_abundant <- PID_patch %>% #filter(metacluster_combined != "tumor_8") %>% 
  mutate(max = max(fra)) %>% filter(fra == max) %>% ungroup() %>% count(metacluster_combined) %>% mutate(fra = n/sum(n), category = "cat") %>% arrange(fra)

PID_patch_abundant$metacluster_combined <- factor(PID_patch_abundant$metacluster_combined, levels = unfactor(PID_patch_abundant$metacluster_combined))

ggplot(PID_patch_abundant) +
  geom_tile(color = "black",aes(category, n, fill = metacluster_combined), position = "stack")+
  scale_fill_manual(values=metadata(sce)$colors$metacluster_combined) + 
  theme_classic(base_size = 15)+
  ylab("Density")+
  xlab("PID")+
  theme(axis.text.x = element_blank())
```

### Overall frequency plot

```{r}
PID_dom_MC8 <- PID_patch %>% mutate(max = max(fra)) %>% filter(fra == max) %>% ungroup() %>% filter(metacluster_combined == "tumor_8") %>% pull(PID)

PID_overall <- colData(sce) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_MC8_overall <- PID_overall %>% group_by(PID) %>% filter(metacluster_combined == "tumor_8" & Freq > 5) %>% mutate(patch = ifelse(PID %in% patch_MC8_PID, TRUE, FALSE)) %>% mutate(dominated = ifelse(PID %in% PID_dom_MC8, TRUE, FALSE)) %>% mutate(PID_category = ifelse(patch && dominated, "dominated", ifelse(patch, "patch", "cells")))                                                                                    
PID_MC8_overall_plot <- PID_MC8_overall %>% group_by(PID_category) %>% count() %>% ungroup() %>% mutate(fra = n/sum(n)) %>% mutate(category = "MC8")


PID_MC8_overall_plot$PID_category <- factor(PID_MC8_overall_plot$PID_category, levels = c("cells","patch","dominated"))

ggplot(PID_MC8_overall_plot) +
  geom_tile(color = "black",aes(category, n, fill = PID_category), position = "stack")+
  scale_fill_brewer(palette = "Blues")+
  theme_classic(base_size = 15)+
  ylab("Density")+
  theme(axis.title.x = element_blank())
```

### Interaction testing 

```{r}
sce_tumor <- sce[,sce$analysis_cat == "tumor"]
sce_tumor$metacluster_combined <- unfactor(sce_tumor$metacluster_combined)

# For PID with at least 5 MC8 cells 
PID_overall <- colData(sce_tumor) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_MC8_overall <- PID_overall %>% group_by(PID) %>% filter(metacluster_combined == "tumor_8" & Freq > 5)

sce_tumor_MC8 <- sce_tumor[,sce_tumor$PID %in% PID_MC8_overall$PID]

interaction_tumor_MC8 <- testInteractions(sce_tumor_MC8, 
                 group_by = "sample_id", 
                 label = "metacluster_combined", 
                 colPairName = "neighborhood", 
                 BPPARAM = MulticoreParam())

# Full tumor compartments
interaction_tumor <- testInteractions(sce_tumor, 
                  group_by = "sample_id", 
                  label = "metacluster_combined", 
                  colPairName = "neighborhood", 
                  BPPARAM = MulticoreParam())
```

Visualization 

```{r}
interaction_tumor_MC8 %>% as_tibble() %>% filter(from_label %in% c("tumor_8")) %>% filter(to_label != "tumor_8") %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE)) %>%
    ggplot() +
        geom_tile(aes(from_label, to_label, fill = sum_sigval)) +
        scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_classic()

interaction_tumor_plot <- interaction_tumor_MC8 %>% as_tibble() %>% filter(from_label %in% c("tumor_8")) %>% filter(to_label != "tumor_8") %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

ggplot(interaction_tumor_plot) +
        geom_tile(aes(reorder(to_label,sum_sigval), from_label, fill = sum_sigval)) +
        scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_classic()
```

### Survival analysis for MC8 patches 

```{r}
# Survival estimations for DFS when patches of MC8 cells are present 

#Assess survival impact 
cur_surv <- colData(sce) %>% as.data.frame() %>% select(PID, patch_tumor_8_binary, patch_tumor_3_binary, patch_tumor_3or8_binary, OS_months, DFS_months, status_DFS, status_OS) %>% unique() 

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ patch_tumor_8_binary, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("#377EB8", "#E41A1C")
)

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ patch_tumor_3or8_binary, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("#377EB8", "#E41A1C")
)
dev.off()

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ patch_tumor_3_binary, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("#377EB8", "#E41A1C")
          )
```

Univariate Cox model 

```{r}
PID_patch_MC8 <- colData(sce) %>% as.data.frame() %>% select(patch_MC8_classification,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_patch_MC8 <- PID_patch_MC8 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_patch_MC8 %>% pivot_wider(id_cols = "PID",names_from = "patch_MC8_classification",values_from = "fra") %>% column_to_rownames("PID")
#colnames(PID_metagroups) <- paste0("CK57_",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -no_patch_MC8) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~", paste(covariates, collapse = " + "),"+no_patch_MC8"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratiDFS
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_manual(values = c("#377EB8"))+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0,5))+
#scale_y_continuous(trans = "log10")
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

Multivariate 

```{r}
PID_patch_MC8 <- colData(sce) %>% as.data.frame() %>% select(patch_MC8_classification,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_patch_MC8 <- PID_patch_MC8 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_patch_MC8 %>% pivot_wider(id_cols = "PID",names_from = "patch_MC8_classification",values_from = "fra") %>% column_to_rownames("PID")
#colnames(PID_metagroups) <- paste0("CK57_",colnames(PID_metagroups))

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -no_patch_MC8) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+no_patch_MC8"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratiDFS
hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0.1,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

#### All tumor patches

```{r}
for(x in c("patch_tumor_9", "patch_tumor_7", "patch_tumor_6", "patch_tumor_11", 
"patch_tumor_3", "patch_tumor_2", "patch_tumor_1", "patch_tumor_8", 
"patch_tumor_5", "patch_tumor_10", "patch_tumor_4")){
patch_PID <- unique(sce$PID[!is.na(colData(sce)[,x])])
colData(sce)[paste0(x, "_binary")] <- ifelse(sce$PID %in% patch_PID,1,0)
}
```

```{r}
# Tumor patch 3 or patch 8
patch_combined <- colData(sce) %>% as.data.frame() %>% select(patch_tumor_3_binary, patch_tumor_8_binary, PID) %>% unique() %>% as.data.frame %>% remove_rownames() %>% column_to_rownames("PID") %>% mutate(sum = rowSums(.)) %>% mutate(patch_tumor_3or8_binary = ifelse(sum > 0, 1, 0)) 

sce$patch_tumor_3or8_binary <- patch_combined$patch_tumor_3or8_binary[match(sce$PID, rownames(patch_combined))]
```

```{r}
all_patches <- paste0(c("patch_tumor_3or8"), "_binary")

Cox_df <- colData(sce) %>% as.data.frame() %>% select(all_of(all_patches),PID) %>% unique() %>% as.data.frame %>% remove_rownames() %>% column_to_rownames("PID")

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratiDFS
hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0.1,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
all_patches <- paste0(c("patch_tumor_9", "patch_tumor_7", "patch_tumor_6", "patch_tumor_11", 
"patch_tumor_3", "patch_tumor_2", "patch_tumor_1", "patch_tumor_8", 
"patch_tumor_5", "patch_tumor_10", "patch_tumor_4"), "_binary")

Cox_df <- colData(sce) %>% as.data.frame() %>% select(all_of(all_patches),PID) %>% unique() %>% as.data.frame %>% remove_rownames() %>% column_to_rownames("PID")

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratiDFS
hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0.1,10))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

# 3.4 Immune scores 

```{r}
##Load sce
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
```

```{r}
## Combine tumor clusters 
sce$metacluster_tumor_combined <- ifelse(str_detect(unfactor(sce$metacluster_combined),"tumor"),"tumor",unfactor(sce$metacluster_combined))

sce <- aggregateNeighbors(sce, colPairName = "neighborhood", aggregate_by = "metadata", count_by = "metacluster_tumor_combined", proportions = FALSE, name = "count_neighbors")
```

## Spatial CD8 Phenotype

```{r}
cur_dat <- sce$count_neighbors %>% as.data.frame
rownames(cur_dat) <- sce$cell_id
cur_dat$PID <- sce$PID
cur_dat$sample_id <- sce$sample_id
cur_dat$metacluster_tumor_combined <- sce$metacluster_tumor_combined

## Step 1: Define CD8+ T cells as infiltrating, bystanding, ignored - SPATIAL PATTERN
cur_dat <- cur_dat %>% filter(metacluster_tumor_combined == "CD8_T_cell") %>% 
  mutate(CD8Tcell_type = ifelse(tumor >= 3, "infiltrating", 
                                ifelse(tumor < 3 & tumor > 0, "bystanding", "ignored")))

## Step 2: Classify Patients based on spatial CD8 T cell scores - SPATIAL GROUP
out <- cur_dat %>% group_by(PID, CD8Tcell_type) %>% dplyr::count() %>% 
  pivot_wider(id_cols = PID, values_from = n, names_from = CD8Tcell_type) %>% 
  mutate(ratio_inf_ex = bystanding / infiltrating, number_inf_ex = bystanding + infiltrating) %>% 
  mutate(spatial_CD8_phenotype = ifelse(ratio_inf_ex >= 10 | infiltrating <= 20, "excluded", "inflamed")) %>% 
  mutate(spatial_CD8_phenotype = ifelse(number_inf_ex <= 10 | is.na(number_inf_ex), "cold", spatial_CD8_phenotype))

# Inflamed: ratio_inf_ex < 8 or > 20 inf CD8
# Excluded: ratio_inf_ex >= 8 
# Cold: number_inf_ex <= 10? 

# Add to sce
sce$spatial_CD8_pheno <- out$spatial_CD8_phenotype[match(sce$PID, out$PID)]
sce[,is.na(sce$spatial_CD8_pheno)]$spatial_CD8_pheno <- "cold"

spatial_CD8_pheno_col <- setNames(c("firebrick", "navy","steelblue1"), unique(sce$spatial_CD8_pheno)) 
metadata(sce)$colors$spatial_CD8_pheno <- spatial_CD8_pheno_col
```

Visualization 

```{r}
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_TNBC_new.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")

# Select images of interest
cur_sample <- (colData(sce) %>% as.data.frame %>% select(sample_id, spatial_CD8_pheno) %>% group_by(sample_id) %>% filter(spatial_CD8_pheno == "inflamed") %>% count() %>% arrange(desc(n)) %>% pull(sample_id))[1:10]

cur_images <- images[mcols(images)$sample_id %in% cur_sample]
cur_masks <- masks[mcols(masks)$sample_id %in% cur_sample]
cur_sce <- sce[,sce$sample_id %in% cur_sample & sce$metacluster_tumor_combined %in% c("tumor", "CD8_T_cell")]

cur_sce$CD8Tcell_tumor_type <- ifelse(is.na(cur_sce$CD8Tcell_type), "tumor", cur_sce$CD8Tcell_type)
```

We will use cytoviewer for image exploration and download

```{r}
app <- cytoviewer(image = cur_images, mask = cur_masks, object = cur_sce, img_id = "sample_id", cell_id = "ObjectNumber")

shiny::runApp(app)
```


```{r}
plotSpatial(
  object = sce[, sce$sample_id == "ZTMA174.2_ZTMA174_B06.22442_10_AX2Y2_93" & sce$metacluster_tumor_combined %in% c("tumor", "CD8_T_cell")], 
  img_id = "sample_id", 
  node_color_by = "CD8Tcell_type", 
  node_shape_by = "metacluster_tumor_combined", 
  colPairName = "neighborhood", 
  draw_edges = TRUE, 
  nodes_first = FALSE, 
  directed = FALSE)+
  scale_color_manual(values = c("navy", "steelblue1", "firebrick"))
```

### Kaplan-Meier 

```{r}
#Assess survival impact 
## Spatial_CD8_pheno
cur_surv <- colData(sce) %>% as.data.frame() %>% select(PID, spatial_CD8_pheno, OS_months, DFS_months, status_DFS, status_OS) %>% unique()

#Plot survival curves
fit <- survfit(Surv(OS_months, status_OS) ~ spatial_CD8_pheno, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("navy","steelblue1","firebrick"))

fit <- survfit(Surv(DFS_months, status_DFS) ~ spatial_CD8_pheno, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("navy","steelblue1","firebrick"))#,


## Spatial_CD8_pheno_merg
cur_surv <- colData(sce) %>% as.data.frame() %>% select(PID, spatial_CD8_pheno_merg, OS_months, DFS_months, status_DFS, status_OS) %>% unique()

#Plot survival curves
fit <- survfit(Surv(OS_months, status_OS) ~ spatial_CD8_pheno_merg, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("steelblue1","firebrick"))#,

fit <- survfit(Surv(DFS_months, status_DFS) ~ spatial_CD8_pheno_merg, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("steelblue1","firebrick"))#,

```

#### Number of T cells comparison 

```{r}
# Compare average number of CD8+ T cells between spatial CD8 T cell groups
out <- colData(sce) %>% as.data.frame %>% select(PID, sample_id, spatial_CD8_pheno, metacluster_tumor_combined) %>% filter(metacluster_tumor_combined == "CD8_T_cell") %>% group_by(sample_id, spatial_CD8_pheno, PID) %>% dplyr::count() %>% group_by(PID, spatial_CD8_pheno) %>% mutate(n = mean(n)) %>% select(PID, spatial_CD8_pheno, n) %>% unique()

out$CD8_category <- ifelse(out$n <= quantile(out$n, probs = seq(0,1,0.33))[2], "low", 
                           ifelse(out$n >= quantile(out$n, probs = seq(0,1,0.33))[3], "high", "medium"))

out_viz <- out %>% group_by(CD8_category, spatial_CD8_pheno) %>% dplyr::count() %>% ungroup() %>% group_by(CD8_category) %>% mutate(fra = n/sum(n))

out_viz$CD8_category <- factor(out_viz$CD8_category, levels = c("high","medium","low"))

ggplot(out_viz, aes(x = CD8_category, y = fra, fill = spatial_CD8_pheno))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = metadata(sce)$colors$spatial_CD8_pheno)

sce$CD8_number_category <- out$CD8_category[match(sce$PID, out$PID)]
sce$CD8_number_category_merg <- ifelse(sce$CD8_number_category == "high", "high","medium/low")
```

```{r}
## Number group
cur_surv <- colData(sce) %>% as.data.frame() %>% select(PID, CD8_number_category, CD8_number_category_merg, OS_months, DFS_months, status_DFS, status_OS) %>% unique()

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ CD8_number_category, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(),
          palette = "Set1")#, # Change ggplot2 theme

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ CD8_number_category_merg, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), 
          palette = "Set1")#, # Change ggplot2 theme

```

Multivariate 

```{r}
PID_CD8 <- colData(sce) %>% as.data.frame() %>% select(CD8_number_category,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_CD8 <- PID_CD8 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_CD8 %>% pivot_wider(id_cols = "PID",names_from = "CD8_number_category",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -high) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+high"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratiDFS
#hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0.1,100))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))
```

### Cox model

Spatial CD8 phenotypes

Multivariate 

```{r}
PID_CD8 <- colData(sce) %>% as.data.frame() %>% select(spatial_CD8_pheno,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_CD8 <- PID_CD8 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_CD8 %>% pivot_wider(id_cols = "PID",names_from = "spatial_CD8_pheno",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -inflamed) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+ inflamed"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

write.csv(plot_df_prop, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/supp_tables/Cox_Multi_SpatialGroup.csv")

#Plot hazard ratiDFS
#hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
coord_cartesian(ylim=c(0.1,63))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

#hr_prop
```

Univariate 

```{r}
PID_CD8 <- colData(sce) %>% as.data.frame() %>% select(spatial_CD8_pheno,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_CD8 <- PID_CD8 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_CD8 %>% pivot_wider(id_cols = "PID",names_from = "spatial_CD8_pheno",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -inflamed) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~", paste(covariates, collapse = " + "),"+ inflamed"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

write.csv(plot_df_prop, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/supp_tables/Cox_Uni_SpatialGroup.csv")

#Plot hazard ratiDFS
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_manual(values = c("#377EB8"))+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
#coord_cartesian(ylim=c(0,8))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

### Differential abundance testing

```{r}
### Differential abundance analysis based on edgeR
sce$spatial_CD8_pheno_merg <- colData(sce) %>% 
  as.data.frame %>% 
  select(spatial_CD8_pheno) %>% 
  mutate(spatial_CD8_pheno_merg = ifelse(spatial_CD8_pheno %in% c("cold","excluded"),"cold/excluded", spatial_CD8_pheno)) %>% pull(spatial_CD8_pheno_merg)

#test_sce <- sce[,sce$spatial_CD8_pheno %in% c("cold","inflamed")]

abundances <- table(sce$metacluster_combined, sce$PID)
abundances <- unclass(abundances) 
head(abundances)

#relevant patient metadata (Relevel so that pT 1/2 is baseline)
meta <- colData(sce)[match(colnames(abundances), sce$PID),]
meta$spatial_CD8_pheno_merg <- relevel(x=factor(meta$spatial_CD8_pheno_merg), ref = "cold/excluded")

#filter
meta <- meta[!is.na(meta$spatial_CD8_pheno_merg),]
abundances <- abundances[,colnames(abundances) %in% meta$PID]

#create DGEList - based on edgeR package
y.ab <- DGEList(abundances, samples=meta, group = meta$spatial_CD8_pheno_merg)

#Design matrix
design <- model.matrix(~spatial_CD8_pheno_merg, data = y.ab$samples)

#Visualize
# logCPM of abundances per cluster
cur_mat <- cpm(abundances,prior.count = 1,log = TRUE)

Heatmap(cur_mat, 
        col = viridis(100),
        column_split = unfactor(meta$spatial_CD8_pheno_merg))%v%
  columnAnnotation(spatial_CD8_pheno_merg_merg = anno_simple(unfactor(meta$spatial_CD8_pheno_merg)))


# Estimate NB and QL dispersion 
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
plotBCV(y.ab, cex=1)

fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)
plotQLDisp(fit.ab, cex=1)

# Test for differential abundance of clusters between tumor stages
res <- glmQLFTest(fit.ab, coef=ncol(design))

res$coefficients
summary(decideTests(res))

DA <- res$table

DA <- topTags(res, n = nrow(DA), adjust.method = "BH")$table

DA <- DA %>% mutate(cluster = rownames(DA), sign = DA$FDR<0.1, result = ifelse(DA$logFC > 0 & DA$FDR < 0.1, "UP",ifelse(DA$logFC < 0 & DA$FDR < 0.1, "DOWN", "NOTSIG"))) %>% arrange(desc(logFC)) %>% mutate(category = ifelse(str_detect(cluster,"tumor"),"tumor","non_tumor"))

DA$cluster <- factor(DA$cluster, levels = DA$cluster)

sig <- DA %>% filter(FDR<0.1) %>% pull(cluster) %>% unfactor

## Visualize
#Barplot
ggplot(DA, aes(cluster, y = logFC))+
  geom_col(aes(fill=result), col = "black")+
  scale_fill_manual(values = c("#377EB8","lightgrey","#E41A1C"))+
  ylab("logFC")+
  xlab("Cluster")+
  labs(fill = "FDR<0.1")+
  ggtitle("Spatial CD8 phenotype (cold vs inflamed)")+
  theme_classic()

#Volcano plot  
ggplot(DA, aes(x = logFC, y = -log10(FDR)))+
  geom_point(aes(col=cluster, size = -log10(FDR)))+
  scale_color_manual(values = metadata(sce)$colors$metacluster_combined)+
  xlab("logFC (cold vs inflamed)")+
  ylab("-log10(FDR)")+
  theme_classic()+
  geom_hline(yintercept = -log10(0.1), linetype = "dashed")+
  geom_vline(xintercept = 0, linetype = "dashed")+
  labs(color = "cluster")

#Abundance vs log fold change (best?)
ggplot(DA, aes(x = logFC, y = logCPM))+
  geom_point(aes(col= cluster, 
                 size = -log10(FDR), shape = category))+
  geom_label_repel(data = DA %>% filter(FDR<0.1), aes(label = cluster))+
  scale_color_manual(values = metadata(sce)$colors$metacluster_combined)+
  xlab("logFC (cold/excluded vs inflamed)")+
  ylab("Abundance (logCPM)")+
  theme_classic()+
  geom_vline(xintercept = 0, linetype = "dashed")+
  labs(color = "cluster")

ggplot(DA, aes(x = cluster, y = category, col = logFC, size = -log10(FDR))) +
  geom_point() +
  theme_classic() +
  scale_color_gradient2(high = "#E41A1C", mid = "white", low = "#377EB8") +
  scale_size(range = c(2, 10)) +
  new_scale_colour() +
  geom_point(data = DA, shape = 1, aes(size = -log10(FDR), col = result), stroke = 1.5) +
  scale_color_manual(values = c("UP" = "#E41A1C", "NOTSIG" = "snow2", "DOWN" = "#377EB8"), name = "Direction") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### Differential interaction testing

```{r}
### First comparison: CD8-T-cell interaction with tumor (phenotypes combined) between "cold/excluded" vs. "inflamed"

# Generate segmentedCells object 
cur_sce <- sce[, sce$metacluster_tumor_combined %in% c("CD8_T_cell","tumor")]

cells <- data.frame(row.names = colnames(cur_sce))
cells$x <- cur_sce$Pos_X
cells$y <- cur_sce$Pos_Y
cells$imageID <- cur_sce$sample_id
cells$imageCellID <- cur_sce$ObjectNumber
cells$cellType <- cur_sce$metacluster_tumor_combined

cells_sc <- SegmentedCells(cells, cellProfiler = TRUE)

# Add condition and patient id
spatial_CD8_factor <- relevel(x=factor(cur_sce$spatial_CD8_pheno_merg), ref = "cold/excluded")

imagePheno(cells_sc) <- DataFrame(imageID = cur_sce$sample_id, 
          PID = cur_sce$PID,
          spatial_CD8 = spatial_CD8_factor
          )

cells_sc

# Run differential interaction analysis 
spicyTest <- spicy(cells_sc, 
      condition = "spatial_CD8", 
      subject = "PID", 
      BPPARAM = BiocParallel::MulticoreParam(),
      includeZeroCells = TRUE)

saveRDS(spicyTest, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_spicyR/spicy_inflamedvscold_CD8vstumor_withzero.rds")

spicyTest <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_spicyR/spicy_inflamedvscold_CD8vstumor_withzero.rds")
topPairs(spicyTest)

signifPlot(spicyTest,
           type = "heatmapbliblablub",
           breaks=c(-3, 3, 0.5),
           fdr = TRUE,
           marksToPlot = c("tumor","CD8_T_cell"))


### Second comparison: CD8-T-cell interaction with tumor phenotypes between "cold/excluded" vs. "inflamed"

# Generate segmentedCells object 
cur_sce <- sce[, sce$metacluster_tumor_combined %in% c("CD8_T_cell","tumor")]

cells <- data.frame(row.names = colnames(cur_sce))
cells$x <- cur_sce$Pos_X
cells$y <- cur_sce$Pos_Y
cells$imageID <- cur_sce$sample_id
cells$imageCellID <- cur_sce$ObjectNumber
cells$cellType <- cur_sce$metacluster_combined

cells_sc <- SegmentedCells(cells, cellProfiler = TRUE)

# Add condition and patient id
spatial_CD8_factor <- relevel(x=factor(cur_sce$spatial_CD8_pheno_merg), ref = "cold/excluded")

imagePheno(cells_sc) <- DataFrame(imageID = cur_sce$sample_id, 
          PID = cur_sce$PID,
          spatial_CD8 = spatial_CD8_factor
          )

cells_sc

# Run differential interaction analysis 
spicyTest <- spicy(cells_sc, 
      condition = "spatial_CD8", 
      subject = "PID", 
      BPPARAM = BiocParallel::MulticoreParam(),
      includeZeroCells = TRUE)

saveRDS(spicyTest, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_spicyR/spicy_inflamedvscold_CD8vstumorphenotypes_withzero.rds")
spicyTest2 <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_spicyR/spicy_inflamedvscold_CD8vstumorphenotypes_withzero.rds")

topPairs(spicyTest2)

signifPlot(spicyTest2,
           type = "heatmapbliblablub",
           breaks=c(-3, 3, 0.5),
           fdr = TRUE,
           marksToPlot = unfactor(unique(cur_sce$metacluster_combined)))


### Third comparison: General phenotype interaction between "cold/excluded" vs. "inflamed"

# Generate segmentedCells object 
cur_sce <- sce

cells <- data.frame(row.names = colnames(cur_sce))
cells$x <- cur_sce$Pos_X
cells$y <- cur_sce$Pos_Y
cells$imageID <- cur_sce$sample_id
cells$imageCellID <- cur_sce$ObjectNumber
cells$cellType <- cur_sce$metacluster_combined

cells_sc <- SegmentedCells(cells, cellProfiler = TRUE)

# Add condition and patient id
unique(cur_sce$spatial_CD8_pheno_merg)

spatial_CD8_factor <- relevel(x=factor(cur_sce$spatial_CD8_pheno_merg), ref = "cold_excluded")

imagePheno(cells_sc) <- DataFrame(imageID = cur_sce$sample_id, 
          PID = cur_sce$PID,
          spatial_CD8 = spatial_CD8_factor
          )

cells_sc

# Run differential interaction analysis 
spicyTest <- spicy(cells_sc, 
      condition = "spatial_CD8", 
      subject = "PID", 
      BPPARAM = BiocParallel::MulticoreParam(),
      includeZeroCells = TRUE)

saveRDS(spicyTest, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_spicyR/spicy_inflamedvscold_allVSall_withzero.rds")

spicyTest3 <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_spicyR/spicy_inflamedvscold_allVSall_withzero.rds")
topPairs(spicyTest3)

signifPlot(spicyTest3,
           type = "heatmapbliblablub",
           breaks=c(-3, 3, 0.5),
           fdr = TRUE,
           marksToPlot = unfactor(unique(sce$metacluster_combined)))
```

```{r}
results <- spicyTest3
marksToPlot <- unfactor(unique(sce$metacluster_combined))
breaks <- NULL
comparisonGroup <- NULL
colours = c("#4575B4", "white", "#D73027")
cutoff = 0.05
fdr = TRUE

if (is.null(comparisonGroup)) {
    coef <- 2
  } else {
    coef <- which(levels(results$condition) == comparisonGroup)
  }
  marks <- unique(results$comparisons$from)
  
  if (is.null(breaks)) breaks <- c(-3, 3, 0.5)
  breaks <- seq(from = breaks[1], to = breaks[2], by = breaks[3])
  pal <- grDevices::colorRampPalette(colours)(length(breaks))
  col <- colorRamp2(breaks, pal)

  pVal <- results$p.value[, coef]

  if (min(pVal, na.rm = TRUE) == 0) {
    pVal[pVal == 0] <-
      pVal[pVal == 0] + 10^floor(log10(min(pVal[pVal > 0], na.rm = TRUE)))
  }

  if (fdr) {
    pVal <- stats::p.adjust(pVal, method = "fdr")
  }

  isGreater <- results$coefficient[, coef] > 0

  results$coefficient[, 1]
  results$coefficient[, coef]
  
  pVal <- log10(pVal)

  pVal[isGreater] <- abs(pVal[isGreater])

  pVal <- matrix(pVal, nrow = length(marks))
  colnames(pVal) <- marks
  rownames(pVal) <- marks
  
  pVal[abs(pVal) < -log10(0.05)] <- 0
  
  anno <- data.frame(celltypes = marksToPlot)
  
  anno$type <- ifelse(str_detect(anno$celltypes, "tumor"),"tumor","TME")
  
Heatmap(pVal,
          col = col,
          name = "log10(p)",
          row_split = anno$type,
          column_split = anno$type,
          cluster_columns = FALSE, 
          cluster_rows = FALSE, 
          border = TRUE, 
          row_title = "From", 
          column_title = "To")
```

### B cell patch analysis 

```{r}
## Patch detection
sce <- patchDetection(sce,
               patch_cells = sce$metacluster_tumor_combined == "B_cell",
               colPairName = "knn_interaction_graph",
               min_patch_size = 10,
               name = "patch_B_cell",
               expand_by = 0, 
               img_id = "sample_id", 
               BPPARAM = MulticoreParam(workers = 20))

## 1. Number of patches per PID 
n_Bpatch_PID <- colData(sce) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  unique() %>% 
  group_by(spatial_CD8_pheno_merg, PID) %>% 
  count()

plot_df <- colData(sce) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg) %>% 
  unique() 
rownames(plot_df) <- NULL
plot_df$n_Bpatch <- n_Bpatch_PID$n[match(plot_df$PID, n_Bpatch_PID$PID)]
plot_df$n_Bpatch[is.na(plot_df$n_Bpatch)] <- 0

plot_df$spatial_CD8_pheno <- sce$spatial_CD8_pheno[match(plot_df$PID, sce$PID)]

ggplot(plot_df, aes(x = spatial_CD8_pheno, y = n_Bpatch))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = spatial_CD8_pheno), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$spatial_CD8_pheno)+
  theme_classic(base_size = 12)+
  coord_cartesian(ylim = c(0,15))+
  theme(axis.title.x = element_blank())+
  ylab("Number of B cell patches per PID")+
  stat_compare_means(method = "kruskal.test")
  

## 2. Size of B cell patches 
size_Bpatch_PID <- colData(sce) %>% as.data.frame %>% select(PID, spatial_CD8_pheno_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  group_by(patch_B_cell, PID) %>% 
  count()

size_Bpatch_PID$spatial_CD8_pheno_merg <- plot_df$spatial_CD8_pheno_merg[match(size_Bpatch_PID$PID, plot_df$PID)]
size_Bpatch_PID$spatial_CD8_pheno <- sce$spatial_CD8_pheno[match(size_Bpatch_PID$PID, sce$PID)]

ggplot(size_Bpatch_PID, aes(x = spatial_CD8_pheno, y = n))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = spatial_CD8_pheno), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$spatial_CD8_pheno)+
  theme_classic(base_size = 12)+
  #coord_cartesian(ylim = c(0,15))+
  theme(axis.title.x = element_blank())+
  #scale_y_log10()+
  ylab("Number of B cells per patch")+
  stat_compare_means(method = "kruskal.test")

# Define cut-off for small and large B cell follicles - use 30 B cells per patch
summary(size_Bpatch_PID$n)

size_Bpatch_PID <- size_Bpatch_PID %>% mutate(patch_score = ifelse(n > 30, "large", "small"))

PID_large_patches <- size_Bpatch_PID %>% filter(patch_score == "large") %>% pull(PID) %>% unique()

size_Bpatch_PID <- size_Bpatch_PID %>% mutate(large_Bpatch = ifelse(PID %in% PID_large_patches, TRUE, FALSE))

## 3. Create B cell score per PID 
# No B cells = 1 
# No B cell patches = 2
# Small B cell patches = 3
# Large B cell patches = 4 

B_cells_PID <- colData(sce) %>% as.data.frame %>% select(PID, metacluster_combined) %>% filter(metacluster_combined == "B_cell") %>% group_by(PID, metacluster_combined) %>% count()

plot_df$n_Bcells <- B_cells_PID$n[match(plot_df$PID, B_cells_PID$PID)]

plot_df$n_Bcells[is.na(plot_df$n_Bcells)] <- 0

plot_df$large_Bpatch <- size_Bpatch_PID$large_Bpatch[match(plot_df$PID, size_Bpatch_PID$PID)]


plot_df <- plot_df %>% mutate(B_cell_score = ifelse(n_Bcells == 0, "No_B_cell", 
                                         ifelse(n_Bpatch == 0, "No_B_patch", 
                                                ifelse(large_Bpatch == FALSE, "Small_B_patch", "Large_B_patch"))))

plot_df$B_cell_score <- factor(plot_df$B_cell_score, levels = c("No_B_cell","No_B_patch", "Small_B_patch", "Large_B_patch"))


plot <- plot_df %>%
  group_by(B_cell_score, spatial_CD8_pheno) %>%
  summarise(n=n()) %>%
  group_by(spatial_CD8_pheno) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup()


plot_test <- plot %>% pivot_wider(names_from = spatial_CD8_pheno, id_cols = B_cell_score, values_from = n) %>% column_to_rownames("B_cell_score")

plot_test_out <- fisher.test(plot_test, simulate.p.value = TRUE)

round(plot_test_out$p.value, 4)

ggplot(plot) + 
  geom_col(aes(y=spatial_CD8_pheno, x=fraction, fill=B_cell_score), color = "black") +
  theme_classic() + 
  xlab("Fraction of PID") +
  ylab("")+
  scale_fill_brewer(palette = "Paired")+
  coord_flip()+
  ggtitle(paste0("p (Fisher) = ", plot_test_out$p.value))


## 4. Add to sce
#sce$B_cell_score <- plot_df$B_cell_score[match(sce$PID, plot_df$PID)]
```

### T cell expression analysis 

```{r}
set.seed(22)

# Focus on Ki67 
# Subset to CD8_T_cells
#cur_sce <- sce[,sce$metacluster_combined == "CD8_T_cell"]


# Define Ki67 positivity cut-off based on GMM (exprs)
GMM_Ki67 <- Mclust(assay(sce, "exprs")["Ki67",], G = 2)

GMM_df <- data.frame(class_Ki67 = ifelse(GMM_Ki67$classification == 2 & GMM_Ki67$uncertainty == 0, 
                                        "Ki67+",
                                        "Ki67-"), 
                     Ki67_exprs = assay(sce, "exprs")["Ki67",],
                     uncertainty = GMM_Ki67$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = Ki67_exprs, fill = class_Ki67))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")
  
#Add GMM epi class to SCE
sce$GMM_Ki67 <- GMM_df$class_Ki67

# Violin plot
cur_dat <- t(assay(cur_sce, "exprs")) %>% as.data.frame
cur_dat$CD8Tcell_type <- sce$CD8Tcell_type[match(rownames(cur_dat),colnames(sce))]
cur_dat$CD8Tcell_type <- factor(cur_dat$CD8Tcell_type, levels = c("ignored","bystanding","infiltrating"))
cur_dat$Ki67_GMM <- sce$GMM_Ki67[match(rownames(cur_dat),colnames(sce))]

cur_out <- cur_dat %>% select(CD8Tcell_type, Ki67_GMM) %>% group_by(CD8Tcell_type, Ki67_GMM) %>% count()

cur_plot <- cur_out %>% group_by(CD8Tcell_type) %>% mutate(fra = n/sum(n))

cur_test <- cur_plot %>% pivot_wider(names_from = CD8Tcell_type, id_cols = Ki67_GMM, values_from = n) %>% column_to_rownames("Ki67_GMM")

cur_test_out <- chisq.test(cur_test)

cur_test_out$p.value

ggplot(cur_plot) +
  geom_tile(color = "black",aes(CD8Tcell_type, fra, fill = Ki67_GMM), position = "stack")+
  scale_fill_brewer(palette = "Paired")+
  theme_classic(base_size = 15)+
  ylab("Density")+
  theme(axis.title.x = element_blank())+
  ggtitle(paste("p (Chi-Square) = ",cur_test_out$p.value))
```

### Tumor expression analysis 

```{r}
## Analyse tumor compartment expression 

# Subset to tumor cells
sce_tumor <- sce[,sce$analysis_cat == "tumor"]

# Aggregate expression per PID 
spatialpheno_mean_PID <- aggregateAcrossCells(sce_tumor, ids = sce_tumor$PID,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_tumor)$tumor_cluster_channel)

# Test for differential expression per category 
test <- t(assay(spatialpheno_mean_PID, "exprs")[rowData(spatialpheno_mean_PID)$tumor_cluster_channel,]) %>% as.data.frame %>% mutate(spatial_CD8_pheno = spatialpheno_mean_PID$spatial_CD8_pheno) 

#test$spatial_CD8_pheno <- unfactor(test$spatial_CD8_pheno)

test_result <- compare_means(c(H3K27me3, CK5, HLADR, CK8_18, CD15, KRT14, Vimentin, 
cMYC, HER2, pH3, ER, PR, p53, AR, GATA3, CAIX, 
Ki67, EGFR, S6, p_mTOR, CK7, c_Cas3_PARP)~spatial_CD8_pheno, method = "kruskal.test",data = test)

test_result$p.adj <- p.adjust(test_result$p, method = "BH") 

test_result$psignif.adj <- as.matrix(symnum(test_result$p.adj,
       cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), 
       symbols = c("****", "***", "**", "*", "ns")))[,1]

rownames(test_result) <- test_result$.y.

sig.poteins <- unfactor(test_result$.y.[test_result$p.adj < 0.05])


# Aggregate expression per spatial phenotype
spatialpheno_mean_sce <- aggregateAcrossCells(sce_tumor, ids = sce_tumor$spatial_CD8_pheno,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_tumor)$tumor_cluster_channel)

mat_scaled<- t(scale(t(assay(spatialpheno_mean_sce, "exprs"))))


highest_cat <- mat_scaled %>% as.data.frame() %>% rownames_to_column("gene") %>% pivot_longer(cols = 2:4, names_to = "spatial_CD8", values_to = "exprs") %>% group_by(gene) %>% mutate(max = max(exprs)) %>% filter(exprs == max) %>% select(gene, spatial_CD8) %>% as.data.frame() %>% column_to_rownames("gene")

## Visualize 

# Heatmap 
h <- Heatmap(assay(spatialpheno_mean_sce, "exprs"), col = viridis(100), name = "exprs", cluster_columns = FALSE, row_names_side = "left", show_row_dend = FALSE)+
  #Heatmap(mat_scaled, name = "scaled exprs", cluster_columns = FALSE)+
  rowAnnotation(cat = anno_simple(highest_cat, col = metadata(sce)$colors$spatial_CD8_pheno, width = unit(10, "mm")))+
  rowAnnotation(test = anno_text(test_result$psignif.adj))

lgd_cat <- Legend(title = "Highest_exprs", at = names(metadata(sce)$colors$spatial_CD8_pheno), 
              legend_gp = gpar(fill = metadata(sce)$colors$spatial_CD8_pheno))


draw(h,annotation_legend_list = list(lgd_cat))

# Violin plot
plotExpression(spatialpheno_mean_PID, 
               features = sig.poteins,
               x = "spatial_CD8_pheno", exprs_values = "exprs", 
               colour_by = "spatial_CD8_pheno") +
    theme(axis.text.x =  element_text(angle = 90))+
    scale_color_manual(values = metadata(sce)$colors$spatial_CD8_pheno)
```

## Immune mixing score - Keren et al., 2020

```{r}
# Patients with less than 1000 immune cells - cold
# Calculate mixing score based on immune-immune and tumor-immune interactions
# Classify rest of patients as mixed or compartmentalized

## Step 1: Count interactions for immune-immune and tumor-immune
cur_dat <- sce$count_neighbors %>% as.data.frame
rownames(cur_dat) <- sce$cell_id
cur_dat$PID <- sce$PID
cur_dat$analysis_cat <- sce$analysis_cat

#Exclude endothelial, immune_low and stromal cells
cur_dat <- cur_dat %>% select(-Endothelial, -Immune_low, -Stromal)
cur_dat <- cur_dat %>% mutate(immune = B_cell+CD4_T_cell+CD8_T_cell+HLA_DR+Monocyte+Neutrophil, .after = Neutrophil)

#calculate number of immune cells
number_immune <- cur_dat %>% filter(analysis_cat == "non_tumor") %>% select(PID, analysis_cat) %>% group_by(PID) %>% count()

## Step 2: Calculate mixing score
cur_out <- cur_dat %>% group_by(PID, analysis_cat) %>% summarise(immune = sum(immune))
cur_out <- cur_out %>% pivot_wider(id_cols = PID, names_from = analysis_cat, values_from = immune) %>% mutate(mixing_score = tumor/non_tumor) %>% filter(tumor != 0)

cur_out$number_immune <- number_immune$n[match(cur_out$PID, number_immune$PID)]

## Step 3: Classify mixing score

#Exclude "cold" tumors
cur_k <- cur_out %>% filter(number_immune > quantile(cur_out$number_immune)[2]) #quantile(cur_out$number_immune)[2]
#cur_k <- cur_out %>% filter(number_immune > 250)

#kmeans
set.seed(22)
#cur_k$mixing_pheno <- kmeans(cur_k$mixing_score, 2, nstart = 1000)$cluster
#cur_k$mixing_pheno <- ifelse(cur_k$mixing_pheno == "1", "mixed","compartmentalized")
cur_k$mixing_pheno <- ifelse(cur_k$mixing_score > 0.22, "mixed","compartmentalized")

#add back
cur_out$mixing_pheno <- cur_k$mixing_pheno[match(cur_out$PID, cur_k$PID)]
cur_out$mixing_pheno[is.na(cur_out$mixing_pheno)] <- "cold"


## Add to sce
sce$mixing_pheno <- cur_out$mixing_pheno[match(sce$PID, cur_out$PID)]

## Create color scheme
mixing_pheno_col <- setNames(c("seagreen","purple","steelblue1"), unique(sce$mixing_pheno))
metadata(sce)$colors$mixing_pheno <- mixing_pheno_col
```

### Kaplan-Meier 

```{r}
#Assess survival impact 
cur_surv <- colData(sce) %>% as.data.frame() %>% select(PID, mixing_pheno, OS_months, DFS_months, status_DFS, status_OS) %>% unique() #%>% filter(!mixing_pheno %in% "cold")

#Plot survival curves
fit <- survfit(Surv(OS_months, status_OS) ~ mixing_pheno, data = cur_surv)
print(fit)

mix_OS <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("steelblue1","purple","seagreen"),
          log.rank.weights = "n")

fit <- survfit(Surv(DFS_months, status_DFS) ~ mixing_pheno, data = cur_surv)
print(fit)

mix_DFS <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("steelblue1","purple","seagreen"),
          log.rank.weights = "n")

mix_OS
mix_DFS
```

```{r}
out_viz <- colData(sce) %>% as.data.frame %>% select(PID, mixing_pheno, spatial_CD8_pheno) %>% unique() %>% group_by(mixing_pheno, spatial_CD8_pheno) %>% dplyr::count() %>% ungroup() %>% group_by(mixing_pheno) %>% mutate(fra = n/sum(n))

out_viz$mixing_pheno <- factor(out_viz$mixing_pheno, levels = c("cold", "mixed", "compartmentalized"))

ggplot(out_viz, aes(x = mixing_pheno, y = fra, fill = spatial_CD8_pheno))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = metadata(sce)$colors$spatial_CD8_pheno)
```


```{r}
PID_CD8 <- colData(sce) %>% as.data.frame() %>% select(mixing_pheno,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_CD8 <- PID_CD8 %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_CD8 %>% pivot_wider(id_cols = "PID",names_from = "mixing_pheno",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -compartmentalized) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "),"+compartmentalized"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#write.csv(plot_df_prop, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/supp_tables/Cox_Multi_NumberGroup.csv")

#Plot hazard ratiDFS
hr_prop <- 
ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0.1,100))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metaclusters")+
ylab("HR for con_prop")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

# 4. Image visualization 

## Extract representative images

```{r}
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_TNBC_new.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")
```

```{r}
#Select images of interest 

## Inflamed patients with B cell patches 
img_ids <- colData(sce) %>% as.data.frame() %>% select(PID, sample_id, B_cell_score, spatial_CD8_pheno) %>% unique() %>% filter(spatial_CD8_pheno == "inflamed" & B_cell_score == "Large_B_patch") %>% pull(sample_id) %>% unique()

#img_ids <- img_ids[img_ids == "ZTMA249.2_ZTMA249_ID_197_BX5Y10_171"]

# Select images
cur_images <- images[img_ids]

#Subset mask and object
cur_mask <- masks[img_ids]
cur_sce <- sce[,sce$sample_id %in% names(cur_images)]
```

```{r}
##Visualize metaclusters on images - select 10 images for each category with highest number of cells
cur_PID <- colData(sce) %>% as.data.frame %>% group_by(metacluster_combined) %>% filter(metacluster_combined == "tumor_8") %>% count(PID) %>% arrange(desc(n)) %>% slice(1:10) %>% pull(PID)

cur_sample <- colData(sce) %>% as.data.frame %>% filter(PID %in% cur_PID) %>% pull(sample_id) %>% unique()

#cur_sample <- "ZTMA249.1_ZTMA249_ID_195_BX3Y10_18"

# Images
cur_images <- images[cur_sample]
cur_masks <- masks[cur_sample]

#Subset sce
cur_sce <- sce[,sce$sample_id %in% names(cur_images)]
```

Interactive visualization with cytoviewer 

```{r}
app <- cytoviewer(image = cur_images, mask = cur_masks, object = cur_sce, cell_id = "ObjectNumber",img_id = "sample_id")

shiny::runApp(app, launch.browser = TRUE)

cur_sample
```

# Save objects

```{r save_objects, message = FALSE, warning=FALSE}
#Save sce
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
```

# Session Info 

```{r}
sessionInfo()
```