---
title: "08_combined_patient_grouping"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: sce_ALL.rds 

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Script 8: Patient group analysis

Here, we will define patient groups based on tumor and TME features.

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scran)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(glmnet)
library(broom) 
library(lisaClust)
library(spicyR)
library(survival)
library(survminer)
library(ggnewscale)
library(scales)
library(ggpubr)
library(wesanderson)
library(cytoviewer)
library(edgeR)
library(doParallel)
library(vegan)
library(easyalluvial)
library(ggalluvial)
```

# Load data

```{r load combined sce}
##Load sce
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
```


# 1. Patient grouping 

```{r}
# Add CK5_7_group
sce_tumor <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor_new.rds")
sce$CK5_7_group <- sce_tumor$CK5_7_group[match(sce$PID, sce_tumor$PID)]

metadata(sce)$colors$CK5_7_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("CK5/7-", "CK5/7+", "CK5+", "CK7+"))
```

## Based on spatial CD8 phenotype and CK5/7 scores 

```{r}
# Give CK5_7 groups interpretable names 

# Merge
sce$CK5_7_group_name <- sce$CK5_7_group

sce$CK5_7_group_name <- str_replace(sce$CK5_7_group_name, "1", "CK5/7+") 
sce$CK5_7_group_name <- str_replace(sce$CK5_7_group_name, "2", "CK5+") 
sce$CK5_7_group_name <- str_replace(sce$CK5_7_group_name, "3", "CK7+") 
sce$CK5_7_group_name <- str_replace(sce$CK5_7_group_name, "4", "CK5/7-") 
```

```{r}
# Combine information from spatial CD8 phenotype and CK5/7 scores
sce$BL_CD8_group <- paste0(sce$CK5_7_group, "_",sce$spatial_CD8_pheno_merg)

# Merge
sce$BL_CD8_group_merg <- sce$BL_CD8_group

sce$BL_CD8_group_merg <- str_replace(sce$BL_CD8_group_merg, "1", "BL") 
sce$BL_CD8_group_merg <- str_replace(sce$BL_CD8_group_merg, "2", "B") 
sce$BL_CD8_group_merg <- str_replace(sce$BL_CD8_group_merg, "3", "L") 
sce$BL_CD8_group_merg <- str_replace(sce$BL_CD8_group_merg, "4", "N") 
sce$BL_CD8_group_merg <- str_replace(sce$BL_CD8_group_merg, "/","_")

# Groups: Inflamed (Good prognostic)
sce$BL_CD8_group_merg[str_detect(sce$BL_CD8_group_merg,"inflamed")] <- "inflamed"

sce$BL_CD8_group_merg[sce$BL_CD8_group_merg == "B_cold_excluded"] <- "B"
sce$BL_CD8_group_merg[sce$BL_CD8_group_merg == "L_cold_excluded"] <- "L"
sce$BL_CD8_group_merg[sce$BL_CD8_group_merg == "BL_cold_excluded"] <- "BL"
sce$BL_CD8_group_merg[sce$BL_CD8_group_merg == "N_cold_excluded"] <- "N"
```

```{r}
# Risk group definition
sce$PID_group_risk <- ifelse(sce$BL_CD8_group_merg == "inflamed", "low", ifelse(sce$BL_CD8_group_merg == "BL", "high", "medium"))

risk_color <- setNames(c(wes_palette("Zissou1")[5], wes_palette("Zissou1")[1],"gold"), 
                       c("high", "low", "medium"))
metadata(sce)$colors$risk_color <- risk_color

# Define factors 
sce$BL_CD8_group_merg <- factor(sce$BL_CD8_group_merg, levels = c("inflamed","N","B","L","BL"))
sce$PID_group_risk <- factor(sce$PID_group_risk, levels = c("low","medium","high"))
```

```{r save_objects, message = FALSE, warning=FALSE}
#Save sce
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/08_sce_PIDgroup.rds")
```

### Alluvial 

```{r}
df_alluvial <- colData(sce) %>% as.data.frame %>% dplyr::select(PID, CK5_7_group_name, spatial_CD8_pheno_merg,BL_CD8_group_merg, PID_group_risk) %>% unique %>% dplyr::select(CK5_7_group_name, spatial_CD8_pheno_merg,BL_CD8_group_merg, PID_group_risk)

alluvial_wide(df_alluvial, 
              colorful_fill_variable_stratum = FALSE,
              col_vector_flow = c("#85D4E3","grey", "#9C964A", "#F4B5BD"), 
              order_levels = c("CK5/7+","CK5/7-","CK7+","CK5+","cold/excluded","BL","N","L","B","high","medium","low"))+
  theme_classic()
```

### Kaplan-Meier 

```{r}
#Assess survival impact 
cur_surv <- colData(sce) %>% as.data.frame() %>% 
  dplyr::select(PID, BL_CD8_group_merg, PID_group_risk, OS_months, DFS_months, status_DFS, status_OS, pretreatment) %>% 
  filter(pretreatment == "FALSE") %>%
  unique()

# Define colors 
group_color <- setNames(c("#3B9AB2", "gold", "pink", "orange","#F21A00"), c("inflamed","N","B","L","BL"))

metadata(sce)$colors$group_color <- group_color

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ BL_CD8_group_merg, data = cur_surv)
print(fit)

PID_DFS_1 <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = unname(group_color))

PID_DFS_1

#Plot survival curves
fit <- survfit(Surv(DFS_months, status_DFS) ~ PID_group_risk, data = cur_surv)
print(fit)

PID_DFS_2 <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, 
          risk.table.col = "strata", 
          linetype = "strata", 
          surv.median.line = "hv", 
          ggtheme = theme_classic(), 
          palette = c(wes_palette("Zissou1")[1] ,"gold", wes_palette("Zissou1")[5]))

PID_DFS_2
```

### Multivariate 

PID groups 

```{r coxph dfs}
### Multivariate cox proportional hazard model on multiple covariates (here PID groups)
BL_CD8_group <- colData(sce) %>% as.data.frame() %>% dplyr::select(BL_CD8_group_merg,PID) %>% group_by(PID) %>% table() %>% as.data.frame

BL_CD8_group <- BL_CD8_group %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

BL_CD8_group <- BL_CD8_group %>% pivot_wider(id_cols = "PID",names_from = "BL_CD8_group_merg",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on SCP assignments
Cox_df <- BL_CD8_group

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% dplyr::select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -inflamed) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "), "+ inflamed"))


multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("SCP")+
ylab("HR for PID groups")+
theme_classic()+
coord_cartesian(ylim=c(0.1,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r coxph dfs}
### Multivariate cox proportional hazard model on multiple covariates (here PID groups)
BL_CD8_group <- colData(sce) %>% as.data.frame() %>% dplyr::select(PID_group_risk,PID) %>% group_by(PID) %>% table() %>% as.data.frame

BL_CD8_group <- BL_CD8_group %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

BL_CD8_group <- BL_CD8_group %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on SCP assignments
Cox_df <- BL_CD8_group

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

Cox_df$age <- sce$age[match(rownames(Cox_df),sce$PID)]
Cox_df$grade <- factor(sce$grade[match(rownames(Cox_df),sce$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce$pT_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pN_simple <- as.factor(sce$pN_simple[match(rownames(Cox_df),sce$PID)])
Cox_df$pM_simple <- as.factor(sce$pM_simple[match(rownames(Cox_df),sce$PID)])

#Based on proportions
covariates <- Cox_df %>% dplyr::select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple, -low) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "), "+ low"))


multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("SCP")+
ylab("HR for PID groups")+
theme_classic()+
coord_cartesian(ylim=c(0.1,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

### Univariate 

PID groups 

```{r}
### Multivariate cox proportional hazard model on multiple covariates (here PID groups)
BL_CD8_group <- colData(sce) %>% as.data.frame() %>% dplyr::select(BL_CD8_group_merg,PID) %>% group_by(PID) %>% table() %>% as.data.frame

BL_CD8_group <- BL_CD8_group %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

BL_CD8_group <- BL_CD8_group %>% pivot_wider(id_cols = "PID",names_from = "BL_CD8_group_merg",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on SCP assignments
Cox_df <- BL_CD8_group

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

#Based on proportions
covariates <- Cox_df %>% dplyr::select(-status_DFS,-DFS_months, -inflamed) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~", paste(covariates, collapse = " + "), "+ inflamed"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("SCP")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

PID risk

```{r}
### Multivariate cox proportional hazard model on multiple covariates (here PID groups)
BL_CD8_group <- colData(sce) %>% as.data.frame() %>% dplyr::select(PID_group_risk,PID) %>% group_by(PID) %>% table() %>% as.data.frame

BL_CD8_group <- BL_CD8_group %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

BL_CD8_group <- BL_CD8_group %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on SCP assignments
Cox_df <- BL_CD8_group

#add covariates for correction
Cox_df$status_DFS <- sce$status_DFS[match(rownames(Cox_df),sce$PID)]
Cox_df$DFS_months <- sce$DFS_months[match(rownames(Cox_df),sce$PID)]

#Based on proportions
covariates <- Cox_df %>% dplyr::select(-status_DFS,-DFS_months, -low) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~", paste(covariates, collapse = " + "), "+ low"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_manual(values = "#377EB8")+
xlab("SCP")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

# 2. Patient group visualization

```{r}
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/08_sce_PIDgroup.rds")
```

## Immune scores

```{r}
spatial_CD8 <- colData(sce) %>% as.data.frame() %>% dplyr::select(BL_CD8_group_merg,spatial_CD8_pheno) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
spatial_CD8 <- spatial_CD8 %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq))

ggplot(spatial_CD8, aes(x = BL_CD8_group_merg, y = fra, fill = spatial_CD8_pheno))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  ylab("Fraction of PID")+
  scale_fill_manual(values = metadata(sce)$colors$spatial_CD8_pheno)
```


```{r}
spatial_CD8 <- colData(sce) %>% as.data.frame() %>% select(BL_CD8_group_merg,B_cell_score) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
spatial_CD8 <- spatial_CD8 %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq))

ggplot(spatial_CD8, aes(x = BL_CD8_group_merg, y = fra, fill = B_cell_score))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  ylab("Fraction of PID")
  #scale_fill_manual(values = metadata(sce)$colors$B_cell_score)
```

## Proliferation scores

```{r}
## Across all cells
# 1. Ki67
set.seed(22)
GMM_Ki67 <- Mclust(assay(sce, "exprs")["Ki67",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_Ki67$classification == 2 & GMM_Ki67$uncertainty == 0, 
                                        "Ki67+",
                                        "Ki67-"), 
                     Ki67_exprs = assay(sce, "exprs")["Ki67",],
                     uncertainty = GMM_Ki67$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = Ki67_exprs, fill = class_epi))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")
  
#Add GMM epi class to SCE
sce$GMM_all_Ki67 <- GMM_df$class_epi

sce$GMM_tumor_Ki67 <- ifelse(sce$analysis_cat == "tumor", sce$GMM_all_Ki67, NA)
```

```{r}
tumor_sce <- sce[,sce$analysis_cat == "tumor"]
```

```{r}
prolif_PID <- colData(sce) %>% as.data.frame() %>% select(PID,GMM_tumor_Ki67) %>% group_by(PID) %>% table() %>% as.data.frame
prolif_PID <- prolif_PID %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
prolif_PID$BL_CD8_group_merg <- sce$BL_CD8_group_merg[match(prolif_PID$PID, sce$PID)]
prolif_PID <- prolif_PID %>% filter(GMM_tumor_Ki67 == "Ki67+")

ggplot(prolif_PID, aes(x = reorder(BL_CD8_group_merg,-fra), y = fra))+
  geom_boxplot(outlier.colour = "black", aes(fill = BL_CD8_group_merg))+
  #geom_jitter(aes(color = BL_CD8_group_merg), size = 3, position=position_jitter(w=0.1,h=0.1))+
  theme_classic()+
  scale_fill_manual(values = metadata(sce)$colors$group_color)+
  stat_compare_means()

ggplot(prolif_PID, aes(x = reorder(BL_CD8_group_merg,-fra), y = fra))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = BL_CD8_group_merg), size = 3, position=position_jitter(w=0.1,h=0.1))+
  theme_classic()+
  scale_color_manual(values = metadata(sce)$colors$group_color)+
  stat_compare_means()
```

## CK5/7 score

```{r}
CK5_7 <- colData(sce) %>% as.data.frame() %>% select(BL_CD8_group_merg,CK5_7_score) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
CK5_7 <- CK5_7 %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq))

ggplot(CK5_7, aes(x = BL_CD8_group_merg, y = fra, fill = CK5_7_score))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  ylab("Fraction of PID")
  #scale_fill_manual(values = metadata(sce)$colors$CK5_7_score)
```

## CATALYST-based visualization 

We will first convert the sce into a CATALYST-compatible format.

```{r CATALYST}
# save spe in CATALYST-compatible object with renamed colData entries and 
# new metadata information
sce_cat <- sce 
sce_cat$sample_id <- factor(sce$PID)
sce_cat$cluster_id <- factor(sce$BL_CD8_group_merg)

#add celltype information to metadata
metadata(sce_cat)$cluster_codes <- data.frame(PID_SC_groups = factor(sce_cat$BL_CD8_group_merg))
```

```{r CATALYST pbmds and clrDR, message=FALSE}
# MDS pseudobulk by metacluster
pbMDS(sce_cat, by = "cluster_id", 
      features = rownames(sce_cat)[rowData(sce_cat)$use_channel], 
      label_by = "cluster_id", k = "PID_SC_groups") +
  scale_color_manual(values = metadata(sce_cat)$colors$group_color)+
  theme_classic()

# MDS pseudobulk by metacluster
pbMDS(sce_cat, by = "both", 
      features = rownames(sce_cat)[rowData(sce_cat)$use_channel], 
      k = "PID_SC_groups",
      size_by = TRUE) +
  #scale_color_manual(values = metadata(sce_cat)$colors$metacluster)+
  theme_classic()

# CLR on metacluster proportions across samples
clrDR(sce_cat, dr = "PCA", 
      by = "cluster_id", k = "PID_SC_groups", 
      label_by = "cluster_id", arrow_col = "PID", 
      #point_pal = metadata(sce_cat)$colors$metacluster, 
      arrows = FALSE) +
  scale_color_manual(values = metadata(sce_cat)$colors$PID)+
  guides(color = "none")+
  theme_classic()
```

## Expression heatmaps 

```{r}
tumor_sce <- sce[,sce$analysis_cat == "tumor"]
cluster_mean_sce <- aggregateAcrossCells(tumor_sce, ids = tumor_sce$BL_CD8_group_merg,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(tumor_sce)$clean_target %in% c("HLADR", "CK5","CK7","KRT14","CK8_18"))

Heatmap(t(assay(cluster_mean_sce, "exprs")), col = viridis(100))
Heatmap(scale(t(assay(cluster_mean_sce, "exprs"))))
```

```{r}
group_mean_sce <- aggregateAcrossCells(sce, ids = sce$BL_CD8_group_merg,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$clean_target %in% c("CK5","CK7","HLADR"))

group_mean_sce_2 <- aggregateAcrossCells(sce, ids = sce$BL_CD8_group_merg,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$clean_target %in% c("CK8_18","KRT14","CD8a", "Ki67"))

Heatmap(t(assay(group_mean_sce, "exprs")), col = viridis(100))+
  Heatmap(t(assay(group_mean_sce_2, "exprs")), col = viridis(100))


Heatmap(scale(t(assay(group_mean_sce, "exprs"))), clustering_method_rows = "ward.D2")+
  Heatmap(scale(t(assay(group_mean_sce_2, "exprs"))), clustering_method_rows = "ward.D2")
```

```{r}
group_mean_sce_prolif <- aggregateAcrossCells(tumor_sce, ids = tumor_sce$BL_CD8_group_merg,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(tumor_sce)$clean_target %in% c("Ki67", "pH3", "CD44"))

Heatmap(t(assay(group_mean_sce_prolif, "exprs")), col = viridis(100))

Heatmap(scale(t(assay(group_mean_sce_prolif, "exprs"))), clustering_method_rows = "ward.D2")
```

```{r}
n_Bpatch_PID <- colData(sce) %>% as.data.frame %>% select(PID, BL_CD8_group_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  unique() %>% 
  group_by(BL_CD8_group_merg, PID) %>% 
  count()

plot_df <- colData(sce) %>% as.data.frame %>% select(PID, BL_CD8_group_merg) %>% 
  unique() 
rownames(plot_df) <- NULL
plot_df$n_Bpatch <- n_Bpatch_PID$n[match(plot_df$PID, n_Bpatch_PID$PID)]
plot_df$n_Bpatch[is.na(plot_df$n_Bpatch)] <- 0

plot_df$BL_CD8_group_merg <- sce$BL_CD8_group_merg[match(plot_df$PID, sce$PID)]

ggplot(plot_df, aes(x = BL_CD8_group_merg, y = n_Bpatch))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = BL_CD8_group_merg), size = 3, position=position_jitter(w=0.1,h=0.1))+
  guides(color = "none")+
  theme_classic(base_size = 12)+
  coord_cartesian(ylim = c(0,15))+
  theme(axis.title.x = element_blank())+
  ylab("Number of B cell patches per PID")+
  stat_compare_means(method = "kruskal.test")
```

## Rich heatmap

```{r heatmap sce_1, message=FALSE, warning=FALSE}
### 1. Heatmap body ###

# Tumor metacluster
PID_spatial_tumor <- colData(sce) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_spatial_tumor <- PID_spatial_tumor %>% filter(str_detect(metacluster_combined, "tumor")) %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_spatial_tumor$BL_CD8_group_merg <- sce$BL_CD8_group_merg[match(PID_spatial_tumor$PID, sce$PID)]
PID_spatial_tumor <- PID_spatial_tumor[!is.na(PID_spatial_tumor$BL_CD8_group_merg),]

hb_tumor <- PID_spatial_tumor %>% ungroup %>% select(metacluster_combined,BL_CD8_group_merg,fra) %>% group_by(BL_CD8_group_merg, metacluster_combined) %>% summarize(mean = mean(fra)) %>% pivot_wider(names_from = metacluster_combined, values_from = mean) %>% column_to_rownames("BL_CD8_group_merg")
hb_tumor_scaled <- scale(as.matrix(hb_tumor))

# TME metacluster
PID_spatial_TME <- colData(sce) %>% as.data.frame() %>% select(metacluster_combined,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_spatial_TME <- PID_spatial_TME %>% filter(!str_detect(metacluster_combined, "tumor")) %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
PID_spatial_TME$BL_CD8_group_merg <- sce$BL_CD8_group_merg[match(PID_spatial_TME$PID, sce$PID)]
PID_spatial_TME <- PID_spatial_TME[!is.na(PID_spatial_TME$BL_CD8_group_merg),]

hb_TME <- PID_spatial_TME %>% ungroup %>% select(metacluster_combined,BL_CD8_group_merg,fra) %>% group_by(BL_CD8_group_merg, metacluster_combined) %>% summarize(mean = mean(fra)) %>% pivot_wider(names_from = metacluster_combined, values_from = mean) %>% column_to_rownames("BL_CD8_group_merg")
hb_TME_scaled <- scale(as.matrix(hb_TME))

### 2. Heatmap annotation ###

###2.1 Row - Metadata features

# Annotation dataframe 
anno <- colData(sce) %>% as.data.frame %>% select(BL_CD8_group_merg) %>% unique() %>% filter(!is.na(BL_CD8_group_merg))
rownames(anno) <- anno$BL_CD8_group_merg
  
# Proportion of grade per cluster
grade <- colData(sce) %>% as.data.frame() %>% select(BL_CD8_group_merg,grade) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
grade <- grade %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq)) 
grade <- grade %>% select(-Freq) %>% pivot_wider(id_cols = BL_CD8_group_merg,names_from = grade,values_from = fra) %>% column_to_rownames("BL_CD8_group_merg")

# Proportion of pT stage per cluster
pT <- colData(sce) %>% as.data.frame() %>% select(BL_CD8_group_merg,pT_simple) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
pT <- pT %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq)) 
pT <- pT %>% select(-Freq) %>% pivot_wider(id_cols = BL_CD8_group_merg,names_from = pT_simple,values_from = fra) %>% column_to_rownames("BL_CD8_group_merg")

# Proportion of pN stage per cluster
pN <- colData(sce) %>% as.data.frame() %>% select(BL_CD8_group_merg,pN_simple) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
pN <- pN %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq)) 
pN <- pN %>% select(-Freq) %>% pivot_wider(id_cols = BL_CD8_group_merg,names_from = pN_simple,values_from = fra) %>% column_to_rownames("BL_CD8_group_merg")

# Number of contributing patients
cluster_PID <- colData(sce) %>% as.data.frame() %>% select(BL_CD8_group_merg,PID) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
n_PID <- cluster_PID %>% filter(Freq>0) %>% group_by(BL_CD8_group_merg) %>% count(name = "n_PID") %>% column_to_rownames("BL_CD8_group_merg")

# Risk group 
anno$BL_CD8_group_merg <- anno$BL_CD8_group_merg[match(rownames(n_PID), rownames(anno))]
rownames(anno) <- anno$BL_CD8_group_merg

anno$PID_group_risk <- sce$PID_group_risk[match(anno$BL_CD8_group_merg, sce$BL_CD8_group_merg)]


# Metadata annotation
ha_group <- HeatmapAnnotation(BL_CD8_group_merg = anno_simple(unfactor(anno$BL_CD8_group_merg),border=TRUE,col=metadata(sce)$colors$group_color),
                              PID_group_risk = anno_simple(unfactor(anno$PID_group_risk),border=TRUE,col=metadata(sce)$colors$risk_color),
                             border = TRUE, 
                             annotation_name_rot = 90,
                             gap = unit(2,"mm"),
                             which = "row")

ha_meta <- HeatmapAnnotation(n_PID = anno_barplot(n_PID, width = unit(10, "mm"),gp = gpar(fill = "#440154FF")),
                             grade = anno_barplot(grade, width = unit(10, "mm"),gp = gpar(fill = metadata(sce)$colors$grade)),
                             pT = anno_barplot(pT, width = unit(10, "mm"),gp = gpar(fill = metadata(sce)$colors$pT_simple)),
                             pN = anno_barplot(pN, width = unit(10, "mm"),gp = gpar(fill = metadata(sce)$colors$pN_simple)),
                             border = TRUE, 
                             annotation_name_rot = 90,
                             gap = unit(2,"mm"),
                             which = "row")


hb_tumor_scaled <- hb_tumor_scaled[match(rownames(anno),rownames(hb_tumor_scaled)),]
hb_TME_scaled <- hb_TME_scaled[match(rownames(anno),rownames(hb_TME_scaled)),]

###2.2 Row - Score features

# Expression of CK5,CK7, CD8a 
cluster_mean_sce <- aggregateAcrossCells(sce, ids = sce$BL_CD8_group_merg,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$clean_target %in% c("CK5","CK7","HLADR"))

exprs_out <- scale(t(assay(cluster_mean_sce, "exprs")))

# Number of B cell patches per PID 
n_Bpatch_PID <- colData(sce) %>% as.data.frame %>% select(PID, BL_CD8_group_merg, patch_B_cell) %>% 
  filter(!is.na(patch_B_cell)) %>% 
  unique() %>% 
  group_by(BL_CD8_group_merg, PID) %>% 
  count()

plot_df <- colData(sce) %>% as.data.frame %>% select(PID, BL_CD8_group_merg) %>% 
  unique() 
rownames(plot_df) <- NULL
plot_df$n_Bpatch <- n_Bpatch_PID$n[match(plot_df$PID, n_Bpatch_PID$PID)]
plot_df$n_Bpatch[is.na(plot_df$n_Bpatch)] <- 0
plot_df$BL_CD8_group_merg <- sce$BL_CD8_group_merg[match(plot_df$PID, sce$PID)]

# Proportion of B cell score per group
Bscore <- colData(sce) %>% as.data.frame() %>% select(BL_CD8_group_merg,B_cell_score) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
Bscore <- Bscore %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq)) 
Bscore <- Bscore %>% select(-Freq) %>% pivot_wider(id_cols = BL_CD8_group_merg,names_from = B_cell_score,values_from = fra) %>% column_to_rownames("BL_CD8_group_merg") 

B_out <- plot_df %>% pivot_wider(names_from = PID, values_from = n_Bpatch) %>% column_to_rownames("BL_CD8_group_merg")
B_out <- B_out[match(rownames(Bscore),rownames(B_out)),]

# Proportion of proliferation score per group
prolif_PID <- colData(sce) %>% as.data.frame() %>% select(PID,GMM_tumor_Ki67) %>% group_by(PID) %>% table() %>% as.data.frame
prolif_PID <- prolif_PID %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
prolif_PID$BL_CD8_group_merg <- sce$BL_CD8_group_merg[match(prolif_PID$PID, sce$PID)]
prolif_PID <- prolif_PID %>% filter(GMM_tumor_Ki67 == "Ki67+")

prolif_out <- prolif_PID %>% select(PID, fra, BL_CD8_group_merg) %>% pivot_wider(names_from = PID, values_from = fra) %>% column_to_rownames("BL_CD8_group_merg")
prolif_out <- prolif_out[match(rownames(Bscore),rownames(prolif_out)),]

# Proportion of CD8 T cell groups per group
CD8group <- colData(sce) %>% as.data.frame() %>% select(BL_CD8_group_merg,spatial_CD8_pheno) %>% group_by(BL_CD8_group_merg) %>% table() %>% as.data.frame
CD8group <- CD8group %>% group_by(BL_CD8_group_merg) %>% mutate(fra = Freq/sum(Freq)) 
CD8group <- CD8group %>% select(-Freq) %>% pivot_wider(id_cols = BL_CD8_group_merg,names_from = spatial_CD8_pheno,values_from = fra) %>% column_to_rownames("BL_CD8_group_merg") 

col_patch <- setNames(brewer.pal(4,"Paired"),colnames(Bscore))
col_CD8 <- c("steelblue1","navy","firebrick")

ha_score <- HeatmapAnnotation(#CD8_group = anno_barplot(CD8group, gp = gpar(fill = col_CD8),width = unit(10, "mm")),
  prolif_fraction = anno_boxplot(as.matrix(prolif_out), gp = gpar(fill = "pink"), which = "row",outline = FALSE,width = unit(15, "mm")),
  B_cell_score = anno_barplot(Bscore, gp = gpar(fill = col_patch),width = unit(10, "mm")),
  B_cell_patch = anno_boxplot(as.matrix(B_out), gp = gpar(fill = "#B2DF8A"), which = "row",outline = FALSE,width = unit(15, "mm")),
  border = TRUE, 
  annotation_name_rot = 90,
  gap = unit(2,"mm"),
  which = "row")

ha_score + NULL

### 3. Plot heatmap ###

order <- c("BL","L","B","N","inflamed")

h <- ha_group+
  Heatmap(hb_tumor_scaled,
        name = "Mean tumor MC fraction",
        show_row_dend = FALSE,
        clustering_method_rows = "ward.D2", 
        clustering_method_columns = "ward.D2",
        row_order = order,
        row_title = NULL)+
  Heatmap(hb_TME_scaled,
          name = "Mean TME MC fraction")+
  ha_meta+
  Heatmap(exprs_out, 
          name = "Mean expression")+
  ha_score

draw(h)

#Create legends for annotation 
lgd <- Legend(title = "Patient groups", at = names(metadata(sce)$colors$group_color), 
              legend_gp = gpar(fill = metadata(sce)$colors$group_color))

lgd_1 <- Legend(title = "Risk groups", at = names(metadata(sce)$colors$risk_color), 
              legend_gp = gpar(fill = metadata(sce)$colors$risk_color))

# Draw heatmap 
draw(h,annotation_legend_list = list(lgd, lgd_1))
```

# 3. Image visualization

```{r}
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_TNBC_new.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")
```

```{r}
# Select images of interest
cur_sample <- colData(sce) %>% as.data.frame %>% select(sample_id, PID_group_risk) %>% group_by(sample_id) %>% filter(PID_group_risk == "high") %>% count() %>% arrange(desc(n)) %>% pull(sample_id)

cur_images <- images[mcols(images)$sample_id %in% cur_sample]
cur_masks <- masks[mcols(masks)$sample_id %in% cur_sample]
cur_sce <- sce[,sce$sample_id %in% cur_sample]
```

```{r}
# Random images
cur_sample <- sample(unique(sce$sample_id), 30)

cur_images <- images[mcols(images)$sample_id %in% cur_sample]
cur_masks <- masks[mcols(masks)$sample_id %in% cur_sample]
cur_sce <- sce[,sce$sample_id %in% cur_sample]
```

We will use cytoviewer for image exploration and download

```{r}
app <- cytoviewer(image = cur_images, mask = cur_masks, object = cur_sce, img_id = "sample_id", cell_id = "ObjectNumber")

shiny::runApp(app)
```


```{r}
PID_groups_select <- unique(sce$PID_SC_groups)

registerDoParallel(cores = 7)

foreach(i = "4")%dopar%{

##Visualize clusters on images - select 10 images for each category with highest number of cells
cur_sample <- colData(sce) %>% as.data.frame %>% group_by(PID_SC_groups) %>% filter(PID_SC_groups == i) %>% pull(sample_id) %>% unique() #slice(1:10) %>% pull(sample_id)

#Prepare images
cur_images <- images[cur_sample]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))


#Subset sce
#cur_sce <- sce[,sce$PID_SC_groups == i]

#Plot 
plotPixels(cur_images,
           mask = NULL,
           object = sce,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           missing_colour = "white",
           outline_by = NULL,
           colour_by = c("CK5","CK7","CK8_18"),
           thick = FALSE,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = paste0("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/08_PIDgroup/PID_group_viz_",i,".png"), scale = 2)
           )
}

i = "inflamed"
##Visualize clusters on images - select 10 images for each category with highest number of cells
cur_sample <- colData(sce) %>% as.data.frame %>% group_by(spatial_CD8_pheno) %>% filter(spatial_CD8_pheno == i) %>% pull(sample_id) %>% unique() #slice(1:10) %>% pull(sample_id)

#Prepare images
cur_images <- images[cur_sample]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

#Subset sce
#cur_sce <- sce[,sce$PID_SC_groups == i]

#Plot 
plotPixels(cur_images,
           mask = NULL,
           object = sce,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           missing_colour = "white",
           outline_by = NULL,
           colour_by = c("panCK","HLADR","CD8a","CD20"),
           thick = FALSE,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           save_plot = list(filename = paste0("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/08_PIDgroup/PID_group_viz_",i,".png"), scale = 2)
           )
}

stopCluster()
```

# Save objects 

```{r}
#Save sce
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/08_sce_PIDgroup.rds")
```

# Session Info 

```{r}
sessionInfo()
```
