---
title: "09_validation_prediction"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: sce_ALL.rds, ML_ranger_JACKSON.rds, ML_ranger_SUB.rds, ML_ranger_ALL.rds 

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Script 9: Machine learning framework

Here, we will train machine learning classifiers to predict tumor metacluster in independent validation datasets. 

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scran)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(caret)
library(caretEnsemble)
library(doParallel)
library(tidyverse)
library(ggridges)
library(ggsci)
library(ggVennDiagram)
library(circlize)
library(imcdatasets)
library(survival)
library(survminer)
library(edgeR)
library(wesanderson)
library(gridExtra)
library(dendextend)
library(Rphenoannoy)
```

# 1. Read in data (train)

```{r}
##Load sce
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
sce <- sce[,sce$analysis_cat == "tumor"]
sce$metacluster_combined <- factor(unfactor(sce$metacluster_combined))
sce
```

## Filter for tumor cells

```{r}
## Add more rowData 
# All tumor cluster channels
rowData(sce)$clean_target[rowData(sce)$tumor_cluster_channel] 

# Sub-select important relevant markers
rowData(sce)$tumor_cluster_sub <- rowData(sce)$clean_target %in% c("CK5","CK7","KRT14","CK8_18","CD15","HLADR","Vimentin","pH3","Ki67","EGFR")

# Sub-select shared tumor markers from Jackson et al.
rowData(sce)$tumor_cluster_jana <- rowData(sce)$clean_target %in% c("H3K27me3", "CK5", "CK8_18", "CD68", "KRT14", "SMA", "Vimentin", "cMYC", "HER2", "pH3", "ER", "PR", "p53", "CD44", "CD45", "GATA3", "CD20", "CAIX", "Ki67", "EGFR", "S6", "p_mTOR", "CK7", "panCK", 
"c_Cas3_PARP")
```

# 2. Train the model

## Register backend 

```{r}
set.seed(22)
# Register parallel backend 
cl <- makePSOCKcluster(5, setup_strategy = "sequential")
registerDoParallel(cl)

#Set seeds for each iteration for parallel processing
num_folds <- 5
num_combinations <- 10  # Default is 10 for ranger
seeds <- vector(mode = "list", length = num_folds + 1)
for (i in 1:num_folds) {
    seeds[[i]] <- sample.int(5000, num_combinations)
}
seeds[[num_folds + 1]] <- sample.int(5000, 1)  # Final model seed
```

## Train/Test 

```{r}
# Randomly split into train and test data
set.seed(22)
trainIndex <- createDataPartition(sce$metacluster_combined, p = 0.75)
train_sce <- sce[,trainIndex$Resample1]
test_sce <- sce[,-trainIndex$Resample1]

colData(train_sce) %>% as.data.frame %>% group_by(metacluster_combined) %>% count() %>% ungroup %>% mutate(fra = n/sum(n))
colData(test_sce) %>% as.data.frame %>% group_by(metacluster_combined) %>% count() %>% ungroup %>% mutate(fra = n/sum(n))
```

## Random forest 

Ranger algorithm implementation

### All markers 

```{r}
# Train data
train_data <- t(assay(train_sce, "exprs")[rowData(train_sce)$tumor_cluster_channel,])

# Set folds for CV steps based on PID
folds <- groupKFold(train_sce$PID, k=5) 
#lapply(folds, function(x, y) table(y[x]), y = groups)

# Specify train parameters for 5-fold cross validation
fitControl <- trainControl(method = "cv",
                           number = 5,
                           seeds = seeds, 
                           index = folds)

# Calculate class weights
class_freq <- table(train_sce$metacluster_combined)
class_weights <- 1 / class_freq
class_weights <- class_weights / sum(class_weights) 
sample_weights <- class_weights[as.character(train_sce$metacluster_combined)]

# Train a random forest model for predicting cell labels
# This call also performs parameter tuning
rffit <- train(x = train_data, 
               y = train_sce$metacluster_combined,
               method = "ranger", 
               metric = 'Accuracy',
               tuneLength = 5,
               trControl = fitControl,
               importance = "impurity",
               #probability = TRUE,
               weights = sample_weights)

rffit
saveRDS(rffit, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_metacluster_pred_folds_allmarkers_withweights.rds")

on.exit(stopCluster(cl))
```

### Sub markers 

```{r}
set.seed(22)
# Register parallel backend 
cl <- makePSOCKcluster(5, setup_strategy = "sequential")
registerDoParallel(cl)
```

```{r}
# Train data
train_data <- t(assay(train_sce, "exprs")[rowData(train_sce)$tumor_cluster_sub,])

# Set folds for CV steps based on PID
folds <- groupKFold(train_sce$PID, k=5) 
#lapply(folds, function(x, y) table(y[x]), y = groups)

# Specify train parameters for 5-fold cross validation
fitControl <- trainControl(method = "cv",
                           number = 5,
                           seeds = seeds, 
                           index = folds)

# Calculate class weights
class_freq <- table(train_sce$metacluster_combined)
class_weights <- 1 / class_freq
class_weights <- class_weights / sum(class_weights) 
sample_weights <- class_weights[as.character(train_sce$metacluster_combined)]

# Train a random forest model for predicting cell labels
# This call also performs parameter tuning
rffit <- train(x = train_data, 
               y = train_sce$metacluster_combined,
               method = "ranger", 
               metric = 'Accuracy',
               tuneLength = 5,
               trControl = fitControl,
               importance = "impurity",
               #probability = TRUE,
               weights = sample_weights)

rffit

saveRDS(rffit, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_metacluster_pred_folds_sub_withweights.rds")

on.exit(stopCluster(cl))
```

### Jana markers 

```{r}
set.seed(22)
# Register parallel backend 
cl <- makePSOCKcluster(5, setup_strategy = "sequential")
registerDoParallel(cl)
```

```{r}
# Train data
train_data <- t(assay(train_sce, "exprs")[rowData(train_sce)$tumor_cluster_jana,])

# Set folds for CV steps based on PID
folds <- groupKFold(train_sce$PID, k=5) 
#lapply(folds, function(x, y) table(y[x]), y = groups)

# Specify train parameters for 5-fold cross validation
fitControl <- trainControl(method = "cv",
                           number = 5,
                           seeds = seeds, 
                           index = folds)

# Calculate class weights
class_freq <- table(train_sce$metacluster_combined)
class_weights <- 1 / class_freq
class_weights <- class_weights / sum(class_weights) 
sample_weights <- class_weights[as.character(train_sce$metacluster_combined)]

# Train a random forest model for predicting cell labels
# This call also performs parameter tuning
rffit <- train(x = train_data, 
               y = train_sce$metacluster_combined,
               method = "ranger", 
               metric = 'Accuracy',
               tuneLength = 5,
               trControl = fitControl, 
               importance = "impurity",
               #probability = TRUE,
               weights = sample_weights)

rffit

saveRDS(rffit, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_metacluster_pred_folds_jana_withweights.rds")

on.exit(stopCluster(cl))
```

```{r}
rffit <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_metacluster_pred_folds_allmarkers_withweights.rds")

rffit_sub <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_metacluster_pred_folds_sub_withweights.rds")

rffit_jana <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_metacluster_pred_folds_jana_withweights.rds")
```

## Assess feature importance

```{r}
# Classification accuracy during parameter tuning
ggplot(rffit) + 
  geom_errorbar(data = rffit$results,
                aes(ymin = Accuracy - AccuracySD,
                    ymax = Accuracy + AccuracySD),
                width = 0.4) +
    theme_classic(base_size = 15)

# Scaled feature importance (Decrease in Gini index)
plot(varImp(rffit))
plot(varImp(rffit_sub))
```


```{r}
out_var <- varImp(rffit, scale = TRUE)
out_var_1 <- varImp(rffit_sub, scale = TRUE)
#out_var_2 <- varImp(rffit_jana, scale = TRUE)

out_comb <- rbind(
  out_var$importance %>% rownames_to_column("marker") %>% mutate(category = "all_markers"),
  out_var_1$importance %>% rownames_to_column("marker") %>% mutate(category = "sub_markers")#,
  #out_var_2$importance %>% rownames_to_column("marker") %>% mutate(category = "jana_markers")
)


ggplot(out_comb, aes(x = reorder(marker, -Overall), y = Overall, fill = category))+
  facet_wrap(~category, scales = "free_x")+
  geom_col()+
  theme_classic()+
  scale_fill_brewer(palette = "Paired")+
  coord_flip()+
  ylab("Scaled feature importance")+
  xlab("Markers")
```


# 3. Predict on test data

```{r}
# Old without weights
#rffit_old <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/randomforest_metacluster_pred_folds_allmarkers.rds")
#rffit_sub_old <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/randomforest_metacluster_pred_folds_sub.rds")

# New with weights
rffit <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_metacluster_pred_folds_allmarkers_withweights.rds")
rffit_sub <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_metacluster_pred_folds_sub_withweights.rds")
rffit_jana <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ranger_metacluster_pred_folds_jana_withweights.rds")

#rffit_xg <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/XGboost_metacluster_pred_folds_allmarkers_withweights.rds")
```


```{r}
set.seed(22)

# Select test data
mat_test <- t(assay(test_sce, "exprs")[rowData(test_sce)$tumor_cluster_channel,])

# Predict cell phenotypes in test data
cur_pred <- predict(rffit, 
                    newdata = mat_test)

# Select test data
mat_test_sub <- t(assay(test_sce, "exprs")[rowData(test_sce)$tumor_cluster_sub,])

# Predict cell phenotypes in test data
cur_pred_sub <- predict(rffit_sub, 
                        newdata = mat_test_sub)

# Select test data
mat_test_jana <- t(assay(test_sce, "exprs")[rowData(test_sce)$tumor_cluster_jana,])

# Predict cell phenotypes in test data
cur_pred_jana <- predict(rffit_jana, 
                        newdata = mat_test_jana)
```

## Assess classifier performance

```{r}
cm <- confusionMatrix(data = cur_pred, 
                      reference = test_sce$metacluster_combined, 
                      mode = "everything")

cm

cm_stats <- as.data.frame(cm$byClass)
mean(cm_stats$`Balanced Accuracy`) 
mean(cm_stats$F1)# Report this
```

```{r}
cm_sub <- confusionMatrix(data = cur_pred_sub, 
                      reference = test_sce$metacluster_combined, 
                      mode = "everything")

cm_sub

cm_sub_stats <- as.data.frame(cm_sub$byClass)
mean(cm_sub_stats$`Balanced Accuracy`) 
mean(cm_sub_stats$F1)# Report this
```

```{r}
cm_jana <- confusionMatrix(data = cur_pred_jana, 
                      reference = test_sce$metacluster_combined, 
                      mode = "everything")
cm_jana

cm_jana_stats <- as.data.frame(cm_jana$byClass)
mean(cm_jana_stats$`Balanced Accuracy`) 
mean(cm_jana_stats$F1)# Report this
```


```{r}
df_out <- 
  rbind(
    data.frame(cm$byClass) %>%
      mutate(class = sub("Class: ", "", rownames(cm$byClass))) %>% mutate(category = "all_markers"),
    data.frame(cm_sub$byClass) %>%
      mutate(class = sub("Class: ", "", rownames(cm_sub$byClass))) %>% mutate(category = "b_sub_markers"),
    data.frame(cm_jana$byClass) %>%
      mutate(class = sub("Class: ", "", rownames(cm_jana$byClass))) %>% mutate(category = "jana_markers")
    )

ggplot(df_out) + 
  facet_wrap(~category) +
  geom_point(aes(1-Specificity, Sensitivity, 
                 size = Detection.Rate,
                 fill = class),
             shape = 21) + 
  scale_fill_manual(values = metadata(sce)$colors$metacluster_combined, limits = force) +
  theme_classic(base_size = 15) + 
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")+
  ggtitle(paste0("Accuracy (all): ",round(cm$overall[1],3)," [",round(cm$overall[3],3)," - ",round(cm$overall[4],3),"]", " Accuracy (sub): ",round(cm_sub$overall[1],3)," [",round(cm_sub$overall[3],3)," - ",round(cm_sub$overall[4],3),"]"))#+
  #coord_cartesian(ylim = c(0,1))

```

```{r}
data.frame(cm_jana$byClass) %>%
  mutate(class = sub("Class: ", "", rownames(cm_jana$byClass))) %>%
  ggplot() + 
  geom_point(aes(1-Specificity, Sensitivity, 
                 size = Detection.Rate,
                 fill = class),
             shape = 21) + 
  scale_fill_manual(values = metadata(sce)$colors$metacluster_combined, limits = force) +
  theme_classic(base_size = 15) + 
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")+
  ggtitle(paste0("Accuracy: ",round(cm_jana$overall[1],3)," [",round(cm_jana$overall[3],3)," - ",round(cm_jana$overall[4],3),"]"))
```

# 4. Read and predict on independent dataset 

## Fischer et al. - Processed with discovery

First, we will read in the `SingleCellExperiment` object containing the cleaned, spillover-corrected single-cell data. 
**Note:** We pre-processed the main cohort together with the validation cohort (including improved segmentation) to avoid strong batch effects. 
We also trained another random forest classifier to predict tumor metaclusters from the original study and use them here for comparison. 

```{r}
sce_fischer <- readRDS("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/ML_sce_fischer_prediction.rds")
```

### Cohort overview 

```{r}
sce_fischer$n_samples <-1
sce_fischer$OS_data <- !is.na(sce_fischer$OS_months)
sce_fischer$DFS_data <- FALSE

meta_fischer_clean <- colData(sce_fischer) %>% as.data.frame %>% select("PID","n_samples","grade","pT_simple","pN_simple","pM_simple","OS_data","DFS_data") %>% unique()
```

```{r}
## Metadata heatmap 

#Filter for columns to visualize
Meta_viz <- meta_fischer_clean
#Convert empty columns and NAs for easier handling with ComplexHeatmap package
Meta_viz[is.na(Meta_viz)] <- "Unknown"

Meta_viz <- Meta_viz %>% arrange(desc(pT_simple),desc(pN_simple), desc(grade))

## Set Color scheme (also for downstream analysis)
color_vectors <- list(
PID = setNames(colorRampPalette(brewer.pal(11, "BrBG"))(length(unique(Meta_viz$PID))), 
                sort(unique(Meta_viz$PID))),
n_samples = setNames(c("blue","navy"),c("1", "2")),
grade = setNames(c("#FEE6CE", "#FDC08C", "#F5904B", "#E6550D"), c("1", "2", "3", "4")
),
pT_simple = setNames(colorRampPalette(brewer.pal(9, "Blues"))(length(unique(Meta_viz$pT_simple))), 
                sort(unique(Meta_viz$pT_simple))),
pN_simple = setNames(c(colorRampPalette(brewer.pal(9, "Purples"))(length(unique(Meta_viz$pN_simple))-1),"black"), 
                sort(unique(Meta_viz$pN_simple))),
pM_simple = setNames(colorRampPalette(brewer.pal(2, "RdPu"))(length(unique(Meta_viz$pM_simple))), 
                sort(unique(Meta_viz$pM_simple))),
DFS_data = setNames("black","FALSE"),
OS_data = setNames(c("grey","black"), c("TRUE", "FALSE"))
)

#Save color list object for downstream usage
#saveRDS(col_list_1,file="/Users/lasmey/Desktop/RCC/Metadaten/color.rds")

#Heatmap Annotation
sample_anno <- HeatmapAnnotation(PID = Meta_viz$PID,
                                 n_samples = Meta_viz$n_samples,
                                 grade = Meta_viz$grade,
                                 pT_simple = Meta_viz$pT_simple,
                                pN_simple = Meta_viz$pN_simple,
                                pM_simple = Meta_viz$pM_simple,
                                OS_data = Meta_viz$OS_data,
                                DFS_data = Meta_viz$DFS_data,
                                col = color_vectors,
                                border = T,
                                na_col = "black",
                                gap = unit(5, "points"),
                                show_annotation_name = T,
                                show_legend = c(PID = FALSE)
)%v%NULL

#Display plot
draw(sample_anno)
```


```{r}
# Compare sets of markers for clustering of epithelial cells 
intersect(rownames(sce_fischer)[rowData(sce_fischer)$tumor_cluster_channel],
rownames(sce_fischer)[rowData(sce_fischer)$tumor_cluster_fischer])

setdiff(rownames(sce_fischer)[rowData(sce_fischer)$tumor_cluster_channel],
rownames(sce_fischer)[rowData(sce_fischer)$tumor_cluster_fischer])

setdiff(rownames(sce_fischer)[rowData(sce_fischer)$tumor_cluster_fischer], rownames(sce_fischer)[rowData(sce_fischer)$tumor_cluster_channel])

ggVennDiagram(list(rownames(sce_fischer)[rowData(sce_fischer)$tumor_cluster_channel], rownames(sce_fischer)[rowData(sce_fischer)$tumor_cluster_fischer]),
              category.names = c("Meyer et al.", "Fischer et al."),
              label_alpha = 0, 
              set_color = "black")+
  scale_fill_gradient(low = "white", high = "grey")+
  scale_color_manual(values = c("black","black","black"))
```

### Full marker set

```{r}
# Select validation expression matrix
mat_valid <- t(assay(sce_fischer, "exprs")[rowData(sce_fischer)$tumor_cluster_channel,])

# Predict metacluster
set.seed(22)
cell_class <- as.character(predict.train(rffit, 
                                         newdata = mat_valid, 
                                         type = "raw"))

names(cell_class) <- rownames(mat_valid)

table(cell_class)
```

Add predicted metacluster to sce and save 

```{r}
# Store labels in SCE object
sce_fischer$metacluster_predict <- factor(cell_class[match(sce_fischer$cell_id, names(cell_class))])

cur_table <- table(sce_fischer$metacluster_predict, sce_fischer$PID)
cur_table
```

### Sub marker set 

```{r}
# Select validation expression matrix
mat_valid <- t(assay(sce_fischer, "exprs")[rowData(sce_fischer)$tumor_cluster_sub,])

# Predict metacluster
set.seed(22)
cell_class <- as.character(predict.train(rffit_sub, 
                                         newdata = mat_valid, 
                                         type = "raw"))

names(cell_class) <- rownames(mat_valid)

table(cell_class)
```

Add predicted metacluster to sce and save 

```{r}
# Store labels in SCE object
sce_fischer$metacluster_predict_sub <- factor(cell_class[match(sce_fischer$cell_id, names(cell_class))])

cur_table <- table(sce_fischer$metacluster_predict_sub, sce_fischer$PID)
cur_table
```


#### Visualization - chordDiagram 

```{r}
## Test for association of categories 

# Extract levels of metacluster_predict and metacluster
metacluster_predict_levels <- unique(sce_fischer$metacluster_predict)
metacluster_levels <- unique(sce_fischer$tumor_metacluster_fischer)

# Create all pairwise combinations of levels between metacluster_predict and metacluster
pairwise_combinations <- expand.grid(metacluster_predict = metacluster_predict_levels,
                                     metacluster = metacluster_levels)

pair <- pairwise_combinations[1,]

# Function to generate 2x2 table and perform Fisher's exact test
pairwise_2x2_fisher <- function(pair) {
  # Create logical vectors for A/notA and B/notB
  A <- sce_fischer$metacluster_predict == pair$metacluster_predict
  B <- sce_fischer$tumor_metacluster_fischer == pair$metacluster
  
  # Construct the 2x2 contingency table
  contingency_table <- matrix(c(sum(A & B),    # A and B
                                sum(A & !B),   # A and notB
                                sum(!A & B),   # notA and B
                                sum(!A & !B)), # notA and notB
                              nrow = 2, byrow = TRUE)
  
  # Perform Fisher's exact test
  fisher_test <- fisher.test(contingency_table)
  
  # Return pair, p-value, and table
  return(data.frame(pair = paste0(pair$metacluster_predict, "__", pair$metacluster),
                    pair_1 = pair$metacluster_predict, pair_2 = pair$metacluster,
                    p_value = fisher_test$p.value, odds_ratio = fisher_test$estimate))
}

# Apply the function to all pairwise combinations
results <- do.call(rbind, apply(pairwise_combinations, 1, function(row) pairwise_2x2_fisher(as.list(row))))

# Adjust p-values using FDR
results$adjusted_p_value <- p.adjust(results$p_value, method = "fdr")

# Print results
print(results)

# Odds ratio threshold
#results$odds_high <- results$odds_ratio > 10
#pair_sig <- results %>% filter(odds_high) %>% pull(pair)

# OR TOP10
pair_sig <- (results %>% arrange(desc(odds_ratio)) %>% pull(pair))[1:10]
```


```{r}
# Visualize in chordDiagram

# Define colors
col_MC <- metadata(sce)$colors$metacluster_combined 
col_MC_Fischer <- setNames(rep("grey", length(unique(sce_fischer$tumor_metacluster_fischer))),unique(sce_fischer$tumor_metacluster_fischer))
col_full <- c(col_MC, col_MC_Fischer)

# Define number of cells per category
anno <- rbind(
  colData(sce_fischer) %>% as.data.frame %>%
    group_by(tumor_metacluster_fischer) %>% count() %>%
    ungroup() %>% transmute(type = tumor_metacluster_fischer, n),
  colData(sce_fischer) %>% as.data.frame %>%
    group_by(metacluster_predict) %>% count() %>%
    ungroup() %>% transmute(type = metacluster_predict, n)
)


# Plot as chordDiagram
df <- as.data.frame(table(sce_fischer$metacluster_predict, sce_fischer$tumor_metacluster_fischer))
df$pair <- paste0(df$Var1, "__", df$Var2)
df$sig_pair <- df$pair %in% pair_sig

## Not scaled
chordDiagram(df, scale = FALSE, grid.col = col_full, link.border = df$sig_pair)
circos.clear()

## Scaled
chordDiagram(df, scale = TRUE, grid.col = col_full, link.border = df$sig_pair)
circos.clear()
```

## Jackson et al. 

```{r}
sce_jackson <- imcdatasets::JacksonFischer_2020_BreastCancer(data_type = c("sce"), full_dataset = TRUE)
```

```{r}
# Subset for tumor metacluster
sce_jackson <- sce_jackson[,sce_jackson$cell_metacluster %in% c(14:27)]
```

```{r}
# Adjust rowData
rowData(sce_jackson)$target_match <- c("Ru96", "Ru98", "Ru99", "Ru100", "Ru101", "Ru102", "Ru104", 
"H3", "H3K27me3", "CK5", "FN1", "KRT19", "CK8_18", "TWIST1", 
"CD68", "KRT14", "SMA", "Vimentin", "cMYC", "HER2", "CD3e", "pH3", 
"SNAI2", "ER", "PR", "p53", "CD44", "EpCAM", "CD45", "GATA3", 
"CD20", "CTNNB", "CAIX", "CDH1", "Ki67", "EGFR", "S6", "SOX9", 
"vWF", "p_mTOR", "CK7", "panCK", "c_Cas3_PARP", "DNA1", "DNA2"
)

# Identify shared tumor markers between our dataset and Jackson et al.
rowData(sce_jackson)$target_match[rowData(sce_jackson)$target_match %in% rowData(sce)$clean_target]

rownames(sce_jackson) <- rowData(sce_jackson)$target_match
```

```{r}
# Give metacluster names
unique(sce_jackson$cell_metacluster)

sce_jackson$tumor_metacluster_jackson <- sce_jackson$cell_metacluster
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(14)] <- "Hypoxic"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(15)] <- "Apoptotic"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(16)] <- "Proliferative"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(17)] <- "p53_pos_EGFR_pos"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(18)] <- "Basal_CKs"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(19)] <- "CK7pos_CKhi_Cadhi"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(20)] <- "CK7pos_CKhi"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(21)] <- "Epithelial_low"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(22)] <- "CKlow_HRlow"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(23)] <- "CKpos_HRhi"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(24)] <- "CKpos_HRpos"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(25)] <- "CKpos_HRlow"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(26)] <- "CKlow_HRhi_p53pos"
sce_jackson$tumor_metacluster_jackson[sce_jackson$tumor_metacluster_jackson %in% c(27)] <- "Myoepithelial"

unique(sce_jackson$tumor_metacluster_jackson)
length(unique(sce_jackson$tumor_metacluster_jackson)) #14 Metacluster
```

```{r}
## TNM staging 
sce_jackson$pT_simple <- str_extract(sce_jackson$tumor_PTNM_T, "[:digit:]")
sce_jackson$pN_simple <- str_extract(sce_jackson$tumor_PTNM_N, "[:digit:]")
sce_jackson$pM_simple <- str_extract(sce_jackson$tumor_PTNM_M, "[:digit:]")

## DFS status 
sce_jackson$status_DFS <- ifelse(sce_jackson$patient_status %in% c("death","death by primary disease", "alive w metastases"), 1, 0)
sce_jackson$status_OS <- ifelse(sce_jackson$patient_status %in% c("death","death by primary disease"), 1, 0)
sce_jackson$DFS_months <- sce_jackson$patient_DFS_months
sce_jackson$OS_months <- sce_jackson$patient_OS_months
```

```{r}
## Filter for TNBC AND tumor cases
sce_jackson <- sce_jackson[, sce_jackson$tumor_clinical_type == "TripleNeg"]
sce_jackson <- sce_jackson[, sce_jackson$patient_disease_status %in% c("tumor")]
```

### Cohort overview 

```{r}
n_images <- colData(sce_jackson) %>% as.data.frame %>% select(patient_id, image_name, patient_cohort) %>% group_by(patient_cohort) %>% unique() %>% count(patient_id)

# Add metadata
sce_jackson$n_samples <- n_images$n[match(sce_jackson$patient_id, n_images$patient_id)]
sce_jackson$OS_data <- !is.na(sce_jackson$OS_months)
sce_jackson$DFS_data <- !is.na(sce_jackson$DFS_months)
sce_jackson$grade <- sce_jackson$tumor_grade
sce_jackson$PID <- sce_jackson$patient_id

meta_jackson_clean <- colData(sce_jackson) %>% as.data.frame %>% select("PID","n_samples","grade","pT_simple","pN_simple","pM_simple","OS_data","DFS_data") %>% unique()
```

```{r}
## Metadata heatmap 

#Filter for columns to visualize
Meta_viz <- meta_jackson_clean
#Convert empty columns and NAs for easier handling with ComplexHeatmap package
Meta_viz[is.na(Meta_viz)] <- "Unknown"

Meta_viz <- Meta_viz %>% arrange(desc(pT_simple),desc(pN_simple), desc(grade))

## Set Color scheme (also for downstream analysis)
color_vectors <- list(
PID = setNames(colorRampPalette(brewer.pal(11, "BrBG"))(length(unique(Meta_viz$PID))), 
                sort(unique(Meta_viz$PID))),
n_samples = setNames(c("blue","navy"),c("1", "2")),
grade = setNames(c("#FEE6CE", "#FDC08C", "#F5904B", "#E6550D"), c("1", "2", "3", "4")
),
pT_simple = setNames(colorRampPalette(brewer.pal(9, "Blues"))(length(unique(Meta_viz$pT_simple))), 
                sort(unique(Meta_viz$pT_simple))),
pN_simple = setNames(c(colorRampPalette(brewer.pal(9, "Purples"))(length(unique(Meta_viz$pN_simple))-1),"black"), 
                sort(unique(Meta_viz$pN_simple))),
pM_simple = setNames(colorRampPalette(brewer.pal(2, "RdPu"))(length(unique(Meta_viz$pM_simple))), 
                sort(unique(Meta_viz$pM_simple))),
OS_data = setNames("grey","TRUE"),
DFS_data = setNames("grey","TRUE")
)

#Save color list object for downstream usage
#saveRDS(col_list_1,file="/Users/lasmey/Desktop/RCC/Metadaten/color.rds")

#Heatmap Annotation
sample_anno <- HeatmapAnnotation(PID = Meta_viz$PID,
                                 n_samples = Meta_viz$n_samples,
                                 grade = Meta_viz$grade,
                                 pT_simple = Meta_viz$pT_simple,
                                pN_simple = Meta_viz$pN_simple,
                                pM_simple = Meta_viz$pM_simple,
                                OS_data = Meta_viz$OS_data,
                                DFS_data = Meta_viz$DFS_data,
                                col = color_vectors,
                                border = T,
                                na_col = "black",
                                gap = unit(5, "points"),
                                show_annotation_name = T,
                                show_legend = c(PID = FALSE)
)%v%NULL

#Display plot
draw(sample_anno)
```

```{r}
# Compare sets of markers for clustering of epithelial cells 
rowData(sce_jackson)$tumor_cluster_og <- !str_detect(rownames(sce_jackson), "Ru")

dput(intersect(rownames(sce)[rowData(sce)$tumor_cluster_channel], rownames(sce_jackson)[rowData(sce_jackson)$tumor_cluster_og]))

setdiff(rownames(sce)[rowData(sce)$tumor_cluster_channel], rownames(sce_jackson)[rowData(sce_jackson)$tumor_cluster_og])

dput(setdiff(rownames(sce_jackson)[rowData(sce_jackson)$tumor_cluster_og],rownames(sce)[rowData(sce)$tumor_cluster_channel]))

ggVennDiagram(list(rownames(sce)[rowData(sce)$tumor_cluster_channel], rownames(sce_jackson)[rowData(sce_jackson)$tumor_cluster_og]),
              category.names = c("Meyer et al.", "Jackson et al."),
              label_alpha = 0, 
              set_color = "black")+
  scale_fill_gradient(low = "white", high = "grey")+
  scale_color_manual(values = c("black","black","black"))
```

### Jackson marker set 

```{r}
# Select validation expression matrix
rowData(sce_jackson)$tumor_cluster_jana <- rowData(sce_jackson)$target_match %in% rownames(sce)

mat_valid <- t(assay(sce_jackson, "exprs")[rowData(sce_jackson)$tumor_cluster_jana,])

# Predict metacluster
set.seed(2201)
cell_class <- as.character(predict.train(rffit_jana, 
                                         newdata = mat_valid, 
                                         type = "raw"))

names(cell_class) <- rownames(mat_valid)

table(cell_class)
```

Add predicted metacluster to sce and save 

```{r}
# Store labels in SCE object
sce_jackson$metacluster_predict <- factor(cell_class[match(sce_jackson$cell_id, names(cell_class))])

cur_table <- table(sce_jackson$metacluster_predict, sce_jackson$patient_id)
cur_table
```

#### Visualization - chordDiagram

```{r}
## Test for association of categories 

# Extract levels of metacluster_predict and metacluster
metacluster_predict_levels <- unique(sce_jackson$metacluster_predict)
metacluster_levels <- unique(sce_jackson$tumor_metacluster_jackson)

# Create all pairwise combinations of levels between metacluster_predict and metacluster
pairwise_combinations <- expand.grid(metacluster_predict = metacluster_predict_levels,
                                     metacluster = metacluster_levels)

pair <- pairwise_combinations[1,]

# Function to generate 2x2 table and perform Fisher's exact test
pairwise_2x2_fisher <- function(pair) {
  # Create logical vectors for A/notA and B/notB
  A <- sce_jackson$metacluster_predict == pair$metacluster_predict
  B <- sce_jackson$tumor_metacluster_jackson == pair$metacluster
  
  # Construct the 2x2 contingency table
  contingency_table <- matrix(c(sum(A & B),    # A and B
                                sum(A & !B),   # A and notB
                                sum(!A & B),   # notA and B
                                sum(!A & !B)), # notA and notB
                              nrow = 2, byrow = TRUE)
  
  # Perform Fisher's exact test
  fisher_test <- fisher.test(contingency_table)
  
  # Return pair, p-value, and table
  return(data.frame(pair = paste0(pair$metacluster_predict, "__", pair$metacluster),
                    pair_1 = pair$metacluster_predict, pair_2 = pair$metacluster,
                    p_value = fisher_test$p.value, odds_ratio = fisher_test$estimate))
}

# Apply the function to all pairwise combinations
results <- do.call(rbind, apply(pairwise_combinations, 1, function(row) pairwise_2x2_fisher(as.list(row))))

# Adjust p-values using FDR
results$adjusted_p_value <- p.adjust(results$p_value, method = "fdr")

# Print results
print(results)

# Odds ratio threshold
#results$odds_high <- results$odds_ratio > 10
#pair_sig <- results %>% filter(odds_high) %>% pull(pair)

# OR TOP10
pair_sig <- (results %>% arrange(desc(odds_ratio)) %>% pull(pair))[1:10]
```


```{r}
# Visualize in chordDiagram

# Define colors
col_MC <- metadata(sce)$colors$metacluster_combined 
col_MC_jackson <- setNames(rep("grey", length(unique(sce_jackson$tumor_metacluster_jackson))),unique(sce_jackson$tumor_metacluster_jackson))
col_full <- c(col_MC, col_MC_jackson)

# Define number of cells per category
anno <- rbind(
  colData(sce_jackson) %>% as.data.frame %>%
    group_by(tumor_metacluster_jackson) %>% count() %>%
    ungroup() %>% transmute(type = tumor_metacluster_jackson, n),
  colData(sce_jackson) %>% as.data.frame %>%
    group_by(metacluster_predict) %>% count() %>%
    ungroup() %>% transmute(type = metacluster_predict, n)
)


# Plot as chordDiagram
df <- as.data.frame(table(sce_jackson$metacluster_predict, sce_jackson$tumor_metacluster_jackson))
df$pair <- paste0(df$Var1, "__", df$Var2)
df$sig_pair <- df$pair %in% pair_sig

## Not scaled
chordDiagram(df, scale = FALSE, grid.col = col_full, link.border = df$sig_pair)
circos.clear()

## Scaled
chordDiagram(df, scale = TRUE, grid.col = col_full, link.border = df$sig_pair)
circos.clear()
```

## Save objects 

```{r}
# Save sce objects after prediction
#saveRDS(sce_fischer,"/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/ML_sce_fischer_prediction.rds")
#sce_fischer <- readRDS("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/ML_sce_fischer_prediction.rds")
#saveRDS(sce_jackson,"/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/ML_sce_jackson_prediction.rds")
```


# 5. Result analysis 

## Comparison discovery and validation 

Compare metacluster fractions between discovery and validation cohort 

```{r}
# A: Fischer et al 
discovery <- colData(sce) %>% as.data.frame %>% select(metacluster_combined) %>% count(metacluster_combined) %>% mutate(fra = n/sum(n))

validation <- colData(sce_fischer) %>% as.data.frame %>% select(metacluster_predict) %>% count(metacluster_predict) %>% mutate(fra = n/sum(n)) 

discovery$fra_valid <- validation$fra[match(discovery$metacluster_combined, validation$metacluster_predict)]

cor(discovery$fra, discovery$fra_valid, method = "spearman")

spearman_test <- cor.test(discovery$fra, discovery$fra_valid, method = "spearman")

ggplot(discovery, aes(x = fra, y = fra_valid))+
  geom_smooth(method='lm', formula= y~x, se = TRUE, color = "darkgrey")+
  geom_point(aes(color = metacluster_combined), size = 3)+
  scale_color_manual(values=metadata(sce)$colors$metacluster_combined, limits = force)+
  ylab("Predicted MC fraction (Fischer et al., 2023)")+
  xlab("MC fraction (Discovery)")+
  theme_classic()+
  ggtitle(paste("Spearman rho = ",round(spearman_test$estimate, digits = 2)))
```

```{r}
# B: Jackson
discovery <- colData(sce) %>% as.data.frame %>% select(metacluster_combined) %>% count(metacluster_combined) %>% mutate(fra = n/sum(n))

validation <- colData(sce_jackson) %>% as.data.frame %>% select(metacluster_predict) %>% count(metacluster_predict) %>% mutate(fra = n/sum(n)) 

discovery$fra_valid <- validation$fra[match(discovery$metacluster_combined, validation$metacluster_predict)]

cor(discovery$fra, discovery$fra_valid, method = "spearman")

spearman_test <- cor.test(discovery$fra, discovery$fra_valid, method = "spearman")

ggplot(discovery, aes(x = fra, y = fra_valid))+
  geom_smooth(method='lm', formula= y~x, se = TRUE, color = "darkgrey")+
  geom_point(aes(color = metacluster_combined), size = 3)+
  scale_color_manual(values=metadata(sce)$colors$metacluster_combined)+
  ylab("Predicted MC fraction (Jackson et al., 2020)")+
  xlab("MC fraction (Discovery)")+
  theme_classic()+
  ggtitle(paste("Spearman rho = ",round(spearman_test$estimate, digits = 2)))
```

Compare mean expression profiles between discovery and validation cohort

```{r}
# Fischer
discovery_exprs <- aggregateAcrossCells(sce, ids = sce$metacluster_combined,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$tumor_cluster_channel)

validation_exprs <- aggregateAcrossCells(sce_fischer, ids = sce_fischer$metacluster_predict,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_fischer)$tumor_cluster_channel)

cor(assay(discovery_exprs, "exprs"),assay(validation_exprs, "exprs"), method = "spearman")

corrplot(cor(assay(discovery_exprs, "exprs"),assay(validation_exprs, "exprs")), method="color", col=brewer.pal(n=8, name="RdBu"),  
         type="upper", order="hclust",hclust.method = "ward.D2", 
         addCoef.col = "black", addCoefasPercent = F,
         tl.col="black", tl.srt=40,
         diag=TRUE)
```

````{r}
# Jackson
discovery_exprs <- aggregateAcrossCells(sce, ids = sce$metacluster_combined,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$tumor_cluster_jana)

validation_exprs <- aggregateAcrossCells(sce_jackson, ids = sce_jackson$metacluster_predict,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_jackson)$tumor_cluster_jana)

cor(assay(discovery_exprs, "exprs"),assay(validation_exprs, "exprs"), method = "spearman")

corrplot(cor(assay(discovery_exprs, "exprs"),assay(validation_exprs, "exprs")), method="color", col=brewer.pal(n=8, name="RdBu"),  
         type="upper", order="hclust",hclust.method = "ward.D2", 
         addCoef.col = "black", addCoefasPercent = F,
         tl.col="black", tl.srt=40,
         diag=TRUE)
```


```{r}
discovery <- colData(sce) %>% as.data.frame %>% select(metacluster_combined) %>% count(metacluster_combined) %>% mutate(fra = n/sum(n))

validation <- colData(sce_jana) %>% as.data.frame %>% select(metacluster_predict) %>% count(metacluster_predict) %>% mutate(fra = n/sum(n)) 
discovery$fra_valid <- validation$fra[match(discovery$metacluster_combined, validation$metacluster_predict)]

cor(discovery$fra, discovery$fra_valid, method = "spearman")

spearman_test <- cor.test(discovery$fra, discovery$fra_valid, method = "spearman")

ggplot(discovery, aes(x = fra, y = fra_valid))+
  geom_smooth(method='lm', formula= y~x, se = TRUE, color = "darkgrey")+
  geom_point(aes(color = metacluster_combined), size = 3)+
  scale_color_manual(values=metadata(sce)$colors$metacluster_combined)+
  ylab("Predicted MC fraction (Fischer et al., 2023)")+
  xlab("MC fraction (Discovery)")+
  theme_classic()+
  ggtitle(paste("Spearman rho = ",round(spearman_test$estimate, digits = 2)))
```


```{r}
discovery_exprs <- aggregateAcrossCells(sce, ids = sce$metacluster_combined,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$tumor_cluster_jana)

validation_exprs <- aggregateAcrossCells(sce_jana, ids = sce_jana$metacluster_predict,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_jana)$tumor_cluster_jana)

cor(assay(discovery_exprs, "exprs"),assay(validation_exprs, "exprs"), method = "spearman")

corrplot(cor(assay(discovery_exprs, "exprs"),assay(validation_exprs, "exprs")), method="color", col=brewer.pal(n=8, name="RdBu"),  
         type="upper", order="hclust",hclust.method = "ward.D2", 
         addCoef.col = "black", addCoefasPercent = F,
         tl.col="black", tl.srt=40,
         diag=TRUE)
```


```{r}
# 2. Metacluster
for_plot_MC <- colData(sce_val) %>% as_tibble() %>%
    group_by(PID, metacluster_predict) %>%
    summarize(count = n()) %>%
    mutate(log_count = log(count)) %>%
    pivot_wider(id_cols = PID, names_from = metacluster_predict, 
                values_from = log_count, values_fill = 0) %>%
    ungroup() %>%
    select(-PID)

Heatmap(for_plot_MC, 
        name = "log(cell_count)",
        col = viridis(100)
        )
```


## Heatmap visualization 

```{r}
#Aggregrate across cells with the mean (median is also possible)
cluster_mean_sce <- aggregateAcrossCells(sce_val, ids = sce_val$metacluster_predict,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce)$tumor_cluster_channel)

col_exprs <- colorRamp2(c(0,1,2,3), c("#440154FF", "#31688EFF", "#35B779FF", "#FDE725FF"))

anno <- colData(cluster_mean_sce) %>% as.data.frame %>% select(metacluster_predict,ncells)

# Proportion of grade per cluster
grade <- colData(sce_val) %>% as.data.frame() %>% select(metacluster_predict,grade) %>% group_by(metacluster_predict) %>% table() %>% as.data.frame
grade <- grade %>% group_by(metacluster_predict) %>% mutate(fra = Freq/sum(Freq)) 
grade <- grade %>% select(-Freq) %>% pivot_wider(id_cols = metacluster_predict,names_from = grade,values_from = fra) %>% column_to_rownames("metacluster_predict")

# Proportion of pT stage per cluster
pT <- colData(sce_val) %>% as.data.frame() %>% select(metacluster_predict,pT_simple) %>% group_by(metacluster_predict) %>% table() %>% as.data.frame
pT <- pT %>% group_by(metacluster_predict) %>% mutate(fra = Freq/sum(Freq)) 
pT <- pT %>% select(-Freq) %>% pivot_wider(id_cols = metacluster_predict,names_from = pT_simple,values_from = fra) %>% column_to_rownames("metacluster_predict")

# Proportion of pN stage per cluster
pN <- colData(sce_val) %>% as.data.frame() %>% select(metacluster_predict,pN_simple) %>% group_by(metacluster_predict) %>% table() %>% as.data.frame
pN <- pN %>% group_by(metacluster_predict) %>% mutate(fra = Freq/sum(Freq)) 
pN <- pN %>% select(-Freq) %>% pivot_wider(id_cols = metacluster_predict,names_from = pN_simple,values_from = fra) %>% column_to_rownames("metacluster_predict")

# Number of contributing patients
cluster_PID <- colData(sce_val) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(metacluster_predict) %>% table() %>% as.data.frame
n_PID <- cluster_PID %>% filter(Freq>0) %>% group_by(metacluster_predict) %>% count(name = "n_PID") %>% column_to_rownames("metacluster_predict")

# Metadata annotation
ha_meta <- HeatmapAnnotation(metacluster_predict = anno_simple(unfactor(anno$metacluster_predict), pch = str_split(unfactor(anno$metacluster_predict),"_",simplify = TRUE)[,2],border=TRUE,col=metadata(sce)$colors$metacluster_combined),
                            n_cells = anno_barplot(anno$ncells, width = unit(10, "mm"),gp = gpar(fill = "#440154FF")),
                            n_PID = anno_barplot(n_PID, width = unit(10, "mm"),gp = gpar(fill = "#440154FF")),
                            #grade = anno_barplot(grade, width = unit(10, "mm"),gp = gpar(fill = metadata(sce_val)$colors$grade)),
                            #pT = anno_barplot(pT, width = unit(10, "mm"),gp = gpar(fill = metadata(sce_val)$colors$pT_simple)),
                            #pN = anno_barplot(pN, width = unit(10, "mm"),gp = gpar(fill = metadata(sce_val)$colors$pN_simple)),
                            #cluster_cat = anno_barplot(cluster_CC, width = unit(10, "mm"),gp = gpar(fill = metadata(sce_val)$colors$cluster_category)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(1,"mm"),
                            which = "row")

Heatmap(t(assay(cluster_mean_sce, "exprs")),
        column_title = "pred_channels",
        col = col_exprs,
        name = "mean exprs",
        row_title = NULL, 
        show_column_names =  TRUE, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2",
        row_split = anno$metacluster_predict,
        cluster_row_slices = FALSE)+
  ha_meta

mat_scaled <- as.data.frame(scale(t(assay(cluster_mean_sce, "exprs"))))

Heatmap(mat_scaled,
        column_title = "pred_channels",
        #col = col_exprs,
        name = "mean exprs",
        row_title = NULL, 
        show_column_names =  TRUE, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2",
        #row_split = anno$metacluster_predict,
        cluster_row_slices = FALSE)+
  ha_meta
```

## Survival - Cox / Kaplan 

```{r}
# A: Fischer et al. - Multivariate
PID_metacluster_predict <- colData(sce_fischer) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster_predict <- PID_metacluster_predict %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_metacluster_predict %>% pivot_wider(id_cols = "PID",names_from = "metacluster_predict",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_OS <- sce_fischer$status_OS[match(rownames(Cox_df),sce_fischer$PID)]
Cox_df$OS_months <- sce_fischer$OS_months[match(rownames(Cox_df),sce_fischer$PID)]

Cox_df$age <- sce_fischer$age[match(rownames(Cox_df),sce_fischer$PID)]
Cox_df$grade <- factor(sce_fischer$grade[match(rownames(Cox_df),sce_fischer$PID)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce_fischer$pT_simple[match(rownames(Cox_df),sce_fischer$PID)])
Cox_df$pN_simple <- as.factor(sce_fischer$pN_simple[match(rownames(Cox_df),sce_fischer$PID)])
Cox_df$pM_simple <- as.factor(sce_fischer$pM_simple[match(rownames(Cox_df),sce_fischer$PID)])

## 2. Based on categories (mean division of cluster proportions)
Cox_df_mean <- foreach(i = Cox_df %>% select(-status_OS,-OS_months, -status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames,.combine = "cbind")%do%{
Cox_df %>% select(i) %>% mutate(mean = ifelse(Cox_df[,i] > mean(Cox_df[,i]),"B","A")) %>% pull(mean)
}

colnames(Cox_df_mean) <- Cox_df %>% select(-status_OS,-OS_months, -status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames
rownames(Cox_df_mean) <- Cox_df %>% select(-status_OS,-OS_months, -status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% rownames
Cox_df_mean <- as.data.frame(Cox_df_mean)

identical(rownames(Cox_df), rownames(Cox_df_mean))

Cox_df_mean <- cbind(Cox_df_mean, Cox_df %>% select(status_OS,OS_months, status_OS, OS_months, age, grade, pT_simple, pN_simple, pM_simple))

#Based on proportions
covariates <- Cox_df_mean %>% select(-status_OS,-OS_months, -age, -grade, -pT_simple, -pN_simple, -pM_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~ age + grade + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_mean)

summary(multi_model)

multi_res <- summary(multi_model)

plot_df_mean <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_mean <- ggplot(plot_df_mean, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metacluster_predicts")+
ylab("HR for mean")+
theme_classic()+
coord_cartesian(ylim=c(0.01,10000))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_mean
```

```{r}
# A: Fischer et al. - Univariate
PID_metacluster_predict <- colData(sce_fischer) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_metacluster_predict <- PID_metacluster_predict %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
PID_metagroups <- PID_metacluster_predict %>% pivot_wider(id_cols = "PID",names_from = "metacluster_predict",values_from = "fra") %>% column_to_rownames("PID")

## 1. Based on continuous cluster proportions
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$status_OS <- sce_fischer$status_OS[match(rownames(Cox_df),sce_fischer$PID)]
Cox_df$OS_months <- sce_fischer$OS_months[match(rownames(Cox_df),sce_fischer$PID)]

## 2. Based on categories (mean division of cluster proportions)
Cox_df_mean <- foreach(i = Cox_df %>% select(-status_OS,-OS_months) %>% colnames,.combine = "cbind")%do%{
Cox_df %>% select(i) %>% mutate(mean = ifelse(Cox_df[,i] > mean(Cox_df[,i]),"B","A")) %>% pull(mean)
}

colnames(Cox_df_mean) <- Cox_df %>% select(-status_OS,-OS_months) %>% colnames
rownames(Cox_df_mean) <- Cox_df %>% select(-status_OS,-OS_months) %>% rownames
Cox_df_mean <- as.data.frame(Cox_df_mean)

identical(rownames(Cox_df), rownames(Cox_df_mean))

Cox_df_mean <- cbind(Cox_df_mean, Cox_df %>% select(status_OS,OS_months))

#Based on proportions
covariates <- Cox_df_mean %>% select(-status_OS,-OS_months) %>% colnames

multi_formula <- as.formula(paste("Surv(OS_months, status_OS) ~ ", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_mean)

summary(multi_model)

multi_res <- summary(multi_model)

plot_df_mean <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_mean <- ggplot(plot_df_mean, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metacluster_predicts")+
ylab("HR for mean")+
theme_classic()+
coord_cartesian(ylim=c(0.01,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_mean
```

```{r}
#Plot survival curves
fit <- survfit(Surv(OS_months, status_OS) ~ tumor_8, data = Cox_df_mean)
print(fit)

p1 <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = "Set1")+
  ggtitle("Mean proportion")

p1
```

```{r}
# Jackson - Multivariate
patient_id_metacluster_predict <- colData(sce_jackson) %>% as.data.frame() %>% select(metacluster_predict,patient_id) %>% group_by(patient_id) %>% table() %>% as.data.frame
patient_id_metacluster_predict <- patient_id_metacluster_predict %>% group_by(patient_id) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
patient_id_metagroups <- patient_id_metacluster_predict %>% pivot_wider(id_cols = "patient_id",names_from = "metacluster_predict",values_from = "fra") %>% column_to_rownames("patient_id")

## 1. Based on continuous cluster proportions
Cox_df <- patient_id_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce_jackson$status_DFS[match(rownames(Cox_df),sce_jackson$patient_id)]
Cox_df$DFS_months <- sce_jackson$DFS_months[match(rownames(Cox_df),sce_jackson$patient_id)]

Cox_df$age <- sce_jackson$patient_age[match(rownames(Cox_df),sce_jackson$patient_id)]
Cox_df$grade <- factor(sce_jackson$tumor_grade[match(rownames(Cox_df),sce_jackson$patient_id)], levels = c("3","2","1"))
Cox_df$pT_simple <- as.factor(sce_jackson$pT_simple[match(rownames(Cox_df),sce_jackson$patient_id)])
Cox_df$pN_simple <- as.factor(sce_jackson$pN_simple[match(rownames(Cox_df),sce_jackson$patient_id)])
#Cox_df$pM_simple <- as.factor(sce_jackson$pM_simple[match(rownames(Cox_df),sce_jackson$patient_id)])

is.na(Cox_df$status_DFS)

## 2. Based on categories (mean division of cluster proportions)
Cox_df_mean <- foreach(i = Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple) %>% colnames,.combine = "cbind")%do%{
Cox_df %>% select(i) %>% mutate(mean = ifelse(Cox_df[,i] > mean(Cox_df[,i]),"B","A")) %>% pull(mean)
}

colnames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple) %>% colnames
rownames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple) %>% rownames
Cox_df_mean <- as.data.frame(Cox_df_mean)

identical(rownames(Cox_df), rownames(Cox_df_mean))

Cox_df_mean <- cbind(Cox_df_mean, Cox_df %>% select(status_DFS,DFS_months, age, grade, pT_simple, pN_simple))

#Based on proportions
covariates <- Cox_df_mean %>% select(-status_DFS,-DFS_months, -age, -grade, -pT_simple, -pN_simple) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ age + grade + pT_simple + pN_simple +", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_mean)

summary(multi_model)

multi_res <- summary(multi_model)

plot_df_mean <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratiDFS
hr_mean <- ggplot(plot_df_mean, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metacluster_predicts")+
ylab("HR for mean")+
theme_classic()+
#coord_cartesian(ylim=c(0.01,1000))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_mean
```

```{r}
# Jackson - Univariate
patient_id_metacluster_predict <- colData(sce_jackson) %>% as.data.frame() %>% select(metacluster_predict,patient_id) %>% group_by(patient_id) %>% table() %>% as.data.frame
patient_id_metacluster_predict <- patient_id_metacluster_predict %>% group_by(patient_id) %>% mutate(fra = Freq/sum(Freq))

### Multivariate cox proportional hazard model on multiple covariates (here proportion of phenotypes)
patient_id_metagroups <- patient_id_metacluster_predict %>% pivot_wider(id_cols = "patient_id",names_from = "metacluster_predict",values_from = "fra") %>% column_to_rownames("patient_id")

## 1. Based on continuous cluster proportions
Cox_df <- patient_id_metagroups

#add covariates for correction
Cox_df$status_DFS <- sce_jackson$status_DFS[match(rownames(Cox_df),sce_jackson$patient_id)]
Cox_df$DFS_months <- sce_jackson$DFS_months[match(rownames(Cox_df),sce_jackson$patient_id)]

#Cox_df$age <- sce_jackson$patient_age[match(rownames(Cox_df),sce_jackson$patient_id)]
#Cox_df$grade <- factor(sce_jackson$tumor_grade[match(rownames(Cox_df),sce_jackson$patient_id)], levels = c("3","2","1"))
#Cox_df$pT_simple <- as.factor(sce_jackson$pT_simple[match(rownames(Cox_df),sce_jackson$patient_id)])
#Cox_df$pN_simple <- as.factor(sce_jackson$pN_simple[match(rownames(Cox_df),sce_jackson$patient_id)])
#Cox_df$pM_simple <- as.factor(sce_jackson$pM_simple[match(rownames(Cox_df),sce_jackson$patient_id)])

is.na(Cox_df$status_DFS)

## 2. Based on categories (mean division of cluster proportions)
Cox_df_mean <- foreach(i = Cox_df %>% select(-status_DFS,-DFS_months) %>% colnames,.combine = "cbind")%do%{
Cox_df %>% select(i) %>% mutate(mean = ifelse(Cox_df[,i] > mean(Cox_df[,i]),"B","A")) %>% pull(mean)
}

colnames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months) %>% colnames
rownames(Cox_df_mean) <- Cox_df %>% select(-status_DFS,-DFS_months) %>% rownames
Cox_df_mean <- as.data.frame(Cox_df_mean)

identical(rownames(Cox_df), rownames(Cox_df_mean))

Cox_df_mean <- cbind(Cox_df_mean, Cox_df %>% select(status_DFS,DFS_months))

#Based on proportions
covariates <- Cox_df_mean %>% select(-status_DFS,-DFS_months) %>% colnames

multi_formula <- as.formula(paste("Surv(DFS_months, status_DFS) ~ ", paste(covariates, collapse = " + ")))

multi_model <- coxph(multi_formula, data = Cox_df_mean)

summary(multi_model)

multi_res <- summary(multi_model)

plot_df_mean <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratiDFS
hr_mean <- ggplot(plot_df_mean, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,50))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Tumor metacluster_predicts")+
ylab("HR for mean")+
theme_classic()+
coord_cartesian(ylim=c(0.01,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_mean
```

```{r}
#Plot survival curves - also show this?
fit <- survfit(Surv(DFS_months, status_DFS) ~ tumor_8, data = Cox_df_mean)
print(fit)

p1 <- ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = "Set1")+
  ggtitle("Mean proportion")

p1
#dev.off()
```

# 6. Image visualization 

```{r}
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/04_sce_TNBC_Clus_com.rds")
```

Filter sce for validation cohort

```{r filter sce, message=FALSE}
sce <- sce[,sce$TMANr %in% c("TMA21", "TMA26")]
sce
```

Load images as CytoImageList 

```{r}
images_jana <- loadImages("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_Jana/h5_img/")
images_jana

# Set channel names 
channelNames(images_jana) <- rownames(sce)

images_BL <- images_jana[names(images_jana) == "ZTMA21_984_230_BX6Y9"]
```

Visualize using cytoviewer 

```{r}
cytoviewer(image = images_BL)
```

# 7. Independent clustering 

**NOTE**: Please be aware that some analyse including unsupervised clustering involves random processes and may not be full reproducible across systems. We did, however, set seeds for all such analysis. To ensure full reproducibility, please use the information stored in the SCE objects available on Zenodo upon publication and the figure generation scripts. 

```{r}
sce_val <- readRDS("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/ML_sce_fischer_prediction.rds")
```

```{r}
rowData(sce_val)$tumor_cluster_sub <- rowData(sce_val)$clean_target %in% c("CK5","CK7","KRT14","CK8_18","CD15","HLADR","Vimentin","pH3","Ki67","EGFR")
```

```{r}
## Run clustering
set.seed(22)

mat <- t(assay(sce_val, "exprs")[rowData(sce_val)$tumor_cluster_sub,])

#k=30
out <- Rphenoannoy(mat, k = 30)

clusters <- factor(membership(out[[2]]))

sce_val$cluster_1_30 <- clusters

unique(sce_val$cluster_1_30)

#Save sce_vals after clustering
saveRDS(sce_val, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/09_sce_val_TNBC_tumor.rds")

gc()

#k=50
out <- Rphenoannoy(mat, k = 50)

clusters <- factor(membership(out[[2]]))

sce_val$cluster_1_50 <- clusters

unique(sce_val$cluster_1_50)

#Save sce_vals after clustering
saveRDS(sce_val, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/09_sce_val_TNBC_tumor.rds")

gc()

#k=70
out <- Rphenoannoy(mat, k = 70)

clusters <- factor(membership(out[[2]]))

sce_val$cluster_1_70 <- clusters

unique(sce_val$cluster_1_70)

#Save sce_vals after clustering
saveRDS(sce_val, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/09_sce_val_TNBC_tumor.rds")

gc()

#k=90
out <- Rphenoannoy(mat, k = 90)

clusters <- factor(membership(out[[2]]))

sce_val$cluster_1_90 <- clusters

unique(sce_val$cluster_1_90)

#Save sce_vals after clustering
saveRDS(sce_val, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/09_sce_val_TNBC_tumor.rds")

gc()

#k=110
out <- Rphenoannoy(mat, k = 110)

clusters <- factor(membership(out[[2]]))

sce_val$cluster_1_110 <- clusters

unique(sce_val$cluster_1_110)

#Save sce_vals after clustering
saveRDS(sce_val, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/09_sce_val_TNBC_tumor.rds")

gc()

#k=130
out <- Rphenoannoy(mat, k = 130)

clusters <- factor(membership(out[[2]]))

sce_val$cluster_1_130 <- clusters

unique(sce_val$cluster_1_130)

#Save sce_vals after clustering
saveRDS(sce_val, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/09_sce_val_TNBC_tumor.rds")

gc()
```

## Cluster stability testing and comparison

```{r}
#Pairwise rand index 
#Agreement between two clustering approaches - ARI > 0.5 = "good" similarity
pairwiseRand(sce_val$cluster_30,sce_val$cluster_50, mode = "index")
pairwiseRand(sce_val$cluster_50,sce_val$cluster_70, mode = "index")
pairwiseRand(sce_val$cluster_70,sce_val$cluster_90, mode = "index")
pairwiseRand(sce_val$cluster_90,sce_val$cluster_110, mode = "index")
pairwiseRand(sce_val$cluster_110,sce_val$cluster_130, mode = "index")
```

```{r clustree, message = FALSE, warning=FALSE}
sce_val <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/09_sce_val_TNBC_tumor.rds")

#Over different resolution parameters - Silhouette width [-1,1,max], Purity[0,1,max], Root-mean-squared difference [min]
my_clusters <- colnames(colData(sce_val))[str_detect(colnames(colData(sce_val)),"cluster_1_[:digit:]")]

df_1 <- foreach(i = seq_along(my_clusters),.combine = rbind) %do% {
num_clus <- length(unique(colData(sce_val)[,my_clusters[i]]))
sil <- approxSilhouette(t(assay(sce_val, "exprs")), clusters=colData(sce_val)[,my_clusters[i]])
rmsd <- clusterRMSD(t(assay(sce_val, "exprs")), clusters=colData(sce_val)[,my_clusters[i]])
data.frame(num_clus = num_clus,
           k = as.numeric(str_split(my_clusters[i],"_",simplify = TRUE)[,3]), 
           sil = mean(sil$width), 
           rmsd = sum(rmsd),
           type = "cluster_sub")
}

df_out <- rbind(df_1)

p1 <- ggplot(df_out, aes(x=k, y=num_clus, color = type))+
  geom_point(aes(k, num_clus), size = 3)+
  geom_line(aes(k, num_clus))+
  theme_classic()+
  scale_color_brewer(palette = "Set1")+
  guides(color = "none")

p2 <- ggplot(df_out, aes(x=k, y=rmsd, color = type))+
  geom_point(aes(k, rmsd), size = 3)+
  geom_line(aes(k, rmsd))+
  theme_classic()+
  scale_color_brewer(palette = "Set1")+
  guides(color = "none")


p3 <- ggplot(df_out, aes(x=k, y=sil, color = type))+
  geom_point(aes(k, sil), size = 3)+
  geom_line(aes(k, sil))+
  theme_classic()+
  scale_color_brewer(palette = "Set1")
  
p1+p2+p3 
```

```{r}
#Heatmap body colors
col_exprs <- colorRamp2(c(0,1,2,3), c("#440154FF", "#31688EFF", "#35B779FF", "#FDE725FF"))

### 1. Mean exprs counts
#Aggregrate across cells with the mean

cluster_mean_sce_val <- aggregateAcrossCells(sce_val, ids = sce_val$cluster_1_90,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_val)$tumor_cluster_channel)

cluster_mean_sce_val_1 <- aggregateAcrossCells(sce_val, ids = sce_val$cluster_130,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_val)$tumor_cluster_channel)

h_110 <- Heatmap(t(assay(cluster_mean_sce_val, "exprs")),
        column_title = "1. mean exprs",
        col = col_exprs,
        name= "mean exprs",
        show_row_names = T, 
        show_column_names =  T, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")
  
h_130 <- Heatmap(t(assay(cluster_mean_sce_val_1, "exprs")),
        column_title = "1. mean exprs",
        col = col_exprs,
        name= "mean exprs",
        show_row_names = T, 
        show_column_names =  T, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2")

h_110
h_130

h_110+
h_130

scale(table(sce_val$cluster_110,sce_val$metacluster_predict))

Heatmap(scale(table(sce_val$cluster_110,sce_val$metacluster_predict)), clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2")

Heatmap(log2(table(sce_val$cluster_110,sce_val$metacluster_predict)+1), 
        col = viridis(100))
```

## Define tumor metaclusters 

```{r}
# Update
sce_val$metacluster_predict <- sce_fischer$metacluster_predict[match(sce_val$cell_id, sce_fischer$cell_id)]
```

Based on manual assignment and accuracy testing 

```{r}
# Extract cluster and metacluster data
clusters <- sce_val$cluster_1_90
metacluster_old <- sce_val$metacluster_predict

# Unique clusters and metaclusters
unique_clusters <- unique(clusters)
unique_metaclusters <- unique(metacluster_old)

# Initialize a dataframe to store accuracies
accuracy_matrix <- as.data.frame(matrix(0, nrow = length(unique_clusters), ncol = length(unique_metaclusters)))
rownames(accuracy_matrix) <- unique_clusters
colnames(accuracy_matrix) <- unique_metaclusters

# Calculate accuracy for each cluster-to-metacluster assignment
for (i in unfactor(unique_clusters)) {
  for (j in unfactor(unique_metaclusters)) {
    
    cluster_indices <- which(clusters == i)
    assigned_metaclusters <- metacluster_old[cluster_indices]
    
    confusion_stats <- confusionMatrix(
      assigned_metaclusters, 
      factor(rep(j, length(assigned_metaclusters)),levels = levels(assigned_metaclusters)), mode = "everything")
    
    #accuracy_matrix[i, j] <- confusion_stats$overall['Accuracy']
    cm_stats_1 <- as.data.frame(confusion_stats$byClass)
    accuracy_matrix[i, j] <- mean(cm_stats_1$F1, na.rm = TRUE) 
    
  }
}

# Assign clusters to metaclusters based on highest accuracy

best_assignments <- apply(accuracy_matrix, 1, function(row) names(row)[which.max(row)])

best_assignments_df <- data.frame(new = unname(best_assignments), old = names(best_assignments))

length(unique(best_assignments_df$new))

# Manual annoation based on similar expression profiles
best_assignments_df$new[23] <- "tumor_8"
best_assignments_df$new[31] <- "tumor_2"
best_assignments_df$new[21] <- "tumor_10"

length(unique(best_assignments_df$new))
```

```{r}
# Update the SingleCellExperiment object with new metacluster assignments
sce_val$metacluster_new <- best_assignments_df$new[match(sce_val$cluster_1_90, best_assignments_df$old)]
sce_val$metacluster_new <- factor(sce_val$metacluster_new, levels = levels(sce_val$metacluster_predict))
```

## Comparison

### Confusion matrix

```{r}
cm <- confusionMatrix(sce_val$metacluster_predict, sce_val$metacluster_new, mode = "everything")

cm_stats <- as.data.frame(cm$byClass)

mean(cm_stats$`Balanced Accuracy`, na.rm = TRUE) 
mean(cm_stats$F1, na.rm = TRUE) # Report this

data.frame(cm$byClass) %>%
  mutate(class = sub("Class: ", "", rownames(cm$byClass))) %>%
  ggplot() + 
  geom_point(aes(1-Specificity, Sensitivity, 
                 size = Detection.Rate,
                 fill = class),
             shape = 21) + 
  scale_fill_manual(values = metadata(sce_fischer)$colors$metacluster_combined, limits = force) +
  theme_classic(base_size = 15) + 
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")+
  ggtitle(paste0("Accuracy: ",round(cm$overall[1],3)," [",round(cm$overall[3],3)," - ",round(cm$overall[4],3),"]"))
```

### Mean expression profile

Compare mean expression profiles between discovery and validation cohort

```{r}
discovery_exprs <- aggregateAcrossCells(sce_val, ids = sce_val$metacluster_predict,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_val)$tumor_cluster_sub)

validation_exprs <- aggregateAcrossCells(sce_val, ids = sce_val$metacluster_new,
                                 statistics = "mean",
                                 use.assay.type = "exprs",
                                 subset.row = rowData(sce_val)$tumor_cluster_sub)

val_exprs <- assay(validation_exprs, "exprs")

dis_exprs <- assay(discovery_exprs, "exprs")[,colnames(assay(discovery_exprs, "exprs")) %in% 
colnames(assay(validation_exprs, "exprs"))]

corrplot(cor(dis_exprs,val_exprs), 
         method="color", 
         col=brewer.pal(n=8, name="RdBu"),  
         type="upper", 
         order="hclust",
         hclust.method = "ward.D2", 
         addCoef.col = "black", 
         addCoefasPercent = F,
         tl.col="black", 
         tl.srt=40,
         diag=TRUE)
```

### Adjusted Rand index

```{r}
pairwiseRand(sce_val$metacluster_new, sce_val$metacluster_predict, mode = "index")
```

### Fraction correlation 

```{r}
validation <- colData(sce_val) %>% as.data.frame %>% select(metacluster_new) %>% count(metacluster_new) %>% mutate(fra = n/sum(n)) 

discovery <- colData(sce) %>% as.data.frame %>% select(metacluster_combined) %>% count(metacluster_combined) %>% mutate(fra = n/sum(n)) %>% filter(metacluster_combined %in% validation$metacluster_new)

discovery$fra_valid <- validation$fra[match(discovery$metacluster_combined, validation$metacluster_new)]

cor(discovery$fra, discovery$fra_valid, method = "spearman")

spearman_test <- cor.test(discovery$fra, discovery$fra_valid, method = "spearman")

ggplot(discovery, aes(x = fra, y = fra_valid))+
  geom_smooth(method='lm', formula= y~x, se = TRUE, color = "darkgrey")+
  geom_point(aes(color = metacluster_combined), size = 3)+
  scale_color_manual(values=metadata(sce)$colors$metacluster_combined)+
  ylab("Independent clustered MC fraction (Fischer et al., 2023)")+
  xlab("MC fraction (Discovery)")+
  theme_classic()+
  ggtitle(paste("Spearman rho = ",round(spearman_test$estimate, digits = 2)))
```

### Rich heatmap

```{r}
### 1. Heatmap body ###
sce_val$cluster_selected <- sce_val$metacluster_new

#Aggregrate across cells with the mean (median is also possible)
cluster_mean_sce_val <- aggregateAcrossCells(sce_val, 
                                             ids = sce_val$cluster_selected,
                                             statistics = "mean",
                                             use.assay.type = "exprs",
                                             subset.row = rowData(sce_val)$tumor_cluster_sub)

# Heatmap body colors 
#col_exprs <- colorRamp2(c(0,1.25,2.5,3.75), c("#440154FF", "#31688EFF", "#35B779FF", "#FDE725FF"))


### 2. Heatmap annotation ###
###2.1 Metacluster 
cluster_mean_sce_val$metacluster_new

###2.2 Metadata features

#Number of cells per cluster 
n_cells <- colData(cluster_mean_sce_val) %>% as.data.frame %>% select(cluster_selected, ncells)

# Proportion of grade per cluster
grade <- colData(sce_val) %>% as.data.frame() %>% select(cluster_selected,grade) %>% group_by(cluster_selected) %>% table() %>% as.data.frame
grade <- grade %>% group_by(cluster_selected) %>% mutate(fra = Freq/sum(Freq)) 
grade <- grade %>% select(-Freq) %>% pivot_wider(id_cols = cluster_selected,names_from = grade,values_from = fra) %>% column_to_rownames("cluster_selected")

# Proportion of pT stage per cluster
pT <- colData(sce_val) %>% as.data.frame() %>% select(cluster_selected,pT_simple) %>% group_by(cluster_selected) %>% table() %>% as.data.frame
pT <- pT %>% group_by(cluster_selected) %>% mutate(fra = Freq/sum(Freq)) 
pT <- pT %>% select(-Freq) %>% pivot_wider(id_cols = cluster_selected,names_from = pT_simple,values_from = fra) %>% column_to_rownames("cluster_selected")

# Proportion of pN stage per cluster
pN <- colData(sce_val) %>% as.data.frame() %>% select(cluster_selected,pN_simple) %>% group_by(cluster_selected) %>% table() %>% as.data.frame
pN <- pN %>% group_by(cluster_selected) %>% mutate(fra = Freq/sum(Freq)) 
pN <- pN %>% select(-Freq) %>% pivot_wider(id_cols = cluster_selected,names_from = pN_simple,values_from = fra) %>% column_to_rownames("cluster_selected")

# Number of contributing patients
cluster_PID <- colData(sce_val) %>% as.data.frame() %>% select(cluster_selected,PID) %>% group_by(cluster_selected) %>% table() %>% as.data.frame
n_PID <- cluster_PID %>% filter(Freq>0) %>% group_by(cluster_selected) %>% count(name = "n_PID") %>% column_to_rownames("cluster_selected")

# Cluster categories contributing to clusters
cluster_CC <- colData(sce_val) %>% as.data.frame() %>% select(metacluster_predict,cluster_category) %>% group_by(metacluster_predict) %>% table() %>% as.data.frame
cluster_CC <- cluster_CC %>% group_by(metacluster_predict) %>% mutate(fra = Freq/sum(Freq)) 
cluster_CC <- cluster_CC %>% select(-Freq) %>% pivot_wider(id_cols = metacluster_predict,names_from = cluster_category,values_from = fra) %>% column_to_rownames("metacluster_predict")

# Metadata annotation
metadata(sce_val)$colors$metacluster_combined <- metadata(sce)$colors$metacluster_combined
col_mc <- metadata(sce_val)$colors$metacluster_combined

ha_meta <- HeatmapAnnotation(#cluster_selected = anno_simple(unfactor(cluster_mean_sce_val$cluster_selected), pch = unfactor(cluster_mean_sce_val$cluster_selected),width = unit(8, "mm"), border=TRUE),
                            metacluster = anno_simple(unfactor(cluster_mean_sce_val$metacluster_new),width = unit(8, "mm"), pch = str_split(unfactor(cluster_mean_sce_val$metacluster_new),"_",simplify = TRUE)[,2], col =  col_mc, border=TRUE),
                            n_cells = anno_barplot(n_cells$ncells, width = unit(12, "mm"),gp = gpar(fill = "#440154FF")),
                            n_PID = anno_barplot(n_PID, width = unit(12, "mm"),gp = gpar(fill = "#440154FF")),
                            #grade = anno_barplot(grade, width = unit(15, "mm"),gp = gpar(fill = metadata(sce_val)$colors$grade)),
                            #pT = anno_barplot(pT, width = unit(15, "mm"),gp = gpar(fill = metadata(sce_val)$colors$pT_simple)),
                            #pN = anno_barplot(pN, width = unit(15, "mm"),gp = gpar(fill = metadata(sce_val)$colors$pN_simple)),
                            border = TRUE, 
                            annotation_name_rot = 90,
                            gap = unit(2,"mm"),
                            which = "row")
                            
### 3. Plot heatmap ###

h_meta <- Heatmap(t(assay(cluster_mean_sce_val, "exprs")),
        column_title = "cluster_channels",
        col = viridis(100),
        name = "mean exprs",
        row_title = NULL, 
        show_column_names =  TRUE, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2",
        row_split = cluster_mean_sce_val$metacluster_new,
        cluster_row_slices = FALSE)+
  ha_meta

draw(h_meta)
```


# 8. Focus on MC8 frequency / patches / dominant PID 

```{r}
# Focus on Patients with high MC8 
PID_MC8 <- (as.data.frame(cur_table) %>% filter(Var1 == "tumor_8") %>% arrange(desc(Freq)) %>% pull(Var2) %>% unfactor)[1:10]
  
colData(sce_val) %>% as.data.frame %>% filter(PID %in% PID_MC8) %>% select(PID, pN_simple, pM_simple, status_OS, OS_months) %>% unique()

plotSpatial(sce_val[,sce_val$PID == "924"], 
            node_color_by = "CK5", 
            assay_type = "exprs",
            img_id = "sample_id", 
            node_size_fix = 2,
            node_shape_by = "analysis_cat")

plotSpatial(sce_val[,sce_val$PID == "924"], 
            node_color_by = "CK7", 
            assay_type = "exprs",
            img_id = "sample_id", 
            node_size_fix = 2,
            node_shape_by = "analysis_cat")

plotSpatial(sce_val[,sce_val$PID == "924"], 
            node_color_by = "metacluster_predict", 
            assay_type = "exprs",
            img_id = "sample_id", 
            node_size_fix = 2,
            node_shape_by = "analysis_cat")+
  scale_color_manual(values = metadata(sce)$colors$metacluster_combined)
```

```{r}
# Focus on basoluminal cells
cur_scat <- t(assay(sce_val,"exprs")) %>% as.data.frame %>% select("CK5","CK7","CK8_18")

cur_scat$metacluster_predict <- sce_val$metacluster_predict[match(rownames(cur_scat),sce_val$cell_id)]
  
cur_scat <- cur_scat %>% filter(metacluster_predict == "tumor_8")

ggplot(cur_scat, aes(x=CK7, y=CK5, color = metacluster_predict)) +
  geom_point(alpha = 0.5)+
  #stat_density_2d(aes(fill = ..level..), geom="polygon", contour_var = "count")+
  theme_classic()

ggplot(cur_scat, aes(x=CK8_18, y=CK5, color = metacluster_predict)) +
  geom_point(alpha = 0.5)+
  #stat_density_2d(aes(fill = ..level..), geom="polygon", contour_var = "count")+
  theme_classic()
```


```{r, message=FALSE}
sce_fischer <- readRDS("/mnt/central_nas/tnbc_volume/TNBC_validation/TNBC_all/data_analysis/ML_sce_fischer_prediction.rds")

sce <- sce_fischer
```

Build knn interaction graph 

```{r}
# Build spatial knn graph
set.seed(22)
sce <- buildSpatialGraph(sce, img_id = "sample_id", type = "knn", k = 20, 
                         coords = c("Pos_X",  "Pos_Y"),
                         #max_dist = 30,
                         BPPARAM = MulticoreParam())

colPairNames(sce) #now includes knn_interaction_graph
```

Patch detection

```{r}
## Patch detection
sce <- patchDetection(sce,
               patch_cells = sce$metacluster_predict == "tumor_8",
               colPairName = "knn_interaction_graph",
               min_patch_size = 10,
               name = "patch_MC8",
               expand_by = 0, 
               img_id = "sample_id", 
               BPPARAM = MulticoreParam(workers = 10))
```

Frequency analysis 

```{r}
# Colors 
colors_sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
```

```{r}
patch_MC8_PID <- colData(sce) %>% as.data.frame() %>% filter(!is.na(patch_MC8)) %>% pull(PID) %>% unique()

cells_MC8_PID <- colData(sce) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(PID) %>% table() %>% as.data.frame

sce$patch_MC8_classification <- ifelse(sce$PID %in% patch_MC8_PID,"patch_MC8","no_patch_MC8")

PID_patch <- colData(sce) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_patch <- PID_patch %>% group_by(PID) %>% filter(PID %in% patch_MC8_PID) %>% filter(str_detect(metacluster_predict,"tumor")) %>% mutate(fra = Freq/sum(Freq)) 
```

```{r}
# most abundant phenotype in patients with MC8 patches 
PID_patch_abundant <- PID_patch %>% #filter(metacluster_predict != "tumor_8") %>% 
  mutate(max = max(fra)) %>% filter(fra == max) %>% ungroup() %>% count(metacluster_predict) %>% mutate(fra = n/sum(n), category = "cat") %>% arrange(fra)

PID_patch_abundant$metacluster_predict <- factor(PID_patch_abundant$metacluster_predict, levels = unfactor(PID_patch_abundant$metacluster_predict))

ggplot(PID_patch_abundant) +
  geom_tile(color = "black",aes(category, n, fill = metacluster_predict), position = "stack")+
  scale_fill_manual(values=metadata(colors_sce)$colors$metacluster_combined) + 
  theme_classic(base_size = 15)+
  ylab("Density")+
  xlab("PID")+
  theme(axis.text.x = element_blank())
```

```{r}
PID_dom_MC8 <- PID_patch %>% mutate(max = max(fra)) %>% filter(fra == max) %>% ungroup() %>% filter(metacluster_predict == "tumor_8") %>% pull(PID)

PID_overall <- colData(sce) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_MC8_overall <- PID_overall %>% group_by(PID) %>% filter(metacluster_predict == "tumor_8" & Freq > 5) %>% mutate(patch = ifelse(PID %in% patch_MC8_PID, TRUE, FALSE)) %>% mutate(dominated = ifelse(PID %in% PID_dom_MC8, TRUE, FALSE)) %>% mutate(PID_category = ifelse(patch && dominated, "dominated", ifelse(patch, "patch", "cells")))                                                                                    
PID_MC8_overall_plot <- PID_MC8_overall %>% group_by(PID_category) %>% count() %>% ungroup() %>% mutate(fra = n/sum(n)) %>% mutate(category = "MC8")


PID_MC8_overall_plot$PID_category <- factor(PID_MC8_overall_plot$PID_category, levels = c("cells","patch","dominated"))

ggplot(PID_MC8_overall_plot) +
  geom_tile(color = "black",aes(category, n, fill = PID_category), position = "stack")+
  scale_fill_brewer(palette = "Blues")+
  theme_classic(base_size = 15)+
  ylab("Number of patients")+
  theme(axis.title.x = element_blank())
```

Interaction testing 

```{r}
sce_tumor <- sce[,sce$metacluster_predict != "Unknown"]

sce_tumor$metacluster_predict <- unfactor(sce_tumor$metacluster_predict)

# For PID with at least 5 MC8 cells 
PID_overall <- colData(sce_tumor) %>% as.data.frame() %>% select(metacluster_predict,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_MC8_overall <- PID_overall %>% group_by(PID) %>% filter(metacluster_predict == "tumor_8" & Freq > 5)

sce_tumor_MC8 <- sce_tumor[,sce_tumor$PID %in% PID_MC8_overall$PID]

set.seed(22)
interaction_tumor <- testInteractions(sce_tumor_MC8, 
                  group_by = "sample_id", 
                  label = "metacluster_predict", 
                  colPairName = "neighborhood", 
                  BPPARAM = MulticoreParam(workers = 10, RNGseed = 22))
```

```{r}
interaction_tumor_plot <- interaction_tumor %>% as_tibble() %>% filter(from_label %in% c("tumor_8")) %>% filter(to_label != "tumor_8") %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

ggplot(interaction_tumor_plot) +
        geom_tile(aes(reorder(to_label,sum_sigval), from_label, fill = sum_sigval)) +
        scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_classic()
```

# Session Info 

```{r}
sessionInfo()
```

