---
title: "TCGA_BRCA_RNAseq_NORMAL_analysis"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: TCGA_COMBINED_DET.rds, TCGA_COMBINED_RES.rds, TCGA_NORMAL_DET.rds, TCGA_NORMAL_RES.rds, GeneSignatures.rds 

# Validation

Data analysis scripts for data analysis in the validation cohorts. 

# Script: TCGA BRCA - Tumor and Normal analysis

Here, we are analyzing RNAseq data from the TCGA BRCA cohort.

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(scuttle)
library(scater)
library(scran)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(SummarizedExperiment)
library(TCGAbiolinks)
library(DESeq2)
library(gridExtra)
library(ggVennDiagram)
library(apeglm)
library(survival)
library(survminer)
library(readxl)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(ExploreModelMatrix)
library(ggvenn)
library(wesanderson)
library(immunedeconv)
library(limma)
library(easyalluvial)
library(ggalluvial)
library(ggpubr)
```

## Load data 

```{r}
query.exp <- GDCquery(
    project = "TCGA-BRCA", 
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification", 
    workflow.type = "STAR - Counts",
    sample.type = c("Primary Tumor", "Solid Tissue Normal")
)

setwd("/mnt/central_nas/tnbc_volume/TNBC_validation/TCGA_BRCA/TU_NT/")

GDCdownload(
    query = query.exp,
    files.per.chunk = 100
)

brca.exp <- GDCprepare(
    query = query.exp, 
    save = TRUE, 
    save.filename = "brcaExp_normal.rda"
)
```

```{r}
brca.clean <- brca.exp
colData <- colData(brca.clean) %>% as.data.frame
```

## 1. Normal sample processing

```{r}
normal.clean <- brca.clean[,brca.clean$shortLetterCode == "NT"]
```

```{r}
saveRDS(normal.clean,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_normal.rds")
saveRDS(brca.exp,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_tumorANDnormal.rds")

normal.clean <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_normal.rds")
normal.clean
```

### DESeq2 processing

113 samples

```{r}
# Create DESeqDataset
ddsSE <- DESeqDataSet(normal.clean, design = ~ 1)
ddsSE

# DESeq analysis includes count normalization based on size factors
res <- DESeq(ddsSE)
rownames(res) <- rowData(res)$gene_name

counts(res, normalized = TRUE) # Access normalized counts 

# Variance-stabilizing transformation
vsd <- DESeq2::vst(res)

# Access normalized and stabilized counts
mat <- assay(vsd)
```


```{r}
saveRDS(vsd,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_Normal_vsd.rds")
saveRDS(res,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_Normal_res.rds")
```

```{r}
normal.clean <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_normal.rds")
vsd <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_Normal_vsd.rds")
res <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_Normal_res.rds")
```


### CK5/CK7 Signature 

GMM on size-factor normalized (NOT VST) counts

```{r}
set.seed(22)

# 1. CK5
GMM_CK5 <- Mclust(counts(res, normalized = TRUE)["KRT5",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK5$classification == 2, 
                                        "KRT5+",
                                        "KRT5-"), 
                     KRT5_exprs = assay(vsd)["KRT5",],
                     uncertainty = GMM_CK5$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = KRT5_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")

#Add GMM epi class to object
vsd$GMM_KRT5 <- GMM_df$class_epi

# 2. CK7
GMM_CK7 <- Mclust(counts(res, normalized = TRUE)["KRT7",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK7$classification == 2, 
                                        "KRT7+",
                                        "KRT7-"), 
                     KRT7_exprs = assay(vsd)["KRT7",],
                     uncertainty = GMM_CK7$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = KRT7_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 17)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")

#Add GMM epi class to object
vsd$GMM_KRT7 <- GMM_df$class_epi

## Combine KRT5/7 GMM scores 
vsd$KRT5_7_score <- ifelse(vsd$GMM_KRT5 == "KRT5+" & vsd$GMM_KRT7 == "KRT7+", "KRT5/7+", 
       ifelse(vsd$GMM_KRT5 == "KRT5+" & vsd$GMM_KRT7 == "KRT7-", "KRT5+",
              ifelse(vsd$GMM_KRT5 == "KRT5-" & vsd$GMM_KRT7 == "KRT7+", "KRT7+","KRT5/7-")))

## Add colors 
metadata(vsd)$colors$KRT57_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("KRT5/7-", "KRT5/7+","KRT5+", "KRT7+"))
metadata(vsd)$colors$PAM50 <- setNames(brewer.pal(6,"Paired"),sort(unique(vsd$PAM50Call_RNAseq)))

## Visualize 

### 1. Normalized counts
plot <- as.data.frame(t(counts(res, normalized = TRUE)[rownames(counts(res, normalized = TRUE)) %in% c("KRT5","KRT7"),]))
plot$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(plot),colnames(vsd))]
#plot$PAM50Call_RNAseq <- vsd$PAM50Call_RNAseq[match(rownames(plot),colnames(vsd))]
#plot$TNBC <- vsd$TNBC[match(rownames(plot),colnames(vsd))]
plot$GMM_KRT5 <- vsd$GMM_KRT5[match(rownames(plot),colnames(vsd))]
plot$GMM_KRT7 <- vsd$GMM_KRT7[match(rownames(plot),colnames(vsd))]
```


```{r}
main <- ggplot(plot, aes(x=KRT7, y=KRT5, color = KRT5_7_score)) +
  geom_point(size = 2, alpha = 0.5)+
  theme_classic()+
  xlab("CK7 normalized counts")+
  ylab("CK5 normalized counts")+
  coord_cartesian()+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  guides(color = "none")

y_side <- ggplot(plot, aes(x = KRT5, fill=GMM_KRT5)) + 
  geom_density(alpha=0.8) + 
  scale_fill_manual(values = c('#999999','#F4B5BD')) + 
  theme_classic()+
  theme(legend.position = "none", axis.title.y = element_blank())+
  coord_flip()

x_side <- ggplot(plot, aes(x = KRT7, fill=GMM_KRT7)) + 
  geom_density(alpha=0.8) + 
  scale_fill_manual(values = c('#999999','#9C964A')) + 
  coord_cartesian()+
  theme_classic()+
  theme(legend.position = "none", axis.title.x = element_blank())

blankPlot <- ggplot()+geom_blank(aes(1,1)) + 
  cowplot::theme_nothing()

grid.arrange(x_side, blankPlot, main, y_side, 
        ncol=2, nrow=2, widths=c(4, 1.4), heights=c(1.4, 4))
```
```{r}
colData(vsd) %>% as.data.frame %>% group_by(shortLetterCode, KRT5_7_score) %>% count()
```



### Keratins

```{r}
# Define keratins of interest
keratins <- c("KRT5","KRT7","KRT15","KRT8","KRT18","KRT19", "KRT14","KRT17","ITGA6","EPCAM")

# Define heatmap based on normalized counts 
## Option 1: size-factor normalized counts 
#df <- t(counts(dds, normalized = TRUE)[keratins,]) %>% as.data.frame %>% rownames_to_column("PID")
## Option 2: vst counts
df <- t(assay(vsd)[keratins,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")

## Create heatmap body 
df$KRT5_7_score <- vsd$KRT5_7_score[match(df$PID, colnames(vsd))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% dplyr::select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(ht, name = "Mean Z-scores")

Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

### Stem-like 

```{r}
# Define stem markers of interest
stem <- c("SOX9","KIT","PROM1","YAP1","HDAC2","NOTCH1","CD44","ALDH1A3","BRCA1","BRCA2", "TGFB2", "ANXA8","MUC1", "CD24", "TNFAIP3", "ERBB3", "TNFRSF11A")#"NT5E","TNF","NFKB1","EZH2", "MDK",
#stem <- c("SOX9","KIT","PROM1","YAP1","HDAC2","NOTCH1","CD44","ALDH1A3","BRCA1","BRCA2", "TGFB2", "ANXA8","MUC1", "NT5E","TNF","NFKB1","EZH2", "ALDH3A1", "ALDH1A1", "TNFAIP3", "CD24", "WNT4")

stem_rel <- stem[stem %in% rownames(assay(vsd))]

# Define heatmap based on normalized counts 
## Option 1: size-factor normalized counts 
#df <- t(counts(dds, normalized = TRUE)[stem,]) %>% as.data.frame %>% rownames_to_column("PID")
## Option 2: vst counts
df <- t(assay(vsd)[stem_rel,]) %>% as.data.frame 
#df <- as.data.frame(scale(df))
df <- df %>% rownames_to_column("PID")

## Create heatmap body 
df$KRT5_7_score <- vsd$KRT5_7_score[match(df$PID, colnames(vsd))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% dplyr::select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(ht, name = "Mean Z-scores")

Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

### Gene signatures 

#### Gray et al signatures

A: Alveolar subtypes 

```{r}
AV_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Lietal_AV_BL_Signatures_reduced.xlsx",sheet = 1)

colnames(AV_sig) <- c("pan_AV", "BL_enriched","AP_enriched")

# Extract unique genes that were up- or down-regulated in each signature 
panAV_sig_up <- AV_sig$pan_AV
panAV_sig_up <- panAV_sig_up[!is.na(panAV_sig_up)]

BL_sig_up <- AV_sig$BL_enriched
BL_sig_up <- BL_sig_up[!is.na(BL_sig_up)]

AP_sig_up <- AV_sig$AP_enriched
AP_sig_up <- AP_sig_up[!is.na(AP_sig_up)]
```

```{r}
## 1. pan AV signature per PID group 
panAV_sig_up_rel <- panAV_sig_up[panAV_sig_up %in% rownames(assay(vsd))]
# Filter for available genes
#TNBC_PID <- colnames(vsd[,vsd$TNBC == "TNBC"])
#vsd_TNBC <- vsd[, colnames(vsd) %in% TNBC_PID]
#df <- t(assay(vsd_TNBC)[panAV_sig_up_rel ,]) %>% as.data.frame
df <- t(assay(vsd)[panAV_sig_up_rel ,]) %>% as.data.frame

# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Pan-Alveolar signature (Gray et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```


```{r}
## 2. BL signature per PID group 
BL_sig_up_rel <- BL_sig_up[BL_sig_up %in% rownames(assay(vsd))]

# Filter for available genes
df <- t(assay(vsd)[BL_sig_up_rel ,]) %>% as.data.frame

#df <- t(assay(vsd)[BL_sig_up_rel ,]) %>% as.data.frame

# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

#compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
#compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Basal-luminal signature (Li et al.)")+
  #geom_signif(
  #  comparisons = list(c("KRT5/7+", "KRT5+"), c("KRT5/7+", "KRT5/7-"),c("KRT5/7+", "KRT7+")),
  #  map_signif_level = FALSE, step_increase = 0.1, test = "t.test"
  #)+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

#### Lim et al signatures

```{r}
lum_progenitor_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/LuminalProgenitor_Signature_Limetal_NatureMedicine_2009.xls")
mature_lum_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/MatureLuminalSignature_Limetal_NatureMedicine_2009.xls")
basal_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Basal_Signature_Limetal_NatureMedicine_2009.xls")

# Extract unique genes that were up- or down-regulated in each signature 
lum_progenitor_sig_up <- lum_progenitor_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
lum_progenitor_sig_up <- lum_progenitor_sig_up[!is.na(lum_progenitor_sig_up)]
lum_progenitor_sig_down <- lum_progenitor_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
lum_progenitor_sig_down <- lum_progenitor_sig_down[!is.na(lum_progenitor_sig_down)]

mature_lum_sig_up <- mature_lum_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
mature_lum_sig_up <- mature_lum_sig_up[!is.na(mature_lum_sig_up)]
mature_lum_sig_down <- mature_lum_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
mature_lum_sig_down <- mature_lum_sig_down[!is.na(mature_lum_sig_down)]

basal_sig_up <- basal_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
basal_sig_up <- basal_sig_up[!is.na(basal_sig_up)]
basal_sig_down <- basal_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
basal_sig_down <- basal_sig_down[!is.na(basal_sig_down)]
```

Upregulated genes per signature 

```{r}
## 1. Luminal progenitor signature per PID group 
lum_progenitor_sig_up_rel <- lum_progenitor_sig_up[lum_progenitor_sig_up %in% rownames(assay(vsd))]
# Filter for available genes
df <- t(assay(vsd)[lum_progenitor_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Luminal progenitor signature")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

```{r}
#Mature luminal signature per PID group 
mature_lum_sig_up_rel <- mature_lum_sig_up[mature_lum_sig_up %in% rownames(assay(vsd))]

df <- t(assay(vsd)[mature_lum_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Mature luminal signature")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```



## 2. Tumor VS Normal 

### Label transfer 

```{r}
# Normal samples
vsd_normal <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_Normal_vsd.rds")
vsd_normal$sample_full <- colnames(vsd_normal)
vsd_normal$KRT5_7_score_anno <- paste0("NT_", vsd_normal$KRT5_7_score)

# Tumor samples from TCGA BRCA analysis       
vsd_tumor <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_PAM50_vsd.rds")
vsd_tumor$sample_full <- colnames(vsd_tumor)
vsd_tumor$KRT5_7_score_anno <- paste0("TU_", vsd_tumor$KRT5_7_score)
```


```{r}
# Load full dataset and add labels
full.clean <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_tumorANDnormal.rds")
full.clean$sample_full <- colnames(full.clean)

full.clean$KRT5_7_score_anno <- vsd_tumor$KRT5_7_score_anno[match(full.clean$sample_full, vsd_tumor$sample_full)]
full.clean$KRT5_7_score_anno <- ifelse(!is.na(vsd_tumor$KRT5_7_score_anno[match(full.clean$sample_full, vsd_tumor$sample_full)]), vsd_tumor$KRT5_7_score_anno[match(full.clean$sample_full, vsd_tumor$sample_full)], vsd_normal$KRT5_7_score_anno[match(full.clean$sample_full, vsd_normal$sample_full)])

unique(full.clean$KRT5_7_score_anno)
full.clean <- full.clean[,!is.na(full.clean$KRT5_7_score_anno)]
full.clean$broad_cat <- ifelse(str_detect(full.clean$KRT5_7_score_anno, "TU"), "TU", "NT")


# Select BL patients from Tumor and Normal
full.selected <- full.clean[,full.clean$KRT5_7_score_anno %in% c("TU_KRT5/7+", "NT_KRT5/7+")]
unique(full.selected$KRT5_7_score_anno)
colData(full.selected) %>% as.data.frame %>% group_by(KRT5_7_score_anno) %>% count()
full.selected$KRT5_7_score_anno <- factor(full.selected$KRT5_7_score_anno)
```

### DESeq2 processing

```{r}
# Create DESeqDataset
ddsSE_1 <- DESeqDataSet(full.selected, design = ~ KRT5_7_score_anno)

# Filter genes
summary(rowSums(counts(ddsSE_1)) >= 10)
keep <- rowSums(counts(ddsSE_1)) >= 10
ddsSE_1 <- ddsSE_1[keep, ]

# DESeq analysis includes count normalization based on size factors
dds <- DESeq(ddsSE_1, )
rownames(dds) <- rowData(dds)$gene_name

# Extract results - also performs independent filtering
res <- DESeq2::results(dds) #alpha = 0.05, lfcThreshold = 1)
summary(res)
res

# Variance-stabilizing transformation
vsd <- DESeq2::vst(dds)
```


```{r}
saveRDS(full.selected,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_TumorANDNormal_BL.rds")
saveRDS(vsd,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_TumorANDNormal_BL_vsd.rds")
saveRDS(res,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_TumorANDNormal_BL_res.rds")
```

```{r}
full.clean <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_tumorANDnormal.rds")
full.selected <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_TumorANDNormal_BL.rds")
vsd <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_TumorANDNormal_BL_vsd.rds")
res <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_TumorANDNormal_BL_res.rds")
```

### GSEA 

With `clusterProfiler`. 

```{r}
organism <- "org.Hs.eg.db"

# Prepare input gene list based on log2foldchanges 
original_gene_list <- res$log2FoldChange
names(original_gene_list) <- rownames(res)
gene_list <- na.omit(original_gene_list)
gene_list <- sort(gene_list, decreasing = TRUE) # sort in decreasing order

# Run GO analysis
gse <- gseGO(geneList=gene_list, 
             ont ="BP", 
             keyType = "SYMBOL", 
             minGSSize = 50, 
             maxGSSize = 1000, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = organism, 
             pAdjustMethod = "BH")

# Visualization
## Dot plot
dotplot(gse, showCategory=5, split=".sign") + facet_grid(~.sign)

## Emap plot
emapplot(pairwise_termsim(gse), showCategory = 10, color = "enrichmentScore", layout = "circle")

## GSEA plot
gseaplot2(gse, title = gse$Description[2], geneSetID = 1)

out <- data.frame(name = gse$Description, score = gse$NES, gse$p.adjust)
```

### KEGG pathway enrichment 

With `clusterProfiler`. 

```{r}
# KEGG requires Entrez IDs 
entrez.ids <- mapIds(org.Hs.eg.db, keys=rownames(res), column="ENTREZID", keytype="SYMBOL")
entrez.ids <- entrez.ids[!duplicated(entrez.ids)]

df <- data.frame(log_change = gene_list, 
           SYMBOL = names(gene_list))

df$ENTREZ <- entrez.ids[match(df$SYMBOL, names(entrez.ids))]
df <- df[!is.na(df$ENTREZ),]

gene_list_kegg <- df$log_change
names(gene_list_kegg) <- df$ENTREZ

# Run KEGG
gseKEGG <- gseKEGG(geneList=gene_list_kegg, 
        organism = "hsa",
        keyType = "kegg", 
        minGSSize = 50, 
        maxGSSize = 500, 
        pvalueCutoff = 0.05, 
        verbose = TRUE, 
        pAdjustMethod = "BH")

# Visualization 
## Dotplot 
dotplot(gseKEGG, showCategory = 5, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)

## EMAP plot 
emapplot(pairwise_termsim(gseKEGG), showCategory = 15,color = "enrichmentScore", layout = "circle")

## GSEA plot
gseaplot2(gseKEGG, title = gseKEGG$Description[4], geneSetID = 1)
```

### Volcano-plot

```{r}
#rownames(reshape) <- rowData(brca.clean)$gene_name[match(rownames(reshape), rownames(brca.clean))]

sig.up <- res %>% as.data.frame %>% filter(log2FoldChange > 1) %>% arrange(padj) %>% slice(1:30) %>% rownames_to_column("gene") %>% pull(gene)
sig.down <- res %>% as.data.frame %>% filter(log2FoldChange < -1) %>% arrange(padj) %>% slice(1:30) %>% rownames_to_column("gene") %>% pull(gene)

plot_vol <- res %>% as.data.frame %>% rownames_to_column("gene")
plot_vol <- plot_vol %>% mutate(type = ifelse(log2FoldChange > 1 & padj < 0.01, "UP", 
                                  ifelse(log2FoldChange < -1 & padj < 0.01, "DOWN","NA")))

col <- c("DOWN" = "#E41A1C", "UP" = "#377EB8", "NA" = "black")

# Combined 
ggplot(plot_vol, aes(x = log2FoldChange, y=-log10(padj)))+
  geom_point(aes(color = type))+
  geom_label_repel(data = plot_vol %>% filter(gene %in% sig.down), aes(label = gene, color = type), max.overlaps = 30)+
  geom_label_repel(data = plot_vol %>% filter(gene %in% sig.up), aes(label = gene, color = type), max.overlaps = 30)+
  xlab("log2FC (CK5/7+ vs Not_CK5/7+)")+
  ylab("-log10(FDR)")+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed")+
  geom_hline(yintercept = -log10(1), linetype = "dashed")+
  geom_vline(xintercept = 1, linetype = "dashed")+
  geom_vline(xintercept = -1, linetype = "dashed")+
  theme_classic()+
  scale_color_manual(values = col)+
  guides(color = "none")

# Down-regulated
ggplot(plot_vol, aes(x = log2FoldChange, y=-log10(padj)))+
  geom_point(aes(color = type))+
  geom_label_repel(data = plot_vol %>% filter(gene %in% sig.down), aes(label = gene, color = type), max.overlaps = 30)+
  xlab("log2FC (CK5/7+ vs Not_CK5/7+)")+
  ylab("-log10(FDR)")+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed")+
  geom_hline(yintercept = -log10(1), linetype = "dashed")+
  geom_vline(xintercept = 1, linetype = "dashed")+
  geom_vline(xintercept = -1, linetype = "dashed")+
  theme_classic()+
  scale_color_manual(values = col)+
  guides(color = "none")

# Up-regulated
ggplot(plot_vol, aes(x = log2FoldChange, y=-log10(padj)))+
  geom_point(aes(color = type))+
  geom_label_repel(data = plot_vol %>% filter(gene %in% sig.up), aes(label = gene, color = type), max.overlaps = 30)+
  xlab("log2FC (CK5/7+ vs Not_CK5/7+)")+
  ylab("-log10(FDR)")+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed")+
  geom_hline(yintercept = -log10(1), linetype = "dashed")+
  geom_vline(xintercept = 1, linetype = "dashed")+
  geom_vline(xintercept = -1, linetype = "dashed")+
  theme_classic()+
  scale_color_manual(values = col)+
  guides(color = "none")
```

### Age comparison 

```{r}
col_NT <- metadata(vsd_tumor)$colors$KRT57_score
names(col_NT) <- paste0("NT_", names(metadata(vsd_tumor)$colors$KRT57_score))

age <- colData(full.selected) %>% as.data.frame %>% dplyr::select(age_at_index, KRT5_7_score_anno) 

ggplot(age, aes(x = KRT5_7_score_anno, y = age_at_index))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score_anno), size = 3, position=position_jitter(w=0.1,h=0.1))+
  stat_compare_means()+
  theme_classic()+
  scale_color_brewer(palette = "Set1")


age <- colData(full.clean) %>% as.data.frame %>% dplyr::select(age_at_index, KRT5_7_score_anno, broad_cat) %>% filter(broad_cat == "NT")

ggplot(age, aes(x = reorder(KRT5_7_score_anno,age_at_index, FUN = mean), y = age_at_index))+
  #facet_grid(~broad_cat)+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score_anno), size = 3, position=position_jitter(w=0.1,h=0.1))+
  stat_compare_means()+
  theme_classic()+
  scale_color_manual(values = col_NT)
```


### Gene signatures

#### Chromosomal instability

```{r}
chrom_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/ChromosomalInstability_Carter_NatGen06.xlsx",sheet = 1, col_names = FALSE)

# Extract unique genes that were up- or down-regulated in each signature 
chrom_sig_up <- chrom_sig$...1 %>% unique()
```

```{r}
## 1. pan AV signature per PID group 
chrom_sig_up_rel <- chrom_sig_up[chrom_sig_up %in% rownames(assay(vsd))]
# Filter for available genes
#TNBC_PID <- colnames(vsd[,vsd$TNBC == "TNBC"])
#vsd_TNBC <- vsd[, colnames(vsd) %in% TNBC_PID]
#df <- t(assay(vsd_TNBC)[chrom_sig_up_rel ,]) %>% as.data.frame
df <- t(assay(vsd)[chrom_sig_up_rel ,]) %>% as.data.frame

# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score_anno[match(rownames(df), colnames(vsd))]
#df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = reorder(KRT5_7_score,signature_sum, FUN = median), y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_brewer(palette = "Set1")+
  theme_classic()+
  ylab("Chromosomal instability (Carter et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```


### Gray et al 

```{r}
## 2. BL signature per PID group 
BL_sig_up_rel <- BL_sig_up[BL_sig_up %in% rownames(assay(vsd))]

# Filter for available genes
df <- t(assay(vsd)[BL_sig_up_rel ,]) %>% as.data.frame

#df <- t(assay(vsd)[BL_sig_up_rel ,]) %>% as.data.frame

# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score_anno[match(rownames(df), colnames(vsd))]
#df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

#compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
#compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_brewer(palette = "Set1")+
  theme_classic()+
  ylab("Basal-luminal signature (Li et al.)")+
  #geom_signif(
  #  comparisons = list(c("KRT5/7+", "KRT5+"), c("KRT5/7+", "KRT5/7-"),c("KRT5/7+", "KRT7+")),
  #  map_signif_level = FALSE, step_increase = 0.1, test = "t.test"
  #)+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

# Save objects

```{r}
saveRDS(full.selected,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_TumorANDNormal_BL.rds")
saveRDS(vsd,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_TumorANDNormal_BL_vsd.rds")
saveRDS(res,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_TumorANDNormal_BL_res.rds")
```

# Session Info 

```{r}
sessionInfo()
```
