---
title: "TCGA_BRCA_RNAseq_analysis"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: TCGA_DET.rds, TCGA_RES.rds, TCGA_DS.rds, TCGA_QS.rds, GeneSignatures.rds 

# Validation

Data analysis scripts for data analysis in the validation cohorts. 

# Script: TCGA BRCA - Tumor analysis

Here, we are analyzing RNA-seq data from the TCGA BRCA cohort.

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(scuttle)
library(scater)
library(scran)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(SummarizedExperiment)
library(TCGAbiolinks)
library(DESeq2)
library(gridExtra)
library(ggVennDiagram)
library(apeglm)
library(survival)
library(survminer)
library(readxl)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(ExploreModelMatrix)
library(ggvenn)
library(wesanderson)
library(immunedeconv)
library(limma)
library(easyalluvial)
library(ggalluvial)
library(ggpubr)
```

# Load data 

```{r}
query.exp <- GDCquery(
    project = "TCGA-BRCA", 
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification", 
    workflow.type = "STAR - Counts",
    sample.type = c("Primary Tumor")
)

GDCdownload(
    query = query.exp,
    files.per.chunk = 100
)

brca.exp <- GDCprepare(
    query = query.exp, 
    save = TRUE, 
    save.filename = "brcaExp.rda"
)
```

## Add clinical/survival info

```{r}
# Clean object
brca.clean <- brca.exp
colData(brca.clean) <- NULL
brca.clean$PID <- brca.exp$PID

# 1. Add clinical information 
## Cleaned clinical information from UCSC Xena
information.subtype.xena <- read_delim("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/BRCA_clinicalinfo.txt")
information.subtype.xena <- information.subtype.xena %>% as.data.frame

## Add relevant data based on PID
for (entry in colnames(information.subtype.xena)[colnames(information.subtype.xena) %in% c("ER_Status_nature2012","PR_Status_nature2012","HER2_Final_Status_nature2012","Metastasis_nature2012","Tumor_nature2012","Node_nature2012","PAM50Call_RNAseq","PAM50_mRNA_nature2012", "Age_at_Initial_Pathologic_Diagnosis_nature2012")]) {
  colData(brca.clean)[,entry] <- information.subtype.xena[,entry][match(colData(brca.clean)[,"PID"],information.subtype.xena$patient_id)]
}

# 2. Add survival information 
## Cleaned survival information from UCSC Xena
information.survival_xena <- read_delim("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/BRCA_survival_Xena.txt")
information.survival_xena$PID <- str_split(information.survival_xena$sample,"-",simplify = TRUE)[,3]
information.survival_xena <- information.survival_xena %>% as.data.frame

for (entry in colnames(information.survival_xena)[!colnames(information.survival_xena) %in% c("sample","_PATIENT","Redaction","PID")]) {
  colData(brca.clean)[,entry] <- information.survival_xena[,entry][match(colData(brca.clean)[,"PID"],information.survival_xena$PID)]
}

colData(brca.clean)
```

### Check dataset for TNBC and basal-like

```{r}
brca.clean$TNBC <- ifelse(brca.clean$ER_Status_nature2012 == "Negative" & brca.clean$PR_Status_nature2012 == "Negative" & brca.clean$HER2_Final_Status_nature2012 == "Negative", "TNBC", "Not_TNBC")
brca.clean$TNBC[is.na(brca.clean$TNBC)] <- "Undefined"

# TNBC 
tnbc.clean <- brca.clean[,brca.clean$TNBC == "TNBC"]
tnbc.clean 

# Basal
brca.clean$PAM50Call_RNAseq[is.na(brca.clean$PAM50Call_RNAseq)] <- "Undefined"
basal.clean <- brca.clean[,brca.clean$PAM50Call_RNAseq == "Basal"]
basal.clean
```

```{r}
saveRDS(brca.clean,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_cleaned_full.rds")

brca.clean <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_cleaned_full.rds")
```


## DESeq2 processing

Full dataset: 1106 patients 

```{r}
brca.clean$PAM50Call_RNAseq[is.na(brca.clean$PAM50Call_RNAseq)] <- "Undefined"

# Create DESeqDataset
ddsSE <- DESeqDataSet(brca.clean, design = ~ PAM50Call_RNAseq)
ddsSE

# DESeq analysis includes count normalization based on size factors
res <- DESeq(ddsSE)
rownames(res) <- rowData(res)$gene_name

DESeq2::plotCounts(res, gene = "KRT7", intgroup = "PAM50Call_RNAseq")
DESeq2::plotCounts(res, gene = "KRT5", intgroup = "PAM50Call_RNAseq")

counts(res, normalized = TRUE) # Access normalized counts 

# Variance-stabilizing transformation
vsd <- DESeq2::vst(res)

# Access normalized and stabilized counts
mat <- assay(vsd)
```

```{r}
res <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_PAM50_res.rds") 
vsd <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_PAM50_vsd.rds")
```

## CK5/7 Signature 

GMM on size-factor normalized (NOT VST) counts

```{r}
# 1. CK5
GMM_CK5 <- Mclust(counts(res, normalized = TRUE)["KRT5",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK5$classification == 2, 
                                        "KRT5+",
                                        "KRT5-"), 
                     KRT5_exprs = assay(vsd)["KRT5",],
                     uncertainty = GMM_CK5$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = KRT5_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")

#Add GMM epi class to object
vsd$GMM_KRT5 <- GMM_df$class_epi

# 2. CK7
GMM_CK7 <- Mclust(counts(res, normalized = TRUE)["KRT7",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK7$classification == 2, 
                                        "KRT7+",
                                        "KRT7-"), 
                     KRT7_exprs = assay(vsd)["KRT7",],
                     uncertainty = GMM_CK7$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = KRT7_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 17)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")

#Add GMM epi class to object
vsd$GMM_KRT7 <- GMM_df$class_epi

## Combine KRT5/7 GMM scores 
vsd$KRT5_7_score <- ifelse(vsd$GMM_KRT5 == "KRT5+" & vsd$GMM_KRT7 == "KRT7+", "KRT5/7+", 
       ifelse(vsd$GMM_KRT5 == "KRT5+" & vsd$GMM_KRT7 == "KRT7-", "KRT5+",
              ifelse(vsd$GMM_KRT5 == "KRT5-" & vsd$GMM_KRT7 == "KRT7+", "KRT7+","KRT5/7-")))

## Add colors 
metadata(vsd)$colors$KRT57_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("KRT5/7-", "KRT5/7+","KRT5+", "KRT7+"))
metadata(vsd)$colors$PAM50 <- setNames(brewer.pal(6,"Paired"),sort(unique(vsd$PAM50Call_RNAseq)))

## Visualize 

### 1. Normalized counts
plot <- as.data.frame(t(counts(res, normalized = TRUE)[rownames(counts(res, normalized = TRUE)) %in% c("KRT5","KRT7"),]))
plot$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(plot),colnames(vsd))]
plot$PAM50Call_RNAseq <- vsd$PAM50Call_RNAseq[match(rownames(plot),colnames(vsd))]
plot$TNBC <- vsd$TNBC[match(rownames(plot),colnames(vsd))]
plot$GMM_KRT5 <- vsd$GMM_KRT5[match(rownames(plot),colnames(vsd))]
plot$GMM_KRT7 <- vsd$GMM_KRT7[match(rownames(plot),colnames(vsd))]
```

```{r}
brca.clean$KRT5_7_score <- vsd$KRT5_7_score[match(brca.clean$PID, vsd$PID)]
brca.clean$KRT5_7_score_combined <- ifelse(brca.clean$KRT5_7_score == "KRT5/7+","KRT5_7_pos","Not_KRT5_7_pos")
```

```{r}
main <- ggplot(plot, aes(x=KRT7, y=KRT5, color = KRT5_7_score)) +
  geom_point(size = 2, alpha = 0.5)+
  theme_classic()+
  xlab("CK7 normalized counts")+
  ylab("CK5 normalized counts")+
  coord_cartesian(xlim = c(0,120000), ylim = c(0,300000))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  guides(color = "none")

y_side <- ggplot(plot, aes(x = KRT5, fill=GMM_KRT5)) + 
  geom_density(alpha=0.8) + 
  scale_fill_manual(values = c('#999999','#F4B5BD')) + 
  theme_classic()+
  theme(legend.position = "none", axis.title.y = element_blank())+
  coord_flip(xlim = c(0,300000), ylim = c(0, 0.00005))

x_side <- ggplot(plot, aes(x = KRT7, fill=GMM_KRT7)) + 
  geom_density(alpha=0.8) + 
  scale_fill_manual(values = c('#999999','#9C964A')) + 
  coord_cartesian(xlim = c(0,120000))+
  theme_classic()+
  theme(legend.position = "none", axis.title.x = element_blank())

blankPlot <- ggplot()+geom_blank(aes(1,1)) + 
  cowplot::theme_nothing()

grid.arrange(x_side, blankPlot, main, y_side, 
        ncol=2, nrow=2, widths=c(4, 1.4), heights=c(1.4, 4))
```


```{r}
saveRDS(vsd,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_PAM50_vsd.rds")
saveRDS(res,"/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_PAM50_res.rds")
#res <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_PAM50_res.rds") 
#vsd <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_PAM50_vsd.rds")
```


## 1. DEA - combined

```{r}
brca.clean <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_cleaned_full.rds")
```

```{r}
# Create DESeqDataset
ddsSE_1 <- DESeqDataSet(brca.clean, design = ~ KRT5_7_score_combined)

# Filter genes
summary(rowSums(counts(ddsSE_1)) >= 10)
keep <- rowSums(counts(ddsSE_1)) >= 10
ddsSE_1 <- ddsSE_1[keep, ]

# DESeq analysis includes count normalization based on size factors
dds <- DESeq(ddsSE_1, )

# Extract results - also performs independent filtering
res <- DESeq2::results(dds) #alpha = 0.05, lfcThreshold = 1)
summary(res)

# Variance-stabilizing transformation
vsd <- DESeq2::vst(dds)
```

```{r}
#Save objects 
saveRDS(res, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_full_CombinedScore.rds")
saveRDS(vsd, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_full_Combined_VST.rds")

vsd <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_full_Combined_VST.rds")

## Add colors 
metadata(vsd)$colors$KRT57_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("KRT5/7-", "KRT5/7+","KRT5+", "KRT7+"))
metadata(vsd)$colors$PAM50 <- setNames(brewer.pal(6,"Paired"),sort(unique(vsd$PAM50Call_RNAseq)))
rownames(vsd) <- rowData(vsd)$gene_name
```

### TNBC only 

```{r}
tnbc.clean <- brca.clean[,brca.clean$TNBC == "TNBC"]

# Create DESeqDataset
ddsSE_1 <- DESeqDataSet(tnbc.clean, design = ~ KRT5_7_score_combined)

# Filter genes
summary(rowSums(counts(ddsSE_1)) >= 10)
keep <- rowSums(counts(ddsSE_1)) >= 10
ddsSE_1 <- ddsSE_1[keep, ]

# DESeq analysis includes count normalization based on size factors
dds <- DESeq(ddsSE_1, )

# Extract results - also performs independent filtering
res <- DESeq2::results(dds) #alpha = 0.05, lfcThreshold = 1)
summary(res)

# Variance-stabilizing transformation
vsd <- DESeq2::vst(dds)
```

```{r}
#Save objects (TNBC ONLY)
saveRDS(res, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_TNBC_CombinedScore.rds")
saveRDS(vsd, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_TNBC_Combined_VST.rds")

## Add colors 
metadata(vsd)$colors$KRT57_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("KRT5/7-", "KRT5/7+","KRT5+", "KRT7+"))
metadata(vsd)$colors$PAM50 <- setNames(brewer.pal(6,"Paired"),sort(unique(vsd$PAM50Call_RNAseq)))
rownames(vsd) <- rowData(vsd)$gene_name
```

#### Heatmap visualization 

```{r}
res <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_full_CombinedScore.rds")
vsd <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_full_Combined_VST.rds")
brca.clean <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_cleaned_full.rds")

res_TNBC <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_TNBC_CombinedScore.rds")
rownames(res_TNBC) <- rowData(brca.clean)$gene_name[match(rownames(res_TNBC), rowData(brca.clean)$gene_id)]
vsd_TNBC <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_TNBC_Combined_VST.rds")
```

```{r}
#res <- res_TNBC
rownames(brca.clean) <- rowData(brca.clean)$gene_name

# Extract Top 20
rownames(res) <- rowData(brca.clean)$gene_name[match(rownames(res), rownames(brca.clean))]

sig.up <- res %>% as.data.frame %>% filter(log2FoldChange > 1) %>% arrange(padj) %>% slice(1:30) %>% rownames_to_column("gene") %>% pull(gene)
sig.down <- res %>% as.data.frame %>% filter(log2FoldChange < -1) %>% arrange(padj) %>% slice(1:30) %>% rownames_to_column("gene") %>% pull(gene)

# Heatmap body 
ht <- assay(vsd) %>% as.data.frame %>% rownames_to_column("gene")
ht$gene <- rowData(brca.clean)$gene_name[match(ht$gene, rownames(brca.clean))]
mat_up <- ht %>% filter(gene %in% sig.up) %>% column_to_rownames("gene")
mat_down <- ht %>% filter(gene %in% sig.down) %>% column_to_rownames("gene")

# Annotation
anno <- as.data.frame(t(mat_up))
anno$KRT5_7_score_combined <- brca.clean$KRT5_7_score_combined[match(rownames(anno),colnames(brca.clean))]
anno$KRT5_7_score <- brca.clean$KRT5_7_score[match(rownames(anno),colnames(brca.clean))]
anno$PAM50 <- brca.clean$PAM50Call_RNAseq[match(rownames(anno),colnames(brca.clean))]
anno$TNBC <- brca.clean$TNBC[match(rownames(anno),colnames(brca.clean))]

# Plot Heatmap
Heatmap(mat_up, 
        col = viridis(100),
        column_split = anno$KRT5_7_score_combined, 
        column_title = NULL)%v%
  columnAnnotation(KRT5_7_score_combined = anno$KRT5_7_score_combined)%v%
  columnAnnotation(KRT5_7_score = anno$KRT5_7_score)%v%
  columnAnnotation(PAM50 = anno$PAM50)%v%
  columnAnnotation(TNBC = anno$TNBC)

Heatmap(mat_down, 
        col = viridis(100),
        column_split = anno$KRT5_7_score_combined, 
        column_title = NULL)%v%
  columnAnnotation(KRT5_7_score_combined = anno$KRT5_7_score_combined)%v%
  columnAnnotation(KRT5_7_score = anno$KRT5_7_score)%v%
  columnAnnotation(PAM50 = anno$PAM50)%v%
  columnAnnotation(TNBC = anno$TNBC)
```

#### Volcano-plot visualization 

```{r}
## LFC shrinkage using apeglm 
DESeq2::plotMA(res, ylim = c(-5, 5))

plot_vol <- res %>% as.data.frame %>% rownames_to_column("gene")

DESeq2::resultsNames(dds)

reshape <- DESeq2::lfcShrink(dds, coef = "KRT5_7_score_combined_Not_KRT5_7_pos_vs_KRT5_7_pos", 
                            type = "apeglm")

DESeq2::plotMA(reshape, ylim = c(-5, 5))
```

```{r}
#rownames(reshape) <- rowData(brca.clean)$gene_name[match(rownames(reshape), rownames(brca.clean))]

plot_vol <- res %>% as.data.frame %>% rownames_to_column("gene")
plot_vol <- plot_vol %>% mutate(type = ifelse(log2FoldChange > 1 & padj < 0.01, "UP", 
                                  ifelse(log2FoldChange < -1 & padj < 0.01, "DOWN","NA")))

col <- c("DOWN" = "#E41A1C", "UP" = "#377EB8", "NA" = "black")

# Combined 
ggplot(plot_vol, aes(x = log2FoldChange, y=-log10(padj)))+
  geom_point(aes(color = type))+
  geom_label_repel(data = plot_vol %>% filter(gene %in% sig.down), aes(label = gene, color = type), max.overlaps = 30)+
  geom_label_repel(data = plot_vol %>% filter(gene %in% sig.up), aes(label = gene, color = type), max.overlaps = 30)+
  xlab("log2FC (CK5/7+ vs Not_CK5/7+)")+
  ylab("-log10(FDR)")+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed")+
  geom_hline(yintercept = -log10(1), linetype = "dashed")+
  geom_vline(xintercept = 1, linetype = "dashed")+
  geom_vline(xintercept = -1, linetype = "dashed")+
  theme_classic()+
  scale_color_manual(values = col)+
  guides(color = "none")

# Down-regulated
ggplot(plot_vol, aes(x = log2FoldChange, y=-log10(padj)))+
  geom_point(aes(color = type))+
  geom_label_repel(data = plot_vol %>% filter(gene %in% sig.down), aes(label = gene, color = type), max.overlaps = 30)+
  xlab("log2FC (CK5/7+ vs Not_CK5/7+)")+
  ylab("-log10(FDR)")+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed")+
  geom_hline(yintercept = -log10(1), linetype = "dashed")+
  geom_vline(xintercept = 1, linetype = "dashed")+
  geom_vline(xintercept = -1, linetype = "dashed")+
  theme_classic()+
  scale_color_manual(values = col)+
  guides(color = "none")

# Up-regulated
ggplot(plot_vol, aes(x = log2FoldChange, y=-log10(padj)))+
  geom_point(aes(color = type))+
  geom_label_repel(data = plot_vol %>% filter(gene %in% sig.up), aes(label = gene, color = type), max.overlaps = 30)+
  xlab("log2FC (CK5/7+ vs Not_CK5/7+)")+
  ylab("-log10(FDR)")+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed")+
  geom_hline(yintercept = -log10(1), linetype = "dashed")+
  geom_vline(xintercept = 1, linetype = "dashed")+
  geom_vline(xintercept = -1, linetype = "dashed")+
  theme_classic()+
  scale_color_manual(values = col)+
  guides(color = "none")
```

#### Boxplot visualization for GOI 

```{r}
p1 <- ggplot(anno, aes(x = KRT5_7_score_combined, y = TBC1D9))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = KRT5_7_score_combined), size = 3, position=position_jitter(w=0.1,h=0.1))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90))+
  xlab("")+
  #stat_compare_means(method = "kruskal.test")+
  scale_color_brewer(palette = "Set1")+
  guides(color = "none")

p2 <- ggplot(anno, aes(x = KRT5_7_score, y = TBC1D9))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = KRT5_7_score_combined), size = 3, position=position_jitter(w=0.1,h=0.1))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("")+
  xlab("")+
  scale_color_brewer(palette = "Set1")+
  guides(color = "none")
  
p3 <- ggplot(anno, aes(x = TNBC, y = TBC1D9))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = KRT5_7_score_combined), size = 3, position=position_jitter(w=0.1,h=0.1))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("")+
  xlab("")+
  scale_color_brewer(palette = "Set1")+
  guides(color = "none")


p1+p2+p3
```

## 2. Survival analysis 

### Kaplan-Meier

```{r}
#Assess survival impact 
cur_surv <- colData(vsd) %>% as.data.frame() %>% dplyr::select(PID, KRT5_7_score, DFI, DFI.time, OS, OS.time, DSS, DSS.time, PFI, PFI.time, TNBC)

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ KRT5_7_score, data = cur_surv)
print(fit)

metadata(vsd)$colors$KRT57_score 

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("grey","#85D4E3","#F4B5BD","#9C964A"),
          log.rank.weights = "n", 
          xlim = c(0, 4000))

#Plot survival curves
fit_1 <- survfit(Surv(PFI.time, PFI) ~ KRT5_7_score, data = cur_surv)
print(fit_1)

ggsurvplot(fit_1,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("grey","#85D4E3","#F4B5BD","#9C964A"),
          log.rank.weights = "n",
          xlim = c(0, 4000))
```

### Cox model 

```{r}
vsd$age <- vsd$Age_at_Initial_Pathologic_Diagnosis_nature2012
```

Multivariate 

```{r}
## 1. Disease-free interval 
### Multivariate cox proportional hazard model on multiple covariates (here KRT5/7 groups)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% dplyr::select(KRT5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "KRT5_7_score",values_from = "fra") %>% column_to_rownames("PID")

colnames(PID_metagroups) <- c("KRT57negative","KRT57positive","KRT5positive","KRT7positive")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$DFI <- vsd$DFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$DFI.time <- vsd$DFI.time[match(rownames(Cox_df),vsd$PID)]

#Cox_df$PAM50Call_RNAseq <- as.factor(vsd$PAM50Call_RNAseq[match(rownames(Cox_df),vsd$PID)])
Cox_df$age <- as.numeric(vsd$age[match(rownames(Cox_df),vsd$PID)])
Cox_df$pT_simple <- as.factor(vsd$Tumor_nature2012[match(rownames(Cox_df),vsd$PID)])
Cox_df$pN_simple <- as.factor(vsd$Node_nature2012[match(rownames(Cox_df),vsd$PID)])
Cox_df$pM_simple <- as.factor(vsd$Metastasis_nature2012[match(rownames(Cox_df),vsd$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-DFI,-DFI.time, -age, -pT_simple, -pN_simple, -pM_simple, - KRT57negative) %>% colnames

multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~ age + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "), "+ KRT57negative"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
#coord_cartesian(ylim=c(0.1,100))+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

Univariate

```{r}
## 1. Disease-free interval 
### Multivariate cox proportional hazard model on multiple covariates (here KRT5/7 groups)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% dplyr::select(KRT5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "KRT5_7_score",values_from = "fra") %>% column_to_rownames("PID")

colnames(PID_metagroups) <- c("KRT57negative","KRT57positive","KRT5positive","KRT7positive")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$DFI <- vsd$DFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$DFI.time <- vsd$DFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-DFI,-DFI.time, - KRT57negative) %>% colnames

multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ KRT57negative"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,5))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
## 2. Progression-free interval 
### Multivariate cox proportional hazard model on multiple covariates (here KRT5/7 groups)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% dplyr::select(KRT5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "KRT5_7_score",values_from = "fra") %>% column_to_rownames("PID")

colnames(PID_metagroups) <- c("KRT57negative","KRT57positive","KRT5positive","KRT7positive")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$PFI <- vsd$PFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$PFI.time <- vsd$PFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-PFI,-PFI.time, - KRT57negative) %>% colnames

multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ KRT57negative"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,5))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

Multivariate 

````{r}
## 2. Progression-free interval 
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% dplyr::select(KRT5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "KRT5_7_score",values_from = "fra") %>% column_to_rownames("PID")

colnames(PID_metagroups) <- c("KRT57negative","KRT57positive","KRT5positive","KRT7positive")

## 1. Based on SCP assignments
Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$PFI <- vsd$PFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$PFI.time <- vsd$PFI.time[match(rownames(Cox_df),vsd$PID)]

Cox_df$age <- as.numeric(vsd$age[match(rownames(Cox_df),vsd$PID)])
Cox_df$pT_simple <- as.factor(vsd$Tumor_nature2012[match(rownames(Cox_df),vsd$PID)])
Cox_df$pN_simple <- as.factor(vsd$Node_nature2012[match(rownames(Cox_df),vsd$PID)])
Cox_df$pM_simple <- as.factor(vsd$Metastasis_nature2012[match(rownames(Cox_df),vsd$PID)])

#Based on proportions
covariates <- Cox_df %>% select(-PFI,-PFI.time, -age, -pT_simple, -pN_simple, -pM_simple, - KRT57negative) %>% colnames

multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~ age + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "), "+ KRT57negative"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratios
hr_prop_PFS <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop_PFS
```

Univariate

```{r}
## 2. Progression-free interval 
### Univariate cox proportional hazard model on multiple covariates (here KRT5/7 groups)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% dplyr::select(KRT5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "KRT5_7_score",values_from = "fra") %>% column_to_rownames("PID")

colnames(PID_metagroups) <- c("KRT57negative","KRT57positive","KRT5positive","KRT7positive")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$PFI <- vsd$PFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$PFI.time <- vsd$PFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-PFI,-PFI.time, -KRT57negative) %>% colnames

multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ KRT57negative"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
coord_cartesian(ylim=c(0,5))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```


## 3. Barplot visualization 

```{r}
dittoBarPlot(vsd, var = "KRT5_7_score", group.by = "TNBC")+
    scale_fill_manual(values = metadata(vsd)$colors$KRT57_score)
dittoBarPlot(vsd, var = "KRT5_7_score", group.by = "TNBC", scale = "count")+
  scale_fill_manual(values = metadata(vsd)$colors$KRT57_score)


dittoBarPlot(vsd, var = "KRT5_7_score", group.by = "PAM50Call_RNAseq")+
    scale_fill_manual(values = metadata(vsd)$colors$KRT57_score)
dittoBarPlot(vsd, var = "KRT5_7_score", group.by = "PAM50Call_RNAseq", scale = "count")+
      scale_fill_manual(values = metadata(vsd)$colors$KRT57_score)

dittoBarPlot(vsd, var = "PAM50Call_RNAseq", group.by = "KRT5_7_score")+
        scale_fill_manual(values = metadata(vsd)$colors$PAM50)
dittoBarPlot(vsd, var = "PAM50Call_RNAseq", group.by = "KRT5_7_score", scale = "count")+
  scale_fill_manual(values = metadata(vsd)$colors$PAM50)

dittoBarPlot(vsd, var = "TNBC", group.by = "KRT5_7_score")
dittoBarPlot(vsd, var = "TNBC", group.by = "KRT5_7_score", scale = "count")
```

```{r}
# Chi-square test to compare between TNBC status between KRT5/7 scores
bar_out <- dittoBarPlot(vsd, var = "KRT5_7_score", group.by = "TNBC", scale = "count", data.out = TRUE)$data

bar_test <- bar_out %>% filter(!grouping %in% "Undefined") %>% pivot_wider(names_from = grouping, id_cols = label, values_from = count) %>% column_to_rownames("label")

bar_test_out <- chisq.test(bar_test)

print(bar_test_out$p.value)
```


## 4. Keratin expression 

```{r}
# Define keratins of interest
keratins <- c("KRT5","KRT7","KRT15","KRT8","KRT18","KRT19", "KRT14","KRT17","ITGA6","EPCAM")

# Define heatmap based on normalized counts 
df <- t(assay(vsd)[keratins,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")

## Create heatmap body 
df$KRT5_7_score <- vsd$KRT5_7_score[match(df$PID, colnames(vsd))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(ht, name = "Mean Z-scores")

Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

## 5. Stem-like marker expression

```{r}
# Define stem markers of interest
stem <- c("SOX9","KIT","PROM1","YAP1","HDAC2","NOTCH1","CD44","ALDH1A3","BRCA1","BRCA2", "TGFB2", "ANXA8","MUC1", "CD24", "TNFAIP3", "ERBB3", "TNFRSF11A")

stem_rel <- stem[stem %in% rownames(assay(vsd))]

# Define heatmap based on normalized counts 
df <- t(assay(vsd)[stem_rel,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")

## Create heatmap body 
df$KRT5_7_score <- vsd$KRT5_7_score[match(df$PID, colnames(vsd))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(ht, name = "Mean Z-scores")

Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

```{r}
## BRCA1 expression
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))
compare_means(BRCA1 ~ KRT5_7_score,df) #pairwise comparisons
compare_means(BRCA1 ~ KRT5_7_score,df, method = "kruskal.test") #global comparison
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

ggplot(df, aes(x = KRT5_7_score, y = SOX9))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("SOX9 expression")
```

## 6. Lim et al signatures

```{r}
lum_progenitor_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/LuminalProgenitor_Signature_Limetal_NatureMedicine_2009.xls")
mature_lum_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/MatureLuminalSignature_Limetal_NatureMedicine_2009.xls")
basal_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Basal_Signature_Limetal_NatureMedicine_2009.xls")

# Extract unique genes that were up- or down-regulated in each signature 
lum_progenitor_sig_up <- lum_progenitor_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
lum_progenitor_sig_up <- lum_progenitor_sig_up[!is.na(lum_progenitor_sig_up)]
lum_progenitor_sig_down <- lum_progenitor_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
lum_progenitor_sig_down <- lum_progenitor_sig_down[!is.na(lum_progenitor_sig_down)]

mature_lum_sig_up <- mature_lum_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
mature_lum_sig_up <- mature_lum_sig_up[!is.na(mature_lum_sig_up)]
mature_lum_sig_down <- mature_lum_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
mature_lum_sig_down <- mature_lum_sig_down[!is.na(mature_lum_sig_down)]

basal_sig_up <- basal_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
basal_sig_up <- basal_sig_up[!is.na(basal_sig_up)]
basal_sig_down <- basal_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
basal_sig_down <- basal_sig_down[!is.na(basal_sig_down)]
```

Upregulated genes per signature 

```{r}
## 1. Luminal progenitor signature per PID group 
lum_progenitor_sig_up_rel <- lum_progenitor_sig_up[lum_progenitor_sig_up %in% rownames(assay(vsd))]
# Filter for available genes
df <- t(assay(vsd)[lum_progenitor_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Luminal progenitor signature")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

```{r}
#Mature luminal signature per PID group 
mature_lum_sig_up_rel <- mature_lum_sig_up[mature_lum_sig_up %in% rownames(assay(vsd))]

df <- t(assay(vsd)[mature_lum_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Mature luminal signature")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

```{r}
#Basal signature per PID group 
basal_sig_up_rel <- basal_sig_up[basal_sig_up %in% rownames(assay(vsd))]

df <- t(assay(vsd)[basal_sig_up_rel ,]) %>% as.data.frame
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Basal signature")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```


## 7. Gray et al signatures for PID groups 

A: Alveolar subtypes 

```{r}
AV_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Lietal_AV_BL_Signatures_reduced.xlsx",sheet = 1)

colnames(AV_sig) <- c("pan_AV", "BL_enriched","AP_enriched")

# Extract unique genes that were up- or down-regulated in each signature 
panAV_sig_up <- AV_sig$pan_AV
panAV_sig_up <- panAV_sig_up[!is.na(panAV_sig_up)]

BL_sig_up <- AV_sig$BL_enriched
BL_sig_up <- BL_sig_up[!is.na(BL_sig_up)]

AP_sig_up <- AV_sig$AP_enriched
AP_sig_up <- AP_sig_up[!is.na(AP_sig_up)]
```

```{r}
## 1. pan AV signature per PID group 
panAV_sig_up_rel <- panAV_sig_up[panAV_sig_up %in% rownames(assay(vsd))]

df <- t(assay(vsd)[panAV_sig_up_rel ,]) %>% as.data.frame

# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Pan-Alveolar signature (Gray et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```


```{r}
## 2. BL signature per PID group 
BL_sig_up_rel <- BL_sig_up[BL_sig_up %in% rownames(assay(vsd))]

# Filter for available genes
TNBC_PID <- colnames(vsd[,vsd$TNBC == "TNBC"])
vsd_TNBC <- vsd[, colnames(vsd) %in% TNBC_PID]
df <- t(assay(vsd_TNBC)[BL_sig_up_rel ,]) %>% as.data.frame

# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Basal-luminal signature (Li et al.)")+
  #geom_signif(
  #  comparisons = list(c("KRT5/7+", "KRT5+"), c("KRT5/7+", "KRT5/7-"),c("KRT5/7+", "KRT7+")),
  #  map_signif_level = FALSE, step_increase = 0.1, test = "t.test"
  #)+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

```{r}
## 3. AP signature per PID group 
AP_sig_up_rel <- AP_sig_up[AP_sig_up %in% rownames(assay(vsd))]
# Filter for available genes
df <- t(assay(vsd)[AP_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("AV progenitor signature (Li et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

B: pan-Basal and pan-HS

```{r}
HS_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Lietal_AV_BL_Signatures_reduced.xlsx",sheet = 2) %>% select("Pan-HS") %>% pull("Pan-HS")
HS_sig_up <- HS_sig[!is.na(HS_sig)]

BA_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Lietal_AV_BL_Signatures_reduced.xlsx",sheet = 3) %>% select("Pan-BA") %>% pull("Pan-BA")
BA_sig_up <- BA_sig[!is.na(BA_sig)]
```

```{r}
## 1. HS signature per PID group 
HS_sig_up_rel <- HS_sig_up[HS_sig_up %in% rownames(assay(vsd))]
# Filter for available genes
df <- t(assay(vsd)[HS_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Hormone-sensing signature (Li et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

```{r}
## 2. BA signature per PID group 
BA_sig_up_rel <- BA_sig_up[BA_sig_up %in% rownames(assay(vsd))]
# Filter for available genes
df <- t(assay(vsd)[BA_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("BA signature (Li et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

### Overlap between gene signatures 

```{r}
ggVennDiagram(gene_signatures,
              category.names = c("LuminalProgenitorLim","panAlveolarGray","BLGray"),
              label_alpha = 0, 
              set_color = "black")+
  scale_fill_gradient(low = "white", high = "red")+
  scale_color_manual(values = c("black","black","black"))
```


## 8. Chromosomal instability signature

```{r}
chrom_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/ChromosomalInstability_Carter_NatGen06.xlsx",sheet = 1, col_names = FALSE)

# Extract unique genes that were up- or down-regulated in each signature 
chrom_sig_up <- chrom_sig$...1 %>% unique()
```

```{r}
## 2. chrom signature per PID group 
chrom_sig_up_rel <- chrom_sig_up[chrom_sig_up %in% rownames(assay(vsd))]
# Filter for available genes
df <- t(assay(vsd)[chrom_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- vsd$KRT5_7_score[match(rownames(df), colnames(vsd))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #glochroml comparison

ggplot(df, aes(x = reorder(KRT5_7_score, signature_sum, FUN = median), y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Chromosomal instability signature (Carter et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

## 9. GSEA 

With `clusterProfiler`. 

```{r}
organism <- "org.Hs.eg.db"

# Prepare input gene list based on log2foldchanges 
original_gene_list <- res$log2FoldChange
names(original_gene_list) <- rownames(res)
gene_list <- na.omit(original_gene_list)
gene_list <- sort(gene_list, decreasing = TRUE) # sort in decreasing order

# Run GO analysis
gse <- gseGO(geneList=gene_list, 
             ont ="BP", 
             keyType = "SYMBOL", 
             minGSSize = 50, 
             maxGSSize = 1000, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = organism, 
             pAdjustMethod = "BH")

# Visualization
## Dot plot
dotplot(gse, showCategory=15, split=".sign") + facet_grid(~.sign)

## Emap plot
emapplot(pairwise_termsim(gse), showCategory = 15, color = "enrichmentScore", layout = "circle")

## GSEA plot
gseaplot2(gse, title = gse$Description[9], geneSetID = 1)
gseaplot2(gse, title = gse$Description[10], geneSetID = 1)
gseaplot2(gse, title = gse$Description[14], geneSetID = 1)
```

```{r}
# TNBC ONLY 
organism <- "org.Hs.eg.db"

# Prepare input gene list based on log2foldchanges 
original_gene_list <- res_TNBC$log2FoldChange
names(original_gene_list) <- rownames(res_TNBC)
gene_list_TNBC <- na.omit(original_gene_list)
gene_list_TNBC <- sort(gene_list_TNBC, decreasing = TRUE) # sort in decreasing order

# Run GO analysis
gse_TNBC <- gseGO(geneList=gene_list_TNBC, 
             ont ="BP", 
             keyType = "SYMBOL", 
             minGSSize = 50, 
             maxGSSize = 1000, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = organism, 
             pAdjustMethod = "BH")

# Visualization
## Dot plot
dotplot(gse_TNBC, showCategory=15, split=".sign") + facet_grid(~.sign)

## Emap plot
emapplot(pairwise_termsim(gse_TNBC), showCategory = 15, color = "enrichmentScore", layout = "circle")

## GSEA plot
gseaplot2(gse_TNBC, title = gse_TNBC$Description[9], geneSetID = 1)
gseaplot2(gse_TNBC, title = gse_TNBC$Description[10], geneSetID = 1)
gseaplot2(gse_TNBC, title = gse_TNBC$Description[14], geneSetID = 1)
```

## 10. KEGG pathway enrichment 

With `clusterProfiler`. 

```{r}
# KEGG requires Entrez IDs 
entrez.ids <- mapIds(org.Hs.eg.db, keys=rownames(res), column="ENTREZID", keytype="SYMBOL")
entrez.ids <- entrez.ids[!duplicated(entrez.ids)]

df <- data.frame(log_change = gene_list, 
           SYMBOL = names(gene_list))

df$ENTREZ <- entrez.ids[match(df$SYMBOL, names(entrez.ids))]
df <- df[!is.na(df$ENTREZ),]

gene_list_kegg <- df$log_change
names(gene_list_kegg) <- df$ENTREZ

# Run KEGG
gseKEGG <- gseKEGG(geneList=gene_list_kegg, 
        organism = "hsa",
        keyType = "kegg", 
        minGSSize = 50, 
        maxGSSize = 500, 
        pvalueCutoff = 0.05, 
        verbose = TRUE, 
        pAdjustMethod = "BH")

# Visualization 
## Dotplot 
dotplot(gseKEGG, showCategory = 10, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)

## EMAP plot 
emapplot(pairwise_termsim(gseKEGG), showCategory = 15,color = "enrichmentScore", layout = "circle")

## GSEA plot
gseaplot2(gseKEGG, title = gseKEGG$Description[4], geneSetID = 1)
gseaplot2(gseKEGG, title = gseKEGG$Description[12], geneSetID = 1)
gseaplot2(gseKEGG, title = gseKEGG$Description[20], geneSetID = 1)
```

```{r}
# TNBC only 
# KEGG requires Entrez IDs 
entrez.ids <- mapIds(org.Hs.eg.db, keys=rownames(res_TNBC), column="ENTREZID", keytype="SYMBOL")
entrez.ids <- entrez.ids[!duplicated(entrez.ids)]

df <- data.frame(log_change = gene_list_TNBC, 
           SYMBOL = names(gene_list_TNBC))

df$ENTREZ <- entrez.ids[match(df$SYMBOL, names(entrez.ids))]
df <- df[!is.na(df$ENTREZ),]

gene_list_kegg <- df$log_change
names(gene_list_kegg) <- df$ENTREZ

# Run KEGG
gseKEGG_TNBC <- gseKEGG(geneList=gene_list_kegg, 
        organism = "hsa",
        keyType = "kegg", 
        minGSSize = 50, 
        maxGSSize = 500, 
        pvalueCutoff = 0.05, 
        verbose = TRUE, 
        pAdjustMethod = "BH")

# Visualization 
## Dotplot 
dotplot(gseKEGG_TNBC, showCategory = 10, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)

## EMAP plot 
emapplot(pairwise_termsim(gseKEGG_TNBC), showCategory = 15,color = "enrichmentScore", layout = "circle")

## GSEA plot
gseaplot2(gseKEGG_TNBC, title = gseKEGG_TNBC$Description[4], geneSetID = 1)
gseaplot2(gseKEGG_TNBC, title = gseKEGG_TNBC$Description[12], geneSetID = 1)
gseaplot2(gseKEGG_TNBC, title = gseKEGG_TNBC$Description[20], geneSetID = 1)
```

## 11. Clustering for PID groups

```{r}
res <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_PAM50_res.rds") 
vsd <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_full_Combined_VST.rds")
brca.clean <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_cleaned_full.rds")
```

Find the same patient groups in TCGA BRCA data as in discovery dataset. 

### HLADR GMM 

```{r}
set.seed(22)

# HLADR
GMM_HLADR <- Mclust(DESeq2::counts(res, normalized = TRUE)["HLA-DRA",], G = 2)


GMM_df <- data.frame(class_epi = ifelse(GMM_HLADR$classification == 2 & GMM_HLADR$uncertainty < 0.01, 
                                        "HLADR+",
                                        "HLADR-"), 
                     HLADR_exprs = counts(res, normalized = TRUE)["HLA-DRA",],
                     uncertainty = GMM_HLADR$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = HLADR_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")

# Add HLA group 
vsd$HLADR_group <- GMM_df$class_epi[match(colnames(vsd), rownames(GMM_df))]

n_cat <- colData(vsd) %>% as.data.frame() %>% group_by(HLADR_group) %>% dplyr::count()
n_cat <- n_cat %>% ungroup() %>% mutate(sum = sum(n)) %>% mutate(fra = (n/sum)*100)
```

```{r}
# Make patient groups 
vsd$PID_group_test <- ifelse(vsd$HLADR_group == "HLADR+", "inflamed", vsd$KRT5_7_score)
vsd$PID_group_risk <- ifelse(vsd$PID_group_test == "inflamed", "low", ifelse(vsd$PID_group_test == "KRT5/7+", "high", "medium"))
```

```{r}
# Define colors 
group_color <- setNames(c("#3B9AB2", "gold", "#F21A00", "pink", "orange"), 
                        c("inflamed","KRT5/7-","KRT5/7+","KRT5+","KRT7+"))
group_color

metadata(vsd)$colors$PID_group_test <- group_color

risk_group_color <- setNames(c("#3B9AB2", "gold", "#F21A00"), 
                        c("low","medium","high"))
risk_group_color
metadata(vsd)$colors$PID_group_risk <- risk_group_color

# Define factors 
vsd$PID_group_test <- factor(vsd$PID_group_test, levels = c("inflamed","KRT5/7-","KRT5+","KRT7+","KRT5/7+"))
vsd$PID_group_risk <- factor(vsd$PID_group_risk, levels = c("low","medium","high"))
```

### Alluvial 

```{r}
df_alluvial <- colData(vsd) %>% as.data.frame %>% dplyr::select(KRT5_7_score, HLADR_group,PID_group_test,PID_group_risk)
alluvial_wide(df_alluvial, 
              colorful_fill_variable_stratum = FALSE,
              col_vector_flow = c("#85D4E3","grey", "#F4B5BD","#9C964A"),
              order_levels = c("KRT5/7+","KRT5/7-","KRT7+","KRT5+","high","medium","low"))+
  theme_classic()

df_alluvial_1 <- colData(vsd) %>% as.data.frame %>% dplyr::select(PID_group_test, PID_group_risk)

alluvial_wide(df_alluvial_1, 
              colorful_fill_variable_stratum = FALSE,
              col_vector_flow = c("#3B9AB2", "gold", "#E1AF00", "#EBCC2A", "#F21A00"),  
              order_levels = c("inflamed","N","L","B","BL", "low", "medium","high")
              )+  
  theme_classic()
```

### Marker heatmap 

```{r}
GOI <- c("KRT5","KRT7","HLA-DRA") #three marker IHC panel?! 

GOI_plus <- c("CD8A","CD274","PDCD1","GZMB","PDCD1LG2","IDO1", "ICOS", "TOX") # additional immune markers

assayNames(vsd) <- "vst"

GOI_mean_sce <- aggregateAcrossCells(vsd, ids = vsd$PID_group_test,
                                 statistics = "mean",
                                 use.assay.type = "vst",
                                 subset.row = rowData(vsd)$gene_name %in% GOI)

GOI_mean_sce_2 <- aggregateAcrossCells(vsd, ids = vsd$PID_group_test,
                                 statistics = "mean",
                                 use.assay.type = "vst",
                                 subset.row = rowData(vsd)$gene_name %in% GOI_plus)

Heatmap(t(assay(GOI_mean_sce, "vst")), col = viridis(100))+
Heatmap(t(assay(GOI_mean_sce_2, "vst")), col = viridis(100))

Heatmap(scale(t(assay(GOI_mean_sce, "vst"))))+
  Heatmap(scale(t(assay(GOI_mean_sce_2, "vst"))))

Heatmap(scale(t(assay(GOI_mean_sce_2, "vst"))))
```

### Survival analysis 

#### Kaplan-Meier 

KAPLAN MEIER CURVES 

Full TCGA dataset 

```{r}
#Assess survival impact 
cur_surv <- colData(vsd) %>% as.data.frame() %>%
  dplyr::select(PID, PID_group_test, PID_group_risk, PAM50Call_RNAseq, DFI, DFI.time, OS, OS.time, DSS, DSS.time, PFI, PFI.time) 


#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ PID_group_test, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"),
          log.rank.weights = "n",
          #conf.int = TRUE,
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ PID_group_risk, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          palette = c("#3B9AB2", "gold","#F21A00"),
          log.rank.weights = "n",
          #conf.int = TRUE,
          xlim = c(0, 4000))

#Plot survival curves
fit_1 <- survfit(Surv(PFI.time, PFI) ~ PID_group_test, data = cur_surv)
print(fit_1)

ggsurvplot(fit_1,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"),
          xlim = c(0, 4000))

#Plot survival curves
fit_1 <- survfit(Surv(PFI.time, PFI) ~ PID_group_risk, data = cur_surv)
print(fit_1)

ggsurvplot(fit_1,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = c(wes_palette("Zissou1")[1],"gold", wes_palette("Zissou1")[5]),
          xlim = c(0, 4000))

#Plot survival curves for PAM50 subtypes
cur_surv_PAM50 <- cur_surv %>% filter(PAM50Call_RNAseq != "Undefined")
fit_2 <- survfit(Surv(DFI.time, DFI) ~ PAM50Call_RNAseq, data = cur_surv_PAM50)
print(fit_1)

ggsurvplot(fit_2,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = "Paired",
          xlim = c(0, 4000))

```

Subset for TNBC 

```{r}
#Assess survival impact 
cur_surv <- colData(vsd) %>% as.data.frame() %>% filter(TNBC == "TNBC") %>% 
  #filter(PID_group_test %in% c("KRT5/7+","inflamed")) %>%
  dplyr::select(PID, PID_group_test, PID_group_risk, DFI, DFI.time, OS, OS.time, DSS, DSS.time, PFI, PFI.time) 


#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ PID_group_test, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"),
          #conf.int = TRUE,
          xlim = c(0, 4000))

fit <- survfit(Surv(DFI.time, DFI) ~ PID_group_risk, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = c(wes_palette("Zissou1")[5], wes_palette("Zissou1")[1],"gold"),
          #conf.int = TRUE,
          xlim = c(0, 4000))

#Plot survival curves
fit_1 <- survfit(Surv(PFI.time, PFI) ~ PID_group_test, data = cur_surv)
print(fit_1)

ggsurvplot(fit_1,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = c("#3B9AB2", "gold", "pink", "orange","#F21A00"),
          #conf.int = TRUE,
          xlim = c(0, 4000))

fit_1 <- survfit(Surv(PFI.time, PFI) ~ PID_group_risk, data = cur_surv)
print(fit_1)

ggsurvplot(fit_1,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = c(wes_palette("Zissou1")[1],"gold",wes_palette("Zissou1")[5]),
          #conf.int = TRUE,
          xlim = c(0, 4000))
```


#### Cox models

##### PID groups 

```{r}
vsd$age <- vsd$Age_at_Initial_Pathologic_Diagnosis_nature2012
```

Multivariate 

```{r}
## 1. Disease-free interval 
### Multivariate cox proportional hazard model on multiple covariates (here PID_group_test)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>%
  dplyr::select(PID_group_test, PID) %>% 
  group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_test",values_from = "fra") %>% column_to_rownames("PID")

colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$DFI <- vsd$DFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$DFI.time <- vsd$DFI.time[match(rownames(Cox_df),vsd$PID)]

Cox_df$age <- as.numeric(vsd$age[match(rownames(Cox_df),vsd$PID)])
Cox_df$pT_simple <- as.factor(vsd$Tumor_nature2012[match(rownames(Cox_df),vsd$PID)])
Cox_df$pN_simple <- as.factor(vsd$Node_nature2012[match(rownames(Cox_df),vsd$PID)])
Cox_df$pM_simple <- as.factor(vsd$Metastasis_nature2012[match(rownames(Cox_df),vsd$PID)])

#Based on proportions
covariates <- Cox_df %>% dplyr::select(-DFI,-DFI.time, -age, -pT_simple, -pN_simple, -pM_simple, - inflamed) %>% colnames

multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~ age + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "), "+ inflamed"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

write.csv(plot_df_prop, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/supp_tables/TCGA_TNBC_Cox_Multi_Groups.csv.csv")

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_log10()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

Univariate

```{r}
## 1. Disease-free interval 
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>% 
  dplyr::select(PID_group_test,PID) %>% 
group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_test",values_from = "fra") %>% column_to_rownames("PID")

colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$DFI <- vsd$DFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$DFI.time <- vsd$DFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-DFI,-DFI.time, - inflamed) %>% colnames

multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ inflamed"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,12))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
## 2. PFS
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>% 
  dplyr::select(PID_group_test,PID) %>% 
group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_test",values_from = "fra") %>% column_to_rownames("PID")

colnames(PID_metagroups) <- c("inflamed","KRT57negative","KRT5positive","KRT7positive","KRT57positive")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$PFI <- vsd$PFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$PFI.time <- vsd$PFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-PFI,-PFI.time, - inflamed) %>% colnames

multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ inflamed"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,12))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

##### PID risk groups 

```{r}
## 1. Disease-free interval 
### Multivariate cox proportional hazard model on multiple covariates (here PID_group_risk)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>% 
  dplyr::select(PID_group_risk, PID) %>% 
  group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$DFI <- vsd$DFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$DFI.time <- vsd$DFI.time[match(rownames(Cox_df),vsd$PID)]

Cox_df$age <- as.numeric(vsd$age[match(rownames(Cox_df),vsd$PID)])
Cox_df$pT_simple <- as.factor(vsd$Tumor_nature2012[match(rownames(Cox_df),vsd$PID)])
Cox_df$pN_simple <- as.factor(vsd$Node_nature2012[match(rownames(Cox_df),vsd$PID)])
Cox_df$pM_simple <- as.factor(vsd$Metastasis_nature2012[match(rownames(Cox_df),vsd$PID)])

#Based on proportions
covariates <- Cox_df %>% dplyr::select(-DFI,-DFI.time, -age, -pT_simple, -pN_simple, -pM_simple, - low) %>% colnames

multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~ age + pT_simple + pN_simple + pM_simple +", paste(covariates, collapse = " + "), "+ low"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],3),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],3),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],3),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],3),
                      wald_test = signif(multi_res$waldtest["pvalue"],3))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,20))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_log10()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

Univariate for TNBC risk groups

```{r}
## 1. Disease-free interval 
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>%
  dplyr::select(PID_group_risk,PID) %>% 
   
  group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$DFI <- vsd$DFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$DFI.time <- vsd$DFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions

covariates <- Cox_df %>% select(-DFI,-DFI.time, - low) %>% colnames
multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ low"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,12))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
## 2. Progression-free interval 
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% 
  filter(TNBC == "TNBC") %>%
  dplyr::select(PID_group_risk,PID) %>% 
   
  group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "PID_group_risk",values_from = "fra") %>% column_to_rownames("PID")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$PFI <- vsd$PFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$PFI.time <- vsd$PFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions

covariates <- Cox_df %>% select(-PFI,-PFI.time, -low) %>% colnames
multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ low"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,12))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
theme_classic()+
scale_y_continuous(trans = "log10")+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

### Barplot visualization 

```{r}
dittoBarPlot(vsd, var = "PID_group_test", group.by = "TNBC",retain.factor.levels = TRUE)+
  scale_fill_manual(values = group_color)+
  ggtitle("")

dittoBarPlot(vsd, var = "PID_group_test", group.by = "PAM50Call_RNAseq",retain.factor.levels = TRUE)+
  scale_fill_manual(values = group_color)+
  ggtitle("")
```

```{r}
# Chi-square test to compare between TNBC status between KRT5/7 scores
bar_out <- dittoBarPlot(vsd, var = "PID_group_test", group.by = "TNBC",retain.factor.levels = TRUE,data.out = TRUE)$data

bar_test <- bar_out %>% filter(!grouping %in% "Undefined") %>% pivot_wider(names_from = grouping, id_cols = label, values_from = count) %>% column_to_rownames("label")

bar_test_out <- chisq.test(bar_test)

print(bar_test_out$p.value)
```

### IFN-G signature 

From Ayers et al., 2017 (JCI) - Predicted response to PD1 blockade in melanoma

```{r}
IFNG_sig <- c("IFNG", "STAT1", "CCR5", "CXCL9", "CXCL10", "CXCL11", "IDO1", "PRF1", "GZMA") #excluded HLA-DRA from signature 

## IFN signature per PID group 
IFNG_sig_rel <- IFNG_sig[IFNG_sig %in% rownames(assay(vsd))]
# Filter for available genes
df <- t(assay(vsd)[IFNG_sig_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_group_test <- vsd$PID_group_test[match(rownames(df), colnames(vsd))]

compare_means(signature_sum ~ PID_group_test,df) #pairwise comparisons
compare_means(signature_sum ~ PID_group_test,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = PID_group_test, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = PID_group_test), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$PID_group_test)+
  theme_classic()+
  ylab("IFN-gamma signature score (Ayers et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_group_test, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_group_test")

Heatmap(ht, name = "Mean signature Z-scores")

INFg <- mat %>% group_by(PID_group_test) %>% summarize(INFg = mean(mean)) %>% column_to_rownames("PID_group_test")

Heatmap(INFg, cluster_rows = FALSE)
```

### CTA signature 

```{r}
CTA_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/CTAlist_PNAS_2019.xlsx")

CTA_sig_up <- CTA_sig$`Cancer Testis Antigens` %>% unique()
CTA_sig_up <- toupper(CTA_sig_up)
```

```{r}
## 1. pan AV signature per PID group 
CTA_sig_up_rel <- CTA_sig_up[CTA_sig_up %in% rownames(assay(vsd))]

df <- t(assay(vsd)[CTA_sig_up_rel ,]) %>% as.data.frame

# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_group_test <- vsd$PID_group_test[match(rownames(df), colnames(vsd))]

compare_means(signature_sum ~ PID_group_test,df) #pairwise comparisons
compare_means(signature_sum ~ PID_group_test,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = reorder(PID_group_test, -signature_sum), y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = PID_group_test), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  coord_cartesian(ylim = c(-50,350))+
  ylab("Pan-Alveolar signature (Gray et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_group_test, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_group_test")

Heatmap(ht, name = "Mean signature Z-scores")

Heatmap(mat %>% group_by(PID_group_test) %>% summarize(CTA = mean(mean)) %>% column_to_rownames("PID_group_test"))

CTA <- mat %>% group_by(PID_group_test) %>% summarize(CTA = mean(mean)) %>% column_to_rownames("PID_group_test")

Heatmap(CTA, cluster_rows = FALSE)
```

### IFNab signature 

```{r}
INFab_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/INFALPHABETA_Signaling_Reactome.xlsx")

INFab_sig_up <- INFab_sig$INF_SIG %>% unique()
INFab_sig_up <- toupper(INFab_sig_up)
```

```{r}
## 1. pan AV signature per PID group 
INFab_sig_up_rel <- INFab_sig_up[INFab_sig_up %in% rownames(assay(vsd))]

df <- t(assay(vsd)[INFab_sig_up_rel ,]) %>% as.data.frame

# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_group_test <- vsd$PID_group_test[match(rownames(df), colnames(vsd))]

compare_means(signature_sum ~ PID_group_test,df) #pairwise comparisons
compare_means(signature_sum ~ PID_group_test,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = reorder(PID_group_test, -signature_sum), y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = PID_group_test), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)+
  theme_classic()+
  ylab("Pan-Alveolar signature (Gray et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_group_test, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_group_test")

Heatmap(ht, name = "Mean signature Z-scores")

INFab <- mat %>% group_by(PID_group_test) %>% summarize(INFab = mean(mean)) %>% column_to_rownames("PID_group_test")

Heatmap(INFab, name = "INFab", cluster_rows = TRUE)
```


```{r}
Heatmap(CTA, name = "CTA", cluster_rows = TRUE)+
  Heatmap(INFab, name = "INFab", cluster_rows = FALSE)+
  Heatmap(INFg, name = "INFg", cluster_rows = FALSE)
```


### Immunosuppressive genes

```{r}
IFNG_sig <- c("PVR", "CD276", "TGFB2", "PROM1", "TGFB1") #"TGFB2"

## IFN signature per PID group 
IFNG_sig_rel <- IFNG_sig[IFNG_sig %in% rownames(assay(vsd))]
# Filter for available genes
df <- t(assay(vsd)[IFNG_sig_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_group_test <- vsd$PID_group_test[match(rownames(df), colnames(vsd))]

compare_means(signature_sum ~ PID_group_test,df) #pairwise comparisons
compare_means(signature_sum ~ PID_group_test,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = PID_group_test, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = PID_group_test), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(vsd)$colors$PID_group_test)+
  theme_classic()+
  ylab("IFN-gamma signature score (Ayers et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_group_test, gene) %>% summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_group_test")

Heatmap(ht, name = "Mean signature Z-scores")
```


## 12. Immune cell deconvolution 

```{r}
deconvolution_methods

brca.clean$PID_group_test <- vsd$PID_group_test[match(brca.clean$PID, vsd$PID)]

# TPM gene expression matrix 
TPM_mat <- assay(brca.clean, "tpm_unstrand")
rownames(TPM_mat) <- rowData(brca.clean)$gene_name
```

### QuantiSeq

```{r}
TCGA_QS <- immunedeconv::deconvolute(TPM_mat, method = "quantiseq")
#TCGA_QS <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_Quantiseq.rds")

# Z-scoring of QS results by cell type
TCGA_QS_1 <- TCGA_QS %>% filter(cell_type != "uncharacterized cell") %>% column_to_rownames("cell_type")
TCGA_QS_z <- as.data.frame(t(scale(t(TCGA_QS_1), center=TRUE, scale=TRUE)))
TCGA_QS_z <- TCGA_QS_z %>% rownames_to_column("cell_type")
TCGA_QS_plot <- TCGA_QS_z %>% pivot_longer(2:ncol(TCGA_QS_z), names_to = "PID", values_to = "score")

TCGA_QS_plot$PID_group_test <- brca.clean$PID_group_test[match(TCGA_QS_plot$PID, colnames(brca.clean))]
  
# Heatmap visualization  
ht_QS <- TCGA_QS_plot %>% group_by(PID_group_test, cell_type) %>% dplyr::summarize(mean = mean(score)) %>% pivot_wider(names_from = PID_group_test, values_from = mean) %>% column_to_rownames("cell_type")

Heatmap(ht_QS, name = "mean QS-Z-score")
```

### MCP-counter 

```{r}
res_MCP <- immunedeconv::deconvolute(TPM_mat, method = "mcp_counter")
#res_MCP <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_MCP.rds")

#Z-scoring of MCP results by cell type
res_MCP_1 <- res_MCP %>% dplyr::select(-cell_type)
rownames(res_MCP_1) <- res_MCP$cell_type
res_MCP_z <- as.data.frame(t(scale(t(res_MCP_1), center=TRUE, scale=TRUE)))
res_MCP_z <- res_MCP_z %>% rownames_to_column("cell_type")

res_MCP_plot <- res_MCP_z %>% pivot_longer(2:ncol(res_MCP_z), names_to = "PID", values_to = "score")

res_MCP_plot$PID_group_test <- brca.clean$PID_group_test[match(res_MCP_plot$PID, colnames(brca.clean))]

res_MCP_plot$TNBC <- brca.clean$TNBC[match(res_MCP_plot$PID, colnames(brca.clean))]

ggplot(res_MCP_plot, aes(x = PID_group_test, y = score))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID_group_test), size = 3, position=position_jitter(w=0.1,h=0.1))+
  facet_wrap(~cell_type, scales = "free_y")+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90))+
  scale_color_brewer(palette = "Set1")+
  guides(color = "none")+
  scale_color_manual(values = metadata(vsd)$colors$KRT57_score)
  #stat_compare_means(method = "t.test")
  

# Heatmap visualization  
ht_MCP <- res_MCP_plot %>% group_by(PID_group_test, cell_type) %>% dplyr::summarize(mean = mean(score)) %>% pivot_wider(names_from = PID_group_test, values_from = mean) %>% column_to_rownames("cell_type")

Heatmap(ht_MCP, name = "mean MCP-Z-score")
```

```{r}
saveRDS(res_MCP, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_MCP.rds")
```


## 13. Lehmann subtype comparison 

```{r}
# Read in data from Lehmann et al. (2016, Plos ONE)
lehmann_TCGA <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/lehmann_refined_TCGA_2016.xlsx", sheet = 1)

lehmann_TCGA$PID <- str_split(lehmann_TCGA$BARCODE, "-", simplify = TRUE)[,3]

# Add data to vsd 
# TNBCtype (2011 study), TNBCtype_4 (2016 study)
vsd$TNBCtype <- lehmann_TCGA$TNBCtype[match(vsd$PID, lehmann_TCGA$PID)]
vsd$TNBCtype_4 <- lehmann_TCGA$TNBCtype_4[match(vsd$PID, lehmann_TCGA$PID)]
vsd$TNBC_assign_lehmann <- lehmann_TCGA$TNBC[match(vsd$PID, lehmann_TCGA$PID)]

# Turn subtypes into NA (if patient was not assigned TNBC in my dataset)
vsd$TNBCtype <- ifelse(vsd$TNBCtype %in% c("BL1", "BL2", "M", "MSL", "LAR", "IM") & vsd$TNBC == "TNBC", vsd$TNBCtype, NA)

vsd$TNBCtype_4 <- ifelse(vsd$TNBCtype_4 %in% c("BL1", "BL2", "M", "LAR") & vsd$TNBC == "TNBC", vsd$TNBCtype_4, NA)
```

### Survival analysis 

```{r}
## DFS 

#Assess survival impact 

cur_surv <- colData(vsd) %>% as.data.frame() %>% filter(TNBC == "TNBC") %>% 
  #filter(PID_group_test %in% c("KRT5/7+","inflamed")) %>%
  dplyr::select(PID, PID_group_test, PID_group_risk, TNBCtype, TNBCtype_4, DFI, DFI.time, OS, OS.time, DSS, DSS.time, PFI, PFI.time) 

## Disease free survival 

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ PID_group_test, data = cur_surv)
print(fit)


ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = c(wes_palette("Zissou1")[1],"gold",wes_palette("Zissou1")[3],wes_palette("Zissou1")[4], wes_palette("Zissou1")[5]),
          #conf.int = TRUE,
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ TNBCtype_4, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = "Set1",
          #conf.int = TRUE,
          xlim = c(0, 4000))

#Plot survival curves
fit <- survfit(Surv(DFI.time, DFI) ~ TNBCtype, data = cur_surv)
print(fit)

ggsurvplot(fit,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = "Set1",
          #conf.int = TRUE,
          xlim = c(0, 4000))
```


```{r}
## PFS
cur_surv <- colData(vsd) %>% as.data.frame() %>% filter(TNBC == "TNBC") %>% 
  #filter(PID_group_test %in% c("KRT5/7+","inflamed")) %>%
  dplyr::select(PID, PID_group_test, PID_group_risk, TNBCtype, TNBCtype_4, DFI, DFI.time, OS, OS.time, DSS, DSS.time, PFI, PFI.time) 

#Plot survival curves
fit_1 <- survfit(Surv(PFI.time, PFI) ~ PID_group_test, data = cur_surv)
print(fit_1)

ggsurvplot(fit_1,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
         palette = c(wes_palette("Zissou1")[1],"gold",wes_palette("Zissou1")[3],wes_palette("Zissou1")[4], wes_palette("Zissou1")[5]),
          xlim = c(0, 4000))

#Plot survival curves
fit_1 <- survfit(Surv(PFI.time, PFI) ~ PID_group_risk, data = cur_surv)
print(fit_1)

ggsurvplot(fit_1,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
         palette = c("#3B9AB2", "gold","#F21A00"),
          xlim = c(0, 4000))

#Plot survival curves
fit_1 <- survfit(Surv(PFI.time, PFI) ~ TNBCtype_4, data = cur_surv)
print(fit_1)

ggsurvplot(fit_1,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = "Set1",
          #conf.int = TRUE,
          xlim = c(0, 4000))

#Plot survival curves
fit_1 <- survfit(Surv(PFI.time, PFI) ~ TNBCtype, data = cur_surv)
print(fit_1)

ggsurvplot(fit_1,
          pval = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_classic(), # Change ggplot2 theme
          log.rank.weights = "n",
          palette = "Set1",
          #conf.int = TRUE,
          xlim = c(0, 4000))
```

```{r}
# TNBC type
dittoBarPlot(vsd, var = "KRT5_7_score", group.by = "TNBCtype")+
    scale_fill_manual(values = metadata(vsd)$colors$KRT57_score)
dittoBarPlot(vsd, var = "KRT5_7_score", group.by = "TNBCtype", scale = "count")+
  scale_fill_manual(values = metadata(vsd)$colors$KRT57_score)

dittoBarPlot(vsd, var = "TNBCtype", group.by = "KRT5_7_score")+
    scale_fill_brewer(palette = "Set1")


dittoBarPlot(vsd, var = "PID_group_test", group.by = "TNBCtype")+
    scale_fill_manual(values = metadata(vsd)$colors$PID_group_test)
dittoBarPlot(vsd, var = "PID_group_test", group.by = "TNBCtype", scale = "count")+
  scale_fill_manual(values = metadata(vsd)$colors$PID_group_test)

# TNBC type 4
dittoBarPlot(vsd, var = "KRT5_7_score", group.by = "TNBCtype_4")+
    scale_fill_manual(values = metadata(vsd)$colors$KRT57_score)
dittoBarPlot(vsd, var = "KRT5_7_score", group.by = "TNBCtype_4", scale = "count")+
  scale_fill_manual(values = metadata(vsd)$colors$KRT57_score)

dittoBarPlot(vsd, var = "PID_group_test", group.by = "TNBCtype_4")+
    scale_fill_manual(values = metadata(vsd)$colors$PID_group_test)
dittoBarPlot(vsd, var = "PID_group_test", group.by = "TNBCtype_4", scale = "count")+
  scale_fill_manual(values = metadata(vsd)$colors$PID_group_test)
```

```{r}
bar_out <- colData(vsd) %>% as.data.frame() %>% group_by(KRT5_7_score, TNBCtype) %>% dplyr::count() %>% ungroup %>% filter(!is.na(TNBCtype)) %>% group_by(KRT5_7_score) %>% mutate(fra = n/sum(n)) %>% arrange(KRT5_7_score)

ggplot(bar_out, aes(x = KRT5_7_score, y = fra, fill = TNBCtype))+
  geom_bar(stat = "identity")

bar_out <- colData(vsd) %>% as.data.frame() %>% group_by(PID_group_test, TNBCtype_4) %>% dplyr::count() %>% ungroup %>% filter(!is.na(TNBCtype_4)) %>% group_by(PID_group_test) %>% mutate(fra = n/sum(n)) %>% arrange(PID_group_test)

ggplot(bar_out, aes(x = PID_group_test, y = fra, fill = TNBCtype_4))+
  geom_bar(stat = "identity")
```

#### Cox models 

Univariate for both TNBCtype and TNBCtype4 

```{r}
## 1. Disease-free interval 
### Univariate cox proportional hazard model on multiple covariates (here KRT5/7 groups)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% dplyr::select(TNBCtype,PID) %>% filter(!is.na(TNBCtype)) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype",values_from = "fra") %>% column_to_rownames("PID")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$DFI <- vsd$DFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$DFI.time <- vsd$DFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-DFI,-DFI.time, - BL1) %>% colnames

multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ BL1"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,5))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
## 1. Disease-free interval 
### Multivariate cox proportional hazard model on multiple covariates (here KRT5/7 groups)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% dplyr::select(TNBCtype_4,PID) %>% filter(!is.na(TNBCtype_4)) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype_4",values_from = "fra") %>% column_to_rownames("PID")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$DFI <- vsd$DFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$DFI.time <- vsd$DFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-DFI,-DFI.time, - BL1) %>% colnames

multi_formula <- as.formula(paste("Surv(DFI.time, DFI) ~", paste(covariates, collapse = " + "), "+ BL1"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,5))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```


```{r}
## TNBCtype 

## 2. PFS 
### Univariate cox proportional hazard model on multiple covariates (here KRT5/7 groups)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% dplyr::select(TNBCtype,PID) %>% filter(!is.na(TNBCtype)) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype",values_from = "fra") %>% column_to_rownames("PID")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$PFI <- vsd$PFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$PFI.time <- vsd$PFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-PFI,-PFI.time, - BL1) %>% colnames

multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ BL1"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))


#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,5))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

```{r}
# TNBCtype4 
## 2. PFS
### Multivariate cox proportional hazard model on multiple covariates (here KRT5/7 groups)
PID_metagroups <- colData(vsd) %>% as.data.frame() %>% dplyr::select(TNBCtype_4,PID) %>% filter(!is.na(TNBCtype_4)) %>% group_by(PID) %>% table() %>% as.data.frame

PID_metagroups <- PID_metagroups %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))

PID_metagroups <- PID_metagroups %>% pivot_wider(id_cols = "PID",names_from = "TNBCtype_4",values_from = "fra") %>% column_to_rownames("PID")

Cox_df <- PID_metagroups

#add covariates for correction
Cox_df$PFI <- vsd$PFI[match(rownames(Cox_df),vsd$PID)]
Cox_df$PFI.time <- vsd$PFI.time[match(rownames(Cox_df),vsd$PID)]

#Based on proportions
covariates <- Cox_df %>% select(-PFI,-PFI.time, - BL1) %>% colnames

multi_formula <- as.formula(paste("Surv(PFI.time, PFI) ~", paste(covariates, collapse = " + "), "+ BL1"))

multi_model <- coxph(multi_formula, data = Cox_df)

multi_res <- summary(multi_model)

plot_df_prop <- data.frame(group = rownames(multi_res$coefficients),
                      p_value = signif(multi_res$coefficients[,"Pr(>|z|)"],2),
                      HR = signif(multi_res$coefficients[,"exp(coef)"],2),
                      CI_low = signif(multi_res$conf.int[,"lower .95"],2),
                      CI_high = signif(multi_res$conf.int[,"upper .95"],2),
                      wald_test = signif(multi_res$waldtest["pvalue"],2))

#Plot hazard ratios
hr_prop <- ggplot(plot_df_prop, aes(x = reorder(group,dplyr::desc(HR)), y = HR, color = p_value < 0.05))+
geom_point(size = 4)+
geom_errorbar(aes(ymax = CI_high, ymin = CI_low))+
#coord_cartesian(ylim=c(0,5))+
geom_hline(yintercept = 1,color = "black",linetype = "dashed")+
scale_color_brewer(palette = "Set1")+
xlab("Covariates")+
ylab("HR for PID groups")+
scale_y_continuous(trans = "log10")+
theme_classic()+
theme(axis.text.x = element_text(angle = 90))

hr_prop
```

# Save objects

```{r}
saveRDS(vsd, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_DESeq_VST_full_Combined_VST.rds")
saveRDS(brca.clean, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/TCGA_BRCA_cleaned_full.rds")
```

# Session Info 

```{r}
sessionInfo()
```
