---
title: "GEO_Microarray_analysis"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: ISPY2_ES.rds, ISPY2_BM.rds, GeneSignatures.rds	

# Validation

Data analysis scripts for data analysis in the validation cohorts. 

# Script : GSE194040 - Microarray data analysis for ISPY2 Neoadjuvant Trial

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scran)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(GEOquery)
library(gridExtra)
library(readxl)
library(survival)
library(survminer)
library(epitools)
library(tidyverse)
library(easyalluvial)
library(ggalluvial)
library(wesanderson)
library(ggnewscale)
```

# 1. Download data

```{r}
assay <- read.delim("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/GSE194040_ISPY2ResID_AgilentGeneExp_990_FrshFrzn_meanCol_geneLevel_n988.txt")

colnames(assay) <- str_split(colnames(assay), "X", simplify = TRUE)[,2]

assay <- as.matrix(assay[,!str_detect(colnames(assay), "GPL")])
```

```{r}
features <- data.frame(gene = rownames(assay))
rownames(features) <- features$gene
```

```{r}
phenos <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/metadata_IPSY2.xlsx")

phenos$PID <- phenos$`Patient Identifier`

phenos <- as.data.frame(phenos[match(colnames(assay), phenos$PID),])

rownames(phenos) <- phenos$PID
```

```{r}
# Create a new ExpressionSet object
new_expset <- ExpressionSet(
  assayData = assay,
  phenoData = AnnotatedDataFrame(phenos),
  featureData = AnnotatedDataFrame(features)
)

# Verify the new ExpressionSet
print(new_expset)
```

## Add metadata 

```{r}
new_expset$TNBC <- ifelse(new_expset$`Receptor Subtype` == "TN", "TNBC", "Non_TNBC")
new_expset$PAM50 <- new_expset$BP_subtype
new_expset$response <- as.character(new_expset$pCR)
new_expset$arm_short <- new_expset$`Arm (short name)`
```

```{r}
pData(new_expset) %>% as.data.frame %>% group_by(TNBC) %>% count()
```

## Save objects 

```{r}
#saveRDS(new_expset, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ISPY2_RNA_analysis_expressionset.rds")

#new_expset <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ISPY2_RNA_analysis_expressionset.rds")

#new_expset <- new_expset[,new_expset$TNBC == "TNBC"]
```

# 2. CK57 signature 

```{r}
set.seed(22)

# 1. CK5
GMM_CK5 <- Mclust(exprs(new_expset)["KRT5",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(exprs(new_expset)["KRT5",] > 10.2, "KRT5+", "KRT5-"),
                     KRT5_exprs = exprs(new_expset)["KRT5",],
                     uncertainty = GMM_CK5$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = KRT5_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")

#Add GMM epi class to object
new_expset$GMM_KRT5 <- GMM_df$class_epi

# 2. CK7
GMM_CK7 <- Mclust(exprs(new_expset)["KRT7",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(exprs(new_expset)["KRT7",] > 13.9, "KRT7+", "KRT7-"),
                     KRT7_exprs = exprs(new_expset)["KRT7",],
                     uncertainty = GMM_CK7$uncertainty)


#Visualize GMM
ggplot(GMM_df, aes(x = KRT7_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 17)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")

#Add GMM epi class to object
new_expset$GMM_KRT7 <- GMM_df$class_epi

## Combine KRT5/7 GMM scores 
new_expset$KRT5_7_score <- ifelse(new_expset$GMM_KRT5 == "KRT5+" & new_expset$GMM_KRT7 == "KRT7+", "KRT5/7+", 
       ifelse(new_expset$GMM_KRT5 == "KRT5+" & new_expset$GMM_KRT7 == "KRT7-", "KRT5+",
              ifelse(new_expset$GMM_KRT5 == "KRT5-" & new_expset$GMM_KRT7 == "KRT7+", "KRT7+","KRT5/7-")))


## Add colors 
col_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("KRT5/7-", "KRT5/7+","KRT5+", "KRT7+"))
```

## Visualization 

```{r}
### 1. Normalized counts
plot <- as.data.frame(t(exprs(new_expset)[rownames(exprs(new_expset)) %in% c("KRT5", "KRT7", "KRT18","KRT14","HLA-DRA"),]))
plot$KRT5_7_score <- new_expset$KRT5_7_score[match(rownames(plot),colnames(new_expset))]
plot$GMM_KRT5 <- new_expset$GMM_KRT5[match(rownames(plot),colnames(new_expset))]
plot$GMM_KRT7 <- new_expset$GMM_KRT7[match(rownames(plot),colnames(new_expset))]

plot$PAM50 <- new_expset$PAM50.Subtype[match(rownames(plot),colnames(new_expset))]
plot$response <- new_expset$response[match(rownames(plot),colnames(new_expset))]
plot$TNBC <- new_expset$TNBC[match(rownames(plot),colnames(new_expset))]


ggplot(plot, aes(x=KRT7, y=KRT5, color = KRT5_7_score)) +
  geom_point(size = 2, alpha = 0.5)+
  theme_classic()+
  xlab("CK7 normalized counts")+
  ylab("CK5 normalized counts")

ggplot(plot, aes(x=KRT7, y=KRT5, color = TNBC)) +
  geom_point(size = 2, alpha = 0.5)+
  theme_classic()+
  xlab("CK7 normalized counts")+
  ylab("CK5 normalized counts")

ggplot(plot, aes(x=KRT7, y=KRT5, color = PAM50)) +
  geom_point(size = 2, alpha = 0.5)+
  theme_classic()+
  xlab("CK7 normalized counts")+
  ylab("CK5 normalized counts")

ggplot(plot, aes(x=KRT7, y=KRT5, color = response)) +
  geom_point(size = 2, alpha = 0.5)+
  theme_classic()+
  xlab("CK7 normalized counts")+
  ylab("CK5 normalized counts")

main <- ggplot(plot, aes(x=KRT7, y=KRT5, color = KRT5_7_score)) +
  geom_point(size = 2, alpha = 0.5)+
  theme_classic()+
  xlab("CK7 log counts")+
  ylab("CK5 log counts")+
  scale_color_manual(values = col_score)+
  guides(color = "none")

y_side <- ggplot(plot, aes(x = KRT5, fill=GMM_KRT5)) +
  geom_density(alpha=0.8) +
  scale_fill_manual(values = c('#999999','#F4B5BD')) +
  theme_classic()+
  theme(legend.position = "none", axis.title.y = element_blank())+
  coord_flip()

x_side <- ggplot(plot, aes(x = KRT7, fill=GMM_KRT7)) +
  geom_density(alpha=0.8) +
  scale_fill_manual(values = c('#999999','#9C964A')) +
  theme_classic()+
  theme(legend.position = "none", axis.title.x = element_blank())

blankPlot <- ggplot()+geom_blank(aes(1,1)) +
  cowplot::theme_nothing()

grid.arrange(x_side, blankPlot, main, y_side,
        ncol=2, nrow=2, widths=c(4, 1.4), heights=c(1.4, 4))

plot %>% as.data.frame() %>% rownames_to_column("PID") %>% group_by(KRT5_7_score) %>% dplyr::count() %>% ungroup %>% mutate(fra = n/sum(n))
```

#### Barplots 

```{r}
## PAM50 subtype 
bar_out <- plot %>% as.data.frame() %>% group_by(KRT5_7_score, PAM50) %>% dplyr::count() %>% ungroup %>% group_by(PAM50) %>% mutate(fra = n/sum(n)) %>% arrange(PAM50)

ggplot(bar_out, aes(x = PAM50, y = fra, fill = KRT5_7_score))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = col_score)+
  theme_classic()

plot %>% as.data.frame() %>% rownames_to_column("PID") %>% group_by(PAM50) %>% dplyr::count() %>% ungroup %>% mutate(fra = n/sum(n))

## TNBC 
bar_out <- plot %>% as.data.frame() %>% group_by(KRT5_7_score, TNBC) %>% dplyr::count() %>% ungroup %>% group_by(TNBC) %>% mutate(fra = n/sum(n)) %>% arrange(TNBC)

ggplot(bar_out, aes(x = TNBC, y = fra, fill = KRT5_7_score))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = col_score)+
  theme_classic()

bar_test <- bar_out %>% filter(TNBC != "NA") %>% pivot_wider(names_from = TNBC, id_cols = KRT5_7_score, values_from = n) %>% column_to_rownames("KRT5_7_score")
bar_test_out <- chisq.test(bar_test)
print(bar_test_out$p.value)

plot %>% as.data.frame() %>% rownames_to_column("PID") %>% group_by(TNBC) %>% dplyr::count() %>% ungroup %>% mutate(fra = n/sum(n))

## Chemotherapy response 
bar_out <- plot %>% as.data.frame() %>% group_by(KRT5_7_score, response) %>% dplyr::count() %>% filter(response != "NA") %>% ungroup %>% group_by(response) %>% mutate(fra = n/sum(n))

ggplot(bar_out, aes(x = response, y = fra, fill = KRT5_7_score))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = col_score)+
  theme_classic()

bar_out <- plot %>% as.data.frame() %>% group_by(KRT5_7_score, response) %>% dplyr::count() %>% filter(response != "NA") %>% ungroup %>% group_by(KRT5_7_score) %>% mutate(fra = n/sum(n))

ggplot(bar_out, aes(x = KRT5_7_score, y = fra, fill = response))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  theme_classic()
```

```{r}
## Chemotherapy response - TNBC
plot$arm_short <- new_expset$arm_short[match(rownames(plot),colnames(new_expset))]

bar_out <- plot %>% as.data.frame() %>% group_by(TNBC, response, arm_short) %>% filter(!arm_short %in% c("Pertuzumab", "TDM1/P")) %>% dplyr::count() %>% filter(response != "NA") %>% ungroup %>% group_by(TNBC, arm_short) %>% mutate(fra = n/sum(n))

p_out <- foreach(i = unique(bar_out$arm_short), .combine = "rbind")%do%{
  bar_test <- bar_out %>% filter(arm_short == i) %>% pivot_wider(names_from = TNBC, id_cols = response, values_from = n) %>% column_to_rownames("response")
  bar_test[is.na(bar_test)] <- 0
  bar_test <- bar_test[rowSums(bar_test) > 0,]
  return(data.frame(p_chi = fisher.test(bar_test)$p.value, cat = i))
}

p_out$FDR_chi <- p.adjust(p_out$p_chi)
p_out$p_chi
bar_out$FDR <- p_out$FDR_chi[match(bar_out$arm_short, p_out$cat)]
  
ggplot(bar_out, aes(x = TNBC, y = fra, fill = response))+
  facet_wrap(~arm_short)+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  geom_text(aes(label = FDR))+
  theme_classic()

ggplot(bar_out, aes(x = TNBC, y = n, fill = response))+
  facet_wrap(~arm_short)+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  geom_text(aes(label = FDR))+
  theme_classic()
```

### Keratin markers 

```{r}
# Define keratins of interest
keratins <- c("KRT5","KRT7","KRT15","KRT8","KRT18","KRT19", "KRT14","KRT17","ITGA6","EPCAM")

keratins_rel <- keratins[keratins %in% rownames(exprs(new_expset))]


# Define heatmap based on normalized counts 
df <- t(exprs(new_expset)[keratins_rel,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")

## Create heatmap body 
df$KRT5_7_score <- new_expset$KRT5_7_score[match(df$PID, colnames(new_expset))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% dplyr::count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(ht, name = "Mean Z-scores")

Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

### Stem markers 

```{r}
# Define stem markers of interest
stem <- c("SOX9","KIT","PROM1","HDAC2","NOTCH1","CD44","ALDH1A3","BRCA1","MUC1","EZH2", "YAP1","ANXA8", "TGFB2", "CD24", "TNFAIP3", "ERBB3", "TNFRSF11A")
stem_rel <- stem[stem %in% rownames(exprs(new_expset))]

# Define heatmap based on normalized counts 
df <- t(exprs(new_expset)[stem_rel,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")

## Create heatmap body 
df$KRT5_7_score <- new_expset$KRT5_7_score[match(df$PID, colnames(new_expset))]
mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% dplyr::summarize(mean = mean(exprs, na.rm = TRUE))

ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

## Create row annotation 
anno <- df %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% dplyr::count()
anno <- anno[match(anno$KRT5_7_score, rownames(ht)),]

## Print heatmap 
Heatmap(ht, name = "Mean Z-scores")

Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

## Gene signatures 

```{r}
lum_progenitor_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/LuminalProgenitor_Signature_Limetal_NatureMedicine_2009.xls")
mature_lum_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/MatureLuminalSignature_Limetal_NatureMedicine_2009.xls")
basal_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Basal_Signature_Limetal_NatureMedicine_2009.xls")

# Extract unique genes that were up- or down-regulated in each signature 
lum_progenitor_sig_up <- lum_progenitor_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
lum_progenitor_sig_up <- lum_progenitor_sig_up[!is.na(lum_progenitor_sig_up)]
lum_progenitor_sig_down <- lum_progenitor_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
lum_progenitor_sig_down <- lum_progenitor_sig_down[!is.na(lum_progenitor_sig_down)]

mature_lum_sig_up <- mature_lum_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
mature_lum_sig_up <- mature_lum_sig_up[!is.na(mature_lum_sig_up)]
mature_lum_sig_down <- mature_lum_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
mature_lum_sig_down <- mature_lum_sig_down[!is.na(mature_lum_sig_down)]

basal_sig_up <- basal_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
basal_sig_up <- basal_sig_up[!is.na(basal_sig_up)]
basal_sig_down <- basal_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
basal_sig_down <- basal_sig_down[!is.na(basal_sig_down)]
```

Upregulated genes per signature 

```{r}
## 1. Luminal progenitor signature per PID group 
lum_progenitor_sig_up_rel <- lum_progenitor_sig_up[lum_progenitor_sig_up %in% rownames(exprs(new_expset))]
# Filter for available genes
df <- t(exprs(new_expset)[lum_progenitor_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- new_expset$KRT5_7_score[match(rownames(df), colnames(new_expset))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = col_score)+
  theme_classic()+
  ylab("Luminal progenitor signature")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

```{r}
AV_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Lietal_AV_BL_Signatures_reduced.xlsx",sheet = 1)

colnames(AV_sig) <- c("pan_AV", "BL_enriched","AP_enriched")

# Extract unique genes that were up- or down-regulated in each signature 
panAV_sig_up <- AV_sig$pan_AV
panAV_sig_up <- panAV_sig_up[!is.na(panAV_sig_up)]

BL_sig_up <- AV_sig$BL_enriched
BL_sig_up <- BL_sig_up[!is.na(BL_sig_up)]

AP_sig_up <- AV_sig$AP_enriched
AP_sig_up <- AP_sig_up[!is.na(AP_sig_up)]
```

```{r}
## 1. pan AV signature per PID group 
panAV_sig_up_rel <- panAV_sig_up[panAV_sig_up %in% rownames(exprs(new_expset))]
# Filter for available genes
df <- t(exprs(new_expset)[panAV_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- new_expset$KRT5_7_score[match(rownames(df), colnames(new_expset))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = col_score)+
  theme_classic()+
  ylab("Pan-Alveolar signature (Li et al.)")+
  stat_compare_means()
```


```{r}
## 2. BL signature per PID group 
BL_sig_up_rel <- BL_sig_up[BL_sig_up %in% rownames(exprs(new_expset))]
# Filter for available genes
df <- t(exprs(new_expset)[BL_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- new_expset$KRT5_7_score[match(rownames(df), colnames(new_expset))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = col_score)+
  theme_classic()+
  ylab("Basal-luminal signature (Li et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

```{r}
# TNBC subset 

## 2. BL signature per PID group 
BL_sig_up_rel <- BL_sig_up[BL_sig_up %in% rownames(exprs(new_expset))]
# Filter for available genes
df <- t(exprs(new_expset[,new_expset$TNBC == "TNBC"])[BL_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$KRT5_7_score <- new_expset$KRT5_7_score[match(rownames(df), colnames(new_expset))]
df$KRT5_7_score <- factor(df$KRT5_7_score, levels = c("KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

compare_means(signature_sum ~ KRT5_7_score,df) #pairwise comparisons
compare_means(signature_sum ~ KRT5_7_score,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = KRT5_7_score, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = KRT5_7_score), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = col_score)+
  theme_classic()+
  ylab("Basal-luminal signature (Li et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(KRT5_7_score, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("KRT5_7_score")

Heatmap(ht, name = "Mean signature Z-scores")
```

# 3. Patient groups 

## HLADR signature 

```{r}
# 1. CK5
GMM_HLA <- Mclust(exprs(new_expset)["HLA-DRA",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(exprs(new_expset)["HLA-DRA",] > 13.9, "HLA-DRA+", "HLA-DRA-"),
                     HLADRA_exprs = exprs(new_expset)["HLA-DRA",],
                     uncertainty = GMM_HLA$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = HLADRA_exprs, fill = class_epi))+
  geom_density()+
  geom_rug(aes(color = class_epi))+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  scale_color_brewer(palette = "Set1")

#Add GMM epi class to object
new_expset$GMM_HLADRA <- GMM_df$class_epi
```

```{r}
new_expset$PID_groups <- ifelse(new_expset$GMM_HLADRA == "HLA-DRA+", "inflamed", new_expset$KRT5_7_score)

new_expset$PID_group_risk <- ifelse(new_expset$PID_groups == "inflamed", "low", ifelse(new_expset$PID_groups %in% c("KRT5/7+"), "high", "medium"))
new_expset$PID_group_risk <- factor(new_expset$PID_group_risk, levels = c("high", "medium", "low"))

# Define color 
group_color <- setNames(c("#3B9AB2", "gold", "pink", "orange","#F21A00"), 
                        c("inflamed","KRT5/7-","KRT5+","KRT7+","KRT5/7+"))
new_expset$PID_groups <- factor(new_expset$PID_groups, levels = c("inflamed","KRT5/7-","KRT5+","KRT7+","KRT5/7+"))
```

# 4. Biomarker association

```{r}
biomarker_df <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/biomarker_ISPY2_def.xlsx")

biomarker_df <- biomarker_df %>% filter(!is.na(biomarker)) %>% select(biomarker, pathway, type) %>% as.data.frame

rownames(biomarker_df) <- biomarker_df$biomarker
```

```{r}
new_expset$MP_index <- new_expset$`MP_index_adj*(-1`
pData(new_expset)["MP_index_adj*(-1)"] <- NULL
```

```{r}
# Heatmap body 
mat <- t(pData(new_expset) %>% select(25:50, MP_index) %>% mutate_all(as.numeric))

test <- as.data.frame(t(mat))

anno <- pData(new_expset) 
test$PID_groups <- anno$PID_groups[match(rownames(test), rownames(anno))]

test_result <- as.data.frame(compare_means(c(ICS5_score, Chemokine12_score, Module5_TcellBcell_score, 
STAT1_sig, Module3_IFN_score, Dendritic_cells, B_cells, 
Mast_cells, Module11_Prolif_score, MP_index, Basal_Index, 
PARPi7_score, PARPi7_plus_MP2, VCpred_TN, STMN1_dat, 
HER2_Index, Mod7_ERBB2, ERBB2.Y1248, EGFR.Y1173, mTOR.S2448, 
IGF1R_dat, TIE2.Y992, Mod10_ECM, RPL24_dat, LYMPHS_PCA_16704732, 
Luminal_Index, ER_PGR_avg)~PID_groups, method = "kruskal.test",data = test))

test_result$p.adj <- p.adjust(test_result$p, method = "BH") 

test_result$psignif.adj <- as.matrix(symnum(test_result$p.adj,
       cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), 
       symbols = c("****", "***", "**", "*", "ns")))[,1]

rownames(test_result) <- test_result$.y.

mat_mean <- test %>% pivot_longer(cols = 1:27, names_to = "cat", values_to = "score") %>% group_by(PID_groups, cat) %>% summarize(mean_score = mean(score, na.rm = TRUE)) %>% pivot_wider(names_from = "cat", values_from = "mean_score") %>% column_to_rownames("PID_groups")

# Row annotation 
highest_cat <- test %>% pivot_longer(cols = 1:27, names_to = "cat", values_to = "score") %>% group_by(PID_groups, cat) %>% summarize(mean_score = mean(score, na.rm = TRUE)) %>% ungroup() %>% group_by(cat) %>% mutate(max = max(mean_score)) %>% filter(mean_score == max) %>% select(cat, PID_groups) %>% as.data.frame %>% column_to_rownames("cat")

biomarker_df$PID_groups <- highest_cat$PID_groups[match(rownames(biomarker_df), rownames(highest_cat))]

# Column annotation 
col_anno <- anno %>% group_by(PID_groups) %>% count() %>% as.data.frame()
rownames(col_anno) <- col_anno$PID_groups 

# Align order
biomarker_df <- biomarker_df[match(colnames(mat_mean), rownames(biomarker_df)),]
test_result <- test_result[match(colnames(mat_mean), rownames(test_result)),]
col_anno <- col_anno[match(rownames(mat_mean), rownames(col_anno)),]

ca <- columnAnnotation(n_PID = anno_barplot(col_anno$n, gp = gpar(fill = "#440154FF")))

## Visualize 

# Heatmap 
h <- 
  rowAnnotation(bio = anno_text(biomarker_df$biomarker))+
  rowAnnotation(path = anno_simple(biomarker_df$pathway, col = setNames(brewer.pal(8, "Paired"), unique(biomarker_df$pathway)), width = unit(5, "mm")))+
  rowAnnotation(type = anno_simple(biomarker_df$type, col = setNames(brewer.pal(3, "Set1")[1:2], unique(biomarker_df$type)), width = unit(5, "mm")))+
  Heatmap(t(mat_mean), name = "mean_score", cluster_columns = TRUE, row_names_side = "left", show_row_dend = FALSE, top_annotation = ca)+
  rowAnnotation(cat = anno_simple(unfactor(biomarker_df$PID_groups), col = group_color, width = unit(5, "mm")))+
  rowAnnotation(test = anno_text(test_result$psignif.adj))


# Legends
lgd_cat <- Legend(title = "Highest_exprs", at = names(group_color), 
                  legend_gp = gpar(fill = group_color))
lgd_path <- Legend(title = "Pathway", at = names(setNames(brewer.pal(8, "Paired"), unique(biomarker_df$pathway))), 
                  legend_gp = gpar(fill = setNames(brewer.pal(8, "Paired"), unique(biomarker_df$pathway))))

lgd_type <- Legend(title = "Highest_exprs", at = names(setNames(brewer.pal(3, "Set1")[1:2], unique(biomarker_df$type))), 
                  legend_gp = gpar(fill = setNames(brewer.pal(3, "Set1")[1:2], unique(biomarker_df$type))))

# FINAL
draw(h,annotation_legend_list = list(lgd_cat, lgd_path, lgd_type))
```

# 5. Treatment association 

```{r}
new_expset$arm_short <- new_expset$`Arm (short name)`

treatment <- pData(new_expset) %>% select(PID, arm_short, response, PID_groups, TNBC)

out_check <- treatment

out_C <- out_check %>% group_by(PID_groups, arm_short, response) %>% dplyr::count() %>% group_by(PID_groups, arm_short) %>% mutate(fra = n/sum(n))

out_C$response <- factor(out_C$response, levels = c("1", "0"))
out_C$PID_groups <- factor(out_C$PID_groups, levels = c("inflamed", "KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

ggplot(out_C, aes(x = PID_groups, y = n, fill = response))+
  facet_wrap(~arm_short, scales = "free")+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = c("#8019C0", "#ECD2FC"))+
  coord_flip()

ggplot(out_C, aes(x = PID_groups, y = fra, fill = response))+
  facet_wrap(~arm_short)+
  geom_bar(stat = "identity", color = "black")+
  geom_hline(yintercept = 0.5, linetype = "dashed")+
  theme_classic()+
  scale_fill_manual(values = c("#8019C0", "#ECD2FC"))+
  coord_flip()

## Using epitab package
res_final <- list()

for(i in unique(out_C$PID_groups)){
  data <- out_C %>% mutate(PID_binary = ifelse(PID_groups == i, i, "Other"))
  for(y in unique(out_C$arm_short)){
    data_test <- data %>% filter(arm_short == y)
    if(length(unique(data_test$PID_binary)) > 1){
      res <- fisher.test(xtabs(n ~ PID_binary + response, data = data_test))
      out <- data.frame(odds_ratio = res$estimate, p = res$p.value, arm_short = y, group = i)
      res_final[[paste0(y, i)]] <- out
    }else{
      res_final[[paste0(y, i)]] <- data.frame(odds_ratio = NA, p = NA, arm_short = y, group = i)
    }
  }
}

res_final <- do.call("rbind", res_final)

isGreater <- res_final$odds_ratio > 1 & !is.na(res_final$odds_ratio)
pVal <- log10(res_final$p)
pVal[isGreater] <- abs(pVal[isGreater])
res_final$p_out <- pVal
colnames(res_final) <- c("odds_ratio", "p_value", "arm","category", "p_out")
```

```{r}
res_final$result <- ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.05, "UP",
                         ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.1, "UP_trend",
                                ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.05, "DOWN",
                                       ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.1, "DOWN_trend", "NOTSIG"))))

res_final$odds_ratio[res_final$odds_ratio == Inf] <- 1

## hierarchical clustering on odds_ratio
cur_df <- res_final %>% select(arm, category, odds_ratio) %>% pivot_wider(values_from = odds_ratio, names_from = arm)

# rows 
dd <- dist(cur_df[2:ncol(cur_df)], method = "euclidean")

hc <- hclust(dd, method = "ward.D2")
plot(hc)
row_sorted <- hc$order
res_final$category <- factor(res_final$category, levels = c("inflamed", "KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

# columns 
dd <- dist(t(cur_df[2:ncol(cur_df)]), method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
plot(hc)
column_sorted <- hc$order
res_final$arm <- factor(res_final$arm, levels = colnames(cur_df[2:ncol(cur_df)][,column_sorted]))
res_final_full <- res_final

ggplot(res_final, aes(y=category, x = arm,col = p_out, size = abs(p_out)))+
  geom_point()+
  theme_classic()+
  scale_color_gradient2(high = "#E41A1C", mid = "white", low = "#377EB8", midpoint = 0)+
  scale_size(range = c(2,10))+
  new_scale_colour() +
  geom_point(data = res_final,shape = 1,aes(size = abs(p_out), col = result), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214", "UP_trend" = "#E41A1C", "NOTSIG" = "snow2","DOWN_trend" = "#377EB8", "DOWN" = "#25567d"), name = "Direction")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
new_expset$arm_short <- new_expset$`Arm (short name)`

treatment <- pData(new_expset) %>% select(PID, arm_short, response, PID_group_risk, TNBC)

out_check <- treatment

out_C <- out_check %>% group_by(PID_group_risk, arm_short, response) %>% dplyr::count() %>% group_by(PID_group_risk, arm_short) %>% mutate(fra = n/sum(n))

out_C$response <- factor(out_C$response, levels = c("1", "0"))

ggplot(out_C, aes(x = PID_group_risk, y = n, fill = response))+
  facet_wrap(~arm_short, scales = "free")+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = c("#8019C0", "#ECD2FC"))+
  coord_flip()

ggplot(out_C, aes(x = PID_group_risk, y = fra, fill = response))+
  facet_wrap(~arm_short)+
  geom_bar(stat = "identity", color = "black")+
  geom_hline(yintercept = 0.5, linetype = "dashed")+
  theme_classic()+
  scale_fill_manual(values = c("#8019C0", "#ECD2FC"))+
  coord_flip()

## Using epitab package
res_final <- list()

for(i in unique(out_C$PID_group_risk)){
  data <- out_C %>% mutate(PID_binary = ifelse(PID_group_risk == i, i, "Other"))
  for(y in unique(out_C$arm_short)){
    data_test <- data %>% filter(arm_short == y)
    if(length(unique(data_test$PID_binary)) > 1){
      res <- fisher.test(xtabs(n ~ PID_binary + response, data = data_test))
      out <- data.frame(odds_ratio = res$estimate, p = res$p.value, arm_short = y, group = i)
      res_final[[paste0(y, i)]] <- out
    }else{
      res_final[[paste0(y, i)]] <- data.frame(odds_ratio = NA, p = NA, arm_short = y, group = i)
    }
  }
}

res_final <- do.call("rbind", res_final)

isGreater <- res_final$odds_ratio > 1 & !is.na(res_final$odds_ratio)
pVal <- log10(res_final$p)
pVal[isGreater] <- abs(pVal[isGreater])
res_final$p_out <- pVal
colnames(res_final) <- c("odds_ratio", "p_value", "arm","category", "p_out")
```


```{r}
res_final$result <- ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.05, "UP",
                         ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.1, "UP_trend",
                                ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.05, "DOWN",
                                       ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.1, "DOWN_trend", "NOTSIG"))))

res_final$odds_ratio[res_final$odds_ratio == Inf] <- 1

## hierarchical clustering on odds_ratio
cur_df <- res_final %>% select(arm, category, odds_ratio) %>% pivot_wider(values_from = odds_ratio, names_from = arm)

# rows 
dd <- dist(cur_df[2:ncol(cur_df)], method = "euclidean")

hc <- hclust(dd, method = "ward.D2")
plot(hc)
row_sorted <- hc$order
res_final$category <- factor(res_final$category, levels = c("low", "medium", "high"))

# columns 
dd <- dist(t(cur_df[2:ncol(cur_df)]), method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
plot(hc)
column_sorted <- hc$order
res_final$arm <- factor(res_final$arm, levels = colnames(cur_df[2:ncol(cur_df)][,column_sorted]))

library(ggnewscale)
ggplot(res_final, aes(x=category, y = arm,col = p_out, size = p_out))+
  geom_point()+
  theme_classic()+
  scale_color_gradient2(high = "#E41A1C", mid = "white", low = "#377EB8", midpoint = 0)+
  scale_size(range = c(2,10))+
  new_scale_colour() +
  geom_point(data = res_final,shape = 1,aes(size = p_out, col = result), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214", "UP_trend" = "#E41A1C", "NOTSIG" = "snow2","DOWN_trend" = "#377EB8", "DOWN" = "#25567d"), name = "Direction")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## TNBC subset

```{r}
new_expset$arm_short <- new_expset$`Arm (short name)`

treatment <- pData(new_expset) %>% select(PID, arm_short, response, PID_groups, TNBC)

treatment <- treatment %>% filter(TNBC == "TNBC")

out_check <- treatment

out_C <- out_check %>% group_by(PID_groups, arm_short, response) %>% dplyr::count() %>% group_by(PID_groups, arm_short) %>% mutate(fra = n/sum(n))

out_C$response <- factor(out_C$response, levels = c("1", "0"))
out_C$PID_groups <- factor(out_C$PID_groups, levels = c("inflamed", "KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))
out_C_TNBC<- out_C

ggplot(out_C, aes(x = PID_groups, y = n, fill = response))+
  facet_wrap(~arm_short, scales = "free")+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = c("#8019C0", "#ECD2FC"))+
  coord_flip()

ggplot(out_C, aes(x = PID_groups, y = fra, fill = response))+
  facet_wrap(~arm_short)+
  geom_bar(stat = "identity", color = "black")+
  geom_hline(yintercept = 0.5, linetype = "dashed")+
  theme_classic()+
  scale_fill_manual(values = c("#8019C0", "#ECD2FC"))+
  coord_flip()

## Using epitab package
res_final <- list()

for(i in unique(out_C$PID_groups)){
  data <- out_C %>% mutate(PID_binary = ifelse(PID_groups == i, i, "Other"))
  for(y in unique(out_C$arm_short)){
    data_test <- data %>% filter(arm_short == y)
    if(length(unique(data_test$PID_binary)) > 1){
      res <- fisher.test(xtabs(n ~ PID_binary + response, data = data_test))
      out <- data.frame(odds_ratio = res$estimate, p = res$p.value, arm_short = y, group = i)
      res_final[[paste0(y, i)]] <- out
    }else{
      res_final[[paste0(y, i)]] <- data.frame(odds_ratio = NA, p = NA, arm_short = y, group = i)
    }
  }
}

res_final <- do.call("rbind", res_final)

isGreater <- res_final$odds_ratio > 1 & !is.na(res_final$odds_ratio)
pVal <- log10(res_final$p)
pVal[isGreater] <- abs(pVal[isGreater])
res_final$p_out <- pVal
colnames(res_final) <- c("odds_ratio", "p_value", "arm","category", "p_out")
```

```{r}
res_final$result <- ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.05, "UP",
                         ifelse(res_final$odds_ratio > 1 & res_final$p_value < 0.1, "UP_trend",
                                ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.05, "DOWN",
                                       ifelse(res_final$odds_ratio < 1 & res_final$p_value < 0.1, "DOWN_trend", "NOTSIG"))))

res_final$odds_ratio[res_final$odds_ratio == Inf] <- 1

## hierarchical clustering on odds_ratio
cur_df <- res_final %>% select(arm, category, odds_ratio) %>% pivot_wider(values_from = odds_ratio, names_from = arm)

# rows 
dd <- dist(cur_df[2:ncol(cur_df)], method = "euclidean")

hc <- hclust(dd, method = "ward.D2")
plot(hc)
row_sorted <- hc$order
res_final$category <- factor(res_final$category, levels = c("inflamed", "KRT5/7-", "KRT7+", "KRT5+", "KRT5/7+"))

# columns 
dd <- dist(t(cur_df[2:ncol(cur_df)]), method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
plot(hc)
column_sorted <- hc$order
res_final$arm <- factor(res_final$arm, levels = colnames(cur_df[2:ncol(cur_df)][,column_sorted]))
res_final_TNBC <- res_final 

ggplot(res_final, aes(y=category, x = arm,col = p_out, size = abs(p_out)))+
  geom_point()+
  theme_classic()+
  scale_color_gradient2(high = "#E41A1C", mid = "white", low = "#377EB8", midpoint = 0)+
  scale_size(range = c(2,10))+
  new_scale_colour() +
  geom_point(data = res_final,shape = 1,aes(size = abs(p_out), col = result), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214", "UP_trend" = "#E41A1C", "NOTSIG" = "snow2","DOWN_trend" = "#377EB8", "DOWN" = "#25567d"), name = "Direction")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
res_final_TNBC$subset <- "TNBC"
res_final_full$subset <- "Full"
```

```{r}
res_final_full <- rbind(res_final_TNBC, res_final_full)

ggplot(res_final_full, aes(y=category, x = arm,col = p_out, size = abs(p_out)))+
  facet_wrap(~subset)+
  geom_point()+
  theme_classic()+
  scale_color_gradient2(high = "#E41A1C", mid = "white", low = "#377EB8", midpoint = 0)+
  scale_size(range = c(2,10))+
  new_scale_colour() +
  geom_point(data = res_final_full,shape = 1,aes(size = abs(p_out), col = result), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214", "UP_trend" = "#E41A1C", "NOTSIG" = "snow2","DOWN_trend" = "#377EB8", "DOWN" = "#25567d"), name = "Direction")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# 6. Visualization 

## Barplot visualization 

```{r}
plot <- as.data.frame(t(exprs(new_expset)[rownames(exprs(new_expset)) %in% c("KRT5", "KRT7", "KRT18","KRT14","HLA-DRA"),]))
plot$KRT5_7_score <- new_expset$KRT5_7_score[match(rownames(plot),colnames(new_expset))]
plot$PAM50 <- new_expset$PAM50.Subtype[match(rownames(plot),colnames(new_expset))]
plot$response <- new_expset$response[match(rownames(plot),colnames(new_expset))]
plot$TNBC <- new_expset$TNBC[match(rownames(plot),colnames(new_expset))]
plot$PID_groups <- new_expset$PID_groups[match(rownames(plot),colnames(new_expset))]
plot$PID_group_risk <- new_expset$PID_group_risk[match(rownames(plot),colnames(new_expset))]

# PID groups 

## PAM50 subtype 
bar_out <- plot %>% as.data.frame() %>% group_by(PID_groups, PAM50) %>% dplyr::count() %>% filter(PAM50 != "NA") %>% ungroup %>% group_by(PAM50) %>% mutate(fra = n/sum(n)) %>% arrange(PAM50)

ggplot(bar_out, aes(x = PAM50, y = fra, fill = PID_groups))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = group_color)+
  theme_classic()

plot %>% as.data.frame() %>% rownames_to_column("PID") %>% group_by(PAM50) %>% dplyr::count() %>% ungroup %>% mutate(fra = n/sum(n))

## TNBC 
bar_out <- plot %>% as.data.frame() %>% group_by(PID_groups, TNBC) %>% dplyr::count() %>% ungroup %>% group_by(TNBC) %>% mutate(fra = n/sum(n)) %>% arrange(TNBC)

ggplot(bar_out, aes(x = TNBC, y = fra, fill = PID_groups))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = group_color)+
  theme_classic()

bar_test <- bar_out %>% filter(TNBC != "NA") %>% pivot_wider(names_from = TNBC, id_cols = PID_groups, values_from = n) %>% column_to_rownames("PID_groups")
bar_test_out <- chisq.test(bar_test)
print(bar_test_out$p.value)

plot %>% as.data.frame() %>% rownames_to_column("PID") %>% group_by(TNBC) %>% dplyr::count() %>% ungroup %>% mutate(fra = n/sum(n))
```
```{r}
## General chemotherapy response 
bar_out <- plot %>% as.data.frame() %>% group_by(PID_groups, response) %>% dplyr::count() %>% filter(response != "NA") %>% ungroup %>% group_by(PID_groups) %>% mutate(fra = n/sum(n))

test <- chisq.test(xtabs(n ~ response + PID_groups, data = bar_out))
test$p.value

# Optional: Individual comparisons
fisher.test(xtabs(n ~ response + PID_groups, data = bar_out %>% 
                    filter(PID_groups %in% c("inflamed", "KRT5/7+")) %>% 
                    mutate(PID_groups = unfactor(PID_groups))))

bar_out$PID_groups <- factor(bar_out$PID_groups, levels = bar_out %>% filter(response == "1") %>% arrange(fra) %>% pull(PID_groups) %>% unfactor())

ggplot(bar_out, aes(x = PID_groups, y = fra, fill = response))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  theme_classic()+
  ggtitle(paste0("p = ", test$p.value))
```

```{r}
## General chemotherapy response 
bar_out <- plot %>% as.data.frame() %>% group_by(PID_group_risk, response) %>% dplyr::count() %>% filter(response != "NA") %>% ungroup %>% group_by(PID_group_risk) %>% mutate(fra = n/sum(n))

test <- chisq.test(xtabs(n ~ response + PID_group_risk, data = bar_out))
test$p.value

bar_out$PID_group_risk <- factor(bar_out$PID_group_risk, levels = bar_out %>% filter(response == "1") %>% arrange(fra) %>% pull(PID_group_risk) %>% unfactor())

ggplot(bar_out, aes(x = PID_group_risk, y = fra, fill = response))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  theme_classic()+
  ggtitle(paste0("p = ", test$p.value))
```

## Alluvial 

```{r}
df_alluvial <- pData(new_expset) %>% as.data.frame %>% dplyr::select(KRT5_7_score,GMM_HLADRA,PID_groups, PID_group_risk) #PID_groups_risk

alluvial_wide(df_alluvial, 
              colorful_fill_variable_stratum = FALSE,
              col_vector_flow = c("#85D4E3","grey", "#9C964A", "#F4B5BD"), 
              order_levels = c("KRT5/7+","KRT5/7-","KRT7+","KRT5+", "high", "medium", "low"))+
  theme_classic()
```

## Key markers and gene signatures

### Key 

```{r}
# Define key markers of interest
key_marker <- c("KRT5","KRT7","HLA-DRA")

# Define heatmap based on normalized counts 
df <- t(exprs(new_expset)[key_marker,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")

## Create heatmap body 
df$PID_groups <- new_expset$PID_groups[match(df$PID, colnames(new_expset))]

mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

## Create row annotation 
anno <- df %>% select(PID_groups) %>% group_by(PID_groups) %>% dplyr::count()
anno <- anno[match(anno$PID_groups, rownames(ht)),]

## Print heatmap 
Heatmap(ht, name = "Mean Z-scores")

Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  #rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$PID_groups))
```

### Immune 

```{r}
# Define key immune markers of interest
#key_marker <- c("CD8A","CD274","PDCD1","GZMB","PDCD1LG2","IDO1")

key_marker <- c("HLA-DRA","PDCD1","GZMB", "ICOS", "TOX") # additional immune markers

key_marker <- key_marker[key_marker %in% rownames(exprs(new_expset))]

# Define heatmap based on normalized counts 
df <- t(exprs(new_expset)[key_marker,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")

## Create heatmap body 
df$PID_groups <- new_expset$PID_groups[match(df$PID, colnames(new_expset))]

mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

## Create row annotation 
anno <- df %>% select(PID_groups) %>% group_by(PID_groups) %>% dplyr::count()
anno <- anno[match(anno$PID_groups, rownames(ht)),]

## Print heatmap 
Heatmap(ht, col = viridis(100), name = "Mean Z-scores")+

Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  #rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$PID_groups))

Heatmap(scale(ht), name = "Scaled mean Z-scores")
```

### Immunosuppression

```{r}
# Define key immune markers of interest
key_marker <- c("TGFB2", "TGFB1", "PROM1", "PVR", "PVRL1","CD276") #"TGFB1"
key_marker <- key_marker[key_marker %in% rownames(exprs(new_expset))]

# Define heatmap based on normalized counts 
df <- t(exprs(new_expset)[key_marker,]) %>% as.data.frame 
df <- df %>% rownames_to_column("PID")

## Create heatmap body 
df$PID_groups <- new_expset$PID_groups[match(df$PID, colnames(new_expset))]

mat <- df %>% pivot_longer(cols = 2:(ncol(df)-1), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

## Create row annotation 
anno <- df %>% select(PID_groups) %>% group_by(PID_groups) %>% dplyr::count()
anno <- anno[match(anno$PID_groups, rownames(ht)),]

## Print heatmap 
Heatmap(ht, name = "Mean Z-scores")

Heatmap(scale(ht), name = "Scaled mean Z-scores")+
  #rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$PID_groups))
```

### IFNg

```{r}
INFg_sig_up <- c("IFNG", "STAT1", "CCR5", "CXCL9", "CXCL10", "CXCL11", "IDO1", "PRF1", "GZMA") #excluded HLA-DRA from signature 
```

```{r}
## 2. BL signature per PID group 
INFg_sig_up_rel <- INFg_sig_up[INFg_sig_up %in% rownames(exprs(new_expset))]
# Filter for available genes
df <- t(exprs(new_expset)[INFg_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_groups <- new_expset$PID_groups[match(rownames(df), colnames(new_expset))]

compare_means(signature_sum ~ PID_groups,df) #pairwise comparisons
compare_means(signature_sum ~ PID_groups,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = PID_groups, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = PID_groups), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = group_color)+
  theme_classic()+
  ylab("Basal-luminal signature (Li et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

Heatmap(ht, name = "Mean signature Z-scores")

INFg <- mat %>% group_by(PID_groups) %>% summarize(INFg = mean(mean)) %>% column_to_rownames("PID_groups")

Heatmap(INFg)
```

### CTA

```{r}
CTA_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/CTAlist_PNAS_2019.xlsx")

CTA_sig_up <- CTA_sig$`Cancer Testis Antigens` %>% unique()
CTA_sig_up <- toupper(CTA_sig_up)
```

```{r}
## 2. BL signature per PID group 
CTA_sig_up_rel <- CTA_sig_up[CTA_sig_up %in% rownames(exprs(new_expset))]
# Filter for available genes
df <- t(exprs(new_expset)[CTA_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_groups <- new_expset$PID_groups[match(rownames(df), colnames(new_expset))]

compare_means(signature_sum ~ PID_groups,df) #pairwise comparisons
compare_means(signature_sum ~ PID_groups,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = PID_groups, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = PID_groups), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = group_color)+
  theme_classic()+
  ylab("Basal-luminal signature (Li et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

Heatmap(ht, name = "Mean signature Z-scores")

CTA <- mat %>% group_by(PID_groups) %>% summarize(CTA = mean(mean)) %>% column_to_rownames("PID_groups")

Heatmap(CTA)
```

### INFab

```{r}
INFab_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/INFALPHABETA_Signaling_Reactome.xlsx")

INFab_sig_up <- INFab_sig$INF_SIG %>% unique()
INFab_sig_up <- toupper(INFab_sig_up)
```

```{r}
## 2. BL signature per PID group 
INFab_sig_up_rel <- INFab_sig_up[INFab_sig_up %in% rownames(exprs(new_expset))]
# Filter for available genes
df <- t(exprs(new_expset)[INFab_sig_up_rel ,]) %>% as.data.frame
# Z-score gene expression 
df <- as.data.frame(scale(df))

df$signature_sum <- rowSums(df,na.rm = TRUE)
df$PID_groups <- new_expset$PID_groups[match(rownames(df), colnames(new_expset))]

compare_means(signature_sum ~ PID_groups,df) #pairwise comparisons
compare_means(signature_sum ~ PID_groups,df, method = "kruskal.test") #global comparison

ggplot(df, aes(x = PID_groups, y = signature_sum))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = PID_groups), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = group_color)+
  theme_classic()+
  ylab("Basal-luminal signature (Li et al.)")+
  stat_compare_means()

mat <- df %>% pivot_longer(cols = 1:(ncol(df)-2), names_to = "gene", values_to = "exprs") %>% group_by(PID_groups, gene) %>% dplyr::summarize(mean = mean(exprs))
mat <- mat %>% filter(!is.na(mean))
ht <- mat %>% pivot_wider(names_from = "gene", values_from = "mean") %>% column_to_rownames("PID_groups")

Heatmap(ht, name = "Mean signature Z-scores")

INFab <- mat %>% group_by(PID_groups) %>% summarize(INFab = mean(mean)) %>% column_to_rownames("PID_groups")

Heatmap(INFab)
```
```{r}
Heatmap(CTA, name = "CTA", cluster_rows = TRUE)+
  Heatmap(INFab, name = "INFab", cluster_rows = FALSE)+
  Heatmap(INFg, name = "INFg", cluster_rows = FALSE)
```

# Save objects 

```{r}
saveRDS(new_expset, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/ISPY2_RNA_analysis_expressionset.rds")
```

# Session Info 

```{r}
sessionInfo()
```
