---
title: "TNBC_Metadata_Update"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Aim: Create updated metasheet for TNBC TMA cohorts and visualize cohort heatmaps**

# 1. ZTMA174

### Load metadata csv files

```{r load and combine meta, message=FALSE,warning=FALSE}
library(tidyverse)
library(readxl)
library(knitr)

### TMA174

#General metadata
Meta_174 <- as.data.frame(read_excel("/Users/lasmey/Desktop/TNBC/Metadaten/ZTMA174_TNBC_LM_new.xlsx",col_names = T, na="NA",col_types = "text")) 

#change columns to numeric
Meta_174 <- Meta_174 %>% mutate(tumor_size_cm = as.numeric(tumor_size_cm)) %>% mutate_if(is.numeric,round,digits=2) %>% filter(!is.na(PID))

#Load filenames for IMC data and check overlap with metadata
TNBC_raw <- list.files("/Volumes/rcc_volume/TNBC/raw/")

TNBC_files_174 <- TNBC_raw[str_detect(TNBC_raw,pattern = "_ZTMA174_")]

## 1. Filter PID for which metadata is missing (should be HER2 cases)
PID_174 <- str_split(TNBC_files_174,"_",simplify = TRUE)[,3] %>% unique()

PID_174_missing <- PID_174[!PID_174 %in% Meta_174$PID] #Patients for which we have samples but no metadata (should be HER2 positive; n=22)

#write_csv(as.data.frame(PID_174_missing),"/Users/lasmey/Desktop/TNBC/Metadaten/ZTMA174_BnummerHER2.csv")

# #move HER2 cases to `exclude` folder
# PID_174_missing_files <- TNBC_files_174[str_split(TNBC_files_174,"_",simplify = TRUE)[,3] %in% PID_174_missing]
# 
# file.copy(from = paste0("/Volumes/rcc_volume/TNBC/raw/",PID_174_missing_files),
#           to = "/Volumes/rcc_volume/TNBC/excluded/")
# 
# file.remove(from = paste0("/Volumes/rcc_volume/TNBC/raw/",PID_174_missing_files))


## 2. Filter PID for which metadata is there but not samples
PID_174_missing_2 <- Meta_174[!Meta_174$PID %in% PID_174,] %>% pull(PID) #Patients for which we have metadata but no samples (n=2)

PID_174_missing_2

#Remove from file
Meta_174 <- Meta_174 %>% filter(!PID %in% PID_174_missing_2)

##double-check that entries are distinct (which is TRUE)
dup <- Meta_174 %>% 
  count(PID) %>% 
  filter(n>1) %>% 
  pull(PID)
```

### Tidy up data for processing and display

```{r tidy, message = F, warning = F}
library(DT)
library(prodlim)

#Count number of samples per PID
n_samples_PID_174 <- str_split(TNBC_files_174,"_",simplify = TRUE)[,3] %>% table(dnn = "PID") %>% as.data.frame()
Meta_174$n_samples <- n_samples_PID_174$Freq[match(Meta_174$PID,n_samples_PID_174$PID)]

#Age bin for patients (five groups)
Meta_174$age <- as.numeric(Meta_174$age)
Meta_174 <- Meta_174 %>% mutate(age_bin = ifelse(age <= 40, "27-40", 
                                   ifelse(40<age & age<=55, "41-55",
                                          ifelse(55<age & age<=70, "56-70",
                                                 ifelse(70<age & age<=85, "71-85",
                                                        ifelse(85<age & age<=90, "85-88",NA))))))

#Simplified grade and TNM staging
Meta_174$grade <- str_extract(Meta_174$grade,"[0-3]")
Meta_174$pT_simple <- str_extract(Meta_174$pT,"[0-4]")
Meta_174$pN_simple <- str_extract(Meta_174$pN,"[0-4]")
Meta_174$pM_simple <- str_extract(Meta_174$pM,"[0-1]")

#Survival data availability
Meta_174 <- Meta_174 %>% mutate(OS_data = ifelse(!is.na(Meta_174$status_OS),TRUE,FALSE)) %>% mutate(DFS_data = ifelse(!is.na(Meta_174$status_DFS),TRUE,FALSE))

#survival data censoring update (label switch for DFS?)
Meta_174 <- Meta_174 %>% mutate(status_OS = ifelse(status_OS == "2" & is.na(death_differentcause), "1","0")) %>% mutate(status_DFS = ifelse(status_DFS == "1", "1", "0"))
```

```{r save processed meta174}
## Save processed metadata
write_excel_csv(Meta_174,"~/Desktop/TNBC/Metadaten/ZTMA174_TNBC_LM_processed.csv")
```

### Print heatmap based overview
based on the [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html) package and the implemented `HeatmapAnnotation` function  

```{r viz, message=F, warning = F,fig.width=14, fig.height=8, fig.align='center'}
library(ComplexHeatmap)
library(foreach)
library(ggsci)
library(RColorBrewer)

#Filter for columns to visualize
Meta_viz <- Meta_174 %>% select(PID,n_samples,age_bin,tumor_size_cm,grade,pT_simple,pN_simple,pM_simple,ER,PR,Her2,OS_data, DFS_data)

#Convert empty columns and NAs for easier handling with ComplexHeatmap package
Meta_viz[is.na(Meta_viz)] <- "Unknown"

Meta_viz <- Meta_viz %>% arrange(desc(pT_simple),desc(pN_simple))

## Set Color scheme (also for downstream analysis)
color_vectors <- list(
PID = setNames(colorRampPalette(brewer.pal(11, "BrBG"))(length(unique(Meta_viz$PID))), 
                sort(unique(Meta_viz$PID))),
n_samples = setNames(colorRampPalette(brewer.pal(2, "Greens"))(length(unique(Meta_viz$n_samples))), 
                sort(unique(Meta_viz$n_samples))),
age_bin = setNames(colorRampPalette(brewer.pal(9, "Greens"))(length(unique(Meta_viz$age_bin))), 
                sort(unique(Meta_viz$age_bin))),
grade = setNames(colorRampPalette(brewer.pal(3, "Oranges"))(length(unique(Meta_viz$grade))), 
                sort(unique(Meta_viz$grade))),
pT_simple = setNames(colorRampPalette(brewer.pal(9, "Blues"))(length(unique(Meta_viz$pT_simple))), 
                sort(unique(Meta_viz$pT_simple))),
pN_simple = setNames(c(colorRampPalette(brewer.pal(9, "Purples"))(length(unique(Meta_viz$pN_simple))-1),"black"), 
                sort(unique(Meta_viz$pN_simple))),
pM_simple = setNames(c(colorRampPalette(brewer.pal(9, "RdPu"))(length(unique(Meta_viz$pM_simple))-1),"black"), 
                sort(unique(Meta_viz$pM_simple))),
ER = setNames("#A6CEE3",sort(unique(Meta_viz$ER))),
PR = setNames(c("#A6CEE3", "#1F78B4"),sort(unique(Meta_viz$PR))),
Her2 = setNames("#A6CEE3",sort(unique(Meta_viz$Her2))),
OS_data = setNames(c("black","grey"),sort(unique(Meta_viz$OS_data))),
DFS_data = setNames(c("black","grey"),sort(unique(Meta_viz$DFS_data)))
)

#Save color list object for downstream usage
#saveRDS(col_list_1,file="/Users/lasmey/Desktop/RCC/Metadaten/color.rds")

#Heatmap Annotation
sample_anno <- HeatmapAnnotation(PID = Meta_viz$PID,
                                 n_samples = Meta_viz$n_samples,
                                 age_bin = Meta_viz$age_bin,
                                 tumor_size_cm = Meta_viz$tumor_size_cm,
                                 grade = Meta_viz$grade,
                                 pT_simple = Meta_viz$pT_simple,
                                pN_simple = Meta_viz$pN_simple,
                                pM_simple = Meta_viz$pM_simple,
                                ER = Meta_viz$ER,
                                PR = Meta_viz$PR,
                                Her2 = Meta_viz$Her2,
                                OS_data = Meta_viz$OS_data,
                                DFS_data = Meta_viz$DFS_data,
                                col = color_vectors,
                                border = T,
                                na_col = "black",
                                gap = unit(5, "points"),
                                show_annotation_name = T,
                                show_legend = c(PID = FALSE)
)%v%NULL

#Display plot
draw(sample_anno)
```

# 2. ZTMA249

### Load metadata csv files

```{r load and combine meta_1, message=FALSE,warning=FALSE}
library(tidyverse)
library(readxl)
library(knitr)

### TMA249

#General metadata
Meta_249 <- as.data.frame(read_excel("/Users/lasmey/Desktop/TNBC/Metadaten/ZTMA249_TNBC_LM_new.xlsx",col_names = T, na="NA")) 

##Load filenames for IMC data and check overlap with metadata
TNBC_raw <- list.files("/Volumes/rcc_volume/TNBC/raw/")
TNBC_files_249 <- TNBC_raw[str_detect(TNBC_raw,pattern = "_ZTMA249_")]

#link spot_numbers with ID in filesnames
ID_numbers <- sort(as.numeric(str_split(TNBC_files_249,"_",simplify = TRUE)[,4]))

#detect breaks
diff(ID_numbers)

#Filter PID for which metadata or samples are missing
ID_meta <- Meta_249$ID_Number %>% unique()
PID_274_missing_2 <- Meta_249 %>% filter(ID_Number == ID_meta[!ID_meta %in% ID_numbers]) %>% pull(PID) #Patients for which we have metadata but no samples ("B14.19170")


ID_meta[!ID_meta %in% ID_numbers]

#Remove from file
Meta_249 <- Meta_249 %>% filter(!PID %in% PID_274_missing_2)

#Generate sample ID from PID and ID
Meta_249 <- Meta_249 %>% unite("sample_ID","PID","ID_Number",sep="_", remove = FALSE)

##double-check that entries are distinct (TRUE)
dup <- Meta_249 %>% 
  count(sample_ID) %>% 
  filter(n>1) %>% 
  pull(sample_ID)
```

### Tidy up data for processing and display

```{r tidy_1,message = F, warning = F}
library(DT)
library(prodlim)

#1. Lehmann subtypes
Meta_249 <- Meta_249 %>% mutate(lehmann_subtype = ifelse(!is.na(Meta_249$basal_like_1),"BL1",
                                             ifelse(!is.na(Meta_249$basal_like_2),"BL2",
                                                    ifelse(!is.na(Meta_249$immunomodulatory),"IM",
                                                           ifelse(!is.na(Meta_249$mesenchymal_like_ML),"M",
                                                                  ifelse(!is.na(Meta_249$mesenchymal_stem_like_MSL),"MSL",
                                                                         ifelse(!is.na(Meta_249$luminal_androgen_receptor_LAR),"LAR",NA))))))) %>%
  unite("Lehmann_Binary","basal_like_1","basal_like_2","immunomodulatory","mesenchymal_like_ML","mesenchymal_stem_like_MSL","luminal_androgen_receptor_LAR",na.rm = TRUE, remove = FALSE)


#Double-check that Lehmann subtypes are unique (not for B14.54051_262)
SID_249_lehmann_unclear <- Meta_249[!Meta_249$Lehmann_Binary %in% c("Ja","ja", ""),] %>% pull(sample_ID)

Meta_249 <- Meta_249 %>% mutate(lehmann_subtype = ifelse(Meta_249$sample_ID == SID_249_lehmann_unclear, "Unclear",Meta_249$lehmann_subtype))
```

```{r save_processed}
write.csv(Meta_249, "/Users/lasmey/Desktop/TNBC/Metadaten/ZTMA249_TNBC_LM_processed_bysample.csv", col.names = TRUE)
```

```{r tidy_2, message=F,warning=F}
## Collapse data by PID (two samples per PID)
Meta_249 <- aggregate(Meta_249,by=list(Meta_249$PID),function(x){paste(x[!is.na(x)], collapse = ",")})
Meta_249 <- Meta_249 %>% mutate(PID = Meta_249$Group.1) %>% select(-"Group.1")

#1. Decode and simplify TNM staging 

decode_pT_df <- data.frame(value = as.character(seq(1,10)), 
                        key = c("NO_TUMOR","ypT1a","ypT2","ypT3","pT1a","pT1c","pT2","pT3","pT4","NODAL_MET_CHEMO"))

Meta_249$pT <- decode_pT_df$key[match(Meta_249$pT, decode_pT_df$value)]

decode_pN_df <- data.frame(value = as.character(seq(1,8)), 
                        key = c("cN1","ypN0","ypN2","pN0","pN1","ypN2","ypN3","pN3"))

Meta_249$pN <- decode_pN_df$key[match(Meta_249$pN, decode_pN_df$value)]

Meta_249$pT_simple <- str_extract(Meta_249$pT,"[0-4]")
Meta_249$pN_simple <- str_extract(Meta_249$pN,"[0-4]")
Meta_249$pM_simple <- Meta_249$pM

#3. Survival data availability and n_samples
Meta_249 <- Meta_249 %>% mutate(OS_data = ifelse(Meta_249$OS_months != "", TRUE,FALSE)) %>% mutate(DFS_data = ifelse(Meta_249$OS_months != "",TRUE,FALSE)) %>% mutate(n_samples = 2)

#4. Perform OS/DFS status censoring for survival analysis
Meta_249 <- Meta_249 %>% mutate(status_OS = ifelse(status_2018 == "4", "1","0")) %>% mutate(status_DFS = ifelse(status_2018 %in% c("2","3","4"), "1", "0"))

Meta_249$status_OS <- ifelse(Meta_249$OS_data == FALSE,NA,Meta_249$status_OS)
Meta_249$status_DFS <- ifelse(Meta_249$OS_data == FALSE,NA,Meta_249$status_DFS)
```

```{r save_processed 2}
write.csv(Meta_249, "/Users/lasmey/Desktop/TNBC/Metadaten/ZTMA249_TNBC_LM_processed.csv")
```


### Print heatmap based overview
based on the [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html) package and the implemented `HeatmapAnnotation` function  

```{r viz_1, message=F, warning = FALSE,fig.width=14, fig.height=8, fig.align='center'}
library(ComplexHeatmap)
library(foreach)
library(ggsci)
library(RColorBrewer)

#Filter for columns to visualize
Meta_viz <- Meta_249 %>% select(PID,n_samples,grade,pT_simple,pN_simple,pM_simple,OS_data, DFS_data)

#Convert empty columns and NAs for easier handling with ComplexHeatmap package
Meta_viz[is.na(Meta_viz)] <- "Unknown"

Meta_viz <- Meta_viz %>% arrange(desc(pT_simple),desc(pN_simple), desc(grade))

## Set Color scheme (also for downstream analysis)
color_vectors <- list(
PID = setNames(colorRampPalette(brewer.pal(11, "BrBG"))(length(unique(Meta_viz$PID))), 
                sort(unique(Meta_viz$PID))),
n_samples = setNames(colorRampPalette(brewer.pal(2, "Greens"))(length(unique(Meta_viz$n_samples))), 
                sort(unique(Meta_viz$n_samples))),
grade = setNames(colorRampPalette(brewer.pal(3, "Oranges"))(length(unique(Meta_viz$grade))), 
                sort(unique(Meta_viz$grade))),
pT_simple = setNames(c(colorRampPalette(brewer.pal(9, "Blues"))(length(unique(Meta_viz$pT_simple))-1),"black"), 
                sort(unique(Meta_viz$pT_simple))),
pN_simple = setNames(c(colorRampPalette(brewer.pal(9, "Purples"))(length(unique(Meta_viz$pN_simple))-1),"black"), 
                sort(unique(Meta_viz$pN_simple))),
pM_simple = setNames(c(colorRampPalette(brewer.pal(2, "RdPu"))(length(unique(Meta_viz$pM_simple))-1),"black"), 
                sort(unique(Meta_viz$pM_simple))),
OS_data = setNames(c("black","grey"),sort(unique(Meta_viz$OS_data))),
DFS_data = setNames(c("black","grey"),sort(unique(Meta_viz$DFS_data)))
)

#Save color list object for downstream usage
#saveRDS(col_list_1,file="/Users/lasmey/Desktop/RCC/Metadaten/color.rds")

#Heatmap Annotation
sample_anno <- HeatmapAnnotation(PID = Meta_viz$PID,
                                 n_samples = Meta_viz$n_samples,
                                 grade = Meta_viz$grade,
                                 pT_simple = Meta_viz$pT_simple,
                                pN_simple = Meta_viz$pN_simple,
                                pM_simple = Meta_viz$pM_simple,
                                OS_data = Meta_viz$OS_data,
                                DFS_data = Meta_viz$DFS_data,
                                col = color_vectors,
                                border = T,
                                na_col = "black",
                                gap = unit(5, "points"),
                                show_annotation_name = T,
                                show_legend = c(PID = FALSE)
)%v%NULL

#Display plot
draw(sample_anno)
```

#3. Combined - ZTMA174 and ZTMA249

### Combine metadata from both 

```{r combine}
#select columns of interest 
Meta_174_comb <- Meta_174 %>% select("TMANr","SpotNr","PID","n_samples","age","age_bin","tumor_size_cm","grade","pT","pT_simple","pN","pN_simple","pM","pM_simple","status_OS","OS_months", "status_DFS","DFS_months","OS_data","DFS_data")

Meta_249$TMANr <- "TMA249"
Meta_249_comb <- Meta_249 %>% select("TMANr","SpotNr","PID","n_samples","lehmann_subtype","grade","pT","pT_simple","pN","pN_simple","pM","pM_simple","status_OS","OS_months", "status_DFS","DFS_months","OS_data","DFS_data")

Meta <- bind_rows(Meta_174_comb,Meta_249_comb)

Meta$n_samples <- as.character(Meta$n_samples)
Meta$tumor_size_cm <- as.numeric(Meta$tumor_size_cm)
```

```{r save combined}
write.csv(Meta, "/Users/lasmey/Desktop/TNBC/Metadaten/Metadata_Combined_TNBC_LM.csv")
```

### Print heatmap based overview
based on the [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html) package and the implemented `HeatmapAnnotation` function

```{r}
library(ComplexHeatmap)
library(foreach)
library(ggsci)
library(RColorBrewer)

#Filter for columns to visualize
Meta_viz <- Meta %>% select(TMANr,PID,n_samples,grade,pT_simple,pN_simple,pM_simple,OS_data, DFS_data, age_bin, tumor_size_cm, lehmann_subtype)

#Convert empty columns and NAs for easier handling with ComplexHeatmap package
Meta_viz[is.na(Meta_viz)] <- "NA"

Meta_viz$tumor_size_cm <- as.numeric(Meta_viz$tumor_size_cm)

#Arrange
Meta_viz <- Meta_viz %>% arrange(desc(pT_simple),desc(pN_simple), desc(grade))

## Set Color scheme (also for downstream analysis)
color_vectors <- list(
TMANr = setNames(colorRampPalette(brewer.pal(3, "Dark2"))(length(unique(Meta_viz$TMANr))), 
                sort(unique(Meta_viz$TMANr))),
PID = setNames(colorRampPalette(brewer.pal(12, "Paired"))(length(unique(Meta_viz$PID))), 
                sort(unique(Meta_viz$PID))),
n_samples = setNames(c("#A6CEE3", "#1F78B4"), 
                sort(unique(Meta_viz$n_samples))),
grade = setNames(colorRampPalette(brewer.pal(3, "Oranges"))(length(unique(Meta_viz$grade))), 
                sort(unique(Meta_viz$grade))),
pT_simple = setNames(c(colorRampPalette(brewer.pal(9, "Blues"))(length(unique(Meta_viz$pT_simple))-1),"black"), 
                sort(unique(Meta_viz$pT_simple), na.last = TRUE)),
pN_simple = setNames(c(colorRampPalette(brewer.pal(9, "Purples"))(length(unique(Meta_viz$pN_simple))-1),"black"), 
                sort(unique(Meta_viz$pN_simple))),
pM_simple = setNames(c(colorRampPalette(brewer.pal(3, "PuRd"))(length(unique(Meta_viz$pM_simple))-1),"black"), 
                sort(unique(Meta_viz$pM_simple))),
OS_data = setNames(c("black","grey"),sort(unique(Meta_viz$OS_data))),
DFS_data = setNames(c("black","grey"),sort(unique(Meta_viz$DFS_data))),
age_bin = setNames(c(colorRampPalette(brewer.pal(11, "BrBG"))(length(unique(Meta_viz$age_bin))-1),"black"), 
                sort(unique(Meta_viz$age_bin))),
lehmann_subtype = setNames(c(colorRampPalette(brewer.pal(9, "Set1"))(length(unique(Meta_viz$lehmann_subtype))-1),"black"), 
                sort(unique(Meta_viz$lehmann_subtype)))
)

#Save color vector  object for downstream usage
saveRDS(color_vectors,file="/Volumes/rcc_volume/TNBC/data_analysis/color_meta.rds")

#Heatmap Annotation
sample_anno <- HeatmapAnnotation(TMANr = Meta_viz$TMANr,
                                 PID = Meta_viz$PID,
                                 n_samples = Meta_viz$n_samples,
                                 grade = Meta_viz$grade,
                                 pT_simple = Meta_viz$pT_simple,
                                pN_simple = Meta_viz$pN_simple,
                                pM_simple = Meta_viz$pM_simple,
                                OS_data = Meta_viz$OS_data,
                                DFS_data = Meta_viz$DFS_data,
                                age_bin = Meta_viz$age_bin,
                                tumor_size_cm = Meta_viz$tumor_size_cm,
                                lehmann_subtype = Meta_viz$lehmann_subtype,
                                col = color_vectors,
                                border = T,
                                na_col = "black",
                                gap = unit(5, "points"),
                                show_annotation_name = T,
                                show_legend = c(PID = FALSE)
)%v%NULL

#Display plot
draw(sample_anno)

pdf("~/Desktop/sample_anno.pdf",width=12,height = 8)
sample_anno
dev.off()
```


