---
title: "SCRNAseq_organoid_analysis"
author: "Haithem Dakhli"
date: "`r Sys.Date()`"
output: html_document
---

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: sce_organoid.rds

# Validation

Data analysis scripts for data analysis in the validation cohorts. 

# Script: Breast cancer organoid scRNAseq preprocessing

Here, we preprocess scRNA-seq data for 6 different breast cancer organoid lines. 

**DISCLAIMER**: This code is related to an ongoing study (Dakhli et al., unpublished) of a larger set of organoid lines and is only fully reproducible with the full set of lines. Here, we describe the main coding steps for the 6 lines analyzed in this study.

# Software requirements

```{r}
library(SingleCellExperiment)
library(SummarizedExperiment)
library(MatrixGenerics)
library(matrixStats)
library(GenomicRanges)
library(stats4)
library(BiocGenerics)
library(S4Vectors)
library(IRanges)
library(GenomeInfoDb)
library(Biobase)
library(DT)
library(DropletUtils)
```

# 1. Read data 

```{r read-data, message=FALSE, warning=FALSE}
file_list <- list()
data_path <- list()
sce_list <- list()

for (x in 1:length(list.files("~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/20230301_scOrganoidseq_HD/"))) {
  
file_list[[x]] <- list.files("~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/20230301_scOrganoidseq_HD/")[x]
data_path[[x]] <- paste0(paste0('~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/20230301_scOrganoidseq_HD/', file_list[[x]]) , '/filtered_feature_bc_matrix')

sce_list[[x]] <- read10xCounts(data_path[[x]])

}

sce_list
```

Save `SingleCellExperiment` object.

```{r save-sce}
saveRDS(sce_list, "~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/scripts/20230301_scOrganoidseq_HD/Robjects/sce_list.rds")
```

# 2. QC

Read single-cell data.

```{r read-data, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(tidyverse)
library(scran)
library(scater)
library(scuttle)
library(scDblFinder)
library(BiocSingular)
library(bluster)
library(BiocNeighbors)
library(dittoSeq)
library(viridis)
library(DT)

sce_list <- readRDS("~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/scripts/20230301_scOrganoidseq_HD/Robjects/sce_list.rds")

file_list <- list()

for (x in 1:length(list.files("~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/20230301_scOrganoidseq_HD/"))) {
file_list[[x]] <- list.files("~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/20230301_scOrganoidseq_HD/")[x]
}

# #Replace Normal Tissue by ID numbers
# file_list[[1]] <- 'N_BB491'
# file_list[[2]] <- 'N_BB505'
# file_list[[3]] <- 'N_BB528'

for (x in 1:length(file_list)) {
  
colnames(sce_list[[x]]) <- paste0(paste0(file_list[[x]], "_"), sce_list[[x]]$Barcode)
rownames(sce_list[[x]]) <- rowData(sce_list[[x]])$Symbol

}
```

## Compute UMAP for visualization

First, we will normalize the raw data. To keep the unnormalized data in place
we will duplicate the object here.

```{r normalize, message=FALSE, warning=FALSE}
set.seed(221215)
cur_clust <- list()
sce_2 <- list()

for (x in 1:length(sce_list)) {
  
cur_clust[[x]] <- quickCluster(sce_list[[x]]) 

sce_2[[x]] <- computeSumFactors(sce_list[[x]], cluster=cur_clust[[x]])

sce_2[[x]] <- logNormCounts(sce_2[[x]])

}

```

Now, we will find the highly variable genes.

```{r HVG}
# Fit the mean variance trend

cur_var <- list()
cur_fit <- list()
cur_mat_final <- data.frame()

for (x in 1:length(sce_list)) {
  
cur_var[[x]] <- modelGeneVar(sce_2[[x]])
cur_fit[[x]] <- metadata(cur_var[[x]])

}

for (y in 1:length(sce_list)) {

cur_mat <- as.data.frame(matrix(ncol = 2, nrow = 21134))
cur_mat[,1] <- cur_fit[[y]]$mean
cur_mat[,2] <- cur_fit[[y]]$var
cur_mat[,3] <- file_list[[y]]

cur_mat_final <- rbind(cur_mat_final, cur_mat)

}

ggplot(cur_mat_final, aes(x = V1, y = V2)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  facet_wrap(~V3) +
  labs(x = "Mean of log-expression", y = "Variance of log-expression")

# Order genes by their biological variance

hvg <- list()
for (x in 1:length(sce_list)) {
  
cur_var[[x]][order(cur_var[[x]]$bio, decreasing=TRUE),] %>%
  as.data.frame() %>%
  head(50)

## Select top 1000 HVG
hvg[[x]] <- getTopHVGs(cur_var[[x]], n=1000)

}
```

And compute the UMAP.

```{r runUMAP, message=FALSE}
set.seed(221215)
for (x in 1:length(sce_list)) {
  
sce_2[[x]] <- runUMAP(sce_2[[x]], subset_row = hvg[[x]])

# Transfer UMAP to original SCE object
reducedDim(sce_list[[x]], "UMAP") <- reducedDim(sce_2[[x]], "UMAP")
}
```

## Compute quality control measures

Here, we use the `scuttle` package to compute the library size and proportion
of mitochondrial counts.

```{r qc-diagnostics, message=FALSE}

is.mito <- list()
df <- list()

for (x in 1:length(sce_list)) {
  
is.mito[[x]] <- grepl("^MT-", rownames(sce_list[[x]]))
rownames(sce_list[[x]])[is.mito[[x]]]

df[[x]] <- perCellQCMetrics(sce_list[[x]], subsets=list(Mito=is.mito[[x]]))

colData(sce_list[[x]]) <- cbind(colData(sce_list[[x]]), df[[x]])

}
```

We will now visualize a number of quality metrics.

### Percent mitochondrial genes

First, we will visualize the fraction of mitochondrial genes against the 
library size.

```{r fraction-mito, message = FALSE, fig.width=5, fig.height=5}

cur_mat_final <- data.frame()

for (x in 1:length(sce_list)) {
  
  cur_mat <- df[[x]]
  cur_mat[,7] <- file_list[[x]]
  cur_mat_final <- rbind(cur_mat_final, as.data.frame(cur_mat))
  
}

ggplot(cur_mat_final) +
  geom_point(aes(log10(sum), subsets_Mito_percent), size = 0.3) +
  geom_abline(slope = 0, intercept = 10, linewidth = 2, color = "dark red") +
  theme_classic(base_size = 15)+
  facet_wrap(~V7)
```

We can now observe if these cells form their own cluster. We take T_BB022 as an example

```{r mito-UMAP, message=FALSE}
  
dittoDimPlot(sce_list[[4]], var = "subsets_Mito_percent") +
  scale_color_viridis()

```

We can now try to find an automatic threshold for mitochondrial reads.

```{r}

mito_thres <- list()

for (x in 1:length(sce_list)) {
  
mito_thres[[x]] <- perCellQCFilters(df[[x]], sub.fields="subsets_Mito_percent")

}

ggplot(as.data.frame(cbind(df[[4]], mito_thres[[4]]))) +
  geom_point(aes(log10(sum), subsets_Mito_percent, color = high_subsets_Mito_percent)) +
  theme_classic(base_size = 15)

```

We observe many cells with high (> 10%) mitochondrial counts.

```{r mitochondrial-percent}
  
cbind(df[[4]], mito_thres[[4]]) %>%
  as_tibble() %>%
  arrange(desc(subsets_Mito_percent)) %>%
  ggplot() +
  geom_point(aes((1:ncol(sce_list[[4]]))/ncol(sce_list[[4]]), subsets_Mito_percent, color = high_subsets_Mito_percent)) +
  geom_abline(intercept = 15, slope = 0, linewidth = 1, color = "dark red") +
  theme_classic(base_size = 15)

sum(df[[4]]$subsets_Mito_percent > 10)/ncol(sce_list[[4]])

sum(df[[4]]$subsets_Mito_percent > 15)/ncol(sce_list[[4]])

sum(mito_thres[[4]]$high_subsets_Mito_percent)/ncol(sce_list[[4]])

```

## Number of genes detected

Next, we will observe the number of genes detected.

```{r}
dittoDimPlot(sce_list[[4]], var = "detected") +
  scale_color_viridis()

ggplot(as.data.frame(cbind(colData(sce_list[[4]]), reducedDim(sce_list[[4]], "UMAP")))) +
  geom_point(aes(V1, V2, color = detected > 1000))
```

And plot them against the library size.

```{r}

ggplot(as.data.frame(df[[4]])) +
  geom_point(aes(log10(sum), detected)) +
  geom_abline(intercept = 1000, slope = 0) +
  theme_classic(base_size = 15)
  
```

## Doublet detection

Now, we will detect doublets by simulating artificial doublets.

```{r}
set.seed(221215)

dbl_dens <- list()

for (x in 1:length(sce_list)) {

dbl_dens[[x]] <- computeDoubletDensity(sce_list[[x]], subset.row=hvg[[x]])

sce_list[[x]]$DoubletScore <- dbl_dens[[x]]

}

dittoDimPlot(sce_list[[4]], var = "DoubletScore") +
  scale_color_viridis()

```

## Cluster detection and DE

Finally, we will detect clusters prior to excluding cells and perform 
differential expression testing. Here, we perform a rather braod clustering.

```{r clustering}
set.seed(221215) 

nn_clusters <- list()

for (x in 1:length(sce_list)) {

sce_2[[x]] <- fixedPCA(sce_2[[x]], subset.row=hvg[[x]])

nn_clusters[[x]] <- clusterCells(sce_2[[x]], use.dimred="PCA", BLUSPARAM = NNGraphParam(k = 20))

sce_2[[x]]$clusters <- nn_clusters[[x]]

}

dittoDimPlot(sce_2[[4]], var = "clusters")

```

Marker detection.

```{r marker-detection}

markers <- list()

for (x in 1:length(sce_list)) {
  
markers[[x]] <- scoreMarkers(sce_2[[x]], sce_2[[x]]$clusters)

}
```

## DE genes {.tabset}

### T_BB056org

```{r}
DT::datatable(head(as.data.frame(
  markers[[8]]$`1`[order(markers[[8]]$`1`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`2`[order(markers[[8]]$`2`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`3`[order(markers[[8]]$`3`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`4`[order(markers[[8]]$`4`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`5`[order(markers[[8]]$`5`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`6`[order(markers[[8]]$`6`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`7`[order(markers[[8]]$`7`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`8`[order(markers[[8]]$`8`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`9`[order(markers[[8]]$`9`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`10`[order(markers[[8]]$`10`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`11`[order(markers[[8]]$`11`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`12`[order(markers[[8]]$`12`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`13`[order(markers[[8]]$`13`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`14`[order(markers[[8]]$`14`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`15`[order(markers[[8]]$`15`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`16`[order(markers[[8]]$`16`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`17`[order(markers[[8]]$`17`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`18`[order(markers[[8]]$`18`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[8]]$`19`[order(markers[[8]]$`19`$mean.AUC, decreasing=TRUE),]), 100))

```

### T_BB171org

```{r}
DT::datatable(head(as.data.frame(
  markers[[11]]$`1`[order(markers[[11]]$`1`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`2`[order(markers[[11]]$`2`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`3`[order(markers[[11]]$`3`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`4`[order(markers[[11]]$`4`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`5`[order(markers[[11]]$`5`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`6`[order(markers[[11]]$`6`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`7`[order(markers[[11]]$`7`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`8`[order(markers[[11]]$`8`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`9`[order(markers[[11]]$`9`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`10`[order(markers[[11]]$`10`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`11`[order(markers[[11]]$`11`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`12`[order(markers[[11]]$`12`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`13`[order(markers[[11]]$`13`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`14`[order(markers[[11]]$`14`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[11]]$`15`[order(markers[[11]]$`15`$mean.AUC, decreasing=TRUE),]), 100))

```

### T_BB225org

```{r}
DT::datatable(head(as.data.frame(
  markers[[12]]$`1`[order(markers[[12]]$`1`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[12]]$`2`[order(markers[[12]]$`2`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[12]]$`3`[order(markers[[12]]$`3`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[12]]$`4`[order(markers[[12]]$`4`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[12]]$`5`[order(markers[[12]]$`5`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[12]]$`6`[order(markers[[12]]$`6`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[12]]$`7`[order(markers[[12]]$`7`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[12]]$`8`[order(markers[[12]]$`8`$mean.AUC, decreasing=TRUE),]), 100))

```

### T_BB375org

```{r}
DT::datatable(head(as.data.frame(
  markers[[13]]$`1`[order(markers[[13]]$`1`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`2`[order(markers[[13]]$`2`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`3`[order(markers[[13]]$`3`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`4`[order(markers[[13]]$`4`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`5`[order(markers[[13]]$`5`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`6`[order(markers[[13]]$`6`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`7`[order(markers[[13]]$`7`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`8`[order(markers[[13]]$`8`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`9`[order(markers[[13]]$`9`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`10`[order(markers[[13]]$`10`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`11`[order(markers[[13]]$`11`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`12`[order(markers[[13]]$`12`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`13`[order(markers[[13]]$`13`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`14`[order(markers[[13]]$`14`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`15`[order(markers[[13]]$`15`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`16`[order(markers[[13]]$`16`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`17`[order(markers[[13]]$`17`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`18`[order(markers[[13]]$`18`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[13]]$`19`[order(markers[[13]]$`19`$mean.AUC, decreasing=TRUE),]), 100))

```

### T_BB420org

```{r}
DT::datatable(head(as.data.frame(
  markers[[17]]$`1`[order(markers[[17]]$`1`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`2`[order(markers[[17]]$`2`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`3`[order(markers[[17]]$`3`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`4`[order(markers[[17]]$`4`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`5`[order(markers[[17]]$`5`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`6`[order(markers[[17]]$`6`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`7`[order(markers[[17]]$`7`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`8`[order(markers[[17]]$`8`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`9`[order(markers[[17]]$`9`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`10`[order(markers[[17]]$`10`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`11`[order(markers[[17]]$`11`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`12`[order(markers[[17]]$`12`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[17]]$`13`[order(markers[[17]]$`13`$mean.AUC, decreasing=TRUE),]), 100))

```

### T_BB475org

```{r}
DT::datatable(head(as.data.frame(
  markers[[20]]$`1`[order(markers[[20]]$`1`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`2`[order(markers[[20]]$`2`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`3`[order(markers[[20]]$`3`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`4`[order(markers[[20]]$`4`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`5`[order(markers[[20]]$`5`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`6`[order(markers[[20]]$`6`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`7`[order(markers[[20]]$`7`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`8`[order(markers[[20]]$`8`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`9`[order(markers[[20]]$`9`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`10`[order(markers[[20]]$`10`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`11`[order(markers[[20]]$`11`$mean.AUC, decreasing=TRUE),]), 100))

DT::datatable(head(as.data.frame(
  markers[[20]]$`12`[order(markers[[20]]$`12`$mean.AUC, decreasing=TRUE),]), 100))

```

## Exclude cells

After observing the different metrics we will exclude doublets and dying cells independently for each patient.
When a cluster contains doublets but also a substantial amount of single cells, we will keep it and exclude doublets later with the DoubletScore.

```{r}
sce_list[[1]] <- sce_list[[1]][,!as.character(sce_2[[1]]$clusters) %in% c("5", "6")]
sce_list[[2]] <- sce_list[[2]][,!as.character(sce_2[[2]]$clusters) %in% c("11","22")]
sce_list[[3]] <- sce_list[[3]][,!as.character(sce_2[[3]]$clusters) %in% c("7", "29")]
sce_list[[4]] <- sce_list[[4]][,!as.character(sce_2[[4]]$clusters) %in% c("9", "11")]
sce_list[[5]] <- sce_list[[5]][,!as.character(sce_2[[5]]$clusters) %in% c("11")]
sce_list[[6]] <- sce_list[[6]][,!as.character(sce_2[[6]]$clusters) %in% c("10")]
sce_list[[7]] <- sce_list[[7]][,!as.character(sce_2[[7]]$clusters) %in% c("14")]
sce_list[[8]] <- sce_list[[8]][,!as.character(sce_2[[8]]$clusters) %in% c("13", "16")]
sce_list[[9]] <- sce_list[[9]][,!as.character(sce_2[[9]]$clusters) %in% c("20" ,"14", "16", "24")]
sce_list[[10]] <- sce_list[[10]][,!as.character(sce_2[[10]]$clusters) %in% c("8", "15")]
sce_list[[11]] <- sce_list[[11]][,!as.character(sce_2[[11]]$clusters) %in% c("11", "13")]
sce_list[[12]] <- sce_list[[12]][,!as.character(sce_2[[12]]$clusters) %in% c("6", "8")]
sce_list[[13]] <- sce_list[[13]][,!as.character(sce_2[[13]]$clusters) %in% c("4", "18", "19")]
sce_list[[14]] <- sce_list[[14]][,!as.character(sce_2[[14]]$clusters) %in% c("7", "14")]
sce_list[[15]] <- sce_list[[15]][,!as.character(sce_2[[15]]$clusters) %in% c("1", "12", "13")]
sce_list[[16]] <- sce_list[[16]][,!as.character(sce_2[[16]]$clusters) %in% c("10", "11")]
sce_list[[17]] <- sce_list[[17]][,!as.character(sce_2[[17]]$clusters) %in% c("3", "6", "7", "9")]
sce_list[[18]] <- sce_list[[18]][,!as.character(sce_2[[18]]$clusters) %in% c("2")]
sce_list[[19]] <- sce_list[[19]][,!as.character(sce_2[[19]]$clusters) %in% c("8")]
sce_list[[20]] <- sce_list[[20]][,!as.character(sce_2[[20]]$clusters) %in% c("1", "10")]

```

Remove doublets that were not excluded in the previous step.

```{r}
for (x in 1:length(sce_list)) {

sce_list[[x]] <- sce_list[[x]][,sce_list[[x]]$DoubletScore < 10]

}

```

We will also exclude cells with high mitochondrial counts.

```{r}

for (x in 1:length(sce_list)) {
  
sce_list[[x]] <- sce_list[[x]][,sce_list[[x]]$subsets_Mito_percent < 20]

}
```

And exclude cells with few detected genes.

```{r}

for (x in 1:length(sce_list)) {
  
sce_list[[x]] <- sce_list[[x]][,sce_list[[x]]$detected > 1000]

}
```

## Save object

```{r save-sce}
saveRDS(sce_list, "~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/scripts/20230301_scOrganoidseq_HD/Robjects/sce_list_preprocessed.rds")
```

# 3. DE

```{r}
sce_pp <- readRDS("~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/scripts/20230301_scOrganoidseq_HD/Robjects/sce_list_preprocessed.rds")
```

## Normalization and HVG detection

In the first instance we will normalize the data and detect HVG again.

```{r normalization, message=FALSE, warning=FALSE}
set.seed(221216)

cur_clust <- list()

for (x in 1:length(sce_pp)) {
  
cur_clust[[x]] <- quickCluster(sce_pp[[x]]) 

sce_pp[[x]] <- computeSumFactors(sce_pp[[x]], cluster=cur_clust[[x]])

sce_pp[[x]] <- logNormCounts(sce_pp[[x]])

}

```

## Detection of highly variable genes.

```{r HVG}
# Fit the mean variance trend

cur_var <- list()
cur_fit <- list()

for (x in 1:length(sce_pp)) {

cur_var[[x]] <- modelGeneVar(sce_pp[[x]])

cur_fit[[x]] <- metadata(cur_var[[x]])

}

#Example with T_BB022

plot(cur_fit[[4]]$mean, cur_fit[[4]]$var, xlab="Mean of log-expression",
    ylab="Variance of log-expression")
curve(cur_fit[[4]]$trend(x), col="dodgerblue", add=TRUE, lwd=2)

# Order genes by their biological variance

hvg <- list()

for (x in 1:length(sce_pp)) {
  
cur_var[[x]][order(cur_var[[x]]$bio, decreasing=TRUE),] %>%
  as.data.frame() %>%
  head(50)

hvg[[x]] <- getTopHVGs(cur_var[[x]], n=1000)

}
```

## Broad clustering

We will now broadly cluster the cells.

```{r broad-clustering}

nn_clusters <- list()

for (x in 1:length(sce_pp)) {
  
set.seed(221216)
sce_pp[[x]] <- fixedPCA(sce_pp[[x]], subset.row=hvg[[x]])

sce_pp[[x]] <- runUMAP(sce_pp[[x]], dimred = "PCA")

nn_clusters[[x]] <- clusterCells(sce_pp[[x]], use.dimred="PCA", 
                            BLUSPARAM = NNGraphParam(k = 20))

sce_pp[[x]]$broad_clusters <- nn_clusters[[x]]

}

#Example with T_BB022

dittoDimPlot(sce_pp[[4]], var = "broad_clusters") 
```

### Cell type annotation

We can now annotate the clusters based on marker gene expression for every sce object individually.
These clusters will be named 'broad clusters'

## T_BB056

```{r}

dittoHeatmap(sce_pp[[8]], genes = genes, annot.by = "broad_clusters", 
             scaled.to.max = TRUE)

celltypes8 <- as.character(sce_pp[[8]]$broad_clusters)
celltypes8[celltypes8 == "1"] <- "basal"
celltypes8[celltypes8 == "2"] <- "mature_luminal"
celltypes8[celltypes8 == "3"] <- "basal"
celltypes8[celltypes8 == "4"] <- "mature_luminal"
celltypes8[celltypes8 == "5"] <- "luminal_progenitor"
celltypes8[celltypes8 == "6"] <- "basal"
celltypes8[celltypes8 == "7"] <- "basal"
celltypes8[celltypes8 == "8"] <- "basal"
celltypes8[celltypes8 == "9"] <- "basal"
celltypes8[celltypes8 == "10"] <- "basal"
celltypes8[celltypes8 == "11"] <- "basal"
celltypes8[celltypes8 == "12"] <- "basal"

sce_pp[[8]]$celltypes <- celltypes8

```

## T_BB171

```{r}

dittoHeatmap(sce_pp[[11]], genes = genes, annot.by = "broad_clusters", 
             scaled.to.max = TRUE)

celltypes11 <- as.character(sce_pp[[11]]$broad_clusters)
celltypes11[celltypes11 == "1"] <- "mature_luminal"
celltypes11[celltypes11 == "2"] <- "basal"
celltypes11[celltypes11 == "3"] <- "luminal_progenitor"
celltypes11[celltypes11 == "4"] <- "luminal_progenitor"
celltypes11[celltypes11 == "5"] <- "basal"
celltypes11[celltypes11 == "6"] <- "mature_luminal"
celltypes11[celltypes11 == "7"] <- "mature_luminal"
celltypes11[celltypes11 == "8"] <- "mature_luminal"
celltypes11[celltypes11 == "9"] <- "mature_luminal"

sce_pp[[11]]$celltypes <- celltypes11

```

## T_BB225

```{r}

dittoHeatmap(sce_pp[[12]], genes = genes, annot.by = "broad_clusters", 
             scaled.to.max = TRUE)

celltypes12 <- as.character(sce_pp[[12]]$broad_clusters)
celltypes12[celltypes12 == "1"] <- "luminal_progenitor"
celltypes12[celltypes12 == "2"] <- "basal"
celltypes12[celltypes12 == "3"] <- "mature_luminal"
celltypes12[celltypes12 == "4"] <- "luminal_progenitor"
celltypes12[celltypes12 == "5"] <- "basal"
celltypes12[celltypes12 == "6"] <- "mature_luminal"
celltypes12[celltypes12 == "7"] <- "basal"
celltypes12[celltypes12 == "8"] <- "mature_luminal"
celltypes12[celltypes12 == "9"] <- "mature_luminal"
celltypes12[celltypes12 == "10"] <- "luminal_progenitor"
celltypes12[celltypes12 == "11"] <- "basal"
celltypes12[celltypes12 == "12"] <- "luminal_progenitor"
celltypes12[celltypes12 == "13"] <- "mature_luminal"


sce_pp[[12]]$celltypes <- celltypes12

```

## T_BB375

```{r}

dittoHeatmap(sce_pp[[13]], genes = genes, annot.by = "broad_clusters", 
             scaled.to.max = TRUE)

celltypes13 <- as.character(sce_pp[[13]]$broad_clusters)
celltypes13[celltypes13 == "1"] <- "mature_luminal"
celltypes13[celltypes13 == "2"] <- "mature_luminal"
celltypes13[celltypes13 == "3"] <- "basal"
celltypes13[celltypes13 == "4"] <- "basal"
celltypes13[celltypes13 == "5"] <- "mature_luminal"
celltypes13[celltypes13 == "6"] <- "mature_luminal"
celltypes13[celltypes13 == "7"] <- "mature_luminal"
celltypes13[celltypes13 == "8"] <- "mature_luminal"
celltypes13[celltypes13 == "9"] <- "mature_luminal"
celltypes13[celltypes13 == "10"] <- "basal"
celltypes13[celltypes13 == "11"] <- "mature_luminal"
celltypes13[celltypes13 == "12"] <- "basal"
celltypes13[celltypes13 == "13"] <- "mature_luminal"
celltypes13[celltypes13 == "14"] <- "mature_luminal"
celltypes13[celltypes13 == "15"] <- "mature_luminal"
celltypes13[celltypes13 == "16"] <- "mature_luminal"
celltypes13[celltypes13 == "17"] <- "mature_luminal"
celltypes13[celltypes13 == "18"] <- "mature_luminal"
celltypes13[celltypes13 == "19"] <- "mature_luminal"
celltypes13[celltypes13 == "20"] <- "mature_luminal"
celltypes13[celltypes13 == "21"] <- "mature_luminal"
celltypes13[celltypes13 == "22"] <- "basal"
celltypes13[celltypes13 == "23"] <- "mature_luminal"
celltypes13[celltypes13 == "24"] <- "mature_luminal"
celltypes13[celltypes13 == "25"] <- "mature_luminal"

sce_pp[[13]]$celltypes <- celltypes13

```

## T_BB420

```{r}

dittoHeatmap(sce_pp[[17]], genes = genes, annot.by = "broad_clusters", 
             scaled.to.max = TRUE)

celltypes17 <- as.character(sce_pp[[17]]$broad_clusters)
celltypes17[celltypes17 == "1"] <- "mature_luminal"
celltypes17[celltypes17 == "2"] <- "mature_luminal"
celltypes17[celltypes17 == "3"] <- "mature_luminal"
celltypes17[celltypes17 == "4"] <- "mature_luminal"
celltypes17[celltypes17 == "5"] <- "mature_luminal"
celltypes17[celltypes17 == "6"] <- "mature_luminal"
celltypes17[celltypes17 == "7"] <- "mature_luminal"
celltypes17[celltypes17 == "8"] <- "mature_luminal"
celltypes17[celltypes17 == "9"] <- "mature_luminal"
celltypes17[celltypes17 == "10"] <- "mature_luminal"
celltypes17[celltypes17 == "11"] <- "mature_luminal"
celltypes17[celltypes17 == "12"] <- "mature_luminal"

sce_pp[[17]]$celltypes <- celltypes17

```

## T_BB475

```{r}

dittoHeatmap(sce_pp[[20]], genes = genes, annot.by = "broad_clusters", 
             scaled.to.max = TRUE)

celltypes20 <- as.character(sce_pp[[20]]$broad_clusters)
celltypes20[celltypes20 == "1"] <- "mature_luminal"
celltypes20[celltypes20 == "2"] <- "mature_luminal"
celltypes20[celltypes20 == "3"] <- "mature_luminal"
celltypes20[celltypes20 == "4"] <- "mature_luminal"
celltypes20[celltypes20 == "5"] <- "mature_luminal"
celltypes20[celltypes20 == "6"] <- "mature_luminal"
celltypes20[celltypes20 == "7"] <- "mature_luminal"
celltypes20[celltypes20 == "8"] <- "mature_luminal"
celltypes20[celltypes20 == "9"] <- "mature_luminal"
celltypes20[celltypes20 == "10"] <- "mature_luminal"
celltypes20[celltypes20 == "11"] <- "mature_luminal"
celltypes20[celltypes20 == "12"] <- "mature_luminal"
celltypes20[celltypes20 == "13"] <- "mature_luminal"
celltypes20[celltypes20 == "14"] <- "mature_luminal"
celltypes20[celltypes20 == "15"] <- "mature_luminal"

sce_pp[[20]]$celltypes <- celltypes20

```

## Save sce object

```{r}

saveRDS(sce_pp, "~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/scripts/20230301_scOrganoidseq_HD/Robjects/sce_list_annotated.rds")

```

# 4. Merge and batch correction

```{r}
sce_pp <- readRDS("~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/scripts/20230301_scOrganoidseq_HD/Robjects/sce_list_annotated.rds")
```

## Merge and correct for batch effect using fastMNN

```{r, message=FALSE, warning=FALSE}

sce_merged <- cbind(sce_pp[[1]], sce_pp[[2]], sce_pp[[3]], sce_pp[[4]], sce_pp[[5]], sce_pp[[6]], sce_pp[[7]], sce_pp[[8]], sce_pp[[9]], sce_pp[[10]], sce_pp[[11]], sce_pp[[12]], sce_pp[[13]], sce_pp[[14]], sce_pp[[15]], sce_pp[[16]], sce_pp[[17]], sce_pp[[18]], sce_pp[[19]], sce_pp[[20]])

set.seed(101)
f_out <- fastMNN(sce_pp[[1]], sce_pp[[2]], sce_pp[[3]], sce_pp[[4]], sce_pp[[5]], sce_pp[[6]], sce_pp[[7]], sce_pp[[8]], sce_pp[[9]], sce_pp[[10]], sce_pp[[11]], sce_pp[[12]], sce_pp[[13]], sce_pp[[14]], sce_pp[[15]], sce_pp[[16]], sce_pp[[17]], sce_pp[[18]], sce_pp[[19]], sce_pp[[20]])

#Store 'reconstructed' and 'corrected' and in the sce_merged object

assay(sce_merged, 'reconstructed') <- assay(f_out, 'reconstructed')
reducedDim(sce_merged, 'corrected') <- reducedDim(f_out, 'corrected')

```

## Include metadata in the new sce file

```{r}

sce_merged$Sample <- c(sce_pp[[1]]$Sample, sce_pp[[2]]$Sample, sce_pp[[3]]$Sample, sce_pp[[4]]$Sample, sce_pp[[5]]$Sample, sce_pp[[6]]$Sample, sce_pp[[7]]$Sample, sce_pp[[8]]$Sample, sce_pp[[9]]$Sample, sce_pp[[10]]$Sample, sce_pp[[11]]$Sample, sce_pp[[12]]$Sample, sce_pp[[13]]$Sample, sce_pp[[14]]$Sample, sce_pp[[15]]$Sample, sce_pp[[16]]$Sample, sce_pp[[17]]$Sample, sce_pp[[18]]$Sample, sce_pp[[19]]$Sample, sce_pp[[20]]$Sample)

sce_merged$barcode <- c(sce_pp[[1]]$barcode, sce_pp[[2]]$barcode, sce_pp[[3]]$barcode, sce_pp[[4]]$barcode, sce_pp[[5]]$barcode, sce_pp[[6]]$barcode, sce_pp[[7]]$barcode, sce_pp[[8]]$barcode, sce_pp[[9]]$barcode, sce_pp[[10]]$barcode, sce_pp[[11]]$barcode, sce_pp[[12]]$barcode, sce_pp[[13]]$barcode, sce_pp[[14]]$barcode, sce_pp[[15]]$barcode, sce_pp[[16]]$barcode, sce_pp[[17]]$barcode, sce_pp[[18]]$barcode, sce_pp[[19]]$barcode, sce_pp[[20]]$barcode)

sce_merged$sum <- c(sce_pp[[1]]$sum, sce_pp[[2]]$sum, sce_pp[[3]]$sum, sce_pp[[4]]$sum, sce_pp[[5]]$sum, sce_pp[[6]]$sum, sce_pp[[7]]$sum, sce_pp[[8]]$sum, sce_pp[[9]]$sum, sce_pp[[10]]$sum, sce_pp[[11]]$sum, sce_pp[[12]]$sum, sce_pp[[13]]$sum, sce_pp[[14]]$sum, sce_pp[[15]]$sum, sce_pp[[16]]$sum, sce_pp[[17]]$sum, sce_pp[[18]]$sum, sce_pp[[19]]$sum, sce_pp[[20]]$sum)

sce_merged$detected <- c(sce_pp[[1]]$detected, sce_pp[[2]]$detected, sce_pp[[3]]$detected, sce_pp[[4]]$detected, sce_pp[[5]]$detected, sce_pp[[6]]$detected, sce_pp[[7]]$detected, sce_pp[[8]]$detected, sce_pp[[9]]$detected, sce_pp[[10]]$detected, sce_pp[[11]]$detected, sce_pp[[12]]$detected, sce_pp[[13]]$detected, sce_pp[[14]]$detected, sce_pp[[15]]$detected, sce_pp[[16]]$detected, sce_pp[[17]]$detected, sce_pp[[18]]$detected, sce_pp[[19]]$detected, sce_pp[[20]]$detected)

sce_merged$subsets_Mito_sum <- c(sce_pp[[1]]$subsets_Mito_sum, sce_pp[[2]]$subsets_Mito_sum, sce_pp[[3]]$subsets_Mito_sum, sce_pp[[4]]$subsets_Mito_sum, sce_pp[[5]]$subsets_Mito_sum, sce_pp[[6]]$subsets_Mito_sum, sce_pp[[7]]$subsets_Mito_sum, sce_pp[[8]]$subsets_Mito_sum, sce_pp[[9]]$subsets_Mito_sum, sce_pp[[10]]$subsets_Mito_sum, sce_pp[[11]]$subsets_Mito_sum, sce_pp[[12]]$subsets_Mito_sum, sce_pp[[13]]$subsets_Mito_sum, sce_pp[[14]]$subsets_Mito_sum, sce_pp[[15]]$subsets_Mito_sum, sce_pp[[16]]$subsets_Mito_sum, sce_pp[[17]]$subsets_Mito_sum, sce_pp[[18]]$subsets_Mito_sum, sce_pp[[19]]$subsets_Mito_sum, sce_pp[[20]]$subsets_Mito_sum)

sce_merged$subsets_Mito_percent <- c(sce_pp[[1]]$subsets_Mito_percent, sce_pp[[2]]$subsets_Mito_percent, sce_pp[[3]]$subsets_Mito_percent, sce_pp[[4]]$subsets_Mito_percent, sce_pp[[5]]$subsets_Mito_percent, sce_pp[[6]]$subsets_Mito_percent, sce_pp[[7]]$subsets_Mito_percent, sce_pp[[8]]$subsets_Mito_percent, sce_pp[[9]]$subsets_Mito_percent, sce_pp[[10]]$subsets_Mito_percent, sce_pp[[11]]$subsets_Mito_percent, sce_pp[[12]]$subsets_Mito_percent, sce_pp[[13]]$subsets_Mito_percent, sce_pp[[14]]$subsets_Mito_percent, sce_pp[[15]]$subsets_Mito_percent, sce_pp[[16]]$subsets_Mito_percent, sce_pp[[17]]$subsets_Mito_percent, sce_pp[[18]]$subsets_Mito_percent, sce_pp[[19]]$subsets_Mito_percent, sce_pp[[20]]$subsets_Mito_percent)


sce_merged$subsets_Mito_detected <- c(sce_pp[[1]]$subsets_Mito_detected, sce_pp[[2]]$subsets_Mito_detected, sce_pp[[3]]$subsets_Mito_detected, sce_pp[[4]]$subsets_Mito_detected, sce_pp[[5]]$subsets_Mito_detected, sce_pp[[6]]$subsets_Mito_detected, sce_pp[[7]]$subsets_Mito_detected, sce_pp[[8]]$subsets_Mito_detected, sce_pp[[9]]$subsets_Mito_detected, sce_pp[[10]]$subsets_Mito_detected, sce_pp[[11]]$subsets_Mito_detected, sce_pp[[12]]$subsets_Mito_detected, sce_pp[[13]]$subsets_Mito_detected, sce_pp[[14]]$subsets_Mito_detected, sce_pp[[15]]$subsets_Mito_detected, sce_pp[[16]]$subsets_Mito_detected, sce_pp[[17]]$subsets_Mito_detected, sce_pp[[18]]$subsets_Mito_detected, sce_pp[[19]]$subsets_Mito_detected, sce_pp[[20]]$subsets_Mito_detected)

sce_merged$total <- c(sce_pp[[1]]$total, sce_pp[[2]]$total, sce_pp[[3]]$total, sce_pp[[4]]$total, sce_pp[[5]]$total, sce_pp[[6]]$total, sce_pp[[7]]$total, sce_pp[[8]]$total, sce_pp[[9]]$total, sce_pp[[10]]$total, sce_pp[[11]]$total, sce_pp[[12]]$total, sce_pp[[13]]$total, sce_pp[[14]]$total, sce_pp[[15]]$total, sce_pp[[16]]$total, sce_pp[[17]]$total, sce_pp[[18]]$total, sce_pp[[19]]$total, sce_pp[[20]]$total)

sce_merged$DoubletScore <- c(sce_pp[[1]]$DoubletScore, sce_pp[[2]]$DoubletScore, sce_pp[[3]]$DoubletScore, sce_pp[[4]]$DoubletScore, sce_pp[[5]]$DoubletScore, sce_pp[[6]]$DoubletScore, sce_pp[[7]]$DoubletScore, sce_pp[[8]]$DoubletScore, sce_pp[[9]]$DoubletScore, sce_pp[[10]]$DoubletScore, sce_pp[[11]]$DoubletScore, sce_pp[[12]]$DoubletScore, sce_pp[[13]]$DoubletScore, sce_pp[[14]]$DoubletScore, sce_pp[[15]]$DoubletScore, sce_pp[[16]]$DoubletScore, sce_pp[[17]]$DoubletScore, sce_pp[[18]]$DoubletScore, sce_pp[[19]]$DoubletScore, sce_pp[[20]]$DoubletScore)

sce_merged$sizeFactor <- c(sce_pp[[1]]$sizeFactor, sce_pp[[2]]$sizeFactor, sce_pp[[3]]$sizeFactor, sce_pp[[4]]$sizeFactor, sce_pp[[5]]$sizeFactor, sce_pp[[6]]$sizeFactor, sce_pp[[7]]$sizeFactor, sce_pp[[8]]$sizeFactor, sce_pp[[9]]$sizeFactor, sce_pp[[10]]$sizeFactor, sce_pp[[11]]$sizeFactor, sce_pp[[12]]$sizeFactor, sce_pp[[13]]$sizeFactor, sce_pp[[14]]$sizeFactor, sce_pp[[15]]$sizeFactor, sce_pp[[16]]$sizeFactor, sce_pp[[17]]$sizeFactor, sce_pp[[18]]$sizeFactor, sce_pp[[19]]$sizeFactor, sce_pp[[20]]$sizeFactor)

sce_merged$broad_clusters <- c(sce_pp[[1]]$broad_clusters, sce_pp[[2]]$broad_clusters, sce_pp[[3]]$broad_clusters, sce_pp[[4]]$broad_clusters, sce_pp[[5]]$broad_clusters, sce_pp[[6]]$broad_clusters, sce_pp[[7]]$broad_clusters, sce_pp[[8]]$broad_clusters, sce_pp[[9]]$broad_clusters, sce_pp[[10]]$broad_clusters, sce_pp[[11]]$broad_clusters, sce_pp[[12]]$broad_clusters, sce_pp[[13]]$broad_clusters, sce_pp[[14]]$broad_clusters, sce_pp[[15]]$broad_clusters, sce_pp[[16]]$broad_clusters, sce_pp[[17]]$broad_clusters, sce_pp[[18]]$broad_clusters, sce_pp[[19]]$broad_clusters, sce_pp[[20]]$broad_clusters)

sce_merged$celltypes <- c(sce_pp[[1]]$celltypes, sce_pp[[2]]$celltypes, sce_pp[[3]]$celltypes, sce_pp[[4]]$celltypes, sce_pp[[5]]$celltypes, sce_pp[[6]]$celltypes, sce_pp[[7]]$celltypes, sce_pp[[8]]$celltypes, sce_pp[[9]]$celltypes, sce_pp[[10]]$celltypes, sce_pp[[11]]$celltypes, sce_pp[[12]]$celltypes, sce_pp[[13]]$celltypes, sce_pp[[14]]$celltypes, sce_pp[[15]]$celltypes, sce_pp[[16]]$celltypes, sce_pp[[17]]$celltypes, sce_pp[[18]]$celltypes, sce_pp[[19]]$celltypes, sce_pp[[20]]$celltypes)

sce_merged$Patient_ID <- substring(colnames(sce_merged), 1,7)

```

## Run and plot t-SNE to see batches

```{r}

set.seed(101)
sce_merged <- runTSNE(sce_merged, dimred="corrected")
plotTSNE(sce_merged, colour_by="Patient_ID")

```

## Run and plot PCA to see batches

```{r, message=FALSE, warning=FALSE}

set.seed(101)
sce_merged <- runPCA(sce_merged, dimred="corrected")
plotPCA(sce_merged, colour_by="Patient_ID")

```

## Run and plot UMAP to see batches

```{r}

set.seed(101)
sce_merged <- runUMAP(sce_merged, dimred="corrected")
plotUMAP(sce_merged, colour_by="Patient_ID")

```

## Save object

```{r}

saveRDS(sce_merged ,"~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/scripts/20230301_scOrganoidseq_HD/Robjects/sce_merged.rds")

```

# 5. Cluster annotate

```{r}
sce_merged <- readRDS("~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/scripts/20230301_scOrganoidseq_HD/Robjects/sce_merged.rds")
```

## Try different settings of graph-based clustering

```{r}
set.seed(101)
nn_clusters_20 <- clusterCells(sce_merged, use.dimred="corrected", 
                            BLUSPARAM = NNGraphParam(k = 20, cluster.fun = "louvain"))

set.seed(101)
nn_clusters_40 <- clusterCells(sce_merged, use.dimred="corrected", 
                            BLUSPARAM = NNGraphParam(k = 40, cluster.fun = "louvain"))

set.seed(101)
nn_clusters_60 <- clusterCells(sce_merged, use.dimred="corrected", 
                            BLUSPARAM = NNGraphParam(k = 60, cluster.fun = "louvain"))

set.seed(101)
nn_clusters_80 <- clusterCells(sce_merged, use.dimred="corrected", 
                            BLUSPARAM = NNGraphParam(k = 80, cluster.fun = "louvain"))

set.seed(101)
nn_clusters_100 <- clusterCells(sce_merged, use.dimred="corrected", 
                            BLUSPARAM = NNGraphParam(k = 100, cluster.fun = "louvain"))

set.seed(101)
nn_clusters_120 <- clusterCells(sce_merged, use.dimred="corrected", 
                            BLUSPARAM = NNGraphParam(k = 120, cluster.fun = "louvain"))

#Compare

pheatmap(table(nn_clusters_20, nn_clusters_40), cluster_rows = F, cluster_cols = F)
pheatmap(table(nn_clusters_60, nn_clusters_80), cluster_rows = F, cluster_cols = F)
pheatmap(table(nn_clusters_100, nn_clusters_120), cluster_rows = F, cluster_cols = F)
pheatmap(table(nn_clusters_20, nn_clusters_120), cluster_rows = F, cluster_cols = F)

#We choose to keep the clustering with k=20
sce_merged$nn_clusters <- nn_clusters_20

```

## Annotate cell types

```{r, fig.width=10, fig.height=20}
genes <- c("KRT8", "KRT18", "KRT19", "KRT5", "KRT14", "EHF", "ESR1", "PGR", "AR", "PTPRC", "ACTA2", "CD68", "PECAM1", "CD3D", "MS4A1", "NCAM1", "CD3E", "CD19", "COL1A1", "CD34", "FAP", "PDPN", "PDGFRA")

dittoHeatmap(sce_merged, genes = genes, annot.by = "nn_clusters", 
             scaled.to.max = TRUE)

#Erase previous patient-level annotation that was made during the quality check

sce_merged$celltypes <- NULL

celltypes <- as.character(sce_merged$nn_clusters)
celltypes[celltypes == "1"] <- "endothelial"
celltypes[celltypes == "2"] <- "B_cell"
celltypes[celltypes == "3"] <- "T_cell"
celltypes[celltypes == "4"] <- "macrophage"
celltypes[celltypes == "5"] <- "fibroblast"
celltypes[celltypes == "6"] <- "endothelial"
celltypes[celltypes == "7"] <- "mature_luminal"
celltypes[celltypes == "8"] <- "basal"
celltypes[celltypes == "9"] <- "basal"
celltypes[celltypes == "10"] <- "mature_luminal"
celltypes[celltypes == "11"] <- "basal"
celltypes[celltypes == "12"] <- "mature_luminal"
celltypes[celltypes == "13"] <- "basal"
celltypes[celltypes == "14"] <- "mature_luminal"
celltypes[celltypes == "15"] <- "mature_luminal"
celltypes[celltypes == "16"] <- "luminal_progenitor"
celltypes[celltypes == "17"] <- "mature_luminal"
celltypes[celltypes == "18"] <- "mature_luminal"
celltypes[celltypes == "19"] <- "mature_luminal"
celltypes[celltypes == "20"] <- "mature_luminal"
celltypes[celltypes == "21"] <- "mature_luminal"
celltypes[celltypes == "22"] <- "luminal_progenitor"

sce_merged$celltypes <- celltypes
```

## Plot UMAP with annotated cell types

```{r}
plotUMAP(sce_merged, colour_by="celltypes")
```

## Save object

```{r}
saveRDS(sce_merged ,"~/data/old_VM_scripts/HD_Organoids/20230301_scOrganoidseq/scripts/20230301_scOrganoidseq_HD/Robjects/sce_final.rds")
```

