---
title: "SCRNAseq_tumor_analysis"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: sce_PalWu_ALL.rds, sce_PalWu.rds, GeneSignatures.rds

# Validation

Data analysis scripts for data analysis in the validation cohorts. 

# Script: Breast cancer patient scRNAseq analysis

Here, we analyse scRNA-seq data from two breast cancer scRNAseq datasets:
  GSE161529 - Pal et al. (2021, EMBO Journal)
  GSE176078 - Wu et al. (2021, Nature Genetics)

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(Seurat)
library(glmGamPoi)
library(future)
library(tidyverse)
library(scater)
library(scran)
library(BiocParallel)
library(Matrix)
library(gridExtra)
library(ggpubr)
library(ComplexHeatmap)
library(viridis)
library(readxl)
library(ggrepel)
library(clusterProfiler)
library(enrichplot)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(mclust)
library(RColorBrewer)
library(wesanderson)
library(dittoSeq)
```

# 1. Load data and preprocessing

## Full dataset

### Wu et al.

```{r}
## Wu et al. dataset

# Load count matrix
counts <- readMM("/mnt/central_nas/tnbc_volume/TNBC_validation/GSE176078/count_matrix_sparse.mtx")

# Load genes and barcodes
genes <- read.table("/mnt/central_nas/tnbc_volume/TNBC_validation/GSE176078/count_matrix_genes.tsv", header = FALSE)
barcodes <- read.table("/mnt/central_nas/tnbc_volume/TNBC_validation/GSE176078/count_matrix_barcodes.tsv", header = FALSE)

# Assign row and column names
rownames(counts) <- genes$V1
colnames(counts) <- barcodes$V1

# Load metadata
metadata <- read.csv("/mnt/central_nas/tnbc_volume/TNBC_validation/GSE176078/metadata.csv", row.names = 1)

# Create Seurat
seurat_wu <- CreateSeuratObject(counts = counts, meta.data = metadata)
seurat_wu

# Preprocessing
seurat_wu <- subset(seurat_wu, subset = nFeature_RNA > 500 & percent.mito < 20 & celltype_major == "Cancer Epithelial" )
seurat_wu

# Identify and remove mitochondrial genes
mt_genes <- grep("^MT-", rownames(seurat_wu), value = TRUE)
seurat_wu <- seurat_wu[!rownames(seurat_wu) %in% mt_genes, ]

# Add PID identifier
seurat_wu$PID <- seurat_wu$orig.ident
seurat_wu$dataset <- "wu"

saveRDS(seurat_wu, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SeuratObject_wu_full.rds")

# Normalization 
seurat_wu_sct <- SCTransform(seurat_wu, assay = "RNA", verbose = TRUE, return.only.var.genes = FALSE)
```

### Pal et al. 

```{r}
seurat_pal <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SeuratObject_pal.rds")

seurat_pal

# Add identifiers
seurat_pal$PID <- seurat_pal$group
seurat_pal$dataset <- "pal"
```

```{r}
saveRDS(seurat_pal, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SeuratObject_pal_full.rds")
```

```{r}
# Normalization
seurat_pal_sct <- SCTransform(seurat_pal, assay = "RNA", verbose = TRUE, return.only.var.genes = FALSE)
```

# 2. Seurat integration

```{r}
seurat_pal <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SeuratObject_pal_full.rds")
seurat_wu <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SeuratObject_wu_full.rds")

# Normalization
seurat_pal_sct <- SCTransform(seurat_pal, assay = "RNA", verbose = TRUE, return.only.var.genes = FALSE)
seurat_wu_sct <- SCTransform(seurat_wu, assay = "RNA", verbose = TRUE, return.only.var.genes = FALSE)

# Subsample for full dataset (100.000 cells total)
seurat_wu_sct <- subset(seurat_wu_sct, cells = sample(Cells(seurat_wu_sct), size = 24489))
seurat_pal_sct <- subset(seurat_pal_sct, cells = sample(Cells(seurat_pal_sct), size = 75511))

# Integration 
seurat_list <- list(seurat_pal_sct, seurat_wu_sct)
seurat_list <- lapply(X = seurat_list, FUN = function(x) {
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

features <- SelectIntegrationFeatures(object.list = seurat_list)

seurat_list <- PrepSCTIntegration(object.list = seurat_list,
                                  assay = "SCT",
                                  anchor.features = features)

integ_anchors <- FindIntegrationAnchors(object.list = seurat_list,
                                        normalization.method = "SCT",
                                        anchor.features = features,
                                        verbose = TRUE)

seurat_integrated <- IntegrateData(anchorset = integ_anchors,
                                   normalization.method = "SCT")

DefaultAssay(seurat_integrated) <- "integrated"

# Save integrated seurat object
saveRDS(seurat_integrated, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SeuratObject_palwu_sct_full.rds")
```

```{r Save}
# Save integrated seurat object
saveRDS(seurat_integrated, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SeuratObject_palwu_sct_full.rds")
#seurat_integrated <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SeuratObject_palwu_sct_full.rds")
```

## SCE conversion

```{r}
seurat_integrated

DefaultAssay(seurat_integrated) <- "integrated"
variable_features <- VariableFeatures(seurat_integrated)
integrated_matrix <- GetAssayData(seurat_integrated, assay = "integrated", slot = "data")

# Change to SCT assay and filter variable features
DefaultAssay(seurat_integrated) <- "SCT"
seurat_integrated@assays <- seurat_integrated@assays["SCT"]

sce <- as.SingleCellExperiment(seurat_integrated)
sce <- sce[variable_features,]
assay(sce, "integrated") <- integrated_matrix
sce$cell_id <- colnames(sce)
sce
```

```{r}
sce$TNBC <- ifelse(sce$orig.ident %in% c("TN") | sce$subtype %in% c("TNBC"), "TNBC", "Not_TNBC")

sce$TNBC <- ifelse(sce$orig.ident %in% c("TN") | sce$subtype %in% c("TNBC"), "TNBC", "Not_TNBC")

sce$subtype <- ifelse(sce$orig.ident %in% c("TN") | sce$subtype %in% c("TNBC"), "TNBC", 
                      ifelse(sce$orig.ident %in% c("ER") | sce$subtype %in% c("ER+"), "ER",
                             ifelse(sce$orig.ident %in% c("HER2") | sce$subtype %in% c("HER2+"), "HER2","Unknown")))
```

# 3. Dimensionality reduction

```{r}
set.seed(22)

#Use 30 PCs for denoising 
sce <- runPCA(sce, exprs_values = "integrated", ncomponents=30)

#Run TSNE on selected PCs
sce <- runTSNE(sce, dimred = "PCA", exprs_values = "integrated", external_neighbors=TRUE, BPPARAM = MulticoreParam())

# Visualize 
plotReducedDim(sce, dimred="TSNE", colour_by="PID")

plotReducedDim(sce, dimred="TSNE", colour_by="dataset", point_size = 1)+
  scale_color_manual(values = c("#1F78B4", "#B2DF8A"))+

plotReducedDim(sce, dimred="TSNE", colour_by="subtype", point_size = 1)+
  scale_color_manual(values = c("#4DAF4A", "#377EB8", "#E41A1C"))


plotReducedDim(sce, dimred = "TSNE", 
               colour_by = "KRT5", 
               by_exprs_values = "logcounts")+
plotReducedDim(sce, dimred = "TSNE", 
               colour_by = "KRT7", 
               by_exprs_values = "logcounts")

plotReducedDim(sce, dimred = "TSNE", 
               colour_by = "KRT8", 
               by_exprs_values = "logcounts")+
plotReducedDim(sce, dimred = "TSNE", 
               colour_by = "KRT19", 
               by_exprs_values = "logcounts")+
plotReducedDim(sce, dimred = "TSNE", 
               colour_by = "KRT18", 
               by_exprs_values = "logcounts")+
plotReducedDim(sce, dimred = "TSNE", 
               colour_by = "KRT14", 
               by_exprs_values = "logcounts")
```

# 4. CK5-7 signature

```{r}
set.seed(22)

## add CK5 and CK7 GMM
# 1. CK5
GMM_CK5 <- Mclust(assay(sce, "logcounts")["KRT5",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK5$classification == 2, "KRT5+", "KRT5-"), 
                     KRT5_exprs = assay(sce, "logcounts")["KRT5",],
                     uncertainty = GMM_CK5$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = KRT5_exprs, fill = class_epi))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")
  
#Add GMM epi class to SCE
sce$GMM_tumor_KRT5 <- GMM_df$class_epi

# 2. CK7
GMM_CK7 <- Mclust(assay(sce, "logcounts")["KRT7",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK7$classification == 2, "KRT7+", "KRT7-"), 
                     KRT7_exprs = assay(sce, "logcounts")["KRT7",],
                     uncertainty = GMM_CK7$uncertainty)

#Visualize GMM
ggplot(GMM_df, aes(x = KRT7_exprs, fill = class_epi))+
  geom_density()+
  theme_classic(base_size = 15)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")
  
#Add GMM epi class to SCE
sce$GMM_tumor_KRT7 <- GMM_df$class_epi

## Combine KRT5/7 GMM scores 
sce$KRT5_7_score <- ifelse(sce$GMM_tumor_KRT5 == "KRT5+" & sce$GMM_tumor_KRT7 == "KRT7+", "KRT5/7+", 
       ifelse(sce$GMM_tumor_KRT5 == "KRT5+" & sce$GMM_tumor_KRT7 == "KRT7-", "KRT5+",
              ifelse(sce$GMM_tumor_KRT5 == "KRT5-" & sce$GMM_tumor_KRT7 == "KRT7+", "KRT7+","KRT5/7-")))

metadata(sce)$colors$KRT57_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("KRT5/7-", "KRT5/7+","KRT5+", "KRT7+"))

## Visualization 
cur_dat <- t(assay(sce,"logcounts")) %>% as.matrix %>% as.data.frame %>% select("KRT5","KRT7")
cur_dat$KRT5_7_score <- sce$KRT5_7_score[match(rownames(cur_dat), sce$cell_id)]

## Scatterplot
cur_dat <- t(assay(sce,"logcounts")) %>% as.matrix %>% as.data.frame %>% dplyr::select("KRT5","KRT7")
cur_dat$KRT5_7_score <- sce$KRT5_7_score[match(rownames(cur_dat), rownames(colData(sce)))]
cur_dat$GMM_tumor_KRT7 <- sce$GMM_tumor_KRT7[match(rownames(cur_dat), rownames(colData(sce)))]
cur_dat$GMM_tumor_KRT5 <- sce$GMM_tumor_KRT5[match(rownames(cur_dat), rownames(colData(sce)))]


main <- ggplot(cur_dat, aes(x=KRT7, y=KRT5, color = KRT5_7_score)) +
  geom_point(alpha = 0.5)+
  theme_classic()+
  xlab("KRT7 logcounts")+
  ylab("KRT5 logcounts")+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce)$colors$KRT57_score)

y_side <- ggplot(cur_dat, aes(x = KRT5, fill=GMM_tumor_KRT5)) + 
  geom_density(alpha=0.8) + 
  scale_fill_manual(values = c('#999999','#F4B5BD')) + 
  theme_classic()+
  theme(legend.position = "none", axis.title.y = element_blank())+
  coord_flip()

x_side <- ggplot(cur_dat, aes(x = KRT7, fill=GMM_tumor_KRT7)) + 
  geom_density(alpha=0.8) + 
  scale_fill_manual(values = c('#999999','#9C964A')) + 
  theme_classic()+
  theme(legend.position = "none", axis.title.x = element_blank())

blankPlot <- ggplot()+geom_blank(aes(1,1)) + 
  cowplot::theme_nothing()

scatter_final <- grid.arrange(x_side, blankPlot, main, y_side, 
        ncol=2, nrow=2, widths=c(4, 1.4), heights=c(1.4, 4))


colData(sce) %>% as.data.frame %>% group_by(KRT5_7_score) %>% dplyr::count() %>% ungroup() %>% mutate(fra = n/sum(n))


# 2. Reduced Dims Plots
plotReducedDim(sce, dimred = "TSNE", 
               colour_by = "KRT5", 
               by_exprs_values = "logcounts", point_size = 1)+
plotReducedDim(sce, dimred = "TSNE", 
               colour_by = "KRT7", 
               by_exprs_values = "logcounts", point_size = 1)

plotReducedDim(sce, dimred="TSNE", colour_by="GMM_tumor_KRT5", point_size = 1)+
  scale_color_manual(values = c('#999999','#F4B5BD'))+

plotReducedDim(sce, dimred="TSNE", colour_by="GMM_tumor_KRT7", point_size = 1)+
  scale_color_manual(values = c('#999999','#9C964A'))

plotReducedDim(sce, dimred="TSNE", colour_by="KRT5_7_score", point_size = 1)+
  scale_color_manual(values = metadata(sce)$colors$KRT57_score)
```

```{r Save}
# Save sce object
#saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SCEObject_palwu_sct.rds")
#sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SCEObject_palwu_sct.rds")
```

## Barplot visualization

```{r}
dittoBarPlot(sce, var = "KRT5_7_score", group.by = "PID")+
  scale_fill_manual(values = metadata(sce)$colors$KRT57_score)+
  guides(fill = "none")
dittoBarPlot(sce, var = "KRT5_7_score", group.by = "PID", scale = "count")+
  scale_fill_manual(values = metadata(sce)$colors$KRT57_score)

dittoBarPlot(sce, var = "KRT5_7_score", group.by = "TNBC")+
  scale_fill_manual(values = metadata(sce)$colors$KRT57_score)+
  guides(fill = "none")
dittoBarPlot(sce, var = "KRT5_7_score", group.by = "TNBC", scale = "count")+
  scale_fill_manual(values = metadata(sce)$colors$KRT57_score)
```


```{r}
## PID KRT5_7_score fraction dataframe
PID_KRT5_7_score <- colData(sce) %>% as.data.frame() %>% dplyr::select(KRT5_7_score,PID) %>% group_by(PID) %>% table() %>% as.data.frame
PID_KRT5_7_score <- PID_KRT5_7_score %>% group_by(PID) %>% mutate(fra = Freq/sum(Freq))
cur_df <- PID_KRT5_7_score %>% pivot_wider(id_cols = "PID", names_from = "KRT5_7_score",values_from = "fra") %>% column_to_rownames("PID")

# Perform hierarchical clustering on fractions
dd <- dist(cur_df, method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
plot(hc)
row_sorted <- hc$order

# Define annotation dataframe
cur_df <- cur_df[row_sorted,]

anno <- colData(sce) %>% as.data.frame %>% dplyr::select(PID, TNBC, subtype, dataset) %>% unique()
anno <- anno[match(rownames(cur_df), anno$PID),]

n_cells_PID <- PID_KRT5_7_score %>% group_by(PID) %>% summarize(n_cells = sum(Freq))  
anno$ncells <- n_cells_PID$n_cells[match(anno$PID, n_cells_PID$PID)]

# Define colors 

sub_col <- setNames(rev(brewer.pal(3, "Set1")), unique(sce$subtype))
dataset_col <- setNames(brewer.pal(3, "Paired")[2:3], unique(sce$dataset))

# Plot annotation dataframe with ComplexHeatmap
fraction_anno <- HeatmapAnnotation(
  n_cells = anno_barplot(anno$ncells, gp=gpar(fill="#440154FF"),
                                   bar_width = 1,
                                   height = unit(3,"cm"), axis_param = list(direction = "reverse")),
  score_fractions = anno_barplot(cur_df, gp=gpar(fill=metadata(sce)$colors$KRT57_score[colnames(cur_df)]),
                                   bar_width = 1,
                                   height = unit(25,"cm"),
                                   width = unit(11,"cm"),
                                   show_row_names = FALSE), 
                          subtype = anno_simple(anno$subtype, col = sub_col, border = TRUE),
                          dataset = anno_simple(anno$dataset, col = dataset_col, border = TRUE),
                          border = TRUE, 
                          #annotation_name_rot = 90,
                          gap = unit(1,"mm"),
                          which = "row")

lgd <- Legend(title = "subtype", at = names(sub_col), 
               legend_gp = gpar(fill = sub_col))               
lgd_1 <- Legend(title = "dataset", at = names(dataset_col), 
               legend_gp = gpar(fill = dataset_col))  

draw(fraction_anno + NULL, annotation_legend_list = list(lgd, lgd_1))
```

## Keratin expression 

```{r}
# Use all marker expression based analysis, we use all features (not just the variable ones)
sce_all <- as.SingleCellExperiment(seurat_integrated)
sce_all$cell_id <- colnames(sce_all)
sce_all

sce_all$KRT5_7_score <- sce_PalWu$KRT5_7_score[match(sce_all$cell_id, sce_PalWu$cell_id)]
metadata(sce_all)$colors$KRT57_score <- metadata(sce_PalWu)$colors$KRT57_score
```

```{r}
keratins <- c("KRT5","KRT7","KRT15","KRT8","KRT18","KRT19", "KRT14","KRT17","ITGA6","EPCAM")

cluster_mean_sce_all <- aggregateAcrossCells(sce_all, ids = sce_all$KRT5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "logcounts",
                                 subset.row = rownames(sce_all) %in% keratins)

scaled_mat <- t(scale(t(assay(cluster_mean_sce_all, "logcounts"))))

anno <- colData(sce_all) %>% as.data.frame() %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% dplyr::count()
anno <- anno[match(anno$KRT5_7_score, rownames(t(scaled_mat))),]

## Print heatmap 
Heatmap(assay(cluster_mean_sce_all, "logcounts"), col = viridis(100),name = "Mean Expression")
Heatmap(t(scaled_mat), name = "Scaled mean expression")

# Add row annotations 
Heatmap(t(scaled_mat), name = "Scaled mean expression", clustering_method_columns = "ward.D2", clustering_method_rows = "ward.D2")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

## Stem marker expression 

```{r}
# Define stem markers of interest
stem <- c("SOX9","KIT","PROM1","YAP1","HDAC2","NOTCH1","CD44","ALDH1A3","BRCA1","BRCA2", "TGFB2", "MUC1","ANXA8", "CD24", "TNFAIP3", "ERBB3", "TNFRSF11A")

stem_rel <- stem[stem %in% rownames(sce_all)]

cluster_mean_sce_all <- aggregateAcrossCells(sce_all, ids = sce_all$KRT5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "logcounts",
                                 subset.row = rownames(sce_all) %in% stem_rel)

scaled_mat <- t(scale(t(assay(cluster_mean_sce_all, "logcounts"))))

anno <- colData(sce_all) %>% as.data.frame() %>% select(KRT5_7_score) %>% group_by(KRT5_7_score) %>% dplyr::count()
anno <- anno[match(anno$KRT5_7_score, rownames(t(scaled_mat))),]

## Print heatmap 
Heatmap(assay(cluster_mean_sce_all, "logcounts"), col = viridis(100),name = "Mean Expression")
Heatmap(scaled_mat, name = "Scaled mean expression")

# Add row annotations 
Heatmap(t(scaled_mat), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(anno$n, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(anno$KRT5_7_score))
```

## Gray et al signatures

A: Alveolar subtypes 

```{r}
AV_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Lietal_AV_BL_Signatures_reduced.xlsx",sheet = 1)

colnames(AV_sig) <- c("pan_AV", "BL_enriched","AP_enriched")

# Extract unique genes that were up- or down-regulated in each signature 
panAV_sig_up <- AV_sig$pan_AV
panAV_sig_up <- panAV_sig_up[!is.na(panAV_sig_up)]

BL_sig_up <- AV_sig$BL_enriched
BL_sig_up <- BL_sig_up[!is.na(BL_sig_up)]

AP_sig_up <- AV_sig$AP_enriched
AP_sig_up <- AP_sig_up[!is.na(AP_sig_up)]
```

```{r}
## 1. pan AV signature per PID group 
panAV_sig_up_rel <- panAV_sig_up[panAV_sig_up %in% rownames(assay(sce_all))]

stem_mean_sce_all <- aggregateAcrossCells(sce_all, ids = DataFrame(sce_all$KRT5_7_score,sce_all$PID), #Aggregate for score AND per line 
                                 statistics = "mean",
                                 use.assay.type = "logcounts",
                                 subset.row = rownames(sce_all) %in% panAV_sig_up_rel)

test <- as.data.frame(scale(t(assay(stem_mean_sce_all, "logcounts")), center = TRUE))

plot_df <- data.frame(signature_score = rowSums(test), cat = stem_mean_sce_all$sce_all.KRT5_7_score, PID = stem_mean_sce_all$sce_all.PID)

plot_df$cat <- factor(plot_df$cat, levels = c("KRT5/7-", "KRT7+", "KRT5+","KRT5/7+"))

#Boxplot 
ggplot(plot_df, aes(x = cat, y = signature_score))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = cat), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce_all)$colors$KRT57_score)+
  theme_classic()+
  ylab("Pan-Alveolar signature (Li et al.)")+
  stat_compare_means()

# Heatmap 
test$cat <- stem_mean_sce_all$sce_all.KRT5_7_score

test_out <- test %>% pivot_longer(cols = 1:41, names_to = "genes", values_to = "exprs") %>% group_by(cat,genes) %>% dplyr::summarize(sum = sum(exprs))

test_plot <- test_out %>% pivot_wider(names_from = genes, values_from = sum) %>% column_to_rownames("cat")

Heatmap(test_plot)

## Option 2: Aggregated by score and per single cell
stem_mean_sce_all <- assay(sce_all, "logcounts")[rownames(sce_all) %in% panAV_sig_up_rel,]

stem_mean_sce_all <- as.data.frame(as.matrix(stem_mean_sce_all))

test <- as.data.frame(scale(t(stem_mean_sce_all)), center = TRUE)

test$KRT5_7_score <- sce_all$KRT5_7_score[match(rownames(test), colnames(sce_all))]

plot_df <- data.frame(signature_score = rowSums(test[1:(ncol(test)-1)]), cat = test$KRT5_7_score) #PID = stem_mean_sce_all$sce_all.PID)

plot_df$cat <- factor(plot_df$cat, levels = c("KRT5/7-", "KRT7+", "KRT5+","KRT5/7+"))

#Boxplot 
ggplot(plot_df, aes(x = cat, y = signature_score))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = cat), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce_all)$colors$KRT57_score)+
  theme_classic()+
  ylab("Basal luminal signature (Li et al.)")+
  stat_compare_means()

#Heatmap
heat <- plot_df %>% group_by(cat) %>% summarise(mean = mean(signature_score)) %>% pivot_wider(names_from = cat, values_from = mean)
Heatmap(heat)
```

```{r}
## 2. BL signature per PID group 
## Option 1: Aggregated by score and PID 
BL_sig_up_rel <- BL_sig_up[BL_sig_up %in% rownames(assay(sce_all))]

stem_mean_sce_all <- aggregateAcrossCells(sce_all, ids = DataFrame(sce_all$KRT5_7_score, sce_all$PID), #Aggregate for score AND per PID
                                 statistics = "mean",
                                 use.assay.type = "logcounts",
                                 subset.row = rownames(sce_all) %in% BL_sig_up_rel)

test <- as.data.frame(scale(t(assay(stem_mean_sce_all))), center = TRUE)

plot_df <- data.frame(signature_score = rowSums(test), cat = stem_mean_sce_all$sce_all.KRT5_7_score, PID = stem_mean_sce_all$sce_all.PID)

plot_df$cat <- factor(plot_df$cat, levels = c("KRT5/7-", "KRT7+", "KRT5+","KRT5/7+"))

#Boxplot 
ggplot(plot_df, aes(x = cat, y = signature_score))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = cat), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce_all)$colors$KRT57_score)+
  theme_classic()+
  ylab("Basal luminal signature (Li et al.)")+
  stat_compare_means()

# Heatmap 
test$cat <- stem_mean_sce_all$sce_all.KRT5_7_score

test_out <- test %>% pivot_longer(cols = 1:ncol(test)-1, names_to = "genes", values_to = "exprs") %>% group_by(cat,genes) %>% dplyr::summarize(sum = sum(exprs))

test_plot <- test_out %>% pivot_wider(names_from = genes, values_from = sum) %>% column_to_rownames("cat")

Heatmap(test_plot)


## Option 2: Aggregated by score and per single cell
stem_mean_sce_all <- assay(sce_all, "logcounts")[rownames(sce_all) %in% BL_sig_up_rel,]

stem_mean_sce_all <- as.data.frame(as.matrix(stem_mean_sce_all))

test <- as.data.frame(scale(t(stem_mean_sce_all)), center = TRUE)

test$KRT5_7_score <- sce_all$KRT5_7_score[match(rownames(test), colnames(sce_all))]

plot_df <- data.frame(signature_score = rowSums(test[1:(ncol(test)-1)]), cat = test$KRT5_7_score) #PID = stem_mean_sce_all$sce_all.PID)

plot_df$cat <- factor(plot_df$cat, levels = c("KRT5/7-", "KRT7+", "KRT5+","KRT5/7+"))

#Boxplot 
ggplot(plot_df, aes(x = cat, y = signature_score))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = cat), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce_all)$colors$KRT57_score)+
  theme_classic()+
  ylab("Basal luminal signature (Li et al.)")+
  stat_compare_means()

#Heatmap
heat <- plot_df %>% group_by(cat) %>% summarise(mean = mean(signature_score)) %>% pivot_wider(names_from = cat, values_from = mean)
Heatmap(heat)
```

## Lim et al signatures

```{r}
lum_progenitor_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/LuminalProgenitor_Signature_Limetal_NatureMedicine_2009.xls")
mature_lum_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/MatureLuminalSignature_Limetal_NatureMedicine_2009.xls")
basal_sig <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Basal_Signature_Limetal_NatureMedicine_2009.xls")

# Extract unique genes that were up- or down-regulated in each signature 
lum_progenitor_sig_up <- lum_progenitor_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
lum_progenitor_sig_up <- lum_progenitor_sig_up[!is.na(lum_progenitor_sig_up)]
lum_progenitor_sig_down <- lum_progenitor_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
lum_progenitor_sig_down <- lum_progenitor_sig_down[!is.na(lum_progenitor_sig_down)]

mature_lum_sig_up <- mature_lum_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
mature_lum_sig_up <- mature_lum_sig_up[!is.na(mature_lum_sig_up)]
mature_lum_sig_down <- mature_lum_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
mature_lum_sig_down <- mature_lum_sig_down[!is.na(mature_lum_sig_down)]

basal_sig_up <- basal_sig %>% filter(`Average log fold-change` > 1) %>% pull(Symbol) %>% unique()
basal_sig_up <- basal_sig_up[!is.na(basal_sig_up)]
basal_sig_down <- basal_sig %>% filter(`Average log fold-change` < -1) %>% pull(Symbol) %>% unique()
basal_sig_down <- basal_sig_down[!is.na(basal_sig_down)]
```

```{r}
## 1. Luminal progenitor signature per PID group 
lum_progenitor_sig_up_rel <- lum_progenitor_sig_up[lum_progenitor_sig_up %in% rownames(assay(sce_all))]

stem_mean_sce_all <- aggregateAcrossCells(sce_all, ids = DataFrame(sce_all$KRT5_7_score,sce_all$PID), #Aggregate for score AND per PID
                                 statistics = "mean",
                                 use.assay.type = "logcounts",
                                 subset.row = rownames(sce_all) %in% lum_progenitor_sig_up_rel)

test <- as.data.frame(scale(t(assay(stem_mean_sce_all, "logcounts")), center = TRUE))

plot_df <- data.frame(signature_score = rowSums(test,na.rm = TRUE), cat = stem_mean_sce_all$sce_all.KRT5_7_score, PID = stem_mean_sce_all$sce_all.PID)

plot_df$cat <- factor(plot_df$cat, levels = c("KRT5/7-", "KRT7+", "KRT5+","KRT5/7+"))

#Boxplot 
ggplot(plot_df, aes(x = cat, y = signature_score))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = cat), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce_all)$colors$KRT57_score)+
  theme_classic()+
  ylab("Luminal progenitor signature (Lim et al.)")+
  stat_compare_means()

# Heatmap 
test$cat <- stem_mean_sce_all$sce_all.KRT5_7_score

test_out <- test %>% pivot_longer(cols = 1:ncol(test)-1, names_to = "genes", values_to = "exprs") %>% group_by(cat,genes) %>% dplyr::summarize(sum = sum(exprs))

test_plot <- test_out %>% pivot_wider(names_from = genes, values_from = sum) %>% column_to_rownames("cat")

Heatmap(test_plot)

## Option 2: Aggregated by score and per single cell
stem_mean_sce_all <- assay(sce_all, "logcounts")[rownames(sce_all) %in% lum_progenitor_sig_up_rel,]

stem_mean_sce_all <- as.data.frame(as.matrix(stem_mean_sce_all))

test <- as.data.frame(scale(t(stem_mean_sce_all)), center = TRUE)

test$KRT5_7_score <- sce_all$KRT5_7_score[match(rownames(test), colnames(sce_all))]

plot_df <- data.frame(signature_score = rowSums(test[1:(ncol(test)-1)], na.rm = TRUE), cat = test$KRT5_7_score) #PID = stem_mean_sce_all$sce_all.PID)

plot_df$cat <- factor(plot_df$cat, levels = c("KRT5/7-", "KRT7+", "KRT5+","KRT5/7+"))


#Boxplot 
ggplot(plot_df, aes(x = cat, y = signature_score))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(aes(color = cat), size = 3, position=position_jitter(w=0.1,h=0.1))+
  scale_color_manual(values = metadata(sce_all)$colors$KRT57_score)+
  theme_classic()+
  ylab("Basal luminal signature (Li et al.)")+
  stat_compare_means()

#Heatmap
heat <- plot_df %>% group_by(cat) %>% summarise(mean = mean(signature_score)) %>% pivot_wider(names_from = cat, values_from = mean)
Heatmap(heat)
```


# 6. Differential gene expression analysis

A) DEG between KRT5/7+ and all others 

```{r}
sce_all$KRT5_7_score_sum <- ifelse(sce_all$KRT5_7_score == "KRT5/7+", "KRT5/7+", "Not_KRT5/7+")
sce_all$KRT5_7_score_sum <- factor(sce_all$KRT5_7_score_sum, levels = c("Not_KRT5/7+","KRT5/7+"))
```

Create pseudobulk-profiles (sample-label combination)

```{r}
summed <- aggregateAcrossCells(sce_all, id=colData(sce_all)[,c("KRT5_7_score_sum", "PID")])
summed
```

```{r}
between.res <- pseudoBulkDGE(summed,
    label=rep("dummy", ncol(summed)),
    design=~factor(PID) + KRT5_7_score_sum,
    coef="KRT5_7_score_sumKRT5/7+")[[1]]

plot_df <- between.res %>% as.data.frame %>% arrange(PValue) %>% rownames_to_column(var = "gene") %>% mutate(sig = ifelse((FDR<0.05 & logFC > 1) | (FDR<0.05 & logFC < -1), TRUE, FALSE))

# Volcano plot
ggplot(plot_df, aes(x = logFC, y = -log10(FDR)))+
  geom_point(aes(col=sig, size = -log10(FDR)))+
  geom_label_repel(data = plot_df %>% filter((FDR<0.05 & logFC > 1) | (FDR<0.05 & logFC < -1)), aes(label = gene))+
  scale_color_brewer(palette = "Set1")+
  xlab("logFC (KRT5_7_score)")+
  ylab("-log10(FDR)")+
  theme_classic()+
  geom_hline(yintercept = -log10(0.05), linetype = "dashed")+
  geom_vline(xintercept = 1, linetype = "dashed")+
  geom_vline(xintercept = -1, linetype = "dashed")+
  labs(color = "cluster")
```

# 7. Gene set enrichment analysis 

With `clusterProfiler`. 

```{r}
organism <- "org.Hs.eg.db"

res <- between.res

# Prepare input gene list based on log2foldchanges 
original_gene_list <- res$logFC
names(original_gene_list) <- rownames(res)
gene_list <- na.omit(original_gene_list)
gene_list <- sort(gene_list, decreasing = TRUE) # sort in decreasing order

# Run GO analysis
gse <- gseGO(geneList=gene_list, 
             ont ="BP", 
             keyType = "SYMBOL", 
             minGSSize = 50, 
             maxGSSize = 1000, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = organism, 
             pAdjustMethod = "BH")

# Visualization
## Dot plot
dotplot(gse, showCategory=5, split=".sign") + facet_grid(~.sign)

## Emap plot
emapplot(pairwise_termsim(gse), showCategory = 10, color = "enrichmentScore", layout = "circle")

## GSEA plot
gseaplot2(gse, title = gse$Description[9], geneSetID = 1)
gseaplot2(gse, title = gse$Description[10], geneSetID = 1)
gseaplot2(gse, title = gse$Description[14], geneSetID = 1)
```

# Save objects

```{r}
#saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SCEObject_palwu_sct.rds")
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/SCEObject_palwu_sct_full.rds")
```

# Session Info 

```{r}
sessionInfo()
```