---
title: "TNBC_Metadata_Update"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: TNBC_discovery_metadata.csv

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Support Script: Clinical metadata processing

Here, we create updated metadata for TNBC cohorts and visualize cohort heatmaps.

# Software requirements 

```{r lib, message=FALSE,warning=FALSE}
library(tidyverse)
library(readxl)
library(knitr)
library(DT)
library(prodlim)
library(ComplexHeatmap)
library(foreach)
library(ggsci)
library(RColorBrewer)
```

# Combined - ZTMA174, ZTMA249

### Load all metadata

```{r}
Meta_174 <- read_csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/ZTMA174_TNBC_LM_processed.csv")
Meta_249 <- read_csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/ZTMA249_TNBC_LM_processed.csv")
#Meta_2126 <- read_csv("/Users/lasmey/Desktop/TNBC/Metadaten/ZTMA21_26_Jana_TNBC_LM_processed.csv")
```

### Combine metadata from all 

```{r combine}
#select columns of interest 
Meta_174_comb <- Meta_174 %>% select("TMANr","PID","n_samples","age","tumor_size_cm","grade","pT","pT_simple","pN","pN_simple","pM","pM_simple","status_OS","OS_months", "status_DFS","DFS_months","OS_data","DFS_data","HER2_case") %>% mutate_all(as.character)

Meta_249$TMANr <- "TMA249"
Meta_249$HER2_case <- FALSE
Meta_249_comb <- Meta_249 %>% select("TMANr","ID_Number","PID","n_samples","age","lehmann_subtype","grade","pT","pT_simple","pN","pN_simple","pM","pM_simple","status_OS","OS_months", "status_DFS","DFS_months","OS_data","DFS_data","HER2_case") %>% mutate_all(as.character)

#Meta_2126_comb <- Meta_2126 %>% select("TMANr","ID_Number","PID","n_samples","age","grade","pT","pT_simple","pN","pN_simple","pM","pM_simple","status_#OS","OS_months", "status_DFS","DFS_months","OS_data","DFS_data","HER2_case") %>% mutate_all(as.character)

Meta <- bind_rows(Meta_174_comb,Meta_249_comb) #Meta_2126_comb)

Meta$tumor_size_cm <- as.numeric(Meta$tumor_size_cm)
```

```{r}
#quality control 
##double-check that PID entries are distinct
dup <- Meta %>% 
  count(PID) %>% 
  filter(n>1) %>% 
  pull(PID)

dup #10 patients are duplicated between the two TMAs (combine metadata information?)

#dup_meta <- Meta %>% filter(PID %in% dup) %>% arrange(PID)
#write.csv(dup_meta, "/Users/lasmey/Desktop/TNBC/Metadaten/Metadata_Duplicated_TNBC_LM.csv")

#read in metadata updates for duplicated cases
#dup_meta_update <- read.csv("/Users/lasmey/Desktop/TNBC/Metadaten/Metadata_Duplicated_TNBC_LM_mod.csv",)

Meta <- Meta %>% filter(!PID %in% dup)

Meta <- rbind(Meta, dup_meta_update)

## Filter patients which are TNBC but metadata is missing 
Meta_meta_missing <- Meta %>% filter(HER2_case %in% c(NA,FALSE) & is.na(pT)) 
#write_csv(Meta_174_meta_missing,"/Users/lasmey/Desktop/TNBC/Metadaten/ZTMA174_MetadataMissing.csv")

PID_missing <- Meta_meta_missing$PID

#meta_add <- read.csv("/Volumes/rcc_volume/TNBC/data_analysis/03_data/Metadata_Combined_TNBC_LM.csv",row.names = 1)
meta_update <- meta_add %>% filter(PID %in% PID_missing)
meta_update$age_bin <- NULL

Meta <- Meta %>% filter(!PID %in% PID_missing)
Meta <- rbind(Meta, meta_update)


##Create age_bin 
summary(as.numeric(Meta$age))
Meta <- Meta %>% mutate(age_bin = ifelse(age <= 40, "27-40", 
                                   ifelse(40<age & age<=55, "41-55",
                                          ifelse(55<age & age<=70, "56-70",
                                                 ifelse(70<age & age<=85, "71-85",
                                                        ifelse(85<age & age<=93, "85-93",NA))))))
```

```{r save combined}
write.csv(Meta, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Metadata_Combined_TNBC_LM.csv")
```

```{r save combined}
Meta <- read.csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Metadata_Combined_TNBC_LM.csv")

Meta <- Meta %>% filter(!TMANr %in% c("TMA21","TMA26"))
```

### Print heatmap based overview
based on the [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html) package and the implemented `HeatmapAnnotation` function

```{r}
# Add neoadjuvant info 
Meta$pretreatment <- str_detect(Meta$pT, "y")

#Filter for columns to visualize
Meta_viz <- Meta %>% #filter(HER2_case == FALSE) %>% 
  select(TMANr,PID,n_samples,grade,pT_simple,pN_simple,pM_simple,OS_data, DFS_data, age_bin, tumor_size_cm, lehmann_subtype, pretreatment) 

#Convert empty columns and NAs for easier handling with ComplexHeatmap package
Meta_viz[is.na(Meta_viz)] <- "NA"

Meta_viz$tumor_size_cm <- as.numeric(Meta_viz$tumor_size_cm)

#Arrange
Meta_viz <- Meta_viz %>% arrange(desc(pT_simple),desc(pN_simple), desc(grade))

## Set Color scheme (also for downstream analysis)
color_vectors <- list(
TMANr = setNames(colorRampPalette(brewer.pal(3, "Dark2"))(length(unique(Meta_viz$TMANr))), 
                sort(unique(Meta_viz$TMANr))),
PID = setNames(colorRampPalette(brewer.pal(12, "Paired"))(length(unique(Meta_viz$PID))), 
                sort(unique(Meta_viz$PID))),
n_samples = setNames(colorRampPalette(brewer.pal(3, "Greens"))(length(unique(Meta_viz$n_samples))), 
                sort(unique(Meta_viz$n_samples))),
grade = setNames(c(colorRampPalette(brewer.pal(3, "Oranges"))(length(unique(Meta_viz$grade))-1),"black"), 
                sort(unique(Meta_viz$grade))),
pT_simple = setNames(c(colorRampPalette(brewer.pal(9, "Blues"))(length(unique(Meta_viz$pT_simple))-1),"black"), 
                sort(unique(Meta_viz$pT_simple), na.last = TRUE)),
pN_simple = setNames(c(colorRampPalette(brewer.pal(9, "Purples"))(length(unique(Meta_viz$pN_simple))-1),"black"), 
                sort(unique(Meta_viz$pN_simple))),
pM_simple = setNames(c(colorRampPalette(brewer.pal(3, "PuRd"))(length(unique(Meta_viz$pM_simple))-1),"black"), 
                sort(unique(Meta_viz$pM_simple))),
pretreatment = setNames(c(colorRampPalette(brewer.pal(4, "Set3")[3:4])(length(unique(Meta_viz$pretreatment))-1),"black"), 
                c("FALSE","TRUE","NA")),
OS_data = setNames(c("black","grey"),sort(unique(Meta_viz$OS_data))),
DFS_data = setNames(c("black","grey"),sort(unique(Meta_viz$DFS_data))),
age_bin = setNames(c(colorRampPalette(brewer.pal(11, "BrBG"))(length(unique(Meta_viz$age_bin))-1),"black"), 
                sort(unique(Meta_viz$age_bin))),
lehmann_subtype = setNames(c(colorRampPalette(brewer.pal(9, "Set1"))(length(unique(Meta_viz$lehmann_subtype))-1),"black"), 
                sort(unique(Meta_viz$lehmann_subtype)))
)



#Save color vector  object for downstream usage
#saveRDS(color_vectors,file="/Volumes/rcc_volume/TNBC/data_analysis/color_meta.rds")

#Heatmap Annotation
sample_anno <- HeatmapAnnotation(TMANr = Meta_viz$TMANr,
                                 PID = Meta_viz$PID,
                                 age_bin = Meta_viz$age_bin,
                                 n_samples = Meta_viz$n_samples,
                                 grade = Meta_viz$grade,
                                 pT_simple = Meta_viz$pT_simple,
                                pN_simple = Meta_viz$pN_simple,
                                pM_simple = Meta_viz$pM_simple,
                                pretreatment = Meta_viz$pretreatment,
                                OS_data = Meta_viz$OS_data,
                                DFS_data = Meta_viz$DFS_data,
                                #tumor_size_cm = Meta_viz$tumor_size_cm,
                                #lehmann_subtype = Meta_viz$lehmann_subtype,
                                col = color_vectors,
                                border = T,
                                na_col = "black",
                                gap = unit(5, "points"),
                                show_annotation_name = T,
                                show_legend = c(PID = FALSE)
)%v%NULL

#Display plot
draw(sample_anno)

sample_anno
```


