---
title: "00_seg_mesmer_QC"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: / 

# Discovery 

Data analysis scripts for IMC data analysis in the discovery cohort. 

# Support Script: Segmentation Quality Control

Here, we used the pre-trained deep-learning mesmer model to perform single-cell segmentation with different channel-combinations for testing.

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(CATALYST)
library(scuttle)
library(scater)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(viridis)
library(tidyverse)
library(openxlsx)
library(stringr)
library(BiocParallel)
```

## Read in images

Read in multi-channel images (TIFF) and segmentation masks into a
`CytoImageList` container using the `loadImages` function from the `cytomapper` package.

```{r read-images}
#images <- loadImages("/mnt/central_nas/tnbc_volume/TNBC/img/",on_disk = FALSE,BPPARAM = MulticoreParam(workers = 2,progressbar = TRUE,stop.on.error = FALSE))
#images <- loadImages("/mnt/central_nas/tnbc_volume/TNBC/img/",on_disk = TRUE,h5FilesPath = "/mnt/central_nas/tnbc_volume/TNBC/h5_img/",BPPARAM = MulticoreParam(workers = 14,progressbar = TRUE))
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_TNBC_spillcor.rds")

#set channel-names for multi-channel images
#double-check that the channel order is correct
panel <- read_csv("/mnt/central_nas/tnbc_volume/TNBC/panel.csv")
channelNames(images) <- panel$clean_target
images

#Load masks
#masks_1 <- loadImages("/mnt/central_nas/tnbc_volume/TNBC/masks_1/", as.is = TRUE)
#masks_2 <- loadImages("/mnt/central_nas/tnbc_volume/TNBC/masks_2/", as.is = TRUE)
#masks_3 <- loadImages("/mnt/central_nas/tnbc_volume/TNBC/masks_3/", as.is = TRUE)
#masks_4 <- loadImages("/mnt/central_nas/tnbc_volume/TNBC/masks_4/", as.is = TRUE)
#masks_5 <- loadImages("/mnt/central_nas/tnbc_volume/TNBC/masks_5/", as.is = TRUE)

masks_1 <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_1.rds")
masks_2 <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_2.rds")
masks_3 <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_3.rds")
masks_4 <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")
masks_5 <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_5.rds")
```

Add additional metadata to the `elementMetadata` slot of the `CytoImageList` (accessible via mcols)

```{r add-image-metadata}
all.equal(names(images), names(masks_1))

PID <- c(str_split(names(images), "_",simplify = TRUE)[1:180,3],paste0(str_split(names(images), "_",simplify = TRUE)[181:length(names(images)),3],str_split(names(images), "_",simplify = TRUE)[181:length(names(images)),4]))

mcols(images) <- mcols(masks_1) <- mcols(masks_2) <- mcols(masks_3) <- mcols(masks_4) <- mcols(masks_5) <- DataFrame(sample_id = names(images), PID = PID)
```

# Save images and masks

```{r save-objects-read-data}
saveRDS(images, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_segtest.rds")

saveRDS(masks_1, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_1.rds")
saveRDS(masks_2, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_2.rds")
saveRDS(masks_3, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_3.rds")
saveRDS(masks_4, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")
saveRDS(masks_5, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_5.rds")
```

## QC

### Image visualization

```{r overlay-masks, message=FALSE}
#Randomly select 20 images
set.seed(22)
img_random <- sample(seq_len(length(images)), 10)

img_files <- list.files("/mnt/central_nas/tnbc_volume/TNBC/img/")
img_selected <- c(which(str_detect(img_files ,"ZTMA249_ID_320_CX8Y10")),which(str_detect(img_files ,"ZTMA249_ID_289_CX1Y7")),which(str_detect(img_files ,"ZTMA249_ID_36_AX4Y5")),which(str_detect(img_files ,"ZTMA249_ID_7_AX7Y1")),which(str_detect(img_files ,"ZTMA249_ID_220_BX4Y13")),which(str_detect(img_files ,"ZTMA249_ID_271_CX7Y4")),which(str_detect(img_files ,"ZTMA249_ID_82_AX2Y11")), which(str_detect(img_files ,"ZTMA249_ID_179_BX3Y8")), which(str_detect(img_files ,"ZTMA249_ID_183_BX7Y8")), which(str_detect(img_files ,"ZTMA174_B08.26868_79_AX7Y10")))

img_ids <- c(img_selected,img_random)

# Normalize and clip image
cur_images <- images[img_ids]
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, inputRange = c(0, 0.2))
channelNames(cur_images) <- panel$clean_target

#by epithelial channels
plotPixels(cur_images,
           mask = masks_1[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "Vimentin", "SMA","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "Vimentin" = c("black", "cyan"),
                         "SMA" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_epi_1.png", scale = 2))

plotPixels(cur_images,
           mask = masks_2[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "Vimentin", "SMA","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "Vimentin" = c("black", "cyan"),
                         "SMA" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_epi_2.png", scale = 2))

plotPixels(cur_images,
           mask = masks_3[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "Vimentin", "SMA","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "Vimentin" = c("black", "cyan"),
                         "SMA" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_epi_3.png", scale = 2))

plotPixels(cur_images,
           mask = masks_4[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "Vimentin", "SMA","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "Vimentin" = c("black", "cyan"),
                         "SMA" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_epi_4.png", scale = 2))

plotPixels(cur_images,
           mask = masks_5[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("panCK", "Ecad", "Vimentin", "SMA","CD44"),
           colour = list("panCK" = c("black", "yellow"),
                         "Ecad" = c("black", "magenta"),
                         "Vimentin" = c("black", "cyan"),
                         "SMA" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_epi_5.png", scale = 2))


#by immune channels
plotPixels(cur_images,
           mask = masks_1[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD3", "CD8a", "CD68", "CD20", "CD44"),
           colour = list("CD3" = c("black", "yellow"),
                         "CD20" = c("black", "magenta"),
                         "CD8a" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_immune_1.png", scale = 2))

plotPixels(cur_images,
           mask = masks_2[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD3", "CD8a", "CD68", "CD20", "CD44"),
           colour = list("CD3" = c("black", "yellow"),
                         "CD20" = c("black", "magenta"),
                         "CD8a" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_immune_2.png", scale = 2))

plotPixels(cur_images,
           mask = masks_3[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD3", "CD8a", "CD68", "CD20", "CD44"),
           colour = list("CD3" = c("black", "yellow"),
                         "CD20" = c("black", "magenta"),
                         "CD8a" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_immune_3.png", scale = 2))

plotPixels(cur_images,
           mask = masks_4[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD3", "CD8a", "CD68", "CD20", "CD44"),
           colour = list("CD3" = c("black", "yellow"),
                         "CD20" = c("black", "magenta"),
                         "CD8a" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_immune_4.png", scale = 2))

plotPixels(cur_images,
           mask = masks_5[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD3", "CD8a", "CD68", "CD20", "CD44"),
           colour = list("CD3" = c("black", "yellow"),
                         "CD20" = c("black", "magenta"),
                         "CD8a" = c("black", "cyan"),
                         "CD68" = c("black", "green"),
                         "CD44" = c("black","blue")
                         ),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
           #save_plot = list(filename = "/Users/lasmey/Desktop/TNBC/Figures/Masks_comparison/TNBC_masks_immune_5.png", scale = 2))
```


### Single-cell visualization

Best performing segmentations 

```{r read and inspect steinbock}
sce_4 <- read_steinbock("/mnt/central_nas/tnbc_volume/TNBC/",intensities_folder = "intensities_4",return_as = "sce",regionprops_folder = NULL,graphs_folder = NULL)

sce_5 <- read_steinbock("/mnt/central_nas/tnbc_volume/TNBC/",intensities_folder = "intensities_5",return_as = "sce",regionprops_folder = NULL,graphs_folder = NULL)

dim(assay(sce_4,"counts"))
dim(assay(sce_5,"counts"))
```

```{r set-colnames}
colnames(sce_4) <- paste0(sce_4$sample_id, "_", sce_4$ObjectNumber)
colnames(sce_5) <- paste0(sce_5$sample_id, "_", sce_5$ObjectNumber)

rownames(sce_4) <- rowData(sce_4)$clean_target
rownames(sce_5) <- rowData(sce_5)$clean_target
```

```{r transform-counts, message=FALSE}
#Transform counts
assay(sce_4, "exprs") <- asinh(counts(sce_4)/1)
assay(sce_5, "exprs") <- asinh(counts(sce_5)/1)
```

```{r select-features}
rowData(sce_4)$use_channel <- !grepl("DNA|Histone", rowData(sce_4)$clean_target)
rowData(sce_5)$use_channel <- !grepl("DNA|Histone", rowData(sce_5)$clean_target)
```

```{r segmentation-heatmap, message=FALSE, warning = FALSE, fig.width=10, fig.height=12}
set.seed(22)
cur_cells <- sample(seq_len(ncol(sce_4)), 2000)

dittoHeatmap(sce_4[,cur_cells], genes = rownames(sce_4)[rowData(sce_4)$use_channel],
             assay = "exprs", cluster_cols = TRUE,scale = "none",
             heatmap.colors = viridis(100))

cur_cells_1 <- sample(seq_len(ncol(sce_5)), 2000)
dittoHeatmap(sce_5[,cur_cells_1], genes = rownames(sce_5)[rowData(sce_5)$use_channel],
             assay = "exprs", cluster_cols = TRUE,scale = "none",
             heatmap.colors = viridis(100))
```

Masks 4 were selected for downstream analysis.

# Session Info 

```{r}
sessionInfo()
```