---
title: "mIF_basoluminal_quant.Rmd"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
#Data containers
library(SpatialExperiment)
library(SingleCellExperiment)

#Data processing
library(scuttle)
library(scater)
library(scran)
library(imcRtools)

#Data visualization
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)

#Utilities
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(psych)
library(wesanderson)
```

# 1. Multiplex immunofluorescence quantification 

Read old sce for colors 

```{r}
sce_old <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/05_sce_TNBC_tumor_new.rds")
metadata(sce_old)$colors$CK5_7_score <- setNames(c("grey",wes_palette("Moonrise3")[1:3]), c("CK5/7-", "CK5/7+", "CK5+", "CK7+"))
```

## Read in data 

```{r}
measurement_IF <- read_csv("/mnt/central_nas/tnbc_volume/TNBC_IF/mIF/measurements.csv") 
```

```{r}
colnames(measurement_IF) <- str_replace(str_replace_all(str_replace(colnames(measurement_IF), "Cell:", ""), " ",""), "mean","")

# Create image, patient and cell identifiers

measurement_IF$Image <- str_split(measurement_IF$Image, "\\.", simplify = TRUE)[,1]

measurement_IF$PID <- paste0(str_split(measurement_IF$Image, "_", simplify = TRUE)[,3], ".", str_split(measurement_IF$Image, "_", simplify = TRUE)[,4])

cell_id <- foreach(i = unique(measurement_IF$Image),.combine = "c")%do%{ 
  cur_dat <- measurement_IF %>% filter(Image == i)
  cur_out <- paste0(cur_dat$Image, "_", seq(rownames(cur_dat)))
  return(cur_out)
  }

measurement_IF$cell_id <- cell_id
```

### Remove bad-quality patients 

```{r}
measurement_IF <- measurement_IF %>% filter(PID != "B14.30715")
```

### Add SOX9 - nuclear and cytoplasmic expression

```{r}
SOX9_IF <- read_csv("/mnt/central_nas/tnbc_volume/TNBC_IF/mIF/SOX9nuc_cyto.csv") 
```

```{r}
colnames(SOX9_IF) <- str_replace(str_replace_all(str_replace(colnames(SOX9_IF), ": ", "_"), " ",""), "mean","")

# Create image, patient and cell identifiers

SOX9_IF$Image <- str_split(SOX9_IF$Image, "\\.", simplify = TRUE)[,1]

SOX9_IF$PID <- paste0(str_split(SOX9_IF$Image, "_", simplify = TRUE)[,3], ".", str_split(SOX9_IF$Image, "_", simplify = TRUE)[,4])

cell_id <- foreach(i = unique(SOX9_IF$Image),.combine = "c")%do%{ 
  cur_dat <- SOX9_IF %>% filter(Image == i)
  cur_out <- paste0(cur_dat$Image, "_", seq(rownames(cur_dat)))
  return(cur_out)
  }

SOX9_IF$cell_id <- cell_id
```

```{r}
# Add info
measurement_IF$SOX9_nuc <- SOX9_IF$Nucleus_SOX9[match(measurement_IF$cell_id, SOX9_IF$cell_id)]
measurement_IF$SOX9_cyto <- SOX9_IF$Cytoplasm_SOX9[match(measurement_IF$cell_id, SOX9_IF$cell_id)]

## Replace H3K27AC cell expression with nuclear expression
measurement_IF$H3K27AC <- SOX9_IF$Nucleus_H3K27AC[match(measurement_IF$cell_id, SOX9_IF$cell_id)]
```


## Create SCE object 


```{r}
# Assays
assay <- measurement_IF %>% select(-Image, -ObjectID, -Name, -ROI, -CentroidXµm, -CentroidYµm, -Area, -PID, -cell_id)

## Remove ANXA8 due to NAs
assay <- assay %>% select(-ANXA8)
rownames(assay) <- measurement_IF$cell_id

# Coldata 
meta <- measurement_IF %>% select(Image, ObjectID, Name, ROI, CentroidXµm, CentroidYµm, Area, PID, cell_id)
rownames(meta) <- measurement_IF$cell_id
```

```{r}
# Create SingleCellExperiment object 
sce <- SingleCellExperiment(assays = list(counts = t(assay)),
                            colData = meta)
sce
```

## Tumor selection 

```{r}
dittoRidgePlot(sce, var = "panCK", assay = "counts", group.by = "Image")
```

### Manual panCK scoring 

Based on Qupath visual feedback and quality control

```{r}
sce$panCK_manual <- ifelse(assay(sce, "counts")["panCK",] > 1200, "panCK+", "panCK-")
sce$CK5_manual <- ifelse(assay(sce, "counts")["CK5",] > 3500, "CK5+", "CK5-") # Robust results between 3000-3500
sce$CK7_manual <- ifelse(assay(sce, "counts")["CK7",] > 625, "CK7+", "CK7-") # Robust results between 600-650
```

### GMM 

```{r}
set.seed(22)

## panCK GMM
GMM_panCK <- Mclust(assay(sce, "counts")["panCK",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_panCK$classification == 2 & GMM_panCK$uncertainty == 0, 
                                        "panCK+",
                                        "panCK-"), 
                     panCK_exprs = assay(sce, "counts")["panCK",],
                     uncertainty = GMM_panCK$uncertainty)


cutoff <- mean(c((GMM_df %>% filter(class_epi == "panCK+") %>% arrange(panCK_exprs) %>% pull(panCK_exprs))[1],
(GMM_df %>% filter(class_epi == "panCK-") %>% arrange(desc(panCK_exprs)) %>% pull(panCK_exprs))[1]))

#Visualize GMM
ggplot(GMM_df, aes(x = panCK_exprs, fill = class_epi))+
  geom_density()+
  geom_rug()+
  theme_classic(base_size = 15)+
  geom_vline(xintercept = cutoff)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  ggtitle(paste0("GMM cutoff = ", cutoff))

# Add to sce
sce$panCK_GMM <- GMM_df$class_epi
```

```{r}
set.seed(22)

## CK5 GMM
GMM_CK5 <- Mclust(assay(sce, "counts")["CK5",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK5$classification == 2 & GMM_CK5$uncertainty == 0, 
                                        "CK5+",
                                        "CK5-"), 
                     CK5_exprs = assay(sce, "counts")["CK5",],
                     uncertainty = GMM_CK5$uncertainty)


cutoff <- mean(c((GMM_df %>% filter(class_epi == "CK5+") %>% arrange(CK5_exprs) %>% pull(CK5_exprs))[1],
(GMM_df %>% filter(class_epi == "CK5-") %>% arrange(desc(CK5_exprs)) %>% pull(CK5_exprs))[1]))

#Visualize GMM
ggplot(GMM_df, aes(x = CK5_exprs, fill = class_epi))+
  geom_density()+
  geom_rug()+
  theme_classic(base_size = 15)+
  geom_vline(xintercept = cutoff)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  ggtitle(paste0("GMM cutoff = ", cutoff))

# Add to sce
sce$CK5_GMM <- GMM_df$class_epi
```

```{r}
set.seed(22)

## CK7 GMM
GMM_CK7 <- Mclust(assay(sce, "counts")["CK7",], G = 2)

GMM_df <- data.frame(class_epi = ifelse(GMM_CK7$classification == 2, 
                                        "CK7+",
                                        "CK7-"), 
                     CK7_exprs = assay(sce, "counts")["CK7",],
                     uncertainty = GMM_CK7$uncertainty)


cutoff <- mean(c((GMM_df %>% filter(class_epi == "CK7+") %>% arrange(CK7_exprs) %>% pull(CK7_exprs))[1],
(GMM_df %>% filter(class_epi == "CK7-") %>% arrange(desc(CK7_exprs)) %>% pull(CK7_exprs))[1]))

#Visualize GMM
ggplot(GMM_df, aes(x = CK7_exprs, fill = class_epi))+
  geom_density()+
  geom_rug()+
  theme_classic(base_size = 15)+
  geom_vline(xintercept = cutoff)+
  theme(axis.title.x = element_blank())+
  scale_fill_brewer(palette = "Set1")+
  ggtitle(paste0("GMM cutoff = ", cutoff))

# Add to sce
sce$CK7_GMM <- GMM_df$class_epi
```

## Intensity transformation

### Min max

0,1 scaling and 99.7 percentile winsorization per marker

```{r}
library(spatialEco)
mat_wins <- t(apply(counts(sce), MARGIN = 1, function(x) winsorize(x,p=c(0,0.997)))) # As original HIFI paper
assay(sce, "min_max") <- mat_wins
assay(sce, "min_max") <- assay(sce, "min_max") - rowMins(assay(sce, "min_max"))
assay(sce, "min_max") <- assay(sce, "min_max") / (rowMaxs(assay(sce, "min_max")) - rowMins(assay(sce, "min_max")))
```

### Dimensionality reduction 

```{r}
# Select channels
rowData(sce)$all_channel <- TRUE
rowData(sce)$dim_channel <- !rownames(sce) %in% c("DAPI", "CD133", "CD49f", "Wnt4", "CD8a", "HLADR")

rowData(sce)$cyto_channel <- rownames(sce) %in% c("panCK", "CK5", "CK7")

# Subsample per image
set.seed(22)
sub <- colData(sce) %>% as.data.frame() %>% group_by("Image") %>% slice_sample(prop = 0.1) %>% arrange("Image") %>% pull(cell_id)
sce_sub <- sce[,sce$cell_id %in% sub]

#Run UMAP/TSNE on cluster channels
sce_sub <- runUMAP(sce_sub[rowData(sce_sub)$cyto_channel,], n_neighbors = 100, pca = 50, exprs_values = "min_max",external_neighbors=TRUE, BPPARAM = MulticoreParam())

sce_sub <- runTSNE(sce_sub[rowData(sce_sub)$cyto_channel,], exprs_values = "min_max",external_neighbors=TRUE, BPPARAM = MulticoreParam())

## 1. Visualize UMAP and TSNE
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "Image")
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "panCK", by_exprs_values = "min_max", point_size = 0.5)
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "CK5", by_exprs_values = "min_max", point_size = 0.5)
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "CK7", by_exprs_values = "min_max", point_size = 0.5)
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "panCK_manual")
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "CK5_manual")
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "CK7_manual")

plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "Image")
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "panCK", by_exprs_values = "min_max", point_size = 0.5)
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "CK5", by_exprs_values = "min_max", point_size = 0.5)
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "CK7", by_exprs_values = "min_max", point_size = 0.5)
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "panCK_manual")
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "CK5_manual")
plotReducedDim(sce_sub, dimred = "TSNE", colour_by = "CK7_manual")

#plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "CD8a", by_exprs_values = "min_max", point_size = 0.5)
```

### Final selection 

```{r}
sce_tumor <- sce[, sce$panCK_manual == "panCK+"]
sce_tumor 
```

### Save objects 

```{r}
saveRDS(sce_tumor, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/mIF_sce_tumor.rds")
saveRDS(sce, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/mIF_sce_full.rds")

#sce_tumor <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/mIF_sce_tumor.rds")
```


# 2. Tumor analysis

```{r}
sce_full <- sce
sce <- sce_tumor
```

## Visualization 

```{r}
dittoRidgePlot(sce, var = "CK7", assay = "counts", group.by = "Image")
dittoRidgePlot(sce, var = "CK7", assay = "min_max", group.by = "Image")

dittoRidgePlot(sce, var = "CK5", assay = "counts", group.by = "Image")
dittoRidgePlot(sce, var = "CK5", assay = "min_max", group.by = "Image")

dittoRidgePlot(sce, var = "ALDH1A3", assay = "counts", group.by = "Image")
dittoRidgePlot(sce, var = "ALDH1A3", assay = "min_max", group.by = "Image")

dittoRidgePlot(sce, var = "SOX9", assay = "counts", group.by = "Image")
dittoRidgePlot(sce, var = "SOX9", assay = "min_max", group.by = "Image")
```

### Dimensionality reduction 

```{r}
# Select channels
rowData(sce)$all_channel <- TRUE
rowData(sce)$tumor_channel <- !rownames(sce) %in% c("DAPI", "panCK", "CD133", "CD49f", "Wnt4", "CD8a", "HLADR")

# Subsample per image
set.seed(22)
sub <- colData(sce) %>% as.data.frame() %>% group_by("Image") %>% slice_sample(prop = 0.1) %>% arrange("Image") %>% pull(cell_id)
sce_sub <- sce[,sce$cell_id %in% sub]

#Run UMAP/TSNE on cluster channels
sce_sub <- runUMAP(sce_sub[rowData(sce_sub)$tumor_channel,], n_neighbors = 50, pca = 30, exprs_values = "min_max",external_neighbors=TRUE, BPPARAM = MulticoreParam())

## 1. Visualize UMAP and TSNE
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "Image")
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "CK5", by_exprs_values = "min_max", point_size = 0.5)
plotReducedDim(sce_sub, dimred = "UMAP", colour_by = "CK7", by_exprs_values = "min_max", point_size = 0.5)
```


## CK5/CK7 manual scoring

Based on Qupath visual feedback and quality control

```{r}
sce$CK5_manual <- ifelse(assay(sce, "counts")["CK5",] > 3500, "CK5+", "CK5-")
sce$CK7_manual <- ifelse(assay(sce, "counts")["CK7",] > 625, "CK7+", "CK7-")
```

```{r}
## Combine CK5/7 GMM scores 
sce$CK5_7_score <- ifelse(sce$CK5_manual == "CK5+" & sce$CK7_manual == "CK7+", "CK5/7+", 
       ifelse(sce$CK5_manual == "CK5+" & sce$CK7_manual == "CK7-", "CK5+",
              ifelse(sce$CK5_manual == "CK5-" & sce$CK7_manual == "CK7+", "CK7+","CK5/7-")))
```

## Quality control 

### Correlation of CK5/CK7 per duplicate 

```{r}
# mIF experiment
cur_mIF <- colData(sce) %>% as.data.frame %>% group_by(Image, CK5_7_score, PID) %>% count() %>% group_by(Image) %>% mutate(fra = n/sum(n))

plot_out <- foreach(i = unique(sce$PID), .combine = "rbind")%do%{
  cur_cor <- cur_mIF %>% filter(PID == i) %>% pivot_wider(id_cols = CK5_7_score,names_from = Image, values_from = fra)
  cur_out <- cor(cur_cor[,2], cur_cor[,3], method = "pearson")
  out <- data.frame("category" = "PID", 
                    "PID" = i, 
                    "cor" = as.numeric(cur_out))
  return(out)
}

plot_out 

ggplot(plot_out, aes(x = category, y = cor))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), position=position_jitter(w=0.1))+
  guides(color = "none")+
  scale_color_manual(values = metadata(sce_old)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  scale_size_continuous()+
  ylab("Cor CK5/CK7 group fractions per PID")+
  ggtitle(paste("Spearman rho = ",round(median(plot_out$cor, na.rm = TRUE), digits = 4)))
```

### Correlation of marker expression 

```{r}
rowData(sce)$stem_channel <- rownames(sce) %in% c("CD24", "SOX9", "CD117", "CD44", "ALDH1A3", "H3K27AC", "YAP1", "ANXA8", "TNFAIP3",  "Notch1", "CD133") #CD133
cor_sce <- as.data.frame(t(assay(sce,"min_max")[rowData(sce)$stem_channel,]))

corrplot(cor(cor_sce,method = "pearson"), method="color", col=brewer.pal(n=8, name="RdBu"),  
         type="upper", order="hclust",hclust.method = "ward.D2", 
         addCoef.col = "black", addCoefasPercent = F,# Add coefficient of correlation
         tl.col="black", tl.srt=40, #Text label color and rotation
         # Combine with significance
         #p.mat = , sig.level = 0.01, insig = "blank", 
         # hide correlation coefficient on the principal diagonal
         diag=F
)
```

### Correlation of CK5/CK7 proportions 

Correlating CK5/CK7 proportions

```{r}
# IF experiment 1
cur_IF_1 <- read.csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/TNBC_IF_run_1celltype_fractions.csv",row.names = 1)

colnames(cur_IF_1) <- c("CK5/7+", "CK5+", "CK7+")
cur_IF_1 

# mIF experiment
cur_mIF <- colData(sce) %>% as.data.frame %>% filter(CK5_7_score %in% c("CK5/7+", "CK5+", "CK7+")) %>% group_by(PID, CK5_7_score) %>% count() %>% group_by(PID) %>% mutate(fra = n/sum(n))

cur_mIF <- cur_mIF %>% pivot_wider(id_cols = PID, names_from = "CK5_7_score", values_from = "fra") %>% column_to_rownames("PID")
cur_mIF


# TO DO
cur_IF_1 <- cur_IF_1[match(rownames(cur_mIF), rownames(cur_IF_1)),] #IF Basolum - Stem (Run 2)
cur_IF_1 <- cur_IF_1[,match(colnames(cur_mIF), colnames(cur_IF_1))] #IF Basolum - Stem (Run 2)

cur_mIF
cur_IF_1

paste0("BL:", cor(cur_IF_1$`CK5/7+`, cur_mIF$`CK5/7+`), " B:", cor(cur_IF_1$`CK5+`, cur_mIF$`CK5+`), " L:", cor(cur_IF_1$`CK7+`, cur_mIF$`CK7+`))

corrplot(cor(cur_IF_1, cur_mIF,method = "pearson"), method="color", col=brewer.pal(n=8, name="RdBu"),  
         type="upper", order="hclust",hclust.method = "ward.D2", 
         addCoef.col = "black", addCoefasPercent = F,
         tl.col="black", tl.srt=40,
         diag=TRUE)
```

## Heatmap visualization 

Based on the high correlation between images, we are selecting one image per patient for downstream visualization. 

Define channels 

```{r}
rowData(sce)$all_channel <- TRUE

rowData(sce)$keratin_channel <- rownames(sce) %in% c("CK19", "CK7", "CK5", "CK8_18", "CD14")
rowData(sce)$stem_channel <- rownames(sce) %in% c("CD24", "SOX9", "CD117", "CD44", "ALDH1A3", "H3K27AC", "YAP1", "ANXA8", "TNFAIP3",  "Notch1", "CD133")
```

### Keratins 

```{r}
# Keratins
# Do not really trust CK14, 8/18, 19 
cluster_mean_sce <- aggregateAcrossCells(sce, ids = sce$CK5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce)$keratin_channel)

Heatmap(t(assay(cluster_mean_sce, "min_max")), col = viridis(100))

Heatmap(scale(t(assay(cluster_mean_sce, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(cluster_mean_sce$ncells, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(cluster_mean_sce$CK5_7_score))
```
```{r}
sce_sub <- sce[,sce$Image %in% c("pyr_final_B08_42871_7_1", "pyr_final_B10_21451_30_2", "pyr_final_B08_33039_7_3", "pyr_final_B16_74059_16_2")]

# Stem markers
cluster_mean_sce <- aggregateAcrossCells(sce_sub, ids = sce_sub$CK5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce_sub)$keratin_channel)

Heatmap(t(assay(cluster_mean_sce, "min_max")), col = viridis(100))

Heatmap(scale(t(assay(cluster_mean_sce, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(cluster_mean_sce$ncells, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(cluster_mean_sce$CK5_7_score))
```

### Stem markers 

#### Across PID and Images 

```{r}
# Stem markers
cluster_mean_sce <- aggregateAcrossCells(sce, ids = sce$CK5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce)$stem_channel)

Heatmap(t(assay(cluster_mean_sce, "min_max")), col = viridis(100))

Heatmap(scale(t(assay(cluster_mean_sce, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(cluster_mean_sce$ncells, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(cluster_mean_sce$CK5_7_score))
```

#### Per PID or Image 

```{r}
heatmap_list <- list()

for(i in unique(sce$PID)){

  sce_PID <- sce[, sce$PID == i]

  # Stem markers
  cluster_mean_sce <- aggregateAcrossCells(sce_PID, ids = sce_PID$CK5_7_score,
                                   statistics = "mean",
                                   use.assay.type = "min_max",
                                   subset.row = rowData(sce_PID)$stem_channel)
  
  #h_raw <- Heatmap(t(assay(cluster_mean_sce, "min_max")), name = paste0(i, "_raw"))
  #h_raw <- draw(h_raw)
  #heatmap_list[paste0(i, "_raw")] <- h_raw 
  
  h <- Heatmap(scale(t(assay(cluster_mean_sce, "min_max"))), name = paste0(i))+
    rowAnnotation(n = anno_barplot(cluster_mean_sce$ncells, gp = gpar(fill = "#440154FF")))+
    rowAnnotation(anno = anno_text(cluster_mean_sce$CK5_7_score))
  h <- draw(h)
  
  heatmap_list[i] <- h
}

```


```{r}
heatmap_list <- list()

for(i in unique(sce$Image)){

  sce_PID <- sce[, sce$Image == i]

  # Stem markers
  cluster_mean_sce <- aggregateAcrossCells(sce_PID, ids = sce_PID$CK5_7_score,
                                   statistics = "mean",
                                   use.assay.type = "min_max",
                                   subset.row = rowData(sce_PID)$stem_channel)
  
  #h_raw <- Heatmap(t(assay(cluster_mean_sce, "min_max")), name = paste0(i, "_raw"))
  #h_raw <- draw(h_raw)
  #heatmap_list[paste0(i, "_raw")] <- h_raw 
  
  h <- Heatmap(scale(t(assay(cluster_mean_sce, "min_max"))), name = paste0(i))+
    rowAnnotation(n = anno_barplot(cluster_mean_sce$ncells, gp = gpar(fill = "#440154FF")))+
    rowAnnotation(anno = anno_text(cluster_mean_sce$CK5_7_score))
  h <- draw(h)
  
  heatmap_list[i] <- h
}

```



#### For one image per PID 

```{r}
sce_sub <- sce[,sce$Image %in% c("pyr_final_B08_42871_7_1", "pyr_final_B10_21451_30_2", "pyr_final_B08_33039_7_3", "pyr_final_B16_74059_16_2")]

# Stem markers
cluster_mean_sce <- aggregateAcrossCells(sce_sub, ids = sce_sub$CK5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce_sub)$stem_channel)


Heatmap(t(assay(cluster_mean_sce, "min_max")), col = viridis(100))
Heatmap(t(assay(SOX9_mean_sce, "min_max")), col = viridis(100))

Heatmap(scale(t(assay(cluster_mean_sce, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(cluster_mean_sce$ncells, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(cluster_mean_sce$CK5_7_score))
```

#### Average per PID 


```{r}
# Stem markers
cluster_PID_mean_sce <- aggregateAcrossCells(sce, ids = DataFrame(sce$CK5_7_score,sce$PID),
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce)$stem_channel)

cluster_mean_sce <- aggregateAcrossCells(cluster_PID_mean_sce, ids = cluster_PID_mean_sce$CK5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(cluster_PID_mean_sce)$stem_channel)

Heatmap(t(assay(cluster_mean_sce, "min_max")), col = viridis(100))

Heatmap(scale(t(assay(cluster_mean_sce, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(anno = anno_text(cluster_mean_sce$CK5_7_score))
```




#### Annotation heatmap 

All 

```{r}
cur_in <- colData(sce_sub) %>% as.data.frame %>% group_by(PID, CK5_7_score) %>% count() %>% group_by(PID) %>% mutate(fra = n/sum(n), total_n = sum(n))

cur_CK <- cur_in %>% pivot_wider(id_cols = "PID", names_from = "CK5_7_score",values_from = "fra") %>% column_to_rownames("PID")

# Option: order by BL cell fraction
row_sorted_fra <- cur_in %>% filter(CK5_7_score == "CK5/7+") %>% arrange(desc(fra)) %>% pull(PID)

# Define annotation dataframe
cur_CK <- cur_CK[row_sorted_fra,]
cur_CK <- cur_CK[,c("CK5/7+", "CK5+", "CK7+", "CK5/7-")]

anno <- cur_in %>% pivot_wider(id_cols = "PID", names_from = "CK5_7_score", values_from = "total_n") %>% column_to_rownames("PID")

anno <- anno[match(rownames(cur_CK), rownames(anno)),]
anno <- anno[,match(colnames(cur_CK), colnames(anno))]

# Color
library(wesanderson)

cur_color_CK <- c("#85D4E3", "#F4B5BD", "#9C964A", "grey")

# Plot annotation dataframe with ComplexHeatmap
fraction_anno <- HeatmapAnnotation(PID = anno_text(rownames(anno)), 
                                   CK_fractions = anno_barplot(cur_CK, gp=gpar(fill=cur_color_CK),
                                                               bar_width = 1, height = unit(25,"cm"), width = unit(11,"cm"),
                                                               show_row_names = TRUE, row_names_side = "left"),
                                   total_n = anno_barplot(rowSums(anno), border = TRUE, gp = gpar(fill = "#440154FF"), width = unit(3,"cm")),
                                   gap = unit(1,"mm"),
                                   which = "row")
h_1 <- fraction_anno + NULL

lgd_1 <- Legend(title = "CK5_7_score", at = colnames(cur_CK), 
              legend_gp = gpar(fill = cur_color_CK))
             
# Plot
draw(h_1,annotation_legend_list = list(lgd_1))
```
TO DO for subset 

## Heterogeneity analysis 

Coefficient of variation for expression per gene per category

```{r}
expression_data <- assay(sce, "min_max")[rowData(sce)$use_channel,]
#expression_data <- assay(sce, "min_max")[rowData(sce)$stem_channel,]
#expression_data <- assay(sce, "min_max")[rowData(sce)$keratin_channel,]

# Define a function to calculate CV
calculate_cv <- function(x) {
  mean_x <- mean(x)
  sd_x <- sd(x)
  if (mean_x == 0) return(NA)  # Handle zero mean case
  return(sd_x / mean_x)
}

# Create a dataframe to store CV values for each category
cv_results <- data.frame(Category = character(), CV = numeric(), stringsAsFactors = FALSE)

# Calculate CV for each category
cv_out <- foreach(cat = unique(sce$CK5_7_score), .combine = "rbind")%do%{
  expression_values <- expression_data[, which(sce$CK5_7_score == cat)]
  # Calculate CV for each gene (row) and take the mean
  cv_per_gene <- apply(expression_values, 1, calculate_cv)
  return(data.frame(cv = cv_per_gene, cat = cat))
  }

ggplot(cv_out, aes(x = cat, y = cv, fill = cat))+
  geom_boxplot(outlier.colour = NA)+
  theme_classic()+
  scale_fill_manual(values = metadata(sce_old)$colors$CK5_7_score)
```



# 3. SOX9 nuclear vs. cytoplasmic staining 

## SOX9 manual scoring 

```{r}
sce$SOX9_manual <- ifelse(assay(sce, "counts")["SOX9",] > 150, "SOX9+", "SOX9-")

# Subset SCE
sce_SOX9 <- sce[,sce$SOX9_manual == "SOX9+"]
sce_SOX9

rowData(sce_SOX9)$SOX9_channel <- rownames(sce) %in% c("SOX9_nuc", "SOX9_cyto")
```

## Heatmap visualization 

```{r}
# Stem markers
cluster_mean_sce_SOX9 <- aggregateAcrossCells(sce_SOX9, ids = sce_SOX9$CK5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce_SOX9)$SOX9_channel)


Heatmap(t(assay(cluster_mean_sce_SOX9, "min_max")), col = viridis(100))

Heatmap(scale(t(assay(cluster_mean_sce_SOX9, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(cluster_mean_sce_SOX9$ncells, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(cluster_mean_sce_SOX9$CK5_7_score))
```


```{r}
sce_SOX9_sub <- sce_SOX9[,sce_SOX9$Image %in% c("pyr_final_B08_42871_7_1", "pyr_final_B10_21451_30_2", "pyr_final_B08_33039_7_3", "pyr_final_B16_74059_16_2")]

# Stem markers
cluster_mean_sce_SOX9 <- aggregateAcrossCells(sce_SOX9_sub, ids = sce_SOX9_sub$CK5_7_score,
                                 statistics = "mean",
                                 use.assay.type = "min_max",
                                 subset.row = rowData(sce_SOX9_sub)$SOX9_channel)


Heatmap(t(assay(cluster_mean_sce_SOX9, "min_max")), col = viridis(100))

Heatmap(scale(t(assay(cluster_mean_sce_SOX9, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(cluster_mean_sce_SOX9$ncells, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(cluster_mean_sce_SOX9$CK5_7_score))
```





```{r}
# Filter for SOX9-positive cells and tumor cells 
# Create ratios and scores
sce_SOX9$SOX9_nuc <- assay(sce_SOX9, "counts")["SOX9_nuc",]
sce_SOX9$SOX9_cyto <- assay(sce_SOX9, "counts")["SOX9_cyto",]

SOX_tumor <- colData(sce_SOX9) %>% as.data.frame %>% mutate(SOX9_nuc_cyto_ratio = (SOX9_nuc) / (SOX9_cyto)) %>% mutate(SOX9_nuc_cyto_score = ifelse(SOX9_nuc_cyto_ratio >= 1, "nuc_high", "cyto_high"), SOX9_nuc_cyto_score_fine = ifelse(SOX9_nuc_cyto_ratio > 1.1, "nuc_high", ifelse(SOX9_nuc_cyto_ratio < 0.9, "cyto_high", "balanced")))
```


```{r}
plot_score <- SOX_tumor %>% group_by(CK5_7_score) %>% count(SOX9_nuc_cyto_score) %>% mutate(fra = n/sum(n))
plot_score$CK5_7_score <- factor(plot_score$CK5_7_score, levels = c("CK5/7-", "CK7+","CK5+", "CK5/7+"))

ggplot(plot_score, aes(x = CK5_7_score, y = fra, fill = SOX9_nuc_cyto_score))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_brewer(palette = "Reds")+
  theme_classic()

plot_score <- SOX_tumor %>% group_by(CK5_7_score) %>% count(SOX9_nuc_cyto_score_fine) %>% mutate(fra = n/sum(n))
plot_score$CK5_7_score <- factor(plot_score$CK5_7_score, levels = c("CK5/7-", "CK7+","CK5+", "CK5/7+"))

ggplot(plot_score, aes(x = CK5_7_score, y = fra, fill = SOX9_nuc_cyto_score_fine))+
  geom_bar(stat = "identity", color = "black")+
  scale_fill_brewer(palette = "Reds")+
  theme_classic()

plot_score_mean <- SOX_tumor %>% group_by(CK5_7_score) %>% summarise(mean = mean(SOX9_nuc_cyto_ratio))
ht <- plot_score_mean %>% pivot_wider(names_from = CK5_7_score, values_from = mean)

Heatmap(t(scale(t(ht))))
Heatmap(ht, col = rev(viridis(100)))
```



# Save plots 

```{r}
pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/mIF_Basoluminal/mIF_CK57score_Heatmap_Keratin.pdf", height = 5, width = 8, onefile = FALSE)
Heatmap(scale(t(assay(cluster_mean_sce, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(cluster_mean_sce$ncells, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(cluster_mean_sce$CK5_7_score))
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/mIF_Basoluminal/mIF_CK57score_Heatmap_Stem.pdf", height = 5, width = 9, onefile = FALSE)
Heatmap(scale(t(assay(cluster_mean_sce, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(cluster_mean_sce$ncells, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(cluster_mean_sce$CK5_7_score))
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/mIF_Basoluminal/mIF_CK57score_Heatmap_SOX9_compartment.pdf", height = 5, width = 5, onefile = FALSE)
Heatmap(scale(t(assay(cluster_mean_sce_SOX9, "min_max"))), name = "Scaled mean expression")+
  rowAnnotation(n = anno_barplot(cluster_mean_sce_SOX9$ncells, gp = gpar(fill = "#440154FF")))+
  rowAnnotation(anno = anno_text(cluster_mean_sce_SOX9$CK5_7_score))
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/mIF_Basoluminal/mIF_CK57score_Correlation_Dup.pdf", height = 5, width = 4, onefile = FALSE)
ggplot(plot_out, aes(x = category, y = cor))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter(aes(color = PID), position=position_jitter(w=0.1), size = 4)+
  scale_color_manual(values = metadata(sce_old)$colors$PID)+
  theme_classic(base_size = 12)+
  theme(axis.title.x = element_blank())+
  ylab("Cor CK5/CK7 group fractions per PID")+
  ggtitle(paste("Spearman rho = ",round(median(plot_out$cor, na.rm = TRUE), digits = 4)))
dev.off()

pdf("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/01_figures/mIF_Basoluminal/mIF_CK57score_Correlation_Exp.pdf", height = 5, width = 5, onefile = FALSE)
corrplot(cor(cur_IF_1, cur_mIF,method = "pearson"), method="color", col=brewer.pal(n=8, name="RdBu"),  
         type="upper", order="hclust",hclust.method = "ward.D2", 
         addCoef.col = "black", addCoefasPercent = F,
         tl.col="black", tl.srt=40,
         diag=TRUE)
dev.off()

```

