---
title: "TNBC_PDL1_IHC_analysis"
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Meyer et al., 2025 

**DISCLAIMER**: Pipeline scripts contain all code used for data presented in the publication but also small exploratory data analysis parts. For reproducing the figures, please refer to the figure-related scripts and cleaned data objects provided on **Zenodo** (<https://zenodo.org/records/>). File paths have to be adjusted by the user. 

Code related to following cleaned **Zenodo** objects: PDL1_IHC.csv

# Validation

Data analysis scripts for data analysis in the validation cohorts. 

# Script: PDL1 IHC analysis

Here, we are analyzing PDL1 IHC data from patients in the discovery cohort to validate immunotherapy eligibility.

# Software requirements

```{r lib, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scran)
library(imcRtools)
library(cytomapper)
library(dittoSeq)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(openxlsx)
library(stringr)
library(patchwork)
library(parallel)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(tiff)
library(ggrepel)
library(EBImage)
library(viridis)
library(mclust)
library(Rphenograph)
library(igraph)
library(RColorBrewer)
library(foreach)
library(clValid)
library(bluster)
library(psych)
library(corrplot)
library(readxl)
```

# Read and analyze data

```{r}
##Load sce
sce <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/07_sce_combined.rds")
```

```{r}
# Patient selection
inflamed_candidates <- read_csv("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/TNBC_PDL1_IHC.csv") 
inflamed_candidates$spatial_CD8_pheno <- sce$spatial_CD8_pheno[match(inflamed_candidates$PID, sce$PID)]
inflamed_candidates$pM_simple <- sce$pM_simple[match(inflamed_candidates$PID, sce$PID)]

# PDL1 IHC results
PDL1_results <- read_excel("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/03_data/Lasse_TNBC_CPS_20240723GU.xlsx", col_names = TRUE) 
PDL1_results$PID <- str_replace(str_split(PDL1_results$ID, "_", simplify = TRUE)[,1], "-", ".")
PDL1_results$CPS_score <- round(PDL1_results$`CPS (continuous variable)`,0)
```

```{r}
# Add results
inflamed_candidates$CPS_score <- PDL1_results$CPS_score[match(inflamed_candidates$PID, PDL1_results$PID)]
inflamed_candidates$CPS_results <- PDL1_results$`CPS(Positive or Negative)`[match(inflamed_candidates$PID, PDL1_results$PID)]

# Add cut-off cases 
inflamed_candidates$CPS_score_fine <- ifelse(inflamed_candidates$CPS_score >= 10, "CPS >= 10", ifelse(inflamed_candidates$CPS_score < 5, "CPS < 5", "CPS 5-10"))
```

```{r}
write_csv(inflamed_candidates, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/PDL1_IHC.csv") 
```

```{r}
# Spatial Pheno
out_PDL1 <- inflamed_candidates %>% group_by(CPS_results, spatial_CD8_pheno) %>% dplyr::count() %>% group_by(spatial_CD8_pheno) %>% mutate(fra = n/sum(n))

fisher_out <- fisher.test(xtabs(n ~ spatial_CD8_pheno + CPS_results, data = out_PDL1))
fisher_out

ggplot(out_PDL1, aes(x = spatial_CD8_pheno, y = fra, fill = CPS_results))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  ggtitle(paste0("p (Fisher) = ", round(fisher_out$p.value, 4)))
```

```{r}
# Spatial Pheno Merg
out_PDL1 <- inflamed_candidates %>% group_by(CPS_results, spatial_CD8_pheno_merg) %>% dplyr::count() %>% group_by(spatial_CD8_pheno_merg) %>% mutate(fra = n/sum(n))

fisher_out <- fisher.test(xtabs(n ~ spatial_CD8_pheno_merg + CPS_results, data = out_PDL1))
fisher_out

ggplot(out_PDL1, aes(x = spatial_CD8_pheno_merg, y = fra, fill = CPS_results))+
  geom_bar(stat = "identity", color = "black")+
  theme_classic()+
  scale_fill_manual(values = c("#ECD2FC", "#8019C0"))+
  ggtitle(paste0("p (Fisher) = ", round(fisher_out$p.value, 4)))
```

```{r}
inflamed_candidates$PID <- factor(inflamed_candidates$PID, levels = inflamed_candidates %>% arrange(desc(CPS_score),desc(spatial_CD8_pheno)) %>% pull(PID))

ggplot(inflamed_candidates, aes(x = PID, y = CPS_score, fill = CPS_score_fine))+
  geom_bar(stat = "identity", color = "black")+
  geom_rug(aes(color = spatial_CD8_pheno), sides = "b", linewidth = 3)+
  geom_hline(yintercept = 10, linetype = "dashed")+
  theme_classic()+
  scale_color_manual(values = metadata(sce)$colors$spatial_CD8_pheno)+
  scale_fill_manual(values = c("#ECD2FC","#8019C0", "#BF8CDF"))+
  theme(axis.text.x = element_text(angle = 90))
```

# Session Info 

```{r}
sessionInfo()
```